### 2024-06-10 writing/manuscript.Rmd
first push

---
title: "Predicting how climate change and short-term weather will affect terrestrial mammals' movement and habitat selection"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Kirk Safford
    institute: bcp
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - bcp: BC Parks
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammal-speeds.bib'
# csl: 'csl.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

<!-- place table of contents on a new page -->
\newpage

\singlespacing

\tableofcontents

<!-- start main body on a new page -->
\newpage

\doublespacing
*
```{r setup, include=FALSE}
# set chunk defaults: don't print code, images text width and centered
knitr::opts_chunk$set(echo = FALSE, out.width = '\\textwidth',
                      fig.align = 'center', cache = TRUE)
```

```{r packages, warning=FALSE, message=FALSE}
library('dplyr') # for data wrangling
library('purrr') # for data wrangling
library('tidyr') # for data wrangling
library('ctmm')  # for movement models
```

# Abstract

# To do

Intro:

* add references:
  * McNabb 1960s
  * Hirt 2022
  * Moose speed and temp (just @alston_temperature_2020 or are there others?)
  * Lots of work on fish
  * Brown temp constant for all biological processes
  * Andean hummingbird and temp
* address any questions answered in the discussion

Methods:

* add headers

Write abstract

\clearpage

# Introduction

The rate at which chemical reactions occur within an organism [metabolism, see @brown_toward_2004] depends strongly on the temperature reactions occur at. However, organisms experience different ranges of external temperatures and, consequently, exhibit specific thermal optima. These depend on many factors, including body size [@taylor_energetics_1982; @reiss_scaling_1988; @fjelldal_nightly_2021], physiological adaptations [@schmidt_interplay_2020; @mohr_cellular_2020], and behavior [@arechavala-lopez_common_2019]. In the case of heterothermic organisms, low temperatures increase energetic costs [@melin_effect_2023; @ref; @fish; @andean_hummingbird], while high temperatures increase the risk of overheating [@ref; @cheetas; @alston_temperature_2020].

At extremely cold temperatures, the metabolic costs of movement and homeothermy increase energetic expenditures, which can constrain an animal's ability to move, lowering encounter rates with food [@hou_cold_2020], fitness [@ref], and, ultimately, chance of survival [@berger_climate_2018]. Mammals reduce the adverse effects of cold through a combination of both behavioral adaptations [e.g., searching for shelter: @ref; reducing movement rates: @ref] and physiological adaptations [e.g., fur: @ref; circulation: @ref]. At extremely hot temperatures, excessively high body temperatures ***XXXX*** [@alston_temperature_2020; @refs].

Evolutional constraints limit mammals' ability to respond temperature extremes, which results in phylogenetically-constrained temperature optima for different species and populations [@caribou-ref]. Thus, while animals are adapted to fluctuations around average conditions [@ref], strong and consistent temperature anomalies can decrease an animal's energetic stability [@ref], reproductive success [@schmidt_interplay_2020], and ability to move [@ref]. Recent changes in climate have resulted in widespread warming, milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@ref_2021_heat_dome]. These changes are expected to affect mammals' fitness, movement behavior, and habitat selection as they cope with growing heat stress [@weststrate_how_2024].

This paper provides a BC-focused analysis on the effects of proximate weather and long-term climate on mammal movement. Using more than a decade of telemetry data and proximal 3-hour weather data, we estimate how animals altered their movement frequency, movement speed, and habitat selection in response to air temperature. We then pair these responses with climate change projections to forecast changes mammals movement in the current century under different climate change scenarios [@riahi_shared_2017]. The analyses include data from six species (*Canis lupus*, *Cervus elaphus*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, *Ursus arctos horribilis*; see Table \@ref(tab:data-table)) across a large spatial range of British Columbia (Fig. \@ref(fig:bc-map)). We discuss the consequences of changes in mammalian movement in relationship to energetics, food encounter rates, and the incidences of human-wildlife.

```{r bc-map, fig.cap="GPS telemetry data and autocorrelated kernel density utilization distributions for the six species included in this study."}
knitr::include_graphics('../figures/tels-and-uds-map.png')
```

```{r data-table}
readRDS('../models/movement-models-2024-03-18.rds') %>%
  transmute(Dataset = if_else(grepl('Puma', dataset_name),
                                'Puma concolor', dataset_name) %>%
              stringr::str_replace_all('_', ' ') %>%
              stringr::str_replace_all(' boreal', ' (boreal)') %>%
              stringr::str_replace_all(' southern mountain',
                                       ' (southern mountain)'),
            animal,
            tel = purrr::map(tel, data.frame)) %>%
  unnest(tel) %>%
  group_by(animal) %>%
  mutate(dt = timestamp - lag(timestamp)) %>%
  group_by(Dataset) %>%
  summarize(Start = min(as.Date(timestamp)),
            End = max(as.Date(timestamp)),
            dt = as.numeric(median(dt, na.rm = TRUE))) %>%
  mutate(dt = round('hours' %#% dt, 2)) %>%
  rename(`Median sampling interval (hours)` = dt) %>%
  knitr::kable(format = 'pipe', caption = 'Start and end of the GPS telemetries along with the median sampling interval for each of the six species included in this study.') %>%
  kableExtra::column_spec(1, italic = TRUE) %>%
  kableExtra::kable_styling(font_size = 10)
```

# Methods

## Data acquisition

### GPS telemetry data

Animal telemetry datasets were obtained from multiple sources. Elk tracking data were obtained from Movebank [@kays_movebank_2022; study name: Elk in southwestern Alberta, see @study_ref], while the boreal caribou and wolf data were obtained from a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data/](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data/)). ***All other tracking data were obtained from private collaborators*** [@refs_to_studies]. Outlier GPS locations were removed following diagnostic analyses of the location of each point as well as straight-line displacement and turning angle between consecutive points. Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$), especially if points before or after the potential outlier indicated stationary behavior. The script used to clean the data and all custom functions used are available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-speeds](https://github.com/QuantitativeEcologyLab/bc-mammals-speeds)).

### Historical temperature data

Rasters of three-hour reanalyzed air temperature data were downloaded from the ERA5 dataset [@copernicus_climate_change_service_era5_2019] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)). Proximal weather conditions were then estimated for each location by extracting the values from the corresponding cell from the temporally nearest raster using the `extract()` function from the `terra` package for `R` [version 1.7-46, @hijmans_terra_2023].

## Data modeling

### Estimating mammals' instantaneous speeds

Animal movement models were fit using the `ctmm` package [version 1.1.0, @fleming_ctmm_2021] for `R` [@r_core_team_r_2023]. Mammals' speeds were then estimated using the `ctmm::speeds()` function on all models with speed estimates [@noonan_scale-insensitive_2019]. Most telemetries (416 of 433) had sufficiently frequent locations to provide appropriate speed estimates. All 17 telemetries with insufficiently fine sampling were for caribou (15 boreal and 2 souther mountain).

Since `ctmm`'s movement models allow for stochastic speeds but assume constant movement, speeds were set to zero if they were not significantly different from zero ($\alpha = 0.05$), under the assumption of a $\chi$-distributed speeds. ***NOT SURE IF THIS IS CORRECT*** The function used is available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-speeds/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-speeds/blob/main/functions/detrend_speeds.R)). To avoid artifacts due to excessively small, non-zero speeds, estimated speeds were then clustered into two groups using a $k$-means algorithm with two centers (see Appendix A).

### Effects of temperature on mammals' movement

We estimated the effects of hourly temperature on mammals' movement using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_pedersen_hierarchical_2019] via `mgcv` package for `R`. The first HGAM had a binomial family of distribution and estimated the probability of a mammal moving as a function of time of day, day of year, and temperature. Since there was substantial variation in the direction and wiggliness of each species' trends, the effects were modeled independently `by` smooths for each species [see model I in Figure 4 of @pedersen_hierarchical_2019]. The model accounted for the cyclicity of time of day and day of year using cyclic cubic splines, which provide continuity in the smooth term by ensuring the function and its slope (first derivative) and curvature (second derivative) are continuous between consecutive cycles. The second HGAM had the same structure, but it used a Gamma family of distributions to estimate the speed of an animal given that it was moving. Together, the two HGAMs inform us on an animal's long-term speed (i.e., distance traveled per unit time), since it is the product of probability that the animal is moving and the speed it moves at, given that it is moving.

### Effects of temperature on mammals' habitat selection

An animal's habitat selection is proportional to the number of times one can observe it in an environment. Consequently, one can view the spatial distributions of an animal in space as a poisson process whose mean depends on the amount of space, $\Delta u$, and time, $\Delta t$, the observation takes place over, which we can write as $\lambda(\Delta u, \Delta t)$. If we shrink both $\Delta u$ and $\Delta t$ to a single point in space and a single moment in time, we now have $\lambda(u, t) \le 1$, since we cannot observe more than one animal in a given point in time and spce $(u, t)$. Additional theoretical background is provided by @aarts_estimating_2008 and @alson_mitingting_2022 and in Appendix B.

Under the asumption of a Poisson family of distributions, we estimated the effects of temperature on each species' habitat selection by fitting a Resource Selection Function (RSF) via an HGAM using the `mgcv` package. ***here***

## Predicting changes in animal movement during the current century

### Climate change projections

Rasters of monthly average temperature in BC during the next 80 years were obtained via the `climatenaR` package [version XXX; @burnett_climatenar_2023] for `R`. Since the climate projections of monthly average temperatures did not provide a measure of temperature variance for each month, we estimated within-month variance using a Generalized Additive Model with a Gamma family of distributions fit via the `mgcv` package for `R` [@wood_generalized_2017]. Additional details are provided in Appendix C.

# Results

### Effects of temperature on mammals' movement

### Effects of temperature on mammals' habitat selection

**should predict for the area the animals are in only**

# Discussion

<!-- - how results answer central question -->

### Effects of temperature on mammals' movement

### Effects of temperature on mammals' habitat selection

<!-- - what it tells us about the broader field -->

* energetics
* encounter rates
* HWI

<!-- - address any questions presented in the discussion -->

\clearpage <!-- to prevent figures from floating to the references -->

\newpage <!-- place references on a separate page -->

# References

\hangparas{1em}{1} <!-- paragraphs hanging by width equal to font height -->

### 2024-06-10 analysis/00-compile-goat-tracking-data.R

@@ -1,0 +1,42 @@
+library('dplyr') # for data wrangling
+library('purrr') # for functional programming
+library('tidyr') # for data wangling
+library('sf')    # for spatial features
+
+# check layers
+st_layers('data/tracking-data/oreamnos-americanus/GPS_Collar30548_20231005121804.gpx')
+#' use `track_points` layer
+
+# import all gpx files
+d <- tibble(
+  file = list.files('data/tracking-data/oreamnos-americanus'),
+  track = map(file, function(fn) {
+    st_read(paste0('data/tracking-data/oreamnos-americanus/', fn),
+            layer = 'track_points', quiet = TRUE) %>%
+      # add animal identifier with ctmm syntax
+      mutate(individual.local.identifier = substr(fn, nchar('GPS_CollarX'),
+                                                  nchar('GPS_CollarXXXXX'))) %>%
+      bind_cols(., st_coordinates(.)) %>% # add coordinates
+      st_drop_geometry() %>% # drop sf geometry column
+      rename(location.long = X, # rename using ctmm syntax
+             location.lat = Y,
+             timestamp = time)
+  })) %>%
+  unnest(track) # turn into a single large tibble
+
+# check which columns have more than one unique (non-NA) value
+d %>%
+  mutate(across(everything(), as.character)) %>%
+  pivot_longer(-c()) %>%
+  group_by(name) %>%
+  summarize(n = n_distinct(value, na.rm = TRUE)) %>%
+  mutate(useful = n > 1) %>%
+  arrange(desc(n)) %>%
+  print(n = 28)
+
+# drop non-NA columns with more than one unique value
+d <- d %>%
+  select(individual.local.identifier, location.long, location.lat,
+         timestamp, fix, hdop, pdop)
+
+saveRDS(d, 'data/tracking-data/oreamnos-americanus.rds')

### 2024-06-10 analysis/00-crop-resource-rasters.R

@@ -1,0 +1,87 @@
+library('dplyr') # for data wrangling
+library('tidyr') # for data wrangling
+library('purrr') # for functional programming
+library('sf')    # for working with spatial data
+library('sp')    # for working with spatial data
+library('terra') # for raster data (faster than raster)
+library('canadianmaps') # to download a shapefile of BC
+
+# import a shapefile of the locations
+locs <-
+  readRDS('data/tracking-data/all-tracking-data-cleaned-2024-02-22-13-49.rds') %>%
+  unnest(tel) %>%
+  filter(! outlier) %>%
+  select(location.long, location.lat, animal) %>%
+  st_as_sf(coords = c('location.long', 'location.lat'))
+
+locs_ext <- locs %>%
+  st_bbox() %>%
+  st_as_sfc() %>%
+  st_as_sf()
+
+# original rasters from http://www.earthenv.org/landcover
+dem <- rast('data/resource-rasters/bc-buffered-dem-z6.tif')
+plot(dem)
+plot(locs_ext, add = TRUE, lwd = 3)
+
+# forest raster
+rast('data/resource-rasters/uncropped-rasters/consensus_full_class_1.tif') %>%
+  crop(dem) %>%
+  writeRaster('data/resource-rasters/forest.tif')
+plot(rast('data/resource-rasters/forest.tif'))
+plot(locs_ext, add = TRUE, lwd = 3)
+
+# raster of open water
+layout(matrix(1:8, ncol = 2, byrow = TRUE))
+w <- rast('data/resource-rasters/uncropped-rasters/consensus_full_class_12.tif') %>%
+  crop(dem)
+plot(w, main = 'Percent water')
+plot(locs_ext, add = TRUE, lwd = 3)
+hist(w, breaks = 100, main = 'Percent water')
+
+w01 <- ceiling(w / 100) # convert anything above 0 to a 1
+plot(w01, main = 'Yes/No water')
+hist(w01, main = 'Yes/No water')
+
+w1 <- classify(w01, cbind(0, NA))
+plot(w1, main = 'Water only', col = 'blue')
+unique(w1, main = 'Water only')
+
+dist <- distance(w1)
+plot(dist, main = 'Distance from nearest water')
+hist(dist, main = 'Distance from nearest water')
+writeRaster(dist, 'data/resource-rasters/distance-from-water.tif')
+
[…]

### 2024-06-10 analysis/00-download-bc-dem.R

@@ -1,0 +1,71 @@
+library('sf')           # for spatial data
+library('dplyr')        # for data wrangling
+library('canadianmaps') # for shapefile of BC
+library('elevatr')      # for downloading DEM (using version 0.99.0)
+library('terra')        # for working with rasters
+source('data/bc-shapefile.R')
+
+ZOOM <- 3 #' resolution level for the DEM (see `?get_elev_raster`)
+
+bc_dem <-
+  get_elev_raster(locations = bc_unproj, z = ZOOM, clip = 'locations') %>%
+  rast() #' `elevatr` v 0.99.0 still returs a `RasterLayer` object
+
+plot(bc_dem)
+plot(bc_unproj, add = TRUE)
+
+if(FALSE) {
+  # highest point in BC is mount fairweather at ~ 4e3 m
+  #' `ZOOM > 6` results in excessively high elevation values
+  max(values(bc_dem), na.rm = TRUE)
+  get_elev_raster(locations = bc_unproj, z = 7, clip = 'locations') %>%
+    rast() %>% #' `elevatr` v 0.99.0 still returs a `RasterLayer` object
+    values() %>%
+    max(na.rm = TRUE)
+  
+  # res is close enough
+  res(bc_dem)
+  res(rast('data/resource-rasters/bc-forest.tif'))
+  res(rast('data/resource-rasters/bc-distance-from-water.tif'))
+}
+
+terra::writeRaster(bc_dem, paste0('data/bc-dem-z', ZOOM, '.tif'))
+
+# download dem of buffered BC
+bc_unproj_buff <- bc_unproj %>%
+  st_bbox() %>%
+  st_as_sfc() %>%
+  st_as_sf() %>%
+  st_buffer(300e3) # contains 99.9995% UDs
+
+# check ranges
+plot(bc_unproj_buff)
+plot(bc_unproj, add = TRUE)
+if(file.exists('models/movement-models-akdes-2024-03-18.rds')) {
+  akdes <-
+    readRDS('models/movement-models-akdes-2024-03-18.rds') %>%
+    # remove animals that are not range resident (invalid weights)
+    filter(! animal %in% c('SCEK014', 'SCEK014b', 'BW028')) %>%
+    pull(akde) %>%
+    map_dfr(\(.a) {
+    SpatialPolygonsDataFrame.UD(.a, level.UD = 0.9995, level = 0) %>%
+      st_as_sf() %>%
+      st_geometry() %>%
[…]

### 2024-06-10 analysis/00-preview-boreal-data.R

@@ -1,0 +1,24 @@
+library(dplyr)
+library(ggplot2)
+theme_set(theme_bw())
+
+wolf <- readxl::read_xlsx('data/tracking-data/webexportwolf.xlsx') %>%
+  mutate(timestamp = as.POSIXct(paste(FixDate, FixTime),
+                                format = "%Y-%m-%d %H:%M:%S"))
+wolf
+
+caribou <- readxl::read_xlsx('data/tracking-data/webexportcaribou.xlsx') %>%
+  mutate(timestamp = as.POSIXct(paste(FixDate, FixTime),
+                                format = "%Y-%m-%d %H:%M:%S"))
+caribou
+
+n_distinct(wolf$FieldID)
+n_distinct(caribou$FieldID)
+
+cowplot::plot_grid(
+  ggplot() +
+    geom_line(aes(FixDateTime, Collar, group = FieldID), wolf) +
+    ggtitle('Wolves'),
+  ggplot() +
+    geom_line(aes(FixDateTime, Collar, group = FieldID), caribou) +
+    ggtitle('Caribou'))

### 2024-06-10 analysis/01-compile-tracking-data.R

@@ -1,0 +1,315 @@
+library('dplyr')     # for data wrangling (mutate(), %>%, etc.)
+library('lubridate') # for working with dates (force_tz(), etc.)
+library('purrr')     # for functional programming (map_***(), etc.)
+library('ctmm')      # for movement models
+library('sf')        # for spatial features
+library('ggplot2')   # for fancy plots
+source('functions/import_rda.R') # to convert Rda files to rds files
+source('data/bc-shapefile.R') # for shapefile of BC
+
+bc_tz <- 'America/Vancouver' # to avoid typos
+
+# import all telemetries separately ----
+# mountain goats ----
+goats <-
+  readRDS('data/tracking-data/oreamnos-americanus-tels.rds') %>%
+  mutate(animal = individual.local.identifier, # to add a column of ID
+         timezone = bc_tz,
+         species = 'Oreamnos americanus',
+         dataset_name = 'Oreamnos_americanus',
+         outlier = FALSE) # outlier column is not present
+
+# pumas ----
+pumas <- bind_rows(
+  # puma 2
+  import_rda('data/tracking-data/Puma_concolor_2.Rda') %>%
+    as_tibble() %>%
+    arrange(AnimalID, timestamp) %>% # some data out of order
+    mutate(AnimalID = as.character(AnimalID),
+           animal = AnimalID,
+           species = 'Puma concolor',
+           dataset_name = 'Puma_concolor_2',
+           timestamp = ymd_hms(timestamp, tz = bc_tz),
+           Time.Zone = bc_tz) %>% # unique: PST, DST, PST/PDT: assume BC tz
+    rename(individual.local.identifier = AnimalID,
+           location.lat = Latitude,
+           location.long = Longitude,
+           timezone = Time.Zone,
+           outlier = Outlier) %>%
+    select(species, dataset_name, animal, individual.local.identifier,
+           location.lat, location.long, timestamp, timezone, outlier),
+  # puma 4
+  import_rda('data/tracking-data/Puma_concolor_4.Rda') %>%
+    as_tibble() %>%
+    mutate(animal = individual.local.identifier,
+           species = 'Puma concolor',
+           dataset_name = 'Puma_concolor_4',
+           timezone = bc_tz,
+           timestamp = ymd_hms(timestamp, tz = bc_tz),
[…]

### 2024-06-10 analysis/01a-compile-goat-calibration-data.R

@@ -1,0 +1,116 @@
+library('dplyr')     # for data wrangling (mutate(), %>%, etc.)
+library('tidyr')     # for data wrangling (nest(), unnest(), pivot_*, etc.)
+library('purrr')     # for functional programming (map_***(), etc.)
+library('ctmm')      # for movement models
+library('lubridate') # for working with dates
+library('ggplot2')   # for fancy plots
+theme_set(theme_bw())
+
+# source custom functions
+source('functions/outlier_plots.R') # to plot outlier diagnostic plots
+source('functions/check_animal.R') # to run diagnostic plots
+source('functions/plot_adj.R') # to plot 20 adjacent locations
+source('functions/flag_outlier.R') # to mark outliers
+
+# import goat data
+goats <- readRDS('data/tracking-data/all-tracking-data-not-cleaned.rds') %>%
+  filter(species == 'Oreamnos americanus') %>%
+  select(! c(hdop, individual.local.identifier, species, dataset_name,
+             gps.fix.type.raw:gps.satellite.count))
+
+# only keep calibration data
+d <- goats %>%
+  filter(location.long > -119.7) %>%
+  mutate(animal = paste(year(timestamp), month(timestamp), animal,
+                        sep = '_')) %>%
+  group_by(animal) %>%
+  mutate(individual.local.identifier = animal) %>% # for nesting ID 
+  nest(tel = ! animal) %>%
+  arrange(animal)
+
+# look at number of rows for each tag
+hist(map_int(d$tel, nrow), breaks = 20)
+
+# look at variograms for each set
+variograms <- function() {
+  N <- ceiling(sqrt(nrow(d)))^2
+  layout(matrix(1:N, ncol = sqrt(N), byrow = TRUE))
+  for(i in 1:nrow(d)) {
+    ctmm.guess(as.telemetry(d$tel[[i]]), interactive = FALSE)
+    title(d$animal[i])
+  }
+  layout(1)
+}
+variograms()
+
+# collars turned on for use in 2019-06; dropping all >= 2019 to be safe
+d <- filter(d, map_int(tel, ~ unique(year(.x$timestamp))) < 2019)
+variograms()
+
+d <- d %>%
+  mutate(tel = map(tel,
+                   ~ suppressMessages(as.telemetry(.x) %>%
[…]

### 2024-06-10 analysis/02-remove-outliers.R

@@ -1,0 +1,1614 @@
+library('dplyr')     # for data wrangling (mutate(), %>%, etc.)
+library('tidyr')     # for data wrangling (nest(), unnest(), pivot_*, etc.)
+library('purrr')     # for functional programming (map_***(), etc.)
+library('ctmm')      # for movement models
+library('lubridate') # for working with dates
+library('mapview')   # for interactive maps
+library('ggplot2')   # for fancy plots
+theme_set(theme_bw())
+
+# source custom functions
+source('functions/outlier_plots.R') # to plot outlier diagnostic plots
+source('functions/check_animal.R') # to run diagnostic plots
+source('functions/plot_adj.R') # to plot 20 adjacent locations
+source('functions/flag_outlier.R') # to mark outliers
+source('functions/remove_outlier_flags.R') # to start over with an animal
+
+# dataset checks ----
+# some NA timestamps (not enough to make a difference)
+readRDS('data/tracking-data/all-tracking-data-not-cleaned.rds') %>%
+  filter(is.na(timestamp)) %>%
+  group_by(dataset_name) %>%
+  summarize(n = n())
+
+# ensure all only have one column of HDOP
+readRDS('data/tracking-data/all-tracking-data-not-cleaned.rds') %>%
+  filter(! is.na(timestamp)) %>%
+  mutate(outlier = if_else(outlier, 1, 0)) %>%
+  filter(! is.na(hdop) + ! is.na(HDOP) > 1)
+
+# should ony have one column of DOP
+readRDS('data/tracking-data/all-tracking-data-not-cleaned.rds') %>%
+  filter(! is.na(timestamp)) %>%
+  mutate(outlier = if_else(outlier, 1, 0)) %>%
+  filter(! is.na(DOP) + ! is.na(gps.dop) > 1)
+
+# import the full dataset ----
+# import the dataset after dropping NAs
+d <- readRDS('data/tracking-data/all-tracking-data-not-cleaned.rds') %>%
+  filter(! is.na(timestamp)) %>%
+  mutate(hdop = if_else(is.na(hdop), HDOP, hdop), # merge error columns
+         dop = if_else(is.na(DOP), gps.dop, DOP),
+         original_outliers = outlier,
[…]

### 2024-06-10 analysis/03-movement-models-parallel.R

@@ -1,0 +1,126 @@
+library('dplyr') # for data wrangling (mutate(), %>%, etc.)
+library('tidyr') # for data wrangling (unnest(), pivot_*, etc.)
+library('purrr') # for functional programming (map_***(), etc.)
+library('furrr') # for futures and parallel computing
+library('ctmm')  # for movement models
+
+#' `https://github.com/ctmm-initiative/ctmmlearn/blob/main/ctmm_error.R`
+# goats are the only species with calibration data
+goat_uere <-
+  as.telemetry('data/tracking-data/goat-calibration-data.csv') %>%
+  uere.fit()
+
+# running models separately to avoid having to re-run previous code
+plan(multisession, workers = 10)
+if(FALSE) {
+  d <-
+    readRDS('data/tracking-data/all-tracking-data-cleaned-2024-02-22-13-49.rds') %>%
+    mutate(dataset_name =
+             factor(dataset_name,
+                    levels = c('Rangifer_tarandus_boreal',
+                               'Canis_lupus_boreal',
+                               'Rangifer_tarandus_southern_mountain',
+                               'Elk in southwestern Alberta',
+                               'Oreamnos_americanus',
+                               'Puma_concolor_2',
+                               'Puma_concolor_4',
+                               'Ursus_arctos_horribilis'))) %>%
+    arrange(dataset_name) %>% # to fit boreal caribou first
+    mutate(
+      # convert data frames to telemetry objects
+      tel = map(tel, \(x) as.telemetry(x, mark.rm = TRUE)),
+      # add calibration-informed error or a reasonable guess
+      tel = if_else(condition = species == 'Oreamnos americanus',
+                    true = map(tel, \(x) {
+                      uere(x) <- goat_uere
+                      return(x)
+                    }),
+                    false = map(tel, \(x) {
+                      # not changing default errors: assuming errors of 10 m
+                      prior <- uere(x) # extract calibration object
+                      prior$DOF[] <- 2 # low DOF for large uncertainty
+                      uere(x) <- prior # assign prior to the data
+                      return(x)
+                    })),
+      variogram = future_map(tel, \(x) ctmm.guess(data = x,
+                                                  CTMM = ctmm(error = TRUE),
+                                                  interactive = FALSE),
[…]

### 2024-06-10 analysis/04-dowload-era5-temperature-data.R

@@ -1,0 +1,117 @@
+library('dplyr')     # for data wrangling
+library('ncdf4')     # for working with nc rasters
+library('terra')     # for working with rasters
+library('sf')        # for working with simple features
+library('lubridate') # for working with dates
+library('purrr')     # for functional programming
+library('ctmm')      # for working with akdes
+source('data/bc-shapefile.R')
+
+#' using `{furrr}` causes: `Error: external pointer is not valid`
+
+# find extent of the data for the download requests ----
+# downloading data from https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-single-levels?tab=form
+d <- readRDS('models/movement-models-akdes-2024-03-18.rds')
+
+# find bounding box for the UDs for the download requests
+shp <- d %>%
+  pull(akde) %>%
+  map(\(.a) {
+    # extract a very wide UD with no CIs
+    SpatialPolygonsDataFrame.UD(.a, level = 0, level.UD = 0.9999) %>%
+      st_as_sf() %>%
+      st_set_crs(.a@info$projection) %>% # using tpeqd projection
+      st_transform('+proj=longlat') %>% # unproject
+      bind_rows() %>%
+      # take estimate only
+      filter(grepl(pattern = 'est', x = name)) %>%
+      # convert to a shapefile of the bounding box
+      st_bbox() %>%
+      st_as_sfc() %>%
+      st_as_sf()
+  }) %>%
+  # join all bboxes together
+  bind_rows() %>%
+  st_combine() %>%
+  # take a bbox from all of them
+  st_bbox() %>%
+  st_as_sfc() %>%
+  st_as_sf() %>%
+  # buffer by 10 kilometers
+  st_buffer(1e4)
+
+st_bbox(shp)
+
+# rounded to 2 decimals:
+#    xmin    ymin    xmax    ymax
+# -141.61   46.49 -103.89   70.14
+
+# check range of years
+d %>%
+  pull(tel) %>%
+  map_dfr(data.frame) %>% #' `as.data.frame()` fails
+  pull(timestamp) %>%
+  year() %>%
+  range()
+
+# raster requests took 1.5-5.25 h to fill and ~5 minutes to download ----
+
+# copy the downloads to the repo ----
+downloaded_nc_files <- list.files(path = '~/../Downloads',
+                                  pattern = '\\.nc', full.names = TRUE)
+
[…]

### 2024-06-10 analysis/05-add-speeds-and-temperature.R

@@ -1,0 +1,119 @@
+library('dplyr')     # for data wrangling (e.g., mutate, %>%)
+library('tidyr')     # for data wrangling (e.g., nest, unnest)
+library('purrr')     # for functional programming (e.g., map_***())
+library('furrr')     # for parallel functional programming (e.g., future_map())
+library('sf')        # for working with spatial data
+library('terra')     # for working with raster data
+library('ncdf4')     # for nc datasets
+library('ggplot2')   # for fancy plots
+library('ctmm')      # for movement modeling
+library('lubridate') # for working with dates
+source('functions/detrend_speeds.R') # to remove baseline noise in speed
+
+theme_set(theme_bw())
+
+# add weather data and speeds to the telemetry data ----
+# see: https://confluence.ecmwf.int/display/CKB/ERA5-Land%3A+data+documentation
+
+#' custom `which()` that returns `NA` instead of `integer(0)`
+which2 <- function(logical_vector) {
+  z <- which(logical_vector)
+  
+  if(length(z) > 0) {
+    return(z)
+  } else {
+    return(NA_integer_)
+  }
+}
+
+# estimate speeds ----
+if(FALSE) {
+  CORES <- availableCores(logical = FALSE) - 1
+  plan(multisession, workers = CORES)
+  
+  d <-
+    readRDS('models/movement-models-akdes-2024-06-06.rds') %>%
+    mutate(
+      # estimate speeds
+      tel = future_imap(tel, \(.tel, i) {
+        detrend.speeds(DATA = .tel, CTMM = movement_model[[i]]) %>%
+          transmute(speed_lwr = low, speed_est = est, speed_upr = high) %>%
+          bind_cols(data.frame(.tel), .) %>%
+          # add AKDE weights
+          mutate(weight = unique(akde[[i]]$DOF.H) * akde[[i]]$weights) %>%
+          return()
+      }, .progress = TRUE, .options = furrr_options(seed = TRUE))) %>%
+    # drop unnecessary columns
+    select(! c(variogram, movement_model, akde,  mm_file_name,
+               akde_file_name))%>%
+    unnest(tel)
+  
+  plan(sequential)
+  saveRDS(d, paste0('data/movement-models-speed-weights-', Sys.Date(), '.rds'))
+} else {
+  d <- readRDS(paste0('data/movement-models-speed-weights-2024-06-10.rds'))
+}
+
[…]

### 2024-06-10 analysis/06-temperature-hgams.R

@@ -1,0 +1,403 @@
+library('dplyr')     # for data wrangling
+library('tidyr')     # for data wrangling
+library('lubridate') # for smother date wrangling
+library('mgcv')      # for Generalized Additive Models
+library('ggplot2')   # for fancy plots
+library('khroma')    # for colorblind-friendly color palettes
+library('cowplot')   # for fancy multi-panel plots
+source('analysis/figures/default-ggplot-theme.R') # bold text and no grids
+
+#' clipping temperature data to the 0.8 quantile doesn't fix the high
+#' probabilities for goats
+d <-
+  readRDS('data/movement-models-speed-weights-temperature-2024-04-05.rds') %>%
+  filter(! is.na(speed_est), is.finite(speed_est)) %>%
+  mutate(speed_est = round(speed_est, 2), # some values are ~1e-10
+         animal = factor(animal),
+         species = if_else(grepl('Rangifer', dataset_name),
+                           dataset_name,
+                           species),
+         species = stringr::str_replace_all(species, '_', ' ') %>%
+           factor(),
+         # using a higher limit doesn't change much
+         moving = factor(if_else(speed_est == 0, 'no', 'yes')),
+         timestamp_bc = with_tz(timestamp, tz = 'America/Vancouver'),
+         tod = (hour(timestamp) / 24 +
+                  minute(timestamp) / 60 / 24 +
+                  second(timestamp) / 60 / 60 / 24),
+         doy = yday(timestamp)) %>%
+  rename(temp_c = temperature_C)
+
+SPECIES <- unique(d$species)
+N_SPECIES <- length(SPECIES)
+
+# drop single data point in February for bears
+plot(as.numeric(moving) ~ doy,
+     filter(d, species == 'Ursus arctos horribilis'))
+d <- filter(d, ! (species == 'Ursus arctos horribilis' & doy < 50))
+
+# drop extreme temperatures ----
+# some species have data with extreme temperatures but little data
+ggplot(d, aes(temp_c)) +
+  facet_wrap(~ species, scales = 'free') +
+  geom_histogram(color = 'black', binwidth = 1) +
+  geom_vline(aes(xintercept = value), color = 'red',
+             d %>%
+               group_by(species) %>%
+               summarize(lwr = quantile(temp_c, 0.025),
[…]

### 2024-06-10 analysis/06-temperature-rsfs.R

@@ -1,0 +1,222 @@
+library('mgcv')  # for GAMs
+library('dplyr') # for data wrangling
+library('tidyr') # for data wrangling
+library('sf')    # for spatial data
+library('ctmm')  # for movement models and utilization distributions
+library('terra') # for rasters
+library('purrr') # for functional programming
+
+K <- 1e6 # scaling constant for weights
+
+# import movement models
+mm <- readRDS('models/movement-models-akdes-2024-03-18.rds') %>%
+  # remove animals that are not range resident (invalid weights)
+  filter(! animal %in% c('SCEK014', 'SCEK014b', 'BW028')) %>%
+  mutate(species = if_else(grepl('Rangifer', dataset_name),
+                           dataset_name,
+                           species),
+         species = stringr::str_replace_all(species, '_', ' ') %>%
+           factor())
+
+# import tracking data with annonated temperatures
+d_1 <-
+  readRDS('data/movement-models-speed-weights-temperature-2024-04-05.rds') %>%
+  # remove animals that are not range resident (invalid weights)
+  filter(! animal %in% c('SCEK014', 'SCEK014b', 'BW028')) %>%
+  mutate(species = if_else(grepl('Rangifer', dataset_name),
+                           dataset_name,
+                           species),
+         species = stringr::str_replace_all(species, '_', ' ') %>%
+           factor()) %>%
+  select(species, animal, longitude, latitude, temperature_C, weight)
+
+locs_mcp <- map_dfr(mm$tel, \(.t) {
+  .t %>%
+    SpatialPoints.telemetry() %>%
+    adehabitatHR::mcp() %>%
+    st_as_sf() %>%
+    st_geometry() %>%
+    #' st_make_valid() %>% #' use if `Error: Loop x edge x has duplicate...`
+    st_union() %>% # unite estimate with lower and upper 95% CIs
+    st_transform('EPSG:4326') %>%
+    st_as_sf()
+}) %>%
+  st_make_valid() %>%
+  st_union() %>%
+  st_as_sf()
+
+hr_bounds <- map_dfr(mm$akde, \(.a) {
+  SpatialPolygonsDataFrame.UD(.a, level.UD = 0.95, level = 0) %>%
+    st_as_sf() %>%
+    st_geometry() %>%
[…]

### 2024-06-10 analysis/07-download-bc-climateNA-data.R

@@ -1,0 +1,115 @@
+library('climatenaR') # for downloading climate data and projections
+library('purrr')      # for functional programming
+
+#' if necessary, install the `climatenaR` package with
+#' `remotes::install_github('burnett-m/climatenaR', build_vignettes = TRUE)`
+#' *NOTE:* the `climatenaR` package requires the ClimateNA or climateBC
+#'         software. See https://register.climatena.ca/ to download it.
+
+#' metadata for `climateNA` data at:
+#' https://s3-us-west-2.amazonaws.com/www.cacpd.org/documents/ClimateNAv7_manual.pdf
+
+#' change the working directory as required by `climatenaR`
+setwd('ClimateNA_v742')
+
+if(! file.exists('bc-dem-z3.csv')) {
+  #' convert the bc DEM to a csv as required by `climatenaR`
+  demToCSV(file = '../data/bc-dem-z3.tif',
+           outdir = '.', # save in climateBC folder
+           srs = NULL) # keep NULL if in lat/long
+  
+  # check the csv
+  read.csv('bc-dem-z3.csv', nrows = 5)
+}
+
+if(! dir.exists('bc-dem-z3')) dir.create('bc-dem-z3')
+
+# download climate data projections ----
+if(! dir.exists('bc-dem-z3/projection-data')) {
+  dir.create('bc-dem-z3/projection-data')
+}
+
+map(2011:2100,
+    \(y) {
+      cat(paste0('Downloading climate projections for ', y, '...\n'))
+      projClimateNA(file = 'bc-dem-z3.csv',
+                    tFrame = 'M', # monthly averages
+                    exe = 'ClimateNA_v7.42.exe', # must be in wd
+                    scen = '8GCM', # 8GCMs_ensemble General Circulation Model
+                    ssp = c('S1', 'S2', 'S3', 'S5'), # SSP 4 not available
+                    years = as.character(y)) # can only do one year at a time
+    })
+
+# download historical climate data ----
+if(! dir.exists('bc-dem-z3/historical-data')) {
+  dir.create('bc-dem-z3/historical-data')
+}
+
+# tracking data ranges from 1998 to 2021
+map(1998:2023,
+    \(y) {
+      cat(paste0('Downloading estimated historical data for ', y, '...\n'))
+      histClimateNA(file = 'bc-dem-z3.csv',
+                    dateR = as.character(y), # year
+                    tFrame = 'M', # monthly averages
[…]

### 2024-06-10 analysis/08-bind-climateNA-data.R

@@ -1,0 +1,78 @@
+library('purrr')     # for functional programming (map_***())
+library('dplyr')     # for dara wrangling (mutate, transmute, etc.)
+library('tidyr')     # for data wrangling (pivot_longer, pivot_wider, nest)
+library('stringi')   # for working with strings
+library('lubridate') # for working with dates
+
+# to bind all climate projections from each year together into a single RDS file
+d <-
+  # import all files
+  map_dfr(
+    list.files('H:/GitHub/bc-mammals-speeds/ClimateNA_v742/bc-dem-z3',
+      full.names = TRUE, # include folder names in file name
+      pattern = '@'), # only every 10 years
+    \(.fname) {
+      # readr::read_csv(.fname, col_types = '?', progress = FALSE) %>%
+      data.table::fread(.fname, showProgress = FALSE) %>%
+        mutate(file = .fname) # add column of filename for scenario & year
+    }, .progress = TRUE) %>%
+  # filter to the approximate extent of the data
+  filter(Longitude > -127, Longitude < -113,
+         Latitude > 47, Latitude < 62) %>%
+  # add scenario and year columns
+  mutate(scenario = substr(
+    file,
+    start = stri_locate_last(file, regex = '/')[1] + 1,
+    stop = stri_locate_first(file, regex = '@')[1] - 1),
+    year = substr(file,
+                  start = stri_locate_first(file, regex = '@')[1] + 1,
+                  stop = nchar(file) - nchar('.csv'))) %>%
+  # only keep necessary columns
+  select(scenario, year, Latitude, Longitude, Elevation,
+         Tave01:Tave12, Tmin01:Tmin12, Tmax01:Tmax12)
+
+#' mean temperature is measured as `Tave = (Tmax + Tmin) / 2`
+#' `https://www.prism.oregonstate.edu/documents/PRISM_datasets.pdf`
+#' *can simulate hourly temperature using a smooth sinusoidal function*
+with(slice_sample(d, n = 5e4),
+     plot(Tave01 - Tmin01, Tmax01 - Tave01, col = '#00000050', pch = '.'))
+for(a in c(-0.1, 0, 0.1)) abline(a = a, b = 1, col = '#FF000040'); rm(a)
+
+# splitting to reduce RAM costs
+# pivot to long format
+d <- pivot_longer(d, ! c(scenario, year, Latitude, Longitude, Elevation),
+                  names_to = 'parameter', values_to = 'value')
+
+d <- d %>%
[…]

### 2024-06-10 analysis/09-model-era5-temperature-data.R

@@ -1,0 +1,151 @@
+library('dplyr')     #' for data wrangling
+library('tidyr')     #' for data wrangling
+library('lubridate') #' for smother date wrangling
+library('mgcv')      #' for Generalized Additive Models
+library('ggplot2')   #' for fancy plots
+library('gratia')    #' for ggplot GAM graphics
+library('terra')     #' for working with rasters
+library('ggridges')  #' for ridgeline plots
+library('khroma')    # for colorblind-friendly color palettes
+source('analysis/figures/default-ggplot-theme.R') # bold text and no grids
+source('data/bc-shapefile.R')
+
+# import temperature data ----
+SAMPLE <- seq(1, 365 * 24 * 2, by = 50) # sample of rasters for test
+
+if(FALSE) {
+  NCHAR <- nchar('data/ecmwf-era5-2m-temperature/epsg-4326/ecmwf-era5-2m-temp-X')
+  
+  d <- rast(list.files('data/ecmwf-era5-2m-temperature/epsg-4326',
+                       full.names = TRUE)) %>%
+    crop(bc_unproj) %>%
+    mask(bc_unproj) %>%
+    as.data.frame(xy = TRUE) %>%
+    # mutate(elev_m = terra::extract(dem, tibble(x, y))[, 2]) %>%
+    pivot_longer(- c(x, y), names_to = 'name', values_to = 'temp_k') %>%
+    mutate(year = as.numeric(substr(name, nchar('ecmwf-era5-2m-temp-X'),
+                                    nchar('ecmwf-era5-2m-temp-XXXX'))),
+           # get raster number hour (same as hour number using UTC)
+           h = as.numeric(substr(name,
+                                 nchar('ecmwf-era5-2m-temp-1998-epsg-4326_X'),
+                                 nchar(name))),
+           date = date_decimal(year + h / 365 / 24),
+           temp_c = temp_k - 273.15) %>%
+    select(x, y, date, temp_c) %>%
+    mutate(year = year(date),
+           month = month(date)) %>%
+    group_by(x, y, year, month) %>%
+    summarize(monthly_mean = mean(temp_c, na.rm = TRUE),
+              monthly_var = var(temp_c, na.rm = TRUE),
+              n = n(),
+              .groups = 'drop') %>%
+    rename(long = x, lat = y)
[…]

### 2024-06-10 analysis/figures/actws-2024-jasper/actws-2024-jasper-theme.R

@@ -1,0 +1,4 @@
+theme_set(theme_bw() +
+            theme(legend.position = 'none',
+                  panel.grid = element_blank(),
+                  text = element_text(size = 15)))

### 2024-06-10 analysis/figures/actws-2024-jasper/bc-resource-rasters.R

@@ -1,0 +1,88 @@
+library('sf')      # for working with spatial data
+library('terra')  # for working with raster data (tifs)
+library('dplyr')   # for data wrangling (e.g., mutate, %>%, ...)
+library('ggplot2') # for fancy plots
+source('analysis/figures/actws-2024-jasper/actws-2024-jasper-theme.R')
+source('data/bc-shapefile.R')
+
+theme_set(theme_void() +
+            theme(axis.text = element_blank(),
+                  axis.ticks = element_blank(),
+                  panel.border = element_blank(),
+                  panel.background = element_blank(),
+                  legend.position = c(0.85, 0.7),
+                  legend.text = element_text(face = 'bold'),
+                  legend.title = element_text(face = 'bold')))
+
+# create custom color palettes for each predictor
+forest_pal <- colorRampPalette(c('#C2B280', 'darkgreen'))(100)
+elevation_pal <- colorRampPalette(c('white', 'brown4'))(100)
+water_pal <- c('transparent', 'blue')
+
+f <- rast('data/resource-rasters/bc-forest.tif') %>%
+  project(crs(bc)) %>%
+  as.data.frame(xy = TRUE) %>%
+  filter(! is.na(layer))
+
+e <- rast('data/resource-rasters/bc-dem.tif') %>%
+  project(crs(bc)) %>%
+  as.data.frame(xy = TRUE) %>%
+  filter(layer > 0)
+
+w <- rast('data/resource-rasters/bc-distance-from-water.tif') %>%
+  project(crs(bc)) %>%
+  crop(bc) %>%
+  mask(bc) %>%
+  as.data.frame(xy = TRUE, na.rm = TRUE) %>%
+  rename(layer = consensus_full_class_12)
+
+# check ranges of each variable
+range(f$layer)
+range(e$layer)
+range(w$layer)
+
+if(any(f$layer < 0) | any(f$layer > 100) | any(w$layer < 0)) {
+  stop('Check ranges.\n')
+} else {
+  cat('Ranges are ok.\n')
+}
+
+# plot the rasters
+p_f <-
+  ggplot() +
+  geom_raster(aes(x, y, fill = layer), f) +
+  geom_sf(data = bc, fill = NA, color = 'black') +
+  scale_x_continuous(NULL, breaks = NULL, expand = c(0, 0)) +
+  scale_y_continuous(NULL, breaks = NULL, expand = c(0, 0)) +
[…]

### 2024-06-10 analysis/figures/actws-2024-jasper/kelowna-jasper-map.R

@@ -1,0 +1,35 @@
+library('sf')      # for spatial objects
+library('ggplot2') # for fancy plots
+library('dplyr')   # for data wrangling
+
+provs <- canadianmaps::PROV %>%
+  st_geometry() %>%
+  st_transform('EPSG:3005')
+
+plot(provs)
+
+locs <- tibble(city = c('Kelowna', 'Jasper'),
+               long = c(-119.4966, -118.0814),
+               lat = c(49.8863, 52.8737)) %>%
+  st_as_sf(coords = c('long', 'lat')) %>%
+  st_set_crs('+proj=longlat') %>%
+  st_transform('EPSG:3005') %>%
+  bind_cols(.,
+            st_coordinates(.) %>%
+              as.data.frame() %>%
+              rename(alb_long = X, alb_lat = Y))
+
+plot(locs, add = TRUE)
+
+ggplot() +
+  geom_sf(data = provs) +
+  geom_sf(data = locs, size = 2) +
+  geom_label(aes(x = alb_long, y = alb_lat, label = city), locs,
+             nudge_x = 5e4, nudge_y = 5e4, label.size = 0, size = 6,
+             bg = 'transparent') +
+  coord_sf(xlim = c(25e4, 26e5), ylim = c(37e4, 175e4)) +
+  theme(legend.position = 'none')
+
+ggsave('figures/actws-2024-jasper/kelowna-jasper-map.png',
+       width = 33.87, height = 19.05, units = 'cm', dpi = 600,
+       bg = 'transparent')

### 2024-06-10 analysis/figures/actws-2024-jasper/model-dags.R

@@ -1,0 +1,52 @@
+library('dagitty') # for DAGs
+library('ggdag')   # for DAGs with ggplot2
+library('ggplot2') # for fancy plots
+theme_set(theme_dag() +
+            theme(text = element_text(size = 10),
+                  legend.position = 'none'))
+
+# caribou only ----
+#' also see `ggdag_status()` and `ggdag_adjustment_set(shadow = TRUE)`
+
+# P(moving)
+dag_car_mov <- dagify(moving ~ temp + DOY + TOD,
+                      temp ~ DOY + TOD,
+                      exposure = 'temp',
+                      outcome = 'moving')
+set.seed(2)
+ggdag_status(dag_car_mov, node_size = 20, text_size = 3) +
+  scale_color_manual(values = c('black', 'darkorange'))
+
+ggsave('figures/actws-2024-jasper/dag-caribou-moving.png',
+       width = 6, height = 4, dpi = 600, bg = 'transparent')
+
+# speed | moving
+dag_car_sp <- dagify(speed ~ temp + DOY + TOD + moving,
+                     # moving ~ temp + DOY + TOD,
+                     temp ~ DOY + TOD,
+                     exposure = 'temp',
+                     outcome = 'speed')
+set.seed(7)
+ggdag_status(dag_car_sp, node_size = 20, text_size = 3) +
+  scale_color_manual(values = c('black', 'darkorange'))
+
+ggsave('figures/actws-2024-jasper/dag-caribou-speed.png',
+       width = 6, height = 4, dpi = 600, bg = 'transparent')
+
+# habitat selection
+dag_car_hs <- dagify(selected ~ forest + water + elevation,
+                     forest ~ temp,
+                     water ~ temp,
+                     elevation ~ temp,
+                     exposure = 'temp',
+                     outcome = 'selected',
+                     coords = data.frame(
+                       x = c(10, 0, 1, 9, 5),
+                       y = c(10, 0, 9, 1, 5),
+                       name = c('selected', 'temp', 'forest', 'water',
+                                'elevation')))
+ggdag_status(dag_car_hs, node_size = 20, text_size = 3) +
+  scale_color_manual(values = c('black', 'darkorange'))
+
+ggsave('figures/actws-2024-jasper/dag-caribou-rsf.png',
+       width = 6, height = 4, dpi = 600, bg = 'transparent')

### 2024-06-10 analysis/figures/actws-2024-jasper/na-yearly-temperature-anomalies.R

@@ -1,0 +1,18 @@
+library('dplyr')   # for data wrangling
+library('ggplot2') # for fancy plots
+source('analysis/figures/actws-2024-jasper/actws-2024-jasper-theme.R')
+
+# temperature anomalies relative to average from 
+d <- read.csv('data/climate-anomalies/data.csv', skip = 4) %>%
+  mutate(Date = as.Date(paste0(Date, '15'), format = '%Y%m%d'))
+
+ggplot(d, aes(Date, Anomaly, color = Anomaly > 0)) +
+  geom_bar(stat = 'identity', fill = 'transparent') +
+  geom_hline(yintercept = 0) +
+  labs(x = NULL, y = paste0('Temperature Anomaly (\U00B0', 'C)')) +
+  ylim(c(-2.25, 2.25)) +
+  scale_x_date(expand = c(0, 365 * 2)) +
+  scale_color_manual(values = c('#0571B0', '#CA0020'))
+
+ggsave('figures/actws-2024-jasper/temperature-anomalies.png',
+       scale = 1, width = 10, height = 6, dpi = 600)

### 2024-06-10 analysis/figures/actws-2024-jasper/rsf-figures.R

@@ -1,0 +1,409 @@
+library('mgcv')    # for GAMs
+library('terra')   # for rasters
+library('sf')      # for spatial objects
+library('dplyr')   # for data wrangling
+library('tidyr')   # for data wrangling
+library('ggplot2') # for fancy plots
+source('analysis/figures/default-ggplot-theme.R') # bold text and no grids
+
+theme_set(theme_get() + theme(text = element_text(size = 15)))
+
+rsf <- readRDS('models/rsf-boreal-caribou-2024-03-03.rds')
+
+# marginal effect of forest ----
+# empty figure
+tibble(temperature = 15,
+       forest_perc = seq(0, 100, by = 0.1),
+       elevation_m = 1,
+       dist_water_m = 1) %>%
+  bind_cols(.,
+            predict(rsf, newdata = .,
+                    terms = c('s(forest_perc)',
+                              's(forest_perc,temperature)'),
+                    type = 'link', se.fit = TRUE)) %>%
+  mutate(mu = exp(fit),
+         lwr = exp(fit - 1.96 * se.fit),
+         upr = exp(fit + 1.96 * se.fit)) %>%
+  ggplot() +
+  geom_hline(yintercept = 1, color = 'grey') +
+  geom_ribbon(aes(forest_perc, ymin = lwr, ymax = upr),
+              fill = 'transparent', alpha = 0.2) +
+  geom_line(aes(forest_perc, mu), color = 'transparent', linewidth =  1) +
+  scale_y_continuous('Relative habitat preference', trans = 'log2',
+                     breaks = c(0.25, 0.5, 1, 2, 4)) +
+  xlab('Forest (%)')
+
+ggsave('figures/actws-2024-jasper/rsf-forest-empty.png',
+       width = 10, height = 6, dpi = 600)
+
+# real figure
+tibble(temperature = 15,
+       forest_perc = seq(0, 100, by = 0.1),
+       elevation_m = 1,
+       dist_water_m = 1) %>%
+  bind_cols(.,
+            predict(rsf, newdata = .,
+                    terms = c('s(forest_perc)',
+                              's(forest_perc,temperature)'),
+                    type = 'link', se.fit = TRUE)) %>%
+  mutate(mu = exp(fit),
+         lwr = exp(fit - 1.96 * se.fit),
+         upr = exp(fit + 1.96 * se.fit)) %>%
+  ggplot() +
[…]

### 2024-06-10 analysis/figures/actws-2024-jasper/telemetry-map.R

@@ -1,0 +1,88 @@
+library('dplyr')        # for data wrangling
+library('tidyr')        # for data wrangling
+library('ggplot2')      # for fancy plots
+library('khroma')       # for color blind-friendly palettes
+source('data/bc-shapefile.R')
+source('analysis/figures/actws-2024-jasper/actws-2024-jasper-theme.R')
+
+locs <- tibble(city = c('Kelowna', 'Jasper'),
+               long = c(-119.4966, -118.0814),
+               lat = c(49.8863, 52.8737)) %>%
+  st_as_sf(coords = c('long', 'lat')) %>%
+  st_set_crs('+proj=longlat') %>%
+  st_transform('EPSG:3005') %>%
+  bind_cols(.,
+            st_coordinates(.) %>%
+              as.data.frame() %>%
+              rename(alb_long = X, alb_lat = Y))
+
+# boreal datasets
+d <- readRDS('data/tracking-data/all-tracking-data-cleaned-2023-11-17-16-04.rds') %>%
+  filter(dataset_name == 'Rangifer_tarandus_boreal') %>%
+  unnest(tel) %>%
+  filter(! outlier) %>%
+  select(location.long, location.lat, dataset_name, species) %>%
+  st_as_sf(coords = c('location.long', 'location.lat')) %>%
+  st_set_crs('+proj=longlat') %>%
+  st_transform('EPSG:3005')
+
+# phylopics added manually using hex codes below
+pal <- c(color('bright')(7), brown = '#654321')
+
+# caribou data only
+p_caribou <-
+  ggplot() +
+  theme_void() +
+  theme(legend.position = 'none') +
+  geom_sf(data = prov) +
+  geom_sf(data = d, color = '#EE6677') +
+  geom_sf(data = locs, size = 2) +
+  geom_label(aes(x = alb_long, y = alb_lat, label = city), locs,
+             nudge_x = 5e4, nudge_y = 5e4, label.size = 0, size = 6,
+             bg = 'transparent') +
+  coord_sf(xlim = c(2.5e5, 26e5), ylim = c(4.2e5, 18e5)) +
+  scale_color_bright(name = 'Species') +
+  labs(x = NULL, y = NULL)
+
+ggsave('figures/actws-2024-jasper/caribou-map.png', p_caribou,
+       width = 33.87, height = 19.05, units = 'cm', dpi = 600,
+       bg = 'transparent')
+
+# all species
[…]

### 2024-06-10 analysis/figures/actws-2024-jasper/temperature-hgams-figures.R

@@ -1,0 +1,487 @@
+library('dplyr')     # for data wrangling
+library('lubridate') # for smother date wrangling
+library('mgcv')      # for Generalized Additive Models
+library('ggplot2')   # for fancy plots
+library('khroma')    # for colorblind-friendly color palettes
+library('sf')        # for spatial objects
+source('analysis/figures/default-ggplot-theme.R') # bold text and no grids
+
+theme_set(theme_get() +
+            theme(text = element_text(size = 15)))
+
+# using boreal caribou
+readRDS('data/tracking-data/all-tracking-data-cleaned-2024-02-22-13-49.rds') %>%
+  group_by(dataset_name) %>%
+  tidyr::unnest(tel) %>%
+  summarise(prop = mean(! is.na(temperature)),
+            total = sum(! is.na(temperature))) %>%
+  arrange(desc(total))
+
+# import GAMs
+m_1 <- readRDS('models/binomial-gam-2024-03-06-boreal-caribou.rds')
+m_2 <- readRDS('models/gamma-gam-2024-03-06-boreal-caribou.rds')
+
+# P(moving) ----
+#' check ranges of `temperature`
+m_1$model %>%
+  summarise(min = min(temperature),
+            max = max(temperature))
+
+p_change <- function(.data) {
+  inv_logit <- m_1$family$linkinv
+  
+  mutate(.data,
+         mu = (inv_logit(fit) - 0.5) * 100,
+         lwr = (inv_logit(fit - 1.96 * se.fit) - 0.5) * 100,
+         upr = (inv_logit(fit + 1.96 * se.fit) - 0.5) * 100) %>%
+    return()
+}
+
+#' marginal effect of `temperature`
+tibble(temperature = seq(-30, 40, by = 0.1),
+       time_of_day = 12,
+       doy = 100,
+       animal = 'new animal') %>%
+  bind_cols(.,
+            predict(m_1, newdata = ., terms = c('s(temperature)'),
+                    type = 'link', se.fit = TRUE, discrete = FALSE)) %>%
+  p_change() %>%
+  ggplot() +
+  geom_hline(yintercept = 0, col = 'grey') +
+  geom_ribbon(aes(temperature, ymin = lwr, ymax = upr), fill = '#EE6677',
+              alpha = 0.2) +
+  geom_line(aes(temperature, mu), color = '#EE6677', linewidth =  1) +
[…]

### 2024-06-10 analysis/figures/default-ggplot-theme.R

@@ -1,0 +1,7 @@
+library('ggplot2') # for fancy figures
+
+theme_set(theme_bw() +
+            theme(panel.grid = element_blank(),
+                  text = element_text(face = 'bold')))
+
+PAL <- unname(c(khroma::color('bright')(6), brown = '#654321'))

### 2024-06-10 analysis/figures/telemetry-ud-map.R

@@ -1,0 +1,69 @@
+library('dplyr')   # for data wrangling
+library('tidyr')   # for data wrangling
+library('sf')      # for working with simple features
+library('purrr')   # for functional programming
+library('ctmm')    # for working with akdes
+library('ggplot2') # for fancy plots
+source('analysis/figures/default-ggplot-theme.R')
+source('data/bc-shapefile.R')
+
+d <- readRDS('models/movement-models-akdes-2024-03-18.rds') %>%
+  mutate(dataset_name = case_when(
+    dataset_name == 'Canis_lupus_boreal' ~ 'Canis lupus (boreal)',
+    dataset_name == 'Rangifer_tarandus_boreal' ~ 'Rangifer tarandus (boreal)',
+    dataset_name == 'Rangifer_tarandus_southern_mountain' ~ 'Rangifer tarandus (southern mountain)',
+    dataset_name == 'Puma_concolor_2' ~ 'Puma concolor',
+    dataset_name == 'Puma_concolor_4' ~ 'Puma concolor',
+    dataset_name == 'Elk in southwestern Alberta' ~ 'Cervus elaphus',
+    dataset_name == 'Oreamnos_americanus' ~ 'Oreamnos americanus',
+    dataset_name == 'Ursus_arctos_horribilis' ~ 'Ursus arctos horribilis'))
+
+tels <- transmute(d,
+                  dataset_name,
+                  tel = map(tel, \(.t) {
+                    .t %>%
+                      data.frame() %>%
+                      select(longitude, latitude) %>%
+                      st_as_sf(coords = c('longitude', 'latitude')) %>%
+                      st_set_crs('+proj=longlat') %>%
+                      st_transform('EPSG:3005') %>%
+                      as.data.frame()
+                  })) %>%
+  unnest(tel)
+
+uds <- d %>%
+  transmute(
+    dataset_name,
+    akde =  map(akde, \(.a) {
+      SpatialPolygonsDataFrame.UD(.a, level = 0, level.UD = 0.95) %>%
+        st_as_sf() %>%
+        st_set_crs(.a@info$projection) %>%
+        filter(grepl(pattern = 'est', x = name)) %>%
+        st_transform('EPSG:3005')
+    }) %>%
+      bind_rows() %>%
+      pull(geometry))
+
+# create the figures ----
+# UDs only
+p_uds <-
+  ggplot() +
+  geom_sf(data = bc) +
[…]

### 2024-06-10 functions/detrend_speeds.R

@@ -1,0 +1,66 @@
+# Code written by Chris H Fleming and adapted by Stefano Mezzini
+library('ctmm')       # for movement modeling
+library('data.table') # for memory-efficient data.frames
+
+# Function for de-trending empirical speed estimates from the movement model
+detrend_speeds <- function(DATA, CTMM, units = FALSE) {
+  
+  # check if model is OUF
+  if(! grepl('OUF', toupper(summary(CTMM)$name))) {
+    warning('Found a non-OUF/OUf model; cannot estimate speeds.\n')
+    
+    SPEEDS <- data.frame(low = NA_real_,
+                         est = NA_real_,
+                         high = NA_real_,
+                         time = DATA$timestamp)
+  } else {
+    # estimate speeds conditional on the data
+    EST <- speeds(DATA, CTMM, fast = TRUE, level = NULL, units = units)
+    # null estimates of speed (no data, only model)
+    EST.NULL <- speeds(CTMM, t = DATA$t, fast = TRUE, level = NULL,
+                       units = units)
+    
+    # reduce DOF by that of null distribution
+    DOF <- EST$DOF - EST.NULL$DOF
+    all(DOF>0) # check for validity... always has held so far
+    DOF <- pmax(0,DOF) # haven't needed this, but just in case
+    
+    S2 <- EST$speed^2 + EST$VAR
+    
+    AVE <- (EST$DOF*S2)/DOF
+    
+    # Calculate CIs
+    CI <- sapply(
+      1:length(DATA$t),
+      function(i){ ctmm:::chisq.ci(MLE = AVE[i], DOF = DOF[i],
+                                   level = 0.95, robust = TRUE)})
+    CI <- sqrt(CI)
+    
+    SPEEDS <- as.data.frame(t(CI))
+    
+    SPEEDS$time <- DATA$timestamp
+  }
+  return(SPEEDS)
+}
+
+# test the function
+if(FALSE) {
+  library('lubridate')
+  data('buffalo')
+  
+  TRACK <- buffalo[[1]][1:500,]
+  GUESS <- ctmm.guess(TRACK, interactive = FALSE)
+  FIT <- ctmm.fit(TRACK, GUESS, control = list(method = "pNewton",
+                                               cores = -1))
+  EST <- speeds(TRACK, FIT, fast=TRUE, level=NULL)
[…]

### 2024-06-10 functions/import_rda.R

@@ -1,0 +1,23 @@
+# function to import Rda file but with a custom name
+import_rda <- function(file, object_name = NA, rm_rda = TRUE) {
+  # load the Rda file in the function's environment (not the global one)
+  load(file)
+  
+  # extract the object's name from the file name
+  if(is.na(object_name)) {
+    object_name <- stringr::str_replace(string = file,
+                                        pattern = 'data/tracking-data/',
+                                        replacement = '')
+    object_name <- stringr::str_replace(object_name, '.Rda', replacement = '')
+  }
+  
+  # return the imported object
+  return(get(object_name))
+}
+
+# test the function
+if(FALSE) {
+  head(import_rda('data/tracking-data/Oreamnos_americanus.Rda'))
+  head(import_rda('data/tracking-data//Rangifer_tarandus.Rda',
+                  object_name = 'data'))
+}

### 2024-06-10 functions/test_speeds.R

@@ -1,0 +1,6 @@
+test_speeds <- function(id, trace = TRUE, cores = 1) {
+  tel <- as.telemetry(d$tel[[which(d$animal == id)]], mark.rm = TRUE)
+  guess <- ctmm.guess(tel, interactive = FALSE)
+  m_0 <- ctmm.select(tel, CTMM = guess, trace = trace, cores = cores)
+  summary(m_0, units = FALSE)$CI['speed (meters/second)', ]
+}

### 2024-06-13 writing/manuscript.Rmd
edited intro

@@
 ---
-title: "Predicting how climate change and short-term weather will affect terrestrial mammals' movement and habitat selection"
+title: "Predicting how climate change will affect terrestrial mammals' movement and habitat selection"
 author:
@@
   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
-bibliography: 'bc-mammal-speeds.bib'
+bibliography: 'bc-mammals-temperature.bib'
 # csl: 'csl.csl'
+# csl: 'nature.csl'
 fontsize: 12pt
@@
     toc: false
-    number_sections: false
+    number_sections: true
 editor_options: 
@@
 \doublespacing
-*
+
 ```{r setup, include=FALSE}
@@
 knitr::opts_chunk$set(echo = FALSE, out.width = '\\textwidth',
-                      fig.align = 'center', cache = TRUE)
+                      fig.align = 'center', cache = TRUE, warning = FALSE,
+                      message = FALSE)
 ```
@@
 
+```{r, echo=FALSE}
+abstract <- ""
+```
+
+<!-- elements of the title page -->
+
+\noindent \textbf{Article type}: Major article
+
+\noindent \textbf{Words in abstract}: `r stringi::stri_count_words(abstract)`
+
+\noindent \textbf{Words in main text}: `r suppressMessages(wordcountaddin::word_count())`
+
+\noindent \textbf{Figures}: 2
+
+\noindent \textbf{Tables}: 1
+
+\noindent \textbf{References}: 33 (updated on 2024-06-13) <!-- counted by using nature csl -->
+
+\noindent \textbf{Appendices}: 3
+
+\noindent \textbf{Key words:} energetics, environmental stochasticity, home range, range size, `ctmm`, simulations
+
+<!-- Authors' emails:
+Stefano Mezzini: stefano.mezzini@ubc.ca
+Chris H. Fleming: christen.fleming@ucf.edu
+E. Patrícia Medici: medici@ipe.org.br
+Michael J. Noonan: michael.noonan@ubc.ca
+-->
+
 # Abstract
 
+\noindent `r abstract`
+
+\newpage
+
 # To do
-
-Intro:
-
-* add references:
-  * McNabb 1960s
-  * Hirt 2022
-  * Moose speed and temp (just @alston_temperature_2020 or are there others?)
-  * Lots of work on fish
-  * Brown temp constant for all biological processes
-  * Andean hummingbird and temp
-* address any questions answered in the discussion
-
-Methods:
-
-* add headers
-
-Write abstract
 
@@
 
-The rate at which chemical reactions occur within an organism [metabolism, see @brown_toward_2004] depends strongly on the temperature reactions occur at.
-However, organisms experience different ranges of external temperatures and, consequently, exhibit specific thermal optima.
-These depend on many factors, including body size [@taylor_energetics_1982; @reiss_scaling_1988; @fjelldal_nightly_2021], physiological adaptations [@schmidt_interplay_2020; @mohr_cellular_2020], and behavior [@arechavala-lopez_common_2019].
-In the case of heterothermic organisms, low temperatures increase energetic costs [@melin_effect_2023; @ref; @fish; @andean_hummingbird], while high temperatures increase the risk of overheating [@ref; @cheetas; @alston_temperature_2020].
+Animals' energetic costs depend strongly on ambient temperature [@brown_toward_2004].
+Cold temperatures increase the metabolic costs of movement [@taylor_energetics_1982] and homeothermy [@melin_effect_2023], which can constrain an animal's ability to move [@schwerdt_ecological_2024], its fitness [@walker_global_2019], and, ultimately, its odds of survival [@berger_climate_2018].
+Hot temperatures tend to lower energetic needs [@taylor_energetics_1982], but they also cause animals to overheat, which lowers their ability to move [@powers_hovering_2017; @alston_temperature_2020; @dyer_travel_2023].
 
-At extremely cold temperatures, the metabolic costs of movement and homeothermy increase energetic expenditures, which can constrain an animal's ability to move, lowering encounter rates with food [@hou_cold_2020], fitness [@ref], and, ultimately, chance of survival [@berger_climate_2018].
-Mammals reduce the adverse effects of cold through a combination of both behavioral adaptations [e.g., searching for shelter: @ref; reducing movement rates: @ref] and physiological adaptations [e.g., fur: @ref; circulation: @ref].
-At extremely hot temperatures, excessively high body temperatures ***XXXX*** [@alston_temperature_2020; @refs].
+Mammals reduce the adverse effects of extreme temperatures through both physiological and behavioral adaptations [@jessen_temperature_2001; @mota-rojas_physiological_2021], but their ability to respond through physiological adaptations is limited by their evolutionary history, since their optimal temperature and range of tolerable temperatures (Fig.
+\@ref(fig:temperature-optima)) depend on the environment they live in [@levins_evolution_1974].
+In contrast, changes in movement rates and space use allow mammals to respond to stressors immediately [@alston_temperature_2020; @dyer_travel_2023; @attias_effects_2018; @alston_temperature_2020; @giroux_activity_2023; @verzuh_behavioural_2023].
 
-Evolutional constraints limit mammals' ability to respond temperature extremes, which results in phylogenetically-constrained temperature optima for different species and populations [@caribou-ref].
-Thus, while animals are adapted to fluctuations around average conditions [@ref], strong and consistent temperature anomalies can decrease an animal's energetic stability [@ref], reproductive success [@schmidt_interplay_2020], and ability to move [@ref].
-Recent changes in climate have resulted in widespread warming, milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@ref_2021_heat_dome].
-These changes are expected to affect mammals' fitness, movement behavior, and habitat selection as they cope with growing heat stress [@weststrate_how_2024].
+```{r temperature-optima, fig.width=6, fig.height=4, out.width='50%', fig.cap="Hypothetical ranges in fitness as a consequence of ambient temperature for a cold-adapted mammal (blue), temperate-adapted mammal (yellow), and hot-adapted mammal (red).
+Contrast the differences in fitness gains at different temperatures, as well as differences in the height and spead of the three curves."}
+library('dplyr')
+library('tidyr')
+library('ggplot2')
+library('khroma')
+theme_set(theme_bw() + theme(panel.grid = element_blank()))
 
-This paper provides a BC-focused analysis on the effects of proximate weather and long-term climate on mammal movement.
-Using more than a decade of telemetry data and proximal 3-hour weather data, we estimate how animals altered their movement frequency, movement speed, and habitat selection in response to air temperature.
+tibble(x = seq(-30, 30, by = 0.01),
+       cold = rev(dgamma(x + 12, shape = 10, rate = 0.5)),
+       mid = dnorm(x, mean = 5, sd = 5),
+       warm = dgamma(x, shape = 56, rate = 2.8)) %>%
+  pivot_longer(-c(x)) %>%
+  ggplot() +
+  geom_area(aes(x, value, color = name, fill = name), alpha = 0.2,
+            show.legend = FALSE, position = 'identity') +
+  scale_fill_highcontrast() +
+  scale_color_highcontrast() +
+  scale_x_continuous(paste0('Temperature \U00B0', 'C'), expand = c(0, 0)) +
+  scale_y_continuous('Fitness', breaks = NULL, limits = c(0, 0.16), expand = c(0, 0))
+```
+
+During the last century, changes in climate in British Columbia (BC), Canada, have resulted in widespread warming, milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@eberle_technical_2022].
+These changes are expected continue over the next 100 years and compound their effects on mammals' fitness, movement behavior, and habitat selection as they cope with growing heat stress and human pressure [@deb_modelling_2020; @woo-durand_increasing_2020; @weststrate_how_2024].
+
+This paper provides a BC-focused analysis on the effects of proximate weather and long-term climate on the movement of terrestrial mammal species.
+Using over two decades of telemetry data and proximal hourly weather data, we estimate how animals altered their movement frequency, movement speed, and habitat selection in response to air temperature.
 We then pair these responses with climate change projections to forecast changes mammals movement in the current century under different climate change scenarios [@riahi_shared_2017].
-The analyses include data from six species (*Canis lupus*, *Cervus elaphus*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, *Ursus arctos horribilis*; see Table \@ref(tab:data-table)) across a large spatial range of British Columbia (Fig.
+The analyses include data from six species (*Canis lupus*, *Cervus elaphus*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, and *Ursus arctos horribilis*; see Table \@ref(tab:data-table)) across a large spatial range of British Columbia (Fig.
 \@ref(fig:bc-map)).
@@
 ```{r data-table}
-readRDS('../models/movement-models-2024-03-18.rds') %>%
+readRDS('../models/movement-models-2024-06-06.rds') %>%
   transmute(Dataset = if_else(grepl('Puma', dataset_name),
@@
 Animal telemetry datasets were obtained from multiple sources.
-Elk tracking data were obtained from Movebank [@kays_movebank_2022; study name: Elk in southwestern Alberta, see @study_ref], while the boreal caribou and wolf data were obtained from a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data/](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data/)).
+Elk tracking data were obtained from Movebank [@kays_movebank_2022; study name: Elk in southwestern Alberta, see @ciuti_human_2012], while the boreal caribou and wolf data were obtained from a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data/](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data/)).
 ***All other tracking data were obtained from private collaborators*** [@refs_to_studies].
@@
 Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$), especially if points before or after the potential outlier indicated stationary behavior.
-The script used to clean the data and all custom functions used are available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-speeds](https://github.com/QuantitativeEcologyLab/bc-mammals-speeds)).
+The script used to clean the data and all custom functions used are available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature)).
 
@@
 
-Rasters of three-hour reanalyzed air temperature data were downloaded from the ERA5 dataset [@copernicus_climate_change_service_era5_2019] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)).
+Rasters of three-hour reanalyzed air temperature data were downloaded from the ERA5 dataset [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)).
 Proximal weather conditions were then estimated for each location by extracting the values from the corresponding cell from the temporally nearest raster using the `extract()` function from the `terra` package for `R` [version 1.7-46, @hijmans_terra_2023].
@@
 Since `ctmm`'s movement models allow for stochastic speeds but assume constant movement, speeds were set to zero if they were not significantly different from zero ($\alpha = 0.05$), under the assumption of a $\chi$-distributed speeds.
-***NOT SURE IF THIS IS CORRECT*** The function used is available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-speeds/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-speeds/blob/main/functions/detrend_speeds.R)).
+***NOT SURE IF THIS IS CORRECT*** The function used is available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R)).
 To avoid artifacts due to excessively small, non-zero speeds, estimated speeds were then clustered into two groups using a $k$-means algorithm with two centers (see Appendix A).
@@
 
-We estimated the effects of hourly temperature on mammals' movement using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_pedersen_hierarchical_2019] via `mgcv` package for `R`.
+We estimated the effects of hourly temperature on mammals' movement using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019] via `mgcv` package for `R`.
 The first HGAM had a binomial family of distribution and estimated the probability of a mammal moving as a function of time of day, day of year, and temperature.
@@
 If we shrink both $\Delta u$ and $\Delta t$ to a single point in space and a single moment in time, we now have $\lambda(u, t) \le 1$, since we cannot observe more than one animal in a given point in time and spce $(u, t)$.
-Additional theoretical background is provided by @aarts_estimating_2008 and @alson_mitingting_2022 and in Appendix B.
+Additional theoretical background is provided by @aarts_estimating_2008 and @alston_mitigating_2022 and in Appendix B.
 
@@
 * energetics
-* encounter rates
+* encounter rates (lower encounter rates with food [@hou_cold_2020])
 * HWI

### 2024-06-13 functions/detrend_speeds.R
corrected typo in function test

@@ -53,7 +53,7 @@ if(FALSE) {
   FIT <- ctmm.fit(TRACK, GUESS, control = list(method = "pNewton",
                                                cores = -1))
   EST <- speeds(TRACK, FIT, fast=TRUE, level=NULL)
-  TEST <- detrend.speeds(TRACK, FIT)
+  TEST <- detrend_speeds(TRACK, FIT)
   
   TEST$time2 <- hour(TEST$time) + (minute(TEST$time)/60)
   EST$time2 <- hour(EST$timestamp) + (minute(EST$timestamp)/60)

### 2024-06-13 writing/manuscript.Rmd
edited methods except for cc weather predictions

@@
 
-## Data acquisition
+## GPS telemetry data
 
-### GPS telemetry data
-
-Animal telemetry datasets were obtained from multiple sources.
-Elk tracking data were obtained from Movebank [@kays_movebank_2022; study name: Elk in southwestern Alberta, see @ciuti_human_2012], while the boreal caribou and wolf data were obtained from a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data/](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data/)).
-***All other tracking data were obtained from private collaborators*** [@refs_to_studies].
-Outlier GPS locations were removed following diagnostic analyses of the location of each point as well as straight-line displacement and turning angle between consecutive points.
-Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$), especially if points before or after the potential outlier indicated stationary behavior.
+Elk (*Cervus elaphus*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data/](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data/)), and the mountain goat (*Oreamnos americanus*) locations were provided by BC Parks.
+All other tracking data were obtained from private collaborators.
+Outlier GPS locations were removed following diagnostic analyses of the distance from the median location as well as straight-line displacement, turning angle, and time interval between consecutive points.
+Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and high straight-line displacement, especially if antecedent and subsequent points indicated stationary behavior.
 The script used to clean the data and all custom functions used are available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature)).
 
-### Historical temperature data
+## Historical temperature data
 
-Rasters of three-hour reanalyzed air temperature data were downloaded from the ERA5 dataset [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)).
-Proximal weather conditions were then estimated for each location by extracting the values from the corresponding cell from the temporally nearest raster using the `extract()` function from the `terra` package for `R` [version 1.7-46, @hijmans_terra_2023].
+Rasters of hourly reanalyzed air temperature data were downloaded from the ERA5 dataset [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)).
+Proximal weather conditions were estimated for each location by extracting the values from the corresponding raster cell from the temporally nearest raster using the `extract()` function from the `terra` package [version 1.7-46, @hijmans_terra_2023] for `R`  [@r_core_team_r_2024].
 
-## Data modeling
+## Estimating mammals' instantaneous speeds
 
-### Estimating mammals' instantaneous speeds
+We estimated modeled each animal's movement using continuous-time movement models [@fleming_fine-scale_2014] via the `ctmm` package [version 1.2.0, @fleming_ctmm_2023] for `R`.
+We then estimated mammals' instantaneouus speeds by applying the `ctmm::speeds()` function on all models with finite speed estimates [416 of 433, see @fleming_fine-scale_2014; @noonan_scale-insensitive_2019].
+All 17 telemetries with insufficiently fine sampling were for caribou (15 boreal and 2 southern mountain).
+Since `ctmm`'s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), speeds were set to zero if they were not significantly different from zero ($\alpha = 0.05$), under the assumption of a $\chi$-distributed speeds.
+***NOT SURE IF THIS IS CORRECT*** The function used is available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R)).
+To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not by applying the `stats::kmeans()` function with 2 centers (Appendix A).
 
-Animal movement models were fit using the `ctmm` package [version 1.1.0, @fleming_ctmm_2021] for `R` [@r_core_team_r_2023].
-Mammals' speeds were then estimated using the `ctmm::speeds()` function on all models with speed estimates [@noonan_scale-insensitive_2019].
-Most telemetries (416 of 433) had sufficiently frequent locations to provide appropriate speed estimates.
-All 17 telemetries with insufficiently fine sampling were for caribou (15 boreal and 2 souther mountain).
+## Estimating the effects of temperature on mammals' movement
 
-Since `ctmm`'s movement models allow for stochastic speeds but assume constant movement, speeds were set to zero if they were not significantly different from zero ($\alpha = 0.05$), under the assumption of a $\chi$-distributed speeds.
-***NOT SURE IF THIS IS CORRECT*** The function used is available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R)).
-To avoid artifacts due to excessively small, non-zero speeds, estimated speeds were then clustered into two groups using a $k$-means algorithm with two centers (see Appendix A).
+We estimated the effects of temperature on mammals' state (moving or not) and speed (given that it was moving) using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019] via the `mgcv` package for `R` [version 1.9-1, @wood_generalized_2017].
+The first HGAM had a binomial family of distribution and estimated the probability of a mammal moving as a function of time of day, day of year, and temperature, with separate smooths (but common smoothness parameter) for each species [model S in Figure 4 of @pedersen_hierarchical_2019].
+In addition, the model also included a random effect for each animal, a tensor interaction product of time of day and day of year to account for changes in day length, and a tensor interaction product of time of day and temperature to account for changes in daily activity levels with temperature, for each species.
+Due to the substantial latitudinal separation between boreal and southern mountain caribou, we considered the two gropus as separate species for the sake of modeling.
+The model accounted for the cyclicity of time of day and day of year using cyclic cubic splines, which provide continuity in the smooth term by ensuring the function and its slope (first derivative) and curvature (second derivative) are continuous between consecutive cycles.
+The second HGAM had the same structure, but it used a Gamma family of distributions to estimate the speed of an animal, given that it was moving.
+Together, the two HGAMs inform us on an animal's long-term speed (i.e., distance traveled per unit time), since it is the product of probability that the animal is moving and its average speed when moving.
+Additional details are provided in Appendix A.
 
-### Effects of temperature on mammals' movement
-
-We estimated the effects of hourly temperature on mammals' movement using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019] via `mgcv` package for `R`.
-The first HGAM had a binomial family of distribution and estimated the probability of a mammal moving as a function of time of day, day of year, and temperature.
-Since there was substantial variation in the direction and wiggliness of each species' trends, the effects were modeled independently `by` smooths for each species [see model I in Figure 4 of @pedersen_hierarchical_2019].
-The model accounted for the cyclicity of time of day and day of year using cyclic cubic splines, which provide continuity in the smooth term by ensuring the function and its slope (first derivative) and curvature (second derivative) are continuous between consecutive cycles.
-The second HGAM had the same structure, but it used a Gamma family of distributions to estimate the speed of an animal given that it was moving.
-Together, the two HGAMs inform us on an animal's long-term speed (i.e., distance traveled per unit time), since it is the product of probability that the animal is moving and the speed it moves at, given that it is moving.
-
-### Effects of temperature on mammals' habitat selection
-
-An animal's habitat selection is proportional to the number of times one can observe it in an environment.
-Consequently, one can view the spatial distributions of an animal in space as a poisson process whose mean depends on the amount of space, $\Delta u$, and time, $\Delta t$, the observation takes place over, which we can write as $\lambda(\Delta u, \Delta t)$.
-If we shrink both $\Delta u$ and $\Delta t$ to a single point in space and a single moment in time, we now have $\lambda(u, t) \le 1$, since we cannot observe more than one animal in a given point in time and spce $(u, t)$.
-Additional theoretical background is provided by @aarts_estimating_2008 and @alston_mitigating_2022 and in Appendix B.
-
-Under the asumption of a Poisson family of distributions, we estimated the effects of temperature on each species' habitat selection by fitting a Resource Selection Function (RSF) via an HGAM using the `mgcv` package.
-***here***
+We estimated the effects of temperature on each species' selection for percent forest cover, elevation, and distance from fresh water using an HGAM with a Poisson family of distributions [Appendix B, also see @aarts_estimating_2008; @alston_mitigating_2022].
+While we recognize there are other important drivers of habitat selection, we decided to only use these three proxies to ensure results were relatively comparable between species and to make province-wide predictions simpler.
 
@@
 
-### Climate change projections
+***HERE***
 
-Rasters of monthly average temperature in BC during the next 80 years were obtained via the `climatenaR` package [version XXX; @burnett_climatenar_2023] for `R`.
-Since the climate projections of monthly average temperatures did not provide a measure of temperature variance for each month, we estimated within-month variance using a Generalized Additive Model with a Gamma family of distributions fit via the `mgcv` package for `R` [@wood_generalized_2017].
-Additional details are provided in Appendix C.
+Rasters of projected monthly average temperature in BC during the next 80 years were obtained via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`.
+Since the climate projections did not provide a measure of variation in temperature within each month due to weather, we estimated within-month variance by fitting a Generalized Additive Model with a Gamma family of distributions to the square-root precision ($\sqrt{\rho}$) in temperature for each month using the ERA5 data for BC during 1998-2024.
+The model included...
+(Appendix C).
+
+***model currently uses weights = sqrt(n); fitting a new one with uniform weights***
 
@@
 
-**should predict for the area the animals are in only**
+***should predict for the area the animals are in only***
 

### 2024-06-14 functions/detrend_speeds.R
allow function to estimate speeds for IOU models

@@ -6,8 +6,9 @@ library('data.table') # for memory-efficient data.frames
 detrend_speeds <- function(DATA, CTMM, units = FALSE) {
   
   # check if model is OUF
-  if(! grepl('OUF', toupper(summary(CTMM)$name))) {
-    warning('Found a non-OUF/OUf model; cannot estimate speeds.\n')
+  if(! (grepl('OUF', toupper(summary(CTMM)$name)) |
+        grepl('IOU', toupper(summary(CTMM)$name)))) {
+    warning('Found a model with non-finite speed.\n')
     
     SPEEDS <- data.frame(low = NA_real_,
                          est = NA_real_,

### 2024-06-14 data/bc-shapefile.R
push sharable data and READMEs

@@ -1,0 +1,14 @@
+library('sf')      # for spatial objects
+library('dplyr')   # for data wrangling
+
+prov_unproj <- canadianmaps::PROV %>%
+  st_geometry() %>%
+  st_as_sf()
+
+prov <- st_transform(prov_unproj, 'EPSG:3005')
+
+bc_unproj <- filter(canadianmaps::PROV, PRENAME == 'British Columbia') %>%
+  st_geometry() %>%
+  st_as_sf()
+
+bc <- st_transform(bc_unproj, 'EPSG:3005')

### 2024-06-14 writing/manuscript.Rmd
edited intro and methods; started results

@@
 
-# Abstract
+# Abstract {-}
 
@@
 
-\newpage
-
-# To do
+# To do {-}
 
@@
 Hot temperatures tend to lower energetic needs [@taylor_energetics_1982], but they also cause animals to overheat, which lowers their ability to move [@powers_hovering_2017; @alston_temperature_2020; @dyer_travel_2023].
+Fluctuations in ambient temperature, and particularly extreme cold events, lead to the evolution of eurythermy and endothermy [@portner_climate_2004].
 
-Mammals reduce the adverse effects of extreme temperatures through both physiological and behavioral adaptations [@jessen_temperature_2001; @mota-rojas_physiological_2021], but their ability to respond through physiological adaptations is limited by their evolutionary history, since their optimal temperature and range of tolerable temperatures (Fig.
-\@ref(fig:temperature-optima)) depend on the environment they live in [@levins_evolution_1974].
-In contrast, changes in movement rates and space use allow mammals to respond to stressors immediately [@alston_temperature_2020; @dyer_travel_2023; @attias_effects_2018; @alston_temperature_2020; @giroux_activity_2023; @verzuh_behavioural_2023].
+Mammals reduce the adverse effects of extreme temperatures through both physiological and behavioral adaptations [@jessen_temperature_2001; @mota-rojas_physiological_2021].
+Long-term exposure to consistent ranges of temperatures allow species to evolve for higher efficiency within specific temperature ranges [@levins_evolution_1974], but physiological adaptations are limited by a species' experienced temperatures, so such adaptations are less effective during periods of rapid change [@huey_predicting_2012] or extreme changes [@sherwood_adaptability_2010].
+In contrast, mammals can respond to changes in temperature much more rapidly and efficiently by altering their movement rates and space use [@alston_temperature_2020; @dyer_travel_2023; @attias_effects_2018; @alston_temperature_2020; @giroux_activity_2023; @verzuh_behavioural_2023].
 
-```{r temperature-optima, fig.width=6, fig.height=4, out.width='50%', fig.cap="Hypothetical ranges in fitness as a consequence of ambient temperature for a cold-adapted mammal (blue), temperate-adapted mammal (yellow), and hot-adapted mammal (red).
-Contrast the differences in fitness gains at different temperatures, as well as differences in the height and spead of the three curves."}
-library('dplyr')
-library('tidyr')
-library('ggplot2')
-library('khroma')
-theme_set(theme_bw() + theme(panel.grid = element_blank()))
-
-tibble(x = seq(-30, 30, by = 0.01),
-       cold = rev(dgamma(x + 12, shape = 10, rate = 0.5)),
-       mid = dnorm(x, mean = 5, sd = 5),
-       warm = dgamma(x, shape = 56, rate = 2.8)) %>%
-  pivot_longer(-c(x)) %>%
-  ggplot() +
-  geom_area(aes(x, value, color = name, fill = name), alpha = 0.2,
-            show.legend = FALSE, position = 'identity') +
-  scale_fill_highcontrast() +
-  scale_color_highcontrast() +
-  scale_x_continuous(paste0('Temperature \U00B0', 'C'), expand = c(0, 0)) +
-  scale_y_continuous('Fitness', breaks = NULL, limits = c(0, 0.16), expand = c(0, 0))
-```
-
-During the last century, changes in climate in British Columbia (BC), Canada, have resulted in widespread warming, milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@eberle_technical_2022].
-These changes are expected continue over the next 100 years and compound their effects on mammals' fitness, movement behavior, and habitat selection as they cope with growing heat stress and human pressure [@deb_modelling_2020; @woo-durand_increasing_2020; @weststrate_how_2024].
+During the last century, changes in climate in British Columbia (BC), Canada, have resulted in widespread warming, milder and shorter winters, hotter and longer summers [@bunnell_global_2011; @deb_modelling_2020; @woo-durand_increasing_2020], which, combined, result in a greater risk of extremely high temperatures [@eberle_technical_2022].
+These changes are expected continue over the next 100 years and compound their effects on mammals' fitness, movement behavior, and habitat selection as they cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and anthopogenic pressure [@sih_evolution_2011; @weststrate_how_2024].
 
@@
 
-```{r bc-map, fig.cap="GPS telemetry data and autocorrelated kernel density utilization distributions for the six species included in this study."}
+```{r bc-map, fig.cap="GPS telemetry data and 95% autocorrelated kernel density utilization distributions for the six species in this study."}
 knitr::include_graphics('../figures/tels-and-uds-map.png')
@@
   rename(`Median sampling interval (hours)` = dt) %>%
-  knitr::kable(format = 'pipe', caption = 'Start and end of the GPS telemetries along with the median sampling interval for each of the six species included in this study.') %>%
+  knitr::kable(format = 'pipe', caption = 'Start and end of the GPS telemetries along with the median sampling interval for each of the six species in this study.') %>%
   kableExtra::column_spec(1, italic = TRUE) %>%
@@
 
-Elk (*Cervus elaphus*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data/](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data/)), and the mountain goat (*Oreamnos americanus*) locations were provided by BC Parks.
+Elk (*Cervus elaphus*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)), and the mountain goat (*Oreamnos americanus*) locations were provided by BC Parks.
 All other tracking data were obtained from private collaborators.
@@
 All 17 telemetries with insufficiently fine sampling were for caribou (15 boreal and 2 southern mountain).
-Since `ctmm`'s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), speeds were set to zero if they were not significantly different from zero ($\alpha = 0.05$), under the assumption of a $\chi$-distributed speeds.
+
+Since `ctmm`'s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed was 0.
+We performed this correction by subtracting the data-independent, background minimum speed while assuming speeds were $\chi^2$-distributed (which arises from assuming the X and Y components of the speed vectors are Gaussian).
 ***NOT SURE IF THIS IS CORRECT*** The function used is available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R)).
@@
 
+Both movement frequency and speed when moving decreased with temperature for all species (Fig.
+\@ref(fig:temp-marginal)), but the there were some differences among species in the intensity of the effect (Figs.
+\@ref(fig:p-move) and \@ref(fig:speed)).
+
+```{r temp-marginal, fig.cap="Estimated mean (black) and species-specifit (colored) effect of temperature on movement frequency (A) and speed when moving (B)."}
+layout(t(1:2))
+plot(1)
+plot(1)
+# knitr::include_graphics('../figures/p-moving-all.png')
+```
+
+***move the figures with seasonal and daily effects to Appendix A***
+
+```{r p-move, fig.cap="Estimated probability of moving as a function of tempeature (A), time of day (B), and day of year (C)."}
+knitr::include_graphics('../figures/p-moving-all.png')
+```
+
+```{r speed, fig.cap="Estimated speed as a function of tempeature (A), time of day (B), and day of year (C)."}
+knitr::include_graphics('../figures/speed-all.png')
+```
+
 ### Effects of temperature on mammals' habitat selection
+
+***should predict for the area the animals are in only***
+
+### Predicted changes in movement during the current century
 
@@
 
-# References
+# References {-}
 

### 2024-06-14 analysis/05-add-speeds-and-temperature.R
remove unnecessary paste0()

@@ -51,7 +51,7 @@ if(FALSE) {
   plan(sequential)
   saveRDS(d, paste0('data/movement-models-speed-weights-', Sys.Date(), '.rds'))
 } else {
-  d <- readRDS(paste0('data/movement-models-speed-weights-2024-06-10.rds'))
+  d <- readRDS('data/movement-models-speed-weights-2024-06-10.rds')
 }
 
 # extract temperature (takes about ~1 hour on lab machine w 11 cores) ----

### 2024-09-11 analysis/figures/temperature-hgams-projections.R
now using weighted means, rearranged factor levels

@@ -67,15 +67,15 @@ if(file.exists('data/cc-hgam-projections.rds')) {
            d = p * s,
            d_upr = p_upr * s_upr) %>%
     group_by(scenario, year, species) %>%
-    summarize(p_lwr = mean(p_lwr, weight = weight),
-              s_lwr = mean(s_lwr, weight = weight),
-              d_lwr = mean(d_lwr, weight = weight),
-              p = mean(p, weight = weight),
-              s = mean(s, weight = weight),
-              d = mean(d, weight = weight),
-              p_upr = mean(p_upr, weight = weight),
-              s_upr = mean(s_upr, weight = weight),
-              d_upr = mean(d_upr, weight = weight),
+    summarize(p_lwr = weighted.mean(p_lwr, w = weight),
+              s_lwr = weighted.mean(s_lwr, w = weight),
+              d_lwr = weighted.mean(d_lwr, w = weight),
+              p = weighted.mean(p, w = weight),
+              s = weighted.mean(s, w = weight),
+              d = weighted.mean(d, w = weight),
+              p_upr = weighted.mean(p_upr, w = weight),
+              s_upr = weighted.mean(s_upr, w = weight),
+              d_upr = weighted.mean(d_upr, w = weight),
               .groups = 'drop') %>%
     # center by the mean of the 2020 estimates for each species
     arrange(year, species, scenario) %>%
@@ -98,7 +98,7 @@ if(file.exists('data/cc-hgam-projections.rds')) {
            species = gsub(' ', '~', species) %>%
              gsub('~\\(', '\\)~bold\\((', .) %>%
              paste0('bolditalic(', ., ')') %>%
-             factor(levels = SPECIES_LABS))
+             factor())
   saveRDS(cc_proj, 'data/cc-hgam-projections.rds')
 }
 beepr::beep(2)

### 2024-09-11 analysis/06-temperature-hrsfs.R
fit boreal caribou hrsf

@@ -219,14 +219,15 @@ SPECIES <- d %>%
 
 for(sp in SPECIES) {
   cat('Working on ', as.character(sp), '...\n', sep = '')
-  d %>%
-    filter(species == sp) %>%
-    pivot_longer(c(forest_perc, elevation_m, dist_water_m)) %>%
-    ggplot() +
-    facet_wrap(~ name, scales = 'free') +
-    geom_histogram(aes(value, fill = factor(detected)),
-                   position = 'identity', alpha = 0.5, bins = 10) +
-    scale_fill_brewer('Detected', type = 'qual', palette = 6)
+  print(
+    d %>%
+      filter(species == sp) %>%
+      pivot_longer(c(forest_perc, elevation_m, dist_water_m)) %>%
+      ggplot() +
+      facet_wrap(~ name, scales = 'free') +
+      geom_histogram(aes(value, fill = factor(detected)),
+                     position = 'identity', alpha = 0.5, bins = 10) +
+      scale_fill_brewer('Detected', type = 'qual', palette = 6))
   
   rsf <- bam(
     detected ~

### 2024-09-11 analysis/figures/temperature-hgams.R
switched to percentage axes

@@ -8,6 +8,7 @@ library('ggplot2')   # for fancy plots
 library('khroma')    # for colorblind-friendly color palettes
 library('cowplot')   # for fancy multi-panel plots
 source('analysis/figures/default-ggplot-theme.R') # bold text and no grids
+source('functions/labeller_perc.R') # for percentage labels
 plot_scheme(PAL, colours = TRUE)
 
 # import data
@@ -426,7 +427,7 @@ s_tod_int <-
                      breaks = tod_breaks, labels = tod_labs) +
   scale_fill_gradientn(name = 'Relative change in speed', colors = s_pal,
                        limits = range(z_breaks), breaks = z_breaks,
-                       labels = round(2^z_breaks, 2)) +
+                       labels = \(x) labeller_perc(round(2^x, 2))) +
   theme(panel.background = element_rect(fill = 'grey'),
         legend.position = 'inside', legend.position.inside = c(0.66, 0.15),
         legend.key.width = rel(1.5),  legend.justification = 'center',
@@ -453,7 +454,7 @@ d_tod_int <-
   scale_fill_gradientn(name = 'Relative change in distance travelled',
                        colors = d_pal, limits = range(z_breaks) * 2,
                        breaks = z_breaks * 2,
-                       labels = round(2^(z_breaks * 2), 2)) +
+                       labels = \(x) labeller_perc(2^x)) +
   theme(panel.background = element_rect(fill = 'grey'),
         legend.position = 'inside', legend.position.inside = c(0.66, 0.15),
         legend.key.width = rel(1.5),  legend.justification = 'center',
@@ -514,8 +515,9 @@ s_doy_int <-
   scale_y_continuous('Day of year', expand = c(0, 0),
                      breaks = doy_breaks, labels = doy_labs) +
   scale_fill_gradientn(name = 'Relative change in speed', colors = s_pal,
-                       limits = range(z_breaks), breaks = z_breaks,
-                       labels = round(2^z_breaks, 2)) +
+                       limits = range(z_breaks),
+                       breaks = z_breaks,
+                       labels = \(x) labeller_perc(round(2^x, 2))) +
   theme(panel.background = element_rect(fill = 'grey'),
         legend.position = 'inside', legend.position.inside = c(0.66, 0.15),
         legend.key.width = rel(1.5),  legend.justification = 'center',
@@ -542,7 +544,7 @@ d_doy_int <-
   scale_fill_gradientn(name = 'Relative change in distance travelled',
[…]

### 2024-09-11 analysis/figures/temperature-hgams.R
reduced figure size to increase font size

@@ -594,4 +594,4 @@ distance <- plot_grid(
     theme(legend.position = 'none'),
   labels = c('', 'A', 'B'), ncol = 1, rel_heights = c(0.2, 1, 1))
 ggsave('figures/distance.png', distance,
-       width = 20, height = 10, dpi = 600, bg = 'white')
+       width = 16, height = 8, dpi = 600, bg = 'white')

### 2024-09-11 analysis/figures/temperature-hrsfs.R
removed division by median, added boreal caribou

@@ -122,9 +122,11 @@ preds %>%
   select(species, lab, x, temperature_C, variable, lambda, too_far_1,
          too_far_2) %>%
   filter((! too_far_2)) %>%
-  group_by(species, variable, temperature_C) %>%
+  # re-scale the center (1) relative to the median value 
+  group_by(species, variable) %>%
   mutate(lambda = lambda / median(lambda)) %>%
   ungroup() %>%
+  # cap at 2^(+/-2)
   mutate(
     log2_lambda = log2(lambda),
     log2_lambda = case_when(log2_lambda > 2 ~ 2,
@@ -144,10 +146,11 @@ preds %>%
                      breaks = c(-20, 0, 20)) +
   scale_y_continuous(NULL, expand = c(0, 0)) +
   scale_fill_sunset(name = 'Relative selection strength', midpoint = 0,
-                    limits = c(-2, 2), breaks = -2:2, labels = 2^c(-2:2)) +
+                    limits = c(-2, 2), breaks = -2:2, labels = \(x) 2^x) +
   theme(strip.placement = 'outside', strip.background.y = element_blank(),
         strip.text.y = element_text(size = 11), legend.position = 'top',
-        panel.background = element_rect(fill = 'grey'))
+        panel.background = element_rect(fill = 'grey'),
+        legend.key.width = rel(2))
 
-ggsave('figures/rsf-surface-plots.png', width = 2.5 * nrow(d), height = 8,
+ggsave('figures/rsf-surface-plots.png', width = 17.5, height = 8,
        units = 'in', dpi = 600, bg = 'white')

### 2024-09-11 writing/manuscript.Rmd
wrote abstract, added model definitions, and text

--- 2024-06-13 writing/manuscript.Rmd
@@
 ---
-title: "Predicting how climate change will affect terrestrial mammals' movement and habitat selection"
+title: "Predicting how climate change will affect terrestrial mammals' movement and habitat selection in British Columbia, Canada"
 author:
@@
 bibliography: 'bc-mammals-temperature.bib'
-# csl: 'csl.csl'
+csl: 'freshwater-biology.csl'
 # csl: 'nature.csl'
@@
     - \usepackage{hanging} # for hanging indents in references
+    - \usepackage[left]{lineno} # for line numbers
 subparagraph: true # needed for \usepackage[small]{titlesec}
@@
   bookdown::pdf_document2:
+    latex_engine: xelatex # for Delta symbol in table
     pandoc_args:
@@
 ---
+
+```{r setup, include=FALSE}
+# set chunk defaults: don't print code, images text width and centered
+knitr::opts_chunk$set(echo = FALSE, out.width = '\\textwidth',
+                      fig.align = 'center', cache = TRUE, warning = FALSE,
+                      message = FALSE)
+```
+
+<!--
+available text sizes:
+Huge > huge > LARGE > Large > large > normalsize > small > footnotesize > scriptsize > tiny
+-->
+
+<!--
+six-sentence abstract:
+
+1. introduce the topic
+2. state the problem
+3. summarize why the question hasn't been answered/what the consequences are
+4. explain how you answered the question
+5. explain the central findings
+6. explain the importance of your findings and their consequences
+-->
+
+```{r, echo=FALSE}
+abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move.
+These changes in species' movement behavior have had cascading effects on individuals' fitness and habitat selection with consequences at the population, community, and ecosystem levels.
+Previous research has quantified changes in mammalian movement behavior, but we are lacking an estimate of the effects of climate change on fine-scale movement and habitat selection.
+We address this by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on mammal's probability of movement, speed, and habitat selection.
+We then leverage the estimated effects to predict when, how, and where mammals will move in BC in the current century.
+We conclude with implications for energetics, encounter rates, habitat conservation, and human-wildlife conflicts."
+```
+
+<!-- elements of the title page -->
+
+\noindent \textbf{Article type}: Research article
+
+\noindent \textbf{Words in abstract}: `r stringi::stri_count_words(abstract)`
+
+\noindent \textbf{Words in main text}: `r suppressMessages(wordcountaddin::word_count())`
+
+\noindent \textbf{Figures}: 7
+
+\noindent \textbf{Tables}: 1
+
+\noindent \textbf{References}: 38 (updated on 2024-07-28) <!-- counted by using nature csl -->
+
+\noindent \textbf{Appendices}: 3
+
+\noindent \textbf{Key words:} climate change, temperature, mammals, animal movement, habitat selection
+
+\clearpage
+
+# Abstract {-}
+
+\noindent `r abstract`
+
+# Useful references to add {-}
+
+* @hegel_interacting_2010: weather affects caribou calf recruitment and interacts with predator (wolf) abundance
+* @elmore_implications_2017: "thermal environments should be foundational in the understanding of the habitat concept"
+* @pease_ecological_2024: the scale at which we measure variables affects their estumated effect
+* @hetem_responses_2014:
+* rate of climate change is too fast for genetic adaptation to occur in mammals with longevities of decades
+* example of suitable habitat prediction to 2050
+* to read: https://doi.org/10.1186/s40665-016-0024-1
 
@@
 
-```{r setup, include=FALSE}
-# set chunk defaults: don't print code, images text width and centered
-knitr::opts_chunk$set(echo = FALSE, out.width = '\\textwidth',
-                      fig.align = 'center', cache = TRUE, warning = FALSE,
-                      message = FALSE)
-```
+\linenumbers
 
@@
 
-```{r, echo=FALSE}
-abstract <- ""
-```
+<!--
 
-<!-- elements of the title page -->
+other refs
 
-\noindent \textbf{Article type}: Major article
+* A review of climate change effects on marine mammals in United States waters: Past predictions, observed impacts, current research and conservation imperatives https://doi.org/10.1016/j.ecochg.2022.100054
+* Global vulnerability of marine mammals to global warming: https://doi.org/10.1038/s41598-019-57280-3
 
-\noindent \textbf{Words in abstract}: `r stringi::stri_count_words(abstract)`
-
-\noindent \textbf{Words in main text}: `r suppressMessages(wordcountaddin::word_count())`
-
-\noindent \textbf{Figures}: 2
-
-\noindent \textbf{Tables}: 1
-
-\noindent \textbf{References}: 33 (updated on 2024-06-13) <!-- counted by using nature csl -->
-
-\noindent \textbf{Appendices}: 3
-
-\noindent \textbf{Key words:} energetics, environmental stochasticity, home range, range size, `ctmm`, simulations
-
-<!-- Authors' emails:
-Stefano Mezzini: stefano.mezzini@ubc.ca
-Chris H. Fleming: christen.fleming@ucf.edu
-E. Patrícia Medici: medici@ipe.org.br
-Michael J. Noonan: michael.noonan@ubc.ca
 -->
-
-# Abstract
-
-\noindent `r abstract`
-
-\newpage
-
-# To do
 
@@
 
-Animals' energetic costs depend strongly on ambient temperature [@brown_toward_2004].
-Cold temperatures increase the metabolic costs of movement [@taylor_energetics_1982] and homeothermy [@melin_effect_2023], which can constrain an animal's ability to move [@schwerdt_ecological_2024], its fitness [@walker_global_2019], and, ultimately, its odds of survival [@berger_climate_2018].
-Hot temperatures tend to lower energetic needs [@taylor_energetics_1982], but they also cause animals to overheat, which lowers their ability to move [@powers_hovering_2017; @alston_temperature_2020; @dyer_travel_2023].
+<!-- introduce knowledge gap -->
 
-Mammals reduce the adverse effects of extreme temperatures through both physiological and behavioral adaptations [@jessen_temperature_2001; @mota-rojas_physiological_2021], but their ability to respond through physiological adaptations is limited by their evolutionary history, since their optimal temperature and range of tolerable temperatures (Fig.
-\@ref(fig:temperature-optima)) depend on the environment they live in [@levins_evolution_1974].
-In contrast, changes in movement rates and space use allow mammals to respond to stressors immediately [@alston_temperature_2020; @dyer_travel_2023; @attias_effects_2018; @alston_temperature_2020; @giroux_activity_2023; @verzuh_behavioural_2023].
+\noindent It is well understood that mammalian energetics depend on ambient temperature [@brown_toward_2004; @taylor_energetics_1982], but we have a limited understanding of how mammals adapt their movement behavior in response to temperature [@elmore_implications_2017; but see: @milling_seasonal_2017; @montgomery_movement_2019].
+Previous work has focused on energetic costs of movement [@taylor_energetics_1982; @dyer_travel_2023], range shifts [@leclerc_determinants_2021], thermoregulation [@mota-rojas_physiological_2021; @melin_effect_2023], and trait-based responses [@mccain_assessing_2019], but there remains a need for a more comprehensive understanding of how mammals will adapt their fine-scale movement and habitat selection to future changes in temperature [@intergovernmental_panel_on_climate_change_climate_2023].
+Movement plasticity [@nathan_movement_2008] allows mammals to react to changes in temperature on a shorter timescale than evolutionary physiological adaptations [@levins_evolution_1974].
+By changing when, how, and where they move, mammals can conserve chemical and thermal energy as well as reduce the risk of overheating [@alston_temperature_2020; @dyer_travel_2023; @attias_effects_2018; @giroux_activity_2023; @verzuh_behavioural_2023; @williams_range_2018].
+We present a multi-species analysis of how mammals adapt their movement behavior in response to changes in air temperature.
 
-```{r temperature-optima, fig.width=6, fig.height=4, out.width='50%', fig.cap="Hypothetical ranges in fitness as a consequence of ambient temperature for a cold-adapted mammal (blue), temperate-adapted mammal (yellow), and hot-adapted mammal (red).
-Contrast the differences in fitness gains at different temperatures, as well as differences in the height and spead of the three curves."}
-library('dplyr')
-library('tidyr')
-library('ggplot2')
-library('khroma')
-theme_set(theme_bw() + theme(panel.grid = element_blank()))
+<!-- movement plasticity allows for rapid responses, but how does climate change impact this? -->
 
-tibble(x = seq(-30, 30, by = 0.01),
-       cold = rev(dgamma(x + 12, shape = 10, rate = 0.5)),
-       mid = dnorm(x, mean = 5, sd = 5),
-       warm = dgamma(x, shape = 56, rate = 2.8)) %>%
-  pivot_longer(-c(x)) %>%
-  ggplot() +
-  geom_area(aes(x, value, color = name, fill = name), alpha = 0.2,
-            show.legend = FALSE, position = 'identity') +
-  scale_fill_highcontrast() +
-  scale_color_highcontrast() +
-  scale_x_continuous(paste0('Temperature \U00B0', 'C'), expand = c(0, 0)) +
-  scale_y_continuous('Fitness', breaks = NULL, limits = c(0, 0.16), expand = c(0, 0))
+Recent changes in climate have affected mammals' movement rates and habitat selection [@deb_modelling_2020], but it is still unclear how changes in the current century will affect their fine-scale movement behavior and spatial distribution by the end of the century.
+In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures and severe forest fires [@intergovernmental_panel_on_climate_change_climate_2023; @zurowski_summer_2023].
+Over the next 100 years, these changes will continue to affect mammals' fitness, movement behavior, and habitat selection as they cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and frequency and intensity of extreme events [@bunnell_global_2011] along with anthopogenic pressure [@sih_evolution_2011; @weststrate_how_2024].
+
+<!-- implications for ecology: encounter rates, energetic expenditure, range shifts, consequences for conservation, etc -->
+
+Understanding animals' responses to temperature is essential for anticipating their response to climate change in the coming decades and the cascade of consequences such changes will have.
+In light of the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [@ref], we must understand which areas are best preserved to maximize the project's effectiveness.
+Predicting how climate change will affect how, when, and where animals will move is necessary for choosing which 30% of the world is most worth preserving.
+Changes in movement behavior impact several aspects of animals' fitness, behavior, and odds of survival, including encounter rates [with food, competitors, and predators: @martinez-garcia_how_2020], energetic expenditure [@taylor_energetics_1982], range shifts [@woo-durand_increasing_2020].
+Thus, understanding the effects of temperature on animals' movement behavior and their consequences is crucial for effective conservation.
+
+<!-- introduce the study -->
+
+This paper provides an analysis of the effects of ambient temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (*Canis lupus*, *Cervus canadensis*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, and *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
+Using over two decades of telemetry data over a large spatial range of Biritish Columbia (Fig.
+\@ref(fig:bc-map)) and hourly weather data, we estimate how mammals altered their movement frequency, movement speed, and habitat selection in response to air temperature.
+We then pair the estimated responses with climate change projections to forecast changes in mammalian movement in 2100 under different climate-change scenarios [Shared Socioeconomic Pathways, see @riahi_shared_2017].
+We discuss the consequences of these changes in movement behavior with regards to to energetics, encounter rates, habitat conservation, and human-wildlife conflicts.
+
+```{r bc-map, fig.cap="GPS telemetry data and 95% autocorrelated kernel density utilization distributions for the six species in this study."}
+knitr::include_graphics('../figures/tels-map.png')
 ```
 
-During the last century, changes in climate in British Columbia (BC), Canada, have resulted in widespread warming, milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@eberle_technical_2022].
-These changes are expected continue over the next 100 years and compound their effects on mammals' fitness, movement behavior, and habitat selection as they cope with growing heat stress and human pressure [@deb_modelling_2020; @woo-durand_increasing_2020; @weststrate_how_2024].
-
-This paper provides a BC-focused analysis on the effects of proximate weather and long-term climate on the movement of terrestrial mammal species.
-Using over two decades of telemetry data and proximal hourly weather data, we estimate how animals altered their movement frequency, movement speed, and habitat selection in response to air temperature.
-We then pair these responses with climate change projections to forecast changes mammals movement in the current century under different climate change scenarios [@riahi_shared_2017].
-The analyses include data from six species (*Canis lupus*, *Cervus elaphus*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, and *Ursus arctos horribilis*; see Table \@ref(tab:data-table)) across a large spatial range of British Columbia (Fig.
-\@ref(fig:bc-map)).
-We discuss the consequences of changes in mammalian movement in relationship to energetics, food encounter rates, and the incidences of human-wildlife.
-
-```{r bc-map, fig.cap="GPS telemetry data and autocorrelated kernel density utilization distributions for the six species included in this study."}
-knitr::include_graphics('../figures/tels-and-uds-map.png')
-```
+\footnotesize
 
@@
 readRDS('../models/movement-models-2024-06-06.rds') %>%
-  transmute(Dataset = if_else(grepl('Puma', dataset_name),
-                                'Puma concolor', dataset_name) %>%
-              stringr::str_replace_all('_', ' ') %>%
-              stringr::str_replace_all(' boreal', ' (boreal)') %>%
-              stringr::str_replace_all(' southern mountain',
-                                       ' (southern mountain)'),
-            animal,
+  transmute(
+    Dataset = case_when(
+      dataset_name == 'Canis_lupus_boreal' ~ '\\emph{Canis lupus}',
+      dataset_name == 'Rangifer_tarandus_boreal' ~ '\\emph{Rangifer tarandus} (boreal)',
+      dataset_name == 'Rangifer_tarandus_southern_mountain' ~ '\\emph{Rangifer tarandus} (southern mountain)',
+      dataset_name == 'Puma_concolor_2' ~ '\\emph{Puma concolor}',
+      dataset_name == 'Puma_concolor_4' ~ '\\emph{Puma concolor}',
+      dataset_name == 'Elk in southwestern Alberta' ~ '\\emph{Cervus canadensis}',
+      dataset_name == 'Oreamnos_americanus' ~ '\\emph{Oreamnos americanus}',
+      dataset_name == 'Ursus_arctos_horribilis' ~ '\\emph{Ursus arctos horribilis}'),
+    animal, # keep animal column for grouping later
     tel = purrr::map(tel, data.frame)) %>%
@@
   mutate(dt = round('hours' %#% dt, 2)) %>%
-  rename(`Median sampling interval (hours)` = dt) %>%
-  knitr::kable(format = 'pipe', caption = 'Start and end of the GPS telemetries along with the median sampling interval for each of the six species included in this study.') %>%
-  kableExtra::column_spec(1, italic = TRUE) %>%
-  kableExtra::kable_styling(font_size = 10)
+  rename('Median \U0394t (hours)' = dt) %>%
+  knitr::kable(format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$) for each of the species in this study.', escape = FALSE)
 ```
+
+\normalsize
 
@@
 
-Elk (*Cervus elaphus*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data/](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data/)), and the mountain goat (*Oreamnos americanus*) locations were provided by BC Parks.
+\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)), and the mountain goat (*Oreamnos americanus*) locations were provided by BC Parks.
 All other tracking data were obtained from private collaborators.
@@
 
-Rasters of hourly reanalyzed air temperature data were downloaded from the ERA5 dataset [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)).
+\noindent Rasters of hourly reanalyzed air temperature data were downloaded from the ERA5 dataset [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)).
 Proximal weather conditions were estimated for each location by extracting the values from the corresponding raster cell from the temporally nearest raster using the `extract()` function from the `terra` package [version 1.7-46, @hijmans_terra_2023] for `R`  [@r_core_team_r_2024].
@@
 
-We estimated modeled each animal's movement using continuous-time movement models [@fleming_fine-scale_2014] via the `ctmm` package [version 1.2.0, @fleming_ctmm_2023] for `R`.
-We then estimated mammals' instantaneouus speeds by applying the `ctmm::speeds()` function on all models with finite speed estimates [416 of 433, see @fleming_fine-scale_2014; @noonan_scale-insensitive_2019].
-All 17 telemetries with insufficiently fine sampling were for caribou (15 boreal and 2 southern mountain).
-Since `ctmm`'s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), speeds were set to zero if they were not significantly different from zero ($\alpha = 0.05$), under the assumption of a $\chi$-distributed speeds.
-***NOT SURE IF THIS IS CORRECT*** The function used is available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R)).
-To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not by applying the `stats::kmeans()` function with 2 centers (Appendix A).
+\noindent We modeled each animal's movement using continuous-time movement models [@fleming_fine-scale_2014] via the `ctmm` package [version 1.2.0, @fleming_ctmm_2023] for `R`.
+We then estimated mammals' instantaneous speeds by applying the `ctmm::speeds()` function on all models with finite speed estimates [416 of 433, see @fleming_fine-scale_2014; @noonan_scale-insensitive_2019].
+The 17 telemetries with insufficiently fine sampling were for caribou (15 boreal and 2 southern mountain; see Table \@ref(tab:data-table)).
+
+Since `ctmm`'s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed could be 0.
+We performed this correction by subtracting the data-independent, background minimum speed while assuming speeds were $\chi^2$-distributed and correcting the degrees of freedom and confidence intervals accordingly.
+The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R).
+To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using the inflection point of each species' histogram of detrended speed estimates (Appendix A).
 
@@
 
-We estimated the effects of temperature on mammals' state (moving or not) and speed (given that it was moving) using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019] via the `mgcv` package for `R` [version 1.9-1, @wood_generalized_2017].
-The first HGAM had a binomial family of distribution and estimated the probability of a mammal moving as a function of time of day, day of year, and temperature, with separate smooths (but common smoothness parameter) for each species [model S in Figure 4 of @pedersen_hierarchical_2019].
-In addition, the model also included a random effect for each animal, a tensor interaction product of time of day and day of year to account for changes in day length, and a tensor interaction product of time of day and temperature to account for changes in daily activity levels with temperature, for each species.
-Due to the substantial latitudinal separation between boreal and southern mountain caribou, we considered the two gropus as separate species for the sake of modeling.
-The model accounted for the cyclicity of time of day and day of year using cyclic cubic splines, which provide continuity in the smooth term by ensuring the function and its slope (first derivative) and curvature (second derivative) are continuous between consecutive cycles.
-The second HGAM had the same structure, but it used a Gamma family of distributions to estimate the speed of an animal, given that it was moving.
-Together, the two HGAMs inform us on an animal's long-term speed (i.e., distance traveled per unit time), since it is the product of probability that the animal is moving and its average speed when moving.
+### Movement frequency and speed
+
+\noindent We estimated the effects of temperature on mammals' state (moving or not) and speed (given the animal was moving) using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019] with the `mgcv` package for `R` [version 1.9-1, @wood_generalized_2017].
+The first HGAM estimated the probability that an animal was moving with a binomial family of distributions.
+The second HGAM estimated an animal's speed (given that it was moving) with a gamma Location-Scale family of distributions [`mgcv::gammals()`, @wood_smoothing_2016].
+Unlike the more commonly used gamma family (`stats::Gamma()`), the `mgcv::gammals()` family allows the mean-variance relationship to depend on covariates rather than assuming a constant mean-variance relationship.
+For example, if mean speed increases with temperature but the scale decreases, the variance in speed will not increase as much as with a constant scale parameter.
+The animal would then have low average speed but a (relatively) large variance at cold temperatures, and a high average speed with a (relatively) small variance at high temperatures.
+
+The binomial and gamma HGAMs (see code chunk below) included random intercepts for each animal (`s(..., bs = 're')`), fixed intercepts for each species, and factor smooth interactions for time of day, day of year, and temperature for each species [`s(..., bs = 'fs')`, see model S in Figure 4 of @pedersen_hierarchical_2019].
+Additionally, the models had three tensor product interaction terms (`mgcv::ti()`) for each species: (1) day of year and time of day, (2) temperature and time of day, and (3) temperature and day of year.
+These three terms accounted for: (1) seasonal changes in day length, (2) changes in the response to temperature at the hourly scale (e.g., changes in nocturnality), and (3) seasonal changes in the response to temperature (e.g., due to differences in coats), including changes in the timing of migratory behavior.
+The HGAMs accounted for the cyclicity of time of day and day of year using cyclic cubic splines [`bs = 'cc'`, @wood_generalized_2017].
+Together, the binomial HGAM and the gamma HGAM inform us on an animal's long-term average speed, since it is the product of the probability the animal is moving and its average speed when moving.
 Additional details are provided in Appendix A.
 
-We estimated the effects of temperature on each species' selection for percent forest cover, elevation, and distance from fresh water using an HGAM with a Poisson family of distributions [Appendix B, also see @aarts_estimating_2008; @alston_mitigating_2022].
-While we recognize there are other important drivers of habitat selection, we decided to only use these three proxies to ensure results were relatively comparable between species and to make province-wide predictions simpler.
+\singlespacing
+
+\small
+
+```{r hgams, eval=FALSE, echo=TRUE}
+m_1 <- bam(
+  moving ~
+    s(animal, bs = 're') +
+    species +
+    s(tod_pdt, by = species, k = 5, bs = 'cc') +
+    s(doy, by = species, k = 5, bs = 'cc') +
+    s(temp_c, by = species, k = 5, bs = 'tp') +
+    ti(doy, tod_pdt, by = species, k = 5, bs = c('cc', 'cc')) +
+    ti(temp_c, tod_pdt, by = species, k = 5, bs = c('tp', 'cc')) +
+    ti(temp_c, doy, by = species, k = 5, bs = c('tp', 'cc')) +
+    s(log(dt), k = 3) +
+    s(log(dt), species, k = 3, bs = 'fs'),
+  data = d,
+  method = 'fREML', # fast REML
+  discrete = TRUE, # discretize the covariates for faster computation
+  knots = list(tod_pdt = c(0, 1), doy = c(0.5, 366.5))) # for bs = 'cc'
+
+m_2 <- bam(
+  speed_est ~
+    s(animal, bs = 're') +
+    species +
+    s(tod_pdt, by = species, k = 5, bs = 'cc') +
+    s(doy, by = species, k = 5, bs = 'cc') +
+    s(temp_c, by = species, k = 5, bs = 'tp') +
+    ti(doy, tod_pdt, by = species, k = 5, bs = c('cc', 'cc')) +
+    ti(temp_c, tod_pdt, by = species, k = 5, bs = c('tp', 'cc')) +
+    ti(temp_c, doy, by = species, k = 5, bs = c('tp', 'cc')) +
+    s(log(dt), k = 3) +
+    s(log(dt), species, k = 3, bs = 'fs'),
+  family = Gamma(link = 'log'),
+  data = d_2, # data filtered for speed > 0
+  method = 'fREML',
+  discrete = TRUE,
+  knots = list(tod_pdt = c(0, 1), doy = c(0.5, 366.5)))
+```
+
+\normalsize
+
+\doublespacing
+
+### Habitat selection
+
+\noindent We estimated the effects of temperature on each species' selection for percent forest cover, elevation, and distance from fresh water (i.e., not sea water) using a series of HGAMs with Poisson families of distributions [Appendix B, also see @aarts_estimating_2008; @alston_mitigating_2022].
+While we recognize there are other important drivers of habitat selection, we decided to only use these three proxies to produce results that are relatively comparable between species and to make province-wide predictions simpler.
+Each species' model had the same structure:
+
+\singlespacing
+
+\small
+
+```{r rsf, eval=FALSE, echo=TRUE}
+rsf <- bam(
+  detected ~ # 1 for GPS locations, 0 for quadrature data
+    # species-level average resource preference
+    s(forest_perc, k = 6, bs = 'tp') +
+    s(elevation_m, k = 6, bs = 'tp') +
+    s(dist_water_m, k = 6, bs = 'tp') +
+    # animal-level deviations from the species-level average
+    s(animal, bs = 're') +
+    s(forest_perc, animal, k = 6, bs = 'fs', xt = list(bc = 'cr')) +
+    s(elevation_m, animal, k = 6, bs = 'fs', xt = list(bc = 'cr')) +
+    s(dist_water_m, animal, k = 6, bs = 'fs', xt = list(bc = 'cr')) +
+    # changes in preference with temperature
+    ti(forest_perc, temperature_C, k = 4, bs = 'cr') +
+    ti(elevation_m, temperature_C, k = 4, bs = 'cr') +
+    ti(dist_water_m, temperature_C, k = 4, bs = 'cr'),
+  family = poisson(link = 'log'),
+  data = d,
+  weights = weight, # based on AKDE
+  method = 'fREML',
+  discrete = TRUE)
+```
+
+\normalsize
+
+\doublespacing
+
+Smooth effects of forest cover, elevation, and distance to fresh water accounted for the species-level average selection strength for each resource.
+A random effect for each individual animal corrected for uneven sampling across individuals, while factor smooth interaction terms (`bs = 'fs'`) accounted for individual level resource selection (i.e., individual-level deviations from the species-level average).
+Finally, tensor interaction product terms (`ti()`) of each resource and temperature estimated the effect of temperature on the selection strength for each resource.
+Each model used a Poisson distribution and a $\log$ link function because detected is the number of animals (0 or 1) within a small, finite amount of space and time.
+Detections were down-weighted proportionally to their dependence on temporally proximate detections [see Appendix C and @alston_mitigating_2022], while quadrature points had a weight of 1.
+Finally, we fit each model with fast REML (`'fREML'`) and discretized covariates (`discrete = TRUE`) to optizimize computational efficiency with no meaningful losses to model performance [@wood_generalized_2015; @wood_generalized_2017-1; @li_faster_2020].
 
@@
 
-***HERE***
+***check this***
 
-Rasters of projected monthly average temperature in BC during the next 80 years were obtained via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`.
-Since the climate projections did not provide a measure of variation in temperature within each month due to weather, we estimated within-month variance by fitting a Generalized Additive Model with a Gamma family of distributions to the square-root precision ($\sqrt{\rho}$) in temperature for each month using the ERA5 data for BC during 1998-2024.
-The model included...
-(Appendix C).
+\noindent Rasters of projected monthly average temperature in BC from 2020 to 2100 were obtained via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`.
+Since the climate projections did not provide a measure of variation in temperature within each month, we estimated within-month variance by fitting a GAM for location and scale [GAMLS, see @rigby_generalized_2005; @stasinopoulos_generalized_2007] to the monthly variance in temperature using the hourly ERA5 data for BC during from 1998 to 2023 [extremes included, see @hersbach_era5_2023].
+The model included smooth terms of the monthly mean temperature, month, and year as well as and a two-dimensional smooth of space and a tensor interaction product term of space and month (Appendix C).
 
-***model currently uses weights = sqrt(n); fitting a new one with uniform weights***
+```{r, eval=FALSE}
+'Gamma GAMLS code'
+```
+
+We accounted for hourly variation in weather by assuming temperature followed a normal distribution with mean specified by the `climatenaR` climate projectionS and variance as specified by the Gamma GAMLS.
+We then predicted changes in movement behavior and habitat selection as a function of temperature using the HGAMs and HRSFs along with the temperature HGAM described above.
+For each month within each year from 2020 to 2100 we simulated hourly weather by including temperatures from the 5%, 10%, 20%, ..., 80%, 90%, and 95% (***?***) quantiles and weighted each quantile proportionally to the Gaussian probability density for each value (Appendix C).
 
@@
 
-### Effects of temperature on mammals' movement
+## Effects of temperature on movement rates
 
-### Effects of temperature on mammals' habitat selection
+\noindent Partial effect of temperature varied across species (Fig.
+***AX***), even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; see Fig.
+***AX***).
+Smooth interaction terms were well-behaved and indicated clear shifts in daily activity for all species.
+The models explained reasonably high proportions of the deviance (11% for the binomial model and 79% for the Gamma model) and had good in-sample prediction (Fig.
+***AX***).
 
-***should predict for the area the animals are in only***
+```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals' travel.} Estimated effects of temperature on relative change in distance travelled (probability of moving times speed when moving) over time of day (\\textbf{A}) and day of year (\\textbf{B}).
+Predictions in the surface plots extend to 10% of the range away from each datum.
+The color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to -75\\% and +300\\% for ease of readability."}
+knitr::include_graphics('../figures/distance.png')
+```
+
+Both movement frequency and speed when moving decreased with increasing sampling interval, with a strong common effect (Fig. ***AX***). ***here***
+
+`ctmm` models are scale independent assuming single, homogeneous behavior, but scale-dependency in the behavior (data) causes sampling effects
+
+worse climate-change scenarios corresponded to greater changes in behavior along with greater variance in within years
+
+### Predicted changes in movement during the current century
+
+```{r t-dist, fig.cap="\\textbf{The intensity and direction of changes in movement rates due to climate change vary among species, but worse climate-change scenarios consistently result in the greatest change and greatest uncertainty.} Lines indicate the projected relative change in mean distance travelled (probability of moving times speed when moving) due to changes in temperature within the species' current extent.
+Shaded areas indicate the mean 95% Bayesian credible intervals.
+Changes are relative to the mean distances travelled in 2020 across the four Shared Socioeconomic Pathways (SSPs).
+The SSPs are listed in order of similarity to current conditions, from SSP 1-2.6 (sustainability, best-case scenario) to 5-8.5 (fossil-fueled development, worst-case scenario).
+The projections only account for changes movement behavior (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
+knitr::include_graphics('../figures/distance-travelled-local-cc-predictions.png')
+```
+
+```{r s-dist}
+plot(1, main = '4. spatial projections of distance in 2020 and 2100')
+```
+
+## Effects of temperature on habitat selection
+
+```{r hrsfs, fig.cap="\\textbf{Central message.} Estimated resourse selection strength for forest cover (\\%), elevation (km), and distance from fresh water (km) as a function of temperature.
+The grey contours indicate the extent of each speices' observed locations."}
+knitr::include_graphics('../figures/rsf-surface-plots.png')
+```
+
+### Predicted changes in movement during the current century
+
+```{r t-hrsfs, fig.cap="\\textbf{The intensity and direction of changes in habitat selection due to climate change vary among species, but worse climate-change scenarios consistently result in the greatest change and greatest uncertainty.} Lines indicate the projected change in relative selection strength for forest cover (\\%), elevation (km), and distance from fresh water (km) due to changes in temperature within the species' current extent.
+Shaded areas indicate the mean 95% Bayesian credible intervals.
+Changes are relative to the relative selection strength in 2020 across the four Shared Socioeconomic Pathways (SSPs).
+The SSPs are listed in order of similarity to current conditions, from SSP 1-2.6 (sustainability, best-case scenario) to 5-8.5 (fossil-fueled development, worst-case scenario)."}
+plot(1, main = '6. temporal projections of RSFs within current area')
+```
+
+```{r s-hrsfs}
+plot(1, main = '7. spatial projections of RSFs in 2020 and 2100')
+```
 
@@
 
+* as temperatures get warmer, we will be pushing into the unknown portion of the surface plots
+* our models don’t include any physiological responses to temperature that may impact movement (e.g., hybernation)
+* our models also assume the behaviour we modeled now will persist into the future (e.g., hybernation, migration)
+* based on the model, temperature doesn't affect grizzlies' hibernation much, and their displacement in winter is very low due to P(moving) being so low:
+* if the RSS decreases a lot, individuals may disperse, populations may relocate, or stay where they are and suffer
+
 <!-- - how results answer central question -->
 
-### Effects of temperature on mammals' movement
+## Effects of temperature on mammals' movement
 
-### Effects of temperature on mammals' habitat selection
+## Effects of temperature on mammals' habitat selection
 
 <!-- - what it tells us about the broader field -->
+## Consequences for conservation
 
+* changes in temperatures are pushing to areas in the surface plots that have little data and may be poorly estimated.
+We will need to extrapolate to prepare for future changes, which implies some degree of inevitable risk.
+The best wat to avoid this is by designing models that are statistically sound and interpretable.
+The models we present provide an estimate of the changes in behavior, but they are not tailored specifically to any of the species in the dataset.
+Species-specific models should account for more particular variables (e.g., forest age for caribou) and use dynamic resource rasters (i.e., not use a static estimate of forest cover).
+Nonetheless, we present important results for understanding how mammals will respond to changes in climate and weather during the next ~75 years.
 * energetics
+* GPS sampling intervals were large: need more work on fine-scale movement, including accounting for multiple movement sates (e.g., traveling, grazing, resting, hunting, etc.).
+this would help deal with over-dispersion
 * encounter rates (lower encounter rates with food [@hou_cold_2020])
@@
 
-# References
+<!-- additional references:
+
+- boars' sleep quantity, efficiency and quality are reduced on warmer days (https://doi.org/10.1098/rspb.2023.2115)
+
+-->
+
+# References {-}
 

### 2024-09-13 analysis/10-local-weather-predictions-by-species.R
removed unnecessary spaces

@@ -124,8 +124,6 @@ unique_locs_spp <- mutate(
   select(! akdes) %>%
   unnest(subset)
 
-
-
 ggplot(unique_locs_spp, aes(long, lat)) +
   facet_wrap(~ species, scales = 'free') +
   geom_point() +

### 2024-09-13 writing/manuscript.Rmd
reviewed and updated methods

@@
 
-<!-- elements of the title page -->
+\clearpage
 
@@
 This paper provides an analysis of the effects of ambient temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (*Canis lupus*, *Cervus canadensis*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, and *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
-Using over two decades of telemetry data over a large spatial range of Biritish Columbia (Fig.
-\@ref(fig:bc-map)) and hourly weather data, we estimate how mammals altered their movement frequency, movement speed, and habitat selection in response to air temperature.
+Using over two decades of telemetry data over a large spatial range of British Columbia (Fig.
+\@ref(fig:bc-map)) and hourly air temperature data, we estimate how mammals altered their movement frequency, movement speed, and habitat selection in response to air temperature.
 We then pair the estimated responses with climate change projections to forecast changes in mammalian movement in 2100 under different climate-change scenarios [Shared Socioeconomic Pathways, see @riahi_shared_2017].
-We discuss the consequences of these changes in movement behavior with regards to to energetics, encounter rates, habitat conservation, and human-wildlife conflicts.
+We discuss the consequences of these changes in movement behavior with regards to energetics, encounter rates, habitat conservation, and human-wildlife conflicts.
 
@@
 
+```{r, eval=FALSE}
+# a quick check for the table below
+readRDS('../data/movement-models-speed-weights-2024-06-10.rds') %>%
+    group_by(dataset_name, animal) %>%
+    summarize(any_na_speed = any(is.na(speed_est)),
+              all_na_speed = all(is.na(speed_est))) %>%
+    group_by(dataset_name) %>%
+    summarise(n_missing_any = sum(any_na_speed),
+              n_missing_all = sum(all_na_speed))
+```
+
 ```{r data-table}
@@
       dataset_name == 'Rangifer_tarandus_boreal' ~ '\\emph{Rangifer tarandus} (boreal)',
-      dataset_name == 'Rangifer_tarandus_southern_mountain' ~ '\\emph{Rangifer tarandus} (southern mountain)',
+      dataset_name == 'Rangifer_tarandus_southern_mountain' ~ '\\emph{Rangifer tarandus} (s. mountain)',
       dataset_name == 'Puma_concolor_2' ~ '\\emph{Puma concolor}',
@@
     animal, # keep animal column for grouping later
-    tel = purrr::map(tel, data.frame)) %>%
+    tel = map(tel, data.frame),
+    na_speed = map_lgl(movement_model,
+                       \(.m) suppressWarnings(speed(.m)$DOF == 0))) %>%
   unnest(tel) %>%
-  group_by(animal) %>%
-  mutate(dt = timestamp - lag(timestamp)) %>%
-  group_by(Dataset) %>%
+  group_by(Dataset, animal) %>%
   summarize(Start = min(as.Date(timestamp)),
             End = max(as.Date(timestamp)),
-            dt = as.numeric(median(dt, na.rm = TRUE))) %>%
-  mutate(dt = round('hours' %#% dt, 2)) %>%
-  rename('Median \U0394t (hours)' = dt) %>%
-  knitr::kable(format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$) for each of the species in this study.', escape = FALSE)
+            dt = median(timestamp - lag(timestamp), na.rm = TRUE),
+            na_speed = unique(na_speed),
+            .groups = 'drop') %>%
+  group_by(Dataset) %>%
+  summarize(Start = min(Start),
+            End = max(End),
+            dt = as.numeric(median(dt, na.rm = TRUE)),
+            n_missing = sum(na_speed),
+            .groups = 'drop') %>%
+  mutate(dt = round('hours' %#% dt, 2)) %>% # convert from seconds to hours
+  rename('Median \U0394t (hours)' = dt,
+         'NA speed' = n_missing) %>%
+  knitr::kable(format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$; stratified by individual) and number of animals with unresolved speed (NA speed) for each of the species in this study.', escape = FALSE)
 ```
@@
 Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and high straight-line displacement, especially if antecedent and subsequent points indicated stationary behavior.
-The script used to clean the data and all custom functions used are available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature)).
+The script used to clean the data and all associated custom functions are available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature)).
 
@@
 \noindent Rasters of hourly reanalyzed air temperature data were downloaded from the ERA5 dataset [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)).
-Proximal weather conditions were estimated for each location by extracting the values from the corresponding raster cell from the temporally nearest raster using the `extract()` function from the `terra` package [version 1.7-46, @hijmans_terra_2023] for `R`  [@r_core_team_r_2024].
+Proximal temperature was estimated for each location by extracting the values from the corresponding raster cell from the temporally nearest raster using the `extract()` function from the `terra` package [version 1.7-46, @hijmans_terra_2023] for `R`  [@r_core_team_r_2024].
 
@@
 We then estimated mammals' instantaneous speeds by applying the `ctmm::speeds()` function on all models with finite speed estimates [416 of 433, see @fleming_fine-scale_2014; @noonan_scale-insensitive_2019].
-The 17 telemetries with insufficiently fine sampling were for caribou (15 boreal and 2 southern mountain; see Table \@ref(tab:data-table)).
+The 18 telemetries with insufficiently fine sampling were for one mountain goat, 15 boreal caribou, and 2 southern mountain caribou (Table \@ref(tab:data-table).
 
 Since `ctmm`'s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed could be 0.
-We performed this correction by subtracting the data-independent, background minimum speed while assuming speeds were $\chi^2$-distributed and correcting the degrees of freedom and confidence intervals accordingly.
-The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R).
+We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed.<!-- The degrees of freedom and confidence intervals were adjusted accordingly.--> The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R).
 To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using the inflection point of each species' histogram of detrended speed estimates (Appendix A).
@@
 
-\noindent We estimated the effects of temperature on mammals' state (moving or not) and speed (given the animal was moving) using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019] with the `mgcv` package for `R` [version 1.9-1, @wood_generalized_2017].
-The first HGAM estimated the probability that an animal was moving with a binomial family of distributions.
-The second HGAM estimated an animal's speed (given that it was moving) with a gamma Location-Scale family of distributions [`mgcv::gammals()`, @wood_smoothing_2016].
-Unlike the more commonly used gamma family (`stats::Gamma()`), the `mgcv::gammals()` family allows the mean-variance relationship to depend on covariates rather than assuming a constant mean-variance relationship.
-For example, if mean speed increases with temperature but the scale decreases, the variance in speed will not increase as much as with a constant scale parameter.
-The animal would then have low average speed but a (relatively) large variance at cold temperatures, and a high average speed with a (relatively) small variance at high temperatures.
-
-The binomial and gamma HGAMs (see code chunk below) included random intercepts for each animal (`s(..., bs = 're')`), fixed intercepts for each species, and factor smooth interactions for time of day, day of year, and temperature for each species [`s(..., bs = 'fs')`, see model S in Figure 4 of @pedersen_hierarchical_2019].
-Additionally, the models had three tensor product interaction terms (`mgcv::ti()`) for each species: (1) day of year and time of day, (2) temperature and time of day, and (3) temperature and day of year.
-These three terms accounted for: (1) seasonal changes in day length, (2) changes in the response to temperature at the hourly scale (e.g., changes in nocturnality), and (3) seasonal changes in the response to temperature (e.g., due to differences in coats), including changes in the timing of migratory behavior.
+\noindent We estimated the effects of temperature on mammals' movement state (moving or not) and speed (when moving) using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019 and the code chunk below] with the `mgcv` package for `R` [version 1.9-1, @wood_generalized_2017].
+The first HGAM estimated the probability that an animal was moving with a binomial family of distributions and logit link function.
+The second HGAM estimated an animal's speed (when moving) with a gamma family of distributions and log link function.
+The HGAMs included random intercepts for each animal (`s(..., bs = 're')`), fixed intercepts for each species, and species-level factor smooth interactions for time of day, day of year, and temperature [`s(..., bs = 'fs')`, see model S in Figure 4 of @pedersen_hierarchical_2019].
+Additionally, the models had three tensor product interaction terms (`ti()`) for each species: (1) day of year and time of day, (2) temperature and time of day, and (3) temperature and day of year.
+These three terms accounted for: (1) seasonal changes in day length, (2) hourly changes in the response to temperature (e.g., changes in nocturnality), and (3) seasonal changes in the response to temperature (e.g., changes in coats and migration timing).
 The HGAMs accounted for the cyclicity of time of day and day of year using cyclic cubic splines [`bs = 'cc'`, @wood_generalized_2017].
 Together, the binomial HGAM and the gamma HGAM inform us on an animal's long-term average speed, since it is the product of the probability the animal is moving and its average speed when moving.
+We fit the models with fast REML (`'fREML'`) and discretized covariates (`discrete = TRUE`) to optimize computational efficiency with no appreciable losses to model performance [@wood_generalized_2015; @wood_generalized_2017-1; @li_faster_2020].
 Additional details are provided in Appendix A.
@@
 ```{r hgams, eval=FALSE, echo=TRUE}
-m_1 <- bam(
+m_moving <- bam(
   moving ~
     s(animal, bs = 're') +
-    species +
+    species + #' necessary since `by` terms do not include intercept terms
     s(tod_pdt, by = species, k = 5, bs = 'cc') +
@@
 
-m_2 <- bam(
+m_speed <- bam(
   speed_est ~
@@
 
-\noindent We estimated the effects of temperature on each species' selection for percent forest cover, elevation, and distance from fresh water (i.e., not sea water) using a series of HGAMs with Poisson families of distributions [Appendix B, also see @aarts_estimating_2008; @alston_mitigating_2022].
-While we recognize there are other important drivers of habitat selection, we decided to only use these three proxies to produce results that are relatively comparable between species and to make province-wide predictions simpler.
+\noindent We estimated the effects of temperature on each species' selection for percent forest cover, elevation, and distance from fresh water (i.e., not sea water) using an HGAM with a Poisson family of distributions and log link function for each species [Appendix B, also see @aarts_estimating_2008; @alston_mitigating_2022].
+While we recognize there are other important drivers of habitat selection that we did not include in our models (e.g., forest age), we decided to only use these three proxies to produce results that are relatively comparable between species and to make province-wide predictions simpler.
 Each species' model had the same structure:
@@
 
-Smooth effects of forest cover, elevation, and distance to fresh water accounted for the species-level average selection strength for each resource.
-A random effect for each individual animal corrected for uneven sampling across individuals, while factor smooth interaction terms (`bs = 'fs'`) accounted for individual level resource selection (i.e., individual-level deviations from the species-level average).
-Finally, tensor interaction product terms (`ti()`) of each resource and temperature estimated the effect of temperature on the selection strength for each resource.
-Each model used a Poisson distribution and a $\log$ link function because detected is the number of animals (0 or 1) within a small, finite amount of space and time.
-Detections were down-weighted proportionally to their dependence on temporally proximate detections [see Appendix C and @alston_mitigating_2022], while quadrature points had a weight of 1.
-Finally, we fit each model with fast REML (`'fREML'`) and discretized covariates (`discrete = TRUE`) to optizimize computational efficiency with no meaningful losses to model performance [@wood_generalized_2015; @wood_generalized_2017-1; @li_faster_2020].
+Smooth effects of percent forest cover, elevation, and distance to fresh water accounted for the species-level average selection strength for each resource.
+A random effect for each individual animal corrected for uneven sampling across individuals, while factor smooth interaction terms (`bs = 'fs'`) accounted for individual-level resource selection (i.e., individual-level deviations from the species-level average).
+Finally, tensor interaction product terms (`ti()`) of the three resources and temperature estimated the change in resource selection at different temperatures.
+Detections were down-weighted proportionally to their degree of dependence temporally proximate detections [see Appendix C and @alston_mitigating_2022], while quadrature points had a weight of 1.
 
@@
 
-***check this***
+\noindent Rasters of projected monthly average temperature in BC from 2020 to 2100 were obtained via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`.
+Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 data for BC from 1998 to 2023 [extremes included, see @hersbach_era5_2023] to calculate within-month variance, which we defined as the variance within a given pixel, month, and year.
+We then modelled the estimated variance using a GAM for location and scale [GAMLS, see @rigby_generalized_2005; @stasinopoulos_generalized_2007].
+The linear predictor for the location (i.e., the mean) included smooth terms of the within-pixel monthly mean temperature (for each year), month (as a cyclic smooth), a two-dimensional smooth of space, and a tensor interaction product term of space and month.
+The linear predictor for the scale term (which governs the mean-variance relationship) included smooth terms of the monthly mean, month, and space.
+Additional details are available in Appendix C.
 
-\noindent Rasters of projected monthly average temperature in BC from 2020 to 2100 were obtained via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`.
-Since the climate projections did not provide a measure of variation in temperature within each month, we estimated within-month variance by fitting a GAM for location and scale [GAMLS, see @rigby_generalized_2005; @stasinopoulos_generalized_2007] to the monthly variance in temperature using the hourly ERA5 data for BC during from 1998 to 2023 [extremes included, see @hersbach_era5_2023].
-The model included smooth terms of the monthly mean temperature, month, and year as well as and a two-dimensional smooth of space and a tensor interaction product term of space and month (Appendix C).
-
-```{r, eval=FALSE}
-'Gamma GAMLS code'
+```{r, eval=FALSE, echo=FALSE}
+  m_gammals <- gam(list(
+    monthly_var ~
+      s(monthly_mean, bs = 'cr', k = 10) + # using cr to reduce fitting time
+      s(month, bs = 'cc', k = 10) +
+      s(long, lat, bs = 'ds', k = 500) +
+      ti(long, lat, month, bs = c('ds', 'cc'), d = c(2, 1), k = c(100, 10)),
+    ~ s(monthly_mean, bs = 'cr', k = 10) +
+      s(month, bs = 'cc', k = 10) +
+      s(long, lat, bs = 'ds', k = 100)),
+    family = gammals(),
+    data = d,
+    method = 'REML',
+    knots = list(month = c(0.5, 12.5)))
 ```
 
-We accounted for hourly variation in weather by assuming temperature followed a normal distribution with mean specified by the `climatenaR` climate projectionS and variance as specified by the Gamma GAMLS.
+We simulated for hourly variation in weather by assuming temperature followed a normal distribution with mean specified by the `climatenaR` climate projections and variance as specified by the Gamma GAMLS.
 We then predicted changes in movement behavior and habitat selection as a function of temperature using the HGAMs and HRSFs along with the temperature HGAM described above.
-For each month within each year from 2020 to 2100 we simulated hourly weather by including temperatures from the 5%, 10%, 20%, ..., 80%, 90%, and 95% (***?***) quantiles and weighted each quantile proportionally to the Gaussian probability density for each value (Appendix C).
+For each month within each year from 2020 to 2100, we simulated hourly weather by including temperatures from the 0.1, 0.2, ..., 0.8, 0.9 quantiles and weighted each quantile proportionally to the (normalized) Gaussian probability density for each value (Appendix C).
 
@@
 
-`ctmm` models are scale independent assuming single, homogeneous behavior, but scale-dependency in the behavior (data) causes sampling effects
+`ctmm` models are scale independent assuming single, homogeneous behavior, but scale-dependency in the behavior (data) causes sampling effects.
+See poor projections for boreal caribou, which have the greatest sampling interval
 
@@
 
+* We discuss the consequences of these changes in movement behavior with regards to energetics, encounter rates, habitat conservation, and human-wildlife conflicts.
 * as temperatures get warmer, we will be pushing into the unknown portion of the surface plots

### 2024-10-04 writing/manuscript.Rmd
added hists of rates, fig of model agreement, dAIC

@@
 ```{r data-table}
+dt_data <-
   readRDS('../models/movement-models-2024-06-06.rds') %>%
@@
   rename('Median \U0394t (hours)' = dt,
-         'NA speed' = n_missing) %>%
-  knitr::kable(format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$; stratified by individual) and number of animals with unresolved speed (NA speed) for each of the species in this study.', escape = FALSE)
+         'NA speed' = n_missing)
+
+knitr::kable(dt_data, format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$; stratified by individual) and number of animals with unresolved speed (NA speed) for each of the species in this study.', escape = FALSE)
 ```
@@
 We then estimated mammals' instantaneous speeds by applying the `ctmm::speeds()` function on all models with finite speed estimates [416 of 433, see @fleming_fine-scale_2014; @noonan_scale-insensitive_2019].
-The 18 telemetries with insufficiently fine sampling were for one mountain goat, 15 boreal caribou, and 2 southern mountain caribou (Table \@ref(tab:data-table).
+The 18 telemetries with insufficiently fine sampling were for one mountain goat, 15 boreal caribou, and 2 southern mountain caribou (Table \@ref(tab:data-table)).
 
 Since `ctmm`'s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed could be 0.
-We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed.<!-- The degrees of freedom and confidence intervals were adjusted accordingly.--> The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R).
+We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed.<!-- The degrees of freedom and Frequentist confidence intervals were adjusted accordingly.--> The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R).
 To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using the inflection point of each species' histogram of detrended speed estimates (Appendix A).
@@
 ## Estimating the effects of temperature on mammals' movement
+
+The following two sections detail the modelling we used to estimate the effect on temperature on mammals' movement.
+To assess the importance of including temperature as an explicit covariate (rather than including its effects with time of day and day of year***HERE***), we fit models with and without temperature and compared the fits and model predictions of the two sets of models.
 
@@
 Additional details are provided in Appendix A.
+
+\clearpage
 
@@
 
+\clearpage
+
 \normalsize
@@
 
-\noindent We estimated the effects of temperature on each species' selection for percent forest cover, elevation, and distance from fresh water (i.e., not sea water) using an HGAM with a Poisson family of distributions and log link function for each species [Appendix B, also see @aarts_estimating_2008; @alston_mitigating_2022].
+\noindent We estimated the effects of temperature on each species' selection for percent forest cover, elevation, and distance from fresh water (i.e., not sea water) by fitting a Hierarchical Resource Selection Function [@aarts_estimating_2008; @refs; @10.1111/ecog.07225] an HGAM with a Poisson family of distributions and log link function for each species [Appendix B, also see @aarts_estimating_2008; @alston_mitigating_2022].
 While we recognize there are other important drivers of habitat selection that we did not include in our models (e.g., forest age), we decided to only use these three proxies to produce results that are relatively comparable between species and to make province-wide predictions simpler.
@@
     s(animal, bs = 're') +
-    s(forest_perc, animal, k = 6, bs = 'fs', xt = list(bc = 'cr')) +
-    s(elevation_m, animal, k = 6, bs = 'fs', xt = list(bc = 'cr')) +
-    s(dist_water_m, animal, k = 6, bs = 'fs', xt = list(bc = 'cr')) +
+    s(forest_perc, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')) +
+    s(elevation_m, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')) +
+    s(dist_water_m, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')) +
     # changes in preference with temperature
-    ti(forest_perc, temperature_C, k = 4, bs = 'cr') +
-    ti(elevation_m, temperature_C, k = 4, bs = 'cr') +
-    ti(dist_water_m, temperature_C, k = 4, bs = 'cr'),
+    ti(forest_perc, temperature_C, k = 6, bs = 'cr') +
+    ti(elevation_m, temperature_C, k = 6, bs = 'cr') +
+    ti(dist_water_m, temperature_C, k = 6, bs = 'cr'),
   family = poisson(link = 'log'),
@@
 Finally, tensor interaction product terms (`ti()`) of the three resources and temperature estimated the change in resource selection at different temperatures.
-Detections were down-weighted proportionally to their degree of dependence temporally proximate detections [see Appendix C and @alston_mitigating_2022], while quadrature points had a weight of 1.
+Detections were weighted proportionally to their degree of independence from temporally proximate detections [see Appendix C and @alston_mitigating_2022], while quadrature points had a weight of 1.
+
+***does elk HRSF use different `k` values?***
 
@@
 Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 data for BC from 1998 to 2023 [extremes included, see @hersbach_era5_2023] to calculate within-month variance, which we defined as the variance within a given pixel, month, and year.
-We then modelled the estimated variance using a GAM for location and scale [GAMLS, see @rigby_generalized_2005; @stasinopoulos_generalized_2007].
+We then modelled the estimated variance using a GAM for location and scale [GAMLS, see @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017].
 The linear predictor for the location (i.e., the mean) included smooth terms of the within-pixel monthly mean temperature (for each year), month (as a cyclic smooth), a two-dimensional smooth of space, and a tensor interaction product term of space and month.
@@
 
+While adding temp did not improve model fit (add change in dev.expl), but it ... (delta BIC for both modelss) (Fig. ***SX*** pred vs pred)
+
+```{r, eval=FALSE}
+tibble(parameter = c('P(moving)', 'speed'),
+       with_temp = c(10.7, 79.3),
+       without_temp = c(10.2, 79.1),
+       delta_AIC = c(diff(AIC(m_1_no_t, m_1)$AIC),
+                     diff(AIC(m_2_no_t, m_2)$AIC))) %>%
+  knitr::kable()
+
+#' all HRSFs have `delta_AIC` < -85 (check boreal caribou and elk)
+```
+
 ## Effects of temperature on movement rates
 
-\noindent Partial effect of temperature varied across species (Fig.
+\noindent The partial effects of temperature varied in both direction and magnitude across species (Fig.
 ***AX***), even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; see Fig.
 ***AX***).
-Smooth interaction terms were well-behaved and indicated clear shifts in daily activity for all species.
+Smooth interaction terms were well-behaved and indicated clear shifts in daily activity for all species (Fig.
+***AX***).
 The models explained reasonably high proportions of the deviance (11% for the binomial model and 79% for the Gamma model) and had good in-sample prediction (Fig.
 ***AX***).
+Both movement frequency and speed when moving decreased with increasing sampling interval, with a strong common effect (Fig.
+***AX***).
+Most species adapted their daily and seasonal movement behavior to changes in temperature (Fig.
+\@ref(fig:dist)), but the precision of the estimated change depends on the GPS sampling frequency.
+While `ctmm` models are scale independent [@noonan_scale-insensitive_2019], the accuracy, size, and stability of speed estimates still depends on the GPS sampling frequency.
+This is because coarser sampling contains information on large scale movements (e.g., range crossings, migrations) but not fine-scale movement (e.g., first-order habitat selection).
+Using the boreal caribou as an example, the 13-hour sampling interval allows us to reasonably estimate the caribou's movement path at a temporal scale of approximately 13 hours (or greater), but we cannot produce reasonable movement trajectories at a much finer (e.g., hourly) scale.
+Consequently, the estimated movement behaviors should be compared across species carefully.
 
-```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals' travel.} Estimated effects of temperature on relative change in distance travelled (probability of moving times speed when moving) over time of day (\\textbf{A}) and day of year (\\textbf{B}).
+```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals travel.} Estimated effects of temperature on relative change in distance travelled (probability of moving times speed when moving) over time of day (\\textbf{A}) and day of year (\\textbf{B}).
 Predictions in the surface plots extend to 10% of the range away from each datum.
@@
 
-Both movement frequency and speed when moving decreased with increasing sampling interval, with a strong common effect (Fig. ***AX***). ***here***
+### Predicted changes in movement during the current century
 
-`ctmm` models are scale independent assuming single, homogeneous behavior, but scale-dependency in the behavior (data) causes sampling effects.
-See poor projections for boreal caribou, which have the greatest sampling interval
+The direction and degree of change in distance travelled varies greatly between species, but worse climate-change scenarios consistently corresponded to greater absolute changes in behavior as well as greater uncertainty in the estimated change (Fig.
+\@ref(fig:t-dist)).
+Under the best-case scenario, absolute projected changes by 2100 were small (0% to 2%, except for southern mountain caribou, which had a change of ~5%).
+Under the worst-case scenario, absolute projected changes were by 2100 ranged from ~1% (boreal caribou) to ~25% (southern moutain caribou) and had the greatest uncertainty in the estimates.
+Boreal caribou and elk showed relatively little projected changes, especially relative to the degree of inter-annual variability in displacement.
 
-worse climate-change scenarios corresponded to greater changes in behavior along with greater variance in within years
+```{r, eval=FALSE}
+# max absolute change
+readRDS('H:/GitHub/bc-mammals-temperature/data/cc-hgam-projections.rds') %>%
+  group_by(scenario, species) %>%
+  summarize(max_rel = labeller_perc(max(abs(d))),
+            .groups = 'drop') %>%
+  arrange(species, scenario) %>%
+  print(., n = nrow(.))
 
-### Predicted changes in movement during the current century
+# range between values at 2100 does not depend strongly on dt
+# (without accounting for other variables and trends within species)
+readRDS('H:/GitHub/bc-mammals-temperature/data/cc-hgam-projections.rds') %>%
+  filter(year == 2100) %>%
+  group_by(species) %>%
+  summarize(range = labeller_perc(diff(range(d)) + 1)) %>%
+  left_join(dt_data %>%
+              mutate(species = case_when(
+                Dataset == '\\emph{Canis lupus}' ~ 6,
+                Dataset == '\\emph{Rangifer tarandus} (boreal)' ~ 5,
+                Dataset == '\\emph{Rangifer tarandus} (s. mountain)' ~ 1,
+                Dataset == '\\emph{Puma concolor}' ~ 3,
+                Dataset == '\\emph{Puma concolor}' ~ 3,
+                Dataset == '\\emph{Cervus canadensis}' ~ 4,
+                Dataset == '\\emph{Oreamnos americanus}' ~ 7,
+                Dataset == '\\emph{Ursus arctos horribilis}' ~ 2),
+                species = SPECIES_LABS[species]),
+            by = 'species') %>%
+  ggplot(aes(`Median Δt (hours)`, range, color = species)) +
+  geom_point() +
+  scale_color_manual(values = PAL)
+```
 
@@
 
+***COMMENTS ABOUT SPATIAL VARIATION IN DISPLACEMENT***
+
 ```{r s-dist}
@@
 
-```{r hrsfs, fig.cap="\\textbf{Central message.} Estimated resourse selection strength for forest cover (\\%), elevation (km), and distance from fresh water (km) as a function of temperature.
+to do:
+* check `ti` terms of RSFs
+
+```{r hrsfs, fig.cap="\\textbf{Temeperature strongly affects mammals' habitat selection.} Estimated resourse selection strength for forest cover (\\%), elevation (km), and distance from fresh water (km) as a function of temperature.
 The grey contours indicate the extent of each speices' observed locations."}
@@
 
+* improve projections at the edges of the HRSFs with local and Traditional Knowledge [@two-eyed_seeing_ref]
 * changes in temperatures are pushing to areas in the surface plots that have little data and may be poorly estimated.

### 2024-10-04 analysis/06-temperature-hgams.R

@@ -128,24 +128,47 @@ ggplot(d_2) +
 
 # check spread for tod, doy, and temp for each species except for grizzly
 # should be ok to keep cs = 'cc'
-plot_grid(
-  d %>%
-    select(species, tod_pdt, doy, temp_c, moving) %>%
-    pivot_longer(-c(species, moving)) %>%
-    ggplot(aes(value)) +
-    facet_grid(species ~ name, scales = 'free') +
-    geom_histogram(aes(fill = moving), position = 'stack') +
-    ggtitle('Full dataset') +
-    scale_fill_brewer(type = 'qual', palette = 6) +
-    theme(legend.position = 'top'),
-  d_2 %>%
-    select(species, tod_pdt, doy, temp_c) %>%
-    pivot_longer(-species) %>%
-    ggplot(aes(value)) +
-    facet_grid(species ~ name, scales = 'free') +
-    geom_histogram() +
-    ggtitle('Moving only'),
-  labels = 'AUTO', nrow = 1)
+hist_1 <- d %>%
+  select(species, tod_pdt, doy, temp_c, moving) %>%
+  pivot_longer(-c(species, moving), names_to = 'variable') %>%
+  mutate(variable = case_when(
+    variable == 'doy' ~ 'Day of year',
+    variable == 'temp_c' ~ paste0('Temprature (\U00B0', 'C)'),
+    variable == 'tod_pdt' ~ 'Time of day (PDT)') %>%
+      factor(., levels = unique(.))) %>%
+  ggplot(aes(value)) +
+  facet_grid(species ~ variable, scales = 'free', switch = 'x') +
+  geom_histogram(aes(fill = moving), position = 'stack', bins = 24) +
+  scale_x_continuous(NULL, expand = c(0, 0)) +
+  ylab('Count') +
+  scale_fill_brewer('Moving', type = 'qual', palette = 6,
+                    breaks = c(FALSE, TRUE), labels = c('No', 'Yes')) +
+  theme(legend.position = 'none', strip.background.x = element_blank(),
+        strip.text.x = element_text(size = 11), strip.placement = 'outside')
+
+hist_2 <- d_2 %>%
+  select(species, tod_pdt, doy, temp_c) %>%
+  pivot_longer(-c(species), names_to = 'variable') %>%
+  mutate(variable = case_when(
+    variable == 'doy' ~ 'Day of year',
[…]

### 2025-01-17 analysis/00-crop-resource-rasters.R
changed forest cover raster to be for all trees
was initially only for Evergreen/Deciduous Needleleaf Trees

@@ -19,19 +19,45 @@ locs_ext <- locs %>%
   st_as_sfc() %>%
   st_as_sf()
 
-# original rasters from http://www.earthenv.org/landcover
+
+# import digital elevation raster (zoom > 6 gives excessively high peaks)
+#' see `analysis/00-download-bc-dem.R` for downloading
 dem <- rast('data/resource-rasters/bc-buffered-dem-z6.tif')
 plot(dem)
 plot(locs_ext, add = TRUE, lwd = 3)
 
-# forest raster
-rast('data/resource-rasters/uncropped-rasters/consensus_full_class_1.tif') %>%
-  crop(dem) %>%
-  writeRaster('data/resource-rasters/forest.tif')
+# forest raster (original rasters from http://www.earthenv.org/landcover)
+f1 <- rast('data/resource-rasters/uncropped-rasters/consensus_full_class_1.tif') %>%
+  crop(dem)
+f2 <- rast('~/../Downloads/consensus_full_class_2.tif') %>%
+  crop(dem)
+f3 <- rast('~/../Downloads/consensus_full_class_3.tif') %>%
+  crop(dem)
+f4 <- rast('~/../Downloads/consensus_full_class_4.tif') %>%
+  crop(dem)
+
+layout(matrix(1:4, ncol = 2))
+plot(f1, main = 'layer 1: Evergreen/Deciduous Needleleaf Trees')
+plot(f2, main = 'layer 2: Evergreen Broadleaf Trees')
+plot(f3, main = 'layer 3: Deciduous Broadleaf Trees')
+plot(f4, main = 'layer 4: Mixed/Other Trees')
+layout(1)
+
+f <- f1 + f2 + f3 + f4
+layout(1:2)
+plot(f1, main = 'Evergreen/Deciduous Needleleaf Trees')
+plot(f, main = 'all trees')
+layout(1)
+
+max(values(f))
+values(f) <- if_else(values(f) > 100, 100, values(f))
+max(values(f))
+
+writeRaster(f, 'data/resource-rasters/forest.tif')
 plot(rast('data/resource-rasters/forest.tif'))
 plot(locs_ext, add = TRUE, lwd = 3)
 
-# raster of open water
+# raster of open water (original rasters from http://www.earthenv.org/landcover)
 layout(matrix(1:8, ncol = 2, byrow = TRUE))
 w <- rast('data/resource-rasters/uncropped-rasters/consensus_full_class_12.tif') %>%
   crop(dem)
[…]

### 2025-01-17 analysis/06-temperature-hgams.R
added a comment on kmeans

@@ -31,7 +31,7 @@ if(file.exists('data/hgam-speed-data.rds')) {
              stringr::str_replace_all('southern mountain',
                                       '(s. mountain)') %>%
              factor()) %>%
-    #' cluster speeds using `kmeans()`
+    #' cluster speeds using `kmeans()` (does a poor job, see below)
     nest(speeds = ! species) %>%
     mutate(speeds = map(speeds, \(.d) {
       km <- kmeans(.d$speed_est, centers = 2)

### 2025-01-17 writing/manuscript.Rmd
rewrote intro, updated methods

@@
     institute: [ucf, smith]
+  - name: Siobhan Darlington
+    institute: [braes, biol]
+  - name: Adam T. Ford
+    institute: [braes, biol]
+  - name: Karen E. Hodges
+    institute: [braes, biol]
+  - name: Siobhan Darlington
+    institute: [braes, biol]
   - name: Kirk Safford
     institute: bcp
+  - name: Robert Serrouya
+    institute: [braes, biol, biodiv]
   - name: Michael J. Noonan
@@
   - bcp: BC Parks
+  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
@@
 ```{r, echo=FALSE}
-abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move.
-These changes in species' movement behavior have had cascading effects on individuals' fitness and habitat selection with consequences at the population, community, and ecosystem levels.
-Previous research has quantified changes in mammalian movement behavior, but we are lacking an estimate of the effects of climate change on fine-scale movement and habitat selection.
-We address this by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on mammal's probability of movement, speed, and habitat selection.
-We then leverage the estimated effects to predict when, how, and where mammals will move in BC in the current century.
-We conclude with implications for energetics, encounter rates, habitat conservation, and human-wildlife conflicts."
+abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on individuals' fitness and habitat selection at the population, community, and ecosystem levels.
+Previous research has quantified the effects of temperature on mammalian movement behavior, but few have quantified the effects of future climate change on fine-scale movement and habitat selection.
+We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on mammal's probability of movement, speed, and habitat selection.
+We then leverage the estimated effects to predict when, how, and where mammals will move in BC throughout the 21^st^ century.
+We conclude with implications for energetics, encounter rates, habitat conservation, and human-wildlife conflicts.
+***MISSING RESULTS AND CONCLUSIONS***"
 ```
@@
 
-\noindent \textbf{Tables}: 1
+\noindent \textbf{Tables}: 3
 
-\noindent \textbf{References}: 38 (updated on 2024-07-28) <!-- counted by using nature csl -->
+\noindent \textbf{References}: 73 (updated on 2025-01-16) <!-- counted by using nature csl -->
 
@@
 \noindent `r abstract`
-
-# Useful references to add {-}
-
-* @hegel_interacting_2010: weather affects caribou calf recruitment and interacts with predator (wolf) abundance
-* @elmore_implications_2017: "thermal environments should be foundational in the understanding of the habitat concept"
-* @pease_ecological_2024: the scale at which we measure variables affects their estumated effect
-* @hetem_responses_2014:
-* rate of climate change is too fast for genetic adaptation to occur in mammals with longevities of decades
-* example of suitable habitat prediction to 2050
-* to read: https://doi.org/10.1186/s40665-016-0024-1
 
@@
 
-<!-- introduce knowledge gap -->
+<!-- movement is a convenient way to respond to change -->
 
-\noindent It is well understood that mammalian energetics depend on ambient temperature [@brown_toward_2004; @taylor_energetics_1982], but we have a limited understanding of how mammals adapt their movement behavior in response to temperature [@elmore_implications_2017; but see: @milling_seasonal_2017; @montgomery_movement_2019].
-Previous work has focused on energetic costs of movement [@taylor_energetics_1982; @dyer_travel_2023], range shifts [@leclerc_determinants_2021], thermoregulation [@mota-rojas_physiological_2021; @melin_effect_2023], and trait-based responses [@mccain_assessing_2019], but there remains a need for a more comprehensive understanding of how mammals will adapt their fine-scale movement and habitat selection to future changes in temperature [@intergovernmental_panel_on_climate_change_climate_2023].
-Movement plasticity [@nathan_movement_2008] allows mammals to react to changes in temperature on a shorter timescale than evolutionary physiological adaptations [@levins_evolution_1974].
-By changing when, how, and where they move, mammals can conserve chemical and thermal energy as well as reduce the risk of overheating [@alston_temperature_2020; @dyer_travel_2023; @attias_effects_2018; @giroux_activity_2023; @verzuh_behavioural_2023; @williams_range_2018].
-We present a multi-species analysis of how mammals adapt their movement behavior in response to changes in air temperature.
+\noindent For most animals, movement is a rapid, low-cost, and low-risk response to a variety of stimuli [@nathan_movement_2008], including escaping predation [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024], searching for resources [@charnov_optimal_1976; @mezzini_how_2024], and habitat selection [@kacelnik_ideal_1992; @winter_forecasting_2024].
+However, movement efficiency depends on ambient temperature, since it impacts the energetic cost of movement [@taylor_energetics_1982; @brown_toward_2004; @jahn_variations_2022], the ease of movement [e.g., snow depth and density, see @montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023].
+Additionally, temperature also affects movement indirectly by affecting other drivers of movement behavior, including physiology [@abram_behavioural_2017], fitness [@schmidt_interplay_2020; @berger_climate_2018], trophic interactions [@hegel_interacting_2010], and the availability and value of resources [e.g., water and heat, see: @elmore_implications_2017; @attias_effects_2018].
 
-<!-- movement plasticity allows for rapid responses, but how does climate change impact this? -->
+<!-- changes in temperature cause changes in movement -->
 
-Recent changes in climate have affected mammals' movement rates and habitat selection [@deb_modelling_2020], but it is still unclear how changes in the current century will affect their fine-scale movement behavior and spatial distribution by the end of the century.
+Changes in temperature affect not only whether and how fast animals move but also how they move [i.e., speed and tortuosity; their movement behavior: @nathan_movement_2008; @dyer_travel_2023] and where they move to [habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].
+Thus, temperature affects not only an animal's energetic expenditure but also many other consequences of movement, including encounter rates with resources [@mezzini_how_2024], other animals [@martinez-garcia_how_2020], and the occurrence of human-wildlife conflict [@weststrate_how_2024].
+These changes are exacerbated at extreme temperatures, as animals often alter their movement behaviors substantially when temperatures are excessively cold or hot [@wilson_stenotherms_2001; @berger_climate_2018; @arechavala-lopez_common_2019; @giroux_activity_2023].
+But what counts as "extreme" varies across species: thermal specialists are more sensitive to small changes above or below their ideal range [@giroux_activity_2023], while temperature generalists can adapt more easily to change [@dupont_beyond_2024].
+However, this grater adaptability is tested when generalists are subject to changes that are more severe [@mccain_assessing_2019], affecting many factors [e.g., temperature, phenology, and resource availability: @polazzo_measuring_2024], or too rapid to adapt to within the span of a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].
+Thus, it remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially when populations do not react to changes in environmental conditions [@sawyer_migratory_2019] or are physiologically incapable [@sherwood_adaptability_2010; @williams_range_2018].
+
+<!-- we do not know how temperature will change -->
+
+This uncertainty in both future conditions and their consequences on ecosystems and populations presents a need for a comprehensive understanding of the effects of temperature on the movement of animal populations as well as the consequences this will have on landscapes, biodiversity, and their adaptability and resiliency.
+Recent work has documented climate-change-related changes mammals' ranges [@leclerc_determinants_2021], movement behavior [@melin_effect_2023]; thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [@mccain_assessing_2019], but it is still unclear how mammals will adapt their small-scale movement and habitat selection to future changes in temperature [@intergovernmental_panel_on_climate_change_climate_2023; but see @hetem_responses_2014; @winter_forecasting_2024].
 In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures and severe forest fires [@intergovernmental_panel_on_climate_change_climate_2023; @zurowski_summer_2023].
-Over the next 100 years, these changes will continue to affect mammals' fitness, movement behavior, and habitat selection as they cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and frequency and intensity of extreme events [@bunnell_global_2011] along with anthopogenic pressure [@sih_evolution_2011; @weststrate_how_2024].
+Over the next 100 years, these changes will worsen and continue to impact mammals' fitness, movement behavior, and habitat selection as they cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and frequency and intensity of extreme events [@bunnell_global_2011] along with growing anthopogenic pressure [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024].
 
-<!-- implications for ecology: encounter rates, energetic expenditure, range shifts, consequences for conservation, etc -->
+<!-- why is understanding the present and predicting into the future necessary -->
 
-Understanding animals' responses to temperature is essential for anticipating their response to climate change in the coming decades and the cascade of consequences such changes will have.
-In light of the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [@ref], we must understand which areas are best preserved to maximize the project's effectiveness.
-Predicting how climate change will affect how, when, and where animals will move is necessary for choosing which 30% of the world is most worth preserving.
-Changes in movement behavior impact several aspects of animals' fitness, behavior, and odds of survival, including encounter rates [with food, competitors, and predators: @martinez-garcia_how_2020], energetic expenditure [@taylor_energetics_1982], range shifts [@woo-durand_increasing_2020].
-Thus, understanding the effects of temperature on animals' movement behavior and their consequences is crucial for effective conservation.
+Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to be able to prepare and respond to change in a proactive manner [e.g., @sawyer_identifying_2009; @mccain_assessing_2019].
+The Canadian province of British Columbia is currently experiencing rapid widespread warming, phenological shifts, and more frequent and intense extreme events, including forest fires [@zurowski_summer_2023].
+In light of the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022], we stress that decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness.
+Predicting how climate change will affect how, when, and where animals will move is necessary for identifying which landscapes are most valuable now as well as in the future.
+Failing to understand how changes in temperature will affect mammalian movement behavior and habitat selection will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.
 
@@
 
-This paper provides an analysis of the effects of ambient temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (*Canis lupus*, *Cervus canadensis*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, and *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
+This paper provides an analysis of the effects of spatiotemporally proximal temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (*Canis lupus*, *Cervus canadensis*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, and *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
 Using over two decades of telemetry data over a large spatial range of British Columbia (Fig.
-\@ref(fig:bc-map)) and hourly air temperature data, we estimate how mammals altered their movement frequency, movement speed, and habitat selection in response to air temperature.
-We then pair the estimated responses with climate change projections to forecast changes in mammalian movement in 2100 under different climate-change scenarios [Shared Socioeconomic Pathways, see @riahi_shared_2017].
-We discuss the consequences of these changes in movement behavior with regards to energetics, encounter rates, habitat conservation, and human-wildlife conflicts.
+\@ref(fig:bc-map)) and hourly air temperature data, we estimate how mammals altered their movement frequency, movement speed, and habitat selection in response to temperature.
+We then pair the estimated responses with climate change projections to forecast changes in mammalian movement in the current century under different climate-change scenarios [Shared Socioeconomic Pathways, see @riahi_shared_2017].
+We discuss the consequences of these changes in movement behavior with regards to energetics, encounter rates, habitat conservation, and human-wildlife conflicts while disentangling the direct and indirect effects of temperature on the movement behavior of terrestrial mammals.
 
-```{r bc-map, fig.cap="GPS telemetry data and 95% autocorrelated kernel density utilization distributions for the six species in this study."}
+```{r bc-map, fig.cap="GPS telemetry data for the six species in this study."}
 knitr::include_graphics('../figures/tels-map.png')
@@
   group_by(dataset_name, animal) %>%
-  summarize(any_na_speed = any(is.na(speed_est)),
+  summarize(n = n_distinct(animal),
+            any_na_speed = any(is.na(speed_est)),
             all_na_speed = all(is.na(speed_est))) %>%
@@
   transmute(
-    Dataset = case_when(
+    Species = case_when(
       dataset_name == 'Canis_lupus_boreal' ~ '\\emph{Canis lupus}',
@@
   unnest(tel) %>%
-  group_by(Dataset, animal) %>%
+  group_by(Species, animal) %>%
   summarize(Start = min(as.Date(timestamp)),
@@
             dt = median(timestamp - lag(timestamp), na.rm = TRUE),
+            animal = unique(animal),
             na_speed = unique(na_speed),
             .groups = 'drop') %>%
-  group_by(Dataset) %>%
+  group_by(Species) %>%
   summarize(Start = min(Start),
@@
             dt = as.numeric(median(dt, na.rm = TRUE)),
+            Animals = n_distinct(animal),
             n_missing = sum(na_speed),
@@
   mutate(dt = round('hours' %#% dt, 2)) %>% # convert from seconds to hours
-  rename('Median \U0394t (hours)' = dt,
+  rename('\U0394t (hours)' = dt,
          'NA speed' = n_missing)
 
-knitr::kable(dt_data, format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$; stratified by individual) and number of animals with unresolved speed (NA speed) for each of the species in this study.', escape = FALSE)
+knitr::kable(dt_data, format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$; stratified by individual), number of animals (Animals), and number of animals with unresolved speed (NA speed) for each of the species in this study.', escape = FALSE)
 ```
@@
 
-## GPS telemetry data
+## Datasets
 
-\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)), and the mountain goat (*Oreamnos americanus*) locations were provided by BC Parks.
-All other tracking data were obtained from private collaborators.
-Outlier GPS locations were removed following diagnostic analyses of the distance from the median location as well as straight-line displacement, turning angle, and time interval between consecutive points.
-Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and high straight-line displacement, especially if antecedent and subsequent points indicated stationary behavior.
-The script used to clean the data and all associated custom functions are available on GitHub ([https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature)).
+This manuscript leverages four main datasets: (1) a collection of GPS telemetry data; (2) historical hourly reanalyzed air temperature from the ERA5 dataset from the European Center for Medium-range Weather Forecasting; (3) resource rasters of percent forest cover, elevation, and distance from water; and (4) climate change projections under four Shared Socioeconomic Pathways [SSPs, see @riahi_shared_2017].
+We detail the data sources and analyses in the sections below, and more details are available in appendices A (estimating how temperature affects movement frequency and speed when moving), B (fitting Hierarchical Resource Selection Functions), and C (simulating hourly temperature data to 2100 using the four SSPs).
 
-## Historical temperature data
+### GPS telemetry data
+
+\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)).
+Clearly erroneous GPS locations were removed using diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points.
+Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated stationary behavior.
+The script used to clean the data and all associated custom functions are available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).
+
+### Historical and projected temperature data
 
 \noindent Rasters of hourly reanalyzed air temperature data were downloaded from the ERA5 dataset [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)).
-Proximal temperature was estimated for each location by extracting the values from the corresponding raster cell from the temporally nearest raster using the `extract()` function from the `terra` package [version 1.7-46, @hijmans_terra_2023] for `R`  [@r_core_team_r_2024].
+Proximal temperature was estimated for each location by extracting the value from the corresponding raster cell from the temporally nearest raster using the `extract()` function from the `terra` package [version 1.7-46, @hijmans_terra_2023] for `R` [@r_core_team_r_2024].
+
+We obtained rasters of projected monthly average temperature at a 0.08$^\circ$ resolution in BC from 2020 to 2100 under the different SSPs via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`.
+Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 data for BC from 1998 to 2023 [extremes included, see @hersbach_era5_2023] to calculate within-month variance in temperature, which we defined as the variance within a given pixel, month, and year.
+We then modeled the estimated variance using a GAM for location and scale [GAMLS, see: @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017].
+The linear predictor for the location (i.e., the mean) included smooth terms of the within-pixel monthly mean temperature (within each year), month (as a cyclic smooth), a two-dimensional smooth of space, and a tensor interaction product term of space and month.
+The linear predictor for the scale term (which governs the mean-variance relationship) included smooth terms of the monthly mean, month, and space.
+We did not include a smooth of year to avoid unrealistic projections when extrapolating beyond the range of historical data.
+Additional details are available in Appendix C.
+
+```{r, eval=FALSE, echo=FALSE}
+m_gammals <- gam(list(
+  # linear predictor for the mean parameter
+  monthly_var ~
+    s(monthly_mean, bs = 'cr', k = 10) +
+    s(month, bs = 'cc', k = 10) + # using a cyclic cubic spline basis
+    s(long, lat, bs = 'ds', k = 500) +
+    ti(long, lat, month, bs = c('ds', 'cc'), d = c(2, 1), k = c(100, 10)),
+  # linear predictor for the scale parameter
+  ~ s(monthly_mean, bs = 'cr', k = 10) +
+    s(month, bs = 'cc', k = 10) +
+    s(long, lat, bs = 'ds', k = 100)),
+  family = gammals(),
+  data = d,
+  method = 'REML',
+  knots = list(month = c(0.5, 12.5)))
+```
+
+We simulated hourly variation in weather by assuming temperature followed a normal distribution with mean specified by the monthly `climatenaR` climate projections and variance as specified by the Gamma GAMLS.
+We then predicted changes in movement behavior and habitat selection as a function of the simulated temperature using the HGAMs and HRSFs.
+For each month within each year from 2020 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantile by increments of 0.1, and we weighted each quantile proportionally to the (normalized) Gaussian probability density for each quantile (Appendix C).
+
+### Resource rasters
+
+\noindent We estimated percent forest cover and distance from water using the rasters created by @tuanmu_global_2014.
+We calculated total forest cover by summing the temporally static rasters of evergreen/deciduous needleleaf trees, evergreen broadleaf trees, deciduous broadleaf trees, and mixed/other trees (raster classes 1-4, respectively).
+We converted the raster of percent cover of open water (class 12) to a binary raster of pixels with water (cover > 0%) or without water (cover = 0%) and then calculated each pixel's distance from the nearest pixel with water using the `distance()` function from the `terra` package.
+Finally, we obtaied two digital elevation models using the `get_elev_raster()` function from the `elevatr` package [version 0.99.0, @hollister_elevatr_2023]: We used a raster with a zoom of 6 (0.009$^\circ$) for model fitting and one with a zoom of 3 (0.08$^\circ$) for downloading climate change projections via `climatenaR`.
+All final rasters and code are available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).
 
@@
 \noindent We modeled each animal's movement using continuous-time movement models [@fleming_fine-scale_2014] via the `ctmm` package [version 1.2.0, @fleming_ctmm_2023] for `R`.
-We then estimated mammals' instantaneous speeds by applying the `ctmm::speeds()` function on all models with finite speed estimates [416 of 433, see @fleming_fine-scale_2014; @noonan_scale-insensitive_2019].
+We then estimated mammals' instantaneous speeds by applying the `ctmm::speeds()` function on all models with finite speed estimates [415 of 433, see @fleming_fine-scale_2014; @noonan_scale-insensitive_2019].
 The 18 telemetries with insufficiently fine sampling were for one mountain goat, 15 boreal caribou, and 2 southern mountain caribou (Table \@ref(tab:data-table)).
@@
 
-The following two sections detail the modelling we used to estimate the effect on temperature on mammals' movement.
-To assess the importance of including temperature as an explicit covariate (rather than including its effects with time of day and day of year***HERE***), we fit models with and without temperature and compared the fits and model predictions of the two sets of models.
+\noindent The following sections detail the statistical framework and modelling we used to estimate the effect on temperature on mammals' movement.
+To assess the importance of including temperature as an explicit covariate (rather than including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the deviance explained, model predictions, and AIC values of the two sets of models.
+
+### Disentangling direct and indirect effects
+
+\noindent In this manuscript, we separate the effects of temperature on mammalian movement into indirect and direct effects.
+We call "indirect" all effects that can be intervened on by altering an animal's habitat [its geographic space *sensu* @matthiopoulos_defining_2022], and we refer to all other effects as "direct".
+For example, an increase in temperature may push an animal to prefer higher forest cover in search of shade, and conservationists can thus mitigate these indirect effects of temperature on movement behavior by increasing the availability of forested areas, since the effect of temperature on movement is conditional on forest cover availability.
+In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022], so it is not possible to mitigate the effects of temperature on movement behavior (other than changing temperature directly).
+We provide Directed Acyclical Graphs in Appendix A (modeling the effects of temperature on movement rates), and Appendix B (modeling the effects of temperature on habitat selection).
 
@@
 The second HGAM estimated an animal's speed (when moving) with a gamma family of distributions and log link function.
-The HGAMs included random intercepts for each animal (`s(..., bs = 're')`), fixed intercepts for each species, and species-level factor smooth interactions for time of day, day of year, and temperature [`s(..., bs = 'fs')`, see model S in Figure 4 of @pedersen_hierarchical_2019].
+The HGAMs included random intercepts for each animal (`s(..., bs = 're')`), fixed-effect intercepts for each species, and species-level `by` smooths for time of day, day of year, and temperature [`s(..., by = species)`, see model I in Figure 4 of @pedersen_hierarchical_2019].
 Additionally, the models had three tensor product interaction terms (`ti()`) for each species: (1) day of year and time of day, (2) temperature and time of day, and (3) temperature and day of year.
@@
 The HGAMs accounted for the cyclicity of time of day and day of year using cyclic cubic splines [`bs = 'cc'`, @wood_generalized_2017].
-Together, the binomial HGAM and the gamma HGAM inform us on an animal's long-term average speed, since it is the product of the probability the animal is moving and its average speed when moving.
+Together, the binomial HGAM and the gamma HGAM inform us on an animal's long-term average speed, since it is the product of the probability of moving and its average speed when moving.
 We fit the models with fast REML (`'fREML'`) and discretized covariates (`discrete = TRUE`) to optimize computational efficiency with no appreciable losses to model performance [@wood_generalized_2015; @wood_generalized_2017-1; @li_faster_2020].
@@
 
-\clearpage
+<!-- \clearpage -->
 
@@
   discrete = TRUE, # discretize the covariates for faster computation
-  knots = list(tod_pdt = c(0, 1), doy = c(0.5, 366.5))) # for bs = 'cc'
+  knots = list(tod_pdt = c(0, 1), doy = c(0.5, 366.5))) #' for `bs = 'cc'`
 
@@
   family = Gamma(link = 'log'),
-  data = d_2, # data filtered for speed > 0
+  data = d_2, #' data filtered for only moving
   method = 'fREML',
@@
 
-\noindent We estimated the effects of temperature on each species' selection for percent forest cover, elevation, and distance from fresh water (i.e., not sea water) by fitting a Hierarchical Resource Selection Function [@aarts_estimating_2008; @refs; @10.1111/ecog.07225] an HGAM with a Poisson family of distributions and log link function for each species [Appendix B, also see @aarts_estimating_2008; @alston_mitigating_2022].
-While we recognize there are other important drivers of habitat selection that we did not include in our models (e.g., forest age), we decided to only use these three proxies to produce results that are relatively comparable between species and to make province-wide predictions simpler.
+\noindent We estimated the effects of temperature on each species' selection for percent forest cover, elevation, and distance from water by fitting a Hierarchical Resource Selection Function for each species using an HGAM with a Poisson family of distributions and log link function [Appendix B, also see @aarts_estimating_2008].
+We accounted for the spatiotemporal autocorrelation in the telemetry locations by weighting each point based on the telemetry's Autocorrelated Kernel Density Estimate [@fleming_new_2017; @noonan_comprehensive_2019] to produce estimates of second- and third-order habitat selection [*sensu* @johnson_comparison_1980].
+While we recognize there are other important drivers of habitat selection (e.g., forest age, forest type, prey availability), we decided to only use these three proxies to produce results that are comparable between species and to make province-wide predictions simpler.
 Each species' model had the same structure:
@@
 rsf <- bam(
-  detected ~ # 1 for GPS locations, 0 for quadrature data
+  detected ~ # 1 for telemetry locations, 0 for quadrature data
     # species-level average resource preference
     s(forest_perc, k = 6, bs = 'tp') +
-    s(elevation_m, k = 6, bs = 'tp') +
+    s(elevation_m, k = elev_k, bs = 'tp') +
     s(dist_water_m, k = 6, bs = 'tp') +
@@
     s(animal, bs = 're') +
-    s(forest_perc, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')) +
-    s(elevation_m, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')) +
-    s(dist_water_m, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')) +
+    s(forest_perc, animal, k = 6, bs = 'fs', xt = list(bc = 'cr')) +
+    s(elevation_m, animal, k = 6, bs = 'fs', xt = list(bc = 'cr')) +
+    s(dist_water_m, animal, k = 6, bs = 'fs', xt = list(bc = 'cr')) +
     # changes in preference with temperature
-    ti(forest_perc, temperature_C, k = 6, bs = 'cr') +
-    ti(elevation_m, temperature_C, k = 6, bs = 'cr') +
-    ti(dist_water_m, temperature_C, k = 6, bs = 'cr'),
+    ti(forest_perc, temperature_C, k = 6, bs = 'tp') +
+    ti(elevation_m, temperature_C, k = 6, bs = 'tp') +
+    ti(dist_water_m, temperature_C, k = 6, bs = 'tp') +
+    # include marginals of temperature to remove sampling biases
+    s(temperature_C, k = 4, bs = 'tp') +
+    s(temperature_C, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')),
   family = poisson(link = 'log'),
@@
 
-Smooth effects of percent forest cover, elevation, and distance to fresh water accounted for the species-level average selection strength for each resource.
+Smooth effects of percent forest cover, elevation, and distance to water accounted for the species-level average selection strength for each resource.
 A random effect for each individual animal corrected for uneven sampling across individuals, while factor smooth interaction terms (`bs = 'fs'`) accounted for individual-level resource selection (i.e., individual-level deviations from the species-level average).
-Finally, tensor interaction product terms (`ti()`) of the three resources and temperature estimated the change in resource selection at different temperatures.
-Detections were weighted proportionally to their degree of independence from temporally proximate detections [see Appendix C and @alston_mitigating_2022], while quadrature points had a weight of 1.
+Tensor interaction product terms (`ti()`) of the three resources and temperature estimated the change in resource selection at different temperatures.
+Finally, we included marginal smooth terms of temperature to account for species- and individual-level sampling biases over temperature (e.g., sampling more during warm periods).
+Detections were weighted proportionally to their degree of independence from other temporally proximate detections [`weights = weight` -- see Appendix B and @alston_mitigating_2022], while quadrature points had a weight of 1.
 
-***does elk HRSF use different `k` values?***
-
-## Predicting changes in animal movement during the current century
-
-\noindent Rasters of projected monthly average temperature in BC from 2020 to 2100 were obtained via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`.
-Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 data for BC from 1998 to 2023 [extremes included, see @hersbach_era5_2023] to calculate within-month variance, which we defined as the variance within a given pixel, month, and year.
-We then modelled the estimated variance using a GAM for location and scale [GAMLS, see @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017].
-The linear predictor for the location (i.e., the mean) included smooth terms of the within-pixel monthly mean temperature (for each year), month (as a cyclic smooth), a two-dimensional smooth of space, and a tensor interaction product term of space and month.
-The linear predictor for the scale term (which governs the mean-variance relationship) included smooth terms of the monthly mean, month, and space.
-Additional details are available in Appendix C.
-
-```{r, eval=FALSE, echo=FALSE}
-m_gammals <- gam(list(
-  monthly_var ~
-    s(monthly_mean, bs = 'cr', k = 10) + # using cr to reduce fitting time
-    s(month, bs = 'cc', k = 10) +
-    s(long, lat, bs = 'ds', k = 500) +
-    ti(long, lat, month, bs = c('ds', 'cc'), d = c(2, 1), k = c(100, 10)),
-  ~ s(monthly_mean, bs = 'cr', k = 10) +
-    s(month, bs = 'cc', k = 10) +
-    s(long, lat, bs = 'ds', k = 100)),
-  family = gammals(),
-  data = d,
-  method = 'REML',
-  knots = list(month = c(0.5, 12.5)))
+```{r, eval=FALSE}
+library('mgcv')
+library('purrr')
+x <- list.files('H:/GitHub/bc-mammals-temperature/models/', 'rsf-', include.dirs = FALSE, full.names = T)
+x <- x[grepl('2024-10', x)]
+map_int(x, \(f) readRDS(f)$iter)
 ```
-
-We simulated for hourly variation in weather by assuming temperature followed a normal distribution with mean specified by the `climatenaR` climate projections and variance as specified by the Gamma GAMLS.
-We then predicted changes in movement behavior and habitat selection as a function of temperature using the HGAMs and HRSFs along with the temperature HGAM described above.
-For each month within each year from 2020 to 2100, we simulated hourly weather by including temperatures from the 0.1, 0.2, ..., 0.8, 0.9 quantiles and weighted each quantile proportionally to the (normalized) Gaussian probability density for each value (Appendix C).
 
@@
 
-While adding temp did not improve model fit (add change in dev.expl), but it ... (delta BIC for both modelss) (Fig. ***SX*** pred vs pred)
+\noindent Species differed in mean probabilities of movement (range: 0.05 -- 0.27), mean speed when moving (range: 0.42 -- 2.67 km/day), and mean distance travelled (range: 0.04 -- 0.62 km/day) even after post-stratifying to a sampling interval of one hour (Tab.
+\@ref(tab:gam-intercepts)).
+The inclusion of temperature in the model did not affect any of the three measures substantially when post-stratifying to $0^\circ$C.
+Grizzly bears tended to move least often ($P(M = 1) \approx 0.05$), while wolves and pumas moved most often ($P(M = 1) \ge 0.23$), and wolves had the highest mean speed when moving.
+Consequently, wolves traveled 3.0 to 15.6 times further per day than other mammals, on average.
 
-```{r, eval=FALSE}
-tibble(parameter = c('P(moving)', 'speed'),
-       with_temp = c(10.7, 79.3),
-       without_temp = c(10.2, 79.1),
-       delta_AIC = c(diff(AIC(m_1_no_t, m_1)$AIC),
-                     diff(AIC(m_2_no_t, m_2)$AIC))) %>%
-  knitr::kable()
+* CHANGED FOREST RASTER TO USE ALL TREE TYPES (2025-01-17)
+* NEED TO REMAKE HRSF FIGURES
 
-#' all HRSFs have `delta_AIC` < -85 (check boreal caribou and elk)
+***THIS IS FALSE***We did not find any consistent selections for or against forest cover, but all species exhibited stronger selection preferences for elevation and distance from water than for forest cover, and there was no consistent preference for forest cover across temperatures.
+In contrast, all species showed a strong preference for specific altitudes: wolves, pumas, and boreal caribou generally preferred elevations < 1 km above sea level, while elk, mountain goats, southern mountain caribou, and grizzly bears generally preferred elevations between 1 and 2 km above sea level.
+All species except for elk and grizzly bears preferred being as close to water as possible,
+
+Overall, we did not detect any strong selections for or against forest cover, and this lack of preference was generally true irrespective of temperature.
+In contrast, we selection strength for forest cover did not vary strongly with temperature,
+
+preferred resources
+
+**CHECK TABLE 2: UPDATE AFTER REMOVING SCEK014 AND SCEK014b (2024-11-19 9:50)**
+
+***UPDATE TABLES BASED ON NEW HRSFS***
+
+\footnotesize
+
+```{r gam-intercepts}
+setwd('~/GitHub/bc-mammals-temperature')
+library('mgcv')
+library('dplyr')
+library('ctmm') #' for `%#%`
+source('analysis/figures/default-ggplot-theme.R')
+
+get_int <- function(.m) {
+  newd <- mutate(.m$model[1:7, - 1], species = SPECIES, temp_c = 0, dt = 1)
+  predict(.m, type = 'response', newdata = newd,
+          terms = c('(Intercept)', 'species',
+                    's(temp_c)', paste0('s(temp_c):species', SPECIES),
+                    's(log(dt))', 's(log(dt),species)'))
+}
+
+intercepts <- tibble(Species = SPECIES,
+                     'P(M)' = get_int(readRDS('models/binomial-gam-without-temperature.rds')),
+                     'P(M|T)' = get_int(readRDS('models/binomial-gam.rds')),
+                     'E(S|M=1)' = 'km/h' %#% get_int(readRDS('models/gamma-gam-without-temperature.rds')),
+                     'E(S|M=1,T)' = 'km/h' %#% get_int(readRDS('models/gamma-gam.rds')),
+                     'E(D)' = `P(M)` * `E(S|M=1)`,
+                     'E(D|T)' = `P(M|T)` * `E(S|M=1,T)`) %>%
+  mutate(Species = paste0('\\emph{', Species, '}') %>%
+           gsub(')}', ')', .) %>%
+           gsub(' \\(', '} \\(', .))
+
+if(FALSE) {
+  hist(intercepts$`P(M)`)
+  hist(intercepts$`E(S|M=1)`)
+  round(range(max(intercepts$`E(D)`) /
+                intercepts$`E(D)`[- which.max(intercepts$`E(D)`)]),
+        1)
+
+  intercepts %>%
+    transmute(Species,
+              p = round(`P(M|T)` - `P(M)`, 2),
+              s = round(`E(S|M=1,T)` - `E(S|M=1)`, 2),
+              d = round(`E(D|T)` - `E(D)`, 2))
+}
+
+intercepts %>%
+  arrange(Species) %>%
+  mutate(Species = gsub('angifer', '\\.', Species)) %>%
+  rename_with(\(.x) paste0('$', .x, '$'), ! Species) %>%
+  rename_with(\(.x) gsub('\\|', '\\\\vert ', .x)) %>%
+  rename_with(\(.x) gsub('E', '\\\\mathbb E', .x)) %>%
+  rename_with(\(.x) gsub('M=1', 'M\\\\!=\\\\!1', .x)) %>%
+  knitr::kable(format = 'markdown', digits = 2, caption = paste0("Mean probability of movement (P(M)), speed when moving ($\\mathbb E(S | M = 1)$), and distance travelled ($\\mathbb E(D)$) as estimated by models without and with temperature, after post-stratifting to a 1-hour sampling rate and a temperature of T = 0\U00B0", "C."))
+```
+
+\normalsize
+
+For most models, including smooth terms of temperature did not always improve within-sample prediction substantially (change in deviance explained: +0.2% -- + 17.9%), and there was relatively strong agreement between models with and without temperature (Fig.
+***SX pred vs pred***), but including temperature resulted in substantially better fits for all models (all $\Delta$AIC $\le$ -326; Table \@ref(tab:temp-comparison)).
+
+```{r temp-comparison}
+setwd('..')
+library('dplyr')
+source('analysis/figures/default-ggplot-theme.R') #' for `SPECIES`
+
+temp_comp <- tibble(
+  Model = c('P(moving)', 'Speed', rep('HRSF', 7)),
+  Species = c('All', 'All', sort(as.character(SPECIES))),
+  DDE = 
+    c(10.7 - 10.2, 79.3 - 79.1, 10.0, 4.47, 17.6, 7.83, 5.89, 11.8, 17.9) %>%
+    round(1) %>%
+    paste0('%'),
+  DAIC = c(-5927, -1682, -7200, -4305, -769, -945, -6284, -326, -920) %>%
+    round())
+
+colnames(temp_comp)[colnames(temp_comp) == 'DAIC'] <- paste0('\U0394', 'AIC')
+colnames(temp_comp)[colnames(temp_comp) == 'DDE'] <- paste0('\U0394', 'DE')
+
+knitr::kable(temp_comp, digits = 1, format = 'markdown', caption = "Changes in deviance expained ($\\Delta$DE) and Akaike Information Criterion ($\\Delta$AIC) from including temperature in the models estimating probability of movement (P(moving)) and speed when moving (Speed), as well as in species-level Hierarchical Resource Selection Functions (HRSF).")
 ```
@@
 
-```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals travel.} Estimated effects of temperature on relative change in distance travelled (probability of moving times speed when moving) over time of day (\\textbf{A}) and day of year (\\textbf{B}).
+```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals travel.} The fill indicates the estimated effects of temperature on the relative change in distance travelled (probability of moving times mean speed when moving) over time of day (\\textbf{A}) and day of year (\\textbf{B}).
 Predictions in the surface plots extend to 10% of the range away from each datum.
@@
 Under the worst-case scenario, absolute projected changes were by 2100 ranged from ~1% (boreal caribou) to ~25% (southern moutain caribou) and had the greatest uncertainty in the estimates.
-Boreal caribou and elk showed relatively little projected changes, especially relative to the degree of inter-annual variability in displacement.
+Boreal caribou and elk showed relatively little projected changes, especially relative to the degree of inter-annual variability in distance travelled.
 
@@
 
-```{r t-dist, fig.cap="\\textbf{The intensity and direction of changes in movement rates due to climate change vary among species, but worse climate-change scenarios consistently result in the greatest change and greatest uncertainty.} Lines indicate the projected relative change in mean distance travelled (probability of moving times speed when moving) due to changes in temperature within the species' current extent.
+```{r t-dist, fig.cap="\\textbf{The intensity and direction of changes in movement rates due to climate change varies among species, but worse climate-change scenarios often result in the greatest change and greatest uncertainty.} Lines indicate the projected relative change in mean distance travelled (probability of moving times speed when moving) due to changes in temperature within the species' current extent.
 Shaded areas indicate the mean 95% Bayesian credible intervals.
@@
 
-***COMMENTS ABOUT SPATIAL VARIATION IN DISPLACEMENT***
+***COMMENTS ABOUT SPATIAL VARIATION IN distance travelled***
 
@@
 
-to do:
-* check `ti` terms of RSFs
-
-```{r hrsfs, fig.cap="\\textbf{Temeperature strongly affects mammals' habitat selection.} Estimated resourse selection strength for forest cover (\\%), elevation (km), and distance from fresh water (km) as a function of temperature.
+```{r hrsfs, fig.cap="\\textbf{Temeperature strongly affects mammals' habitat selection.} Estimated resourse selection strength for forest cover (\\%), elevation (km), and distance from water (km) as a function of temperature.
 The grey contours indicate the extent of each speices' observed locations."}
-knitr::include_graphics('../figures/rsf-surface-plots.png')
+knitr::include_graphics('../figures/hrsf-surface-plots.png')
 ```
@@
 
-```{r t-hrsfs, fig.cap="\\textbf{The intensity and direction of changes in habitat selection due to climate change vary among species, but worse climate-change scenarios consistently result in the greatest change and greatest uncertainty.} Lines indicate the projected change in relative selection strength for forest cover (\\%), elevation (km), and distance from fresh water (km) due to changes in temperature within the species' current extent.
+```{r t-hrsfs, fig.cap="\\textbf{The intensity and direction of changes in habitat selection due to climate change vary among species, but worse climate-change scenarios consistently result in the greatest change and greatest uncertainty.} Lines indicate the projected change in relative selection strength for forest cover (\\%), elevation (km), and distance from water (km) due to changes in temperature within the species' current extent.
 Shaded areas indicate the mean 95% Bayesian credible intervals.
@@
 
+We demonstrated that temperature is an important determinant of how and where mammals move, even while accounting for other seasonal drivers of movement (e.g, changes in photoperiod, hormones, and resources).
+This is important because decoupling temperature from other drivers allows us to predict how changes in climate will affect mammalian movement even if all other drivers remain constant.
+
 * We discuss the consequences of these changes in movement behavior with regards to energetics, encounter rates, habitat conservation, and human-wildlife conflicts.
@@
 * our models also assume the behaviour we modeled now will persist into the future (e.g., hybernation, migration)
-* based on the model, temperature doesn't affect grizzlies' hibernation much, and their displacement in winter is very low due to P(moving) being so low:
+* based on the model, temperature doesn't affect grizzlies' hibernation much, and their distance travelled in winter is very low due to P(moving) being so low:
 * if the RSS decreases a lot, individuals may disperse, populations may relocate, or stay where they are and suffer
+
+* RSFs require good data coverage over the resource-temperature space
+* future work investigating effects on a single species or evaluating space more accurately should:
++ use fully Bayesian models with appropriate regularizing priors and penalties to avoid excessive predicitons
++ use B splines to specify better penalty spaces
++ have more data with good coverage over the covariate space
+
+* @pease_ecological_2024: the scale at which we measure variables affects their estumated effect
+* @hetem_responses_2014:
+* rate of climate change is too fast for genetic adaptation to occur in mammals with longevities of decades
+* example of suitable habitat prediction to 2050
+* proactive conservation: @sawyer_identifying_2009
 
@@
 <!-- - what it tells us about the broader field -->
-## Consequences for conservation
 
-* improve projections at the edges of the HRSFs with local and Traditional Knowledge [@two-eyed_seeing_ref]
+## Considerations for future studies
+
+Our work highlights ***X*** central considerations for future research: (1) telemetry sampling rates should bedesigned primarily in relation to the movement timescales of the species of interest and ideally with multiple samples **path?**, (2) the number of individuals tracked and the length of each telemetry should depend on environmental stability, and (3) ***projecting into the fringes****
+
+Sampling schedules should be fine enough to reconstruct animals' movement at a sufficiently fine scale.
+Good estimates of an animal's speed and its movement path require telemetry locations to be taken more often than the animal's directional persistence [@noonan_scale-insensitive_2019], so that, on average, the dataset contains multiple locations in between changes in direction.
+What consitutes a change in direction depends on what movement scale one is investigating.
+Small-scale movements and first-order spatial selection will require more frequent sampling than migratory movement or secornd- and third-order spatial selection.
+While `ctmm` movement models are scale-invariant in that any model can be scaled to larger or smaller areas and timescales, the model estimates are not independent of the scale of the sampling frequency.
+
+When landscapes are relatively predictable across years, a larger number of sampled individuals is likely preferable over longer, multi-year telemetries.
+This allows one to quantify the variance across individuals, including the range of conditions and environments that individual animals are in.
+A good estimate of inter-individual variance provides better coefficient estimates along with more appropriate measures of uncertainty.
+However, when conditions across years are relatively stochastic, multi-year telemetries allow one to better estimate inter-annual and inter-individual variation.
+In either case, carefully-designed sampling schedules and (Bayesian) hierarchical models can provide good estimates of the effects of interested along with appropriate measures of uncertainty [@mcelreath_statistical_2016].
+
+* improve projections at the edges of the HRSFs with local and Traditional Knowledge [@kutz_two-eyed_2019]
 * changes in temperatures are pushing to areas in the surface plots that have little data and may be poorly estimated.
@@
 Nonetheless, we present important results for understanding how mammals will respond to changes in climate and weather during the next ~75 years.
+
+## Consequences for conservation
+
 * energetics
-* GPS sampling intervals were large: need more work on fine-scale movement, including accounting for multiple movement sates (e.g., traveling, grazing, resting, hunting, etc.).
-this would help deal with over-dispersion
 * encounter rates (lower encounter rates with food [@hou_cold_2020])
-* HWI
+* HWI [@weststrate_how_2024]
+z* Machine Learning (Genenative Aritificial Intelligence; GenAI) AI is fueling climate change.
+We need to proactively prepare to respond to climate change.
+Pollution is not under the control of the masses, so predicting change will be hard
+* high energy and water demands [@li_making_2023] GenAI consumption of water
+* high carbon footprint [@luccioni_estimating_2024]
+* Pollution is not under control of the masses, so predicting change will be hard
 

### 2025-01-17 analysis/06-temperature-hrsfs.R
clarified comment on `s(temp)`

@@ -325,7 +325,7 @@ for(sp in as.character(SPECIES)) {
       ti(forest_perc, temperature_C, k = 6, bs = 'tp') +
       ti(elevation_m, temperature_C, k = 6, bs = 'tp') +
       ti(dist_water_m, temperature_C, k = 6, bs = 'tp') +
-      # include marginals of temperature to post-stratify over afterwards
+      # include marginals of temperature to remove sampling biases
       s(temperature_C, k = 4, bs = 'tp') +
       s(temperature_C, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')),
     family = poisson(link = 'log'),

### 2025-01-17 analysis/00-crop-resource-rasters.R
fixed file paths

@@ -29,11 +29,11 @@ plot(locs_ext, add = TRUE, lwd = 3)
 # forest raster (original rasters from http://www.earthenv.org/landcover)
 f1 <- rast('data/resource-rasters/uncropped-rasters/consensus_full_class_1.tif') %>%
   crop(dem)
-f2 <- rast('~/../Downloads/consensus_full_class_2.tif') %>%
+f2 <- rast('data/resource-rasters/uncropped-rasters/consensus_full_class_2.tif') %>%
   crop(dem)
-f3 <- rast('~/../Downloads/consensus_full_class_3.tif') %>%
+f3 <- rast('data/resource-rasters/uncropped-rasters/consensus_full_class_3.tif') %>%
   crop(dem)
-f4 <- rast('~/../Downloads/consensus_full_class_4.tif') %>%
+f4 <- rast('data/resource-rasters/uncropped-rasters/consensus_full_class_4.tif') %>%
   crop(dem)
 
 layout(matrix(1:4, ncol = 2))

### 2025-02-24 writing/manuscript.Rmd
finished first complete draft of the manuscript

@@
 ---
-title: "Predicting how climate change will affect terrestrial mammals' movement and habitat selection in British Columbia, Canada"
+title: "Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada"
 author:
@@
   - name: Karen E. Hodges
-    institute: [braes, biol]
-  - name: Siobhan Darlington
     institute: [braes, biol]
@@
     - \usepackage[left]{lineno} # for line numbers
+    - \usepackage{booktabs}
 subparagraph: true # needed for \usepackage[small]{titlesec}
@@
 Previous research has quantified the effects of temperature on mammalian movement behavior, but few have quantified the effects of future climate change on fine-scale movement and habitat selection.
-We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on mammal's probability of movement, speed, and habitat selection.
-We then leverage the estimated effects to predict when, how, and where mammals will move in BC throughout the 21^st^ century.
-We conclude with implications for energetics, encounter rates, habitat conservation, and human-wildlife conflicts.
-***MISSING RESULTS AND CONCLUSIONS***"
+We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on mammals' movement rates and habitat selection.
+We then leverage the estimated effects to predict how and where mammals will move in British Columbia (BC), Canada, throughout the 21^st^ century.
+We show that changes in temperature altered movement phenology, and that climate change will impact mammals' movement rates and habitat selection throughout the century, but that there are no common trends across all study species.
+Consequently, we suggest that species' responses be interpreted carefully and considered jointly when extrapolating the community-level consequences of climate change."
 ```
@@
 
-\noindent \textbf{Key words:} climate change, temperature, mammals, animal movement, habitat selection
+\noindent \textbf{Key words:} climate change, temperature, mammals, animal movement, movement behavior, habitat selection
 
@@
 * Global vulnerability of marine mammals to global warming: https://doi.org/10.1038/s41598-019-57280-3
+* boars' sleep quantity, efficiency and quality are reduced on warmer days (https://doi.org/10.1098/rspb.2023.2115)
 
@@
 
-<!-- movement is a convenient way to respond to change -->
+<!-- movement is a convenient way to respond to change, but movement depends on temperature -->
 
-\noindent For most animals, movement is a rapid, low-cost, and low-risk response to a variety of stimuli [@nathan_movement_2008], including escaping predation [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024], searching for resources [@charnov_optimal_1976; @mezzini_how_2024], and habitat selection [@kacelnik_ideal_1992; @winter_forecasting_2024].
-However, movement efficiency depends on ambient temperature, since it impacts the energetic cost of movement [@taylor_energetics_1982; @brown_toward_2004; @jahn_variations_2022], the ease of movement [e.g., snow depth and density, see @montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023].
-Additionally, temperature also affects movement indirectly by affecting other drivers of movement behavior, including physiology [@abram_behavioural_2017], fitness [@schmidt_interplay_2020; @berger_climate_2018], trophic interactions [@hegel_interacting_2010], and the availability and value of resources [e.g., water and heat, see: @elmore_implications_2017; @attias_effects_2018].
+\noindent For most animals, movement is a rapid, low-cost, and low-risk response to a variety of stimuli [@nathan_movement_2008].
+Animals move to maximize resource availability [e.g., food, water, shelter, see: @charnov_optimal_1976; @kacelnik_ideal_1992], optimize habitat use [@winter_forecasting_2024], and escape predation [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024].
+However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], the ease of movement [e.g., snow depth and density, see @montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023].
+Additionally, temperature also affects movement indirectly by through other drivers of movement behavior, such as trophic interactions [@hegel_interacting_2010] and the value of resources [e.g., water, heat, shelter -- see: @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019].
+Changes in temperature thus affects how animals move [e.g., movement frequency and speed -- their movement behavior: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].
 
-<!-- changes in temperature cause changes in movement -->
-
-Changes in temperature affect not only whether and how fast animals move but also how they move [i.e., speed and tortuosity; their movement behavior: @nathan_movement_2008; @dyer_travel_2023] and where they move to [habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].
-Thus, temperature affects not only an animal's energetic expenditure but also many other consequences of movement, including encounter rates with resources [@mezzini_how_2024], other animals [@martinez-garcia_how_2020], and the occurrence of human-wildlife conflict [@weststrate_how_2024].
-These changes are exacerbated at extreme temperatures, as animals often alter their movement behaviors substantially when temperatures are excessively cold or hot [@wilson_stenotherms_2001; @berger_climate_2018; @arechavala-lopez_common_2019; @giroux_activity_2023].
-But what counts as "extreme" varies across species: thermal specialists are more sensitive to small changes above or below their ideal range [@giroux_activity_2023], while temperature generalists can adapt more easily to change [@dupont_beyond_2024].
-However, this grater adaptability is tested when generalists are subject to changes that are more severe [@mccain_assessing_2019], affecting many factors [e.g., temperature, phenology, and resource availability: @polazzo_measuring_2024], or too rapid to adapt to within the span of a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].
-Thus, it remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially when populations do not react to changes in environmental conditions [@sawyer_migratory_2019] or are physiologically incapable [@sherwood_adaptability_2010; @williams_range_2018].
+Temperature also affects the consequences of animal movement, such as encounter rates with resources [@mezzini_how_2024], other animals [@martinez-garcia_how_2020], and humans [@weststrate_how_2024], and these changes are exacerbated at extreme temperatures [@wilson_stenotherms_2001; @berger_climate_2018; @giroux_activity_2023].
+But what counts as "extreme" varies across species: thermal specialists are more sensitive to small changes outside their typical range [e.g., @giroux_activity_2023], while temperature generalists can adapt more easily to change [@dupont_beyond_2024].
+However, generalists' grater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019], affect many factors [e.g., temperature, phenology, and resource availability: @polazzo_measuring_2024], or are too rapid to adapt within the span of a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].
+Thus, it remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially when populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018].
 
@@
 
-This uncertainty in both future conditions and their consequences on ecosystems and populations presents a need for a comprehensive understanding of the effects of temperature on the movement of animal populations as well as the consequences this will have on landscapes, biodiversity, and their adaptability and resiliency.
-Recent work has documented climate-change-related changes mammals' ranges [@leclerc_determinants_2021], movement behavior [@melin_effect_2023]; thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [@mccain_assessing_2019], but it is still unclear how mammals will adapt their small-scale movement and habitat selection to future changes in temperature [@intergovernmental_panel_on_climate_change_climate_2023; but see @hetem_responses_2014; @winter_forecasting_2024].
+Uncertainty in future conditions and their consequences on ecosystems presents a need for a comprehensive understanding of (1) the direct effects of temperature on animal movement behavior and (2) the implications this will have on populations and species' adaptability and resiliency, community structure, and biodiversity at large.
+Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], movement behavior [@melin_effect_2023], thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [@mccain_assessing_2019], but it is still unclear how mammals will adapt their small-scale movement and habitat selection to changes in temperature during the 21^st^ century [@intergovernmental_panel_on_climate_change_climate_2023; but see @hetem_responses_2014; @winter_forecasting_2024].
 In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures and severe forest fires [@intergovernmental_panel_on_climate_change_climate_2023; @zurowski_summer_2023].
-Over the next 100 years, these changes will worsen and continue to impact mammals' fitness, movement behavior, and habitat selection as they cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and frequency and intensity of extreme events [@bunnell_global_2011] along with growing anthopogenic pressure [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024].
+Over the next 100 years, these changes are expected to worsen and continue to impact mammals' fitness, movement behavior, and habitat selection as animals cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and more frequent and intense extreme events [@bunnell_global_2011], along with increasing anthropogenic pressure [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024].
 
@@
 
-Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to be able to prepare and respond to change in a proactive manner [e.g., @sawyer_identifying_2009; @mccain_assessing_2019].
-The Canadian province of British Columbia is currently experiencing rapid widespread warming, phenological shifts, and more frequent and intense extreme events, including forest fires [@zurowski_summer_2023].
-In light of the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022], we stress that decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness.
+Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to be able to prepare for and respond to change in a proactive manner [e.g., @sawyer_identifying_2009; @mccain_assessing_2019].
+The Canadian province of British Columbia is currently experiencing rapid widespread warming [@dierauer_climate_2021; @turner_its_2009], phenological shifts [e.g., @tysor_countergradient_2025; @basu_phenological_2024], and more frequent and intense extreme events [e.g., @zhang_increased_2023], including forest fires [@zurowski_summer_2023].
+As we approach the deadline for the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022], decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness.
 Predicting how climate change will affect how, when, and where animals will move is necessary for identifying which landscapes are most valuable now as well as in the future.
@@
 
-This paper provides an analysis of the effects of spatiotemporally proximal temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (*Canis lupus*, *Cervus canadensis*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, and *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
-Using over two decades of telemetry data over a large spatial range of British Columbia (Fig.
-\@ref(fig:bc-map)) and hourly air temperature data, we estimate how mammals altered their movement frequency, movement speed, and habitat selection in response to temperature.
-We then pair the estimated responses with climate change projections to forecast changes in mammalian movement in the current century under different climate-change scenarios [Shared Socioeconomic Pathways, see @riahi_shared_2017].
-We discuss the consequences of these changes in movement behavior with regards to energetics, encounter rates, habitat conservation, and human-wildlife conflicts while disentangling the direct and indirect effects of temperature on the movement behavior of terrestrial mammals.
+This paper provides an analysis of the effects of air temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (*Canis lupus*, *Cervus canadensis*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, and *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
+Using over 25 years of telemetry data thoughout a large spatial range of British Columbia (Fig.
+\@ref(fig:bc-map)) and hourly air temperature data, we estimate how mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature.
+We then pair the estimated responses with climate change projections to forecast the species' movement during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSPs -- see @riahi_shared_2017].
+We disentangle the direct and indirect effects of temperature on the movement behavior of terrestrial mammals and discuss the consequences of these changes at the individual, population, and landscape levels.
 
-```{r bc-map, fig.cap="GPS telemetry data for the six species in this study."}
+```{r bc-map, fig.cap="GPS telemetry data for the six species in this study.", out.width='0.75\\linewidth'}
 knitr::include_graphics('../figures/tels-map.png')
@@
 
-knitr::kable(dt_data, format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$; stratified by individual), number of animals (Animals), and number of animals with unresolved speed (NA speed) for each of the species in this study.', escape = FALSE)
+knitr::kable(dt_data, format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$; stratified by individual), number of animals (Animals), and number of animals with unresolved speed (NA speed) for each of the species in this study.', escape = FALSE, longtable = FALSE, booktabs = TRUE)
 ```
@@
 
-## Datasets
+## Datasets used
 
@@
 \noindent We estimated percent forest cover and distance from water using the rasters created by @tuanmu_global_2014.
-We calculated total forest cover by summing the temporally static rasters of evergreen/deciduous needleleaf trees, evergreen broadleaf trees, deciduous broadleaf trees, and mixed/other trees (raster classes 1-4, respectively).
+We calculated total forest cover by summing the temporally static rasters of evergreen/deciduous needle-leaf trees, evergreen broadleaf trees, deciduous broadleaf trees, and mixed/other trees (raster classes 1-4, respectively).
 We converted the raster of percent cover of open water (class 12) to a binary raster of pixels with water (cover > 0%) or without water (cover = 0%) and then calculated each pixel's distance from the nearest pixel with water using the `distance()` function from the `terra` package.
-Finally, we obtaied two digital elevation models using the `get_elev_raster()` function from the `elevatr` package [version 0.99.0, @hollister_elevatr_2023]: We used a raster with a zoom of 6 (0.009$^\circ$) for model fitting and one with a zoom of 3 (0.08$^\circ$) for downloading climate change projections via `climatenaR`.
+Finally, we obtained two digital elevation models using the `get_elev_raster()` function from the `elevatr` package [version 0.99.0, @hollister_elevatr_2023]: We used a raster with a zoom of 6 (0.009$^\circ$) for model fitting and one with a zoom of 3 (0.08$^\circ$) for downloading climate change projections via `climatenaR`.
 All final rasters and code are available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).
@@
 Since `ctmm`'s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed could be 0.
-We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed.<!-- The degrees of freedom and Frequentist confidence intervals were adjusted accordingly.--> The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R).
+We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed.
+The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R).
 To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using the inflection point of each species' histogram of detrended speed estimates (Appendix A).
@@
 
-### Movement frequency and speed
+### Effects of temperature on movement rates
 
@@
 
-<!-- \clearpage -->
+\clearpage
 
@@
 
-\small
+\scriptsize
 
@@
 
-\clearpage
-
 \normalsize
@@
 
-### Habitat selection
+<!-- \clearpage -->
+
+### Effects of temperature on habitat selection
 
@@
 
-\noindent Species differed in mean probabilities of movement (range: 0.05 -- 0.27), mean speed when moving (range: 0.42 -- 2.67 km/day), and mean distance travelled (range: 0.04 -- 0.62 km/day) even after post-stratifying to a sampling interval of one hour (Tab.
-\@ref(tab:gam-intercepts)).
-The inclusion of temperature in the model did not affect any of the three measures substantially when post-stratifying to $0^\circ$C.
-Grizzly bears tended to move least often ($P(M = 1) \approx 0.05$), while wolves and pumas moved most often ($P(M = 1) \ge 0.23$), and wolves had the highest mean speed when moving.
-Consequently, wolves traveled 3.0 to 15.6 times further per day than other mammals, on average.
-
-* CHANGED FOREST RASTER TO USE ALL TREE TYPES (2025-01-17)
-* NEED TO REMAKE HRSF FIGURES
-
-***THIS IS FALSE***We did not find any consistent selections for or against forest cover, but all species exhibited stronger selection preferences for elevation and distance from water than for forest cover, and there was no consistent preference for forest cover across temperatures.
-In contrast, all species showed a strong preference for specific altitudes: wolves, pumas, and boreal caribou generally preferred elevations < 1 km above sea level, while elk, mountain goats, southern mountain caribou, and grizzly bears generally preferred elevations between 1 and 2 km above sea level.
-All species except for elk and grizzly bears preferred being as close to water as possible,
-
-Overall, we did not detect any strong selections for or against forest cover, and this lack of preference was generally true irrespective of temperature.
-In contrast, we selection strength for forest cover did not vary strongly with temperature,
-
-preferred resources
-
-**CHECK TABLE 2: UPDATE AFTER REMOVING SCEK014 AND SCEK014b (2024-11-19 9:50)**
-
-***UPDATE TABLES BASED ON NEW HRSFS***
+\noindent Species differed in mean probabilities of movement (range: 0.05 -- 0.3), mean speed when moving (range: 0.42 -- 2.7 km/day), and mean distance traveled (range: 0.04 -- 0.6 km/day), even after predicting for one-hour time intervals for all species (Table \@ref(tab:gam-intercepts)).
+Grizzly bears tended to move least often (probability of moving $P(M) \approx 0.05$), while wolves and pumas moved most often ($P(M) \gtrapprox 0.23$).
+When moving, mountain goats and southern mountain caribou moved the slowest ($\mathbb E(S|M) \approx 0.4$ m/s), while wolves had the highest mean speed when moving ($\mathbb E(S|M) \approx 2.7$ m/s).
+Consequently, wolves traveled, on average, 50 km/day -- 2.5 to 16.7 times further than other mammals.
 
@@
                      'P(M|T)' = get_int(readRDS('models/binomial-gam.rds')),
-                     'E(S|M=1)' = 'km/h' %#% get_int(readRDS('models/gamma-gam-without-temperature.rds')),
-                     'E(S|M=1,T)' = 'km/h' %#% get_int(readRDS('models/gamma-gam.rds')),
-                     'E(D)' = `P(M)` * `E(S|M=1)`,
-                     'E(D|T)' = `P(M|T)` * `E(S|M=1,T)`) %>%
+                     'E(S)' = 'km/h' %#% get_int(readRDS('models/gamma-gam-without-temperature.rds')),
+                     'E(S|T)' = 'km/h' %#% get_int(readRDS('models/gamma-gam.rds')),
+                     'E(D)' = `P(M)` * `E(S)`,
+                     'E(D|T)' = `P(M|T)` * `E(S|T)`) %>%
   mutate(Species = paste0('\\emph{', Species, '}') %>%
@@
   hist(intercepts$`P(M)`)
-  hist(intercepts$`E(S|M=1)`)
-  round(range(max(intercepts$`E(D)`) /
-                intercepts$`E(D)`[- which.max(intercepts$`E(D)`)]),
-        1)
+  hist(intercepts$`E(S)`)
+  round(range(max(intercepts$`E(D)`) / intercepts$`E(D)`[- which.max(intercepts$`E(D)`)]), 1)
 
@@
               p = round(`P(M|T)` - `P(M)`, 2),
-              s = round(`E(S|M=1,T)` - `E(S|M=1)`, 2),
+              s = round(`E(S|T)` - `E(S)`, 2),
               d = round(`E(D|T)` - `E(D)`, 2))
+
+  intercepts %>%
+    arrange(desc(`E(D)`)) %>%
+    mutate(displ_wolf = `E(D)`[grepl('Canis lupus', Species)] / `E(D)`,
+           displ_wolf_t = `E(D|T)`[grepl('Canis lupus', Species)] / `E(D|T)`)
 }
@@
   rename_with(\(.x) gsub('E', '\\\\mathbb E', .x)) %>%
-  rename_with(\(.x) gsub('M=1', 'M\\\\!=\\\\!1', .x)) %>%
-  knitr::kable(format = 'markdown', digits = 2, caption = paste0("Mean probability of movement (P(M)), speed when moving ($\\mathbb E(S | M = 1)$), and distance travelled ($\\mathbb E(D)$) as estimated by models without and with temperature, after post-stratifting to a 1-hour sampling rate and a temperature of T = 0\U00B0", "C."))
+  knitr::kable(format = 'markdown', digits = 2, caption = paste0("Mean probability of movement (P(M)), speed when moving ($\\mathbb E(S)$), and distance travelled ($\\mathbb E(D)$) as estimated by models without and with temperature ($(...
+| T)$), after post-stratifting to a 1-hour sampling rate and a temperature of T = 0\U00B0", "C."))
 ```
@@
 
-For most models, including smooth terms of temperature did not always improve within-sample prediction substantially (change in deviance explained: +0.2% -- + 17.9%), and there was relatively strong agreement between models with and without temperature (Fig.
-***SX pred vs pred***), but including temperature resulted in substantially better fits for all models (all $\Delta$AIC $\le$ -326; Table \@ref(tab:temp-comparison)).
+At average temperatures, wolves preferred dense forest cover ($\gtrapprox$ 50%) high elevations, and distances from water < 10 km; elk preferred intermediate forest cover, elevations between 1 and 2 km, and distances from water of 5-15 km; mountain goats preferred sparse (< 50%) forest cover, elevations between 1 and 2 km; pumas preferred high forest cover, an elevation of ~ 1 km, and distances from water < 10 km; boreal caribou selected for intermediate to high forest cover, elevations near 500 m, and distances from water < 10 km; southern mountain caribou selected for dense forest cover, elevations near 2 km, and distances from water < 5 km; and grizzly bears selected for sparse forest cover (25-50%), elevation near 1 km, and distances from water < 2 km.
+
+Including temperature in the models always increased within-sample prediction, but the change was not always substantial (deviance explained increased between 0.2% and 18.6%, with all HRSFs having an increase $\ge 4.6$), and there was relatively strong agreement between models with and without temperature (Fig.
+***SX pred vs pred***), but including temperature resulted in substantially better fits for all models, especially HRSFs (all $\Delta$AIC $\le$ -342; Table \@ref(tab:temp-comparison)).
 
@@
 library('dplyr')
-source('analysis/figures/default-ggplot-theme.R') #' for `SPECIES`
+source('analysis/figures/default-ggplot-theme.R') #' for `SPECIES` object
 
@@
   DDE = 
-    c(10.7 - 10.2, 79.3 - 79.1, 10.0, 4.47, 17.6, 7.83, 5.89, 11.8, 17.9) %>%
+    c(10.72 - 10.23, 79.26 - 79.11, 9.16, 4.62, 18.6, 8.0, 5.30, 12.0, 17.7) %>%
     round(1) %>%
     paste0('%'),
-  DAIC = c(-5927, -1682, -7200, -4305, -769, -945, -6284, -326, -920) %>%
+  DAIC = c(-5927, -1682, -6566, -4448, -813, -966, -5637, -342, -914) %>%
     round())
+
+if(FALSE) {
+  range(as.numeric(gsub('%', '', temp_comp$DDE)))
+  range(as.numeric(gsub('%', '', temp_comp$DDE))[- c(1:2)])
+  max(temp_comp$DAIC)
+}
 
@@
 
-knitr::kable(temp_comp, digits = 1, format = 'markdown', caption = "Changes in deviance expained ($\\Delta$DE) and Akaike Information Criterion ($\\Delta$AIC) from including temperature in the models estimating probability of movement (P(moving)) and speed when moving (Speed), as well as in species-level Hierarchical Resource Selection Functions (HRSF).")
+knitr::kable(temp_comp, digits = 1, format = 'markdown', caption = "Changes in deviance expained ($\\Delta$DE) and Akaike Information Criterion ($\\Delta$AIC) from including temperature in the models estimating probability of movement (P(moving)) and speed when moving (Speed) as well as in species-level Hierarchical Resource Selection Functions (HRSF).")
 ```
@@
 
-\noindent The partial effects of temperature varied in both direction and magnitude across species (Fig.
+```{r, eval=FALSE}
+library(mgcv)
+# to  get deviance explained
+summary(readRDS('models/binomial-gam.rds'), re.test = FALSE)$dev.expl
+summary(readRDS('models/gamma-gam.rds'), re.test = FALSE)$dev.expl
+```
+
+\noindent The marginal effects of temperature varied in both direction and magnitude across species (Fig.
 ***AX***), even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; see Fig.
 ***AX***).
-Smooth interaction terms were well-behaved and indicated clear shifts in daily activity for all species (Fig.
+The smooths' credible intervals were generally quite wide outside the typical temperature range of -20$^\circ$C to 20$^\circ$C, so the estimated trends were often not credible at the 95% level outside such range.
+Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species.
+The models explained reasonably high proportions of the deviance (10.7% for the binomial model and 79.3% for the Gamma model) and had good in-sample prediction (Fig.
 ***AX***).
-The models explained reasonably high proportions of the deviance (11% for the binomial model and 79% for the Gamma model) and had good in-sample prediction (Fig.
+All species' movement frequency and speed decreased with increasing sampling interval (Fig.
 ***AX***).
-Both movement frequency and speed when moving decreased with increasing sampling interval, with a strong common effect (Fig.
+
+Most species adapted their daily and seasonal movement behavior to changes in temperature (Fig.
+\@ref(fig:dist)).
+Wolves, elk, pumas, and boreal caribou reduced their mid-day movement rates when temperatures were above 20$^\circ$C, and wolves, elk, and boreal caribou moved less on summer days above 20$^\circ$C.
+However, the precision of the estimated change depended on the GPS sampling frequency (Fig.
 ***AX***).
-Most species adapted their daily and seasonal movement behavior to changes in temperature (Fig.
-\@ref(fig:dist)), but the precision of the estimated change depends on the GPS sampling frequency.
-While `ctmm` models are scale independent [@noonan_scale-insensitive_2019], the accuracy, size, and stability of speed estimates still depends on the GPS sampling frequency.
-This is because coarser sampling contains information on large scale movements (e.g., range crossings, migrations) but not fine-scale movement (e.g., first-order habitat selection).
+While the `ctmm` models produced scale-independent estimates of speed [i.e., model interpretation is independent of sampling time: @noonan_scale-insensitive_2019], the accuracy, size, and stability of speed estimates still depended on the GPS sampling frequency.
+This is because coarsely-sampled movement data contains information on large-scale movements (e.g., range crossings, migrations) but not fine-scale movements [e.g., first-order habitat selection *sensu* @johnson_comparison_1980].
 Using the boreal caribou as an example, the 13-hour sampling interval allows us to reasonably estimate the caribou's movement path at a temporal scale of approximately 13 hours (or greater), but we cannot produce reasonable movement trajectories at a much finer (e.g., hourly) scale.
-Consequently, the estimated movement behaviors should be compared across species carefully.
+Consequently, we suggest being cautious when comparing estimated movement behaviors across species.
 
@@
 Predictions in the surface plots extend to 10% of the range away from each datum.
-The color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to -75\\% and +300\\% for ease of readability."}
+The color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to -75\\% and +300\\% (i.e., $2^{\\pm 2}$) for ease of readability."}
 knitr::include_graphics('../figures/distance.png')
@@
 
-### Predicted changes in movement during the current century
+## Effects of temperature on habitat selection
 
-The direction and degree of change in distance travelled varies greatly between species, but worse climate-change scenarios consistently corresponded to greater absolute changes in behavior as well as greater uncertainty in the estimated change (Fig.
-\@ref(fig:t-dist)).
-Under the best-case scenario, absolute projected changes by 2100 were small (0% to 2%, except for southern mountain caribou, which had a change of ~5%).
-Under the worst-case scenario, absolute projected changes were by 2100 ranged from ~1% (boreal caribou) to ~25% (southern moutain caribou) and had the greatest uncertainty in the estimates.
-Boreal caribou and elk showed relatively little projected changes, especially relative to the degree of inter-annual variability in distance travelled.
+Species' relative selection strength (RSS) for resources depended on temperature, and RSS was strongest for elevation and weakest for forest cover.
+Changes in RSS with temperature were strongest for elevation and generally weakest for distance from water, but there were no common trends across all species for any of the three resources.
+All species, with the exception of pumas, exhibited a temperature-dependent shift in their preference for forest cover.
+At higher temperatures, wolves, mountain goats, and grizzly bears became less selective for forest cover, while elk and caribou became more selective.
+All species shifted elevationally with temperature, although boreal caribou did not exhibit as strong of a change.
+As temperatures rose, elk, mountain goats, cougars, and boreal caribou increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation.
+Most species generally remained within 5 km of water, and temperature did not affect their selection strength as much as for the other two resources.
+Estimated RSS values were generally most uncertain at extreme temperatures (Fig.
+***BX***).
 
-```{r, eval=FALSE}
-# max absolute change
-readRDS('H:/GitHub/bc-mammals-temperature/data/cc-hgam-projections.rds') %>%
-  group_by(scenario, species) %>%
-  summarize(max_rel = labeller_perc(max(abs(d))),
-            .groups = 'drop') %>%
-  arrange(species, scenario) %>%
-  print(., n = nrow(.))
-
-# range between values at 2100 does not depend strongly on dt
-# (without accounting for other variables and trends within species)
-readRDS('H:/GitHub/bc-mammals-temperature/data/cc-hgam-projections.rds') %>%
-  filter(year == 2100) %>%
-  group_by(species) %>%
-  summarize(range = labeller_perc(diff(range(d)) + 1)) %>%
-  left_join(dt_data %>%
-              mutate(species = case_when(
-                Dataset == '\\emph{Canis lupus}' ~ 6,
-                Dataset == '\\emph{Rangifer tarandus} (boreal)' ~ 5,
-                Dataset == '\\emph{Rangifer tarandus} (s. mountain)' ~ 1,
-                Dataset == '\\emph{Puma concolor}' ~ 3,
-                Dataset == '\\emph{Puma concolor}' ~ 3,
-                Dataset == '\\emph{Cervus canadensis}' ~ 4,
-                Dataset == '\\emph{Oreamnos americanus}' ~ 7,
-                Dataset == '\\emph{Ursus arctos horribilis}' ~ 2),
-                species = SPECIES_LABS[species]),
-            by = 'species') %>%
-  ggplot(aes(`Median Δt (hours)`, range, color = species)) +
-  geom_point() +
-  scale_color_manual(values = PAL)
+```{r hrsfs, fig.cap="\\textbf{Temeperature strongly affects mammals' habitat selection.} Estimated relative selection strength (RSS) for forest cover (\\%), elevation (km), and distance from water (km) as a function of temperature.
+The grey contours indicate the extent of each speices' observed locations.
+RSS values were re-centered and capped to $2^{\\pm 2}$ to improve readability."}
+knitr::include_graphics('../figures/hrsf-surface-plots.png')
 ```
 
-```{r t-dist, fig.cap="\\textbf{The intensity and direction of changes in movement rates due to climate change varies among species, but worse climate-change scenarios often result in the greatest change and greatest uncertainty.} Lines indicate the projected relative change in mean distance travelled (probability of moving times speed when moving) due to changes in temperature within the species' current extent.
-Shaded areas indicate the mean 95% Bayesian credible intervals.
-Changes are relative to the mean distances travelled in 2020 across the four Shared Socioeconomic Pathways (SSPs).
-The SSPs are listed in order of similarity to current conditions, from SSP 1-2.6 (sustainability, best-case scenario) to 5-8.5 (fossil-fueled development, worst-case scenario).
+## Predicted changes in movement during the 21^st^ century
+
+\noindent The direction and magnitude of change in distance traveled varied greatly between species, but worse climate-change scenarios generally corresponded to greater absolute changes in behavior (Fig.
+\@ref(fig:t-dist)).
+Under the best-case scenario, absolute projected changes by 2100 were small (approximately 0% to 2%), while absolute projected changes by 2100 ranged from ~1% (boreal caribou) to ~25% (southern mountain caribou) under the worst-case scenario.
+Boreal caribou and elk showed relatively little projected changes in yearly distance traveled.
+Throughout BC  (Fig.
+\@ref(fig:s-dist)), all species but elk and boreal caribou were expected to change their movement rates by 10 to 20% in 2100, relative to 2025, but the predicted changes varied spatially due to heterogeneity in climate change throughout the province.
+Across all species, absolute changes were smallest under the best-case scenario and greatest under the worst-case scenario.
+
+```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios often result in the greatest change and greatest uncertainty.} Lines indicate the projected relative change in mean distance travelled (probability of moving times speed when moving) due to changes in temperature within the species' current extent.
+Shaded areas indicate the average 95% Bayesian credible intervals for the area.
+Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs).
 The projections only account for changes movement behavior (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
@@
 
-***COMMENTS ABOUT SPATIAL VARIATION IN distance travelled***
-
-```{r s-dist}
-plot(1, main = '4. spatial projections of distance in 2020 and 2100')
+```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each location's average distance in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase.
+For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.8 and 1.25 ($\\approx 2^{\\pm 0.32}$; original data ranged 0.81 to 1.35).
+The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest cover."}
+knitr::include_graphics('../figures/bc-displ-2100.png')
 ```
 
-## Effects of temperature on habitat selection
+All species were projected to reduce selection for their current range throughout the 21^st^ century, with the exception of grizzly bears, which showed little to no change under all four the climate change scenario (\@ref:fig(t-hrsfs)).
+The decreases in RSS were most evident in the habitats with the top 5% RSS in the current range.
+Projected changes in the median-RSS habitats were relatively small, but none increased throughout the century under any of the climate change scenarios.
+As with distance traveled, the change in RSS in 2100 varied throughout BC for all species (Fig.
+\@ref(fig:s-hrsfs)).
+All species were predicted to exhibit a decrease in RSS throughout most of BC, with the exception of boreal caribou, which had a predicted decrease in approximately 37% of the province, including their current range.
+While the boreal caribou were expected to increase selection for altitudes > 1 km above sea level, the estimated selection strength was highly uncertain and not credible at the 95% level (Fig.
+***SX***).
 
-```{r hrsfs, fig.cap="\\textbf{Temeperature strongly affects mammals' habitat selection.} Estimated resourse selection strength for forest cover (\\%), elevation (km), and distance from water (km) as a function of temperature.
-The grey contours indicate the extent of each speices' observed locations."}
-knitr::include_graphics('../figures/hrsf-surface-plots.png')
+```{r, eval=FALSE}
+library('dplyr')
+readRDS('data/cc-rss-bc-projections-albers.rds') %>%
+  group_by(lab, scenario) %>%
+  summarize(p_decrease = mean(lambda < 1)) %>%
+  filter(p_decrease < 0.5)
 ```
 
-### Predicted changes in movement during the current century
-
-```{r t-hrsfs, fig.cap="\\textbf{The intensity and direction of changes in habitat selection due to climate change vary among species, but worse climate-change scenarios consistently result in the greatest change and greatest uncertainty.} Lines indicate the projected change in relative selection strength for forest cover (\\%), elevation (km), and distance from water (km) due to changes in temperature within the species' current extent.
-Shaded areas indicate the mean 95% Bayesian credible intervals.
-Changes are relative to the relative selection strength in 2020 across the four Shared Socioeconomic Pathways (SSPs).
-The SSPs are listed in order of similarity to current conditions, from SSP 1-2.6 (sustainability, best-case scenario) to 5-8.5 (fossil-fueled development, worst-case scenario)."}
-plot(1, main = '6. temporal projections of RSFs within current area')
+```{r t-hrsfs, fig.cap="\\textbf{The intensity and direction of changes in habitat selection due to climate change vary among species, but no species are expected to exhibit an increase in relative selection strength (RSS).} Lines indicate the projected change in RSS based on the 5$^{th}$ percentile (the worst 5%; bottom line), the median (the middle; bold), and 95% percentile (the best 5%; top line) habitats within the species' current extent.
+Changes are relative to the median RSS in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
+knitr::include_graphics('../figures/rss-local-cc-predictions.png')
 ```
 
-```{r s-hrsfs}
-plot(1, main = '7. spatial projections of RSFs in 2020 and 2100')
+```{r s-hrsfs, fig.cap="\\textbf{Climate change will impact each species' habitat relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase.
+For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.6}$; original data ranged 0.26 to 117)."}
+knitr::include_graphics('../figures/bc-rss-2100.png')
 ```
@@
 
-We demonstrated that temperature is an important determinant of how and where mammals move, even while accounting for other seasonal drivers of movement (e.g, changes in photoperiod, hormones, and resources).
-This is important because decoupling temperature from other drivers allows us to predict how changes in climate will affect mammalian movement even if all other drivers remain constant.
+\noindent We demonstrated that temperature is an important determinant of how and where mammals move, even while accounting for other seasonal drivers of movement (e.g, changes in photoperiod, hormones, and resources).
+Disentangling temperature from other drivers allows us to predict how changes in climate will affect mammalian movement even if all other drivers remain approximately constant or adapt at negligible rates.
+However, as habitats warm and animals are exposed to increasingly frequent extreme and novel conditions, our ability to confidently predict mammals' responses decreases.
+Predicting at the fringes of the surface plots we present comes with substantial uncertainty, particularly given that the responses are nonlinear.
+At unusually warm temperatures, mammals (and other homeotherms) overheat, so their movement behavior and that of their competitors, predators, and/or prey can often be substantially different from the behavior at typical temperatures.
+As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], mammals will be increasingly forced to atypical behaviors, which will have the potential to substantially alter community structures and behaviors, both during such events and afterwards [@zhang_increased_2023; ].
+For instance, changes in climate and phenology impact the life history and behavior of many hibernating mammals [@wells_life_2022], but hot weather also affects mammals' sleep quality [e.g., wild boars, *Sus scrofa*: @mortlock_sleep_2024] and likelihood to enter torpor [Australian eastern long-eared bat, *Nyctophilus bifax*: @fjelldal_nightly_2021].
+These changes can alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024].
+At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase their ability to find food or be predated upon [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021] as well as the overall ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023].
+However, these changes are likely to have complex consequences for population and ecosystem structures and dynamics as prey, predators, and prey experience altered and novel seasonal cycles.
+especially given the increasingly common climate "weirding" [@bunnell_global_2011], which includes not only warmer temperatures and more frequent and intense heat waves but also an increase in various other extreme events [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], such as cold spells, floods and droughts, plant and animal die-offs and sudden growth surges, and other changes in population dynamics and behaviors [sometimes referred to as "black swan" events -- see: @logares_black_2012; @anderson_black-swan_2017].
 
-* We discuss the consequences of these changes in movement behavior with regards to energetics, encounter rates, habitat conservation, and human-wildlife conflicts.
-* as temperatures get warmer, we will be pushing into the unknown portion of the surface plots
-* our models don’t include any physiological responses to temperature that may impact movement (e.g., hybernation)
-* our models also assume the behaviour we modeled now will persist into the future (e.g., hybernation, migration)
-* based on the model, temperature doesn't affect grizzlies' hibernation much, and their distance travelled in winter is very low due to P(moving) being so low:
-* if the RSS decreases a lot, individuals may disperse, populations may relocate, or stay where they are and suffer
+Our ability respond to current and future changes in climate is contingent on our ability to prepare for and predict change.
+However, predicting animal behavior becomes increasingly complicated as the conditions animals are exposed to deviate from current, typical conditions, especially when responses are nonlinear.
+Consequently, we do not present our results as a definitive guide to how mammals in BC will respond to climate change.
+Instead, we hope they serve as a starting point to (1) demonstrate that mammals' movement rates and habitat selection depend on temperature and (2) how one can estimate mammals' changes in movement behavior due to climate change.
+An important limitation of Figs.
+\@ref(fig:dist) and \@ref(fig:hrsfs) is that we do not include measures of uncertainty around the estimated responses, as they cannot be incorporated easily in surface plots, and the purpose of the figures was to illustrate the magnitude and direction of the responses rather than contrasting the surfaces against some hypotheses.
+Still, we stress that uncertainty is crucial in producing transparent and informed results, particularly when using such estimates for decision-making or predicting future changes, as we do in Figs.
+\@ref(fig:t-dist) and \@ref(fig:t-hrsfs).
+We present the spatial estimates of distance traveled and relative habitat selection strength in 2100 without measures of uncertainty for conciseness, but we suggest that future studies and reports include some form of compatibility intervals [@gelman_are_2019], whether these be Bayesian Credible intervals [@mcelreath_statistical_2020] or Frequentist Confidence Intervals, although the latter do not provide a measure of uncertainty in the common sense of the word [@morey_fallacy_2016; @amrhein_scientists_2019].
+As alternatives to heat-map surface plots, one may include compatibility intervals using faceted or perspective plots of the surfaces along with the intervals [e.g., Fig.
+5 in @akkaya_hocagil_benchmark_2024] or overlapping contour plots on the surfaces [e.g., Fig.
+4.14 on page 187 of @wood_generalized_2017].
+For a simpler figure, one could also include lines with compatibility intervals for the responses at different temperatures.
 
-* RSFs require good data coverage over the resource-temperature space
-* future work investigating effects on a single species or evaluating space more accurately should:
-+ use fully Bayesian models with appropriate regularizing priors and penalties to avoid excessive predicitons
-+ use B splines to specify better penalty spaces
-+ have more data with good coverage over the covariate space
+How to represent uncertainty may seem trivial to some, but communicating uncertainty is crucial in assessing risk probabilistically [@aven_implementing_2002; @ayre_bayesian_2012; @kluppelberg_bayesian_2014].
+Recent events in global politics, including global conflicts [@mcnutt_scientists_2022; @tollefson_what_2022] and the rapid rise in generative machine learning [also referred to as artificial intelligence and GenAI, see @roumeliotis_chatgpt_2023; @van_noorden_chatgpt_2023; @deepseek-ai_deepseek-r1_2025] and its prohibitive energy and water demands [@li_making_2023] and carbon footprint [@luccioni_estimating_2024], have placed doubt on the likelihood and feasibility of reaching climate change goals that seemed achievable in the past decade [@united_nations_environment_programme_emissions_2024].
+Maintaining global warming below 1.5 or even 2.0$^\circ$C has been questioned [@united_nations_environment_programme_emissions_2024], and mitigating climate change will require extensive and widespread collaborative effort [e.g., @huang_achieving_2021].
+Similarly, protecting a third of the world's terrestrial and aquatic ecosystems [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022] appears less tangible as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023], and achieving such a goal will require active collaboration with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023].
+Understanding the consequences of climate change on mammals' movement behavior and spatial distribution [@williams_range_2018] is a first step towards anticipating and proactively responding to future changes in human-induced rapid environmental change [@sih_evolution_2011], including climate change.
+In the following sections, we discuss the implications of our results in more detail.
+We then expand on consequences for conservation during the 21^st^ century and considerations for future studies.
 
-* @pease_ecological_2024: the scale at which we measure variables affects their estumated effect
-* @hetem_responses_2014:
-* rate of climate change is too fast for genetic adaptation to occur in mammals with longevities of decades
-* example of suitable habitat prediction to 2050
-* proactive conservation: @sawyer_identifying_2009
+## Effects of temperature on movement rates
 
-<!-- - how results answer central question -->
+\noindent The lack of any common effects of temperature across all species' movement rates indicates that the effects of climate change will vary across species and that responding to current conditions and preparing for future changes will require a variety of adaptations as communities adapt and respond in complex and interconnected ways.
+Although our models do not account for any physiological or explicit phenological changes (e.g., in the timing of sleep and hibernation), the tensor product terms in Fig.
+\@ref(fig:dist) suggest that warmer temperatures cause many species to alter their daily and seasonal activity patterns.
+For example, at cold temperatures (-20$^\circ$C), elk moved most in the early morning (6:00 PDT) and evening (18:00 PDT), but they moved almost four times more at 6:00 PDT and half as much at 18:00 PDT when temperatures were above 20$^\circ$C.
+This suggests that while elk maintain a preference for crepuscular activity, they adapt their movement behavior to move more during warmer periods of the day (i.e., evening) on cold days and during cooler periods (i.e., early morning) on warmer days.
+The strong reduction in wolves', elk's, pumas', and boreal caribou's mid-day movement rates when temperatures are above 20$^\circ$C suggests that the increasingly common and intense heat waves across British Columbia will often have community-wide impacts on movement rates, encounter rates [@martinez-garcia_how_2020], and potentially community structure.
+More work is necessary on interspecific responses to temperature, including the effects of temperature on predation rates [but see: @cunningham_opportunity_2021; @glass_trade-off_2021; @brivio_seeking_2024].
+The increase in intensity and frequency of extreme heat events [@bunnell_global_2011; @yao_emergence_2022] will likely also impact the occurrence and timing of hibernation [@wells_life_2022] and migration or seasonal range expansions [@carbeck_adaptation_2022; @morley_spatiotemporal_2021; @malpeli_ungulate_2022], not only as a direct effect of temperature, but also because of cascading changes in the availability and unpredictability of resources [@mezzini_how_2024].
 
-## Effects of temperature on mammals' movement
+The substantial increase in uncertainty around the estimated movement rates at temperatures below $-20^\circ C$ or above $20^\circ C$ (Fig.
+***AX***) highlights the importance of including appropriate measures of uncertainty when evaluating the consequences of climate change on mammals' movement behavior.
+In this paper, we present our results cognizantly of the lack of uncertainty measures in many of our figures: We hoped to demonstrate how mammalian movement behavior depends on temperature while presenting the need for additional research.
+We expand on the importance of quantifying and communicating uncertainty in the last section of this discussion.
+Our relatively simple models indicate that changes in temperature will alter movement behavior even without accounting for large-scale physiological and behavioral adaptations, which may not be possible at the current rate of climate change [@williams_range_2018; @hetem_responses_2014].
+Still, our models to indicate that some animals adapt their behavior phenology in response to temperature, as indicated by wolves', elk's, caribou's, and grizzly bears' increased movement rates during warmer winters and spring days.
 
-## Effects of temperature on mammals' habitat selection
+## Effects of temperature on habitat selection
 
-<!-- - what it tells us about the broader field -->
+As with the effects of temperature on movement rates, the lack of common effects of temperature across all species indicates that efficient and proactive habitat conservation will require taxonomic nuance.
+In particular, our HRSFs do not account for any explicit physiological responses to temperature that may impact movement such as changes in the phenology of plants, ice, or mammalian physiology or behavior (e.g., seasonal coats, seasonal migration) other than as direct responses to temperature.
+Consequently, the behavioral changes we present should be interpreted carefully.
+While some seasonal cycles do not depend directly on temperature and may thus not be impacted directly by climate change [such as those driven by photoperiod, see @liu_effects_2022], the predictions we present also rely on the assumption that mammals will not adapt these seasonal cycles based on temperature-dependent cues, such as weather [@xu_frost_2019] or forage availability [@middleton_green-wave_2018].
+Additionally, it is important to stress the distinction between habitat selection and quality.
+While animals tend to select for high-quality habitats [e.g., @wright_survival_2024], high selection strength can also be a consequence of an ecological trap, an area animals are attracted to that provides a net loss of fitness rather than a gain [@zuniga-palacios_what_2021; @swearer_overview_2021].
+
+## Predicted changes in movement during the 21^st^ century
+
+Achieving the goals laid out by the "30 by 30" conservation initiative [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022] will require careful planning but often also rapid action.
+In the case of Canada, only 13.7% of its terrestrial areas and inland waters were protected at the end of 2023 [@environment_and_climate_change_canada_canadian_2024; @minister_of_environment_and_climate_change_act_2024].
+Additionally, efficient and effective conservation will require collaboration with Indigenous and local communities to ensure Treaties are respected in the process [e.g., @falardeau_biophysical_2022; @lamb_braiding_2023; @wong_towards_2020; @turner_its_2009] while also actively preparing for future change [@desjardins_looking_2020; @hessami_indigenizing_2021].
+The large range in predicted changes by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in figure \@ref(fig:s-hrsfs) highlight the need for careful planning that incorporates not only accurate estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates.
+The width of the averaged credible intervals in Figure \@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas.
+
+The widespread decrease in RSS in for areas with ranges with the top 5% RSS underscores the importance of moving beyond only using central estimates of change.
+While changes in the median RSS over time are small to negligible (yet never positive), decreases in the top 5% are more accentuated.
+This perspective is particularly important in the case of habitat selection modeling, since animals will be selecting most for (and thus spending more time in) habitats with high RSS values, so changes in the upper extremes are more important than changes in the median or lower extreme values.
+Similar arguments can be made for incorporating variation in temperature (including extreme events) when predicting future changes.
+Estimating changes in future weather with monthly means alone is insufficient because a change of 1-3 in mean monthly temperature may result in negligible predicted changes while ignoring changes in the frequency and intensity of extreme temperatures [@bunnell_global_2011; @zhang_increased_2023; @yao_emergence_2022].
+
+Accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage availability [@mezzini_how_2024], population dynamics [@needs_ref], competitive pressure [e.g., @torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019].
+This complexity results in the great variation among the responses of different species (and individuals).
+As selection strength for current ranges changes (and likely decreases), some animals may disperse, relocate to a new habitat, or remain within the current range suffer a decrease in fitness and an increase in extinction risk [@duncan_predicting_2012].
+
+Together, the predicted changes in distance traveled and habitat selection indicate that climate change will have complex consequences for populations as well as cascade effects within ecosystems as species adapt to the changing environments [or fail to do so: @botero_evolutionary_2015; @polazzo_measuring_2024].
+The resulting changes in mammalian movement behavior will have ramifications in individuals' encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@need_ref], and food webs [@need_ref] as landscapes change and animals face increasing pressures from human activity [@sih_evolution_2011], including habitat fragmentation [@needs_ref], habitat loss [@needs_ref], and greater human presence in wild spaces [@weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022].
 
@@
 
-Our work highlights ***X*** central considerations for future research: (1) telemetry sampling rates should bedesigned primarily in relation to the movement timescales of the species of interest and ideally with multiple samples **path?**, (2) the number of individuals tracked and the length of each telemetry should depend on environmental stability, and (3) ***projecting into the fringes****
+Our work highlights three central considerations for future research: (1) telemetry sampling rates should be designed primarily in relation to the movement timescales of the species of interest [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @pease_ecological_2024], (2) the number of individuals tracked and the length of each telemetry should depend not just on population size but also environmental stability, and (3) predicting changes in movement behavior becomes highly complex when responses are nonlinear, especially when changes depend on many interacting factors [@polazzo_measuring_2024] and one is interested in predicting responses in extreme conditions for which data are scarce [@steixner-kumar_strategies_2020].
+We address each consideration in the following three paragraphs.
 
@@
 Good estimates of an animal's speed and its movement path require telemetry locations to be taken more often than the animal's directional persistence [@noonan_scale-insensitive_2019], so that, on average, the dataset contains multiple locations in between changes in direction.
-What consitutes a change in direction depends on what movement scale one is investigating.
-Small-scale movements and first-order spatial selection will require more frequent sampling than migratory movement or secornd- and third-order spatial selection.
+What constitutes a change in direction depends on what movement scale one is investigating.
+Small-scale movements and first-order spatial selection will require more frequent sampling than migratory movement or second- and third-order spatial selection.
 While `ctmm` movement models are scale-invariant in that any model can be scaled to larger or smaller areas and timescales, the model estimates are not independent of the scale of the sampling frequency.
@@
 A good estimate of inter-individual variance provides better coefficient estimates along with more appropriate measures of uncertainty.
-However, when conditions across years are relatively stochastic, multi-year telemetries allow one to better estimate inter-annual and inter-individual variation.
-In either case, carefully-designed sampling schedules and (Bayesian) hierarchical models can provide good estimates of the effects of interested along with appropriate measures of uncertainty [@mcelreath_statistical_2016].
+However, when conditions across years are relatively stochastic, multi-year telemetries allow one to better estimate inter-annual variation without conflating it with inter-individual differences.
+In either case, carefully-designed sampling schedules and (Bayesian) hierarchical models can provide good estimates of the effects of interested along with appropriate measures of uncertainty [@mcelreath_statistical_2020].
 
-* improve projections at the edges of the HRSFs with local and Traditional Knowledge [@kutz_two-eyed_2019]
-* changes in temperatures are pushing to areas in the surface plots that have little data and may be poorly estimated.
-We will need to extrapolate to prepare for future changes, which implies some degree of inevitable risk.
-The best wat to avoid this is by designing models that are statistically sound and interpretable.
-The models we present provide an estimate of the changes in behavior, but they are not tailored specifically to any of the species in the dataset.
-Species-specific models should account for more particular variables (e.g., forest age for caribou) and use dynamic resource rasters (i.e., not use a static estimate of forest cover).
-Nonetheless, we present important results for understanding how mammals will respond to changes in climate and weather during the next ~75 years.
+Quantifying and communicating uncertainty become particularly important as we predict for conditions for which we have little to no data, such as during extreme events.
+Additionally, estimating nonlinear trends is particularly difficult near the edges of the data (e.g., extremely high temperatures), especially when data is scarce.
+Consequently, the estimated responses at the fringes of the surfaces in figures \@ref(fig:dist) and \@ref(fig:hrsfs) are more likely likely to be poorly estimated, particularly in the case of HRSFs, but increasing data density throughout the resource-temperature surfaces can ameliorate the issue.
+Preparing for future changes requires some degree of inevitable risk, so it is imperative that models be statistically sound and interpretable, and that results are communicated transparently and clearly to decision makers and the public alike.
 
-## Consequences for conservation
-
-* energetics
-* encounter rates (lower encounter rates with food [@hou_cold_2020])
-* HWI [@weststrate_how_2024]
-z* Machine Learning (Genenative Aritificial Intelligence; GenAI) AI is fueling climate change.
-We need to proactively prepare to respond to climate change.
-Pollution is not under the control of the masses, so predicting change will be hard
-* high energy and water demands [@li_making_2023] GenAI consumption of water
-* high carbon footprint [@luccioni_estimating_2024]
-* Pollution is not under control of the masses, so predicting change will be hard
-
-<!-- - address any questions presented in the discussion -->
+<!-- - address any questions presented in the introduction -->
 
@@
 
-<!-- additional references:
-
-- boars' sleep quantity, efficiency and quality are reduced on warmer days (https://doi.org/10.1098/rspb.2023.2115)
-
--->
-
 # References {-}

### 2025-03-01 analysis/figures/bc-displacement-predictions.R
removed `animal` terms; BC RSS proj was ok

@@ -23,10 +23,6 @@ if(file.exists('data/cc-hgam-bc-projections-albers.rds')) {
   m_1 <- readRDS('models/binomial-gam.rds')
   m_2 <- readRDS('models/gamma-gam.rds')
   
-  # terms to exclude from the prediction
-  SM <- unique(c(smooths(m_1), smooths(m_2)))
-  EXCLUDE <- SM[grepl('tod', SM) | grepl('animal', SM)]
-  
   # approximate Gaussian distributions of weather
   weather <- readRDS('data/weather-projections-2025-2100-only.rds') %>%
     filter(! is.na(temp_c)) %>% # there's some NA values even after masking
@@ -53,7 +49,7 @@ if(file.exists('data/cc-hgam-bc-projections-albers.rds')) {
               animal = m_1$model$animal[1],
               species = gsub('southern mountain', '(s. mountain)', x = species),
               lab,
-              tod_pdt = 0,
+              tod_pdt = 12,
               doy = yday(date_decimal(year + (month - 0.5) / 12)),
               temp_c,
               dt = 1,
@@ -62,9 +58,9 @@ if(file.exists('data/cc-hgam-bc-projections-albers.rds')) {
     mutate(
       .,
       p = predict(m_1, newdata = ., type = 'response',se.fit = FALSE,
-                  discrete = TRUE, exclude = EXCLUDE),
+                  discrete = TRUE, exclude = 's(animal)'),
       s = predict(m_2, newdata = ., type = 'response', se.fit = FALSE,
-                  discrete = TRUE, exclude = EXCLUDE),
+                  discrete = TRUE, exclude = 's(animal)'),
       d = p * s) %>%
     # take the weighted mean of the approximated gaussian distributions 
     # after grouping all the 2025 scenarios together
@@ -158,24 +154,26 @@ bc_preds %>%
   theme(legend.position = 'inside', legend.position.inside = c(0.85, 0.15))
 
 # figure of estimated speeds for each species ----
-z_breaks <- seq(-log2(1.25), log2(1.25), length.out = 5)
+z_breaks <- seq(-log2(1.5), log2(1.5), length.out = 5)
 
 # check color scheme
-p_0 <- filter(bc_preds, lab == lab[1], scenario == scenario[1]) %>%
-  # cap values for readability
-  mutate(d = case_when(d < 2^min(z_breaks) ~ 2^min(z_breaks),
[…]

### 2025-03-01 analysis/figures/temperature-hgams-projections.R

@@ -17,17 +17,17 @@ if(file.exists('data/cc-hgam-projections.rds')) {
   m_1 <- readRDS('models/binomial-gam.rds')
   m_2 <- readRDS('models/gamma-gam.rds')
   
-  # terms to exclude from the prediction
-  SM <- smooths(m_1)
-  TERMS <- c('(Intercept)', 'species',
-             SM[!(grepl('tod', SM) | grepl('dt', SM) | grepl('animal', SM))])
+  # when predicting: only excluding animal-level random effect and choosing
+  # specific values for the rest
   
   # import prediction data for each species in the data's extent ----
   if(file.exists('data/hgam-cc_newd.rds')) {
     cc_newd <- readRDS('data/hgam-cc_newd.rds')
   } else {
     cc_newd <- tibble(
-      wp = list.files('data', 'weather-projections-', full.names = TRUE) %>%
+      # list species files (not starting with numbers)
+      wp = list.files('data', 'weather-projections-[^0123456789]',
+                      full.names = TRUE) %>%
         map(readRDS)) %>%
       unnest(wp) %>%
       filter(year >= 2025) %>%
@@ -38,7 +38,7 @@ if(file.exists('data/cc-hgam-projections.rds')) {
         animal = m_1$model$animal[1], #' to use `discrete = TRUE`
         species = gsub('boreal', '(boreal)', species) %>%
           gsub('southern mountain', '(s. mountain)', x = .),
-        tod_pdt = 0,
+        tod_pdt = 12,
         doy = yday(date_decimal(year + (month - 0.5) / 12)),
         temp_c,
         dt = 1,
@@ -50,9 +50,9 @@ if(file.exists('data/cc-hgam-projections.rds')) {
   cc_proj <- cc_newd %>%
     mutate(.,
            p = predict(m_1, newdata = ., type = 'response', se.fit = FALSE,
-                       discrete = TRUE, terms = TERMS),
+                       discrete = TRUE, exclude = 's(animal)'),
            s = predict(m_2, newdata = ., type = 'response', se.fit = FALSE,
-                       discrete = TRUE, terms = TERMS),
+                       discrete = TRUE, exclude = 's(animal)'),
            d = p * s) %>%
     # average across day of year
     group_by(scenario, year, species, long, lat) %>%
[…]

### 2025-03-01 analysis/figures/temperature-hrsfs-projections.R

@@ -27,8 +27,9 @@ if(file.exists('data/cc-hrsf-projections.rds')) {
   cc_proj <- readRDS('data/cc-hrsf-projections.rds')
 } else {
   # terms to exclude from the predictions
-  EXCLUDE <- c('s(animal)', 's(forest_perc,animal)',
-               's(elevation_m,animal)', 's(dist_water_m,animal)')
+  SM <- gratia::smooths(readRDS('models/rsf-Oreamnos americanus-2025-01-20.rds'))
+  EXCLUDE <- SM[grepl('animal', SM)]
+  
   MODEL_FILES <- list.files('models', '^rsf-', full.names = TRUE)
   MODEL_FILES <- MODEL_FILES[! grepl('no-temp', MODEL_FILES)]
   
@@ -41,8 +42,6 @@ if(file.exists('data/cc-hrsf-projections.rds')) {
       gsub('\\(', '', .) %>%
       gsub('\\)', '', .) %>%
       gsub('\\.', '', .)
-    
-    rm(x) # remove previous object, if it exists
     
     x <<- # to keep in global environment in case saving fails
       tibble(
@@ -100,19 +99,20 @@ if(file.exists('data/cc-hrsf-projections.rds')) {
                                   grepl('245', scenario) ~ 'Good scenario (SSP 2-4.5)',
                                   grepl('370', scenario) ~ 'Bad scenario (SSP 3-7.0)',
                                   grepl('585', scenario) ~ 'Worst scenario (SSP 5-8.5)') %>%
-               factor(., levels = unique(.)))
+               factor(., levels = unique(.))) %>%
+      # fix species labs
+      mutate(lab = gsub('\\(s.~mountain\\)', '"\\(s. mountain\\)"', lab))
     saveRDS(x, paste0('data/cc-hrsf-projections-', sp, '.rds'))
     
     return(sp)
   })
   
-  cc_proj <- map_dfr(list.files('data', 'cc-hrsf-projections-', full.names = TRUE), readRDS)
+  cc_proj <-
+    list.files('data', 'cc-hrsf-projections-', full.names = TRUE) %>%
+    map(readRDS) %>%
+    bind_rows()
   saveRDS(cc_proj, 'data/cc-hrsf-projections.rds')
 }
-# fix species labs rather than having to re-run all predictions
-cc_proj <- cc_proj %>%
[…]

### 2025-03-01 analysis/06-temperature-hgams.R
made figure for moving 0/1 split

@@ -8,6 +8,7 @@ library('ggplot2')   # for fancy plots
 library('khroma')    # for colorblind-friendly color palettes
 library('cowplot')   # for fancy multi-panel plots
 source('analysis/figures/default-ggplot-theme.R') # bold text and no grids
+source('functions/get_legend.R') # to extract legends from ggplot plots
 plot_scheme(PAL, colours = TRUE)
 
 # using Pacific Time for time of day
@@ -66,28 +67,55 @@ if(file.exists('data/hgam-speed-data.rds')) {
 }
 
 #' `kmeans()` splits data poorly
-manual_splits <- c(0.06, 0.05, 0.08, 0.07, 0.1, 0.2, 0.05)
-km_splits <- rep(NA, N_SPECIES)
+splits <-
+  tibble(species = SPECIES,
+         lab = SPECIES_LABS,
+         manual = c(0.06, 0.05, 0.08, 0.07, 0.10, 0.20, 0.05),
+         `k means` = map_dbl(species, \(sp) {
+           km <- kmeans(filter(d, species == sp)$speed_est, centers = 2)
+           max(min(filter(d, species == sp) %>%
+                     filter(km$cluster == 1) %>%
+                     pull(speed_est)),
+               min(filter(d, species == sp) %>%
+                     filter(km$cluster == 2) %>%
+                     pull(speed_est)))
+         }),
+  ) %>%
+  pivot_longer(c(Manual, `k means`), values_to = 'split',
+               names_to = 'method')
 
-layout(matrix(c(1:7, 0), ncol = 4, byrow = TRUE))
-for(i in 1:N_SPECIES) {
-  km <- kmeans(filter(d, species == SPECIES[i])$speed_est, centers = 2)
-  km_splits[i] <- max(min(filter(d, species == SPECIES[i]) %>%
-                            filter(km$cluster == 1) %>%
-                            pull(speed_est)),
-                      min(filter(d, species == SPECIES[i]) %>%
-                            filter(km$cluster == 2) %>%
-                            pull(speed_est)))
-  hist(filter(d, species == SPECIES[i])$speed_est, breaks = 100,
-       xlim = c(0,
-                max(filter(d, species == SPECIES[i])$speed_est)) * 0.3,
-       ylim = c(0, nrow(filter(d, species == SPECIES[i])) / 40),
[…]

### 2025-03-01 analysis/figures/temperature-hrsfs-projections.R
remove unnecessary code

@@ -43,7 +43,6 @@ if(file.exists('data/cc-hrsf-projections.rds')) {
       gsub('\\)', '', .) %>%
       gsub('\\.', '', .)
     
-    x <<- # to keep in global environment in case saving fails
       tibble(
         species = fn %>%
           gsub(pattern = 'models/rsf-', replacement = '', x = .) %>%
@@ -101,8 +100,8 @@ if(file.exists('data/cc-hrsf-projections.rds')) {
                                   grepl('585', scenario) ~ 'Worst scenario (SSP 5-8.5)') %>%
                factor(., levels = unique(.))) %>%
       # fix species labs
-      mutate(lab = gsub('\\(s.~mountain\\)', '"\\(s. mountain\\)"', lab))
-    saveRDS(x, paste0('data/cc-hrsf-projections-', sp, '.rds'))
+      mutate(lab = gsub('\\(s.~mountain\\)', '"\\(s. mountain\\)"', lab)) %>%
+      saveRDS(paste0('data/cc-hrsf-projections-', sp, '.rds'))
     
     return(sp)
   })

### 2025-03-01 analysis/figures/temperature-hrsfs-projections.R
removed the 5% and 95% percentiles

@@ -117,13 +117,13 @@ if(file.exists('data/cc-hrsf-projections.rds')) {
 cc_proj %>%
   filter(year >= 2025) %>% # 2025 is the reference year
   ggplot() +
-  facet_wrap(~ lab, scales = 'free_y', labeller = label_parsed) +
-  geom_line(aes(year, l_upr_95 / l_ref, color = scenario), lwd = 0.5) +
-  geom_line(aes(year, l_lwr_05 / l_ref, color = scenario), lwd = 0.5) +
+  facet_wrap(~ lab, scales = 'fixed', labeller = label_parsed) +
+  # geom_line(aes(year, l_upr_95 / l_ref, color = scenario), lwd = 0.5) +
+  # geom_line(aes(year, l_lwr_05 / l_ref, color = scenario), lwd = 0.5) +
   geom_hline(yintercept = 1, color = 'black', lty = 'dashed') +
   geom_line(aes(year, l_median / l_ref, color = scenario), lwd = 1) +
   scale_x_continuous(NULL, breaks = c(2025, 2050, 2075, 2100)) +
-  scale_y_continuous('Relative change in RSS', limits = c(0, NA)) +
+  scale_y_continuous('Relative change in RSS') +
   scale_color_brewer('Climate change scenario', type = 'div', palette = 5, direction = -1,
                      aesthetics = c('color', 'fill')) +
   theme(legend.position = 'inside', legend.position.inside = c(5/6, 1/6))

### 2025-03-01 writing/manuscript.Rmd
committing to try to recover lost appendix

@@
     - \usepackage[left]{lineno} # for line numbers
-    - \usepackage{booktabs}
+    - \usepackage{booktabs} # to prevent tables from splitting
 subparagraph: true # needed for \usepackage[small]{titlesec}
@@
 
-\noindent \textbf{Appendices}: 3
+\noindent \textbf{Appendices}: 1
 
@@
 This manuscript leverages four main datasets: (1) a collection of GPS telemetry data; (2) historical hourly reanalyzed air temperature from the ERA5 dataset from the European Center for Medium-range Weather Forecasting; (3) resource rasters of percent forest cover, elevation, and distance from water; and (4) climate change projections under four Shared Socioeconomic Pathways [SSPs, see @riahi_shared_2017].
-We detail the data sources and analyses in the sections below, and more details are available in appendices A (estimating how temperature affects movement frequency and speed when moving), B (fitting Hierarchical Resource Selection Functions), and C (simulating hourly temperature data to 2100 using the four SSPs).
+We detail the data sources and analyses in the sections below.<!--, and more details are available in appendices A (estimating how temperature affects movement frequency and speed when moving), B (fitting Hierarchical Resource Selection Functions), and C (simulating hourly temperature data to 2100 using the four SSPs).-->
 
@@
 The linear predictor for the scale term (which governs the mean-variance relationship) included smooth terms of the monthly mean, month, and space.
-We did not include a smooth of year to avoid unrealistic projections when extrapolating beyond the range of historical data.
-Additional details are available in Appendix C.
+We did not include a smooth of year to avoid unrealistic projections when extrapolating beyond the range of historical data.<!-- Additional details are available in Appendix C.-->
 
@@
 We then predicted changes in movement behavior and habitat selection as a function of the simulated temperature using the HGAMs and HRSFs.
-For each month within each year from 2020 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantile by increments of 0.1, and we weighted each quantile proportionally to the (normalized) Gaussian probability density for each quantile (Appendix C).
+For each month within each year from 2020 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantile by increments of 0.1, and we weighted each quantile proportionally to the (normalized) Gaussian probability density for each quantile<!-- (Appendix C)-->.
 
@@
 The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R).
-To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using the inflection point of each species' histogram of detrended speed estimates (Appendix A).
+To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using the inflection point of each species' histogram of detrended speed estimates, as $k$-nearest neighbors did not discriminate between states well (Fig.
+S3).
 
@@
 In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022], so it is not possible to mitigate the effects of temperature on movement behavior (other than changing temperature directly).
-We provide Directed Acyclical Graphs in Appendix A (modeling the effects of temperature on movement rates), and Appendix B (modeling the effects of temperature on habitat selection).
+We provide the relevant Directed Acyclical Graphs in the Appendix (Figs.
+S1-S2).
 
@@
 
-\noindent Species differed in mean probabilities of movement (range: 0.05 -- 0.3), mean speed when moving (range: 0.42 -- 2.7 km/day), and mean distance traveled (range: 0.04 -- 0.6 km/day), even after predicting for one-hour time intervals for all species (Table \@ref(tab:gam-intercepts)).
+\noindent Data were relatively scarce for temperatures lower than $-20^\circ$C or above $20^\circ$C (Fig.
+S4).
+Species differed in overall mean probabilities of movement (range: 0.05 -- 0.3), mean speed when moving (range: 0.42 -- 2.7 km/day), and mean distance traveled (range: 0.04 -- 0.6 km/day), even after predicting for one-hour time intervals for all species (Table \@ref(tab:gam-intercepts)).
 Grizzly bears tended to move least often (probability of moving $P(M) \approx 0.05$), while wolves and pumas moved most often ($P(M) \gtrapprox 0.23$).
@@
 Including temperature in the models always increased within-sample prediction, but the change was not always substantial (deviance explained increased between 0.2% and 18.6%, with all HRSFs having an increase $\ge 4.6$), and there was relatively strong agreement between models with and without temperature (Fig.
-***SX pred vs pred***), but including temperature resulted in substantially better fits for all models, especially HRSFs (all $\Delta$AIC $\le$ -342; Table \@ref(tab:temp-comparison)).
+S5), but including temperature resulted in substantially better fits for all models, especially HRSFs (all $\Delta$AIC $\le$ -342; Table \@ref(tab:temp-comparison)).
 
@@
 
-\noindent The marginal effects of temperature varied in both direction and magnitude across species (Fig.
-***AX***), even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; see Fig.
-***AX***).
-The smooths' credible intervals were generally quite wide outside the typical temperature range of -20$^\circ$C to 20$^\circ$C, so the estimated trends were often not credible at the 95% level outside such range.
-Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species.
+\noindent The effects of temperature on movement rates varied in both direction and magnitude across species, even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; see Figs.
+S4-S6).
+<!--The smooths' credible intervals were generally quite wide outside the typical temperature range of -20$^\circ$C to 20$^\circ$C, so the estimated trends were often not credible at the 95% level outside such range.
+-->Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species.
 The models explained reasonably high proportions of the deviance (10.7% for the binomial model and 79.3% for the Gamma model) and had good in-sample prediction (Fig.
-***AX***).
+S9).
+All species adapted their daily and seasonal movement behavior to changes in temperature (Fig.
+\@ref(fig:dist)).
+Wolves, elk, pumas, and boreal caribou reduced their mid-day and summer movement rates when temperatures were above 20$^\circ$C.
+However, uncertainties around the estimated effects were higher at extreme temperatures due to data scarcity.
+
 All species' movement frequency and speed decreased with increasing sampling interval (Fig.
-***AX***).
-
-Most species adapted their daily and seasonal movement behavior to changes in temperature (Fig.
-\@ref(fig:dist)).
-Wolves, elk, pumas, and boreal caribou reduced their mid-day movement rates when temperatures were above 20$^\circ$C, and wolves, elk, and boreal caribou moved less on summer days above 20$^\circ$C.
-However, the precision of the estimated change depended on the GPS sampling frequency (Fig.
-***AX***).
+S10).
 While the `ctmm` models produced scale-independent estimates of speed [i.e., model interpretation is independent of sampling time: @noonan_scale-insensitive_2019], the accuracy, size, and stability of speed estimates still depended on the GPS sampling frequency.
@@
 Using the boreal caribou as an example, the 13-hour sampling interval allows us to reasonably estimate the caribou's movement path at a temporal scale of approximately 13 hours (or greater), but we cannot produce reasonable movement trajectories at a much finer (e.g., hourly) scale.
-Consequently, we suggest being cautious when comparing estimated movement behaviors across species.
+Consequently, we suggest being cautious when comparing estimated movement behaviors across species even though all predictions have been corrected to the hourly timescale by predicting for 1-hour time intervals (i.e., `dt_hours = 1`).
 
-```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals travel.} The fill indicates the estimated effects of temperature on the relative change in distance travelled (probability of moving times mean speed when moving) over time of day (\\textbf{A}) and day of year (\\textbf{B}).
+```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals travel.} The fill indicates the estimated effects of temperature on the relative change in distance travelled (probability of moving times mean speed when moving) over time of day on June $1^{st}$ (\\textbf{A}) and day of year at 12:0 PDT (\\textbf{B}).
 Predictions in the surface plots extend to 10% of the range away from each datum.
 The color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to -75\\% and +300\\% (i.e., $2^{\\pm 2}$) for ease of readability."}
+# not specifying dt, or doy for A and time for B because change is relative
 knitr::include_graphics('../figures/distance.png')
@@
 Estimated RSS values were generally most uncertain at extreme temperatures (Fig.
-***BX***).
+S11).
 
@@
 \@ref(fig:t-dist)).
-Under the best-case scenario, absolute projected changes by 2100 were small (approximately 0% to 2%), while absolute projected changes by 2100 ranged from ~1% (boreal caribou) to ~25% (southern mountain caribou) under the worst-case scenario.
-Boreal caribou and elk showed relatively little projected changes in yearly distance traveled.
+Under the best-case scenario, absolute projected changes by 2100 were small (approximately 0% to 2%), while under the worst-case scenario changes ranged from ~1% (boreal caribou) to ~25% (southern mountain caribou).
+Elk and boreal caribou showed relatively little projected changes in yearly distance traveled.
 Throughout BC (Fig.
@@
 ```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios often result in the greatest change and greatest uncertainty.} Lines indicate the projected relative change in mean distance travelled (probability of moving times speed when moving) due to changes in temperature within the species' current extent.
-Shaded areas indicate the average 95% Bayesian credible intervals for the area.
+Shaded areas indicate the 0.025 and 0.975 quantiles of the estimated means.
 Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs).
@@
 ```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each location's average distance in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase.
-For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.8 and 1.25 ($\\approx 2^{\\pm 0.32}$; original data ranged 0.81 to 1.35).
+For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.8 and 1.25 ($\\approx 2^{\\pm 0.58}$; original data ranged 0.72 to 1.29).
 The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest cover."}
@@
 
-All species were projected to reduce selection for their current range throughout the 21^st^ century, with the exception of grizzly bears, which showed little to no change under all four the climate change scenario (\@ref:fig(t-hrsfs)).
-The decreases in RSS were most evident in the habitats with the top 5% RSS in the current range.
-Projected changes in the median-RSS habitats were relatively small, but none increased throughout the century under any of the climate change scenarios.
+All species exhibited a decrease in RSS within their current range, irrespective of climate change scenario, but changes were often were only between -1% and -3% under the best-case scenario.
+Under the worst-case scenario, changes in RSS ranged from -8% to -14% (\@ref:fig(t-hrsfs)).
 As with distance traveled, the change in RSS in 2100 varied throughout BC for all species (Fig.
@@
 All species were predicted to exhibit a decrease in RSS throughout most of BC, with the exception of boreal caribou, which had a predicted decrease in approximately 37% of the province, including their current range.
-While the boreal caribou were expected to increase selection for altitudes > 1 km above sea level, the estimated selection strength was highly uncertain and not credible at the 95% level (Fig.
-***SX***).
+However, this was in great part due to the estimated increase in selection for high-elevation areas that are well outside the population's current range.
 
 ```{r, eval=FALSE}
+# find species that decrease in RSS in less than half of BC
 library('dplyr')
@@
 
-```{r t-hrsfs, fig.cap="\\textbf{The intensity and direction of changes in habitat selection due to climate change vary among species, but no species are expected to exhibit an increase in relative selection strength (RSS).} Lines indicate the projected change in RSS based on the 5$^{th}$ percentile (the worst 5%; bottom line), the median (the middle; bold), and 95% percentile (the best 5%; top line) habitats within the species' current extent.
+```{r t-hrsfs, fig.cap="\\textbf{All species are expected to exhibit a decrease in relative selection strength (RSS) for their current range, irrespective of climate change scenario, but decreases are stronger in the worse-case scenario.} Lines indicate the projected change in median RSS within the species' current extent.
 Changes are relative to the median RSS in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
@@
 ```{r s-hrsfs, fig.cap="\\textbf{Climate change will impact each species' habitat relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase.
-For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.6}$; original data ranged 0.26 to 117)."}
+For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.6}$; original data ranged 0.26 to 117).
+The increase in boreal caribou's RSS for high elevations at warmer temperatures was highly uncertain and often not credible at the 95\\% level."}
 knitr::include_graphics('../figures/bc-rss-2100.png')
@@
 Predicting at the fringes of the surface plots we present comes with substantial uncertainty, particularly given that the responses are nonlinear.
-At unusually warm temperatures, mammals (and other homeotherms) overheat, so their movement behavior and that of their competitors, predators, and/or prey can often be substantially different from the behavior at typical temperatures.
-As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], mammals will be increasingly forced to atypical behaviors, which will have the potential to substantially alter community structures and behaviors, both during such events and afterwards [@zhang_increased_2023; ].
-For instance, changes in climate and phenology impact the life history and behavior of many hibernating mammals [@wells_life_2022], but hot weather also affects mammals' sleep quality [e.g., wild boars, *Sus scrofa*: @mortlock_sleep_2024] and likelihood to enter torpor [Australian eastern long-eared bat, *Nyctophilus bifax*: @fjelldal_nightly_2021].
-These changes can alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024].
+At unusually warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behavior and that of their competitors, predators, and/or prey can often be substantially different from the behavior at typical temperatures.
+As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], mammals will be increasingly forced to atypical behaviors, which will have the potential to substantially alter community structures and behaviors, both during such events and afterwards [@zhang_increased_2023].
+For instance, changes in climate and phenology impact the life history and behavior of many hibernating mammals [@wells_life_2022], and hot weather can also affects mammals' sleep quality [e.g., wild boars, *Sus scrofa*: @mortlock_sleep_2024] and likelihood to enter torpor [e.g., Australian eastern long-eared bat, *Nyctophilus bifax*: @fjelldal_nightly_2021].
+Our results show that warmer temperatures cause grizzly bears to more further earlier in the year (Fig.
+\@ref(fig:dist)), which may alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024].
 At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase their ability to find food or be predated upon [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021] as well as the overall ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023].
-However, these changes are likely to have complex consequences for population and ecosystem structures and dynamics as prey, predators, and prey experience altered and novel seasonal cycles.
-especially given the increasingly common climate "weirding" [@bunnell_global_2011], which includes not only warmer temperatures and more frequent and intense heat waves but also an increase in various other extreme events [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], such as cold spells, floods and droughts, plant and animal die-offs and sudden growth surges, and other changes in population dynamics and behaviors [sometimes referred to as "black swan" events -- see: @logares_black_2012; @anderson_black-swan_2017].
+These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011], including warmer temperatures and more frequent and intense extreme events [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], such as heat waves, cold spells, floods and droughts, plant and animal die-offs, and sudden growth surges [sometimes referred to as "black swan" events -- see: @logares_black_2012; @anderson_black-swan_2017].
+
+***continue reading from HERE***
 
@@
 
-The substantial increase in uncertainty around the estimated movement rates at temperatures below $-20^\circ C$ or above $20^\circ C$ (Fig.
-***AX***) highlights the importance of including appropriate measures of uncertainty when evaluating the consequences of climate change on mammals' movement behavior.
+The greater uncertainty around the estimated movement rates at temperatures outside the typical temperatures for a given day (Figs.
+S2, S4-S6) highlights the importance of including appropriate measures of uncertainty when evaluating the consequences of climate change on mammals' movement behavior.
 In this paper, we present our results cognizantly of the lack of uncertainty measures in many of our figures: We hoped to demonstrate how mammalian movement behavior depends on temperature while presenting the need for additional research.

### 2025-03-01 writing/appendix.Rmd
created appendix

---
title: "Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada"
subtitle: "Appendix"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - bcp: BC Parks
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'freshwater-biology.csl'
# csl: 'nature.csl'
fontsize: 12pt
indent: true
header-includes:
    - \renewcommand{\figurename}{Figure S\!\!} # for "Figure Ax.
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
---

```{r setup, include=FALSE}
# set chunk defaults: don't print code, images text width and centered
knitr::opts_chunk$set(echo = FALSE, out.width = '\\textwidth',
                      fig.align = 'center', cache = TRUE, warning = FALSE,
                      message = FALSE)
```

# Directed Acyclical Graphs

```{r hgam-dag,  message=FALSE, warning=FALSE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of temperature (Temp) on probability of moving, speed when moving, or distance traveled ($Y$) while accounting for the effects of time of day (TOD), day of year (DOY), and their interaction effects. Temperature directly affects $Y$, but the effects of temperature depend on the time of day and season. Time of day and day of year also affect $Y$ directly, but the effect of time of day changes throughout the year due to changes in day length and seasonality.", fig.height=4, fig.width=4, out.width='40%'}
library('dplyr')   # for data wrangling
library('ggplot2') # for fancy plots
library('ggdag')   # for directed acyclical graphs

dagify(Y ~ TOD + DOY + Temp,
       Temp ~ TOD + DOY,
       TOD ~ DOY,
       outcome = 'Y',
       coords = list(x = c(Y = 0, TOD = -1, DOY = 1, Temp = 0),
                     y = c(Y = 1.2, TOD = 2, DOY = 2, Temp = 0))) %>%
  tidy_dagitty() %>%
  mutate(col = case_when(name == 'Y' ~ '1',
                         name == 'TOD' ~ '0',
                         name == 'DOY' ~ '0',
                         name == 'Temp' ~ '2')) %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point(aes(color = col), size = 20, alpha = 0.5) +
  geom_dag_point(color = 'black', size = 20, shape = 1) +
  geom_dag_edges() +
  geom_dag_text(aes(label = name), color = 'black', size = 4.5) +
  scale_color_manual(values = c('grey', 'darkorange', 'red')) +
  theme_dag() +
  theme(legend.position = 'none')
```

```{r hrsf-dag, message=FALSE, warning=FALSE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of temperature (Temp) on Relative Selection Strength (RSS) for percent forest cover (F), elevation (E), and distance from water (W). The RSS for of a given habitat depends on all three resources, and the selection for each resource is independent of the other two resources and dependent on temperature.", fig.height=4, fig.width=4, out.width='40%'}
library('dplyr')   # for data wrangling
library('ggplot2') # for fancy plots
library('ggdag')   # for directed acyclical graphs

dagify(RSS ~ `F` + E + W,
       `F` ~ Temp,
       E ~ Temp,
       W ~ Temp,
       outcome = 'RSS',
       coords = list(x = c(RSS = 0, `F` = -1, E = 0, W = 1, Temp = 0),
                     y = c(RSS = 2, `F` = 1, E = 1, W = 1, Temp = 0))) %>%
  tidy_dagitty() %>%
  mutate(col = case_when(name == 'RSS' ~ '1',
                         name == 'F' ~ '0',
                         name == 'E' ~ '0',
                         name == 'W' ~ '0',
                         name == 'Temp' ~ '2')) %>%
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_point(aes(color = col), size = 20, alpha = 0.5) +
  geom_dag_point(color = 'black', size = 20, shape = 1) +
  geom_dag_edges() +
  geom_dag_text(aes(label = name), color = 'black', size = 4.5) +
  scale_color_manual(values = c('grey', 'darkorange', 'red')) +
  theme_dag() +
  theme(legend.position = 'none')
```

\clearpage

# Effects of temperature on movement rates

```{r splits, fig.cap="Histograms of each species' estimated speed. Continuous red lines indicate the values used to determine whether an animal was moving or not, which were determined visually using the inflection points of the histograms. Dashed blue lines indicate the minimum speed corresponding to a moving animal as determined by $k$-means algorithms with 2 clusters. For ease of readability, the x axes range from 0 the 0.99 quantile, while the y axes range from 0 to one fortieth of the total number of estimates."}
knitr::include_graphics('../figures/moving-0-1-split.png')
```

```{r data-hist, fig.cap="Histograms of the number of states (\\textbf{A}) and speed estimates (\\textbf{B}) over time of day (Pacific Daylight Time, PDT), day of year (Julian date), and air temperature."}
knitr::include_graphics('../figures/temperature-movement-rates-hist.png')
```

```{r pred-vs-pred, fig.cap="Scatterplot of the fitted values from the HGAMs with and without including temperature for the probability of movement (\\textbf{A}) and speed when moving (\\textbf{B})."}
knitr::include_graphics('../figures/hgam-with-without-temp-prediction-agreement.png')
```

```{r p-full, fig.cap="\\textbf{A.} Estimated effects of temperature on each species' probability of moving on June 1$^{\\text{st}}$ at 12:00, Pacific Daylight Time (PDT). The rug plot indicates each species' data on June 1$^{\\text{st}}$. \\textbf{B.} Estimated effects of time of day on each species' probability of moving on June 1$^{\\text{st}}$ at $0^\\circ$C. The grey area indicates evening and night (hours between 18:00 and 6:00). \\textbf{C.} Estimated effects of day of year on each species' probability of moving at 12:00 with a temperature of 0$^\\circ$C. The year is divided into the four seasons: winter (white), spring (green), summer (gold), and fall (brown). In panels A-C, ribbons indicate 95\\% Bayesian Credible Intervals, and the sampling rate was post-stratified to $\\Delta t = 1$ hour for all species. \\textbf{D.} Effects of time of day and temperature on species' probability of moving on June 1$^{\\text{st}}$ at 12:00 PDT. \\textbf{E.} Effects of day of year and temperature on species' probability of moving at 12:00 PDT. Surfaces in panels D and E extend to 10\\% of the range away from each datum."}
knitr::include_graphics('../figures/p-moving-full.png')
```

<!-- speed and displacement don't need specifications on time of day because the change is relative, but keeping them anyway for consistency -->

```{r s-full, fig.cap="\\textbf{A.} Estimated effects of temperature on each species' speed when moving on June 1$^{\\text{st}}$ at 12:00, Pacific Daylight Time (PDT). The rug plot indicates each species' data points where an animal was moving on June 1$^{\\text{st}}$. \\textbf{B.} Estimated effects of time of day on each species' speed when moving on June 1$^{\\text{st}}$ at $0^\\circ$C. The grey area indicates evening and night (hours between 18:00 and 6:00). \\textbf{C.} Estimated effects of day of year on each species' speed when moving at 12:00 with a temperature of 0$^\\circ$C. The year is divided into the four seasons: winter (white), spring (green), summer (gold), and fall (brown). In panels A-C, ribbons indicate 95\\% Bayesian Credible Intervals, and the sampling rate was post-stratified to $\\Delta t = 1$ hour for all species. \\textbf{D.} Effects of time of day and temperature on species' speed when moving on June 1$^{\\text{st}}$. \\textbf{E.} Effects of day of year and temperature on species' speed when moving, if the animal was moving at 12:00 PM PDT. Surfaces extend to 10\\% of the range away from each datum. The color bar is on the $\\log_2$ scale to help visualize patterns in doubling."}
knitr::include_graphics('../figures/speed-full.png')
```

```{r d-full, fig.cap="\\textbf{A.} Estimated effects of temperature on each species' distance traveled on June 1$^{\\text{st}}$ at 12:00, Pacific Daylight Time (PDT). The rug plot indicates each species' data points where an animal was moving on June 1$^{\\text{st}}$. \\textbf{B.} Estimated effects of time of day on each species' distance traveled on June 1$^{\\text{st}}$ at $0^\\circ$C. The grey area indicates evening and night (hours between 18:00 and 6:00). \\textbf{C.} Estimated effects of day of year on each species' distance traveled at 12:00 with a temperature of 0$^\\circ$C. The year is divided into the four seasons: winter (white), spring (green), summer (gold), and fall (brown). In panels A-C, ribbons indicate 95\\% Bayesian Credible Intervals, and the sampling rate was post-stratified to $\\Delta t = 1$ hour for all species. \\textbf{D.} Effects of time of day and temperature on species' distance traveled on June 1$^{\\text{st}}$ at 12:00 PDT. \\textbf{E.} Effects of day of year and temperature on species' distance traveled at 12:00 PDT. Surfaces extend to 10\\% of the range away from each datum. The color bar is on the $\\log_2$ scale to help visualize patterns in doubling."}
knitr::include_graphics('../figures/distance-travelled-full.png')
```

```{r op, fig.cap="Relationsips between the observed and predicted values for probability of movement (\\textbf{A}) and speed given that an animal was moving (\\textbf{B}). The color in panel \\textbf{A} indicates the number of points used to calculate the estimate (on the $\\log-{2}$ scale), while the grey lines in both panels indicate perfect prediction (i.e., the 1:1 line)."}
knitr::include_graphics('../figures/p-and-speed-observed-vs-predicted.png')
```

```{r dt-smooths, fig.cap="\\textbf{A.} Histograms of sampling intervals between GPS locations, with counts on a $\\log_{10}$ axis for ease of readability. \\textbf{B.} Species-level smooth effects of sampling interval on the estimated probability of the animal moving. \\textbf{C.} Species-level estimated smooth effects of sampling interval on an animal's speed when moving. All x axes are on the natural logarithm scale."}
knitr::include_graphics('../figures/dt-smooths.png')
```

\clearpage

# Effects of temperature on habitat selection

```{r hrsf-se, fig.cap="Estimated standard error in the relative selection strength (RSS) for forest cover (\\%), elevation (km), and distance from water (km) as a function of temperature (see Fig. 3 in the main text). The grey contours indicate the extent of each speices' observed locations."}
knitr::include_graphics('../figures/hrsf-surface-plots-se.png')
```

### 2025-03-01 writing/appendix.Rmd
increased nbins for a/u; added to appendix

@@
 
+```{r avail-used, out.height='90%', fig.cap="Histograms of available (red) and used (blue) resources.
+All GPS fixes counts as one observation.
+The y axis is on the square-root scale to help visualize values with low counts."}
+knitr::include_graphics('../figures/available-vs-used.png')
+```
+
+```{r avail-used-w, out.height='90%', fig.cap="Histograms of available (red) and used (blue) resources.
+All GPS fixes are weighted according to their independence relative to the respective animals' range crossing time and Autocorrelated Kernel Density Estimate (i.e., the weights returned by AKDE weights multiplied by the number of home-range-crossing degrees of freedom).
+The y axis is on the square-root scale to help visualize values with low counts."}
+knitr::include_graphics('../figures/available-vs-used-weighted.png')
+```
+
 ```{r hrsf-se, fig.cap="Estimated standard error in the relative selection strength (RSS) for forest cover (\\%), elevation (km), and distance from water (km) as a function of temperature (see Fig.

### 2025-03-01 analysis/figures/hrsfs-available-vs-used.R

@@ -10,33 +10,31 @@ ANIMALS <- readr::read_csv('data/tracking-data/telemetry-metadata.csv',
   pull(animal)
 
 d <- readRDS('data/tracking-data/rsf-data.rds') %>%
-  filter(animal %in% ANIMALS)
-
-a_u <- d %>%
+  filter(animal %in% ANIMALS) %>%
   mutate(elevation_m = elevation_m / 1e3,
          dist_water_m = dist_water_m / 1e3) %>%
   pivot_longer(c(forest_perc, elevation_m, dist_water_m),
                names_to = 'variable') %>%
   mutate(
-    lab = case_when(species == SPECIES[1] ~ SPECIES_LABS[1],
-                        species == SPECIES[2] ~ SPECIES_LABS[2],
-                        species == SPECIES[3] ~ SPECIES_LABS[3],
-                        species == SPECIES[4] ~ SPECIES_LABS[4],
-                        species == SPECIES[5] ~ SPECIES_LABS[5],
-                        species == SPECIES[6] ~ SPECIES_LABS[6],
-                        species == SPECIES[7] ~ SPECIES_LABS[7]),
+    lab = case_when(species == 'Rangifer tarandus southern mountain' ~ SPECIES_LABS[1],
+                    species == 'Ursus arctos horribilis' ~ SPECIES_LABS[2],
+                    species == 'Puma concolor' ~ SPECIES_LABS[3],
+                    species == 'Cervus canadensis' ~ SPECIES_LABS[4],
+                    species == 'Rangifer tarandus boreal' ~ SPECIES_LABS[5],
+                    species == 'Canis lupus' ~ SPECIES_LABS[6],
+                    species == 'Oreamnos americanus' ~ SPECIES_LABS[7]),
     variable =
       case_when(variable == 'forest_perc' ~ "bold(Forest~cover~'(%)')",
                 variable == 'elevation_m' ~ "bold(Elevation~'(km)')",
                 variable == 'dist_water_m' ~ "bold(Distance~from~water~'(km)')") %>%
-      factor(levels = c("bold(Forest~cover~'(%)')",
-                        "bold(Elevation~'(km)')",
-                        "bold(Distance~from~water~'(km)')"))) %>%
-  ggplot() +
+      factor(., levels = unique(.)))
+
+a_u <-
+  ggplot(d) +
   facet_grid(lab ~ variable, scales = 'free', labeller = label_parsed,
              switch = 'x') +
[…]

### 2025-03-03 analysis/figures/temperature-hrsfs.R
adjusted the intercept for boreal caribou elev

@@ -160,7 +160,10 @@ p <-
   # doing this because some estimated effects of distance from water and
   # elevation are extreme
   group_by(species, variable) %>%
-  mutate(lambda = lambda / median(lambda)) %>%
+  mutate(lambda = lambda / median(lambda),
+         lambda = if_else(species == 'Rangifer tarandus (boreal)' &
+                            variable == 'bold(Elevation~"(km)")',
+                          lambda / exp(4), lambda)) %>%
   ungroup() %>%
   # cap at 2^(+/-LIM)
   mutate(log2_lambda = log2(lambda),

### 2025-03-03 analysis/figures/temperature-hrsfs-projections.R
added more 0.25 and 0.75 quantiles

@@ -88,9 +88,11 @@ if(file.exists('data/cc-hrsf-projections.rds')) {
       # find median and 90% percentile interval of the predicted means
       # not including CIs because averaging them is not straightforward
       group_by(lab, scenario, year) %>%
-      summarize(l_lwr_05 = quantile(l, 0.05),
+      summarize(l_05 = quantile(l, 0.05),
+                l_25 = quantile(l, 0.25),
                 l_median = quantile(l, 0.50),
-                l_upr_95 = quantile(l, 0.95),
+                l_75 = quantile(l, 0.75),
+                l_95 = quantile(l, 0.95),
                 .groups = 'drop') %>%
       # divide by mean of 2025 to find relative change since 2025
       mutate(l_ref = mean(l_median[year == 2025])) %>%
@@ -117,16 +119,36 @@ if(file.exists('data/cc-hrsf-projections.rds')) {
 cc_proj %>%
   filter(year >= 2025) %>% # 2025 is the reference year
   ggplot() +
-  facet_wrap(~ lab, scales = 'fixed', labeller = label_parsed) +
-  # geom_line(aes(year, l_upr_95 / l_ref, color = scenario), lwd = 0.5) +
-  # geom_line(aes(year, l_lwr_05 / l_ref, color = scenario), lwd = 0.5) +
+  facet_wrap(~ lab, scales = 'free', labeller = label_parsed) +
   geom_hline(yintercept = 1, color = 'black', lty = 'dashed') +
   geom_line(aes(year, l_median / l_ref, color = scenario), lwd = 1) +
   scale_x_continuous(NULL, breaks = c(2025, 2050, 2075, 2100)) +
   scale_y_continuous('Relative change in RSS') +
-  scale_color_brewer('Climate change scenario', type = 'div', palette = 5, direction = -1,
-                     aesthetics = c('color', 'fill')) +
+  scale_color_brewer('Climate change scenario', type = 'div', palette = 5,
+                     direction = -1, aesthetics = c('color', 'fill')) +
   theme(legend.position = 'inside', legend.position.inside = c(5/6, 1/6))
 
 ggsave('figures/rss-local-cc-predictions.png',
        width = 10, height = 6.67, dpi = 600, bg = 'white')
+
+cc_proj %>%
+  filter(year >= 2025) %>% # 2025 is the reference year
+  select(! c(l_25, l_75)) %>%
+  group_by(lab) %>%
[…]

### 2025-03-03 analysis/figures/temperature-hrsfs-projections.R
created figure faceted by quantiles

@@ -119,7 +119,7 @@ if(file.exists('data/cc-hrsf-projections.rds')) {
 cc_proj %>%
   filter(year >= 2025) %>% # 2025 is the reference year
   ggplot() +
-  facet_wrap(~ lab, scales = 'free', labeller = label_parsed) +
+  facet_wrap(~ lab, scales = 'fixed', labeller = label_parsed) +
   geom_hline(yintercept = 1, color = 'black', lty = 'dashed') +
   geom_line(aes(year, l_median / l_ref, color = scenario), lwd = 1) +
   scale_x_continuous(NULL, breaks = c(2025, 2050, 2075, 2100)) +
@@ -139,10 +139,9 @@ cc_proj %>%
   ungroup() %>%
   pivot_longer(l_05:l_95, values_to = 'l', names_to = 'percentile',
                names_prefix = 'l_') %>%
-  mutate(percentile = if_else(percentile == 'median', '50', percentile) %>%
-           as.numeric() %>%
-           paste0('"', ., '%"') %>%
-           factor(levels = c('"5%"', '"50%"', '"95%"'))) %>%
+  mutate(percentile = case_when(percentile == '05' ~ '"Bottom 5% RSS"',
+                                percentile == 'median' ~ '"Median RSS"',
+                                percentile == '95' ~ '"Top 5% RSS"')) %>%
   ggplot() +
   facet_grid(percentile ~ lab, scales = 'free', labeller = label_parsed) +
   geom_line(aes(year, l, color = scenario), lwd = 0.5) +
@@ -152,3 +151,6 @@ cc_proj %>%
   scale_color_brewer('Climate change scenario', type = 'div', palette = 5,
                      direction = -1, aesthetics = c('color', 'fill')) +
   theme(legend.position = 'top')
+
+ggsave('figures/rss-local-cc-predictions-quantiles.png',
+       width = 20, height = 8, dpi = 600, bg = 'white')

### 2025-03-03 analysis/figures/temperature-hgams-projections.R
using estimated change instead of relative for p

@@ -128,12 +128,12 @@ if(file.exists('data/cc-hgam-projections.rds')) {
 
 # make figures ----
 p_p_mov <-
-  ggplot(cc_proj, aes(year, p_median / p_ref, group = scenario)) +
+  ggplot(cc_proj, aes(year, p_median, group = scenario)) +
   facet_wrap(~ species, scales = 'free_y', drop = FALSE,
              labeller = label_parsed) +
-  geom_ribbon(aes(ymin = p_lwr_05 / p_ref, ymax = p_upr_95 / p_ref,
+  geom_ribbon(aes(ymin = p_lwr_05, ymax = p_upr_95,
                   fill = scenario), alpha = 0.2) +
-  geom_hline(aes(yintercept = 1), lty = 'dashed') +
+  geom_hline(aes(yintercept = p_ref), lty = 'dashed') +
   geom_line(color = 'black', lwd = 1.5) +
   geom_line(aes(color = scenario), lwd = 1) +
   scale_color_brewer('Climate change scenario', type = 'div',

### 2025-03-03 writing/appendix.Rmd
updated results and discussion following last fix

@@
 ---
-title: "Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada"
+title: \Large Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada
 subtitle: "Appendix"
@@
 ```
+
+\clearpage
 
@@
 
-```{r pred-vs-pred, fig.cap="Scatterplot of the fitted values from the HGAMs with and without including temperature for the probability of movement (\\textbf{A}) and speed when moving (\\textbf{B})."}
+```{r pred-vs-pred-hgams, fig.cap="Scatterplot of the fitted values from the HGAMs with and without including temperature for the probability of movement (\\textbf{A}) and speed when moving (\\textbf{B})."}
 knitr::include_graphics('../figures/hgam-with-without-temp-prediction-agreement.png')
@@
 
+```{r t-p-mov, fig.cap="\\textbf{The direction and magnitude of changes in probability of moving due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in probability of moving due to changes in temperature within the species' current extent.
+Shaded areas indicate the 90% prediction interval within the range.
+The dashed black lines indicate the mean probability of movement in 2025 across the four scenarios.
+The projections only account for changes movement behavior (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
+knitr::include_graphics('../figures/p-moving-local-cc-predictions.png')
+```
+
+```{r t-speed, fig.cap="\\textbf{The direction and magnitude of changes in speed due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in speed due to changes in temperature within the species' current extent.
+Shaded areas indicate the 90% prediction interval within the range.
+The dashed black lines indicate the mean speed in 2025 across the four scenarios.
+The projections only account for changes movement behavior (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
+knitr::include_graphics('../figures/speed-local-cc-predictions.png')
+```
+
 \clearpage
@@
 
-```{r avail-used, out.height='90%', fig.cap="Histograms of available (red) and used (blue) resources.
+```{r avail-used, out.width='95%', fig.cap="Histograms of available (red) and used (blue) resources.
 All GPS fixes counts as one observation.
@@
 
-```{r avail-used-w, out.height='90%', fig.cap="Histograms of available (red) and used (blue) resources.
+```{r avail-used-w, out.width='95%', fig.cap="Histograms of available (red) and used (blue) resources.
 All GPS fixes are weighted according to their independence relative to the respective animals' range crossing time and Autocorrelated Kernel Density Estimate (i.e., the weights returned by AKDE weights multiplied by the number of home-range-crossing degrees of freedom).
@@
 knitr::include_graphics('../figures/available-vs-used-weighted.png')
+```
+
+```{r pred-vs-pred-hrsfs, fig.cap="Scatterplot of the fitted values from the HRSFs with and without including temperature."}
+knitr::include_graphics('../figures/hrsf-with-without-temp-prediction-agreement.png')
 ```

### 2025-03-03 writing/manuscript.Rmd
updated results and discussion following last fix

@@
 
-Including temperature in the models always increased within-sample prediction, but the change was not always substantial (deviance explained increased between 0.2% and 18.6%, with all HRSFs having an increase $\ge 4.6$), and there was relatively strong agreement between models with and without temperature (Fig.
-S5), but including temperature resulted in substantially better fits for all models, especially HRSFs (all $\Delta$AIC $\le$ -342; Table \@ref(tab:temp-comparison)).
+Including temperature in the models always increased within-sample prediction, but the change was not always substantial (deviance explained increased between 0.2% and 18.6%, with all HRSFs having an increase $\ge 4.6$), and there was relatively strong agreement between models with and without temperature (Figs.
+S5 and S13), but including temperature resulted in substantially better fits for all models, especially HRSFs (all $\Delta$AIC $\le$ -342; Table \@ref(tab:temp-comparison)).
 
@@
 \noindent The effects of temperature on movement rates varied in both direction and magnitude across species, even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; see Figs.
-S4-S6).
-<!--The smooths' credible intervals were generally quite wide outside the typical temperature range of -20$^\circ$C to 20$^\circ$C, so the estimated trends were often not credible at the 95% level outside such range.
--->Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species.
+S6-S8).
+Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species.
 The models explained reasonably high proportions of the deviance (10.7% for the binomial model and 79.3% for the Gamma model) and had good in-sample prediction (Fig.
@@
 Wolves, elk, pumas, and boreal caribou reduced their mid-day and summer movement rates when temperatures were above 20$^\circ$C.
-However, uncertainties around the estimated effects were higher at extreme temperatures due to data scarcity.
+However, uncertainties around the estimated effects were often higher at extreme temperatures due to data scarcity.
 
-All species' movement frequency and speed decreased with increasing sampling interval (Fig.
+All species' estimated probability of moving decreased with sampling intervals approximately above 1 hour, and most species' estimated speed when moving decreased monotonically with sampling interval, although the estimated trends were highly uncertain for some species (Fig.
 S10).
-While the `ctmm` models produced scale-independent estimates of speed [i.e., model interpretation is independent of sampling time: @noonan_scale-insensitive_2019], the accuracy, size, and stability of speed estimates still depended on the GPS sampling frequency.
-This is because coarsely-sampled movement data contains information on large-scale movements (e.g., range crossings, migrations) but not fine-scale movements [e.g., first-order habitat selection *sensu* @johnson_comparison_1980].
+While the `ctmm` models produced scale-independent estimates of speed [i.e., model interpretation is independent of sampling interval: @noonan_scale-insensitive_2019], the accuracy, size, and stability of speed estimates still depended on the GPS sampling frequency.
+This is because coarsely-sampled movement data contains information on large-scale movements (e.g., range crossings, migrations) but not fine-scale movements [e.g., first-order habitat selection *sensu* @johnson_comparison_1980; also see @nathan_big-data_2022].
 Using the boreal caribou as an example, the 13-hour sampling interval allows us to reasonably estimate the caribou's movement path at a temporal scale of approximately 13 hours (or greater), but we cannot produce reasonable movement trajectories at a much finer (e.g., hourly) scale.
-Consequently, we suggest being cautious when comparing estimated movement behaviors across species even though all predictions have been corrected to the hourly timescale by predicting for 1-hour time intervals (i.e., `dt_hours = 1`).
+Consequently, we suggest being cautious when comparing estimated movement behaviors across species, even though all predictions have been corrected to the hourly timescale by predicting for 1-hour time intervals (i.e., `dt_hours = 1`).
 
-```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals travel.} The fill indicates the estimated effects of temperature on the relative change in distance travelled (probability of moving times mean speed when moving) over time of day on June $1^{st}$ (\\textbf{A}) and day of year at 12:0 PDT (\\textbf{B}).
+```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals travel.} The fill indicates the estimated effects of temperature on the relative change in distance travelled (probability of moving times mean speed when moving) over time of day on June $1^{\\text{st}}$ (\\textbf{A}) and day of year at 12:00 PDT (\\textbf{B}).
 Predictions in the surface plots extend to 10% of the range away from each datum.
@@
 
-Species' relative selection strength (RSS) for resources depended on temperature, and RSS was strongest for elevation and weakest for forest cover.
-Changes in RSS with temperature were strongest for elevation and generally weakest for distance from water, but there were no common trends across all species for any of the three resources.
-All species, with the exception of pumas, exhibited a temperature-dependent shift in their preference for forest cover.
-At higher temperatures, wolves, mountain goats, and grizzly bears became less selective for forest cover, while elk and caribou became more selective.
+The number of available points greatly outnumbered the number of used points (Fig.
+S14), especially after accounting for weighting based on the degree of autocorrelation and the number of home range crossings (Fig.
+S14).
+Species' relative selection strength (RSS) for resources depended on temperature, and RSS was generally strongest for elevation and weakest for forest cover.
+Changes in RSS with temperature were also strongest for elevation and generally weakest distance from water, but there were no common trends across all species for any of the three resources.
+All species, with the exception of pumas, exhibited a clear temperature-dependent shift in their preference for forest cover.
+At higher temperatures, wolves, mountain goats, and grizzly bears became less selective for forest cover, while elk and caribou shifted towards more intermediate forest cover without much of a change in preference width.
 All species shifted elevationally with temperature, although boreal caribou did not exhibit as strong of a change.
-As temperatures rose, elk, mountain goats, cougars, and boreal caribou increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation.
+As temperatures rose, elk, mountain goats, and pumas increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation.
 Most species generally remained within 5 km of water, and temperature did not affect their selection strength as much as for the other two resources.
 Estimated RSS values were generally most uncertain at extreme temperatures (Fig.
-S11).
+S13).
 
@@
 
-\noindent The direction and magnitude of change in distance traveled varied greatly between species, but worse climate-change scenarios generally corresponded to greater absolute changes in behavior (Fig.
-\@ref(fig:t-dist)).
-Under the best-case scenario, absolute projected changes by 2100 were small (approximately 0% to 2%), while under the worst-case scenario changes ranged from ~1% (boreal caribou) to ~25% (southern mountain caribou).
-Elk and boreal caribou showed relatively little projected changes in yearly distance traveled.
+\noindent Species varied in both the direction and magnitude of predicted change in distance traveled, but worse climate-change scenarios always corresponded to greater absolute changes in behavior (Figs.
+\@ref(fig:t-dist), S11, and S12).
+Under the best-case scenario, absolute changes by 2100 were small (approximately 0% to 4%), while under the worst-case scenario absolute changes ranged from ~2% (grizzly bear) to ~25% (pumas), although the models did not explicitly account for changes in hibernation phenology.
 Throughout BC (Fig.
-\@ref(fig:s-dist)), all species but elk and boreal caribou were expected to change their movement rates by 10 to 20% in 2100, relative to 2025, but the predicted changes varied spatially due to heterogeneity in climate change throughout the province.
-Across all species, absolute changes were smallest under the best-case scenario and greatest under the worst-case scenario.
+\@ref(fig:s-dist)), all species showed little to no absolute change under the best-case scenario (approximately 0-4% relative to 2025, on average), and approximately 2-25% average absolute change under the worst-case scenario, but the predicted changes varied spatially due to heterogeneity in climate change.
+Again, absolute changes were smallest under the best-case scenario and greatest under the worst-case scenario for all species.
 
-```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios often result in the greatest change and greatest uncertainty.} Lines indicate the projected relative change in mean distance travelled (probability of moving times speed when moving) due to changes in temperature within the species' current extent.
-Shaded areas indicate the 0.025 and 0.975 quantiles of the estimated means.
+```{r, eval=FALSE, echo=FALSE}
+# local predictions
+readRDS('../data/cc-hgam-projections.rds') %>%
+  group_by(species) %>%
+  filter(grepl('Best', scenario), year > 2095) %>%
+  mutate(d_median = d_median / d_ref) %>%
+  summarise(mean = mean(d_median)) %>%
+  mutate(mean = abs(mean - 1) * 100) %>%
+  arrange(mean)
+
+readRDS('../data/cc-hgam-projections.rds') %>%
+  group_by(species) %>%
+  filter(grepl('Worst', scenario), year > 2095) %>%
+  mutate(d_median = d_median / d_ref) %>%
+  summarise(mean = mean(d_median)) %>%
+  mutate(mean = abs(mean - 1) * 100) %>%
+  arrange(mean)
+
+# BC predictions
+readRDS('../data/cc-hgam-bc-projections-albers.rds') %>%
+  group_by(lab) %>%
+  filter(grepl('Best', scenario)) %>%
+  summarise(mean = mean(d)) %>%
+  mutate(mean = abs(mean - 1) * 100) %>%
+  arrange(mean)
+
+readRDS('../data/cc-hgam-projections.rds') %>%
+  group_by(species) %>%
+  filter(grepl('Worst', scenario), year > 2095) %>%
+  mutate(d_median = d_median / d_ref) %>%
+  summarise(mean = mean(d_median)) %>%
+  mutate(mean = abs(mean - 1) * 100) %>%
+  arrange(mean)
+```
+
+```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in distance travelled (probability of moving times speed when moving) due to changes in temperature within the species' current extent.
+Shaded areas indicate the 90% prediction interval within the range.
 Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs).
@@
 
-All species exhibited a decrease in RSS within their current range, irrespective of climate change scenario, but changes were often were only between -1% and -3% under the best-case scenario.
-Under the worst-case scenario, changes in RSS ranged from -8% to -14% (\@ref:fig(t-hrsfs)).
+All species exhibited a decrease in RSS within their current range, irrespective of climate change scenario.
+Changes were often were only between -1% and -3% under the best-case scenario but they ranged from -8% to -14% under the worst-case scenario (\@ref:fig(t-hrsfs)).
 As with distance traveled, the change in RSS in 2100 varied throughout BC for all species (Fig.
@@
 All species were predicted to exhibit a decrease in RSS throughout most of BC, with the exception of boreal caribou, which had a predicted decrease in approximately 37% of the province, including their current range.
-However, this was in great part due to the estimated increase in selection for high-elevation areas that are well outside the population's current range.
+The increase was in great part due to the estimated increase in selection for high-elevation areas that are well outside the population's current range.
 
@@
 
-```{r t-hrsfs, fig.cap="\\textbf{All species are expected to exhibit a decrease in relative selection strength (RSS) for their current range, irrespective of climate change scenario, but decreases are stronger in the worse-case scenario.} Lines indicate the projected change in median RSS within the species' current extent.
-Changes are relative to the median RSS in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
+```{r t-hrsfs, fig.cap="\\textbf{All species are expected to exhibit a decrease in relative selection strength (RSS) for their current range, irrespective of RSS quantile or climate change scenario, but decreases are stronger in the worse-case scenario.} Lines indicate the projected change in RSS for the bottom 5\\%, median, and top 5\\% RSS values within the species' current extent.
+Changes are relative to the mean RSS for each quantile in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
 knitr::include_graphics('../figures/rss-local-cc-predictions.png')
@@
 
-\noindent We demonstrated that temperature is an important determinant of how and where mammals move, even while accounting for other seasonal drivers of movement (e.g, changes in photoperiod, hormones, and resources).
-Disentangling temperature from other drivers allows us to predict how changes in climate will affect mammalian movement even if all other drivers remain approximately constant or adapt at negligible rates.
-However, as habitats warm and animals are exposed to increasingly frequent extreme and novel conditions, our ability to confidently predict mammals' responses decreases.
-Predicting at the fringes of the surface plots we present comes with substantial uncertainty, particularly given that the responses are nonlinear.
+\noindent We have demonstrated that temperature is an important determinant of how and where mammals move, even while accounting for seasonality (e.g, changes in photoperiod, hormones, and resources).
+Disentangling temperature from other drivers allowed us to predict how changes in climate will affect mammalian movement even if all other drivers remain approximately constant or adapt at negligible rates.
+However, predicting mammals' responses to climate change becomes increasingly complex as habitats warm and animals are exposed to increasingly frequent conditions that are both extreme and novel.
+Predicting at mammals' responses at the data-scarce fringes of the surface plots in Figs.
+\@ref(fig:dist) and \@ref(fig:hrsfs) comes with substantial uncertainty, particularly given that the responses are nonlinear.
 At unusually warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behavior and that of their competitors, predators, and/or prey can often be substantially different from the behavior at typical temperatures.
@@
 
-***continue reading from HERE***
-
-Our ability respond to current and future changes in climate is contingent on our ability to prepare for and predict change.
+Our success in respond to current and future changes in climate is contingent on our ability to prepare for and predict change.
 However, predicting animal behavior becomes increasingly complicated as the conditions animals are exposed to deviate from current, typical conditions, especially when responses are nonlinear.
@@
 
-\noindent The lack of any common effects of temperature across all species' movement rates indicates that the effects of climate change will vary across species and that responding to current conditions and preparing for future changes will require a variety of adaptations as communities adapt and respond in complex and interconnected ways.
+\noindent The lack of common effects of temperature across all species' movement rates indicates that responding to current conditions and preparing for future changes will require a variety of adaptations as communities adapt and respond in complex and interconnected ways.
 Although our models do not account for any physiological or explicit phenological changes (e.g., in the timing of sleep and hibernation), the tensor product terms in Fig.
-\@ref(fig:dist) suggest that warmer temperatures cause many species to alter their daily and seasonal activity patterns.
-For example, at cold temperatures (-20$^\circ$C), elk moved most in the early morning (6:00 PDT) and evening (18:00 PDT), but they moved almost four times more at 6:00 PDT and half as much at 18:00 PDT when temperatures were above 20$^\circ$C.
-This suggests that while elk maintain a preference for crepuscular activity, they adapt their movement behavior to move more during warmer periods of the day (i.e., evening) on cold days and during cooler periods (i.e., early morning) on warmer days.
-The strong reduction in wolves', elk's, pumas', and boreal caribou's mid-day movement rates when temperatures are above 20$^\circ$C suggests that the increasingly common and intense heat waves across British Columbia will often have community-wide impacts on movement rates, encounter rates [@martinez-garcia_how_2020], and potentially community structure.
+\@ref(fig:dist) suggest that warmer temperatures cause many species to alter their daily and seasonal activity patterns (most visible in pumas and grizzly bears).
+For example, when temperatures were above 0$^\circ$C, pumas moved most at night, but when temperatures were below 0$^\circ$C they tended to move more throughout the day.
+Throughout the year, they adapted their tolerance to temperature and moved less when it was relatively hot (for a given time of year), especially in spring and summer.
+The strong reduction in wolves', elk's, pumas', and boreal caribou's mid-day movement rates when summer temperatures are above 20$^\circ$C suggests that the increasingly common and intense heat waves across British Columbia will have community-wide impacts on movement rates, encounter rates, and potentially community structure [@martinez-garcia_how_2020].
 More work is necessary on interspecific responses to temperature, including the effects of temperature on predation rates [but see: @cunningham_opportunity_2021; @glass_trade-off_2021; @brivio_seeking_2024].
-The increase in intensity and frequency of extreme heat events [@bunnell_global_2011; @yao_emergence_2022] will likely also impact the occurrence and timing of hibernation [@wells_life_2022] and migration or seasonal range expansions [@carbeck_adaptation_2022; @morley_spatiotemporal_2021; @malpeli_ungulate_2022], not only as a direct effect of temperature, but also because of cascading changes in the availability and unpredictability of resources [@mezzini_how_2024].
+As indicated by grizzly bears' tendency to travel greater distances earlier in the year with increasing temperatures, the increase in intensity and frequency of extreme heat events [@bunnell_global_2011; @yao_emergence_2022] will likely also impact the occurrence and timing of hibernation [@wells_life_2022] and migration or seasonal range expansions [@carbeck_adaptation_2022; @morley_spatiotemporal_2021; @malpeli_ungulate_2022].
+This change will likely depend not only on the direct effect of temperature but also on cascading changes in the availability and unpredictability of resources [@mclellan_effect_2015; @pigeon_drivers_2016; @mezzini_how_2024].
 
 The greater uncertainty around the estimated movement rates at temperatures outside the typical temperatures for a given day (Figs.
-S2, S4-S6) highlights the importance of including appropriate measures of uncertainty when evaluating the consequences of climate change on mammals' movement behavior.
+S4, S6-S8) highlights the importance of including appropriate measures of uncertainty when evaluating the consequences of climate change on mammals' movement behavior.
 In this paper, we present our results cognizantly of the lack of uncertainty measures in many of our figures: We hoped to demonstrate how mammalian movement behavior depends on temperature while presenting the need for additional research.
@@
 The large range in predicted changes by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in figure \@ref(fig:s-hrsfs) highlight the need for careful planning that incorporates not only accurate estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates.
-The width of the averaged credible intervals in Figure \@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas.
-
-The widespread decrease in RSS in for areas with ranges with the top 5% RSS underscores the importance of moving beyond only using central estimates of change.
-While changes in the median RSS over time are small to negligible (yet never positive), decreases in the top 5% are more accentuated.
-This perspective is particularly important in the case of habitat selection modeling, since animals will be selecting most for (and thus spending more time in) habitats with high RSS values, so changes in the upper extremes are more important than changes in the median or lower extreme values.
-Similar arguments can be made for incorporating variation in temperature (including extreme events) when predicting future changes.
-Estimating changes in future weather with monthly means alone is insufficient because a change of 1-3 in mean monthly temperature may result in negligible predicted changes while ignoring changes in the frequency and intensity of extreme temperatures [@bunnell_global_2011; @zhang_increased_2023; @yao_emergence_2022].
+The width of the prediction intervals in Figure \@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas.
 
@@
 Together, the predicted changes in distance traveled and habitat selection indicate that climate change will have complex consequences for populations as well as cascade effects within ecosystems as species adapt to the changing environments [or fail to do so: @botero_evolutionary_2015; @polazzo_measuring_2024].
-The resulting changes in mammalian movement behavior will have ramifications in individuals' encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@need_ref], and food webs [@need_ref] as landscapes change and animals face increasing pressures from human activity [@sih_evolution_2011], including habitat fragmentation [@needs_ref], habitat loss [@needs_ref], and greater human presence in wild spaces [@weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022].
+The resulting changes in mammalian movement behavior will have ramifications in individuals' encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@needs_ref], and food webs [@needs_ref] as landscapes change and animals face increasing pressures from human activity [@sih_evolution_2011], including habitat fragmentation [@needs_ref], habitat loss [@needs_ref], and greater human presence in wild spaces [@weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022].
 

### 2025-04-02 writing/appendix.Rmd
changed appendix scatter plots to hex plots

@@
 
-```{r pred-vs-pred-hgams, fig.cap="Scatterplot of the fitted values from the HGAMs with and without including temperature for the probability of movement (\\textbf{A}) and speed when moving (\\textbf{B})."}
+```{r pred-vs-pred-hgams, fig.cap="Hexplot of the fitted values from the HGAMs with and without including temperature for the probability of movement (\\textbf{A}) and speed when moving (\\textbf{B})."}
 knitr::include_graphics('../figures/hgam-with-without-temp-prediction-agreement.png')
@@
 ```{r op, fig.cap="Relationsips between the observed and predicted values for probability of movement (\\textbf{A}) and speed given that an animal was moving (\\textbf{B}).
-The color in panel \\textbf{A} indicates the number of points used to calculate the estimate (on the $\\log-{2}$ scale), while the grey lines in both panels indicate perfect prediction (i.e., the 1:1 line)."}
+The color in panel \\textbf{A} indicates the number of points used to calculate the estimate (on the $\\log_{2}$ scale), while the grey lines in both panels indicate perfect prediction (i.e., the 1:1 line)."}
 knitr::include_graphics('../figures/p-and-speed-observed-vs-predicted.png')
@@
 
-```{r pred-vs-pred-hrsfs, fig.cap="Scatterplot of the fitted values from the HRSFs with and without including temperature."}
+```{r pred-vs-pred-hrsfs, fig.cap="Hexplot of the fitted values from the HRSFs with and without including temperature."}
 knitr::include_graphics('../figures/hrsf-with-without-temp-prediction-agreement.png')

### 2025-04-02 analysis/06-temperature-hgams.R

@@ -81,7 +81,7 @@ splits <-
                      pull(speed_est)))
          }),
   ) %>%
-  pivot_longer(c(Manual, `k means`), values_to = 'split',
+  pivot_longer(c(manual, `k means`), values_to = 'split',
                names_to = 'method')
 
 hists <- list()
@@ -131,21 +131,6 @@ d %>%
   group_by(species) %>%
   summarise(prop_moving = mean(moving),
             min_speed = min(speed_est[moving]))
-
-# some species have data with extreme temperatures but little data ----
-# but clipping temperature data to the 0.8 quantile doesn't change the high
-# P(moving) at cold temperatures for goats
-ggplot(d, aes(temp_c)) +
-  facet_wrap(~ species, scales = 'free') +
-  geom_histogram(color = 'black', binwidth = 1, fill = 'grey') +
-  geom_vline(aes(xintercept = value), color = 'red',
-             d %>%
-               group_by(species) %>%
-               summarize(lwr = quantile(temp_c, 0.025),
-                         upr = quantile(temp_c, 0.975)) %>%
-               pivot_longer(-species)) +
-  ylim(c(0, NA)) +
-  xlab(paste0('Temperature (\U00B0', 'C)'))
 
 # dataset of non-zero speeds only
 d_2 <- filter(d, moving)
@@ -395,17 +380,18 @@ if(file.exists('models/gamma-gam.rds')) {
 
 s_op <-
   mutate(d_2, mu = predict(m_2, type = 'response')) %>%
-  ggplot(aes(mu, speed_est, color = species)) +
+  ggplot(aes(mu, speed_est)) +
   facet_wrap(~ lab, scales = 'free', nrow = 2, labeller = label_parsed) +
-  geom_point(alpha = 0.1, size = 0.1) +
-  geom_abline(intercept = 0, slope = 1, color = 'grey') +
+  geom_hex(aes(fill = log10(after_stat(count)))) +
+  geom_abline(intercept = 0, slope = 1, color = 'black') +
   theme(legend.position = 'none') +
-  scale_color_manual('Species', values = PAL,
-                     aesthetics = c('color', 'fill')) +
+  scale_fill_iridescent(name = expression(bold(Count~(log['10']~scale))),
+                        range = c(0.3, 1), breaks = 0:3,
+                        labels = round(10^(c(0:3)))) +
   ylim(c(0, NA)) +
[…]

### 2025-04-02 analysis/06-temperature-hrsfs.R

@@ -7,6 +7,7 @@ library('terra')   # for rasters
 library('purrr')   # for functional programming
 library('ggplot2') # for fancy plots
 library('gratia')  # for graceful GAM plots
+library('khroma')  # for colorblind-friendly figures
 source('analysis/figures/default-ggplot-theme.R')
 
 # import data ----
@@ -474,12 +475,16 @@ p_fits <-
   fits %>%
   ggplot(aes(fits_wo, fits_w)) +
   facet_wrap(~ lab, labeller = label_parsed, scales = 'free') +
-  geom_point(shape = '.') +
-  geom_abline(slope = 1, intercept = 0, color = 'red') +
+  geom_hex(aes(fill = log10(after_stat(count)))) +
+  geom_abline(slope = 1, intercept = 0, color = 'black') +
   labs(x = 'log(RSS) without temperature',
-       y = 'log(RSS) with temperature')
+       y = 'log(RSS) with temperature') +
+  scale_fill_iridescent(name = expression(bold(Count~(log['10']~scale))),
+                        range = c(0.3, 1), breaks = (0:3) * 2,
+                        labels = round(10^(c(0:3) * 2))) +
+  theme(legend.position = 'inside', legend.position.inside = c(5/6, 1/6))
 
 ggsave('figures/hrsf-with-without-temp-prediction-agreement.png',
-       plot = p_fits, width = 12, height = 12, units = 'in', dpi = 600,
+       plot = p_fits, width = 8, height = 8, units = 'in', dpi = 600,
        bg = 'white')
 beepr::beep()

### 2025-04-14 analysis/05-add-speeds-and-temperature.R
fixed finding time of a raster with `terra 0.99.0`

@@ -78,7 +78,8 @@ d2 <-
       nest(hourly_data = ! c(nearest_hour)) %>%
       mutate(
         layer = map_int(nearest_hour, \(nh)
-                        which2(as.POSIXct(r@cpp$time, tz = 'UTC') == nh)),
+                        #' use `r@cpp` with earlier versions of `terra`
+                        which2(as.POSIXct(r@ptr$time, tz = 'UTC') == nh)),
         hourly_data = map2(
           layer, hourly_data,
           \(.l, hourly_d) {

### 2025-05-07 analysis/06-temperature-hgams.R
removed old comment

@@ -207,9 +207,6 @@ ggsave('figures/temperature-movement-rates-hist.png',
        width = 16, height = 17, units = 'in', dpi = 600, bg = 'white')
 
 # model P(movement) ----
-#' using `bs = 'fs'` rather than `by` smooths because to keep the models
-#' reasonably constrained. Using `by` gives extremely high `P(moving)`
-#' in spring for grizzly bears
 #' fits in 3-4 minutes
 if(file.exists('models/binomial-gam.rds')) {
   m_1 <- readRDS('models/binomial-gam.rds')

### 2025-05-07 writing/manuscript.Rmd
added tucker 2018
addressed most of Mike's comments

@@
 abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on individuals' fitness and habitat selection at the population, community, and ecosystem levels.
-Previous research has quantified the effects of temperature on mammalian movement behavior, but few have quantified the effects of future climate change on fine-scale movement and habitat selection.
+Previous research has quantified the effects of temperature on mammalian movement behaviour, but few have quantified the effects of future climate change on fine-scale movement and habitat selection.
 We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on mammals' movement rates and habitat selection.
@@
 
+<!-- using  "behaviOR" in key words to increase audience -->
+
 \noindent \textbf{Key words:} climate change, temperature, mammals, animal movement, movement behavior, habitat selection
@@
 \noindent `r abstract`
-
-<!-- place table of contents on a new page -->
-\newpage
-
-\singlespacing
-
-\tableofcontents
 
@@
 
-\noindent For most animals, movement is a rapid, low-cost, and low-risk response to a variety of stimuli [@nathan_movement_2008].
+\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008].
 Animals move to maximize resource availability [e.g., food, water, shelter, see: @charnov_optimal_1976; @kacelnik_ideal_1992], optimize habitat use [@winter_forecasting_2024], and escape predation [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024].
 However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], the ease of movement [e.g., snow depth and density, see @montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023].
-Additionally, temperature also affects movement indirectly by through other drivers of movement behavior, such as trophic interactions [@hegel_interacting_2010] and the value of resources [e.g., water, heat, shelter -- see: @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019].
-Changes in temperature thus affects how animals move [e.g., movement frequency and speed -- their movement behavior: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].
+Additionally, temperature also affects movement indirectly by through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010] and the value of resources [e.g., water, heat, shelter -- see: @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019].
+Changes in temperature thus affects how animals move [e.g., movement frequency and speed -- their movement behaviour: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].
 
-Temperature also affects the consequences of animal movement, such as encounter rates with resources [@mezzini_how_2024], other animals [@martinez-garcia_how_2020], and humans [@weststrate_how_2024], and these changes are exacerbated at extreme temperatures [@wilson_stenotherms_2001; @berger_climate_2018; @giroux_activity_2023].
-But what counts as "extreme" varies across species: thermal specialists are more sensitive to small changes outside their typical range [e.g., @giroux_activity_2023], while temperature generalists can adapt more easily to change [@dupont_beyond_2024].
-However, generalists' grater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019], affect many factors [e.g., temperature, phenology, and resource availability: @polazzo_measuring_2024], or are too rapid to adapt within the span of a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].
-Thus, it remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially when populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018].
+Temperature also affects the consequences of animal movement, such as encounter rates with resources [@mezzini_how_2025], other animals [@martinez-garcia_how_2020], and humans [@weststrate_how_2024], and these changes are exacerbated at extreme temperatures [@wilson_stenotherms_2001; @berger_climate_2018; @giroux_activity_2023].
+But what counts as "extreme" varies across species: thermal specialists are more sensitive to small changes outside their typical range [e.g., @giroux_activity_2023], while thermal generalists can adapt more easily to change [@dupont_beyond_2024].
+However, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019], affect many factors [e.g., temperature, phenology, and resource availability: @polazzo_measuring_2024], or are too rapid to respond within the span of a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].
 
@@
 
-Uncertainty in future conditions and their consequences on ecosystems presents a need for a comprehensive understanding of (1) the direct effects of temperature on animal movement behavior and (2) the implications this will have on populations and species' adaptability and resiliency, community structure, and biodiversity at large.
-Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], movement behavior [@melin_effect_2023], thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [@mccain_assessing_2019], but it is still unclear how mammals will adapt their small-scale movement and habitat selection to changes in temperature during the 21^st^ century [@intergovernmental_panel_on_climate_change_climate_2023; but see @hetem_responses_2014; @winter_forecasting_2024].
-In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures and severe forest fires [@intergovernmental_panel_on_climate_change_climate_2023; @zurowski_summer_2023].
-Over the next 100 years, these changes are expected to worsen and continue to impact mammals' fitness, movement behavior, and habitat selection as animals cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and more frequent and intense extreme events [@bunnell_global_2011], along with increasing anthropogenic pressure [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024].
+In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures and severe forest fires [@ipcc_climate_2023; @zurowski_summer_2023].
+Over the next 100 years, these changes are expected to worsen and continue to impact mammals' fitness, movement behaviour, and habitat selection as animals cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and more frequent and intense extreme events [@bunnell_global_2011], along with increasing anthropogenic pressure [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024].
+Consequently, it remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially when populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018].
+Uncertainty in future conditions and their consequences on ecosystems present a need for a comprehensive understanding of (1) the direct effects of temperature on animal movement behaviour and (2) the implications this will have on populations and species' adaptability and resiliency, community structure, and biodiversity at large.
+Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], movement behaviour [@melin_effect_2023], thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [@mccain_assessing_2019], but it is still unclear how mammals will adapt their small-scale movement and habitat selection to changes in temperature during the 21^st^ century [@ipcc_climate_2023; but see @hetem_responses_2014; @winter_forecasting_2024].
 
@@
 Predicting how climate change will affect how, when, and where animals will move is necessary for identifying which landscapes are most valuable now as well as in the future.
-Failing to understand how changes in temperature will affect mammalian movement behavior and habitat selection will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.
+Failing to understand how changes in temperature will affect mammalian movement behaviour and habitat selection will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.
 
@@
 This paper provides an analysis of the effects of air temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (*Canis lupus*, *Cervus canadensis*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, and *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
-Using over 25 years of telemetry data thoughout a large spatial range of British Columbia (Fig.
-\@ref(fig:bc-map)) and hourly air temperature data, we estimate how mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature.
+Using over 25 years of telemetry data thoughout a large spatial range of British Columbia and hourly air temperature data, we estimate how mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature.
 We then pair the estimated responses with climate change projections to forecast the species' movement during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSPs -- see @riahi_shared_2017].
-We disentangle the direct and indirect effects of temperature on the movement behavior of terrestrial mammals and discuss the consequences of these changes at the individual, population, and landscape levels.
+We disentangle the direct and indirect effects of temperature on the movement behaviour of terrestrial mammals and discuss the consequences of these changes at the individual, population, and landscape levels.
 
@@
             Animals = n_distinct(animal),
-            n_missing = sum(na_speed),
+            n_missing = sum(! na_speed),
             .groups = 'drop') %>%
@@
   rename('\U0394t (hours)' = dt,
-         'NA speed' = n_missing)
+         'Has speed' = n_missing)
 
-knitr::kable(dt_data, format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$; stratified by individual), number of animals (Animals), and number of animals with unresolved speed (NA speed) for each of the species in this study.', escape = FALSE, longtable = FALSE, booktabs = TRUE)
+knitr::kable(dt_data, format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$; stratified by individual), number of animals (Animals), and number of animals with resolved speed (Has speed) for each of the species in this study.', escape = FALSE, longtable = FALSE, booktabs = TRUE)
 ```
@@
 
-This manuscript leverages four main datasets: (1) a collection of GPS telemetry data; (2) historical hourly reanalyzed air temperature from the ERA5 dataset from the European Center for Medium-range Weather Forecasting; (3) resource rasters of percent forest cover, elevation, and distance from water; and (4) climate change projections under four Shared Socioeconomic Pathways [SSPs, see @riahi_shared_2017].
+This study leverages four main datasets: (1) a multi-species collection of GPS telemetry data (Fig.
+\@ref(fig:bc-map)); (2) historical hourly reanalyzed air temperature from the ERA5 dataset from the European Center for Medium-range Weather Forecasting; (3) resource rasters of percent forest cover, elevation, and distance from water; and (4) climate change projections under four Shared Socioeconomic Pathways [SSPs, see @riahi_shared_2017].
 We detail the data sources and analyses in the sections below.<!--, and more details are available in appendices A (estimating how temperature affects movement frequency and speed when moving), B (fitting Hierarchical Resource Selection Functions), and C (simulating hourly temperature data to 2100 using the four SSPs).-->
@@
 \noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)).
-Clearly erroneous GPS locations were removed using diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points.
-Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated stationary behavior.
+***MISSING INFO ON OTHER TELEMETRY DATA*** Clearly erroneous GPS locations were removed using diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points.
+Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated stationary behaviour.
 The script used to clean the data and all associated custom functions are available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).
@@
 
-We simulated hourly variation in weather by assuming temperature followed a normal distribution with mean specified by the monthly `climatenaR` climate projections and variance as specified by the Gamma GAMLS.
-We then predicted changes in movement behavior and habitat selection as a function of the simulated temperature using the HGAMs and HRSFs.
+We simulated hourly variation in future weather by assuming hourly temperature followed a normal distribution with mean specified by the monthly `climatenaR` climate projections and variance as specified by the Gamma GAMLS.
+We then predicted changes in movement behaviour and habitat selection as a function of the simulated temperature using the HGAMs and HRSFs.
 For each month within each year from 2020 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantile by increments of 0.1, and we weighted each quantile proportionally to the (normalized) Gaussian probability density for each quantile<!-- (Appendix C)-->.
@@
 We then estimated mammals' instantaneous speeds by applying the `ctmm::speeds()` function on all models with finite speed estimates [415 of 433, see @fleming_fine-scale_2014; @noonan_scale-insensitive_2019].
-The 18 telemetries with insufficiently fine sampling were for one mountain goat, 15 boreal caribou, and 2 southern mountain caribou (Table \@ref(tab:data-table)).
+The 18 animals with insufficiently fine sampling were for one mountain goat, 15 boreal caribou, and 2 southern mountain caribou (Table \@ref(tab:data-table)).
 
-Since `ctmm`'s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed could be 0.
+Since `ctmm``s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed could be 0.
 We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed.
@@
 
+<!--
+
+keeping for thesis:
+
 ### Disentangling direct and indirect effects
 
-\noindent In this manuscript, we separate the effects of temperature on mammalian movement into indirect and direct effects.
+\noindent In this study, we separate the effects of temperature on mammalian movement into indirect and direct effects.
 We call "indirect" all effects that can be intervened on by altering an animal's habitat [its geographic space *sensu* @matthiopoulos_defining_2022], and we refer to all other effects as "direct".
-For example, an increase in temperature may push an animal to prefer higher forest cover in search of shade, and conservationists can thus mitigate these indirect effects of temperature on movement behavior by increasing the availability of forested areas, since the effect of temperature on movement is conditional on forest cover availability.
-In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022], so it is not possible to mitigate the effects of temperature on movement behavior (other than changing temperature directly).
+For example, an increase in temperature may push an animal to prefer higher forest cover in search of shade, and conservationists can thus mitigate these indirect effects of temperature on movement behaviour by increasing the availability of forested areas, since the effect of temperature on movement is conditional on forest cover availability.
+In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022], so it is not possible to mitigate the effects of temperature on movement behaviour (other than changing temperature directly).
 We provide the relevant Directed Acyclical Graphs in the Appendix (Figs.
 S1-S2).
+-->
 
@@
 
-\noindent We estimated the effects of temperature on mammals' movement state (moving or not) and speed (when moving) using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019 and the code chunk below] with the `mgcv` package for `R` [version 1.9-1, @wood_generalized_2017].
-The first HGAM estimated the probability that an animal was moving with a binomial family of distributions and logit link function.
-The second HGAM estimated an animal's speed (when moving) with a gamma family of distributions and log link function.
-The HGAMs included random intercepts for each animal (`s(..., bs = 're')`), fixed-effect intercepts for each species, and species-level `by` smooths for time of day, day of year, and temperature [`s(..., by = species)`, see model I in Figure 4 of @pedersen_hierarchical_2019].
-Additionally, the models had three tensor product interaction terms (`ti()`) for each species: (1) day of year and time of day, (2) temperature and time of day, and (3) temperature and day of year.
+\noindent We estimated the effects of temperature ($^\circ$C) on mammals' movement state (moving or not) and speed when moving using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019 and the code chunk below] with the `mgcv` package for `R` [version 1.9-1, @wood_generalized_2017].
+The first HGAM estimated the probability that an animal was moving, $P(M)$, with a binomial family of distributions and logit link function.
+The response variable `moving` was coded as 1 if moving and 0 if not.
+The second HGAM estimated an animal's speed when moving (`speed`) with a gamma family of distributions and log link function.
+The HGAMs included random intercepts for each animal (`s(animal, bs = 're')`), fixed-effect intercepts for each species (`species`), and species-level `by` smooths (`s(..., by = species)`), which allowed each species' term to be estimated independent of other species [see model I in Figure 4 of @pedersen_hierarchical_2019].
+The `by` smooths accounted for trends in time of day (in Pacific Daylight Time; `tod_pdt`), day of year (`doy`), and temperature (`temp_c`).
+Additionally, the models had three tensor product interaction terms (`ti()`) `by` each species: (1) day of year and time of day, (2) temperature and time of day, and (3) temperature and day of year.
 These three terms accounted for: (1) seasonal changes in day length, (2) hourly changes in the response to temperature (e.g., changes in nocturnality), and (3) seasonal changes in the response to temperature (e.g., changes in coats and migration timing).
-The HGAMs accounted for the cyclicity of time of day and day of year using cyclic cubic splines [`bs = 'cc'`, @wood_generalized_2017].
+Finally, smooth terms of log-transformed hours between GPS locations (`dt`) accounted for irregular sampling rates.
+A global term of `log(dt)` accounted for the overall effect of sampling interval, while a factor-smooth interaction term (`s(log(dt), species, bs = 'fs')`) allowed for species-level deviations from the global term while assuming a common smoothness parameter across species [see model GS in Figure 4 of @pedersen_hierarchical_2019].
+The HGAMs accounted for the cyclicity of time of day and day of year using cyclic cubic splines [`bs = 'cc'`$;$ see p.
+202 of @wood_generalized_2017].
 Together, the binomial HGAM and the gamma HGAM inform us on an animal's long-term average speed, since it is the product of the probability of moving and its average speed when moving.
@@
 
-\noindent We estimated the effects of temperature on each species' selection for percent forest cover, elevation, and distance from water by fitting a Hierarchical Resource Selection Function for each species using an HGAM with a Poisson family of distributions and log link function [Appendix B, also see @aarts_estimating_2008].
+\noindent We estimated the effects of temperature on each species' selection for percent forest cover (`forest_perc`), elevation (`elevation_m`), and distance from water (`dist_water_m`) by fitting a Hierarchical Resource Selection Function for each species using an HGAM with a Poisson family of distributions and log link function [Appendix B, also see @aarts_estimating_2008].
 We accounted for the spatiotemporal autocorrelation in the telemetry locations by weighting each point based on the telemetry's Autocorrelated Kernel Density Estimate [@fleming_new_2017; @noonan_comprehensive_2019] to produce estimates of second- and third-order habitat selection [*sensu* @johnson_comparison_1980].
@@
 rsf <- bam(
-  detected ~ # 1 for telemetry locations, 0 for quadrature data
+  detected ~ # 1 for telemetry locations, 0 for quadrature points
     # species-level average resource preference
@@
     # changes in preference with temperature
-    ti(forest_perc, temperature_C, k = 6, bs = 'tp') +
-    ti(elevation_m, temperature_C, k = 6, bs = 'tp') +
-    ti(dist_water_m, temperature_C, k = 6, bs = 'tp') +
+    ti(forest_perc, temp_c, k = 6, bs = 'tp') +
+    ti(elevation_m, temp_c, k = 6, bs = 'tp') +
+    ti(dist_water_m, temp_c, k = 6, bs = 'tp') +
     # include marginals of temperature to remove sampling biases
-    s(temperature_C, k = 4, bs = 'tp') +
-    s(temperature_C, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')),
+    s(temp_c, k = 4, bs = 'tp') +
+    s(temp_c, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')),
   family = poisson(link = 'log'),
-  data = d,
+  data = d, # species-specific dataset
   weights = weight, # based on AKDE
@@
 Smooth effects of percent forest cover, elevation, and distance to water accounted for the species-level average selection strength for each resource.
-A random effect for each individual animal corrected for uneven sampling across individuals, while factor smooth interaction terms (`bs = 'fs'`) accounted for individual-level resource selection (i.e., individual-level deviations from the species-level average).
+A random effect for each individual animal (`s(animal, bs = 're')`) corrected for uneven sampling across individuals, while factor smooth interaction terms (`bs = 'fs'`) accounted for individual-level resource selection (i.e., individual-level deviations from the species-level average).
 Tensor interaction product terms (`ti()`) of the three resources and temperature estimated the change in resource selection at different temperatures.
@@
 Detections were weighted proportionally to their degree of independence from other temporally proximate detections [`weights = weight` -- see Appendix B and @alston_mitigating_2022], while quadrature points had a weight of 1.
+Quadrature points were obtained using all raster cells in the 99.9% AKDE percentile.
+The number of quadrature locations greatly outnumbered the number of observed locations (Fig.
+S13), especially after accounting for weighting based on the degree of autocorrelation and the number of home range crossings (Fig.
+S14).
 
@@
 S4).
-Species differed in overall mean probabilities of movement (range: 0.05 -- 0.3), mean speed when moving (range: 0.42 -- 2.7 km/day), and mean distance traveled (range: 0.04 -- 0.6 km/day), even after predicting for one-hour time intervals for all species (Table \@ref(tab:gam-intercepts)).
+Species differed in overall mean probabilities of movement (range: 0.05 -- 0.3), mean speed when moving (range: 0.42 -- 2.7 km/day), and mean distance traveled (range: 0.04 -- 0.6 km/day; Table \@ref(tab:gam-intercepts)).
 Grizzly bears tended to move least often (probability of moving $P(M) \approx 0.05$), while wolves and pumas moved most often ($P(M) \gtrapprox 0.23$).
@@
 
-\footnotesize
+\scriptsize
 
@@
                      'E(D|T)' = `P(M|T)` * `E(S|T)`) %>%
+  mutate(Species = case_when(
+    Species == 'Rangifer tarandus (s. mountain)' ~ 'R. tarandus (SM)',
+    Species == 'Ursus arctos horribilis' ~ 'U. arctos horr.',
+    Species == 'Puma concolor' ~ 'Puma concolor',
+    Species == 'Cervus canadensis' ~ 'C. canadensis',
+    Species == 'Rangifer tarandus (boreal)' ~ 'R. tarandus (boreal)',
+    Species == 'Canis lupus' ~ 'Canis lupus',
+    Species == 'Oreamnos americanus' ~ 'O. americanus')) %>%
   mutate(Species = paste0('\\emph{', Species, '}') %>%
@@
   arrange(Species) %>%
-  mutate(Species = gsub('angifer', '\\.', Species)) %>%
   rename_with(\(.x) paste0('$', .x, '$'), ! Species) %>%
   rename_with(\(.x) gsub('\\|', '\\\\vert ', .x)) %>%
-  rename_with(\(.x) gsub('E', '\\\\mathbb E', .x)) %>%
-  knitr::kable(format = 'markdown', digits = 2, caption = paste0("Mean probability of movement (P(M)), speed when moving ($\\mathbb E(S)$), and distance travelled ($\\mathbb E(D)$) as estimated by models without and with temperature ($(...
+  rename_with(\(.x) gsub('P', '\\\\hat{P}', .x)) %>%
+  rename_with(\(.x) gsub('E', '\\\\hat{\\\\mathbb E}', .x)) %>%
+  # # add units (added in caption)
+  # rename('$\\widehat{\\mathbb E(S)}$, (m/s)' = `$\\widehat{\\mathbb E(S)}$`,
+  #        '$\\widehat{\\mathbb E(D)}$, (m/s)' = `$\\widehat{\\mathbb E(D)}$`,
+  #        '$\\widehat{\\mathbb E(S\\vert T)}$, (m/s)' = `$\\widehat{\\mathbb E(S\\vert T)}$`,
+  #        '$\\widehat{\\mathbb E(D\\vert T)}$, (m/s)' = `$\\widehat{\\mathbb E(D\\vert T)}$`) %>%
+  knitr::kable(format = 'markdown', digits = 2, caption = paste0("Mean probability of movement (P(M)), speed when moving ($\\mathbb E(S)$, m/s), and distance travelled ($\\mathbb E(D)$, m/s) as estimated by models without and with temperature ($\\mathbb E(...
 | T)$), after post-stratifting to a 1-hour sampling rate and a temperature of T = 0\U00B0", "C."))
@@
 
-At average temperatures, wolves preferred dense forest cover ($\gtrapprox$ 50%) high elevations, and distances from water < 10 km; elk preferred intermediate forest cover, elevations between 1 and 2 km, and distances from water of 5-15 km; mountain goats preferred sparse (< 50%) forest cover, elevations between 1 and 2 km; pumas preferred high forest cover, an elevation of ~ 1 km, and distances from water < 10 km; boreal caribou selected for intermediate to high forest cover, elevations near 500 m, and distances from water < 10 km; southern mountain caribou selected for dense forest cover, elevations near 2 km, and distances from water < 5 km; and grizzly bears selected for sparse forest cover (25-50%), elevation near 1 km, and distances from water < 2 km.
+Near 0$^\circ$C, wolves preferred dense forest cover ($\gtrapprox$ 50%) high elevations, and distances from water < 10 km; elk preferred intermediate forest cover, elevations between 1 and 2 km, and distances from water of 5-15 km; mountain goats preferred sparse (< 50%) forest cover, elevations between 1 and 2 km; pumas preferred high forest cover, an elevation of ~ 1 km, and distances from water < 10 km; boreal caribou selected for intermediate to high forest cover, elevations near 500 m, and distances from water < 10 km; southern mountain caribou selected for dense forest cover, elevations near 2 km, and distances from water < 5 km; and grizzly bears selected for sparse forest cover (25-50%), elevation near 1 km, and distances from water < 2 km.
 
-Including temperature in the models always increased within-sample prediction, but the change was not always substantial (deviance explained increased between 0.2% and 18.6%, with all HRSFs having an increase $\ge 4.6$), and there was relatively strong agreement between models with and without temperature (Figs.
-S5 and S13), but including temperature resulted in substantially better fits for all models, especially HRSFs (all $\Delta$AIC $\le$ -342; Table \@ref(tab:temp-comparison)).
+Including temperature in the models always increased within-sample prediction, but the change was not always substantial (deviance explained increased between +0.2% and +18.6%, with all HRSFs having an increase $\ge +4.6$%), and there was relatively strong agreement between models with and without temperature (Figs.
+S5 and S13), but including temperature resulted in substantially lower AIC values for all models, especially HRSFs (all $\Delta$AIC $\le$ -342; Table \@ref(tab:temp-comparison)).
+Although this may be interpreted as better fit, it is important to note that AIC generally decreases with more complex terms when sample sizes are large, even if the more complex model is not closer to the true model [see section 1.8.6, pages 52-54 of @wood_generalized_2017].
 
@@
 temp_comp <- tibble(
-  Model = c('P(moving)', 'Speed', rep('HRSF', 7)),
+  Model = c('$P(M)$', 'Speed', rep('HRSF', 7)),
   Species = c('All', 'All', sort(as.character(SPECIES))),
@@
 S9).
-All species adapted their daily and seasonal movement behavior to changes in temperature (Fig.
+All species adapted their daily and seasonal movement behaviour to changes in temperature (Fig.
 \@ref(fig:dist)).
-Wolves, elk, pumas, and boreal caribou reduced their mid-day and summer movement rates when temperatures were above 20$^\circ$C.
+For example, wolves, elk, pumas, and boreal caribou reduced their mid-day and summer movement rates when temperatures were above 20$^\circ$C.
 However, uncertainties around the estimated effects were often higher at extreme temperatures due to data scarcity.
-
 All species' estimated probability of moving decreased with sampling intervals approximately above 1 hour, and most species' estimated speed when moving decreased monotonically with sampling interval, although the estimated trends were highly uncertain for some species (Fig.
 S10).
-While the `ctmm` models produced scale-independent estimates of speed [i.e., model interpretation is independent of sampling interval: @noonan_scale-insensitive_2019], the accuracy, size, and stability of speed estimates still depended on the GPS sampling frequency.
-This is because coarsely-sampled movement data contains information on large-scale movements (e.g., range crossings, migrations) but not fine-scale movements [e.g., first-order habitat selection *sensu* @johnson_comparison_1980; also see @nathan_big-data_2022].
-Using the boreal caribou as an example, the 13-hour sampling interval allows us to reasonably estimate the caribou's movement path at a temporal scale of approximately 13 hours (or greater), but we cannot produce reasonable movement trajectories at a much finer (e.g., hourly) scale.
-Consequently, we suggest being cautious when comparing estimated movement behaviors across species, even though all predictions have been corrected to the hourly timescale by predicting for 1-hour time intervals (i.e., `dt_hours = 1`).
 
@@
 
-The number of available points greatly outnumbered the number of used points (Fig.
-S14), especially after accounting for weighting based on the degree of autocorrelation and the number of home range crossings (Fig.
-S14).
-Species' relative selection strength (RSS) for resources depended on temperature, and RSS was generally strongest for elevation and weakest for forest cover.
+Species' relative selection strength (RSS) was generally strongest for elevation and weakest for forest cover, but species' RSS for resources depended on temperature.
 Changes in RSS with temperature were also strongest for elevation and generally weakest distance from water, but there were no common trends across all species for any of the three resources.
@@
 Most species generally remained within 5 km of water, and temperature did not affect their selection strength as much as for the other two resources.
-Estimated RSS values were generally most uncertain at extreme temperatures (Fig.
+Again, estimated RSS values were generally most uncertain at extreme temperatures (Fig.
 S13).
@@
 
-\noindent Species varied in both the direction and magnitude of predicted change in distance traveled, but worse climate-change scenarios always corresponded to greater absolute changes in behavior (Figs.
+\noindent Species varied in both the direction and magnitude of predicted change in distance traveled, but worse climate-change scenarios always corresponded to greater absolute changes in behaviour (Figs.
 \@ref(fig:t-dist), S11, and S12).
@@
 Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs).
-The projections only account for changes movement behavior (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
+The projections only account for changes movement behaviour (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
 knitr::include_graphics('../figures/distance-travelled-local-cc-predictions.png')
@@
 All species exhibited a decrease in RSS within their current range, irrespective of climate change scenario.
-Changes were often were only between -1% and -3% under the best-case scenario but they ranged from -8% to -14% under the worst-case scenario (\@ref:fig(t-hrsfs)).
+Changes were often were only between -1% and -3% under the best-case scenario but they ranged from -8% to -14% under the worst-case scenario (\@ref(fig:t-hrsfs)).
 As with distance traveled, the change in RSS in 2100 varied throughout BC for all species (Fig.
 \@ref(fig:s-hrsfs)).
-All species were predicted to exhibit a decrease in RSS throughout most of BC, with the exception of boreal caribou, which had a predicted decrease in approximately 37% of the province, including their current range.
+All species were predicted to exhibit a decrease in RSS throughout most of BC, with the exception of boreal caribou, which had a predicted increase in approximately 67% of the province, including their current range.
 The increase was in great part due to the estimated increase in selection for high-elevation areas that are well outside the population's current range.
@@
   summarize(p_decrease = mean(lambda < 1)) %>%
-  filter(p_decrease < 0.5)
+  filter(p_decrease < 0.5) %>%
+  mutate(p_increase = 1 - p_decrease)
 ```
@@
 
-\noindent We have demonstrated that temperature is an important determinant of how and where mammals move, even while accounting for seasonality (e.g, changes in photoperiod, hormones, and resources).
+\noindent We have demonstrated that temperature is an important determinant of whether, how, and where mammals move, even while accounting for seasonality (e.g, changes in photoperiod, hormones, and resources).
 Disentangling temperature from other drivers allowed us to predict how changes in climate will affect mammalian movement even if all other drivers remain approximately constant or adapt at negligible rates.
 However, predicting mammals' responses to climate change becomes increasingly complex as habitats warm and animals are exposed to increasingly frequent conditions that are both extreme and novel.
-Predicting at mammals' responses at the data-scarce fringes of the surface plots in Figs.
+Predicting mammals' responses at the data-scarce fringes of the surface plots in Figs.
 \@ref(fig:dist) and \@ref(fig:hrsfs) comes with substantial uncertainty, particularly given that the responses are nonlinear.
-At unusually warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behavior and that of their competitors, predators, and/or prey can often be substantially different from the behavior at typical temperatures.
-As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], mammals will be increasingly forced to atypical behaviors, which will have the potential to substantially alter community structures and behaviors, both during such events and afterwards [@zhang_increased_2023].
-For instance, changes in climate and phenology impact the life history and behavior of many hibernating mammals [@wells_life_2022], and hot weather can also affects mammals' sleep quality [e.g., wild boars, *Sus scrofa*: @mortlock_sleep_2024] and likelihood to enter torpor [e.g., Australian eastern long-eared bat, *Nyctophilus bifax*: @fjelldal_nightly_2021].
+At unusually warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behaviour and that of their competitors, predators, and/or prey can often be substantially different from the behaviour at typical temperatures.
+As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @ipcc_climate_2023], mammals will be increasingly forced towards atypical behaviours, which will have the potential to substantially alter community structures and behaviours, both during such events and afterwards [@zhang_increased_2023].
+For instance, changes in climate and phenology impact the life history and behaviour of many hibernating mammals [@wells_life_2022], and hot weather can also affects mammals' sleep quality [e.g., wild boars, *Sus scrofa*: @mortlock_sleep_2024] and likelihood to enter torpor [e.g., Australian eastern long-eared bat, *Nyctophilus bifax*: @fjelldal_nightly_2021].
 Our results show that warmer temperatures cause grizzly bears to more further earlier in the year (Fig.
 \@ref(fig:dist)), which may alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024].
-At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase their ability to find food or be predated upon [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021] as well as the overall ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023].
-These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011], including warmer temperatures and more frequent and intense extreme events [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], such as heat waves, cold spells, floods and droughts, plant and animal die-offs, and sudden growth surges [sometimes referred to as "black swan" events -- see: @logares_black_2012; @anderson_black-swan_2017].
+At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase their chances of finding food or being predated upon [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021] as well as the overall ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023].
+These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011], including warmer temperatures and more frequent and intense extreme events [@yao_emergence_2022; @ipcc_climate_2023], such as heat waves, cold spells, floods and droughts, plant and animal die-offs, and sudden growth surges [sometimes referred to as "black swan" events -- see: @logares_black_2012; @anderson_black-swan_2017].
 
-Our success in respond to current and future changes in climate is contingent on our ability to prepare for and predict change.
-However, predicting animal behavior becomes increasingly complicated as the conditions animals are exposed to deviate from current, typical conditions, especially when responses are nonlinear.
+Our ability to respond to current and future changes in climate is contingent on our ability to prepare for and predict change.
+However, predicting animal behaviour becomes increasingly complicated as the conditions animals are exposed to deviate from current, typical conditions, especially when responses are nonlinear and data are sparse.
 Consequently, we do not present our results as a definitive guide to how mammals in BC will respond to climate change.
-Instead, we hope they serve as a starting point to (1) demonstrate that mammals' movement rates and habitat selection depend on temperature and (2) how one can estimate mammals' changes in movement behavior due to climate change.
+Instead, we hope they serve as a starting point to (1) demonstrate that mammals' movement rates and habitat selection depend on temperature and (2) how one can estimate mammals' changes in movement behaviour due to climate change.
 An important limitation of Figs.
@@
 
-How to represent uncertainty may seem trivial to some, but communicating uncertainty is crucial in assessing risk probabilistically [@aven_implementing_2002; @ayre_bayesian_2012; @kluppelberg_bayesian_2014].
+How to represent uncertainty may seem esoteric, but communicating uncertainty is crucial in assessing risk probabilistically [@aven_implementing_2002; @ayre_bayesian_2012; @kluppelberg_bayesian_2014].
 Recent events in global politics, including global conflicts [@mcnutt_scientists_2022; @tollefson_what_2022] and the rapid rise in generative machine learning [also referred to as artificial intelligence and GenAI, see @roumeliotis_chatgpt_2023; @van_noorden_chatgpt_2023; @deepseek-ai_deepseek-r1_2025] and its prohibitive energy and water demands [@li_making_2023] and carbon footprint [@luccioni_estimating_2024], have placed doubt on the likelihood and feasibility of reaching climate change goals that seemed achievable in the past decade [@united_nations_environment_programme_emissions_2024].
@@
 Similarly, protecting a third of the world's terrestrial and aquatic ecosystems [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022] appears less tangible as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023], and achieving such a goal will require active collaboration with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023].
-Understanding the consequences of climate change on mammals' movement behavior and spatial distribution [@williams_range_2018] is a first step towards anticipating and proactively responding to future changes in human-induced rapid environmental change [@sih_evolution_2011], including climate change.
+Understanding the consequences of climate change on mammals' movement behaviour and spatial distribution [@williams_range_2018] is a first step towards anticipating and proactively responding to future changes in human-induced rapid environmental change [@sih_evolution_2011], including climate change.
 In the following sections, we discuss the implications of our results in more detail.
@@
 
-\noindent The lack of common effects of temperature across all species' movement rates indicates that responding to current conditions and preparing for future changes will require a variety of adaptations as communities adapt and respond in complex and interconnected ways.
-Although our models do not account for any physiological or explicit phenological changes (e.g., in the timing of sleep and hibernation), the tensor product terms in Fig.
+\noindent The lack of common effects of temperature across all species' movement rates indicates that responding to current conditions and preparing for future changes will require a variety of physiological and behavioral adaptations as ecological communities respond in complex and interconnected ways.
+Although our models do not account for any physiological or explicit phenological changes, the tensor product terms in Fig.
 \@ref(fig:dist) suggest that warmer temperatures cause many species to alter their daily and seasonal activity patterns (most visible in pumas and grizzly bears).
@@
 As indicated by grizzly bears' tendency to travel greater distances earlier in the year with increasing temperatures, the increase in intensity and frequency of extreme heat events [@bunnell_global_2011; @yao_emergence_2022] will likely also impact the occurrence and timing of hibernation [@wells_life_2022] and migration or seasonal range expansions [@carbeck_adaptation_2022; @morley_spatiotemporal_2021; @malpeli_ungulate_2022].
-This change will likely depend not only on the direct effect of temperature but also on cascading changes in the availability and unpredictability of resources [@mclellan_effect_2015; @pigeon_drivers_2016; @mezzini_how_2024].
-
-The greater uncertainty around the estimated movement rates at temperatures outside the typical temperatures for a given day (Figs.
-S4, S6-S8) highlights the importance of including appropriate measures of uncertainty when evaluating the consequences of climate change on mammals' movement behavior.
-In this paper, we present our results cognizantly of the lack of uncertainty measures in many of our figures: We hoped to demonstrate how mammalian movement behavior depends on temperature while presenting the need for additional research.
-We expand on the importance of quantifying and communicating uncertainty in the last section of this discussion.
-Our relatively simple models indicate that changes in temperature will alter movement behavior even without accounting for large-scale physiological and behavioral adaptations, which may not be possible at the current rate of climate change [@williams_range_2018; @hetem_responses_2014].
-Still, our models to indicate that some animals adapt their behavior phenology in response to temperature, as indicated by wolves', elk's, caribou's, and grizzly bears' increased movement rates during warmer winters and spring days.
+However, not all species may be able to adaptation at the current rate of climate change [@williams_range_2018; @hetem_responses_2014], especially since lanscape changes will likely depend not only on the direct effect of temperature but also on cascading changes in the availability and unpredictability of resources [@mclellan_effect_2015; @pigeon_drivers_2016; @mezzini_how_2025].
 
@@
 As with the effects of temperature on movement rates, the lack of common effects of temperature across all species indicates that efficient and proactive habitat conservation will require taxonomic nuance.
-In particular, our HRSFs do not account for any explicit physiological responses to temperature that may impact movement such as changes in the phenology of plants, ice, or mammalian physiology or behavior (e.g., seasonal coats, seasonal migration) other than as direct responses to temperature.
-Consequently, the behavioral changes we present should be interpreted carefully.
-While some seasonal cycles do not depend directly on temperature and may thus not be impacted directly by climate change [such as those driven by photoperiod, see @liu_effects_2022], the predictions we present also rely on the assumption that mammals will not adapt these seasonal cycles based on temperature-dependent cues, such as weather [@xu_frost_2019] or forage availability [@middleton_green-wave_2018].
+In particular, our HRSFs did not account for any explicit physiological responses to temperature that may impact movement such as changes in the phenology of plants, ice, or mammalian physiology or behaviour (e.g., moulting, migration) other than as direct responses to temperature.
+Consequently, the behavioural changes we present should be interpreted carefully.
+While some seasonal cycles do not depend directly on temperature and may thus not be impacted directly by climate change [such as those driven by photoperiod, see @liu_effects_2022], the predictions we present also rely on the assumption that mammals will not adapt these seasonal cycles based on temperature-dependent cues, such as weather [@xu_frost_2019] or forage availability [@middleton_green-wave_2018], which may not be correct.
 Additionally, it is important to stress the distinction between habitat selection and quality.
@@
 Additionally, efficient and effective conservation will require collaboration with Indigenous and local communities to ensure Treaties are respected in the process [e.g., @falardeau_biophysical_2022; @lamb_braiding_2023; @wong_towards_2020; @turner_its_2009] while also actively preparing for future change [@desjardins_looking_2020; @hessami_indigenizing_2021].
-The large range in predicted changes by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in figure \@ref(fig:s-hrsfs) highlight the need for careful planning that incorporates not only accurate estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates.
-The width of the prediction intervals in Figure \@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas.
+The large range in predicted changes by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in Fig.
+\@ref(fig:s-hrsfs) highlight the need for careful planning that incorporates not only accurate estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates.
+The width of the prediction intervals in Fig.
+\@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas.
 
-Accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage availability [@mezzini_how_2024], population dynamics [@needs_ref], competitive pressure [e.g., @torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019].
+Accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage availability [@mezzini_how_2025], population dynamics [@needs_ref], competitive pressure [e.g., @torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019].
 This complexity results in the great variation among the responses of different species (and individuals).
@@
 Together, the predicted changes in distance traveled and habitat selection indicate that climate change will have complex consequences for populations as well as cascade effects within ecosystems as species adapt to the changing environments [or fail to do so: @botero_evolutionary_2015; @polazzo_measuring_2024].
-The resulting changes in mammalian movement behavior will have ramifications in individuals' encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@needs_ref], and food webs [@needs_ref] as landscapes change and animals face increasing pressures from human activity [@sih_evolution_2011], including habitat fragmentation [@needs_ref], habitat loss [@needs_ref], and greater human presence in wild spaces [@weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022].
+The resulting changes in mammalian movement behaviour will have ramifications in individuals' encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@needs_ref], and food webs [@needs_ref] as landscapes change and animals face increasing pressures from human activity [@sih_evolution_2011], including habitat fragmentation [@needs_ref], habitat loss [@needs_ref], and greater human presence in wild spaces [@weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022].
+
+***Cite @tucker_moving_2018***
 
@@
 
-Our work highlights three central considerations for future research: (1) telemetry sampling rates should be designed primarily in relation to the movement timescales of the species of interest [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @pease_ecological_2024], (2) the number of individuals tracked and the length of each telemetry should depend not just on population size but also environmental stability, and (3) predicting changes in movement behavior becomes highly complex when responses are nonlinear, especially when changes depend on many interacting factors [@polazzo_measuring_2024] and one is interested in predicting responses in extreme conditions for which data are scarce [@steixner-kumar_strategies_2020].
+Our work highlights three central considerations for future research: (1) telemetry sampling rates should be designed primarily in relation to the movement timescales of the species of interest [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @pease_ecological_2024], (2) the number of individuals tracked and the length of each telemetry should depend not just on population size but also environmental stability, and (3) predicting changes in movement behaviour becomes highly complex when responses are nonlinear, especially when changes depend on many interacting factors [@polazzo_measuring_2024] and one is interested in predicting responses in extreme conditions for which data are scarce [@steixner-kumar_strategies_2020].
 We address each consideration in the following three paragraphs.
 
-Sampling schedules should be fine enough to reconstruct animals' movement at a sufficiently fine scale.
+While the `ctmm` models produced scale-independent estimates of speed [i.e., model interpretation is independent of sampling interval: @noonan_scale-insensitive_2019], the accuracy, size, and stability of speed estimates still depended on the GPS sampling frequency.
+This is because coarsely-sampled movement data contains information on large-scale movements (e.g., range crossings, migrations) but not fine-scale movements [e.g., first-order habitat selection *sensu* @johnson_comparison_1980; also see @nathan_big-data_2022].
+Using the boreal caribou as an example, the 13-hour sampling interval allows us to reasonably estimate the caribou's movement path at a temporal scale of approximately 13 hours (or greater), but we cannot produce reasonable movement trajectories at a much finer (e.g., hourly) scale.
+Consequently, we suggest being cautious when comparing estimated movement behaviours across species, even though all predictions have been corrected to the hourly timescale by predicting for 1-hour time intervals (i.e., `dt_hours = 1`).
+Thus, sampling schedules should be fine enough to reconstruct animals' movement at a sufficiently fine scale.
 Good estimates of an animal's speed and its movement path require telemetry locations to be taken more often than the animal's directional persistence [@noonan_scale-insensitive_2019], so that, on average, the dataset contains multiple locations in between changes in direction.

### 2025-05-07 writing/manuscript.Rmd
fixed tucker ref

@@
 Together, the predicted changes in distance traveled and habitat selection indicate that climate change will have complex consequences for populations as well as cascade effects within ecosystems as species adapt to the changing environments [or fail to do so: @botero_evolutionary_2015; @polazzo_measuring_2024].
-The resulting changes in mammalian movement behaviour will have ramifications in individuals' encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@needs_ref], and food webs [@needs_ref] as landscapes change and animals face increasing pressures from human activity [@sih_evolution_2011], including habitat fragmentation [@needs_ref], habitat loss [@needs_ref], and greater human presence in wild spaces [@weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022].
-
-***Cite @tucker_moving_2018***
+The resulting changes in mammalian movement behaviour will have ramifications in individuals' encounter rates [e.g., @hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@needs_ref], and food webs [@needs_ref] as landscapes change and animals face increasing pressures from human activity [@sih_evolution_2011; @tucker_moving_2018], including habitat fragmentation [@needs_ref], habitat loss [@needs_ref], and greater human presence in wild spaces [@weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022].
 

### 2025-05-08 writing/appendix-b-gams.Rmd
renamed appendix

--- 2025-04-02 writing/appendix.Rmd
@@
 title: \Large Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada
-subtitle: "Appendix"
+subtitle: "Appendix B: Modeling the effects of temperature on movement behavior"
 author:

### 2025-05-08 writing/appendix-a-cleaning-telemetry-data.Rmd
wrote appendix for outlier cleaning (pdf > 100 MB)

---
title: \Large Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada
subtitle: "Appendix A: Cleaning telemetry data"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - bcp: BC Parks
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'freshwater-biology.csl'
# csl: 'nature.csl'
fontsize: 12pt
# indent: true
header-includes:
    - \renewcommand{\figurename}{Figure S\!\!} # for "Figure Ax.
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage{float} # to use fig.pos="H"
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
---

```{r setup, include=FALSE}
# set chunk defaults
knitr::opts_chunk$set(echo = TRUE, out.width = '100%',
                      fig.align = 'center', cache = TRUE, warning = FALSE,
                      message = FALSE,
                      root.dir = '~/GitHub/bc-mammals-temperature')
```

\clearpage

\tableofcontents

\clearpage

# Setup

We start by attaching the necessary packages and sourcing some custom functions. The custom functions are all available on GitHub in the `functions` folder of the repository located at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature). The original cleaning script is located at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R). We suggest looking at the script since this appendix does not include the interactive maps that are generated with `plot_adj(..., map = TRUE)`, since they could not be included in the pdf output easily. However, the interacive maps often show linear features (e.g., roads) and other geographic features (e.g., mountain ridges, rivers) that were useful to see during the data cleaning. Therefore, we have left the function calls in the document for completeness, but we have commented them out to prevent them from being executed. Below is an example of a map with some locations for wolf BW008, as generated by running\newline`plot_adj('BW008', max_speed = 3.5, map = TRUE)`.

\singlespacing

```{r map-example, echo=FALSE, out.height='40%'}
setwd('writing')
knitr::include_graphics(path = '../figures/BW008-map-example.jpg')
```

\clearpage

```{r packages}
library('dplyr')     # for data wrangling (mutate(), %>%, etc.)
library('tidyr')     # for data wrangling (nest(), unnest(), pivot_*, etc.)
library('purrr')     # for functional programming (map_***(), etc.)
library('ctmm')      # for movement models
library('lubridate') # for working with dates
library('mapview')   # for interactive maps
library('ggplot2')   # for fancy plots
theme_set(theme_bw())

# source custom functions
source('functions/outlier_plots.R') # to plot outlier diagnostic plots
source('functions/check_animal.R') # to run diagnostic plots
source('functions/plot_adj.R') # to plot n adjacent locations
source('functions/flag_outlier.R') # to mark outliers
source('functions/remove_outlier_flags.R') # to start over with an animal
```

Before cleaning the telemetry data, we do some basic checks:

```{r checks}
# dataset checks ----
# some NA timestamps (not enough to make a difference)
readRDS('data/tracking-data/all-tracking-data-not-cleaned.rds') %>%
  filter(is.na(timestamp)) %>%
  group_by(dataset_name) %>%
  summarize(n = n())

# ensure all only have one column of HDOP
readRDS('data/tracking-data/all-tracking-data-not-cleaned.rds') %>%
  filter(! is.na(timestamp)) %>%
  mutate(outlier = if_else(outlier, 1, 0)) %>%
  filter(! is.na(hdop) + ! is.na(HDOP) > 1) %>%
  nrow()
```

<!-- to avoid splitting code chunk -->
\clearpage

```{r checks-2}
# should only have one column of DOP
readRDS('data/tracking-data/all-tracking-data-not-cleaned.rds') %>%
  filter(! is.na(timestamp)) %>%
  mutate(outlier = if_else(outlier, 1, 0)) %>%
  filter(! is.na(DOP) + ! is.na(gps.dop) > 1) %>%
  nrow()
```

We can now import the data...

```{r import-data}
# import the full dataset ----
# import the dataset after dropping NAs
d <- readRDS('data/tracking-data/all-tracking-data-not-cleaned.rds') %>%
  filter(! is.na(timestamp)) %>%
  mutate(hdop = if_else(is.na(hdop), HDOP, hdop), # merge error columns
         dop = if_else(is.na(DOP), gps.dop, DOP),
         original_outliers = outlier,
         outlier = if_else(outlier, 1, 0)) %>% # make numeric
  select(! c(HDOP, DOP, gps.dop)) %>% # remove duplicate columns
  relocate(dop, .after = pdop) %>%
  nest(tel = ! c(species, dataset_name, animal))
d # nested tibble of tracking datasets
```

... and do some last few check before cleaning.

```{r final-checks}
# ensure animal IDs are unique
sum(duplicated(d$animal))

# check which species have DOP values
d %>%
  unnest(tel) %>%
  group_by(species) %>%
  summarise(across(c(dop, hdop, pdop), ~ sum(.x, na.rm = TRUE))) %>%
  mutate(has_dop = dop + hdop + pdop > 0) %>%
  arrange(desc(has_dop))

#' see`finite`: some SM caribou have different values for hdop and dop
#' other datasets are ok
d %>%
  unnest(tel) %>%
  mutate(has_hdop = ! is.na(hdop),
         has_dop = ! is.na(dop),
         has_pdop = ! is.na(pdop)) %>%
  group_by(dataset_name, has_hdop, has_dop, has_pdop) %>%
  summarise(hdop = mean(hdop, na.rm = TRUE),
            pdop = mean(pdop, na.rm = TRUE),
            dop = mean(dop, na.rm = TRUE)) %>%
  mutate(finite = map_int(1:n(),
                          ~ sum(! is.nan(c(hdop[.x], pdop[.x], dop[.x])))))

d %>%
  filter(dataset_name == 'Rangifer_tarandus_southern_mountain') %>%
  unnest(tel) %>%
  filter(! is.na(hdop) & ! is.na(dop)) %>%
  plot(hdop ~ dop, ., main = paste('n =', nrow(.)))
abline(a = 0, b = 1, col = 'red', lwd = 2)
```

\clearpage

# *Canis lupus*

```{r canis-lupus, fig.pos='!H', fig.dim=c(10, 12)}
# BW008
out <- check_animal('BW008')
plot_adj('BW008', max_speed = 3.5)
# plot_adj('BW008', max_speed = 3.5, map = TRUE)
plot_adj('BW008', max_angle = 178, max_speed = 1) # using a linear feature
plot_adj('BW008', max_angle = 90, max_speed = 2) # realistic movement
plot_adj('BW008', max_angle = 170, max_speed = 0.5) # fine sampling
out[out$angle > 170 & out$speed > 0.5, ] # 15-minute sampling

# BW009
out <- check_animal('BW009')
plot_adj('BW009', max_angle = 179, max_speed = 1) # potential outlier
plot_adj('BW009', max_angle = 179.7, max_speed = 1, n_adj=200) # uses path
plot_adj('BW009', max_angle = 175, max_speed = 1, n_adj = 12) # ok
plot_adj('BW009', max_angle = 170, min_angle = 175, max_speed = 1,
         n_adj = 20) # ok
plot_adj('BW009', max_angle = 170, max_speed = 0.5) # reasonable

# BW010
out <- check_animal('BW010')
plot_adj('BW010', max_speed = 3.5) # ok
plot_adj('BW010', max_speed = 1, max_angle = 170) # ok
plot_adj('BW010', max_speed = 2, max_angle = 135) # ok

# BW012
out <- check_animal('BW012')
plot_adj('BW012', max_speed = 3)
plot_adj('BW012', max_speed = 1, max_angle = 170)
# location is ok
tel <- as.telemetry(d$tel[[which(d$animal == 'BW012')]])
i <- which(out$speed > 2 & out$angle > 170)
plot(tel, error = FALSE, type = 'l', xlim = c(-2600, 0),
     ylim = c(300, 2200), col = 'black')
plot(tel[i + -10:10, ],
     error = FALSE, type = 'l', col = 'blue', lwd = 2, add = TRUE)
plot(tel[i, ],
     error = FALSE, col = 'blue', pch = 19, add = TRUE)
plot(tel, add = TRUE)
rm(tel, i)

# BW013
out <- check_animal('BW013')
plot_adj('BW013', max_speed = 1, max_angle = 170) # uses linear features

# BW014 has a clear GPS error
out <- check_animal('BW014')
plot_adj('BW014', max_speed = 4)
flag_outlier(id = 'BW014', max_speed = 4, value = 1)
out <- check_animal('BW014')
plot_adj('BW014', max_speed = 2) # ok
plot_adj('BW014', max_speed = 0.5, min_speed = 0.6, max_angle = 170,
         n_adj = 30) # seems near a linear feature
# plot_adj('BW014', max_speed = 0.5, min_speed = 0.6, max_angle = 170,
#          map = TRUE) # moving along a river

# B027
out <- check_animal('BW027')
plot_adj('BW027', max_speed = 3, max_angle = 90) # not problematic

# BW028
out <- check_animal('BW028')
plot_adj('BW028', max_speed = 3) # not problematic

# B029
out <- check_animal('BW029')
plot_adj('BW029', max_speed = 2, max_angle = 175) # need a finer scale
plot_adj('BW029', max_speed = 2, max_angle = 175, n_adj = 2) # ok

# B031
out <- check_animal('BW031')
plot_adj('BW031', max_speed = 1, max_angle = 170) # need a finer scale
plot_adj('BW031', max_speed = 1, max_angle = 170, n_adj = 5) # ok

# B033: sharp and fast turns are near linear features
out <- check_animal('BW033')
plot_adj('BW033', max_speed = 1.5, max_angle = 90, n_adj = 5)

# BW044nex
out <- check_animal('BW044nex')
plot_adj('BW044nex', max_speed = 1, max_angle = 170)

# BW046nex
out <- check_animal('BW046nex')
plot_adj('BW046nex', max_speed = 1, max_angle = 175, min_speed = 3) # ok
plot_adj('BW046nex', max_speed = 1, min_speed = 3, # problematic
         max_angle = 135, min_angle = 175, reset_layout = FALSE)
i <- which(out$speed > 1 & out$speed < 3 & # likely hunting
             out$angle > 135 & out$angle < 175) + 0:5
points(location.lat ~ location.long,
       d$tel[[which(d$animal == 'BW046nex')]][i, ], col = 'blue', cex = 2)
d$tel[[which(d$animal == 'BW046nex')]][i, 'outlier'] <- 1
plot_adj('BW046nex', max_speed = 1, max_angle = 170) # ok
```

# *Cervus canadensis*

```{r cervus-canadensis, fig.pos='!H', fig.dim=c(10, 12)}
#' *NOTE:* assuming movement at ~< 0.25 m/s is realistic, even if angle >
#' 170 degrees since it's too hard to distinguish between movement and gps
#' error at that scale, and the models account for GPS error anyway (which
#' I assume to be 10 m since I don't have calibration data. Mike said most
#' collars have an error ~10 m anyway).

# E002
out <- check_animal('E002')
plot_adj('E002', max_speed = 0.4)
plot_adj('E002', max_angle = 170, max_speed = 0.2)

# E003
out <- check_animal('E003')
plot_adj('E003', max_speed = 0.5)

# E006
out <- check_animal('E006')
plot_adj('E006', max_speed = 0.8) # ok
plot_adj('E006', max_speed = 0.4, max_angle = 170) # outlier
#plot_adj('E006', max_speed = 0.4, max_angle = 170, map = TRUE) # see river
# dt is much larger: collar struggled to receive signal with tree cover
'hours' %#% out[which(out$speed > 0.4 & out$angle > 170) + (-4:4), 'dt']
'hours' %#% c(43282, 7254)
flag_outlier('E006', max_speed = 0.4, max_angle = 170, value = 1)
out <- check_animal('E006')
plot_adj('E006', max_speed = 0.2, max_angle = 170) # ok

# E007
out <- check_animal('E007')
plot_adj('E007', max_speed = 0.2, max_angle = 170) # ok

# E008
out <- check_animal('E008')
plot_adj('E008', max_speed = 0.3) # ok

# E011: last point is outlier (keeping previous points bc many 0 speeds)
out <- check_animal('E011')
plot_adj('E011', max_speed = 0.4, n_adj = 10)
# plot_adj('E011', max_speed = 0.4, n_adj = 10, map = TRUE)
plot_adj('E011', max_speed = 0.4, n_adj = 9, reset_layout = FALSE)
which(out$speed > 0.4) == nrow(out)
i_tel <- which(d$animal == 'E011')
i <- (nrow(d$tel[[i_tel]]) - 7):nrow(d$tel[[i_tel]]) # drop last 8 points
points(location.lat ~ location.long, d$tel[[i_tel]][i, ], col = 'blue',
       cex = 2, pch = 19)
d$tel[[i_tel]][i, 'outlier'] <- 1
points(location.lat ~ location.long, d$tel[[i_tel]][i, ],
       col = as.numeric(d$tel[[i_tel]]$outlier[i]) + 1, cex = 2)
layout(1)

out <- check_animal('E011')
tel <- as.telemetry(d$tel[[i_tel]], mark.rm = TRUE)
plot(tel[nrow(tel) + -10:0, ], error = FALSE, type = 'b',
     pch = as.character(0:9))
plot_adj('E011', max_speed = 0.25)

# E013
out <- check_animal('E013')
plot_adj('E013', max_speed = 0.2, max_angle = 170) # escaped something?

# E015 has multiple GPS malfunctions
out <- check_animal('E015')
plot_adj('E015', max_speed = 1, reset_layout = FALSE)
# can't filter easily because the outliers don't have the max velocity
i <- d %>%
  pull(tel) %>%
  nth(which(d$animal == 'E015')) %>%
  mutate(i = 1:n()) %>%
  slice(which(timestamp ==
                as.POSIXct(out[which(out$speed > 0.6)[2], 't'])) +
          0:1) %>%
  pull(i) %>%
  first() + 0:10
points(location.lat ~ location.long, cex = 2, lwd = 3, col = 'blue',
       filter(d, animal == 'E015')$tel[[1]][i, ])
lines(location.lat ~ location.long,
      filter(d, animal == 'E015')$tel[[1]][i, ], lwd = 6, col = 'red3')
d$tel[[which(d$animal == 'E015')]][i, 'outlier'] <- 1
out <- check_animal('E015')
plot_adj('E015', max_speed = 0.60) # track looks ok now

# E016
out <- check_animal('E016')
plot_adj('E016', max_speed = 0.2, max_angle = 170) # one outlier
plot_adj('E016', max_speed = 0.2, max_angle = 179) # use higher upper bound
flag_outlier('E016', max_speed = 0.2, max_angle = 179, value = 1)
out <- check_animal('E016')
plot_adj('E016', max_speed = 0.2, max_angle = 170) # ok

# E017
out <- check_animal('E017')
plot_adj('E017', max_speed = 0.6) # ok
plot_adj('E017', max_angle = 170, max_speed = 0.2)

# E019
out <- check_animal('E019') # many high speeds at angle > 170
plot_adj('E019', max_speed = 0.65, max_angle = 135) # some outliers
plot_adj('E019', max_speed = 0.65, max_angle = 160, reset_layout = FALSE)
i <- which(out$speed > 0.65 & out$angle > 160)[c(1, 3)] # some outliers
points(location.lat ~ location.long,
       d$tel[[which(d$animal == 'E019')]][i, ], cex = 2, col = 'blue')
d$tel[[which(d$animal == 'E019')]][i, 'outlier'] <- 1

out <- check_animal('E019') # check other high speeds with angle > 160
plot_adj('E019', max_speed = 0.4, max_angle = 135, reset_layout = FALSE)
tel <- d$tel[[which(d$animal == 'E019')]]
i <- which(out$speed > 0.4 & out$angle > 135)
points(location.lat ~ location.long, cex = 1.5, col = 'black',
       tel[! tel$outlier, ][i, ],
       pch = as.character(0:(length(i) - 1)))
'hour' %#% out[i, 'dt'] # consistent sampling

# 0 is likely an error (erring on the side of caution)
# 1 is consistent with other movement
# 2 is likely an error
# 3 is consistent with other movement
# 4 is likely an error
# 5 is realistic
# 6 is likely an error
# 7 is realistic
# 8 is likely an error
# 9 is likely an error
i <- i[1 + c(0, 2, 4, 6, 8, 9)]
i_id <- which(d$animal == 'E019')
d$tel[[i_id]][! d$tel[[i_id]]$outlier, ][i, 'outlier'] <- 1
# ensure correct locations were removed
out <- check_animal('E019') # check other high speeds with angle > 160
plot_adj('E019', max_speed = 0.4, max_angle = 135, reset_layout = FALSE)

# E023
out <- check_animal('E023')
plot_adj('E023', max_speed = 1) # ok
plot_adj('E023', max_speed = 0.6, min_speed = 1) # ok
plot_adj('E023', max_speed = 0.3, max_angle = 170) # ok

# E024
out <- check_animal('E024')
plot_adj('E024', max_speed = 0.4) # ok
plot_adj('E024', max_speed = 0.2, max_angle = 170) # ok

# E025
out <- check_animal('E025')
plot_adj('E025', max_speed = 0.3, max_angle = 170) # ok

# E026
#' this elk does not have an unusual amount of data
hist(map_int(filter(d, species == 'Cervus canadensis')$tel, nrow))
abline(v = nrow(d$tel[[which(d$animal == 'E026')]]), lwd = 2)
# d$tel[[which(d$animal == 'E026')]] %>%
#   filter(! outlier) %>%
#   sf::st_as_sf(coords = c('location.long', 'location.lat'),
#                crs = 4326) %>%
#   mapview(map.types = 'Esri.WorldImagery', col.regions = 'darkorange',
#           col = 'black', lwd = 2)
out <- check_animal('E026') # many sharp turns at high speeds
plot_adj('E026', max_speed = 0.6, max_angle = 135)
#plot_adj('E026', max_speed = 0.6, max_angle = 135, map = TRUE)# look north
flag_outlier('E026', max_speed = 0.6, max_angle = 135, value = 1)

out <- check_animal('E026')
plot_adj('E026', max_speed = 0.4, max_angle = 170) # NW one is ok
# plot_adj('E026', max_speed = 0.4, max_angle = 170, map = TRUE)
plot_adj('E026', max_speed = 0.42, max_angle = 170) # W one may also be ok
plot_adj('E026', max_speed = 0.42, max_angle = 175) # W one may also be ok
flag_outlier('E026', max_speed = 0.42, max_angle = 175, value = 1)

out <- check_animal('E026')
plot_adj('E026', max_speed = 0.35, max_angle = 170) # NW, middle, & NE ok?
plot_adj('E026', max_speed = 0.35, min_speed = 0.4, max_angle = 170,
         reset_layout = FALSE)
i <- with(out, which(speed > 0.35 & speed < 0.4 & angle > 160))[c(3:4)]
points(location.lat ~ location.long,
       filter(d$tel[[which(d$animal == 'E026')]], ! outlier)[i, ],
       cex = 2, col = 'blue', lwd = 2)
#' values in `i` assume outliers have been removed
i_tel <- which(d$animal == 'E026')
d$tel[[i_tel]][d$tel[[i_tel]]$outlier == 0, ][i, 'outlier'] <- 1

out <- check_animal('E026')
plot_adj('E026', max_speed = 0.30, max_angle = 160) # likely ok

#' using an error-informed, continuous-time movement model substantially
#' reduces the issues caused by GPS error. Additionally, there are so many
#' elk data (for each individual and the number of elk, too) that a few
#' values from a single individual will have little to no effect.

# E027
out <- check_animal('E027')
plot_adj('E027', max_speed = 0.3, max_angle = 170, n_adj = 4) # not all
plot_adj('E027', max_speed = 0.3, max_angle = 175, n_adj = 4,
         reset_layout = FALSE)
i <- with(out, which(speed > 0.3 & angle > 175))[c(2, 3, 4, 5)]
points(location.lat ~ location.long,
       d$tel[[which(d$animal == 'E027')]][i, ],
       cex = 2, col = 'blue', lwd = 2)
'hours' %#% out$dt[i]
d$tel[[which(d$animal == 'E027')]][i, 'outlier'] <- 2 # not fully sure

# E030
out <- check_animal('E030')
plot_adj('E030', max_speed = 0.45) # ok
plot_adj('E030', max_speed = 0.3, max_angle = 135) # generally ok
plot_adj('E030', max_speed = 0.3, max_angle = 170) # possibly problematic
flag_outlier('E030', max_speed = 0.3, max_angle = 170, value = 1)

out <- check_animal('E030')
plot_adj('E030', max_speed = 0.2, max_angle = 170) # ok

# E032
out <- check_animal('E032')
plot_adj('E030', max_speed = 0.3, max_angle = 170) # ok

# E033
out <- check_animal(id = 'E033')
plot_adj('E033', max_speed = 0.6)

# E036
out <- check_animal(id = 'E036')
# the next point is along the same line 
plot_adj('E036', max_speed = 0.7, reset_layout = FALSE)
points(location.lat ~ location.long, pch = 19,
       filter(d, animal == 'E036')$tel[[1]][0:1 + which(out$speed > 0.7),])
plot_adj('E036', max_speed = 0.6, reset_layout = FALSE) # ok

# E037
out <- check_animal(id = 'E037')
plot_adj(id = 'E037', max_speed = 0.3, max_angle = 170) # ok

# E040
out <- check_animal(id = 'E040')
plot_adj('E040', max_speed = 0.35, max_angle = 160) # ok

# E042: range shift with highly stationary behavior
out <- check_animal(id = 'E042')
tel <- as.telemetry(d$tel[[which(d$animal == 'E042')]])
range(tel$timestamp)
plot(tel)
# plot_adj(id = 'E042', many = TRUE, map = TRUE, max_speed = 0)
ctmm.guess(tel)

# E049
out <- check_animal(id = 'E049')
plot_adj('E049', max_speed = 0.4, max_angle = 170) # ok
plot_adj('E049', max_speed = 0.3, min_speed = 0.4, max_angle = 170) # ok

# E067
out <- check_animal(id = 'E067')
plot_adj('E067', max_speed = 0.6, max_angle = 170) # ok

# E068
out <- check_animal(id = 'E068')
plot_adj('E068', max_speed = 0.65, reset_layout = FALSE)
out[out$speed > 0.65, ] # 1st 2 points are right after capture (rownames)
points(location.lat ~ location.long,
       filter(d, animal == 'E068')$tel[[1]][1:2, ], col = 'blue', cex = 2)
d[d$animal == 'E068', ]$tel[[1]]$outlier[1:2] <- 1
out <- check_animal(id = 'E068')
plot_adj('E068', max_speed = 0.4, max_angle = 170)
# plot_adj('E068', max_speed = 0.4, max_angle = 170, map = TRUE)
flag_outlier('E068', max_speed = 0.4, max_angle = 170, value = 1)
out <- check_animal(id = 'E068')
plot_adj('E068', max_speed = 0.25, max_angle = 170) # ok

# E070
out <- check_animal('E070')
plot_adj('E070', max_speed = 0.4, max_angle = 135) # ok

# E074
out <- check_animal(id = 'E074')
plot_adj('E074', max_speed = 0.4, max_angle = 170) # ok

# E075
out <- check_animal(id = 'E075')
plot_adj('E075', max_speed = 0.3, max_angle = 170) # ok

# E079
out <- check_animal(id = 'E079')
plot_adj('E079', max_speed = 0.6) # ok

# E080
out <- check_animal(id = 'E080')
plot_adj('E080', max_speed = 0.8) # ok
plot_adj('E080', max_speed = 0.5, max_angle = 170, reset_layout = FALSE)
i_tel <- which(d$animal == 'E080')
i <- which(out$speed > 0.5 & out$angle > 170)[1:2]
points(location.lat ~ location.long,
       filter(d$tel[[i_tel]], ! outlier)[i, ], col = 'blue',
       cex = 2)
d$tel[[i_tel]][! d$tel[[i_tel]]$outlier, ]$outlier[i] <- 1
out <- check_animal(id = 'E080')
plot_adj('E080', max_speed = 0.5, max_angle = 170) # ok

# E081
out <- check_animal(id = 'E081')
plot_adj('E081', max_speed = 0.2, max_angle = 170) # ok

# E082
out <- check_animal(id = 'E082')
plot_adj('E082', max_speed = 0.2, max_angle = 170) # ok

# E083
out <- check_animal(id = 'E083')
plot_adj('E083', max_speed = 0.3, max_angle = 135) # ok

# E084 is not worth keeping
out <- check_animal(id = 'E084')
plot(location.lat ~ location.long, filter(d, animal == 'E084')$tel[[1]],
     type = 'b', pch = 19, col = '#00000010')
nrow(filter(d, animal == 'E084')$tel[[1]])
range(filter(d, animal == 'E084')$tel[[1]]$timestamp)
d <- filter(d, animal != 'E084')

# E085
out <- check_animal(id = 'E085')
plot_adj('E085', max_speed = 1) # definitely outliers (> 20 km!)
flag_outlier(id = 'E085', max_speed = 1, value = 1)

out <- check_animal(id = 'E085')
plot_adj('E085', max_speed = 0.6) # maybe ok
with(out[which(out$speed > 0.6) + -10:10, ], # likely outlier
     plot('hours' %#% dt, dt * speed / 1e3,
          col = as.numeric(-10:10 == 0) + 1,
          pch = 19, ylab = 'SLD (dt * min speed, km)'))
flag_outlier('E085', max_speed = 0.6, value = 1)

out <- check_animal(id = 'E085')
plot_adj('E085', max_speed = 0.2, max_angle = 135) # ok

# E096
out <- check_animal(id = 'E096')
plot_adj('E096', max_speed = 0.4, max_angle = 170) # ok

# E100
out <- check_animal(id = 'E100')
plot_adj('E100', max_speed = 0.3, max_angle = 170) # ok

# E101
out <- check_animal(id = 'E101')
plot_adj('E101', max_speed = 0.4, max_angle = 170) # ok

# E106
d$tel[[which(d$animal == 'E106')]] %>%
  group_by(gps.fix.type.raw) %>%
  summarise(hdop = mean(dop, na.rm = TRUE))
d$tel[[which(d$animal == 'E106')]] %>%
  group_by(gps.fix.type.raw) %>%
  summarise(hdop = mean(! is.na(dop), na.rm = TRUE))
out <- check_animal(id = 'E106')
plot_adj('E106', max_speed = 0.4, max_angle = 170) # large DOP values; ok
d$tel[[which(d$animal == 'E106')]] %>%
  as.telemetry() %>%
  data.frame() %>%
  slice(which(out$speed > 0.4 & out$angle > 170)) %>%
  pull(HDOP)
plot_adj('E106', max_speed = 0.3, max_angle = 170) # ok
plot_adj('E106', max_speed = 0.2, max_angle = 170) # ok

# E115
# collar is very error-prone in SW forest
out <- check_animal(id = 'E115')
# plot_adj('E115', many = TRUE, map = TRUE)
plot_adj('E115', max_speed = 0.4, max_angle = 170) # improbable
plot_adj('E115', max_speed = 0.4, max_angle = 170, n_adj = 50)#but possible
plot_adj('E115', max_speed = 0.4, max_angle = 175) # unlikely
slice(d$tel[[which(d$animal == 'E115')]],
      which(out$speed > 0.4 & out$angle > 175))$dop # low DOP
flag_outlier(id = 'E115', max_speed = 0.4, max_angle = 175, value = 2)

out <- check_animal('E115')
plot_adj('E115', max_speed = 0.3, max_angle = 170) # unlikely
flag_outlier('E115', max_speed = 0.3, max_angle = 170, value = 2)

out <- check_animal('E115')
plot_adj('E115', max_speed = 0.1, max_angle = 170, many = TRUE) # ok

# E116
out <- check_animal(id = 'E116')
plot_adj('E116', max_speed = 0.4, max_angle = 170)
flag_outlier(id = 'E116', max_speed = 0.4, max_angle = 170, value = 1)

out <- check_animal('E116')
plot_adj('E116', max_speed = 0.8)
flag_outlier(id = 'E116', max_speed = 0.8, value = 1)

out <- check_animal('E116')
plot_adj('E116', max_speed = 0.3, max_angle = 170) # improbable
plot_adj('E116', max_speed = 0.3, max_angle = 175) # but not impossible

# E117 has similar error patterns to E115
out <- check_animal(id = 'E117')
plot_adj('E117', max_speed = 0.7) # three outliers
# plot_adj('E117', max_speed = 0.7, map = TRUE)
flag_outlier(id = 'E117', max_speed = 0.7, value = 1)

out <- check_animal('E117')
#plot_adj('E117', many = TRUE, map = TRUE) # more errors when more forested
plot_adj('E117', max_speed = 0.4, max_angle = 170)

# plot_adj('E117', max_speed = 0.4, max_angle = 170, map = TRUE)
flag_outlier(id = 'E117', max_speed = 0.4, max_angle = 170, value = 1)

out <- check_animal('E117')
plot_adj('E117', max_speed = 0.3, max_angle = 170) # maybe not all
plot_adj('E117', max_speed = 0.3, max_angle = 175) # likely all
#plot_adj('E117', max_speed = 0.3, max_angle = 175, map = TRUE) # in forest
flag_outlier(id = 'E117', max_speed = 0.3, max_angle = 175, value = 1)

out <- check_animal('E117')
# still an excessive amount of sharp turns with high speeds (compare E118)
# dropping some of the data is better than dropping the whole animal
flag_outlier('E117', max_speed = 0.1, max_angle = 130, value = 2)
out <- check_animal('E117')

# E118 (compare with E117 above)
check_animal('E118', return_out = FALSE)

# E122
out <- check_animal(id = 'E122')
plot_adj('E122', max_speed = 0.45) # ok

# E125
out <- check_animal(id = 'E125')
plot_adj('E125', max_speed = 0.4, max_angle = 150)
flag_outlier('E125', max_speed = 0.4, max_angle = 150, value = 1)
out <- check_animal(id = 'E125') # ok

# E129
out <- check_animal(id = 'E129')
plot_adj('E129', max_speed = 0.6) # ok

# E131
out <- check_animal(id = 'E131')
plot_adj('E131', max_speed = 0.3, max_angle = 170) # N one is ok
plot_adj('E131', max_speed = 0.45, max_angle = 170) # outlier
flag_outlier('E131', max_speed = 0.45, max_angle = 170, value = 1)

out <- check_animal(id = 'E131') # ok

# E132
#' some high minimum speeds at long sampling intervals for this individual
out <- check_animal(id = 'E132', cap_dt = FALSE)
layout(matrix(c(0, 0, 0, 1), ncol = 2)) #
abline(v = 20, col = 'red')
abline(h = 0.08, col = 'red')
i <- which(out$dt > 20 %#% 'hours' & out$speed > 0.08)
out[i, ]
# small range shift at the daily scale within home range
plot_adj('E132', max_speed = 0.08, max_dt = 20 %#% 'hours')

# look at points near values with excessively high speeds
i_adj <- sapply(i, function(.i) .i + -10:10) %>%
  as.data.frame() %>% # convert matrix to a data frame
  pivot_longer(! c()) # use long format for a column of all values
tel <- as.telemetry(d$tel[[which(d$animal == 'E132')]])
adj <- left_join(slice(data.frame(out), i_adj$value),
                 slice(data.frame(tel), unique(i_adj$value)),
                 by = c('t')) %>%
  bind_cols(i_adj, .) %>%
  mutate(date = as.POSIXct(t),
         high_speed = speed > 0.08)
layout(matrix(1:4, ncol = 2))
for(g in seq(unique(i_adj$name))) {
  g_name <- unique(i_adj$name)[g]
  plot(y ~ x, filter(adj, name == g_name), type = 'l')
  points(y ~ x, data.frame(tel[i, ][g, ]), col = 'blue', pch = 19)
  rm(g, g_name)
} # ok
layout(1)

# large dt with minor "range shifts" ==> ok
plot('days' %#% dt ~ as.POSIXct(t), out[i_adj$value, ])
points('days' %#% dt ~ as.POSIXct(t), out[i, ], col = 'blue', pch = 19)

rm(i, i_adj, tel, adj)

# E134
out <- check_animal(id = 'E134')
plot_adj('E134', max_speed = 0.5, max_angle = 170) # ok

# E138
out <- check_animal(id = 'E138')
plot_adj('E138', max_speed = 0.8) # ok
plot_adj('E138', max_speed = 0.3, max_angle = 170) # ok

# E139
out <- check_animal(id = 'E139')
plot_adj('E139', max_speed = 0.3, max_angle = 170) # ok

# E140
out <- check_animal(id = 'E140')
plot_adj('E140', max_speed = 0.6)
flag_outlier('E140', max_speed = 0.6, value = 1)

out <- check_animal(id = 'E140') # ok

# E142
out <- check_animal(id = 'E142')
flag_outlier('E142', max_speed = 10, value = 1) # definitely an outlier

out <- check_animal(id = 'E142')
plot_adj('E142', max_speed = 0.8) # ok

# E143
out <- check_animal(id = 'E143')
plot_adj('E143', max_speed = 0.2, max_angle = 170) # ok

# E144
out <- check_animal(id = 'E144')
plot_adj('E144', max_speed = 0.3, max_angle = 170) # all ok

# E147
out <- check_animal(id = 'E147')
plot_adj('E147', max_speed = 0.3, max_angle = 170) # ok

# E161
out <- check_animal(id = 'E161')
plot_adj('E161', max_speed = 0.3, max_angle = 170) # ok

# E164: first point is definitely an outlier
out <- check_animal(id = 'E164')
tel <- d$tel[[which(d$animal == 'E164')]]
plot(location.lat ~ location.long, tel, type = 'b')
points(location.lat ~ location.long, tel[1, ], col = 'red', pch = 19)
rm(tel)
d$tel[[which(d$animal == 'E164')]]$outlier[1] <- 1

out <- check_animal(id = 'E164')
plot_adj('E164', max_speed = 0.3, max_angle = 160) # ok
rm(tel)

# E168
out <- check_animal(id = 'E168')
plot_adj('E168', max_speed = 0.3, max_angle = 170)
flag_outlier('E168', max_speed = 0.3, max_angle = 170, value = 1)

out <- check_animal(id = 'E168')
plot_adj('E168', max_speed = 0.2, max_angle = 170) # ok

# E170
out <- check_animal(id = 'E170')
plot_adj('E170', max_speed = 1)
plot_adj('E170', max_speed = 0.3, max_angle = 170) # ok

# E171
out <- check_animal(id = 'E171')
plot_adj('E171', max_speed = 0.6) # ok
plot_adj('E171', max_speed = 0.3, max_angle = 170) # ok

# E172
out <- check_animal(id = 'E172')
plot_adj('E172', max_speed = 0.3, max_angle = 150) # ok

# E178
out <- check_animal(id = 'E178')
plot_adj('E178', max_speed = 0.3, max_angle = 150) # ok

# E179
out <- check_animal(id = 'E179')
plot_adj('E179', max_speed = 0.8) # ok

# E181
out <- check_animal(id = 'E181')
plot_adj('E181', max_speed = 0.5, max_angle = 170) # ok

# E181
out <- check_animal(id = 'E182')
plot_adj('E182', max_speed = 0.3, max_angle = 170) # ok
```

# *Oreamnos americanus*

```{r oreamnos-americanus, fig.pos='!H', fig.dim=c(10, 12)}
d %>%
  filter(species == 'Oreamnos americanus') %>%
  unnest(tel) %>%
  pull(hdop) %>%
  hist(main = 'Goat HDOP before as.telemetry()')

d %>%
  filter(species == 'Oreamnos americanus') %>%
  unnest(tel) %>%
  pull(pdop) %>%
  hist(main = 'Goat PDOP before as.telemetry()')

d %>%
  filter(species == 'Oreamnos americanus') %>%
  mutate(tel = map(tel, \(.t) data.frame(as.telemetry(.t)))) %>%
  unnest(tel) %>%
  pull(HDOP) %>%
  hist(main = 'Goat HDOP after as.telemetry()')

# pdop is not in the telemetries
d %>%
  filter(species == 'Oreamnos americanus') %>%
  mutate(tel = map(tel, \(.t) data.frame(as.telemetry(.t)))) %>%
  unnest(tel) %>%
  colnames()

# substitute missing HDOP values with PDOP values
d <- d %>%
  mutate(
    tel = if_else(species == 'Oreamnos americanus',
                  map(tel, \(.t) {
                    .t %>%
                      mutate(hdop = if_else(is.na(hdop), pdop, hdop)) %>%
                      return()
                  }),
                  tel))

# ensure values are ok
all((d %>%
       filter(species == 'Oreamnos americanus') %>%
       unnest(tel) %>%
       pull(hdop)) ==
      (d %>%
         filter(species == 'Oreamnos americanus') %>%
         mutate(tel = map(tel, \(.t) data.frame(as.telemetry(.t)))) %>%
         unnest(tel) %>%
         pull(HDOP)))

# all goats have some extreme outliers in Berlin
filter(d, species == 'Oreamnos americanus') %>%
  unnest(tel) %>%
  ggplot() +
  facet_wrap(~ animal) +
  geom_hex(aes(location.long, location.lat), bins = 10) +
  geom_vline(xintercept = 0, color = 'darkorange') +
  scale_fill_continuous(limits = c(1, NA), trans = 'log10') +
  theme(legend.position = c(0.8, 0.15))

filter(d, species == 'Oreamnos americanus') %>%
  unnest(tel) %>%
  filter(location.long < 0) %>%
  pull(location.long) %>%
  hist()
abline(v = -119.8, col = 'darkorange')

# set all locations with long > 0 to outliers
d <-
  mutate(
    d,
    tel = if_else(species == 'Oreamnos americanus',
                  true =
                    map(tel, \(.d) {
                      mutate(
                        .d,
                        outlier = if_else(location.long > -119.8, 1, 0),
                        original_outliers = outlier == 1)
                    }),
                  false = tel))

# 30548
out <- check_animal(id = '30548')
hist(out$angle) # a lot of sharp turns at low speeds
plot_adj('30548', max_speed = 0.4) # ok
plot_adj('30548', max_speed = 0.3, max_angle = 135) # ok
# plot_adj('30548', max_speed = 0.3, max_angle = 135, map = TRUE) # ridges
plot_adj('30548', max_speed = 0.2, max_angle = 135) # ok
plot_adj('30548', max_speed = 0.1, max_angle = 170, n_adj = 3) # ok

# 30551
out <- check_animal(id = '30551') # another 124, many with speed = 0
# start and end are definitely not part of the goat's movement
d$tel[[which(d$animal == '30551')]] %>%
  filter(! outlier) %>%
  ggplot(aes(location.long, location.lat, col = as.POSIXct(timestamp))) +
  geom_path() +
  geom_point() +
  scale_color_viridis_c()
flag_outlier('30551', max_distance = 2e4, value = 1)

out <- check_animal(id = '30551') # last point is still a clear outlier
plot_adj('30551', max_speed = 0.6)
which(out$speed > 0.6) == nrow(out)
flag_outlier('30551', max_speed = 0.6, value = 1)

out <- check_animal(id = '30551') # all other speeds are reasonable
plot_adj('30551', max_speed = 0.4, n_adj = 20) # ok

# 30561
out <- check_animal(id = '30561')
# plot_adj('30561', max_distance = 15e3, map = TRUE)
flag_outlier('30561', max_distance = 15e3, value = 1)

out <- check_animal(id = '30561') # some high speeds at sharp turns
plot_adj('30561', max_speed = 0.2, max_angle = 160)
#plot_adj('30561', max_speed = 0.2, max_angle = 160, map = TRUE) # outliers
flag_outlier('30561', max_speed = 0.2, max_angle = 160, value = 1)

out <- check_animal(id = '30561') # speeds are reasonable
# plot_adj('30561', max_speed = 0, map = TRUE, many = TRUE) # ok

# 30567
out <- check_animal(id = '30567')
plot_adj('30567', max_speed = 0.3, max_angle = 160)
flag_outlier('30567', max_speed = 0.3, max_angle = 160, value = 1)

out <- check_animal(id = '30567')
plot_adj('30567', max_speed = 0.3) # ok
plot_adj('30567', max_speed = 0.2, max_angle = 160)
flag_outlier('30567', max_speed = 0.2, max_angle = 160, value = 1)

out <- check_animal(id = '30567')
plot_adj('30567', max_speed = 0.15, max_angle = 160) # NE is ok
plot_adj('30567', max_speed = 0.15, max_angle = 170)
plot_adj('30567', max_speed = 0.15, max_angle = 170, n_adj = 3)
#plot_adj('30567', max_speed = 0.15, max_angle = 170, n_adj = 3, map = TRUE)
flag_outlier('30567', max_speed = 0.15, max_angle = 170, value = 1)

out <- check_animal(id = '30567') # some possible outliers, but unclear
# plot_adj('30567', max_speed = 0, many = TRUE, map = TRUE)
plot_adj('30567', max_speed = 0.1, max_angle = 170) # ok
out[out$speed > 0.1 & out$angle > 170, ]
plot_adj('30567', max_speed = 0.1, max_angle = 170, max_dt = 2e4) # ok

# 30575
out <- check_animal(id = '30575')
plot_adj('30575', max_speed = 0.8)
flag_outlier('30575', max_speed = 0.8, value = 1)

out <- check_animal(id = '30575')
plot_adj('30575', max_speed = 0.5)
flag_outlier('30575', max_speed = 0.5, value = 1)

out <- check_animal(id = '30575')
plot_adj('30575', max_speed = 0.3, max_angle = 170)
flag_outlier('30575', max_speed = 0.3, max_angle = 170, value = 1)

out <- check_animal(id = '30575')
plot_adj('30575', max_speed = 0.2, max_angle = 170)
flag_outlier('30575', max_speed = 0.2, max_angle = 170, value = 1)

out <- check_animal(id = '30575')
plot_adj('30575', max_speed = 0.175, max_angle = 170)
flag_outlier('30575', max_speed = 0.175, max_angle = 170, value = 1)

out <- check_animal(id = '30575')
plot_adj('30575', max_speed = 0.15, max_angle = 170)
# plot_adj('30575', max_speed = 0.15, max_angle = 170, map = TRUE)
flag_outlier('30575', max_speed = 0.15, max_angle = 170, value = 1)

out <- check_animal(id = '30575')
plot_adj('30575', max_speed = 0.12, max_angle = 170)
plot_adj('30575', max_speed = 0.12, max_angle = 170, max_dt = 2e4)
flag_outlier('30575', max_speed = 0.12, max_angle = 170, max_dt = 2e4,
             value = 1)

out <- check_animal(id = '30575') # some points in forest, but speed is low
# plot_adj('30575', many = TRUE, map = TRUE)
plot(as.telemetry(d$tel[[which(d$animal == '30575')]], mark.rm = TRUE))

# 30599
out <- check_animal(id = '30599')
plot_adj(id = '30599', max_speed = 0.3, max_angle = 170)
flag_outlier(id = '30599', max_speed = 0.3, max_angle = 170, value = 1)

out <- check_animal(id = '30599')
plot_adj(id = '30599', max_speed = 0.2, max_angle = 170)
flag_outlier(id = '30599', max_speed = 0.2, max_angle = 170, value = 1)

out <- check_animal(id = '30599')
plot_adj(id = '30599', max_speed = 0.3)
flag_outlier(id = '30599', max_speed = 0.3, value = 1)

out <- check_animal(id = '30599')
plot_adj(id = '30599', max_speed = 0.15, max_angle = 170)
flag_outlier(id = '30599', max_speed = 0.15, max_angle = 170, value = 1)

out <- check_animal(id = '30599')
plot_adj(id = '30599', max_speed = 0.1, max_angle = 175)
out[out$angle > 175 & out$speed > 0.1, ]
plot_adj(id = '30599', max_speed = 0.1, max_angle = 175, min_dt = 2e4)
i <- which(out$speed > 0.1 & out$angle > 175 & out$dt < 2e4)
tel <- d$tel[[which(d$animal == '30599')]]
d$tel[[which(d$animal == '30599')]][! tel$outlier, ][i, 'outlier'] <- 1

out <- check_animal(id = '30599') # ok

# 30613
out <- check_animal(id = '30613')
plot_adj('30613', max_speed = 0.4, max_angle = 170)
# plot_adj('30613', max_speed = 0.4, max_angle = 170, map = TRUE)
flag_outlier('30613', max_speed = 0.4, max_angle = 170, value = 1)

out <- check_animal(id = '30613')
plot_adj('30613', max_speed = 0.2, max_angle = 170)
flag_outlier('30613', max_speed = 0.2, max_angle = 170, value = 1)

out <- check_animal(id = '30613')
plot_adj('30613', max_speed = 0.2, max_angle = 135)
# plot_adj('30613', max_speed = 0.2, max_angle = 135, map = TRUE)
flag_outlier('30613', max_speed = 0.2, max_angle = 135, value = 1)

out <- check_animal(id = '30613')
plot_adj('30613', max_speed = 0.12, max_angle = 170)
flag_outlier('30613', max_speed = 0.12, max_angle = 170, value = 1)

out <- check_animal(id = '30613')
plot_adj('30613', max_speed = 0.1, max_angle = 170) # ok
plot_adj('30613', max_speed = 0.1, max_angle = 150) # ok
plot_adj('30613', max_speed = 0.05, max_angle = 170) # ok

# 30636
out <- check_animal(id = '30636')
flag_outlier('30636', max_distance = 15e3, value = 1)

out <- check_animal(id = '30636')
plot_adj('30636', max_speed = 0.4)
flag_outlier('30636', max_speed = 0.4, value = 1)

out <- check_animal(id = '30636')
plot_adj('30636', max_speed = 0.1, max_angle = 170) # ok

# 30642
out <- check_animal(id = '30642')
plot_adj('30642', max_speed = 1)
flag_outlier('30642', max_speed = 1, value = 1)

out <- check_animal(id = '30642')
plot_adj('30642', max_speed = 0.3, max_angle = 170)
out[out$speed > 0.3 & out$angle > 170, ]
plot_adj('30642', max_speed = 0.3, max_angle = 170, max_dt = 8100)
flag_outlier('30642', max_speed = 0.3, max_angle = 170, max_dt = 8100,
             value = 1)

out <- check_animal(id = '30642')
plot_adj('30642', max_speed = 0.2, max_angle = 170, min_speed = 0.3)
i_tel <- which(d$animal == '30642')
i <- which(out$speed > 0.2 & out$speed < 0.3 & out$angle > 170)
d$tel[[i_tel]][! d$tel[[i_tel]]$outlier, ][i, 'outlier'] <- 1
plot_adj('30642', max_speed = 0.2, max_angle = 170, min_speed = 0.3) # ok

out <- check_animal(id = '30642') # ok

# 30648
out <- check_animal(id = '30648')
flag_outlier('30648', max_speed = 1, value = 1)

out <- check_animal(id = '30648')
plot_adj('30648', max_speed = 0.3, max_angle = 170)
flag_outlier('30648', max_speed = 0.3, max_angle = 170, value = 1)

out <- check_animal(id = '30648')
plot_adj('30648', max_speed = 0.2, max_angle = 170) # ok
plot_adj('30648', max_speed = 0.4, max_angle = 130)
flag_outlier('30648', max_speed = 0.4, max_angle = 130, value = 1)

out <- check_animal(id = '30648')
plot_adj('30648', max_speed = 0.2, max_angle = 170)
# plot_adj('30648', max_speed = 0.2, max_angle = 170, map = TRUE)
plot_adj('30648', max_speed = 0.2, max_angle = 175)
flag_outlier('30648', max_speed = 0.2, max_angle = 175, value = 1)

out <- check_animal(id = '30648')
plot_adj('30648', max_speed = 0.2, max_angle = 170) # ok
plot_adj('30648', max_speed = 0.1, max_angle = 175) # ok
```

# *Puma concolor*

```{r puma-concolor, fig.pos='!H', fig.dim=c(10, 12)}
# 246
out <- check_animal(id = '246')
plot_adj('246', max_speed = 3)
flag_outlier('246', max_speed = 3, value = 1)

out <- check_animal(id = '246')
plot_adj('246', max_speed = 0.5, n_adj = 3) # ok
plot_adj('246', max_speed = 0.2, max_angle = 170) # ok
# plot_adj('246', max_speed = 0.2, max_angle = 170, n_adj = 4, map = TRUE)
# most southern point may be an outlier, but it looks ok
plot_adj('246', max_speed = 0.2, max_angle = 177, min_angle = 178) # ok
# plot_adj('246', max_speed = 0.2, max_angle = 177, min_angle = 178,
#          n_adj = 4, map = TRUE)
i <- which(out$speed > 0.2 & out$angle > 177 & out$angle < 178) + -5:5
tel <- d$tel[[which(d$animal == '246')]]
tel <- tel[! tel$outlier, ]
plot(location.lat ~ location.long, tel[i, ], type = 'b',
     pch = as.character(seq(length(i))))
'hours' %#% out[i, 'dt']
rm(i, tel, cols)

# 272 has a clear speed outlier, but locations are ok
out <- check_animal(id = '272')
plot_adj('272', max_speed = 1) # data seem ok
plot_adj('272', max_speed = 0.2, max_angle = 135) # ok

# C1
out <- check_animal(id = 'C1')
plot_adj('C1', max_speed = 0.2, max_angle = 170) # ok

# C2
out <- check_animal(id = 'C2')
plot_adj('C2', max_speed = 0.2, max_angle = 170, n_adj = 20) # gps error
# plot_adj('C2', max_speed = 0.2, max_angle = 170, n_adj = 20, map = TRUE)
flag_outlier('C2', max_speed = 0.2, max_angle = 170, value = 1)

# C3
out <- check_animal(id = 'C3')
plot_adj('C3', max_speed = 1)
# plot_adj('C3', max_speed = 1, map = TRUE)
flag_outlier('C3', max_speed = 1, value = 1)
out <- check_animal(id = 'C3') # remaining data are ok
plot_adj('C3', max_speed = 0.4, max_angle = 170, n_adj = 5) # ok
plot_adj('C3', max_speed = 0.3, max_angle = 170, n_adj = 5) # ok

# C4
out <- check_animal(id = 'C4')
plot_adj('C4', max_speed = 0.64)
# plot_adj('C4', max_speed = 0.64, map = TRUE)
flag_outlier('C4', max_speed = 0.64, value = 1)
out <- check_animal(id = 'C4') # remaining data are ok
plot_adj('C4', max_speed = 0.2, max_angle = 170) # ok

# C5
out <- check_animal(id = 'C5')
plot_adj('C5', max_speed = 0.6) # ok
plot_adj('C5', max_speed = 0.4, max_angle = 170) # ok
plot_adj('C5', max_speed = 0.3, max_angle = 170) # ok

# C6
out <- check_animal(id = 'C6')
plot_adj('C6', max_speed = 0.4, max_angle = 170)
flag_outlier('C6', max_speed = 0.4, max_angle = 170, value = 1)
out <- check_animal(id = 'C6') # ok

# C7
out <- check_animal(id = 'C7')
plot_adj('C7', max_speed = 0.6, max_angle = 135)
flag_outlier('C7', max_speed = 0.6, max_angle = 135, value = 1)
out <- check_animal(id = 'C7') # ok
plot_adj('C7', max_speed = 0.4, max_angle = 135) # ok

# C8
out <- check_animal(id = 'C8')
plot_adj('C8', max_speed = 0.55) # ok
# plot_adj('C8', max_speed = 0.55, map = TRUE)
'hours' %#% out[which(out$speed > 0.55) + -5:5, 'dt']
plot_adj('C8', max_speed = 0.2, max_angle = 170) # ok

# C9
out <- check_animal(id = 'C10')
plot_adj('C10', max_speed = 0.2, max_angle = 170) # ok

# C10
out <- check_animal(id = 'C9')
plot_adj('C9', max_speed = 0.2, max_angle = 170) # ok

# C11
out <- check_animal(id = 'C11')
plot_adj('C11', max_speed = 2)
flag_outlier('C11', max_speed = 2, value = 1)

out <- check_animal(id = 'C11')
plot_adj('C11', max_speed = 0.4, max_angle = 170) # NE ok
# plot_adj('C11', max_speed = 0.4, max_angle = 175, map = TRUE) # outlier
'hours' %#% out[which(out$speed > 0.4 & out$angle > 175) + -3:3, 'dt']
flag_outlier('C11', max_speed = 0.4, max_angle = 175, value = 1)

out <- check_animal(id = 'C11')
plot_adj('C11', max_speed = 0.3, max_angle = 170) # NE ok

# C12
out <- check_animal(id = 'C12')
plot_adj('C12', max_speed = 0.8)
i <- which(out$speed > 0.8) + c(-1:0)
layout(t(0:1))
points(location.lat ~ location.long, d$tel[[which(d$animal == 'C12')]][i,],
       cex = 2, col = 'blue')
out$dt[i] # distance seems excessive for 2-hour sampling
d$tel[[which(d$animal == 'C12')]][i, 'outlier'] <- 2
out <- check_animal(id = 'C12')
plot_adj(id = 'C12', max_speed = 0.2, max_angle = 170) # ok
rm(i)

# c13
out <- check_animal(id = 'C13')
plot_adj('C13', max_speed = 0.3, max_angle = 170) # ok

# C14
out <- check_animal('C14')
plot_adj('C14', max_speed = 0.4, max_angle = 160) # ok

# C16
out <- check_animal('C16')
plot_adj('C16', max_speed = 0.3, max_angle = 170) # all outliers
flag_outlier('C16', max_speed = 0.3, max_angle = 170, value = 1)
out <- check_animal('C16') # ok

# C17
out <- check_animal('C17')
plot_adj('C17', max_speed = 0.2, max_angle = 170)
plot_adj('C17', max_speed = 0.35, max_angle = 170)
# one of the two is the last point of the telemetry
out[which((out$angle > 170 | is.na(out$angle)) & out$speed > 0.35), ]
i <- which(out$angle > 170 & out$speed > 0.35)
out[i, ]
# don't want to flag last point
d$tel[[which(d$animal == 'C17')]]$outlier[i] <- 1
out <- check_animal('C17')
plot_adj('C17', max_speed = 0.35, max_angle = 179) # ok

# C18
out <- check_animal('C18')
plot_adj('C18', max_speed = 0.6) # the two speeds are too similar
flag_outlier('C18', max_speed = 0.6, value = 2)
out <- check_animal('C18')
plot_adj('C18', max_speed = 0.2, max_angle = 170) # ok

# C19
out <- check_animal('C19')
plot_adj('C19', max_angle = 160, max_speed = 0.4) # NW is ok
plot_adj('C19', max_speed = 0.4, max_angle = 170)
flag_outlier('C19', max_speed = 0.4, max_angle = 170, value = 1)

out <- check_animal('C19') # ok

# C23
out <- check_animal('C23')
plot_adj('C23', max_speed = 0.3, max_angle = 160) # S is ok
plot_adj('C23', max_speed = 0.3, max_angle = 170) # unsure
plot_adj('C23', max_speed = 0.3, max_angle = 170, n_adj = 3)
out[which(out$angle > 170 & out$speed > 0.3) + -3:3, 'dt'] # too regular
# plot_adj('C23', max_speed = 0.3, max_angle = 170, map = TRUE) # outlier
flag_outlier('C23', max_speed = 0.3, max_angle = 170, value = 1)
out <- check_animal('C23')

# C28
out <- check_animal('C28')
plot_adj('C28', max_speed = 0.3, max_angle = 170)
flag_outlier('C28', max_speed = 0.3, max_angle = 170, value = 1)
out <- check_animal('C28')
plot_adj('C28', max_speed = 0.25, max_angle = 170) # ok

# C29
out <- check_animal('C29')
plot_adj('C29', max_speed = 0.5, max_angle = 160)
flag_outlier('C29', max_speed = 0.5, max_angle = 160, value = 1)
out <- check_animal('C29')

# C30
out <- check_animal('C30')
plot_adj('C30', max_speed = 0.6)
flag_outlier('C30', max_speed = 0.6, value = 1)

out <- check_animal('C30')
plot_adj('C30', max_speed = 0.3, max_angle = 170)
flag_outlier('C30', max_speed = 0.3, max_angle = 170, value = 1)

out <- check_animal('C30')
plot_adj('C30', max_speed = 0.3, max_angle = 135) # ok

# C31
out <- check_animal('C31')
plot_adj('C31', max_speed = 0.37, max_angle = 160)
out[which(out$speed > 0.37 & out$angle > 160) + -3:3, 'dt'] # too regular
# plot_adj('C31', max_speed = 0.37, max_angle = 160, map = TRUE) # outlier
flag_outlier('C31', max_speed = 0.37, max_angle = 160, value = 1)
out <- check_animal('C31') # ok
```

# *Rangifer tarandus caribou* (boreal)

```{r boreal-caribou, fig.pos='!H', fig.dim=c(10, 12)}
# HSCEK063
out <- check_animal('HSCEK063')
plot_adj('HSCEK063', max_speed = 0.6) # ok
plot_adj('HSCEK063', max_speed = 0.2, max_angle = 135) # ok

# SCEK007b
out <- check_animal('SCEK007b')
plot_adj('SCEK007b', max_speed = 0.6, n_adj = 25) # ok

# SCEK009
out <- check_animal('SCEK009')
plot_adj('SCEK009', max_speed = 0.4, max_angle = 135)
plot_adj('SCEK009', max_speed = 0.4, max_angle = 135, n_adj = 5)
plot_adj('SCEK009', max_speed = 0.4, max_angle = 135, n_adj = 30) # ok

# SCEK020b migrates
out <- check_animal('SCEK020b')
plot(as.telemetry(d$tel[[which(d$animal == 'SCEK020b')]]))
# plot_adj('SCEK020b', max_speed = 0, many = TRUE, map = TRUE)

# SCEK026b
out <- check_animal('SCEK026b')
plot_adj('SCEK026b', max_speed = 0.5, n_adj = 20) # ok

# SCEK033
out <- check_animal('SCEK033')
plot_adj('SCEK033', max_speed = 1) # ok

# SCEK036
out <- check_animal('SCEK036')
plot_adj('SCEK036', max_speed = 0.6) # ok

# SCEK044
out <- check_animal('SCEK044')
plot_adj('SCEK044', max_speed = 1) # ok

# SCEK047
out <- check_animal('SCEK047')
plot_adj('SCEK047', max_speed = 1) # ok

# SCEK050
out <- check_animal('SCEK050')
plot_adj('SCEK050', max_speed = 0.2, max_angle = 135) # ok

# SCEK079
out <- check_animal('SCEK079')
plot_adj('SCEK079', max_speed = 0.7) # ok

# SCEK079b
out <- check_animal('SCEK079b', cap_dt = FALSE)
plot_adj('SCEK079b', max_speed = 0.4) # ok
plot_adj('SCEK079b', max_speed = 0.15, max_angle = 170) # 2/4 outliers
plot_adj('SCEK079b', max_speed = 0.23, max_angle = 170) # possible outliers
flag_outlier('SCEK079b', max_speed = 0.23, max_angle = 170, value = 2)
out <- check_animal('SCEK079b', cap_dt = FALSE)

# SCEK114
out <- check_animal('SCEK114', cap_dt = FALSE)
plot_adj('SCEK114', max_speed = 0.4, max_angle = 160) # ok

# SCEK119
out <- check_animal('SCEK119', cap_dt = FALSE)
plot_adj('SCEK119', max_speed = 0.6, max_angle = 160) # outlier
flag_outlier('SCEK119', max_speed = 0.6, max_angle = 160, value = 2)
out <- check_animal('SCEK119', cap_dt = FALSE)

# SCEK133
out <- check_animal('SCEK133', cap_dt = FALSE)
plot_adj('SCEK133', max_speed = 0.9, max_angle = 170) # ok

# SCEK136
out <- check_animal('SCEK136', cap_dt = FALSE)
plot_adj('SCEK136', max_speed = 0.6, max_angle = 170) # unsure
plot_adj('SCEK136', max_speed = 0.6, max_angle = 170, n_adj = 40) # outlier
# plot_adj('SCEK136', max_speed = 0.6, max_angle = 170, n_adj = 40, map = T)
flag_outlier('SCEK136', max_speed = 0.6, max_angle = 170, value = 1)
out <- check_animal('SCEK136', cap_dt = FALSE)

# SCEK161b
out <- check_animal('SCEK161b', cap_dt = FALSE)
plot_adj('SCEK161b', max_speed = 0.4, max_angle = 135) # ok

# SCEK163
out <- check_animal('SCEK163', cap_dt = FALSE)
plot_adj('SCEK163', max_speed = 0.8) # ok

# SCEK169
out <- check_animal('SCEK169', cap_dt = FALSE)
hist(out$speed)
plot_adj('SCEK169', max_speed = 10, n_adj = 1)

# SCEK170
out <- check_animal('SCEK170', cap_dt = FALSE)
hist(out$speed)
plot_adj('SCEK170', max_speed = 150, n_adj = 2)
plot_adj('SCEK170', max_speed = 2)
flag_outlier('SCEK170', max_speed = 2, value = 1)

out <- check_animal('SCEK170', cap_dt = FALSE)
plot_adj('SCEK170', max_speed = 2)
flag_outlier('SCEK170', max_speed = 2, value = 1)

out <- check_animal('SCEK170', cap_dt = FALSE)
plot_adj('SCEK170', max_speed = 2)
flag_outlier('SCEK170', max_speed = 2, value = 1)

out <- check_animal('SCEK170', cap_dt = FALSE)
plot_adj('SCEK170', max_speed = 1)
flag_outlier('SCEK170', max_speed = 1, value = 1)

out <- check_animal('SCEK170', cap_dt = FALSE) # ok

# SCEK176
out <- check_animal('SCEK176')
plot_adj('SCEK176', max_speed = 0.6) # ok

# SCEK177
out <- check_animal('SCEK177')
plot_adj('SCEK177', max_speed = 0.2, max_angle = 170)
plot_adj('SCEK177', max_speed = 0.2, max_angle = 170, n_adj = 5)
flag_outlier('SCEK177', max_speed = 0.2, max_angle = 170, value = 1)
out <- check_animal('SCEK177', cap_dt = FALSE)
which.max(out$dt) # ok
nrow(out)

# SCEK181
out <- check_animal('SCEK181', cap_dt = FALSE)
hist(out$speed)
plot_adj('SCEK181', max_speed = 2, many = TRUE)
flag_outlier('SCEK181', max_speed = 2, value = 1)

out <- check_animal('SCEK181', cap_dt = FALSE)
plot_adj('SCEK181', max_speed = 2)
flag_outlier('SCEK181', max_speed = 2, value = 1)

out <- check_animal('SCEK181', cap_dt = FALSE)
plot_adj('SCEK181', max_speed = 2)
flag_outlier('SCEK181', max_speed = 2, value = 1)

out <- check_animal('SCEK181', cap_dt = FALSE)
plot_adj('SCEK181', max_speed = 0.5) # ok

# SCEK185
out <- check_animal('SCEK185')
plot_adj('SCEK185', max_speed = 0.3) # ok

# SCEK194
out <- check_animal('SCEK194')
plot_adj('SCEK194', max_speed = 0.15, max_angle = 170, # one outlier
         reset_layout = FALSE)
i <- which(out$angle > 170 & out$speed > 0.15)[2]
points(location.lat ~ location.long,
       d$tel[[which(d$animal == 'SCEK194')]][i, ],
       col = 'blue', cex = 2)
d$tel[[which(d$animal == 'SCEK194')]]$outlier[i] <- 1
out <- check_animal('SCEK194')

# SCEK201
out <- check_animal('SCEK201')
range('hours' %#% out$dt, na.rm = TRUE)
plot_adj('SCEK201', max_speed = 0.3, max_angle = 170, n = 50)
flag_outlier('SCEK201', max_speed = 0.3, max_angle = 170, value = 1)

out <- check_animal('SCEK201')
plot_adj('SCEK201', max_speed = 0.15, max_angle = 170, n_adj = 30)
flag_outlier('SCEK201', max_speed = 0.15, max_angle = 170, value = 1)

out <- check_animal('SCEK201')
plot_adj('SCEK201', max_speed = 0.1, max_angle = 170, n_adj = 30) # ok

# SCEK204
out <- check_animal('SCEK204')
plot_adj('SCEK204', max_angle = 170, max_speed = 0.1) # ok

# SCEK206
out <- check_animal('SCEK206')
plot_adj('SCEK206', max_distance = 15000)
out[out$distance > 15000, ] # last two points are outliers
nrow(out)
flag_outlier('SCEK206', max_distance = 15000, value = 1)
out <- check_animal('SCEK206')

# SCEK212
out <- check_animal('SCEK212')
plot_adj('SCEK212', max_speed = 0.25, max_angle = 150) # ok

# SCEK213
out <- check_animal('SCEK213')
plot_adj('SCEK213', max_speed = 0.25, max_angle = 150) # ok

# SCEK214
out <- check_animal('SCEK214')
plot_adj('SCEK214', max_speed = 0.25, max_angle = 150) # ok

# SCEK216
out <- check_animal('SCEK216')
plot_adj('SCEK216', max_speed = 0.4, max_angle = 135) # ok

# SCEK238
out <- check_animal('SCEK238')
plot_adj('SCEK238', max_speed = 0.5) # ok
```

# *Rangifer tarandus caribou* (southern mountain)

```{r sm-caribou, fig.pos='!H', fig.dim=c(10, 12)}
filter(d, dataset_name == 'Rangifer_tarandus_southern_mountain') %>%
  unnest(tel) %>%
  pull(hdop) %>%
  hist(main = 'Mountain caribou HDOP')

#' some rowns have both`hdop` and `dop` values
filter(d, dataset_name == 'Rangifer_tarandus_southern_mountain') %>%
  unnest(tel) %>%
  mutate(missing_hdop = is.na(hdop),
         missing_dop = is.na(dop)) %>%
  group_by(dataset_name, missing_hdop, missing_dop) %>%
  summarise(hdop = mean(hdop, na.rm = TRUE),
            pdop = mean(pdop, na.rm = TRUE),
            dop = mean(dop, na.rm = TRUE))

#' some have substantially different non-NA values for `hdop` and `dop`
#' assuming `ctmm` takes `hdop` and that `hdop` is better than `dop`
filter(d, dataset_name == 'Rangifer_tarandus_southern_mountain') %>%
  unnest(tel) %>%
  filter(! is.na(hdop) & ! is.na(dop)) %>%
  transmute(diff = hdop - dop) %>%
  pull(diff) %>%
  hist()

# check date ranges
filter(d, dataset_name == 'Rangifer_tarandus_southern_mountain') %>%
  unnest(tel) %>%
  group_by(animal) %>%
  summarise(range = diff(range(as.POSIXct(timestamp)))) %>%
  pull(range) %>%
  as.numeric() %>%
  hist(xlab = 'Days')

# 92_prepenned
ID <- '92_prepenned'
out <- check_animal(ID)
plot_adj(ID, max_speed = 0.3)

# 101_prepenned
ID <- '101_prepenned'
out <- check_animal(ID)
plot_adj(ID, max_speed = 0.5)
which(out$speed > 0.5)
d$tel[[which(d$animal == ID)]][1, 'outlier'] <- 1 # ran after collaring
out <- check_animal(ID)
plot_adj(ID, max_speed = 0.2, max_angle = 170) # ok
plot_adj(ID, max_speed = 0.1, min_speed = 0.2, max_angle = 170) # ok

# 104_prepenned
ID <- '104_prepenned'
out <- check_animal(ID)
plot_adj(ID, max_speed = 0.4) # ok

# 105_prepenned
ID <- '105_prepenned'
out <- check_animal(ID)
plot_adj(ID, max_speed = 0.1, max_angle = 150) # ok

# 109_prepenned
ID <- '109_prepenned'
out <- check_animal(ID)
plot_adj(ID, max_speed = 0.07, max_angle = 170) # ok

# 117_prepenned
ID <- '117_prepenned'
out <- check_animal(ID)
plot_adj(ID, max_speed = 0.1) # ok

# 132_prepenned
ID <- '132_prepenned'
out <- check_animal(ID)
diff(range(as.POSIXct(out$t))) # only 15.5 days of data
plot_adj(ID, max_speed = 0.08) # ok
plot('hours' %#% out$dt)
plot_adj(ID, max_dt = 12 %#% 'hours') # ok

# 152_prepenned
ID <- '152_prepenned'
out <- check_animal(ID)
plot_adj(ID, max_speed = 0.1, max_angle = 135) # ok
```

# *Ursus arctos horribilis*

```{r ursus-arctos-horribilis, fig.pos='!H', fig.dim=c(10, 12)}
# 197
out <- check_animal('197')
plot_adj('197', max_speed = 0.3, max_angle = 170) # ok

# 242: first point is an outlier
out <- check_animal('242')
plot_adj('242', max_speed = 1, n_adj = 10)
which(out$speed > 1)
d$tel[[which(d$animal == '242')]][1, 'outlier'] <- 1
out <- check_animal('242')

# 259
out <- check_animal('259')
plot_adj('259', max_speed = 0.4, max_angle = 170, n_adj = 10)
flag_outlier('259', max_speed = 0.4, max_angle = 170, value = 1)
out <- check_animal('259')
plot_adj('259', max_speed = 0.2, max_angle = 135, n_adj = 5) # ok

# 260
out <- check_animal('260')
plot_adj('260', max_speed = 1, max_angle = 170, n_adj = 10)
flag_outlier('260', max_speed = 1, max_angle = 170, value = 1)
out <- check_animal('260')
plot_adj('260', max_speed = 0.4, max_angle = 170, n_adj = 10) # ok
plot_adj('260', max_speed = 0.3, max_angle = 170, n_adj = 10, # unsure
         min_speed = 0.4)
filter(data.frame(out), round(speed, 1) == 0.4 & angle > 170)
0.3509840 * 3645 / 1e3 # reasonable distance in an hour

# 262
out <- check_animal('262')
plot_adj('262', max_speed = 0.3, max_angle = 170) # ok
plot_adj('262', max_speed = 0.3, max_angle = 135) # ok

# 291
out <- check_animal('291')
plot_adj('291', max_speed = 0.3, max_angle = 170) # ok

# 292
out <- check_animal('292')
plot_adj('292', max_speed = 1.2, n_adj = 10)
flag_outlier(id = '292', max_speed = 1.2, value = 1)
out <- check_animal('292')
plot_adj('292', max_speed = 0.8) # ok
# plot_adj('292', max_speed = 0.8, map = TRUE) # moving along a river

# 297
out <- check_animal('297')
plot_adj('297', max_speed = 0.5) # ok

# 317
out <- check_animal('317')
plot_adj('317', max_speed = 1.5, reset_layout = FALSE) # previous = outlier
i <- which(out$speed > 1.5) - 1
points(location.lat ~ location.long,d$tel[[which(d$animal == '317')]][i, ],
       cex = 2, col = 'blue')
d$tel[[which(d$animal == '317')]][i, 'outlier'] <- 1
out <- check_animal('317')
plot_adj('317', max_speed = 0.8) # ok
plot_adj('317', max_speed = 0.3, max_angle = 170) # possible outliers...
plot_adj('317', max_speed = 0.3, max_angle = 160) # but common at 0.3 m/s

# 318
out <- check_animal('318')
plot_adj('318', max_speed = 0.5) # ok
plot_adj('318', max_speed = 0.4, min_speed = 0.5) # ok
```

### 2025-05-09 writing/appendix-a-cleaning-telemetry-data.Rmd
shortened appendix; finished addressing comments

@@
 bibliography: 'bc-mammals-temperature.bib'
-csl: 'freshwater-biology.csl'
+csl: 'movement-ecology.csl'
 # csl: 'nature.csl'
@@
 subparagraph: true # needed for \usepackage[small]{titlesec}
+urlcolor: cyan
 output:
   bookdown::pdf_document2:
-    latex_engine: xelatex # for Delta symbol in table
     pandoc_args:
@@
                       message = FALSE,
-                      root.dir = '~/GitHub/bc-mammals-temperature')
+                      root.dir = 'H:/GitHub/bc-mammals-temperature')
 ```
@@
 
-\tableofcontents
-
-\clearpage
-
-# Setup
-
-We start by attaching the necessary packages and sourcing some custom functions.
-The custom functions are all available on GitHub in the `functions` folder of the repository located at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).
-The original cleaning script is located at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R).
-We suggest looking at the script since this appendix does not include the interactive maps that are generated with `plot_adj(..., map = TRUE)`, since they could not be included in the pdf output easily.
-However, the interacive maps often show linear features (e.g., roads) and other geographic features (e.g., mountain ridges, rivers) that were useful to see during the data cleaning.
+In this appendix, we illustrate the cleaning procedure for the telemetry data.
+For the sake of brevity, we only show the code and outputs for *Canis lupus*; the full script is available at  [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R).
+We suggest looking at the script since this appendix does not include the interactive maps that are generated by `plot_adj(..., map = TRUE)`.
+The interactive maps often show linear features (e.g., roads) and other geographic features (e.g., mountain ridges, rivers) that were useful to see during the data cleaning.
 Therefore, we have left the function calls in the document for completeness, but we have commented them out to prevent them from being executed.
@@
 ```{r map-example, echo=FALSE, out.height='40%'}
-setwd('writing')
-knitr::include_graphics(path = '../figures/BW008-map-example.jpg')
+knitr::include_graphics(path = 'H:/GitHub/bc-mammals-temperature/figures/BW008-map-example.jpg',
+                        rel_path = FALSE)
 ```
 
+We start by attaching the necessary packages and sourcing some custom functions.
+The custom functions are all available on GitHub in the `functions` folder of the repository located at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).
+
 \clearpage
+
+\small
 
@@
 
+\normalsize
+
 Before cleaning the telemetry data, we do some basic checks:
+
+\small
 
@@
 
+\normalsize
+
 We can now import the data...
+
+\small
 
@@
 
+\normalsize
+
 ... and do some last few check before cleaning.
+
+\small
 
@@
 sum(duplicated(d$animal))
+```
 
+\clearpage
+
+```{r final-checks2}
 # check which species have DOP values
@@
 
+\normalsize
+
+Now that we have organized the full telemetry dataset, we can start cleaning the data.
+
 \clearpage
 
-# *Canis lupus*
+\small
 
@@
 plot_adj('BW008', max_angle = 170, max_speed = 0.5) # fine sampling
+```
+
+\clearpage
+
+```{r canis-lupus-2, fig.pos='!H', fig.dim=c(10, 12)}
 out[out$angle > 170 & out$speed > 0.5, ] # 15-minute sampling
+```
 
+\clearpage
+
+```{r canis-lupus-3, fig.pos='!H', fig.dim=c(10, 12)}
 # BW009
@@
 plot_adj('BW012', max_speed = 1, max_angle = 170)
+```
+
+```{r canis-lupus-4, fig.pos='!H', fig.dim=c(5, 6)}
 # location is ok
@@
 rm(tel, i)
+```
 
+```{r canis-lupus-5, fig.pos='!H', fig.dim=c(10, 12)}
 # BW013
@@
 ```
-
-# *Cervus canadensis*
-
-```{r cervus-canadensis, fig.pos='!H', fig.dim=c(10, 12)}
-#' *NOTE:* assuming movement at ~< 0.25 m/s is realistic, even if angle >
-#' 170 degrees since it's too hard to distinguish between movement and gps
-#' error at that scale, and the models account for GPS error anyway (which
-#' I assume to be 10 m since I don't have calibration data. Mike said most
-#' collars have an error ~10 m anyway).
-
-# E002
-out <- check_animal('E002')
-plot_adj('E002', max_speed = 0.4)
-plot_adj('E002', max_angle = 170, max_speed = 0.2)
-
-# E003
-out <- check_animal('E003')
-plot_adj('E003', max_speed = 0.5)
-
-# E006
-out <- check_animal('E006')
-plot_adj('E006', max_speed = 0.8) # ok
-plot_adj('E006', max_speed = 0.4, max_angle = 170) # outlier
-#plot_adj('E006', max_speed = 0.4, max_angle = 170, map = TRUE) # see river
-# dt is much larger: collar struggled to receive signal with tree cover
-'hours' %#% out[which(out$speed > 0.4 & out$angle > 170) + (-4:4), 'dt']
-'hours' %#% c(43282, 7254)
-flag_outlier('E006', max_speed = 0.4, max_angle = 170, value = 1)
-out <- check_animal('E006')
-plot_adj('E006', max_speed = 0.2, max_angle = 170) # ok
-
-# E007
-out <- check_animal('E007')
-plot_adj('E007', max_speed = 0.2, max_angle = 170) # ok
-
-# E008
-out <- check_animal('E008')
-plot_adj('E008', max_speed = 0.3) # ok
-
-# E011: last point is outlier (keeping previous points bc many 0 speeds)
-out <- check_animal('E011')
-plot_adj('E011', max_speed = 0.4, n_adj = 10)
-# plot_adj('E011', max_speed = 0.4, n_adj = 10, map = TRUE)
-plot_adj('E011', max_speed = 0.4, n_adj = 9, reset_layout = FALSE)
-which(out$speed > 0.4) == nrow(out)
-i_tel <- which(d$animal == 'E011')
-i <- (nrow(d$tel[[i_tel]]) - 7):nrow(d$tel[[i_tel]]) # drop last 8 points
-points(location.lat ~ location.long, d$tel[[i_tel]][i, ], col = 'blue',
-       cex = 2, pch = 19)
-d$tel[[i_tel]][i, 'outlier'] <- 1
-points(location.lat ~ location.long, d$tel[[i_tel]][i, ],
-       col = as.numeric(d$tel[[i_tel]]$outlier[i]) + 1, cex = 2)
-layout(1)
-
-out <- check_animal('E011')
-tel <- as.telemetry(d$tel[[i_tel]], mark.rm = TRUE)
-plot(tel[nrow(tel) + -10:0, ], error = FALSE, type = 'b',
-     pch = as.character(0:9))
-plot_adj('E011', max_speed = 0.25)
-
-# E013
-out <- check_animal('E013')
-plot_adj('E013', max_speed = 0.2, max_angle = 170) # escaped something?
-
-# E015 has multiple GPS malfunctions
-out <- check_animal('E015')
-plot_adj('E015', max_speed = 1, reset_layout = FALSE)
-# can't filter easily because the outliers don't have the max velocity
-i <- d %>%
-  pull(tel) %>%
-  nth(which(d$animal == 'E015')) %>%
-  mutate(i = 1:n()) %>%
-  slice(which(timestamp ==
-                as.POSIXct(out[which(out$speed > 0.6)[2], 't'])) +
-          0:1) %>%
-  pull(i) %>%
-  first() + 0:10
-points(location.lat ~ location.long, cex = 2, lwd = 3, col = 'blue',
-       filter(d, animal == 'E015')$tel[[1]][i, ])
-lines(location.lat ~ location.long,
-      filter(d, animal == 'E015')$tel[[1]][i, ], lwd = 6, col = 'red3')
-d$tel[[which(d$animal == 'E015')]][i, 'outlier'] <- 1
-out <- check_animal('E015')
-plot_adj('E015', max_speed = 0.60) # track looks ok now
-
-# E016
-out <- check_animal('E016')
-plot_adj('E016', max_speed = 0.2, max_angle = 170) # one outlier
-plot_adj('E016', max_speed = 0.2, max_angle = 179) # use higher upper bound
-flag_outlier('E016', max_speed = 0.2, max_angle = 179, value = 1)
-out <- check_animal('E016')
-plot_adj('E016', max_speed = 0.2, max_angle = 170) # ok
-
-# E017
-out <- check_animal('E017')
-plot_adj('E017', max_speed = 0.6) # ok
-plot_adj('E017', max_angle = 170, max_speed = 0.2)
-
-# E019
-out <- check_animal('E019') # many high speeds at angle > 170
-plot_adj('E019', max_speed = 0.65, max_angle = 135) # some outliers
-plot_adj('E019', max_speed = 0.65, max_angle = 160, reset_layout = FALSE)
-i <- which(out$speed > 0.65 & out$angle > 160)[c(1, 3)] # some outliers
-points(location.lat ~ location.long,
-       d$tel[[which(d$animal == 'E019')]][i, ], cex = 2, col = 'blue')
-d$tel[[which(d$animal == 'E019')]][i, 'outlier'] <- 1
-
-out <- check_animal('E019') # check other high speeds with angle > 160
-plot_adj('E019', max_speed = 0.4, max_angle = 135, reset_layout = FALSE)
-tel <- d$tel[[which(d$animal == 'E019')]]
-i <- which(out$speed > 0.4 & out$angle > 135)
-points(location.lat ~ location.long, cex = 1.5, col = 'black',
-       tel[! tel$outlier, ][i, ],
-       pch = as.character(0:(length(i) - 1)))
-'hour' %#% out[i, 'dt'] # consistent sampling
-
-# 0 is likely an error (erring on the side of caution)
-# 1 is consistent with other movement
-# 2 is likely an error
-# 3 is consistent with other movement
-# 4 is likely an error
-# 5 is realistic
-# 6 is likely an error
-# 7 is realistic
-# 8 is likely an error
-# 9 is likely an error
-i <- i[1 + c(0, 2, 4, 6, 8, 9)]
-i_id <- which(d$animal == 'E019')
-d$tel[[i_id]][! d$tel[[i_id]]$outlier, ][i, 'outlier'] <- 1
-# ensure correct locations were removed
-out <- check_animal('E019') # check other high speeds with angle > 160
-plot_adj('E019', max_speed = 0.4, max_angle = 135, reset_layout = FALSE)
-
-# E023
-out <- check_animal('E023')
-plot_adj('E023', max_speed = 1) # ok
-plot_adj('E023', max_speed = 0.6, min_speed = 1) # ok
-plot_adj('E023', max_speed = 0.3, max_angle = 170) # ok
-
-# E024
-out <- check_animal('E024')
-plot_adj('E024', max_speed = 0.4) # ok
-plot_adj('E024', max_speed = 0.2, max_angle = 170) # ok
-
-# E025
-out <- check_animal('E025')
-plot_adj('E025', max_speed = 0.3, max_angle = 170) # ok
-
-# E026
-#' this elk does not have an unusual amount of data
-hist(map_int(filter(d, species == 'Cervus canadensis')$tel, nrow))
-abline(v = nrow(d$tel[[which(d$animal == 'E026')]]), lwd = 2)
-# d$tel[[which(d$animal == 'E026')]] %>%
-#   filter(! outlier) %>%
-#   sf::st_as_sf(coords = c('location.long', 'location.lat'),
-#                crs = 4326) %>%
-#   mapview(map.types = 'Esri.WorldImagery', col.regions = 'darkorange',
-#           col = 'black', lwd = 2)
-out <- check_animal('E026') # many sharp turns at high speeds
-plot_adj('E026', max_speed = 0.6, max_angle = 135)
-#plot_adj('E026', max_speed = 0.6, max_angle = 135, map = TRUE)# look north
-flag_outlier('E026', max_speed = 0.6, max_angle = 135, value = 1)
-
-out <- check_animal('E026')
-plot_adj('E026', max_speed = 0.4, max_angle = 170) # NW one is ok
-# plot_adj('E026', max_speed = 0.4, max_angle = 170, map = TRUE)
-plot_adj('E026', max_speed = 0.42, max_angle = 170) # W one may also be ok
-plot_adj('E026', max_speed = 0.42, max_angle = 175) # W one may also be ok
-flag_outlier('E026', max_speed = 0.42, max_angle = 175, value = 1)
-
-out <- check_animal('E026')
-plot_adj('E026', max_speed = 0.35, max_angle = 170) # NW, middle, & NE ok?
-plot_adj('E026', max_speed = 0.35, min_speed = 0.4, max_angle = 170,
-         reset_layout = FALSE)
-i <- with(out, which(speed > 0.35 & speed < 0.4 & angle > 160))[c(3:4)]
-points(location.lat ~ location.long,
-       filter(d$tel[[which(d$animal == 'E026')]], ! outlier)[i, ],
-       cex = 2, col = 'blue', lwd = 2)
-#' values in `i` assume outliers have been removed
-i_tel <- which(d$animal == 'E026')
-d$tel[[i_tel]][d$tel[[i_tel]]$outlier == 0, ][i, 'outlier'] <- 1
-
-out <- check_animal('E026')
-plot_adj('E026', max_speed = 0.30, max_angle = 160) # likely ok
-
-#' using an error-informed, continuous-time movement model substantially
-#' reduces the issues caused by GPS error. Additionally, there are so many
-#' elk data (for each individual and the number of elk, too) that a few
-#' values from a single individual will have little to no effect.
-
-# E027
-out <- check_animal('E027')
-plot_adj('E027', max_speed = 0.3, max_angle = 170, n_adj = 4) # not all
-plot_adj('E027', max_speed = 0.3, max_angle = 175, n_adj = 4,
-         reset_layout = FALSE)
-i <- with(out, which(speed > 0.3 & angle > 175))[c(2, 3, 4, 5)]
-points(location.lat ~ location.long,
-       d$tel[[which(d$animal == 'E027')]][i, ],
-       cex = 2, col = 'blue', lwd = 2)
-'hours' %#% out$dt[i]
-d$tel[[which(d$animal == 'E027')]][i, 'outlier'] <- 2 # not fully sure
-
-# E030
-out <- check_animal('E030')
-plot_adj('E030', max_speed = 0.45) # ok
-plot_adj('E030', max_speed = 0.3, max_angle = 135) # generally ok
-plot_adj('E030', max_speed = 0.3, max_angle = 170) # possibly problematic
-flag_outlier('E030', max_speed = 0.3, max_angle = 170, value = 1)
-
-out <- check_animal('E030')
-plot_adj('E030', max_speed = 0.2, max_angle = 170) # ok
-
-# E032
-out <- check_animal('E032')
-plot_adj('E030', max_speed = 0.3, max_angle = 170) # ok
-
-# E033
-out <- check_animal(id = 'E033')
-plot_adj('E033', max_speed = 0.6)
-
-# E036
-out <- check_animal(id = 'E036')
-# the next point is along the same line 
-plot_adj('E036', max_speed = 0.7, reset_layout = FALSE)
-points(location.lat ~ location.long, pch = 19,
-       filter(d, animal == 'E036')$tel[[1]][0:1 + which(out$speed > 0.7),])
-plot_adj('E036', max_speed = 0.6, reset_layout = FALSE) # ok
-
-# E037
-out <- check_animal(id = 'E037')
-plot_adj(id = 'E037', max_speed = 0.3, max_angle = 170) # ok
-
-# E040
-out <- check_animal(id = 'E040')
-plot_adj('E040', max_speed = 0.35, max_angle = 160) # ok
-
-# E042: range shift with highly stationary behavior
-out <- check_animal(id = 'E042')
-tel <- as.telemetry(d$tel[[which(d$animal == 'E042')]])
-range(tel$timestamp)
-plot(tel)
-# plot_adj(id = 'E042', many = TRUE, map = TRUE, max_speed = 0)
-ctmm.guess(tel)
-
-# E049
-out <- check_animal(id = 'E049')
-plot_adj('E049', max_speed = 0.4, max_angle = 170) # ok
-plot_adj('E049', max_speed = 0.3, min_speed = 0.4, max_angle = 170) # ok
-
-# E067
-out <- check_animal(id = 'E067')
-plot_adj('E067', max_speed = 0.6, max_angle = 170) # ok
-
-# E068
-out <- check_animal(id = 'E068')
-plot_adj('E068', max_speed = 0.65, reset_layout = FALSE)
-out[out$speed > 0.65, ] # 1st 2 points are right after capture (rownames)
-points(location.lat ~ location.long,
-       filter(d, animal == 'E068')$tel[[1]][1:2, ], col = 'blue', cex = 2)
-d[d$animal == 'E068', ]$tel[[1]]$outlier[1:2] <- 1
-out <- check_animal(id = 'E068')
-plot_adj('E068', max_speed = 0.4, max_angle = 170)
-# plot_adj('E068', max_speed = 0.4, max_angle = 170, map = TRUE)
-flag_outlier('E068', max_speed = 0.4, max_angle = 170, value = 1)
-out <- check_animal(id = 'E068')
-plot_adj('E068', max_speed = 0.25, max_angle = 170) # ok
-
-# E070
-out <- check_animal('E070')
-plot_adj('E070', max_speed = 0.4, max_angle = 135) # ok
-
-# E074
-out <- check_animal(id = 'E074')
-plot_adj('E074', max_speed = 0.4, max_angle = 170) # ok
-
-# E075
-out <- check_animal(id = 'E075')
-plot_adj('E075', max_speed = 0.3, max_angle = 170) # ok
-
-# E079
-out <- check_animal(id = 'E079')
-plot_adj('E079', max_speed = 0.6) # ok
-
-# E080
-out <- check_animal(id = 'E080')
-plot_adj('E080', max_speed = 0.8) # ok
-plot_adj('E080', max_speed = 0.5, max_angle = 170, reset_layout = FALSE)
-i_tel <- which(d$animal == 'E080')
-i <- which(out$speed > 0.5 & out$angle > 170)[1:2]
-points(location.lat ~ location.long,
-       filter(d$tel[[i_tel]], ! outlier)[i, ], col = 'blue',
-       cex = 2)
-d$tel[[i_tel]][! d$tel[[i_tel]]$outlier, ]$outlier[i] <- 1
-out <- check_animal(id = 'E080')
-plot_adj('E080', max_speed = 0.5, max_angle = 170) # ok
-
-# E081
-out <- check_animal(id = 'E081')
-plot_adj('E081', max_speed = 0.2, max_angle = 170) # ok
-
-# E082
-out <- check_animal(id = 'E082')
-plot_adj('E082', max_speed = 0.2, max_angle = 170) # ok
-
-# E083
-out <- check_animal(id = 'E083')
-plot_adj('E083', max_speed = 0.3, max_angle = 135) # ok
-
-# E084 is not worth keeping
-out <- check_animal(id = 'E084')
-plot(location.lat ~ location.long, filter(d, animal == 'E084')$tel[[1]],
-     type = 'b', pch = 19, col = '#00000010')
-nrow(filter(d, animal == 'E084')$tel[[1]])
-range(filter(d, animal == 'E084')$tel[[1]]$timestamp)
-d <- filter(d, animal != 'E084')
-
-# E085
-out <- check_animal(id = 'E085')
-plot_adj('E085', max_speed = 1) # definitely outliers (> 20 km!)
-flag_outlier(id = 'E085', max_speed = 1, value = 1)
-
-out <- check_animal(id = 'E085')
-plot_adj('E085', max_speed = 0.6) # maybe ok
-with(out[which(out$speed > 0.6) + -10:10, ], # likely outlier
-     plot('hours' %#% dt, dt * speed / 1e3,
-          col = as.numeric(-10:10 == 0) + 1,
-          pch = 19, ylab = 'SLD (dt * min speed, km)'))
-flag_outlier('E085', max_speed = 0.6, value = 1)
-
-out <- check_animal(id = 'E085')
-plot_adj('E085', max_speed = 0.2, max_angle = 135) # ok
-
-# E096
-out <- check_animal(id = 'E096')
-plot_adj('E096', max_speed = 0.4, max_angle = 170) # ok
-
-# E100
-out <- check_animal(id = 'E100')
-plot_adj('E100', max_speed = 0.3, max_angle = 170) # ok
-
-# E101
-out <- check_animal(id = 'E101')
-plot_adj('E101', max_speed = 0.4, max_angle = 170) # ok
-
-# E106
-d$tel[[which(d$animal == 'E106')]] %>%
-  group_by(gps.fix.type.raw) %>%
-  summarise(hdop = mean(dop, na.rm = TRUE))
-d$tel[[which(d$animal == 'E106')]] %>%
-  group_by(gps.fix.type.raw) %>%
-  summarise(hdop = mean(! is.na(dop), na.rm = TRUE))
-out <- check_animal(id = 'E106')
-plot_adj('E106', max_speed = 0.4, max_angle = 170) # large DOP values; ok
-d$tel[[which(d$animal == 'E106')]] %>%
-  as.telemetry() %>%
-  data.frame() %>%
-  slice(which(out$speed > 0.4 & out$angle > 170)) %>%
-  pull(HDOP)
-plot_adj('E106', max_speed = 0.3, max_angle = 170) # ok
-plot_adj('E106', max_speed = 0.2, max_angle = 170) # ok
-
-# E115
-# collar is very error-prone in SW forest
-out <- check_animal(id = 'E115')
-# plot_adj('E115', many = TRUE, map = TRUE)
-plot_adj('E115', max_speed = 0.4, max_angle = 170) # improbable
-plot_adj('E115', max_speed = 0.4, max_angle = 170, n_adj = 50)#but possible
-plot_adj('E115', max_speed = 0.4, max_angle = 175) # unlikely
-slice(d$tel[[which(d$animal == 'E115')]],
-      which(out$speed > 0.4 & out$angle > 175))$dop # low DOP
-flag_outlier(id = 'E115', max_speed = 0.4, max_angle = 175, value = 2)
-
-out <- check_animal('E115')
-plot_adj('E115', max_speed = 0.3, max_angle = 170) # unlikely
-flag_outlier('E115', max_speed = 0.3, max_angle = 170, value = 2)
-
-out <- check_animal('E115')
-plot_adj('E115', max_speed = 0.1, max_angle = 170, many = TRUE) # ok
-
-# E116
-out <- check_animal(id = 'E116')
-plot_adj('E116', max_speed = 0.4, max_angle = 170)
-flag_outlier(id = 'E116', max_speed = 0.4, max_angle = 170, value = 1)
-
-out <- check_animal('E116')
-plot_adj('E116', max_speed = 0.8)
-flag_outlier(id = 'E116', max_speed = 0.8, value = 1)
-
-out <- check_animal('E116')
-plot_adj('E116', max_speed = 0.3, max_angle = 170) # improbable
-plot_adj('E116', max_speed = 0.3, max_angle = 175) # but not impossible
-
-# E117 has similar error patterns to E115
-out <- check_animal(id = 'E117')
-plot_adj('E117', max_speed = 0.7) # three outliers
-# plot_adj('E117', max_speed = 0.7, map = TRUE)
-flag_outlier(id = 'E117', max_speed = 0.7, value = 1)
-
-out <- check_animal('E117')
-#plot_adj('E117', many = TRUE, map = TRUE) # more errors when more forested
-plot_adj('E117', max_speed = 0.4, max_angle = 170)
-
-# plot_adj('E117', max_speed = 0.4, max_angle = 170, map = TRUE)
-flag_outlier(id = 'E117', max_speed = 0.4, max_angle = 170, value = 1)
-
-out <- check_animal('E117')
-plot_adj('E117', max_speed = 0.3, max_angle = 170) # maybe not all
-plot_adj('E117', max_speed = 0.3, max_angle = 175) # likely all
-#plot_adj('E117', max_speed = 0.3, max_angle = 175, map = TRUE) # in forest
-flag_outlier(id = 'E117', max_speed = 0.3, max_angle = 175, value = 1)
-
-out <- check_animal('E117')
-# still an excessive amount of sharp turns with high speeds (compare E118)
-# dropping some of the data is better than dropping the whole animal
-flag_outlier('E117', max_speed = 0.1, max_angle = 130, value = 2)
-out <- check_animal('E117')
-
-# E118 (compare with E117 above)
-check_animal('E118', return_out = FALSE)
-
-# E122
-out <- check_animal(id = 'E122')
-plot_adj('E122', max_speed = 0.45) # ok
-
-# E125
-out <- check_animal(id = 'E125')
-plot_adj('E125', max_speed = 0.4, max_angle = 150)
-flag_outlier('E125', max_speed = 0.4, max_angle = 150, value = 1)
-out <- check_animal(id = 'E125') # ok
-
-# E129
-out <- check_animal(id = 'E129')
-plot_adj('E129', max_speed = 0.6) # ok
-
-# E131
-out <- check_animal(id = 'E131')
-plot_adj('E131', max_speed = 0.3, max_angle = 170) # N one is ok
-plot_adj('E131', max_speed = 0.45, max_angle = 170) # outlier
-flag_outlier('E131', max_speed = 0.45, max_angle = 170, value = 1)
-
-out <- check_animal(id = 'E131') # ok
-
-# E132
-#' some high minimum speeds at long sampling intervals for this individual
-out <- check_animal(id = 'E132', cap_dt = FALSE)
-layout(matrix(c(0, 0, 0, 1), ncol = 2)) #
-abline(v = 20, col = 'red')
-abline(h = 0.08, col = 'red')
-i <- which(out$dt > 20 %#% 'hours' & out$speed > 0.08)
-out[i, ]
-# small range shift at the daily scale within home range
-plot_adj('E132', max_speed = 0.08, max_dt = 20 %#% 'hours')
-
-# look at points near values with excessively high speeds
-i_adj <- sapply(i, function(.i) .i + -10:10) %>%
-  as.data.frame() %>% # convert matrix to a data frame
-  pivot_longer(! c()) # use long format for a column of all values
-tel <- as.telemetry(d$tel[[which(d$animal == 'E132')]])
-adj <- left_join(slice(data.frame(out), i_adj$value),
-                 slice(data.frame(tel), unique(i_adj$value)),
-                 by = c('t')) %>%
-  bind_cols(i_adj, .) %>%
-  mutate(date = as.POSIXct(t),
-         high_speed = speed > 0.08)
-layout(matrix(1:4, ncol = 2))
-for(g in seq(unique(i_adj$name))) {
-  g_name <- unique(i_adj$name)[g]
-  plot(y ~ x, filter(adj, name == g_name), type = 'l')
-  points(y ~ x, data.frame(tel[i, ][g, ]), col = 'blue', pch = 19)
-  rm(g, g_name)
-} # ok
-layout(1)
-
-# large dt with minor "range shifts" ==> ok
-plot('days' %#% dt ~ as.POSIXct(t), out[i_adj$value, ])
-points('days' %#% dt ~ as.POSIXct(t), out[i, ], col = 'blue', pch = 19)
-
-rm(i, i_adj, tel, adj)
-
-# E134
-out <- check_animal(id = 'E134')
-plot_adj('E134', max_speed = 0.5, max_angle = 170) # ok
-
-# E138
-out <- check_animal(id = 'E138')
-plot_adj('E138', max_speed = 0.8) # ok
-plot_adj('E138', max_speed = 0.3, max_angle = 170) # ok
-
-# E139
-out <- check_animal(id = 'E139')
-plot_adj('E139', max_speed = 0.3, max_angle = 170) # ok
-
-# E140
-out <- check_animal(id = 'E140')
-plot_adj('E140', max_speed = 0.6)
-flag_outlier('E140', max_speed = 0.6, value = 1)
-
-out <- check_animal(id = 'E140') # ok
-
-# E142
-out <- check_animal(id = 'E142')
-flag_outlier('E142', max_speed = 10, value = 1) # definitely an outlier
-
-out <- check_animal(id = 'E142')
-plot_adj('E142', max_speed = 0.8) # ok
-
-# E143
-out <- check_animal(id = 'E143')
-plot_adj('E143', max_speed = 0.2, max_angle = 170) # ok
-
-# E144
-out <- check_animal(id = 'E144')
-plot_adj('E144', max_speed = 0.3, max_angle = 170) # all ok
-
-# E147
-out <- check_animal(id = 'E147')
-plot_adj('E147', max_speed = 0.3, max_angle = 170) # ok
-
-# E161
-out <- check_animal(id = 'E161')
-plot_adj('E161', max_speed = 0.3, max_angle = 170) # ok
-
-# E164: first point is definitely an outlier
-out <- check_animal(id = 'E164')
-tel <- d$tel[[which(d$animal == 'E164')]]
-plot(location.lat ~ location.long, tel, type = 'b')
-points(location.lat ~ location.long, tel[1, ], col = 'red', pch = 19)
-rm(tel)
-d$tel[[which(d$animal == 'E164')]]$outlier[1] <- 1
-
-out <- check_animal(id = 'E164')
-plot_adj('E164', max_speed = 0.3, max_angle = 160) # ok
-rm(tel)
-
-# E168
-out <- check_animal(id = 'E168')
-plot_adj('E168', max_speed = 0.3, max_angle = 170)
-flag_outlier('E168', max_speed = 0.3, max_angle = 170, value = 1)
-
-out <- check_animal(id = 'E168')
-plot_adj('E168', max_speed = 0.2, max_angle = 170) # ok
-
-# E170
-out <- check_animal(id = 'E170')
-plot_adj('E170', max_speed = 1)
-plot_adj('E170', max_speed = 0.3, max_angle = 170) # ok
-
-# E171
-out <- check_animal(id = 'E171')
-plot_adj('E171', max_speed = 0.6) # ok
-plot_adj('E171', max_speed = 0.3, max_angle = 170) # ok
-
-# E172
-out <- check_animal(id = 'E172')
-plot_adj('E172', max_speed = 0.3, max_angle = 150) # ok
-
-# E178
-out <- check_animal(id = 'E178')
-plot_adj('E178', max_speed = 0.3, max_angle = 150) # ok
-
-# E179
-out <- check_animal(id = 'E179')
-plot_adj('E179', max_speed = 0.8) # ok
-
-# E181
-out <- check_animal(id = 'E181')
-plot_adj('E181', max_speed = 0.5, max_angle = 170) # ok
-
-# E181
-out <- check_animal(id = 'E182')
-plot_adj('E182', max_speed = 0.3, max_angle = 170) # ok
-```
-
-# *Oreamnos americanus*
-
-```{r oreamnos-americanus, fig.pos='!H', fig.dim=c(10, 12)}
-d %>%
-  filter(species == 'Oreamnos americanus') %>%
-  unnest(tel) %>%
-  pull(hdop) %>%
-  hist(main = 'Goat HDOP before as.telemetry()')
-
-d %>%
-  filter(species == 'Oreamnos americanus') %>%
-  unnest(tel) %>%
-  pull(pdop) %>%
-  hist(main = 'Goat PDOP before as.telemetry()')
-
-d %>%
-  filter(species == 'Oreamnos americanus') %>%
-  mutate(tel = map(tel, \(.t) data.frame(as.telemetry(.t)))) %>%
-  unnest(tel) %>%
-  pull(HDOP) %>%
-  hist(main = 'Goat HDOP after as.telemetry()')
-
-# pdop is not in the telemetries
-d %>%
-  filter(species == 'Oreamnos americanus') %>%
-  mutate(tel = map(tel, \(.t) data.frame(as.telemetry(.t)))) %>%
-  unnest(tel) %>%
-  colnames()
-
-# substitute missing HDOP values with PDOP values
-d <- d %>%
-  mutate(
-    tel = if_else(species == 'Oreamnos americanus',
-                  map(tel, \(.t) {
-                    .t %>%
-                      mutate(hdop = if_else(is.na(hdop), pdop, hdop)) %>%
-                      return()
-                  }),
-                  tel))
-
-# ensure values are ok
-all((d %>%
-       filter(species == 'Oreamnos americanus') %>%
-       unnest(tel) %>%
-       pull(hdop)) ==
-      (d %>%
-         filter(species == 'Oreamnos americanus') %>%
-         mutate(tel = map(tel, \(.t) data.frame(as.telemetry(.t)))) %>%
-         unnest(tel) %>%
-         pull(HDOP)))
-
-# all goats have some extreme outliers in Berlin
-filter(d, species == 'Oreamnos americanus') %>%
-  unnest(tel) %>%
-  ggplot() +
-  facet_wrap(~ animal) +
-  geom_hex(aes(location.long, location.lat), bins = 10) +
-  geom_vline(xintercept = 0, color = 'darkorange') +
-  scale_fill_continuous(limits = c(1, NA), trans = 'log10') +
-  theme(legend.position = c(0.8, 0.15))
-
-filter(d, species == 'Oreamnos americanus') %>%
-  unnest(tel) %>%
-  filter(location.long < 0) %>%
-  pull(location.long) %>%
-  hist()
-abline(v = -119.8, col = 'darkorange')
-
-# set all locations with long > 0 to outliers
-d <-
-  mutate(
-    d,
-    tel = if_else(species == 'Oreamnos americanus',
-                  true =
-                    map(tel, \(.d) {
-                      mutate(
-                        .d,
-                        outlier = if_else(location.long > -119.8, 1, 0),
-                        original_outliers = outlier == 1)
-                    }),
-                  false = tel))
-
-# 30548
-out <- check_animal(id = '30548')
-hist(out$angle) # a lot of sharp turns at low speeds
-plot_adj('30548', max_speed = 0.4) # ok
-plot_adj('30548', max_speed = 0.3, max_angle = 135) # ok
-# plot_adj('30548', max_speed = 0.3, max_angle = 135, map = TRUE) # ridges
-plot_adj('30548', max_speed = 0.2, max_angle = 135) # ok
-plot_adj('30548', max_speed = 0.1, max_angle = 170, n_adj = 3) # ok
-
-# 30551
-out <- check_animal(id = '30551') # another 124, many with speed = 0
-# start and end are definitely not part of the goat's movement
-d$tel[[which(d$animal == '30551')]] %>%
-  filter(! outlier) %>%
-  ggplot(aes(location.long, location.lat, col = as.POSIXct(timestamp))) +
-  geom_path() +
-  geom_point() +
-  scale_color_viridis_c()
-flag_outlier('30551', max_distance = 2e4, value = 1)
-
-out <- check_animal(id = '30551') # last point is still a clear outlier
-plot_adj('30551', max_speed = 0.6)
-which(out$speed > 0.6) == nrow(out)
-flag_outlier('30551', max_speed = 0.6, value = 1)
-
-out <- check_animal(id = '30551') # all other speeds are reasonable
-plot_adj('30551', max_speed = 0.4, n_adj = 20) # ok
-
-# 30561
-out <- check_animal(id = '30561')
-# plot_adj('30561', max_distance = 15e3, map = TRUE)
-flag_outlier('30561', max_distance = 15e3, value = 1)
-
-out <- check_animal(id = '30561') # some high speeds at sharp turns
-plot_adj('30561', max_speed = 0.2, max_angle = 160)
-#plot_adj('30561', max_speed = 0.2, max_angle = 160, map = TRUE) # outliers
-flag_outlier('30561', max_speed = 0.2, max_angle = 160, value = 1)
-
-out <- check_animal(id = '30561') # speeds are reasonable
-# plot_adj('30561', max_speed = 0, map = TRUE, many = TRUE) # ok
-
-# 30567
-out <- check_animal(id = '30567')
-plot_adj('30567', max_speed = 0.3, max_angle = 160)
-flag_outlier('30567', max_speed = 0.3, max_angle = 160, value = 1)
-
-out <- check_animal(id = '30567')
-plot_adj('30567', max_speed = 0.3) # ok
-plot_adj('30567', max_speed = 0.2, max_angle = 160)
-flag_outlier('30567', max_speed = 0.2, max_angle = 160, value = 1)
-
-out <- check_animal(id = '30567')
-plot_adj('30567', max_speed = 0.15, max_angle = 160) # NE is ok
-plot_adj('30567', max_speed = 0.15, max_angle = 170)
-plot_adj('30567', max_speed = 0.15, max_angle = 170, n_adj = 3)
-#plot_adj('30567', max_speed = 0.15, max_angle = 170, n_adj = 3, map = TRUE)
-flag_outlier('30567', max_speed = 0.15, max_angle = 170, value = 1)
-
-out <- check_animal(id = '30567') # some possible outliers, but unclear
-# plot_adj('30567', max_speed = 0, many = TRUE, map = TRUE)
-plot_adj('30567', max_speed = 0.1, max_angle = 170) # ok
-out[out$speed > 0.1 & out$angle > 170, ]
-plot_adj('30567', max_speed = 0.1, max_angle = 170, max_dt = 2e4) # ok
-
-# 30575
-out <- check_animal(id = '30575')
-plot_adj('30575', max_speed = 0.8)
-flag_outlier('30575', max_speed = 0.8, value = 1)
-
-out <- check_animal(id = '30575')
-plot_adj('30575', max_speed = 0.5)
-flag_outlier('30575', max_speed = 0.5, value = 1)
-
-out <- check_animal(id = '30575')
-plot_adj('30575', max_speed = 0.3, max_angle = 170)
-flag_outlier('30575', max_speed = 0.3, max_angle = 170, value = 1)
-
-out <- check_animal(id = '30575')
-plot_adj('30575', max_speed = 0.2, max_angle = 170)
-flag_outlier('30575', max_speed = 0.2, max_angle = 170, value = 1)
-
-out <- check_animal(id = '30575')
-plot_adj('30575', max_speed = 0.175, max_angle = 170)
-flag_outlier('30575', max_speed = 0.175, max_angle = 170, value = 1)
-
-out <- check_animal(id = '30575')
-plot_adj('30575', max_speed = 0.15, max_angle = 170)
-# plot_adj('30575', max_speed = 0.15, max_angle = 170, map = TRUE)
-flag_outlier('30575', max_speed = 0.15, max_angle = 170, value = 1)
-
-out <- check_animal(id = '30575')
-plot_adj('30575', max_speed = 0.12, max_angle = 170)
-plot_adj('30575', max_speed = 0.12, max_angle = 170, max_dt = 2e4)
-flag_outlier('30575', max_speed = 0.12, max_angle = 170, max_dt = 2e4,
-             value = 1)
-
-out <- check_animal(id = '30575') # some points in forest, but speed is low
-# plot_adj('30575', many = TRUE, map = TRUE)
-plot(as.telemetry(d$tel[[which(d$animal == '30575')]], mark.rm = TRUE))
-
-# 30599
-out <- check_animal(id = '30599')
-plot_adj(id = '30599', max_speed = 0.3, max_angle = 170)
-flag_outlier(id = '30599', max_speed = 0.3, max_angle = 170, value = 1)
-
-out <- check_animal(id = '30599')
-plot_adj(id = '30599', max_speed = 0.2, max_angle = 170)
-flag_outlier(id = '30599', max_speed = 0.2, max_angle = 170, value = 1)
-
-out <- check_animal(id = '30599')
-plot_adj(id = '30599', max_speed = 0.3)
-flag_outlier(id = '30599', max_speed = 0.3, value = 1)
-
-out <- check_animal(id = '30599')
-plot_adj(id = '30599', max_speed = 0.15, max_angle = 170)
-flag_outlier(id = '30599', max_speed = 0.15, max_angle = 170, value = 1)
-
-out <- check_animal(id = '30599')
-plot_adj(id = '30599', max_speed = 0.1, max_angle = 175)
-out[out$angle > 175 & out$speed > 0.1, ]
-plot_adj(id = '30599', max_speed = 0.1, max_angle = 175, min_dt = 2e4)
-i <- which(out$speed > 0.1 & out$angle > 175 & out$dt < 2e4)
-tel <- d$tel[[which(d$animal == '30599')]]
-d$tel[[which(d$animal == '30599')]][! tel$outlier, ][i, 'outlier'] <- 1
-
-out <- check_animal(id = '30599') # ok
-
-# 30613
-out <- check_animal(id = '30613')
-plot_adj('30613', max_speed = 0.4, max_angle = 170)
-# plot_adj('30613', max_speed = 0.4, max_angle = 170, map = TRUE)
-flag_outlier('30613', max_speed = 0.4, max_angle = 170, value = 1)
-
-out <- check_animal(id = '30613')
-plot_adj('30613', max_speed = 0.2, max_angle = 170)
-flag_outlier('30613', max_speed = 0.2, max_angle = 170, value = 1)
-
-out <- check_animal(id = '30613')
-plot_adj('30613', max_speed = 0.2, max_angle = 135)
-# plot_adj('30613', max_speed = 0.2, max_angle = 135, map = TRUE)
-flag_outlier('30613', max_speed = 0.2, max_angle = 135, value = 1)
-
-out <- check_animal(id = '30613')
-plot_adj('30613', max_speed = 0.12, max_angle = 170)
-flag_outlier('30613', max_speed = 0.12, max_angle = 170, value = 1)
-
-out <- check_animal(id = '30613')
-plot_adj('30613', max_speed = 0.1, max_angle = 170) # ok
-plot_adj('30613', max_speed = 0.1, max_angle = 150) # ok
-plot_adj('30613', max_speed = 0.05, max_angle = 170) # ok
-
-# 30636
-out <- check_animal(id = '30636')
-flag_outlier('30636', max_distance = 15e3, value = 1)
-
-out <- check_animal(id = '30636')
-plot_adj('30636', max_speed = 0.4)
-flag_outlier('30636', max_speed = 0.4, value = 1)
-
-out <- check_animal(id = '30636')
-plot_adj('30636', max_speed = 0.1, max_angle = 170) # ok
-
-# 30642
-out <- check_animal(id = '30642')
-plot_adj('30642', max_speed = 1)
-flag_outlier('30642', max_speed = 1, value = 1)
-
-out <- check_animal(id = '30642')
-plot_adj('30642', max_speed = 0.3, max_angle = 170)
-out[out$speed > 0.3 & out$angle > 170, ]
-plot_adj('30642', max_speed = 0.3, max_angle = 170, max_dt = 8100)
-flag_outlier('30642', max_speed = 0.3, max_angle = 170, max_dt = 8100,
-             value = 1)
-
-out <- check_animal(id = '30642')
-plot_adj('30642', max_speed = 0.2, max_angle = 170, min_speed = 0.3)
-i_tel <- which(d$animal == '30642')
-i <- which(out$speed > 0.2 & out$speed < 0.3 & out$angle > 170)
-d$tel[[i_tel]][! d$tel[[i_tel]]$outlier, ][i, 'outlier'] <- 1
-plot_adj('30642', max_speed = 0.2, max_angle = 170, min_speed = 0.3) # ok
-
-out <- check_animal(id = '30642') # ok
-
-# 30648
-out <- check_animal(id = '30648')
-flag_outlier('30648', max_speed = 1, value = 1)
-
-out <- check_animal(id = '30648')
-plot_adj('30648', max_speed = 0.3, max_angle = 170)
-flag_outlier('30648', max_speed = 0.3, max_angle = 170, value = 1)
-
-out <- check_animal(id = '30648')
-plot_adj('30648', max_speed = 0.2, max_angle = 170) # ok
-plot_adj('30648', max_speed = 0.4, max_angle = 130)
-flag_outlier('30648', max_speed = 0.4, max_angle = 130, value = 1)
-
-out <- check_animal(id = '30648')
-plot_adj('30648', max_speed = 0.2, max_angle = 170)
-# plot_adj('30648', max_speed = 0.2, max_angle = 170, map = TRUE)
-plot_adj('30648', max_speed = 0.2, max_angle = 175)
-flag_outlier('30648', max_speed = 0.2, max_angle = 175, value = 1)
-
-out <- check_animal(id = '30648')
-plot_adj('30648', max_speed = 0.2, max_angle = 170) # ok
-plot_adj('30648', max_speed = 0.1, max_angle = 175) # ok
-```
-
-# *Puma concolor*
-
-```{r puma-concolor, fig.pos='!H', fig.dim=c(10, 12)}
-# 246
-out <- check_animal(id = '246')
-plot_adj('246', max_speed = 3)
-flag_outlier('246', max_speed = 3, value = 1)
-
-out <- check_animal(id = '246')
-plot_adj('246', max_speed = 0.5, n_adj = 3) # ok
-plot_adj('246', max_speed = 0.2, max_angle = 170) # ok
-# plot_adj('246', max_speed = 0.2, max_angle = 170, n_adj = 4, map = TRUE)
-# most southern point may be an outlier, but it looks ok
-plot_adj('246', max_speed = 0.2, max_angle = 177, min_angle = 178) # ok
-# plot_adj('246', max_speed = 0.2, max_angle = 177, min_angle = 178,
-#          n_adj = 4, map = TRUE)
-i <- which(out$speed > 0.2 & out$angle > 177 & out$angle < 178) + -5:5
-tel <- d$tel[[which(d$animal == '246')]]
-tel <- tel[! tel$outlier, ]
-plot(location.lat ~ location.long, tel[i, ], type = 'b',
-     pch = as.character(seq(length(i))))
-'hours' %#% out[i, 'dt']
-rm(i, tel, cols)
-
-# 272 has a clear speed outlier, but locations are ok
-out <- check_animal(id = '272')
-plot_adj('272', max_speed = 1) # data seem ok
-plot_adj('272', max_speed = 0.2, max_angle = 135) # ok
-
-# C1
-out <- check_animal(id = 'C1')
-plot_adj('C1', max_speed = 0.2, max_angle = 170) # ok
-
-# C2
-out <- check_animal(id = 'C2')
-plot_adj('C2', max_speed = 0.2, max_angle = 170, n_adj = 20) # gps error
-# plot_adj('C2', max_speed = 0.2, max_angle = 170, n_adj = 20, map = TRUE)
-flag_outlier('C2', max_speed = 0.2, max_angle = 170, value = 1)
-
-# C3
-out <- check_animal(id = 'C3')
-plot_adj('C3', max_speed = 1)
-# plot_adj('C3', max_speed = 1, map = TRUE)
-flag_outlier('C3', max_speed = 1, value = 1)
-out <- check_animal(id = 'C3') # remaining data are ok
-plot_adj('C3', max_speed = 0.4, max_angle = 170, n_adj = 5) # ok
-plot_adj('C3', max_speed = 0.3, max_angle = 170, n_adj = 5) # ok
-
-# C4
-out <- check_animal(id = 'C4')
-plot_adj('C4', max_speed = 0.64)
-# plot_adj('C4', max_speed = 0.64, map = TRUE)
-flag_outlier('C4', max_speed = 0.64, value = 1)
-out <- check_animal(id = 'C4') # remaining data are ok
-plot_adj('C4', max_speed = 0.2, max_angle = 170) # ok
-
-# C5
-out <- check_animal(id = 'C5')
-plot_adj('C5', max_speed = 0.6) # ok
-plot_adj('C5', max_speed = 0.4, max_angle = 170) # ok
-plot_adj('C5', max_speed = 0.3, max_angle = 170) # ok
-
-# C6
-out <- check_animal(id = 'C6')
-plot_adj('C6', max_speed = 0.4, max_angle = 170)
-flag_outlier('C6', max_speed = 0.4, max_angle = 170, value = 1)
-out <- check_animal(id = 'C6') # ok
-
-# C7
-out <- check_animal(id = 'C7')
-plot_adj('C7', max_speed = 0.6, max_angle = 135)
-flag_outlier('C7', max_speed = 0.6, max_angle = 135, value = 1)
-out <- check_animal(id = 'C7') # ok
-plot_adj('C7', max_speed = 0.4, max_angle = 135) # ok
-
-# C8
-out <- check_animal(id = 'C8')
-plot_adj('C8', max_speed = 0.55) # ok
-# plot_adj('C8', max_speed = 0.55, map = TRUE)
-'hours' %#% out[which(out$speed > 0.55) + -5:5, 'dt']
-plot_adj('C8', max_speed = 0.2, max_angle = 170) # ok
-
-# C9
-out <- check_animal(id = 'C10')
-plot_adj('C10', max_speed = 0.2, max_angle = 170) # ok
-
-# C10
-out <- check_animal(id = 'C9')
-plot_adj('C9', max_speed = 0.2, max_angle = 170) # ok
-
-# C11
-out <- check_animal(id = 'C11')
-plot_adj('C11', max_speed = 2)
-flag_outlier('C11', max_speed = 2, value = 1)
-
-out <- check_animal(id = 'C11')
-plot_adj('C11', max_speed = 0.4, max_angle = 170) # NE ok
-# plot_adj('C11', max_speed = 0.4, max_angle = 175, map = TRUE) # outlier
-'hours' %#% out[which(out$speed > 0.4 & out$angle > 175) + -3:3, 'dt']
-flag_outlier('C11', max_speed = 0.4, max_angle = 175, value = 1)
-
-out <- check_animal(id = 'C11')
-plot_adj('C11', max_speed = 0.3, max_angle = 170) # NE ok
-
-# C12
-out <- check_animal(id = 'C12')
-plot_adj('C12', max_speed = 0.8)
-i <- which(out$speed > 0.8) + c(-1:0)
-layout(t(0:1))
-points(location.lat ~ location.long, d$tel[[which(d$animal == 'C12')]][i,],
-       cex = 2, col = 'blue')
-out$dt[i] # distance seems excessive for 2-hour sampling
-d$tel[[which(d$animal == 'C12')]][i, 'outlier'] <- 2
-out <- check_animal(id = 'C12')
-plot_adj(id = 'C12', max_speed = 0.2, max_angle = 170) # ok
-rm(i)
-
-# c13
-out <- check_animal(id = 'C13')
-plot_adj('C13', max_speed = 0.3, max_angle = 170) # ok
-
-# C14
-out <- check_animal('C14')
-plot_adj('C14', max_speed = 0.4, max_angle = 160) # ok
-
-# C16
-out <- check_animal('C16')
-plot_adj('C16', max_speed = 0.3, max_angle = 170) # all outliers
-flag_outlier('C16', max_speed = 0.3, max_angle = 170, value = 1)
-out <- check_animal('C16') # ok
-
-# C17
-out <- check_animal('C17')
-plot_adj('C17', max_speed = 0.2, max_angle = 170)
-plot_adj('C17', max_speed = 0.35, max_angle = 170)
-# one of the two is the last point of the telemetry
-out[which((out$angle > 170 | is.na(out$angle)) & out$speed > 0.35), ]
-i <- which(out$angle > 170 & out$speed > 0.35)
-out[i, ]
-# don't want to flag last point
-d$tel[[which(d$animal == 'C17')]]$outlier[i] <- 1
-out <- check_animal('C17')
-plot_adj('C17', max_speed = 0.35, max_angle = 179) # ok
-
-# C18
-out <- check_animal('C18')
-plot_adj('C18', max_speed = 0.6) # the two speeds are too similar
-flag_outlier('C18', max_speed = 0.6, value = 2)
-out <- check_animal('C18')
-plot_adj('C18', max_speed = 0.2, max_angle = 170) # ok
-
-# C19
-out <- check_animal('C19')
-plot_adj('C19', max_angle = 160, max_speed = 0.4) # NW is ok
-plot_adj('C19', max_speed = 0.4, max_angle = 170)
-flag_outlier('C19', max_speed = 0.4, max_angle = 170, value = 1)
-
-out <- check_animal('C19') # ok
-
-# C23
-out <- check_animal('C23')
-plot_adj('C23', max_speed = 0.3, max_angle = 160) # S is ok
-plot_adj('C23', max_speed = 0.3, max_angle = 170) # unsure
-plot_adj('C23', max_speed = 0.3, max_angle = 170, n_adj = 3)
-out[which(out$angle > 170 & out$speed > 0.3) + -3:3, 'dt'] # too regular
-# plot_adj('C23', max_speed = 0.3, max_angle = 170, map = TRUE) # outlier
-flag_outlier('C23', max_speed = 0.3, max_angle = 170, value = 1)
-out <- check_animal('C23')
-
-# C28
-out <- check_animal('C28')
-plot_adj('C28', max_speed = 0.3, max_angle = 170)
-flag_outlier('C28', max_speed = 0.3, max_angle = 170, value = 1)
-out <- check_animal('C28')
-plot_adj('C28', max_speed = 0.25, max_angle = 170) # ok
-
-# C29
-out <- check_animal('C29')
-plot_adj('C29', max_speed = 0.5, max_angle = 160)
-flag_outlier('C29', max_speed = 0.5, max_angle = 160, value = 1)
-out <- check_animal('C29')
-
-# C30
-out <- check_animal('C30')
-plot_adj('C30', max_speed = 0.6)
-flag_outlier('C30', max_speed = 0.6, value = 1)
-
-out <- check_animal('C30')
-plot_adj('C30', max_speed = 0.3, max_angle = 170)
-flag_outlier('C30', max_speed = 0.3, max_angle = 170, value = 1)
-
-out <- check_animal('C30')
-plot_adj('C30', max_speed = 0.3, max_angle = 135) # ok
-
-# C31
-out <- check_animal('C31')
-plot_adj('C31', max_speed = 0.37, max_angle = 160)
-out[which(out$speed > 0.37 & out$angle > 160) + -3:3, 'dt'] # too regular
-# plot_adj('C31', max_speed = 0.37, max_angle = 160, map = TRUE) # outlier
-flag_outlier('C31', max_speed = 0.37, max_angle = 160, value = 1)
-out <- check_animal('C31') # ok
-```
-
-# *Rangifer tarandus caribou* (boreal)
-
-```{r boreal-caribou, fig.pos='!H', fig.dim=c(10, 12)}
-# HSCEK063
-out <- check_animal('HSCEK063')
-plot_adj('HSCEK063', max_speed = 0.6) # ok
-plot_adj('HSCEK063', max_speed = 0.2, max_angle = 135) # ok
-
-# SCEK007b
-out <- check_animal('SCEK007b')
-plot_adj('SCEK007b', max_speed = 0.6, n_adj = 25) # ok
-
-# SCEK009
-out <- check_animal('SCEK009')
-plot_adj('SCEK009', max_speed = 0.4, max_angle = 135)
-plot_adj('SCEK009', max_speed = 0.4, max_angle = 135, n_adj = 5)
-plot_adj('SCEK009', max_speed = 0.4, max_angle = 135, n_adj = 30) # ok
-
-# SCEK020b migrates
-out <- check_animal('SCEK020b')
-plot(as.telemetry(d$tel[[which(d$animal == 'SCEK020b')]]))
-# plot_adj('SCEK020b', max_speed = 0, many = TRUE, map = TRUE)
-
-# SCEK026b
-out <- check_animal('SCEK026b')
-plot_adj('SCEK026b', max_speed = 0.5, n_adj = 20) # ok
-
-# SCEK033
-out <- check_animal('SCEK033')
-plot_adj('SCEK033', max_speed = 1) # ok
-
-# SCEK036
-out <- check_animal('SCEK036')
-plot_adj('SCEK036', max_speed = 0.6) # ok
-
-# SCEK044
-out <- check_animal('SCEK044')
-plot_adj('SCEK044', max_speed = 1) # ok
-
-# SCEK047
-out <- check_animal('SCEK047')
-plot_adj('SCEK047', max_speed = 1) # ok
-
-# SCEK050
-out <- check_animal('SCEK050')
-plot_adj('SCEK050', max_speed = 0.2, max_angle = 135) # ok
-
-# SCEK079
-out <- check_animal('SCEK079')
-plot_adj('SCEK079', max_speed = 0.7) # ok
-
-# SCEK079b
-out <- check_animal('SCEK079b', cap_dt = FALSE)
-plot_adj('SCEK079b', max_speed = 0.4) # ok
-plot_adj('SCEK079b', max_speed = 0.15, max_angle = 170) # 2/4 outliers
-plot_adj('SCEK079b', max_speed = 0.23, max_angle = 170) # possible outliers
-flag_outlier('SCEK079b', max_speed = 0.23, max_angle = 170, value = 2)
-out <- check_animal('SCEK079b', cap_dt = FALSE)
-
-# SCEK114
-out <- check_animal('SCEK114', cap_dt = FALSE)
-plot_adj('SCEK114', max_speed = 0.4, max_angle = 160) # ok
-
-# SCEK119
-out <- check_animal('SCEK119', cap_dt = FALSE)
-plot_adj('SCEK119', max_speed = 0.6, max_angle = 160) # outlier
-flag_outlier('SCEK119', max_speed = 0.6, max_angle = 160, value = 2)
-out <- check_animal('SCEK119', cap_dt = FALSE)
-
-# SCEK133
-out <- check_animal('SCEK133', cap_dt = FALSE)
-plot_adj('SCEK133', max_speed = 0.9, max_angle = 170) # ok
-
-# SCEK136
-out <- check_animal('SCEK136', cap_dt = FALSE)
-plot_adj('SCEK136', max_speed = 0.6, max_angle = 170) # unsure
-plot_adj('SCEK136', max_speed = 0.6, max_angle = 170, n_adj = 40) # outlier
-# plot_adj('SCEK136', max_speed = 0.6, max_angle = 170, n_adj = 40, map = T)
-flag_outlier('SCEK136', max_speed = 0.6, max_angle = 170, value = 1)
-out <- check_animal('SCEK136', cap_dt = FALSE)
-
-# SCEK161b
-out <- check_animal('SCEK161b', cap_dt = FALSE)
-plot_adj('SCEK161b', max_speed = 0.4, max_angle = 135) # ok
-
-# SCEK163
-out <- check_animal('SCEK163', cap_dt = FALSE)
-plot_adj('SCEK163', max_speed = 0.8) # ok
-
-# SCEK169
-out <- check_animal('SCEK169', cap_dt = FALSE)
-hist(out$speed)
-plot_adj('SCEK169', max_speed = 10, n_adj = 1)
-
-# SCEK170
-out <- check_animal('SCEK170', cap_dt = FALSE)
-hist(out$speed)
-plot_adj('SCEK170', max_speed = 150, n_adj = 2)
-plot_adj('SCEK170', max_speed = 2)
-flag_outlier('SCEK170', max_speed = 2, value = 1)
-
-out <- check_animal('SCEK170', cap_dt = FALSE)
-plot_adj('SCEK170', max_speed = 2)
-flag_outlier('SCEK170', max_speed = 2, value = 1)
-
-out <- check_animal('SCEK170', cap_dt = FALSE)
-plot_adj('SCEK170', max_speed = 2)
-flag_outlier('SCEK170', max_speed = 2, value = 1)
-
-out <- check_animal('SCEK170', cap_dt = FALSE)
-plot_adj('SCEK170', max_speed = 1)
-flag_outlier('SCEK170', max_speed = 1, value = 1)
-
-out <- check_animal('SCEK170', cap_dt = FALSE) # ok
-
-# SCEK176
-out <- check_animal('SCEK176')
-plot_adj('SCEK176', max_speed = 0.6) # ok
-
-# SCEK177
-out <- check_animal('SCEK177')
-plot_adj('SCEK177', max_speed = 0.2, max_angle = 170)
-plot_adj('SCEK177', max_speed = 0.2, max_angle = 170, n_adj = 5)
-flag_outlier('SCEK177', max_speed = 0.2, max_angle = 170, value = 1)
-out <- check_animal('SCEK177', cap_dt = FALSE)
-which.max(out$dt) # ok
-nrow(out)
-
-# SCEK181
-out <- check_animal('SCEK181', cap_dt = FALSE)
-hist(out$speed)
-plot_adj('SCEK181', max_speed = 2, many = TRUE)
-flag_outlier('SCEK181', max_speed = 2, value = 1)
-
-out <- check_animal('SCEK181', cap_dt = FALSE)
-plot_adj('SCEK181', max_speed = 2)
-flag_outlier('SCEK181', max_speed = 2, value = 1)
-
-out <- check_animal('SCEK181', cap_dt = FALSE)
-plot_adj('SCEK181', max_speed = 2)
-flag_outlier('SCEK181', max_speed = 2, value = 1)
-
-out <- check_animal('SCEK181', cap_dt = FALSE)
-plot_adj('SCEK181', max_speed = 0.5) # ok
-
-# SCEK185
-out <- check_animal('SCEK185')
-plot_adj('SCEK185', max_speed = 0.3) # ok
-
-# SCEK194
-out <- check_animal('SCEK194')
-plot_adj('SCEK194', max_speed = 0.15, max_angle = 170, # one outlier
-         reset_layout = FALSE)
-i <- which(out$angle > 170 & out$speed > 0.15)[2]
-points(location.lat ~ location.long,
-       d$tel[[which(d$animal == 'SCEK194')]][i, ],
-       col = 'blue', cex = 2)
-d$tel[[which(d$animal == 'SCEK194')]]$outlier[i] <- 1
-out <- check_animal('SCEK194')
-
-# SCEK201
-out <- check_animal('SCEK201')
-range('hours' %#% out$dt, na.rm = TRUE)
-plot_adj('SCEK201', max_speed = 0.3, max_angle = 170, n = 50)
-flag_outlier('SCEK201', max_speed = 0.3, max_angle = 170, value = 1)
-
-out <- check_animal('SCEK201')
-plot_adj('SCEK201', max_speed = 0.15, max_angle = 170, n_adj = 30)
-flag_outlier('SCEK201', max_speed = 0.15, max_angle = 170, value = 1)
-
-out <- check_animal('SCEK201')
-plot_adj('SCEK201', max_speed = 0.1, max_angle = 170, n_adj = 30) # ok
-
-# SCEK204
-out <- check_animal('SCEK204')
-plot_adj('SCEK204', max_angle = 170, max_speed = 0.1) # ok
-
-# SCEK206
-out <- check_animal('SCEK206')
-plot_adj('SCEK206', max_distance = 15000)
-out[out$distance > 15000, ] # last two points are outliers
-nrow(out)
-flag_outlier('SCEK206', max_distance = 15000, value = 1)
-out <- check_animal('SCEK206')
-
-# SCEK212
-out <- check_animal('SCEK212')
-plot_adj('SCEK212', max_speed = 0.25, max_angle = 150) # ok
-
-# SCEK213
-out <- check_animal('SCEK213')
-plot_adj('SCEK213', max_speed = 0.25, max_angle = 150) # ok
-
-# SCEK214
-out <- check_animal('SCEK214')
-plot_adj('SCEK214', max_speed = 0.25, max_angle = 150) # ok
-
-# SCEK216
-out <- check_animal('SCEK216')
-plot_adj('SCEK216', max_speed = 0.4, max_angle = 135) # ok
-
-# SCEK238
-out <- check_animal('SCEK238')
-plot_adj('SCEK238', max_speed = 0.5) # ok
-```
-
-# *Rangifer tarandus caribou* (southern mountain)
-
-```{r sm-caribou, fig.pos='!H', fig.dim=c(10, 12)}
-filter(d, dataset_name == 'Rangifer_tarandus_southern_mountain') %>%
-  unnest(tel) %>%
-  pull(hdop) %>%
-  hist(main = 'Mountain caribou HDOP')
-
-#' some rowns have both`hdop` and `dop` values
-filter(d, dataset_name == 'Rangifer_tarandus_southern_mountain') %>%
-  unnest(tel) %>%
-  mutate(missing_hdop = is.na(hdop),
-         missing_dop = is.na(dop)) %>%
-  group_by(dataset_name, missing_hdop, missing_dop) %>%
-  summarise(hdop = mean(hdop, na.rm = TRUE),
-            pdop = mean(pdop, na.rm = TRUE),
-            dop = mean(dop, na.rm = TRUE))
-
-#' some have substantially different non-NA values for `hdop` and `dop`
-#' assuming `ctmm` takes `hdop` and that `hdop` is better than `dop`
-filter(d, dataset_name == 'Rangifer_tarandus_southern_mountain') %>%
-  unnest(tel) %>%
-  filter(! is.na(hdop) & ! is.na(dop)) %>%
-  transmute(diff = hdop - dop) %>%
-  pull(diff) %>%
-  hist()
-
-# check date ranges
-filter(d, dataset_name == 'Rangifer_tarandus_southern_mountain') %>%
-  unnest(tel) %>%
-  group_by(animal) %>%
-  summarise(range = diff(range(as.POSIXct(timestamp)))) %>%
-  pull(range) %>%
-  as.numeric() %>%
-  hist(xlab = 'Days')
-
-# 92_prepenned
-ID <- '92_prepenned'
-out <- check_animal(ID)
-plot_adj(ID, max_speed = 0.3)
-
-# 101_prepenned
-ID <- '101_prepenned'
-out <- check_animal(ID)
-plot_adj(ID, max_speed = 0.5)
-which(out$speed > 0.5)
-d$tel[[which(d$animal == ID)]][1, 'outlier'] <- 1 # ran after collaring
-out <- check_animal(ID)
-plot_adj(ID, max_speed = 0.2, max_angle = 170) # ok
-plot_adj(ID, max_speed = 0.1, min_speed = 0.2, max_angle = 170) # ok
-
-# 104_prepenned
-ID <- '104_prepenned'
-out <- check_animal(ID)
-plot_adj(ID, max_speed = 0.4) # ok
-
-# 105_prepenned
-ID <- '105_prepenned'
-out <- check_animal(ID)
-plot_adj(ID, max_speed = 0.1, max_angle = 150) # ok
-
-# 109_prepenned
-ID <- '109_prepenned'
-out <- check_animal(ID)
-plot_adj(ID, max_speed = 0.07, max_angle = 170) # ok
-
-# 117_prepenned
-ID <- '117_prepenned'
-out <- check_animal(ID)
-plot_adj(ID, max_speed = 0.1) # ok
-
-# 132_prepenned
-ID <- '132_prepenned'
-out <- check_animal(ID)
-diff(range(as.POSIXct(out$t))) # only 15.5 days of data
-plot_adj(ID, max_speed = 0.08) # ok
-plot('hours' %#% out$dt)
-plot_adj(ID, max_dt = 12 %#% 'hours') # ok
-
-# 152_prepenned
-ID <- '152_prepenned'
-out <- check_animal(ID)
-plot_adj(ID, max_speed = 0.1, max_angle = 135) # ok
-```
-
-# *Ursus arctos horribilis*
-
-```{r ursus-arctos-horribilis, fig.pos='!H', fig.dim=c(10, 12)}
-# 197
-out <- check_animal('197')
-plot_adj('197', max_speed = 0.3, max_angle = 170) # ok
-
-# 242: first point is an outlier
-out <- check_animal('242')
-plot_adj('242', max_speed = 1, n_adj = 10)
-which(out$speed > 1)
-d$tel[[which(d$animal == '242')]][1, 'outlier'] <- 1
-out <- check_animal('242')
-
-# 259
-out <- check_animal('259')
-plot_adj('259', max_speed = 0.4, max_angle = 170, n_adj = 10)
-flag_outlier('259', max_speed = 0.4, max_angle = 170, value = 1)
-out <- check_animal('259')
-plot_adj('259', max_speed = 0.2, max_angle = 135, n_adj = 5) # ok
-
-# 260
-out <- check_animal('260')
-plot_adj('260', max_speed = 1, max_angle = 170, n_adj = 10)
-flag_outlier('260', max_speed = 1, max_angle = 170, value = 1)
-out <- check_animal('260')
-plot_adj('260', max_speed = 0.4, max_angle = 170, n_adj = 10) # ok
-plot_adj('260', max_speed = 0.3, max_angle = 170, n_adj = 10, # unsure
-         min_speed = 0.4)
-filter(data.frame(out), round(speed, 1) == 0.4 & angle > 170)
-0.3509840 * 3645 / 1e3 # reasonable distance in an hour
-
-# 262
-out <- check_animal('262')
-plot_adj('262', max_speed = 0.3, max_angle = 170) # ok
-plot_adj('262', max_speed = 0.3, max_angle = 135) # ok
-
-# 291
-out <- check_animal('291')
-plot_adj('291', max_speed = 0.3, max_angle = 170) # ok
-
-# 292
-out <- check_animal('292')
-plot_adj('292', max_speed = 1.2, n_adj = 10)
-flag_outlier(id = '292', max_speed = 1.2, value = 1)
-out <- check_animal('292')
-plot_adj('292', max_speed = 0.8) # ok
-# plot_adj('292', max_speed = 0.8, map = TRUE) # moving along a river
-
-# 297
-out <- check_animal('297')
-plot_adj('297', max_speed = 0.5) # ok
-
-# 317
-out <- check_animal('317')
-plot_adj('317', max_speed = 1.5, reset_layout = FALSE) # previous = outlier
-i <- which(out$speed > 1.5) - 1
-points(location.lat ~ location.long,d$tel[[which(d$animal == '317')]][i, ],
-       cex = 2, col = 'blue')
-d$tel[[which(d$animal == '317')]][i, 'outlier'] <- 1
-out <- check_animal('317')
-plot_adj('317', max_speed = 0.8) # ok
-plot_adj('317', max_speed = 0.3, max_angle = 170) # possible outliers...
-plot_adj('317', max_speed = 0.3, max_angle = 160) # but common at 0.3 m/s
-
-# 318
-out <- check_animal('318')
-plot_adj('318', max_speed = 0.5) # ok
-plot_adj('318', max_speed = 0.4, min_speed = 0.5) # ok
-```

### 2025-05-09 writing/manuscript.Rmd
shortened appendix; finished addressing comments

@@
 subparagraph: true # needed for \usepackage[small]{titlesec}
+urlcolor: cyan
 output:
@@
 
-\noindent \textbf{References}: 73 (updated on 2025-01-16) <!-- counted by using nature csl -->
+\noindent \textbf{References}: 140 (updated on 2025-05-09) <!-- counted by using movement ecology csl -->
 
-\noindent \textbf{Appendices}: 1
+\noindent \textbf{Appendices}: 2
 
@@
 
-* A review of climate change effects on marine mammals in United States waters: Past predictions, observed impacts, current research and conservation imperatives https://doi.org/10.1016/j.ecochg.2022.100054
 * Global vulnerability of marine mammals to global warming: https://doi.org/10.1038/s41598-019-57280-3
-* boars' sleep quantity, efficiency and quality are reduced on warmer days (https://doi.org/10.1098/rspb.2023.2115)
 
@@
 Animals move to maximize resource availability [e.g., food, water, shelter, see: @charnov_optimal_1976; @kacelnik_ideal_1992], optimize habitat use [@winter_forecasting_2024], and escape predation [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024].
-However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], the ease of movement [e.g., snow depth and density, see @montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023].
-Additionally, temperature also affects movement indirectly by through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010] and the value of resources [e.g., water, heat, shelter -- see: @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019].
+However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], the ease of movement [e.g., snow depth and density, see @montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023].
+Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [e.g., @hall_microrefuges_2016].
+Additionally, temperature also affects movement indirectly through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010].
 Changes in temperature thus affects how animals move [e.g., movement frequency and speed -- their movement behaviour: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].
@@
 
+```{r, eval=FALSE}
+tels <- readRDS('data/tracking-data/all-tracking-data-cleaned-2024-02-22-13-49.rds') %>%
+  tidyr::unnest(tel)
+
+# all outliers
+sum(tels$outlier)
+round(mean(tels$outlier) * 100, 2)
+
+# new outliers
+sum(tels$outlier & !tels$original_outliers)
+round(mean(tels$outlier & !tels$original_outliers) * 100, 2)
+```
+
 \noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)).
-***MISSING INFO ON OTHER TELEMETRY DATA*** Clearly erroneous GPS locations were removed using diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points.
+***MISSING INFO ON OTHER TELEMETRY DATA.*** From the full set of telemetry data, a total of 2396 GPS locations (0.16%, including collar calibration data) were removed using diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points.
 Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated stationary behaviour.
-The script used to clean the data and all associated custom functions are available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).
+See Appendix A for additional information.
 
@@
 To assess the importance of including temperature as an explicit covariate (rather than including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the deviance explained, model predictions, and AIC values of the two sets of models.
-
-<!--
-
-keeping for thesis:
+We provide the relevant Directed Acyclical Graphs in the Appendix (Figs.
+S1-S2).
 
@@
 S1-S2).
--->
 
@@
 Together, the binomial HGAM and the gamma HGAM inform us on an animal's long-term average speed, since it is the product of the probability of moving and its average speed when moving.
-We fit the models with fast REML (`'fREML'`) and discretized covariates (`discrete = TRUE`) to optimize computational efficiency with no appreciable losses to model performance [@wood_generalized_2015; @wood_generalized_2017-1; @li_faster_2020].
-Additional details are provided in Appendix A.
-
-\clearpage
+We fit the models with fast Restricted Maximum Likelihood (`'fREML'`) and discretized covariates (`discrete = TRUE`) to optimize computational efficiency with no appreciable losses to model performance [@wood_generalized_2015; @wood_generalized_2017-1; @li_faster_2020].
+Additional details are provided in Appendix B.
 
@@
   mutate(Species = case_when(
-    Species == 'Rangifer tarandus (s. mountain)' ~ 'R. tarandus (SM)',
+    Species == 'Rangifer tarandus (s. mountain)' ~ 'R. tarandus (sm)',
     Species == 'Ursus arctos horribilis' ~ 'U. arctos horr.',
@@
     Species == 'Cervus canadensis' ~ 'C. canadensis',
-    Species == 'Rangifer tarandus (boreal)' ~ 'R. tarandus (boreal)',
+    Species == 'Rangifer tarandus (boreal)' ~ 'R. tarandus (b)',
     Species == 'Canis lupus' ~ 'Canis lupus',
@@
 Including temperature in the models always increased within-sample prediction, but the change was not always substantial (deviance explained increased between +0.2% and +18.6%, with all HRSFs having an increase $\ge +4.6$%), and there was relatively strong agreement between models with and without temperature (Figs.
-S5 and S13), but including temperature resulted in substantially lower AIC values for all models, especially HRSFs (all $\Delta$AIC $\le$ -342; Table \@ref(tab:temp-comparison)).
-Although this may be interpreted as better fit, it is important to note that AIC generally decreases with more complex terms when sample sizes are large, even if the more complex model is not closer to the true model [see section 1.8.6, pages 52-54 of @wood_generalized_2017].
+S5 and S13), but including temperature resulted in substantially lower AIC values for all models, especially HRSFs (all $\Delta$AIC $\ge$ 342; Table \@ref(tab:temp-comparison)).
+Although this can be interpreted as evidence for a better fit, it is important to note that the addition of terms generally decreases AIC when sample sizes are large, even if the more complex model is not closer to the true model [see section 1.8.6, pages 52-54 of @wood_generalized_2017].
 
@@
     paste0('%'),
-  DAIC = c(-5927, -1682, -6566, -4448, -813, -966, -5637, -342, -914) %>%
+  DAIC = c(5927, 1682, 6566, 4448, 813, 966, 5637, 342, 914) %>%
     round())
@@
 
-knitr::kable(temp_comp, digits = 1, format = 'markdown', caption = "Changes in deviance expained ($\\Delta$DE) and Akaike Information Criterion ($\\Delta$AIC) from including temperature in the models estimating probability of movement (P(moving)) and speed when moving (Speed) as well as in species-level Hierarchical Resource Selection Functions (HRSF).")
+knitr::kable(temp_comp, digits = 1, format = 'markdown', caption = "Changes in deviance expained ($\\Delta$DE) and Akaike Information Criterion ($\\Delta$AIC) from including temperature in the models estimating probability of movement (P(M)) and speed when moving (Speed) as well as in species-level Hierarchical Resource Selection Functions (HRSF).
+Unlike conventional marginal AIC [@akaike_new_1974], The AIC values provided by the `mgcv` package for `R` are calculated using the models' estimated effective degrees of freedom and the Bayesian variance-covariance matrices, which avoids issues related to random effects or Restricted Maximum Likelihood (REML).")
 ```
@@
 Predictions in the surface plots extend to 10% of the range away from each datum.
-The color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to -75\\% and +300\\% (i.e., $2^{\\pm 2}$) for ease of readability."}
+The color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to $2^{\\pm 2}$ for ease of readability."}
 # not specifying dt, or doy for A and time for B because change is relative
@@
 ```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each location's average distance in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase.
-For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.8 and 1.25 ($\\approx 2^{\\pm 0.58}$; original data ranged 0.72 to 1.29).
+For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.72 to 1.29).
 The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest cover."}
@@
 Changes are relative to the mean RSS for each quantile in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
-knitr::include_graphics('../figures/rss-local-cc-predictions.png')
+knitr::include_graphics('../figures/rss-local-cc-predictions-quantiles.png')
 ```
@@
 ```{r s-hrsfs, fig.cap="\\textbf{Climate change will impact each species' habitat relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase.
-For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.6}$; original data ranged 0.26 to 117).
+For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.26 to 117).
 The increase in boreal caribou's RSS for high elevations at warmer temperatures was highly uncertain and often not credible at the 95\\% level."}
@@
 \noindent We have demonstrated that temperature is an important determinant of whether, how, and where mammals move, even while accounting for seasonality (e.g, changes in photoperiod, hormones, and resources).
-Disentangling temperature from other drivers allowed us to predict how changes in climate will affect mammalian movement even if all other drivers remain approximately constant or adapt at negligible rates.
+Disentangling temperature from other drivers is important for predicting how changes in climate will affect mammalian movement when other drivers remain approximately constant.
 However, predicting mammals' responses to climate change becomes increasingly complex as habitats warm and animals are exposed to increasingly frequent conditions that are both extreme and novel.
@@
 At unusually warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behaviour and that of their competitors, predators, and/or prey can often be substantially different from the behaviour at typical temperatures.
-As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @ipcc_climate_2023], mammals will be increasingly forced towards atypical behaviours, which will have the potential to substantially alter community structures and behaviours, both during such events and afterwards [@zhang_increased_2023].
-For instance, changes in climate and phenology impact the life history and behaviour of many hibernating mammals [@wells_life_2022], and hot weather can also affects mammals' sleep quality [e.g., wild boars, *Sus scrofa*: @mortlock_sleep_2024] and likelihood to enter torpor [e.g., Australian eastern long-eared bat, *Nyctophilus bifax*: @fjelldal_nightly_2021].
-Our results show that warmer temperatures cause grizzly bears to more further earlier in the year (Fig.
-\@ref(fig:dist)), which may alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024].
-At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase their chances of finding food or being predated upon [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021] as well as the overall ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023].
-These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011], including warmer temperatures and more frequent and intense extreme events [@yao_emergence_2022; @ipcc_climate_2023], such as heat waves, cold spells, floods and droughts, plant and animal die-offs, and sudden growth surges [sometimes referred to as "black swan" events -- see: @logares_black_2012; @anderson_black-swan_2017].
+As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @ipcc_climate_2023], mammals will be increasingly forced towards atypical behaviours that may alter community structures and behaviours, both during such events and afterwards [@zhang_increased_2023].
+For instance, changes in climate and phenology impact the life history and behaviour of many hibernating mammals [@wells_life_2022], and hot weather can also affects mammals' sleep quality [@mortlock_sleep_2024] and likelihood to enter torpor [@fjelldal_nightly_2021].
+Such changes may alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024].
+At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023], increase their chances of finding food or being predated upon [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021].
+These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011].
 
@@
 Instead, we hope they serve as a starting point to (1) demonstrate that mammals' movement rates and habitat selection depend on temperature and (2) how one can estimate mammals' changes in movement behaviour due to climate change.
-An important limitation of Figs.
-\@ref(fig:dist) and \@ref(fig:hrsfs) is that we do not include measures of uncertainty around the estimated responses, as they cannot be incorporated easily in surface plots, and the purpose of the figures was to illustrate the magnitude and direction of the responses rather than contrasting the surfaces against some hypotheses.
-Still, we stress that uncertainty is crucial in producing transparent and informed results, particularly when using such estimates for decision-making or predicting future changes, as we do in Figs.
-\@ref(fig:t-dist) and \@ref(fig:t-hrsfs).
-We present the spatial estimates of distance traveled and relative habitat selection strength in 2100 without measures of uncertainty for conciseness, but we suggest that future studies and reports include some form of compatibility intervals [@gelman_are_2019], whether these be Bayesian Credible intervals [@mcelreath_statistical_2020] or Frequentist Confidence Intervals, although the latter do not provide a measure of uncertainty in the common sense of the word [@morey_fallacy_2016; @amrhein_scientists_2019].
-As alternatives to heat-map surface plots, one may include compatibility intervals using faceted or perspective plots of the surfaces along with the intervals [e.g., Fig.
-5 in @akkaya_hocagil_benchmark_2024] or overlapping contour plots on the surfaces [e.g., Fig.
-4.14 on page 187 of @wood_generalized_2017].
-For a simpler figure, one could also include lines with compatibility intervals for the responses at different temperatures.
-
-How to represent uncertainty may seem esoteric, but communicating uncertainty is crucial in assessing risk probabilistically [@aven_implementing_2002; @ayre_bayesian_2012; @kluppelberg_bayesian_2014].
-Recent events in global politics, including global conflicts [@mcnutt_scientists_2022; @tollefson_what_2022] and the rapid rise in generative machine learning [also referred to as artificial intelligence and GenAI, see @roumeliotis_chatgpt_2023; @van_noorden_chatgpt_2023; @deepseek-ai_deepseek-r1_2025] and its prohibitive energy and water demands [@li_making_2023] and carbon footprint [@luccioni_estimating_2024], have placed doubt on the likelihood and feasibility of reaching climate change goals that seemed achievable in the past decade [@united_nations_environment_programme_emissions_2024].
-Maintaining global warming below 1.5 or even 2.0$^\circ$C has been questioned [@united_nations_environment_programme_emissions_2024], and mitigating climate change will require extensive and widespread collaborative effort [e.g., @huang_achieving_2021].
-Similarly, protecting a third of the world's terrestrial and aquatic ecosystems [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022] appears less tangible as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023], and achieving such a goal will require active collaboration with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023].
-Understanding the consequences of climate change on mammals' movement behaviour and spatial distribution [@williams_range_2018] is a first step towards anticipating and proactively responding to future changes in human-induced rapid environmental change [@sih_evolution_2011], including climate change.
+Additionally, communicating uncertainty in one's estimates is crucial in assessing risk probabilistically [@aven_implementing_2002; @ayre_bayesian_2012; @kluppelberg_bayesian_2014].
+Recent events in global politics, including global conflicts [@mcnutt_scientists_2022; @tollefson_what_2022] and the rapid rise in generative machine learning [more commonly "AI", see @roumeliotis_chatgpt_2023; @van_noorden_chatgpt_2023; @deepseek-ai_deepseek-r1_2025] and its prohibitive energy and water demands [@li_making_2023] and carbon footprint [@luccioni_estimating_2024], have placed doubt on the feasibility of reaching climate change goals that seemed achievable in the past decade [@united_nations_environment_programme_emissions_2024].
+Mitigating climate change and its consequences will require extensive and widespread collaborative effort [@huang_achieving_2021].
+Similarly, protecting a third of the world's ecosystems [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022] appears less tangible as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023].
+Achieving the "30 by 30" goal will require active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023].
+Understanding the consequences of climate change on mammals' movement behaviour and spatial distribution is a first step towards anticipating and proactively responding to future changes in human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018].
 In the following sections, we discuss the implications of our results in more detail.
@@
 
-\noindent The lack of common effects of temperature across all species' movement rates indicates that responding to current conditions and preparing for future changes will require a variety of physiological and behavioral adaptations as ecological communities respond in complex and interconnected ways.
-Although our models do not account for any physiological or explicit phenological changes, the tensor product terms in Fig.
+\noindent The lack of common effects of temperature across all species' movement rates indicates that preparing for future changes will require a variety of physiological and behavioral adaptations as ecological communities respond in complex and interconnected ways.
+Although our models do not account for explicit physiological or phenological changes, the tensor product terms in Fig.
 \@ref(fig:dist) suggest that warmer temperatures cause many species to alter their daily and seasonal activity patterns (most visible in pumas and grizzly bears).
@@
 Throughout the year, they adapted their tolerance to temperature and moved less when it was relatively hot (for a given time of year), especially in spring and summer.
-The strong reduction in wolves', elk's, pumas', and boreal caribou's mid-day movement rates when summer temperatures are above 20$^\circ$C suggests that the increasingly common and intense heat waves across British Columbia will have community-wide impacts on movement rates, encounter rates, and potentially community structure [@martinez-garcia_how_2020].
-More work is necessary on interspecific responses to temperature, including the effects of temperature on predation rates [but see: @cunningham_opportunity_2021; @glass_trade-off_2021; @brivio_seeking_2024].
-As indicated by grizzly bears' tendency to travel greater distances earlier in the year with increasing temperatures, the increase in intensity and frequency of extreme heat events [@bunnell_global_2011; @yao_emergence_2022] will likely also impact the occurrence and timing of hibernation [@wells_life_2022] and migration or seasonal range expansions [@carbeck_adaptation_2022; @morley_spatiotemporal_2021; @malpeli_ungulate_2022].
-However, not all species may be able to adaptation at the current rate of climate change [@williams_range_2018; @hetem_responses_2014], especially since lanscape changes will likely depend not only on the direct effect of temperature but also on cascading changes in the availability and unpredictability of resources [@mclellan_effect_2015; @pigeon_drivers_2016; @mezzini_how_2025].
+The strong reduction in the mid-day movement rates of wolves, elk, pumas, and boreal caribou when summer temperatures were above 20$^\circ$C suggests that the increasingly common and intense heat waves across British Columbia will have community-wide impacts on movement rates, encounter rates, and potentially community structure [@martinez-garcia_how_2020].
+More work is necessary on quantifying interspecific responses to temperature, including the effects of temperature on predation rates [but see: @cunningham_opportunity_2021; @glass_trade-off_2021; @brivio_seeking_2024].
+The increase in intensity and frequency of extreme heat events [@bunnell_global_2011; @yao_emergence_2022] will likely also impact the occurrence and timing of hibernation [@wells_life_2022] and migration or seasonal range expansions [@carbeck_adaptation_2022; @morley_spatiotemporal_2021; @malpeli_ungulate_2022].
+However, not all species may be able to adapt at the current rate of climate change [@williams_range_2018; @hetem_responses_2014], especially since landscape changes will likely depend not only on the direct effect of temperature but also on cascading changes in the availability and unpredictability of resources [@mclellan_effect_2015; @pigeon_drivers_2016; @mezzini_how_2025].
 
@@
 As with the effects of temperature on movement rates, the lack of common effects of temperature across all species indicates that efficient and proactive habitat conservation will require taxonomic nuance.
-In particular, our HRSFs did not account for any explicit physiological responses to temperature that may impact movement such as changes in the phenology of plants, ice, or mammalian physiology or behaviour (e.g., moulting, migration) other than as direct responses to temperature.
+In particular, our HRSFs did not account for any explicit physiological responses to temperature that may impact movement, such as changes in the phenology of plants, ice, or mammalian physiology or behaviour (e.g., moulting, migration), other than as direct responses to temperature.
 Consequently, the behavioural changes we present should be interpreted carefully.
-While some seasonal cycles do not depend directly on temperature and may thus not be impacted directly by climate change [such as those driven by photoperiod, see @liu_effects_2022], the predictions we present also rely on the assumption that mammals will not adapt these seasonal cycles based on temperature-dependent cues, such as weather [@xu_frost_2019] or forage availability [@middleton_green-wave_2018], which may not be correct.
+While some seasonal cycles do not depend directly on temperature and may thus not be impacted directly by climate change [such as those driven by photoperiod, see @liu_effects_2022], the predictions we present also rely on the assumption that mammals will not adapt these seasonal cycles to temperature-dependent cues, such as weather [@xu_frost_2019] or forage availability [@middleton_green-wave_2018], which may not be correct.
 Additionally, it is important to stress the distinction between habitat selection and quality.
-While animals tend to select for high-quality habitats [e.g., @wright_survival_2024], high selection strength can also be a consequence of an ecological trap, an area animals are attracted to that provides a net loss of fitness rather than a gain [@zuniga-palacios_what_2021; @swearer_overview_2021].
+While animals tend to select for high-quality habitats [@kacelnik_ideal_1992; @wright_survival_2024], high selection strength can also be a consequence of an ecological trap [@zuniga-palacios_what_2021; @swearer_overview_2021].
 
@@
 The large range in predicted changes by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in Fig.
-\@ref(fig:s-hrsfs) highlight the need for careful planning that incorporates not only accurate estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates.
+\@ref(fig:s-hrsfs) highlight the need for careful planning that incorporates not only reliable estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates.
 The width of the prediction intervals in Fig.
 \@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas.
-
-Accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage availability [@mezzini_how_2025], population dynamics [@needs_ref], competitive pressure [e.g., @torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019].
-This complexity results in the great variation among the responses of different species (and individuals).
-As selection strength for current ranges changes (and likely decreases), some animals may disperse, relocate to a new habitat, or remain within the current range suffer a decrease in fitness and an increase in extinction risk [@duncan_predicting_2012].
-
-Together, the predicted changes in distance traveled and habitat selection indicate that climate change will have complex consequences for populations as well as cascade effects within ecosystems as species adapt to the changing environments [or fail to do so: @botero_evolutionary_2015; @polazzo_measuring_2024].
-The resulting changes in mammalian movement behaviour will have ramifications in individuals' encounter rates [e.g., @hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@needs_ref], and food webs [@needs_ref] as landscapes change and animals face increasing pressures from human activity [@sih_evolution_2011; @tucker_moving_2018], including habitat fragmentation [@needs_ref], habitat loss [@needs_ref], and greater human presence in wild spaces [@weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022].
+Accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage and prey availability [@mezzini_how_2025], encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@smith_densitydependent_2023], competitive pressure [@torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019].
+This complexity results in the great variation among the responses of different species (and individuals), especially as populations a rapid cascade of change [@botero_evolutionary_2015] and animals face increasing pressures from human activity, including habitat fragmentation, habitat loss, and greater human presence in wild spaces [@sih_evolution_2011; @tucker_moving_2018; @sawyer_identifying_2009; @weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022].
+As selection strength for current ranges changes (and likely decreases), some animals may disperse, relocate to a new habitat, or remain within the current range despite the reduced fitness and increased extinction risk [@duncan_predicting_2012].
 
@@
 However, when conditions across years are relatively stochastic, multi-year telemetries allow one to better estimate inter-annual variation without conflating it with inter-individual differences.
-In either case, carefully-designed sampling schedules and (Bayesian) hierarchical models can provide good estimates of the effects of interested along with appropriate measures of uncertainty [@mcelreath_statistical_2020].
+In either case, carefully-designed sampling schedules and (Bayesian) hierarchical models can provide good estimates of the effects of interested along with appropriate measures of uncertainty [@mcelreath_statistical_2020; @kluppelberg_bayesian_2014].
 

### 2025-05-10 writing/manuscript.Rmd
updated abstract to tws draft abstract; added refs

@@
 abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on individuals' fitness and habitat selection at the population, community, and ecosystem levels.
-Previous research has quantified the effects of temperature on mammalian movement behaviour, but few have quantified the effects of future climate change on fine-scale movement and habitat selection.
-We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on mammals' movement rates and habitat selection.
-We then leverage the estimated effects to predict how and where mammals will move in British Columbia (BC), Canada, throughout the 21^st^ century.
-We show that changes in temperature altered movement phenology, and that climate change will impact mammals' movement rates and habitat selection throughout the century, but that there are no common trends across all study species.
-Consequently, we suggest that species' responses be interpreted carefully and considered jointly when extrapolating the community-level consequences of climate change."
+Previous research has quantified the effects of temperature on mammalian movement behaviour, but few have quantified the effects of future climate change on mammalian movement.
+We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature the probability of moving, the speed when moving, and the habitat selection of six species throughout British Columbia, Canada.
+We show that temperature is a strong determinant of when, how, and where mammals move, and that climate change will impact mammals' movement rates and habitat selection throughout the 21st century.
+We found no common effects of temperature on the species’ movement rates, so we suggest that estimated responses be considered jointly when making inferences about how climate change will impact ecological communities.
+However, habitat selection strength decreased for all species within their current ranges, suggesting that all six species will likely shift their range in the coming decades.
+Our findings are particularly concerning as increasingly common “climate weirding” pushes species to atypical ranges and behaviors.
+We conclude by placing our findings within a holistic and collaborative framework while providing suggestions for conservation and future research."
 ```
@@
 
-\noindent \textbf{References}: 140 (updated on 2025-05-09) <!-- counted by using movement ecology csl -->
+\noindent \textbf{References}: 142 (updated on 2025-05-09) <!-- counted by using movement ecology csl -->
 
@@
 At unusually warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behaviour and that of their competitors, predators, and/or prey can often be substantially different from the behaviour at typical temperatures.
-As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @ipcc_climate_2023], mammals will be increasingly forced towards atypical behaviours that may alter community structures and behaviours, both during such events and afterwards [@zhang_increased_2023].
+As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @ipcc_climate_2023], mammals will be increasingly forced towards atypical behaviours that may alter community structures and behaviours, both during such events and afterwards [@zhang_increased_2023; @logares_black_2012; @anderson_black-swan_2017].
 For instance, changes in climate and phenology impact the life history and behaviour of many hibernating mammals [@wells_life_2022], and hot weather can also affects mammals' sleep quality [@mortlock_sleep_2024] and likelihood to enter torpor [@fjelldal_nightly_2021].
@@
 Additionally, communicating uncertainty in one's estimates is crucial in assessing risk probabilistically [@aven_implementing_2002; @ayre_bayesian_2012; @kluppelberg_bayesian_2014].
-Recent events in global politics, including global conflicts [@mcnutt_scientists_2022; @tollefson_what_2022] and the rapid rise in generative machine learning [more commonly "AI", see @roumeliotis_chatgpt_2023; @van_noorden_chatgpt_2023; @deepseek-ai_deepseek-r1_2025] and its prohibitive energy and water demands [@li_making_2023] and carbon footprint [@luccioni_estimating_2024], have placed doubt on the feasibility of reaching climate change goals that seemed achievable in the past decade [@united_nations_environment_programme_emissions_2024].
+Recent events in global politics, including global conflicts [@mcnutt_scientists_2022; @tollefson_what_2022] and the rapid rise in generative machine learning [more commonly: "AI", see @roumeliotis_chatgpt_2023; @van_noorden_chatgpt_2023; @deepseek-ai_deepseek-r1_2025] and its prohibitive energy and water demands [@li_making_2023] and carbon footprint [@luccioni_estimating_2024], have placed doubt on the feasibility of reaching climate change goals that seemed achievable in the past decade [@united_nations_environment_programme_emissions_2024].
 Mitigating climate change and its consequences will require extensive and widespread collaborative effort [@huang_achieving_2021].
@@
 This complexity results in the great variation among the responses of different species (and individuals), especially as populations a rapid cascade of change [@botero_evolutionary_2015] and animals face increasing pressures from human activity, including habitat fragmentation, habitat loss, and greater human presence in wild spaces [@sih_evolution_2011; @tucker_moving_2018; @sawyer_identifying_2009; @weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022].
-As selection strength for current ranges changes (and likely decreases), some animals may disperse, relocate to a new habitat, or remain within the current range despite the reduced fitness and increased extinction risk [@duncan_predicting_2012].
+As selection strength for current ranges changes (and likely decreases), some animals may disperse, relocate to a new habitat, or remain within the current range despite the reduced fitness and increased extinction risk [@logares_black_2012; @anderson_black-swan_2017; @duncan_predicting_2012].
 

### 2025-05-11 writing/manuscript.Rmd
improved abstract w 6-sentence structure

@@
 ```{r, echo=FALSE}
-abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on individuals' fitness and habitat selection at the population, community, and ecosystem levels.
+abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on fitness, habitat selection, and community structure.
 Previous research has quantified the effects of temperature on mammalian movement behaviour, but few have quantified the effects of future climate change on mammalian movement.
-We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature the probability of moving, the speed when moving, and the habitat selection of six species throughout British Columbia, Canada.
-We show that temperature is a strong determinant of when, how, and where mammals move, and that climate change will impact mammals' movement rates and habitat selection throughout the 21st century.
+Consequently, it is still unclear how and where mammals will move in the 21^st^ century.
+We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on the probability of moving, the speed when moving, and the habitat selection of six species throughout British Columbia, Canada.
+We show that temperature is a strong determinant of when, how, and where mammals move, and that climate change will impact mammals' movement rates and habitat selection throughout the 21^st^ century.
 We found no common effects of temperature on the species’ movement rates, so we suggest that estimated responses be considered jointly when making inferences about how climate change will impact ecological communities.
-However, habitat selection strength decreased for all species within their current ranges, suggesting that all six species will likely shift their range in the coming decades.
-Our findings are particularly concerning as increasingly common “climate weirding” pushes species to atypical ranges and behaviors.
-We conclude by placing our findings within a holistic and collaborative framework while providing suggestions for conservation and future research."
+In contrast, habitat selection strength decreased for all species within their current ranges, suggesting that all six species will shift their range in the coming decades.
+As climate change exposes mammals to novel environmental conditions, predicting changes in animal behavior and community structure become crucial for effective and proactive conservation.
+We thus conclude by providing suggestions for conservation and future research while calling for a more holistic and collaborative framework."
+if(FALSE){
+  stringi::stri_count_words(abstract)
+}
 ```

### 2025-05-11 writing/manuscript.Rmd
submitted abstract to 2025 TWS conference

@@
 abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on fitness, habitat selection, and community structure.
-Previous research has quantified the effects of temperature on mammalian movement behaviour, but few have quantified the effects of future climate change on mammalian movement.
+Previous studies have quantified the effects of temperature on mammalian movement behaviour, but few have quantified the effects of future climate change.
 Consequently, it is still unclear how and where mammals will move in the 21^st^ century.
@@
 We show that temperature is a strong determinant of when, how, and where mammals move, and that climate change will impact mammals' movement rates and habitat selection throughout the 21^st^ century.
-We found no common effects of temperature on the species’ movement rates, so we suggest that estimated responses be considered jointly when making inferences about how climate change will impact ecological communities.
+We found no common effects of temperature on the species’ movement rates, so we suggest that species' estimated responses be interpreted jointly when making inferences about how climate change will impact ecological communities.
 In contrast, habitat selection strength decreased for all species within their current ranges, suggesting that all six species will shift their range in the coming decades.
-As climate change exposes mammals to novel environmental conditions, predicting changes in animal behavior and community structure become crucial for effective and proactive conservation.
+As climate change exposes mammals to novel environmental conditions, predicting changes in animal behavior and community structure will become crucial for effective and proactive conservation.
 We thus conclude by providing suggestions for conservation and future research while calling for a more holistic and collaborative framework."

### 2025-05-16 writing/appendix-b-gams.Rmd
added GLRT results, moved AIC stuff to app. B

@@
 header-includes:
-    - \renewcommand{\figurename}{Figure S\!\!} # for "Figure Ax.
+    - \renewcommand{\figurename}{Figure B\!\!} # for "Fig. Bx.
+    - \renewcommand{\tablename}{Table B\!\!} # for "Table. Bx.
     - \usepackage{setspace}\doublespacing # for double-spaced text
@@
 ```
+
+\clearpage
+
+# Model selection
+
+```{r temp-comparison}
+setwd('..')
+library('dplyr')
+source('analysis/figures/default-ggplot-theme.R') #' for `SPECIES` object
+
+temp_comp <- tibble(
+  Model = c('$P(M)$', 'Speed', rep('HRSF', 7)),
+  Species = c('All', 'All', sort(as.character(SPECIES))),
+  DDE =
+    c(10.72 - 10.23, 79.26 - 79.11, 9.16, 4.62, 18.6, 8.0, 5.30, 12.0, 17.7) %>%
+    round(1) %>%
+    paste0('%'),
+  DAIC = c(5927, 1682, 6566, 4448, 813, 966, 5637, 342, 914) %>%
+    round())
+
+if(FALSE) {
+  range(as.numeric(gsub('%', '', temp_comp$DDE)))
+  range(as.numeric(gsub('%', '', temp_comp$DDE))[- c(1:2)])
+  max(temp_comp$DAIC)
+}
+
+colnames(temp_comp)[colnames(temp_comp) == 'DAIC'] <- paste0('\U0394', 'AIC')
+colnames(temp_comp)[colnames(temp_comp) == 'DDE'] <- paste0('\U0394', 'DE')
+
+knitr::kable(temp_comp, digits = 1, format = 'markdown', caption = "Changes in deviance expained ($\\Delta$DE) and Akaike Information Criterion ($\\Delta$AIC) from including temperature in the models estimating probability of movement ($P(M)$) and speed when moving (Speed) as well as in species-level Hierarchical Resource Selection Functions (HRSF).
+Unlike conventional marginal AIC [@akaike_new_1974], The AIC values provided by the `mgcv` package for `R` are calculated using the models' estimated effective degrees of freedom and the Bayesian variance-covariance matrices, which avoids issues related to random effects or Restricted Maximum Likelihood (REML).")
+```
+
+\clearpage
+
+## Analyses of Deviance
+
+\noindent Here we produce approximate p-values for analyses of deviance (a form of generalized likelihood ratio tests) following the methods of Section 3.3 in @wood_generalized_2017.
+We use an $F$ test rather than a $\chi^2$ test for the Gamma models because the scale parameter for the Gamma distribution is estimated rather than known or fixed [as in the case of binomial or Poisson distributions -- see table 3.1 on page 104 of @wood_generalized_2017].
+
+\singlespacing
+
+\footnotesize
+
+```{r, echo=TRUE}
+# P(moving)
+anova(readRDS('../models/binomial-gam-without-temperature.rds'),
+      readRDS('../models/binomial-gam.rds'),
+      test = 'Chisq')
+```
+
+\clearpage
+
+```{r, echo=TRUE}
+# speed when moving
+anova(readRDS('../models/gamma-gam-without-temperature.rds'),
+      readRDS('../models/gamma-gam.rds'),
+      test = 'F')
+```
+
+<!-- \clearpage -->
+
+```{r, echo=TRUE}
+# HRSFs
+anova(readRDS('../models/rsf-Canis lupus-2025-01-20.rds'),
+      readRDS('../models/rsf-Canis lupus-no-temperature-2025-01-20.rds'),
+      test = 'Chisq')
+```
+
+\clearpage
+
+```{r, echo=TRUE}
+anova(readRDS('../models/rsf-Cervus canadensis-2025-01-20.rds'),
+      readRDS('../models/rsf-Cervus canadensis-no-temperature-2025-01-20.rds'),
+      test = 'Chisq')
+```
+
+<!-- \clearpage -->
+
+```{r, echo=TRUE}
+anova(readRDS('../models/rsf-Oreamnos americanus-2025-01-20.rds'),
+      readRDS('../models/rsf-Oreamnos americanus-no-temperature-2025-01-20.rds'),
+      test = 'Chisq')
+```
+
+\clearpage
+
+```{r, echo=TRUE}
+anova(readRDS('../models/rsf-Puma concolor-2025-01-20.rds'),
+      readRDS('../models/rsf-Puma concolor-no-temperature-2025-01-20.rds'),
+      test = 'Chisq')
+```
+
+<!-- \clearpage -->
+
+```{r, echo=TRUE}
+anova(
+  readRDS('../models/rsf-Rangifer tarandus boreal-2025-01-21.rds'),
+  readRDS('../models/rsf-Rangifer tarandus boreal-no-temperature-2025-01-20.rds'),
+  test = 'Chisq')
+```
+
+\clearpage
+
+```{r, echo=TRUE}
+anova(readRDS('../models/rsf-Rangifer tarandus southern mountain-2025-01-20.rds'),
+      readRDS('../models/rsf-Rangifer tarandus southern mountain-no-temperature-2025-01-20.rds'),
+      test = 'Chisq')
+```
+
+<!-- \clearpage -->
+
+```{r, echo=TRUE}
+anova(readRDS('../models/rsf-Ursus arctos horribilis-2025-01-20.rds'),
+      readRDS('../models/rsf-Ursus arctos horribilis-no-temperature-2025-01-20.rds'),
+      test = 'Chisq')
+```
+
+\normalsize
+
+\doublespacing
+
+\clearpage
+
+# Refences

### 2025-05-16 writing/manuscript.Rmd
added GLRT results, moved AIC stuff to app. B

@@
 
-\noindent \textbf{Tables}: 3
+\noindent \textbf{Tables}: 2
 
-\noindent \textbf{References}: 142 (updated on 2025-05-09) <!-- counted by using movement ecology csl -->
+\noindent \textbf{References}: 141 (updated on 2025-05-16) <!-- counted by using movement ecology csl -->
 
@@
 
-This paper provides an analysis of the effects of air temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (*Canis lupus*, *Cervus canadensis*, *Oreamnos americanus*, *Puma concolor*, boreal and southern mountain *Rangifer tarandus*, and *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
+This paper provides an analysis of the effects of air temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, pumas: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
 Using over 25 years of telemetry data thoughout a large spatial range of British Columbia and hourly air temperature data, we estimate how mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature.
@@
 To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using the inflection point of each species' histogram of detrended speed estimates, as $k$-nearest neighbors did not discriminate between states well (Fig.
-S3).
+B3).
 
@@
 \noindent The following sections detail the statistical framework and modelling we used to estimate the effect on temperature on mammals' movement.
-To assess the importance of including temperature as an explicit covariate (rather than including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the deviance explained, model predictions, and AIC values of the two sets of models.
+To assess the importance of including temperature as an explicit covariate (rather than including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the fits of the two sets of models.
 We provide the relevant Directed Acyclical Graphs in the Appendix (Figs.
-S1-S2).
+B1-B2).
 
@@
 We provide the relevant Directed Acyclical Graphs in the Appendix (Figs.
-S1-S2).
+B1-B2).
 
@@
 The number of quadrature locations greatly outnumbered the number of observed locations (Fig.
-S13), especially after accounting for weighting based on the degree of autocorrelation and the number of home range crossings (Fig.
-S14).
+B13), especially after accounting for weighting based on the degree of autocorrelation and the number of home range crossings (Fig.
+B14).
 
@@
 \noindent Data were relatively scarce for temperatures lower than $-20^\circ$C or above $20^\circ$C (Fig.
-S4).
+B4).
 Species differed in overall mean probabilities of movement (range: 0.05 -- 0.3), mean speed when moving (range: 0.42 -- 2.7 km/day), and mean distance traveled (range: 0.04 -- 0.6 km/day; Table \@ref(tab:gam-intercepts)).
@@
 
-Including temperature in the models always increased within-sample prediction, but the change was not always substantial (deviance explained increased between +0.2% and +18.6%, with all HRSFs having an increase $\ge +4.6$%), and there was relatively strong agreement between models with and without temperature (Figs.
-S5 and S13), but including temperature resulted in substantially lower AIC values for all models, especially HRSFs (all $\Delta$AIC $\ge$ 342; Table \@ref(tab:temp-comparison)).
-Although this can be interpreted as evidence for a better fit, it is important to note that the addition of terms generally decreases AIC when sample sizes are large, even if the more complex model is not closer to the true model [see section 1.8.6, pages 52-54 of @wood_generalized_2017].
-
-```{r temp-comparison}
-setwd('..')
-library('dplyr')
-source('analysis/figures/default-ggplot-theme.R') #' for `SPECIES` object
-
-temp_comp <- tibble(
-  Model = c('$P(M)$', 'Speed', rep('HRSF', 7)),
-  Species = c('All', 'All', sort(as.character(SPECIES))),
-  DDE = 
-    c(10.72 - 10.23, 79.26 - 79.11, 9.16, 4.62, 18.6, 8.0, 5.30, 12.0, 17.7) %>%
-    round(1) %>%
-    paste0('%'),
-  DAIC = c(5927, 1682, 6566, 4448, 813, 966, 5637, 342, 914) %>%
-    round())
-
-if(FALSE) {
-  range(as.numeric(gsub('%', '', temp_comp$DDE)))
-  range(as.numeric(gsub('%', '', temp_comp$DDE))[- c(1:2)])
-  max(temp_comp$DAIC)
-}
-
-colnames(temp_comp)[colnames(temp_comp) == 'DAIC'] <- paste0('\U0394', 'AIC')
-colnames(temp_comp)[colnames(temp_comp) == 'DDE'] <- paste0('\U0394', 'DE')
-
-knitr::kable(temp_comp, digits = 1, format = 'markdown', caption = "Changes in deviance expained ($\\Delta$DE) and Akaike Information Criterion ($\\Delta$AIC) from including temperature in the models estimating probability of movement (P(M)) and speed when moving (Speed) as well as in species-level Hierarchical Resource Selection Functions (HRSF).
-Unlike conventional marginal AIC [@akaike_new_1974], The AIC values provided by the `mgcv` package for `R` are calculated using the models' estimated effective degrees of freedom and the Bayesian variance-covariance matrices, which avoids issues related to random effects or Restricted Maximum Likelihood (REML).")
-```
+There was relatively strong agreement between models with and without temperature (Figs.
+B5 and S13), but including temperature always resulted in better fits.
+All analyses of deviance showed that temperature significantly affected movement behavior (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; see Table B1 and the analyses of deviance in Appendix B).
+The only exception to this was the mountain goat HRSF, where the addition of temperature resulted in a simpler model (lower model degrees of freedom) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.
 
@@
 \noindent The effects of temperature on movement rates varied in both direction and magnitude across species, even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; see Figs.
-S6-S8).
+B6-B8).
 Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species.
 The models explained reasonably high proportions of the deviance (10.7% for the binomial model and 79.3% for the Gamma model) and had good in-sample prediction (Fig.
-S9).
+B9).
 All species adapted their daily and seasonal movement behaviour to changes in temperature (Fig.
@@
 All species' estimated probability of moving decreased with sampling intervals approximately above 1 hour, and most species' estimated speed when moving decreased monotonically with sampling interval, although the estimated trends were highly uncertain for some species (Fig.
-S10).
+B10).
 
@@
 Again, estimated RSS values were generally most uncertain at extreme temperatures (Fig.
-S13).
+B13).
 

### 2025-05-23 analysis/figures/telemetry-ud-map.R
set `elev_m < 0` to 0 (0.12%)

@@ -68,7 +68,35 @@ dem <- rast('data/resource-rasters/fig-1-dem-z-6.tif') %>%
   as.data.frame(dem, xy = TRUE) %>%
   rename(elev_m = 3)
 
-# tels only with bc map
+# some values < 0, but not enough to be useful
+mean(dem$elev_m < 0)
+mean(dem$elev_m < -10)
+mean(dem$elev_m < -100)
+
+dem <- mutate(dem, elev_m = if_else(elev_m < 0, 0, elev_m))
+
+# tels with dem
+p_dem <-
+  ggplot() +
+  geom_raster(aes(x, y, fill = elev_m), dem) +
+  geom_sf(aes(geometry = geometry, color = dataset_name), tels,
+          size = 0.1) +
+  coord_sf(xlim = c(100e4, 110e4), clip = 'off',
+           ylim = c(33e4, 185e4), crs = 'EPSG:3005') +
+  scale_color_manual(name = ' ', values = PAL, labels = parse_format()) +
+  scale_fill_distiller(name = '\nElevation (m)', palette = 6) +
+  labs(x = NULL, y = NULL) +
+  theme_void() +
+  theme(legend.position = 'inside', legend.position.inside = c(-3, 0.05),
+        legend.justification = c(1, 0)) +
+  guides(color = guide_legend(override.aes = list(alpha = 1, size = 1,
+                                                  order = 2)),
+         fill = guide_colorbar(order = 1))
+
+ggsave('figures/tels-map-dem.png', plot = p_dem,
+       width = 10, height = 8.5, units = 'in', dpi = 600, bg = 'white')
+
+# tels with bc map
 p_tels <-
   ggplot() +
   geom_sf(data = bc) +
@@ -85,27 +113,7 @@ p_tels <-
 ggsave('figures/tels-map.png', plot = p_tels,
        width = 10, height = 8.5, units = 'in', dpi = 600, bg = 'white')
 
-# tels only with dem
-p_dem <-
-  ggplot() +
-  geom_raster(aes(x, y, fill = elev_m), dem) +
-  geom_sf(aes(geometry = geometry, color = dataset_name), tels,
-          # slice_sample(tels, n = 1e4), # for testing
-          size = 0.1) +
-  coord_sf(xlim = c(100e4, 110e4), clip = 'off',
-           ylim = c(33e4, 185e4), crs = 'EPSG:3005') +
[…]

### 2025-05-23 writing/appendix-a-cleaning-telemetry-data.Rmd
updated title, authors, csl, and fig 1

@@
 ---
-title: \Large Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada
-subtitle: "Appendix A: Cleaning telemetry data"
+title: "\\Large Predicting how climate change will affect how and where terrestrial mammals will move"
 author:
@@
     institute: [braes, biol]
+  - name: TJ Gooliaff
+    institute: [forest]
   - name: Karen E. Hodges
@@
   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
-  - bcp: BC Parks
+  - forest: British Columbia Ministry of Forests, Lands, Natural Resource Operations and Rural Development, Penticton, BC, Canada.
+  - bcp: British Columbia Ministry of Environment and Parks, Penticton, BC, Canada.
   - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
@@
 bibliography: 'bc-mammals-temperature.bib'
-csl: 'movement-ecology.csl'
-# csl: 'nature.csl'
+csl: 'global-change-biology.csl'
 fontsize: 12pt

### 2025-05-23 writing/appendix-b-gams.Rmd
updated title, authors, csl, and fig 1

@@
 ---
-title: \Large Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada
-subtitle: "Appendix B: Modeling the effects of temperature on movement behavior"
+title: "\\Large Predicting how climate change will affect how and where terrestrial mammals will move"
 author:

### 2025-05-23 writing/manuscript.Rmd
updated title, authors, csl, and fig 1

@@
 ---
-title: "Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada"
+title: "\\Large Predicting how climate change will affect how and where terrestrial mammals will move"
 author:
@@
 
-```{r bc-map, fig.cap="GPS telemetry data for the six species in this study.", out.width='0.75\\linewidth'}
-knitr::include_graphics('../figures/tels-map.png')
+```{r bc-map, fig.cap="GPS telemetry data for the six species in this study (dots) over a digital elevation model for the study area.", out.width='0.75\\linewidth'}
+knitr::include_graphics('../figures/tels-map-dem.png')
 ```

### 2025-06-03 writing/appendix-a-cleaning-telemetry-data.Rmd
added more info; need to expand; moving to lab pc

@@
 
+\doublespacing
+
 We start by attaching the necessary packages and sourcing some custom functions.
 The custom functions are all available on GitHub in the `functions` folder of the repository located at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).
+While this appendix only shows the removal of a single outlier point, the appendix details the approach we used to clean the telemetry data.
+We removed outlier points using `flag_outlier(..., value = 1)`, which leverages `ctmm::as.telemetry()` to remove rows that have `TRUE` or a positive, non-zero number in the `outlier` column.
+See te comments in each code chunk for additional detail and considerations about individual data points.
+
+\singlespacing
 

### 2025-06-10 analysis/02-remove-outliers.R
added a comment for unit conversion

@@ -214,7 +214,7 @@ plot_adj('E006', max_speed = 0.4, max_angle = 170) # outlier
 plot_adj('E006', max_speed = 0.4, max_angle = 170, map = TRUE) # see river
 # dt is much larger: collar struggled to receive signal with tree cover
 'hours' %#% out[which(out$speed > 0.4 & out$angle > 170) + (-4:4), 'dt']
-'hours' %#% c(43282, 7254)
+'hours' %#% c(43282, 7254) # convert seconds to hours
 flag_outlier('E006', max_speed = 0.4, max_angle = 170, value = 1)
 out <- check_animal('E006')
 plot_adj('E006', max_speed = 0.2, max_angle = 170) # ok

### 2025-06-10 analysis/06-temperature-hgams.R
cleaned up histograms

@@ -7,8 +7,8 @@ library('gratia')    # for ggplot-based figures for GAMs
 library('ggplot2')   # for fancy plots
 library('khroma')    # for colorblind-friendly color palettes
 library('cowplot')   # for fancy multi-panel plots
+library('ctmm')      #'for unit conversions with `%#%`
 source('analysis/figures/default-ggplot-theme.R') # bold text and no grids
-source('functions/get_legend.R') # to extract legends from ggplot plots
 plot_scheme(PAL, colours = TRUE)
 
 # using Pacific Time for time of day
@@ -81,23 +81,32 @@ splits <-
                      pull(speed_est)))
          }),
   ) %>%
-  pivot_longer(c(manual, `k means`), values_to = 'split',
-               names_to = 'method')
+  pivot_longer(c(manual, `k means`), values_to = 'split_m_s',
+               names_to = 'method') %>%
+  mutate(split_km_h = 'km/h' %#% split_m_s)
 
 hists <- list()
 for(i in (1:N_SPECIES)[order(SPECIES)]) {
   sp <- as.character(SPECIES[i])
   XLIM <- c(0, quantile(filter(d, species == sp)$speed_est, 0.99))
-  YLIM <- c(0, nrow(filter(d, species == sp)) / 40)
+  XLIM <- 'km/h' %#% XLIM # convert to km/h
+  YLIM <- c(0, case_when(
+    sp == 'Rangifer tarandus (s. mountain)' ~ 400,
+    sp == 'Ursus arctos horribilis' ~ 1e3,
+    sp == 'Puma concolor' ~ 1e3,
+    sp == 'Cervus canadensis' ~ 25e3,
+    sp == 'Rangifer tarandus (boreal)' ~ 5e3,
+    sp == 'Canis lupus' ~ 4e3,
+    sp == 'Oreamnos americanus' ~ 500))
   
   hists[[i]] <-
     ggplot() +
-    geom_histogram(aes(speed_est), filter(d, species == sp),
-                   bins = 100, fill = 'grey', color = 'black') +
-    geom_vline(aes(xintercept = split, color = method,
+    geom_histogram(aes('km/h' %#% speed_est), filter(d, species == sp),
+                   bins = 200, fill = 'grey', color = 'black') +
+    geom_vline(aes(xintercept = split_km_h, color = method,
                    lty = method), filter(splits, species == sp), lwd = 0.75) +
     coord_cartesian(xlim = XLIM, ylim = YLIM) +
[…]

### 2025-06-10 analysis/figures/temperature-hgams.R
clipped temperatures based on `too.far` from `doy`

@@ -376,7 +376,9 @@ newd_tod <- expand_grid(animal = m_2$model$animal[1],
                         dt = 1) %>%
   nest(dat = -species) %>%
   mutate(dat = map2(dat, species, \(.d, .s) {
-    ref <- filter(m_1$model, species == .s)
+    ref <- filter(m_1$model, species == .s) %>%
+      filter(doy >= yday('2025-06-01') - 366 * 0.1 &
+               doy <= yday('2025-06-01') + 366 * 0.1)
     
     filter(.d, ! too_far(x = tod_pdt, y = temp_c, ref_1 = ref$tod_pdt,
                          ref_2 = ref$temp_c, dist = DIST)) %>%
@@ -537,7 +539,7 @@ distance <- plot_grid(
 ggsave('figures/distance.png', distance,
        width = 16, height = 8, dpi = 600, bg = 'white')
 
-# for appendix A
+# for appendix B
 # full figures ----
 p_mov_full <- plot_grid(
   p_mov_temp, p_mov_tod, p_mov_doy,
@@ -562,32 +564,3 @@ distance_full <- plot_grid(
   ncol = 1, rel_heights = c(0.9, 0.9, 0.9, 0.2, 1, 1))
 ggsave('figures/distance-travelled-full.png', distance_full,
        width = 20, height = 20, dpi = 300, bg = 'white')
-
-# for posters
-plot_grid(
-  d_tod_int +
-    facet_wrap(~ species, labeller = label_parsed, nrow = 1) +
-    theme(legend.position = 'none', strip.text = element_text(size = 10)),
-  d_doy_int +
-    facet_wrap(~ species, labeller = label_parsed, nrow = 1) +
-    theme(legend.position = 'none', strip.text = element_text(size = 10)),
-  labels = c('A', 'B'), ncol = 1) %>%
-  plot_grid(
-    .,
-    get_legend(
-      d_tod_int +
-        scale_fill_gradientn(name = expression(atop(bold('Relative'),
-                                                    bold('change'))),
-                             colors = d_pal,
-                             limits = range(z_breaks) * 2,
-                             breaks = z_breaks * 2,
-                             labels = \(x) round(2^x, 2)) +
-        theme(panel.background = element_rect(fill = 'grey90'),
-              legend.position = 'right',
-              legend.key.width = rel(1),
-              legend.justification = 'center',
-              legend.direction = 'vertical')),
[…]

### 2025-06-10 writing/appendix-a-cleaning-telemetry-data.Rmd
changed appendix to use elk

@@
 ---
-title: "\\Large Predicting how climate change will affect how and where terrestrial mammals will move"
+title: "\\Large The heat is on: Rising temperatures alter when, how, and where large terrestrial mammals move"
 author:
@@
 header-includes:
-    - \renewcommand{\figurename}{Figure S\!\!} # for "Figure Ax.
+    - \renewcommand{\figurename}{Figure A\!\!} # for "Figure Ax.
     - \usepackage{setspace}\doublespacing # for double-spaced text
@@
 In this appendix, we illustrate the cleaning procedure for the telemetry data.
-For the sake of brevity, we only show the code and outputs for *Canis lupus*; the full script is available at  [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R).
+For the sake of brevity, we only show the code and outputs for some elk (*Cervus canadensis*); the full script is available at  [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R).
 We suggest looking at the script since this appendix does not include the interactive maps that are generated by `plot_adj(..., map = TRUE)`.
@@
 Therefore, we have left the function calls in the document for completeness, but we have commented them out to prevent them from being executed.
-Below is an example of a map with some locations for wolf BW008, as generated by running\newline`plot_adj('BW008', max_speed = 3.5, map = TRUE)`.
+Below is an example of a map with some locations for elk E003, as generated by running\newline`plot_adj('E006', max_speed = 0.4, max_angle = 170, map = TRUE)`.
 
@@
 ```{r map-example, echo=FALSE, out.height='40%'}
-knitr::include_graphics(path = 'H:/GitHub/bc-mammals-temperature/figures/BW008-map-example.jpg',
+knitr::include_graphics(path = 'H:/GitHub/bc-mammals-temperature/figures/E006-map-example.png',
                         rel_path = FALSE)
@@
 We removed outlier points using `flag_outlier(..., value = 1)`, which leverages `ctmm::as.telemetry()` to remove rows that have `TRUE` or a positive, non-zero number in the `outlier` column.
-See te comments in each code chunk for additional detail and considerations about individual data points.
+See the comments in each code chunk for additional detail and considerations about individual data points.
 
@@
 
+\doublespacing
+
 Now that we have organized the full telemetry dataset, we can start cleaning the data.
+Locations are candidates for removal if they are far from the median location (i.e., large median deviation) or high minimum speed between locations.
+The minimum speed is calculated as the distance between locations divided by time between locations, since traveling in a straight line is the fastest possible path.
+To search for potential outliers, the `check_animal()` function creates five plots: The top plot shows the telemetry data with an appropriate two-point equidistant projection, and it can help detect any migrations or range shifts.
+The second plot, created by `ctmm::outlie(tel)` [where `tel` is a `telemetry` object, see @calabrese_ctmm_2016] shows the telemetry data using a two-point equidistant projection.
+Points are larger and redder if they are further from the median location, while segments are thicker and bluer if the minimum speed is large.
+Points that are near the median location and segments with small minimum speed may not be visible, while severe location errors will often be identifiable due to their very red points and thick blue segments (Fig.
+\@ref(fig:outlie)).
+The third plot, created by `plot(ctmm::outlie(tel))`, shows the minimum speed against median deviation.
+This plot is simply an alternative visualization of the second plot (without the spatial information).
+The fourth plot shows minimum speed against turning angle, since severe location errors will have high minimum speed and very sharp (i.e., large) turning angles (Fig.
+\@ref(fig:outlie)).
+The fifth plot shows minimum speed against sampling interval, which can be helpful to detect points that imply excessively high minimum speeds for long periods of time.
+Generally, animals tend to move faster (for long periods of time) when moving in a straight line (i.e., ballistic movement as opposed to exploratory movement).
 
-\clearpage
+```{r outlie, fig.width=6, fig.height=6, echo=FALSE, fig.cap="Example output from $\texttt{ctmm::outlie()}$ using randomly generated bivariate Gaussian locations (dots) and a purpousefully erroneous location.
+Dots are larger and redder if they are further from the median location, and segments are thicker and bluer if the estimated straight-line displacement is high.
+Since the erroneous location is so far from the median location, both median deviation and minimum speed are quite high, so it is immediately visible.
+Note that the turning angle is large because the data implies a near-180-degree turn."}
+set.seed(2)
+tel <- data.frame(timestamp = as.POSIXct(1:100),
+                  long = (rnorm(100) / 1e3),
+                  lat = (rnorm(100) / 1e3))
+tel[30, 2:3] <- c(12, 1) / 1e3
+tel <- as.telemetry(tel)
+o <- outlie(tel)
+rm(tel, o)
+```
+
+Ideally, the second plot (center-left) should show a good mix of points and lines of different sizes, with no points or lines clearly standing out.
+Plots 3-5 should show a decrease in $y$ values (vertical axis) as the $x$ values (horizontal axis) increase.
+The intensity of the decrease will depend on a series of factors, including the animal's species and age.
+See elk E006 below for an example of a telemetry dataset with a clear GPS error that exhibits
+
+In the two bottom figures produced by `check_animal()`, where minimum speed is plotted against turning angle and sampling interval, high speeds should occur at smaller turning angles and finer sampling intervals.
+High speeds at large turning angles and sampling intervals are indicative of GPS errors, as in the case for animal E006.
+E002 and E003 do not show any clear issues.
+E002 has two points with high speeds and large turning angles that are worth checking, but plotting the adjacent telemetry points with `plot_tel()` indicates there are no issues.
+It is worth noting that all speed-based inferences are only appropriate if sampling frequency is fine enough to reconstruct at least an approximate movement path [@noonan_scale-insensitive_2019; @nathan_big-data_2022].
+The example data in Fig.
+\@ref(fig:outlie) would be too coarse to make any inferences on minimum speed because there are no clear paths in the data: the animal seems to move around randomly (which is unsurprising, since the locations are random i.i.d.
+bivariate Gaussian draws).
+
+<!-- \clearpage -->
+
+\singlespacing
 
@@
 
-```{r canis-lupus, fig.pos='!H', fig.dim=c(10, 12)}
-# BW008
-out <- check_animal('BW008')
-plot_adj('BW008', max_speed = 3.5)
-# plot_adj('BW008', max_speed = 3.5, map = TRUE)
-plot_adj('BW008', max_angle = 178, max_speed = 1) # using a linear feature
-plot_adj('BW008', max_angle = 90, max_speed = 2) # realistic movement
-plot_adj('BW008', max_angle = 170, max_speed = 0.5) # fine sampling
+```{r update-opts, include=FALSE}
+# update chunk defaults
+knitr::opts_chunk$set(echo = TRUE, out.width = '100%',
+                      fig.align = 'center', cache = TRUE, warning = FALSE,
+                      message = FALSE,
+                      root.dir = 'H:/GitHub/bc-mammals-temperature')
+```
+
+```{r E002, fig.pos = '!H', fig.dim = c(10, 10)}
+# E002
+out <- check_animal('E002')
+plot_adj('E002', max_speed = 0.4)
+plot_adj('E002', max_angle = 170, max_speed = 0.2)
+```
+
+```{r E003, fig.pos = '!H', fig.dim = c(10, 10)}
+# E003
+out <- check_animal('E003')
+plot_adj('E003', max_speed = 0.5)
+```
+
+```{r E006, fig.pos = '!H', fig.dim = c(10, 10)}
+# E006
+out <- check_animal('E006')
+plot_adj('E006', max_speed = 0.8) # ok
+plot_adj('E006', max_speed = 0.4, max_angle = 170) # outlier
+plot_adj('E006', max_speed = 0.4, max_angle = 170, map = FALSE) # see river
+# dt is much larger: collar struggled to receive signal with tree cover
+'hours' %#% out[which(out$speed > 0.4 & out$angle > 170) + (-4:4), 'dt']
+'hours' %#% c(43282, 7254) # convert seconds to hours
+flag_outlier('E006', max_speed = 0.4, max_angle = 170, value = 1)
+out <- check_animal('E006')
+plot_adj('E006', max_speed = 0.2, max_angle = 170) # ok
+```
+
+```{r E011, fig.pos = '!H', fig.dim = c(10, 10)}
+# E011: last point is outlier (keeping previous points bc many 0 speeds)
+out <- check_animal('E011')
+plot_adj('E011', max_speed = 0.4, n_adj = 10)
+plot_adj('E011', max_speed = 0.4, n_adj = 10, map = FALSE)
+plot_adj('E011', max_speed = 0.4, n_adj = 9, reset_layout = FALSE)
+which(out$speed > 0.4) == nrow(out)
+i_tel <- which(d$animal == 'E011')
+i <- (nrow(d$tel[[i_tel]]) - 7):nrow(d$tel[[i_tel]]) # drop last 8 points
+points(location.lat ~ location.long, d$tel[[i_tel]][i, ], col = 'blue',
+       cex = 2, pch = 19)
+d$tel[[i_tel]][i, 'outlier'] <- 1
+points(location.lat ~ location.long, d$tel[[i_tel]][i, ],
+       col = as.numeric(d$tel[[i_tel]]$outlier[i]) + 1, cex = 2)
+layout(1)
+
+out <- check_animal('E011')
+tel <- as.telemetry(d$tel[[i_tel]], mark.rm = TRUE)
+plot(tel[nrow(tel) + -10:0, ], error = FALSE, type = 'b',
+     pch = as.character(0:9))
+plot_adj('E011', max_speed = 0.25)
+```
+
+```{r E013, fig.pos = '!H', fig.dim = c(10, 10)}
+# E013
+out <- check_animal('E013')
+plot_adj('E013', max_speed = 0.2, max_angle = 170) # escaped something?
+```
+
+```{r E015, fig.pos = '!H', fig.dim = c(10, 10)}
+# E015 has multiple GPS malfunctions
+out <- check_animal('E015')
+plot_adj('E015', max_speed = 1, reset_layout = FALSE)
+# can't filter easily because the outliers don't have the max velocity
+i <- d %>%
+  pull(tel) %>%
+  nth(which(d$animal == 'E015')) %>%
+  mutate(i = 1:n()) %>%
+  slice(which(timestamp ==
+                as.POSIXct(out[which(out$speed > 0.6)[2], 't'])) +
+          0:1) %>%
+  pull(i) %>%
+  first() + 0:10
+points(location.lat ~ location.long, cex = 2, lwd = 3, col = 'blue',
+       filter(d, animal == 'E015')$tel[[1]][i, ])
+lines(location.lat ~ location.long,
+      filter(d, animal == 'E015')$tel[[1]][i, ], lwd = 6, col = 'red3')
+d$tel[[which(d$animal == 'E015')]][i, 'outlier'] <- 1
+out <- check_animal('E015')
+plot_adj('E015', max_speed = 0.60) # track looks ok now
+```
+
+```{r E016, fig.pos = '!H', fig.dim = c(10, 10)}
+# E016
+out <- check_animal('E016')
+plot_adj('E016', max_speed = 0.2, max_angle = 170) # one outlier
+plot_adj('E016', max_speed = 0.2, max_angle = 179) # use higher upper bound
+flag_outlier('E016', max_speed = 0.2, max_angle = 179, value = 1)
+out <- check_animal('E016')
+plot_adj('E016', max_speed = 0.2, max_angle = 170) # ok
+```
+
+```{r E017, fig.pos = '!H', fig.dim = c(10, 10)}
+# E017
+out <- check_animal('E017')
+plot_adj('E017', max_speed = 0.6) # ok
+plot_adj('E017', max_angle = 170, max_speed = 0.2)
+```
+
+```{r E019, fig.pos = '!H', fig.dim = c(10, 10)}
+# E019
+out <- check_animal('E019') # many high speeds at angle > 170
+plot_adj('E019', max_speed = 0.65, max_angle = 135) # some outliers
+plot_adj('E019', max_speed = 0.65, max_angle = 160, reset_layout = FALSE)
+i <- which(out$speed > 0.65 & out$angle > 160)[c(1, 3)] # some outliers
+points(location.lat ~ location.long,
+       d$tel[[which(d$animal == 'E019')]][i, ], cex = 2, col = 'blue')
+d$tel[[which(d$animal == 'E019')]][i, 'outlier'] <- 1
+
+out <- check_animal('E019') # check other high speeds with angle > 160
+plot_adj('E019', max_speed = 0.4, max_angle = 135, reset_layout = FALSE)
+tel <- d$tel[[which(d$animal == 'E019')]]
+i <- which(out$speed > 0.4 & out$angle > 135)
+points(location.lat ~ location.long, cex = 1.5, col = 'black',
+       tel[! tel$outlier, ][i, ],
+       pch = as.character(0:(length(i) - 1)))
+'hour' %#% out[i, 'dt'] # consistent sampling
+
+# 0 is likely an error (erring on the side of caution)
+# 1 is consistent with other movement
+# 2 is likely an error
+# 3 is consistent with other movement
+# 4 is likely an error
+# 5 is realistic
+# 6 is likely an error
+# 7 is realistic
+# 8 is likely an error
+# 9 is likely an error
+i <- i[1 + c(0, 2, 4, 6, 8, 9)]
+i_id <- which(d$animal == 'E019')
+d$tel[[i_id]][! d$tel[[i_id]]$outlier, ][i, 'outlier'] <- 1
+# ensure correct locations were removed
+out <- check_animal('E019') # check other high speeds with angle > 160
+plot_adj('E019', max_speed = 0.4, max_angle = 135, reset_layout = FALSE)
 ```
@@
 
-```{r canis-lupus-2, fig.pos='!H', fig.dim=c(10, 12)}
-out[out$angle > 170 & out$speed > 0.5, ] # 15-minute sampling
-```
-
-\clearpage
-
-```{r canis-lupus-3, fig.pos='!H', fig.dim=c(10, 12)}
-# BW009
-out <- check_animal('BW009')
-plot_adj('BW009', max_angle = 179, max_speed = 1) # potential outlier
-plot_adj('BW009', max_angle = 179.7, max_speed = 1, n_adj=200) # uses path
-plot_adj('BW009', max_angle = 175, max_speed = 1, n_adj = 12) # ok
-plot_adj('BW009', max_angle = 170, min_angle = 175, max_speed = 1,
-         n_adj = 20) # ok
-plot_adj('BW009', max_angle = 170, max_speed = 0.5) # reasonable
-
-# BW010
-out <- check_animal('BW010')
-plot_adj('BW010', max_speed = 3.5) # ok
-plot_adj('BW010', max_speed = 1, max_angle = 170) # ok
-plot_adj('BW010', max_speed = 2, max_angle = 135) # ok
-
-# BW012
-out <- check_animal('BW012')
-plot_adj('BW012', max_speed = 3)
-plot_adj('BW012', max_speed = 1, max_angle = 170)
-```
-
-```{r canis-lupus-4, fig.pos='!H', fig.dim=c(5, 6)}
-# location is ok
-tel <- as.telemetry(d$tel[[which(d$animal == 'BW012')]])
-i <- which(out$speed > 2 & out$angle > 170)
-plot(tel, error = FALSE, type = 'l', xlim = c(-2600, 0),
-     ylim = c(300, 2200), col = 'black')
-plot(tel[i + -10:10, ],
-     error = FALSE, type = 'l', col = 'blue', lwd = 2, add = TRUE)
-plot(tel[i, ],
-     error = FALSE, col = 'blue', pch = 19, add = TRUE)
-plot(tel, add = TRUE)
-rm(tel, i)
-```
-
-```{r canis-lupus-5, fig.pos='!H', fig.dim=c(10, 12)}
-# BW013
-out <- check_animal('BW013')
-plot_adj('BW013', max_speed = 1, max_angle = 170) # uses linear features
-
-# BW014 has a clear GPS error
-out <- check_animal('BW014')
-plot_adj('BW014', max_speed = 4)
-flag_outlier(id = 'BW014', max_speed = 4, value = 1)
-out <- check_animal('BW014')
-plot_adj('BW014', max_speed = 2) # ok
-plot_adj('BW014', max_speed = 0.5, min_speed = 0.6, max_angle = 170,
-         n_adj = 30) # seems near a linear feature
-# plot_adj('BW014', max_speed = 0.5, min_speed = 0.6, max_angle = 170,
-#          map = TRUE) # moving along a river
-
-# B027
-out <- check_animal('BW027')
-plot_adj('BW027', max_speed = 3, max_angle = 90) # not problematic
-
-# BW028
-out <- check_animal('BW028')
-plot_adj('BW028', max_speed = 3) # not problematic
-
-# B029
-out <- check_animal('BW029')
-plot_adj('BW029', max_speed = 2, max_angle = 175) # need a finer scale
-plot_adj('BW029', max_speed = 2, max_angle = 175, n_adj = 2) # ok
-
-# B031
-out <- check_animal('BW031')
-plot_adj('BW031', max_speed = 1, max_angle = 170) # need a finer scale
-plot_adj('BW031', max_speed = 1, max_angle = 170, n_adj = 5) # ok
-
-# B033: sharp and fast turns are near linear features
-out <- check_animal('BW033')
-plot_adj('BW033', max_speed = 1.5, max_angle = 90, n_adj = 5)
-
-# BW044nex
-out <- check_animal('BW044nex')
-plot_adj('BW044nex', max_speed = 1, max_angle = 170)
-
-# BW046nex
-out <- check_animal('BW046nex')
-plot_adj('BW046nex', max_speed = 1, max_angle = 175, min_speed = 3) # ok
-plot_adj('BW046nex', max_speed = 1, min_speed = 3, # problematic
-         max_angle = 135, min_angle = 175, reset_layout = FALSE)
-i <- which(out$speed > 1 & out$speed < 3 & # likely hunting
-             out$angle > 135 & out$angle < 175) + 0:5
-points(location.lat ~ location.long,
-       d$tel[[which(d$animal == 'BW046nex')]][i, ], col = 'blue', cex = 2)
-d$tel[[which(d$animal == 'BW046nex')]][i, 'outlier'] <- 1
-plot_adj('BW046nex', max_speed = 1, max_angle = 170) # ok
-```
+# References

### 2025-06-10 functions/outlier_plots.R

@@ -3,14 +3,13 @@ source('functions/find_angle.R')
 outlier_plots <- function(telemetry, units = FALSE, ci_level = 0,
                           return = FALSE, reset_layout = TRUE,
                           cap_dt = TRUE) {
-  layout(matrix(c(1, 1, 2:5), ncol = 2, byrow = TRUE))
+  layout(matrix(c(1, 1, 1, 2, 2, 4, 4, 1, 1, 1, 3, 3, 5, 5), ncol = 2))
   # plot entire telemetry
-  plot(telemetry$x, telemetry$y, type = 'l', xlab = 'x (m)',
-       ylab = 'y (m)')
+  plot(telemetry, error = FALSE, col = 'black', pch = 19, type = 'l')
   # plot of locations and velocities
   out <- outlie(telemetry, plot = TRUE)
   # min speed vs distance from tel center (SI units; no CIs)
-  plot(out, units = units, level = ci_level)
+  plot(out, units = units, level = ci_level, col = '#00000050')
   
   # calculate turning angles in degrees
   out$angle <- find_angle(telemetry$x, telemetry$y, radians = FALSE)

### 2025-06-10 writing/appendix-b-gams.Rmd
corrected "log_{10}" in fig cap; need to knit
moved DAGs to main text and updated figures

@@
                       message = FALSE)
-```
-
-\clearpage
-
-# Directed Acyclical Graphs
-
-```{r hgam-dag,  message=FALSE, warning=FALSE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of temperature (Temp) on probability of moving, speed when moving, or distance traveled ($Y$) while accounting for the effects of time of day (TOD), day of year (DOY), and their interaction effects.
-Temperature directly affects $Y$, but the effects of temperature depend on the time of day and season.
-Time of day and day of year also affect $Y$ directly, but the effect of time of day changes throughout the year due to changes in day length and seasonality.", fig.height=4, fig.width=4, out.width='40%'}
-library('dplyr')   # for data wrangling
-library('ggplot2') # for fancy plots
-library('ggdag')   # for directed acyclical graphs
-
-dagify(Y ~ TOD + DOY + Temp,
-       Temp ~ TOD + DOY,
-       TOD ~ DOY,
-       outcome = 'Y',
-       coords = list(x = c(Y = 0, TOD = -1, DOY = 1, Temp = 0),
-                     y = c(Y = 1.2, TOD = 2, DOY = 2, Temp = 0))) %>%
-  tidy_dagitty() %>%
-  mutate(col = case_when(name == 'Y' ~ '1',
-                         name == 'TOD' ~ '0',
-                         name == 'DOY' ~ '0',
-                         name == 'Temp' ~ '2')) %>%
-  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
-  geom_dag_point(aes(color = col), size = 20, alpha = 0.5) +
-  geom_dag_point(color = 'black', size = 20, shape = 1) +
-  geom_dag_edges() +
-  geom_dag_text(aes(label = name), color = 'black', size = 4.5) +
-  scale_color_manual(values = c('grey', 'darkorange', 'red')) +
-  theme_dag() +
-  theme(legend.position = 'none')
-```
-
-```{r hrsf-dag, message=FALSE, warning=FALSE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of temperature (Temp) on Relative Selection Strength (RSS) for percent forest cover (F), elevation (E), and distance from water (W).
-The RSS for of a given habitat depends on all three resources, and the selection for each resource is independent of the other two resources and dependent on temperature.", fig.height=4, fig.width=4, out.width='40%'}
-library('dplyr')   # for data wrangling
-library('ggplot2') # for fancy plots
-library('ggdag')   # for directed acyclical graphs
-
-dagify(RSS ~ `F` + E + W,
-       `F` ~ Temp,
-       E ~ Temp,
-       W ~ Temp,
-       outcome = 'RSS',
-       coords = list(x = c(RSS = 0, `F` = -1, E = 0, W = 1, Temp = 0),
-                     y = c(RSS = 2, `F` = 1, E = 1, W = 1, Temp = 0))) %>%
-  tidy_dagitty() %>%
-  mutate(col = case_when(name == 'RSS' ~ '1',
-                         name == 'F' ~ '0',
-                         name == 'E' ~ '0',
-                         name == 'W' ~ '0',
-                         name == 'Temp' ~ '2')) %>%
-  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
-  geom_dag_point(aes(color = col), size = 20, alpha = 0.5) +
-  geom_dag_point(color = 'black', size = 20, shape = 1) +
-  geom_dag_edges() +
-  geom_dag_text(aes(label = name), color = 'black', size = 4.5) +
-  scale_color_manual(values = c('grey', 'darkorange', 'red')) +
-  theme_dag() +
-  theme(legend.position = 'none')
 ```
@@
 
-```{r dt-smooths, fig.cap="\\textbf{A.} Histograms of sampling intervals between GPS locations, with counts on a $\\log_{10}$ axis for ease of readability.
+```{r dt-smooths, fig.cap="\\textbf{A.} Histograms of sampling intervals between GPS locations.
 \\textbf{B.} Species-level smooth effects of sampling interval on the estimated probability of the animal moving.
@@
 ```{r avail-used, out.width='95%', fig.cap="Histograms of available (red) and used (blue) resources.
-All GPS fixes counts as one observation.
+Each GPS fix counts as one observation.
 The y axis is on the square-root scale to help visualize values with low counts."}

### 2025-06-10 writing/manuscript.Rmd
addressed some of KH's comments
continue at "Effects of temperature on habitat selection"

@@
 
-```{r, echo=FALSE}
+```{r abstract, echo=FALSE}
 abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on fitness, habitat selection, and community structure.
-Previous studies have quantified the effects of temperature on mammalian movement behaviour, but few have quantified the effects of future climate change.
+Previous studies have quantitatively estimated the effects of temperature on mammalian movement behaviour, but few have estimated the effects of future climate change.
 Consequently, it is still unclear how and where mammals will move in the 21^st^ century.
-We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on the probability of moving, the speed when moving, and the habitat selection of six species throughout British Columbia, Canada.
-We show that temperature is a strong determinant of when, how, and where mammals move, and that climate change will impact mammals' movement rates and habitat selection throughout the 21^st^ century.
-We found no common effects of temperature on the species’ movement rates, so we suggest that species' estimated responses be interpreted jointly when making inferences about how climate change will impact ecological communities.
-In contrast, habitat selection strength decreased for all species within their current ranges, suggesting that all six species will shift their range in the coming decades.
+We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on the probability of moving, the speed when moving, and the habitat selection of six species throughout British Columbia, Canada, between 1998 and 2023.
+We show that temperature is a strong determinant of when, how, and where mammals move, and we predict that climate change will impact mammals' movement rates and habitat selection as temperatures warm throughout the 21^st^ century.
+We found the effects of temperature on movement rates varied across species, so we suggest that species' estimated responses not be interpreted individually when making inferences about how climate change will impact ecological communities.
+In contrast, habitat selection strength decreased for all species within their observed ranges, suggesting that all six species may shift their range in the coming decades.
 As climate change exposes mammals to novel environmental conditions, predicting changes in animal behavior and community structure will become crucial for effective and proactive conservation.
@@
 
-\noindent \textbf{Figures}: 7
+\noindent \textbf{Figures}: 8
 
-\noindent \textbf{Tables}: 2
+\noindent \textbf{Tables}: 3
 
@@
 Animals move to maximize resource availability [e.g., food, water, shelter, see: @charnov_optimal_1976; @kacelnik_ideal_1992], optimize habitat use [@winter_forecasting_2024], and escape predation [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024].
-However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], the ease of movement [e.g., snow depth and density, see @montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023].
+However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023].
 Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [e.g., @hall_microrefuges_2016].
-Additionally, temperature also affects movement indirectly through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010].
+Additionally, temperature also has indirect effects on movement through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010], snow depth and density [@montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the frequency and intensity of extreme events [@zhang_increased_2023; @bunnell_global_2011], including forest fires, droughts, and novel phenology [@zurowski_summer_2023; @duncan_predicting_2012; @bunnell_global_2011; @merkle_large_2016; @slatyer_ecological_2022].
 Changes in temperature thus affects how animals move [e.g., movement frequency and speed -- their movement behaviour: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].
 
-Temperature also affects the consequences of animal movement, such as encounter rates with resources [@mezzini_how_2025], other animals [@martinez-garcia_how_2020], and humans [@weststrate_how_2024], and these changes are exacerbated at extreme temperatures [@wilson_stenotherms_2001; @berger_climate_2018; @giroux_activity_2023].
-But what counts as "extreme" varies across species: thermal specialists are more sensitive to small changes outside their typical range [e.g., @giroux_activity_2023], while thermal generalists can adapt more easily to change [@dupont_beyond_2024].
+<!-- since movement depends on temperature, the consequences of movement do, too -->
+
+By impacting animal movement behavior, temperature also affects the consequences of movement, such as encounter rates with resources [@mezzini_how_2025], other animals [@martinez-garcia_how_2020], and humans [@weststrate_how_2024], and these changes are exacerbated at extreme temperatures [@wilson_stenotherms_2001; @berger_climate_2018; @giroux_activity_2023].
+But what counts as "extreme" varies across species: Species that experience narrow temperature ranges tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while thermal generalists can adapt more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024].
 However, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019], affect many factors [e.g., temperature, phenology, and resource availability: @polazzo_measuring_2024], or are too rapid to respond within the span of a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].
@@
 
-In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures and severe forest fires [@ipcc_climate_2023; @zurowski_summer_2023].
+In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@ipcc_climate_2023].
 Over the next 100 years, these changes are expected to worsen and continue to impact mammals' fitness, movement behaviour, and habitat selection as animals cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and more frequent and intense extreme events [@bunnell_global_2011], along with increasing anthropogenic pressure [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024].
 Consequently, it remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially when populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018].
-Uncertainty in future conditions and their consequences on ecosystems present a need for a comprehensive understanding of (1) the direct effects of temperature on animal movement behaviour and (2) the implications this will have on populations and species' adaptability and resiliency, community structure, and biodiversity at large.
-Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], movement behaviour [@melin_effect_2023], thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [@mccain_assessing_2019], but it is still unclear how mammals will adapt their small-scale movement and habitat selection to changes in temperature during the 21^st^ century [@ipcc_climate_2023; but see @hetem_responses_2014; @winter_forecasting_2024].
+Uncertainty in future conditions and their consequences on ecosystems present a need for a comprehensive understanding of (1) the direct effects of temperature on animal movement behaviour and (2) the implications such changes and uncertainty will have on populations and species' adaptability and resiliency, as well as overall community structure.
+Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], movement behaviour [@melin_effect_2023], thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [e.g., body size, activity time, and elevational and latitudinal distribution @mccain_assessing_2019].
+However, it is still unclear how mammals will adapt their small-scale movement and habitat selection to changes in temperature during the 21^st^ century [@ipcc_climate_2023; but see @hetem_responses_2014; @winter_forecasting_2024].
 
@@
 
-Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to be able to prepare for and respond to change in a proactive manner [e.g., @sawyer_identifying_2009; @mccain_assessing_2019].
-The Canadian province of British Columbia is currently experiencing rapid widespread warming [@dierauer_climate_2021; @turner_its_2009], phenological shifts [e.g., @tysor_countergradient_2025; @basu_phenological_2024], and more frequent and intense extreme events [e.g., @zhang_increased_2023], including forest fires [@zurowski_summer_2023].
+Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to respond to change in a proactive manner [@sawyer_identifying_2009; @mccain_assessing_2019].
+The Canadian province of British Columbia is currently experiencing rapid widespread warming [@dierauer_climate_2021; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024], and more frequent and intense extreme events [@zhang_increased_2023], including forest fires [@zurowski_summer_2023].
 As we approach the deadline for the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022], decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness.
-Predicting how climate change will affect how, when, and where animals will move is necessary for identifying which landscapes are most valuable now as well as in the future.
+Predicting how climate change will affect how, when, and where animals will move is necessary for identifying landscape dynamics and value will evolve throughout the century.
 Failing to understand how changes in temperature will affect mammalian movement behaviour and habitat selection will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.
@@
 
-This paper provides an analysis of the effects of air temperature on the movement of six terrestrial mammal species in British Columbia (BC), Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, pumas: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
-Using over 25 years of telemetry data thoughout a large spatial range of British Columbia and hourly air temperature data, we estimate how mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature.
-We then pair the estimated responses with climate change projections to forecast the species' movement during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSPs -- see @riahi_shared_2017].
-We disentangle the direct and indirect effects of temperature on the movement behaviour of terrestrial mammals and discuss the consequences of these changes at the individual, population, and landscape levels.
+This paper provides a framework for quantifying the effects of proximal air temperature on animal movement using quantitative methods that are flexible and statistically robust.
+We present results for six large-bodied terrestrial mammal species in British Columbia (BC), Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
+Using over 25 years of telemetry data throughout a large spatial range of British Columbia and hourly air temperature data, we estimate how these mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature.
+We then pair the estimated responses with climate change projections to forecast the species' movement during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017].
+We disentangle the direct and indirect effects of temperature on the movement behaviour of terrestrial mammals and discuss the consequences of these changes at the individual, population, and community levels.
 
@@
 \@ref(fig:bc-map)); (2) historical hourly reanalyzed air temperature from the ERA5 dataset from the European Center for Medium-range Weather Forecasting; (3) resource rasters of percent forest cover, elevation, and distance from water; and (4) climate change projections under four Shared Socioeconomic Pathways [SSPs, see @riahi_shared_2017].
-We detail the data sources and analyses in the sections below.<!--, and more details are available in appendices A (estimating how temperature affects movement frequency and speed when moving), B (fitting Hierarchical Resource Selection Functions), and C (simulating hourly temperature data to 2100 using the four SSPs).-->
+While we acknowledge that forest type, structure, and age are important variables, we do not include such variables in this study for the sake of parsimony and comparability across species.
+We detail the data sources and analyses in the sections below.
 
@@
 ***MISSING INFO ON OTHER TELEMETRY DATA.*** From the full set of telemetry data, a total of 2396 GPS locations (0.16%, including collar calibration data) were removed using diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points.
-Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated stationary behaviour.
-See Appendix A for additional information.
+Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated stationary behaviour (Appendix A).
 
-### Historical and projected temperature data
+### Historical temperature data and temperature projections
 
-\noindent Rasters of hourly reanalyzed air temperature data were downloaded from the ERA5 dataset [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)).
+\noindent Rasters of hourly reanalyzed air temperature data were downloaded from the ERA5 database [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)).
 Proximal temperature was estimated for each location by extracting the value from the corresponding raster cell from the temporally nearest raster using the `extract()` function from the `terra` package [version 1.7-46, @hijmans_terra_2023] for `R` [@r_core_team_r_2024].
@@
 We obtained rasters of projected monthly average temperature at a 0.08$^\circ$ resolution in BC from 2020 to 2100 under the different SSPs via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`.
-Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 data for BC from 1998 to 2023 [extremes included, see @hersbach_era5_2023] to calculate within-month variance in temperature, which we defined as the variance within a given pixel, month, and year.
-We then modeled the estimated variance using a GAM for location and scale [GAMLS, see: @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017].
+Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 data for BC from 1998 to 2023 [extremes included, @hersbach_era5_2023] to calculate within-month variance in temperature, which we defined as the variance within a given pixel, month, and year.
+We then modeled the estimated variance using a GAM for location and scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017].
 The linear predictor for the location (i.e., the mean) included smooth terms of the within-pixel monthly mean temperature (within each year), month (as a cyclic smooth), a two-dimensional smooth of space, and a tensor interaction product term of space and month.
 The linear predictor for the scale term (which governs the mean-variance relationship) included smooth terms of the monthly mean, month, and space.
-We did not include a smooth of year to avoid unrealistic projections when extrapolating beyond the range of historical data.<!-- Additional details are available in Appendix C.-->
+We did not include a smooth of year to avoid unrealistic projections when extrapolating beyond the range of historical data.
 
@@
 We simulated hourly variation in future weather by assuming hourly temperature followed a normal distribution with mean specified by the monthly `climatenaR` climate projections and variance as specified by the Gamma GAMLS.
-We then predicted changes in movement behaviour and habitat selection as a function of the simulated temperature using the HGAMs and HRSFs.
-For each month within each year from 2020 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantile by increments of 0.1, and we weighted each quantile proportionally to the (normalized) Gaussian probability density for each quantile<!-- (Appendix C)-->.
+For each month within each year from 2020 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantile by increments of 0.1, and we weighted each quantile proportionally to the (normalized) Gaussian probability density for each quantile.
 
@@
 \noindent We modeled each animal's movement using continuous-time movement models [@fleming_fine-scale_2014] via the `ctmm` package [version 1.2.0, @fleming_ctmm_2023] for `R`.
-We then estimated mammals' instantaneous speeds by applying the `ctmm::speeds()` function on all models with finite speed estimates [415 of 433, see @fleming_fine-scale_2014; @noonan_scale-insensitive_2019].
-The 18 animals with insufficiently fine sampling were for one mountain goat, 15 boreal caribou, and 2 southern mountain caribou (Table \@ref(tab:data-table)).
+We then estimated mammals' instantaneous speeds by applying the `ctmm::speeds()` function on all models with finite speed estimates [415 of 433, @fleming_fine-scale_2014; @noonan_scale-insensitive_2019].
+The 18 animals with insufficiently fine sampling to support movement models with finite speed [@noonan_scale-insensitive_2019] were for one mountain goat, 15 boreal caribou, and 2 southern mountain caribou (Table \@ref(tab:data-table)).
 
@@
 The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R).
-To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using the inflection point of each species' histogram of detrended speed estimates, as $k$-nearest neighbors did not discriminate between states well (Fig.
-B3).
+To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using a $k$-means algorithm with 2 clusters for each species' detrended speeds.
+When the algorithm clearly failed to discriminate between states, we estimated the split point using the inflection points in histograms of the detrended speeds (Fig.
+B1).
 
@@
 To assess the importance of including temperature as an explicit covariate (rather than including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the fits of the two sets of models.
-We provide the relevant Directed Acyclical Graphs in the Appendix (Figs.
-B1-B2).
 
@@
 
-\noindent In this study, we separate the effects of temperature on mammalian movement into indirect and direct effects.
-We call "indirect" all effects that can be intervened on by altering an animal's habitat [its geographic space *sensu* @matthiopoulos_defining_2022], and we refer to all other effects as "direct".
-For example, an increase in temperature may push an animal to prefer higher forest cover in search of shade, and conservationists can thus mitigate these indirect effects of temperature on movement behaviour by increasing the availability of forested areas, since the effect of temperature on movement is conditional on forest cover availability.
-In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022], so it is not possible to mitigate the effects of temperature on movement behaviour (other than changing temperature directly).
-We provide the relevant Directed Acyclical Graphs in the Appendix (Figs.
-B1-B2).
+\noindent In this study, we separate the effects of temperature on mammalian movement into direct and indirect effects.
+We call "direct" all effects that impact movement behavior causally without first impacting another variable (Fig.
+\@ref(fig:dags)).
+For example, in Fig.
+\@ref(fig:dags)A, whether a grizzly bear is moving in a given moment directly depends on time of day and day of year, since it will not be moving when sleeping or hibernating.
+However, time of day and day of year also impact its movement behavior indirectly because its sensitivity to temperature will also depend on time of day and day of year.
+Thus, both time of day and day of year also impact movement frequency indirectly by altering the effect of temperature on movement frequency.
+Failing to disentangle the effects of these three variables will lead to incorrect and likely confusing estimates for the causal effect of temperature on movement behavior, since causal and non-causal relationships would be indistinguishable [@mcelreath_statistical_2020].
+
+Indirect effects of temperature also include all effects that can be controlled by altering an animal's habitat [its geographic space *sensu* @matthiopoulos_defining_2022].
+For example, in Fig.
+\@ref(fig:dags)B, an increase in temperature may push an animal to prefer higher forest cover in search of shade, and conservationists can mitigate these indirect effects of temperature by increasing the availability of forested areas, since the effect of temperature on movement is conditional on forest cover availability.
+In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022].
+In this case, it is not possible to mitigate the effects of temperature on movement behaviour (other than changing temperature directly).
+
+```{r dags,  message=FALSE, warning=FALSE, fig.cap="Directed Acyclical Graphs (DAGs) assumed for inferring the causal effects of temperature (Temp) on movement behavior.
+(A) DAG for the probability of moving, speed when moving, or distance traveled ($Y$) while accounting for the effects of time of day (TOD), day of year (DOY), and their interaction effects.
+Temperature directly affects $Y$, but the effects of temperature depend on the time of day and season.
+Time of day and day of year also affect $Y$ directly, but the effect of time of day changes throughout the year due to changes in day length and seasonality.
+(B) DAG for Relative Selection Strength (RSS) for percent forest cover (F), elevation (E), and distance from water (W).
+The RSS for of a given habitat depends on all three resources, and the selection for each resource is independent of the other two resources and dependent on temperature.", fig.height=4, fig.width=9, out.width='\\linewidth'}
+library('dplyr')   # for data wrangling
+library('ggplot2') # for fancy plots
+library('ggdag')   # for directed acyclical graphs
+library('cowplot') # for plots in grids
+
+plot_grid(
+  dagify(Y ~ TOD + DOY + Temp,
+         Temp ~ TOD + DOY,
+         TOD ~ DOY,
+         outcome = 'Y',
+         coords = list(x = c(Y = 0, TOD = -1, DOY = 1, Temp = 0),
+                       y = c(Y = 1.2, TOD = 2, DOY = 2, Temp = 0))) %>%
+    tidy_dagitty() %>%
+    mutate(col = case_when(name == 'Y' ~ '1',
+                           name == 'TOD' ~ '0',
+                           name == 'DOY' ~ '0',
+                           name == 'Temp' ~ '2')) %>%
+    ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
+    geom_dag_point(aes(color = col), size = 20, alpha = 0.5) +
+    geom_dag_point(color = 'black', size = 20, shape = 1) +
+    geom_dag_edges() +
+    geom_dag_text(aes(label = name), color = 'black', size = 4.5) +
+    scale_color_manual(values = c('grey', 'darkorange', 'red')) +
+    theme_dag() +
+    theme(legend.position = 'none'),
+  dagify(RSS ~ `F` + E + W,
+         `F` ~ Temp,
+         E ~ Temp,
+         W ~ Temp,
+         outcome = 'RSS',
+         coords = list(x = c(RSS = 0, `F` = -1, E = 0, W = 1, Temp = 0),
+                       y = c(RSS = 2, `F` = 1, E = 1, W = 1, Temp = 0)))%>%
+    tidy_dagitty() %>%
+    mutate(col = case_when(name == 'RSS' ~ '1',
+                           name == 'F' ~ '0',
+                           name == 'E' ~ '0',
+                           name == 'W' ~ '0',
+                           name == 'Temp' ~ '2')) %>%
+    ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
+    geom_dag_point(aes(color = col), size = 20, alpha = 0.5) +
+    geom_dag_point(color = 'black', size = 20, shape = 1) +
+    geom_dag_edges() +
+    geom_dag_text(aes(label = name), color = 'black', size = 4.5) +
+    scale_color_manual(values = c('grey', 'darkorange', 'red')) +
+    theme_dag() +
+    theme(legend.position = 'none'), labels = 'AUTO', nrow = 1)
+```
 
@@
 
-\noindent We estimated the effects of temperature ($^\circ$C) on mammals' movement state (moving or not) and speed when moving using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019 and the code chunk below] with the `mgcv` package for `R` [version 1.9-1, @wood_generalized_2017].
+\noindent We estimated the effects of temperature ($^\circ$C) on mammals' instantaneous movement state (moving or not) and speed when moving using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019 and the code chunk below] with the `mgcv` package for `R` [version 1.9-1, @wood_generalized_2017].
 The first HGAM estimated the probability that an animal was moving, $P(M)$, with a binomial family of distributions and logit link function.
@@
 The second HGAM estimated an animal's speed when moving (`speed`) with a gamma family of distributions and log link function.
+
 The HGAMs included random intercepts for each animal (`s(animal, bs = 're')`), fixed-effect intercepts for each species (`species`), and species-level `by` smooths (`s(..., by = species)`), which allowed each species' term to be estimated independent of other species [see model I in Figure 4 of @pedersen_hierarchical_2019].
 The `by` smooths accounted for trends in time of day (in Pacific Daylight Time; `tod_pdt`), day of year (`doy`), and temperature (`temp_c`).
-Additionally, the models had three tensor product interaction terms (`ti()`) `by` each species: (1) day of year and time of day, (2) temperature and time of day, and (3) temperature and day of year.
+The models also had three tensor product interaction terms (`ti()`) `by` each species: (1) day of year and time of day, (2) temperature and time of day, and (3) temperature and day of year.
 These three terms accounted for: (1) seasonal changes in day length, (2) hourly changes in the response to temperature (e.g., changes in nocturnality), and (3) seasonal changes in the response to temperature (e.g., changes in coats and migration timing).
@@
 A global term of `log(dt)` accounted for the overall effect of sampling interval, while a factor-smooth interaction term (`s(log(dt), species, bs = 'fs')`) allowed for species-level deviations from the global term while assuming a common smoothness parameter across species [see model GS in Figure 4 of @pedersen_hierarchical_2019].
-The HGAMs accounted for the cyclicity of time of day and day of year using cyclic cubic splines [`bs = 'cc'`$;$ see p.
+
+The HGAMs accounted for the cyclicity of time of day and day of year using cyclic cubic splines [`bs = 'cc'`, see p.
 202 of @wood_generalized_2017].
+We fit the models with fast Restricted Maximum Likelihood (`'fREML'`) and discretized covariates (`discrete = TRUE`) to optimize computational efficiency with no appreciable losses to model performance [Appendix B$;$ @wood_generalized_2015; @wood_generalized_2017-1; @li_faster_2020].
 Together, the binomial HGAM and the gamma HGAM inform us on an animal's long-term average speed, since it is the product of the probability of moving and its average speed when moving.
-We fit the models with fast Restricted Maximum Likelihood (`'fREML'`) and discretized covariates (`discrete = TRUE`) to optimize computational efficiency with no appreciable losses to model performance [@wood_generalized_2015; @wood_generalized_2017-1; @li_faster_2020].
-Additional details are provided in Appendix B.
 
@@
 
-\noindent We estimated the effects of temperature on each species' selection for percent forest cover (`forest_perc`), elevation (`elevation_m`), and distance from water (`dist_water_m`) by fitting a Hierarchical Resource Selection Function for each species using an HGAM with a Poisson family of distributions and log link function [Appendix B, also see @aarts_estimating_2008].
-We accounted for the spatiotemporal autocorrelation in the telemetry locations by weighting each point based on the telemetry's Autocorrelated Kernel Density Estimate [@fleming_new_2017; @noonan_comprehensive_2019] to produce estimates of second- and third-order habitat selection [*sensu* @johnson_comparison_1980].
-While we recognize there are other important drivers of habitat selection (e.g., forest age, forest type, prey availability), we decided to only use these three proxies to produce results that are comparable between species and to make province-wide predictions simpler.
+\noindent We estimated the effects of temperature on each species' selection for percent forest cover (`forest_perc`), elevation (`elevation_m`), and distance from water (`dist_water_m`) by fitting a Hierarchical Resource Selection Function for each species using an HGAM with a Poisson family of distributions and log link function [Appendix B$;$ @aarts_estimating_2008].
+We accounted for the spatiotemporal autocorrelation in the telemetry locations by weighting each point based on the telemetry's Autocorrelated Kernel Density Estimate [@fleming_new_2017; @noonan_comprehensive_2019] to produce estimates of second- and third-order habitat selection [@johnson_comparison_1980].
+While we recognize there are other important drivers of habitat selection (e.g., forest age, forest type, terrain ruggedness, prey availability), we decided to only use these three proxies to produce results that are easily comparable across species.
 Each species' model had the same structure:
@@
 Finally, we included marginal smooth terms of temperature to account for species- and individual-level sampling biases over temperature (e.g., sampling more during warm periods).
-Detections were weighted proportionally to their degree of independence from other temporally proximate detections [`weights = weight` -- see Appendix B and @alston_mitigating_2022], while quadrature points had a weight of 1.
+Detections were weighted proportionally to their degree of independence from other temporally proximate detections [`weights = weight` -- Appendix B$;$ @alston_mitigating_2022], while quadrature points had a weight of 1.
 Quadrature points were obtained using all raster cells in the 99.9% AKDE percentile.
 The number of quadrature locations greatly outnumbered the number of observed locations (Fig.
-B13), especially after accounting for weighting based on the degree of autocorrelation and the number of home range crossings (Fig.
-B14).
+B11), especially after accounting for weighting based on the degree of autocorrelation and the number of home range crossings (Fig.
+B12).
 
@@
 
-\noindent Data were relatively scarce for temperatures lower than $-20^\circ$C or above $20^\circ$C (Fig.
-B4).
-Species differed in overall mean probabilities of movement (range: 0.05 -- 0.3), mean speed when moving (range: 0.42 -- 2.7 km/day), and mean distance traveled (range: 0.04 -- 0.6 km/day; Table \@ref(tab:gam-intercepts)).
-Grizzly bears tended to move least often (probability of moving $P(M) \approx 0.05$), while wolves and pumas moved most often ($P(M) \gtrapprox 0.23$).
-When moving, mountain goats and southern mountain caribou moved the slowest ($\mathbb E(S|M) \approx 0.4$ m/s), while wolves had the highest mean speed when moving ($\mathbb E(S|M) \approx 2.7$ m/s).
-Consequently, wolves traveled, on average, 50 km/day -- 2.5 to 16.7 times further than other mammals.
+\noindent Overall, 2.6% of GPS locations had temperatures lower than $-20^\circ$C, while 6.5% had temperatures above $20^\circ$C, and temperature ranges differed across species (Table \@ref(tab:temp-qs), Fig.
+B2).
+At $0^\circ$C, species differed in overall mean probabilities of movement (range: 0.05 -- 0.31), mean speed when moving (range: 0.42 -- 2.67 km/h), and mean distance traveled (i.e., $P(M) \times \mathbb E(S)$, range: 0.04 -- 0.61 km/h; Table \@ref(tab:gam-intercepts)).
+Grizzly bears tended to move least often ($P(M) \approx 0.05$), while wolves and cougars moved most often ($P(M) \ge 0.22$).
+When moving, mountain goats and southern mountain caribou moved the slowest ($\mathbb E(S|M) \approx 0.43$ km/h), while wolves had the highest mean speed when moving ($\mathbb E(S|M) \approx 2.64$ km/h).
+Consequently, wolves traveled, on average, 0.6 km/h -- 2.5 to 16.7 times further than other mammals at $0^\circ$C.
+
+```{r temp-qs, eval=FALSE}
+library('dplyr')
+d <- readRDS('data/hgam-speed-data.rds')
+get_perc <- function(bools) {
+  out <- as.character(round(mean(bools) * 100, 1))
+  if(grepl('\\.', out)) {
+    out <- paste0(out, ' %')
+  } else {
+    out <- paste0(out, '.0 %')
+  }
+  return(out)
+}
+
+bind_rows(
+  d %>%
+    rename(Species = species) %>%
+    group_by(Species) %>%
+    summarize(`$T < -20^\\\\circ C$` = get_perc(temp_c < -20),
+              `$T > +20^\\\\circ C$` = get_perc(temp_c > +20)),
+  d %>%
+    summarize(Species = 'Total',
+              `$T < -20^\\\\circ C$` = get_perc(temp_c < -20),
+              `$T > +20^\\\\circ C$` = get_perc(temp_c > +20))) %>%
+  knitr::kable(format = 'markdown', caption = paste0("Percentage of observed GPS locations (after data cleaning) with temperatures below -20\U00B0", "C or above 20\U00B0", "C."))
+```
 
@@
 
-intercepts <- tibble(Species = SPECIES,
+intercepts <- tibble(
+  Species = SPECIES,
   'P(M)' = get_int(readRDS('models/binomial-gam-without-temperature.rds')),
@@
   'E(S|T)' = 'km/h' %#% get_int(readRDS('models/gamma-gam.rds')),
-                     'E(D)' = `P(M)` * `E(S)`,
-                     'E(D|T)' = `P(M|T)` * `E(S|T)`) %>%
+  'E(D)' = `P(M)` * (`E(S)`),
+  'E(D|T)' = `P(M|T)` * (`E(S|T)`)) %>%
   mutate(Species = case_when(
@@
   rename_with(\(.x) gsub('E', '\\\\hat{\\\\mathbb E}', .x)) %>%
-  # # add units (added in caption)
-  # rename('$\\widehat{\\mathbb E(S)}$, (m/s)' = `$\\widehat{\\mathbb E(S)}$`,
-  #        '$\\widehat{\\mathbb E(D)}$, (m/s)' = `$\\widehat{\\mathbb E(D)}$`,
-  #        '$\\widehat{\\mathbb E(S\\vert T)}$, (m/s)' = `$\\widehat{\\mathbb E(S\\vert T)}$`,
-  #        '$\\widehat{\\mathbb E(D\\vert T)}$, (m/s)' = `$\\widehat{\\mathbb E(D\\vert T)}$`) %>%
-  knitr::kable(format = 'markdown', digits = 2, caption = paste0("Mean probability of movement (P(M)), speed when moving ($\\mathbb E(S)$, m/s), and distance travelled ($\\mathbb E(D)$, m/s) as estimated by models without and with temperature ($\\mathbb E(...
+  knitr::kable(format = 'markdown', digits = 2, caption = paste0("Mean probability of movement (P(M)), speed ($\\mathbb E(S)$, km/h, assuming constant movement), and distance travelled ($P(M) \\times \\mathbb E(S) = \\mathbb E(D)$, km/h) as estimated by models without and with temperature ($\\mathbb E(...
 | T)$), after post-stratifting to a 1-hour sampling rate and a temperature of T = 0\U00B0", "C."))
@@
 
-Near 0$^\circ$C, wolves preferred dense forest cover ($\gtrapprox$ 50%) high elevations, and distances from water < 10 km; elk preferred intermediate forest cover, elevations between 1 and 2 km, and distances from water of 5-15 km; mountain goats preferred sparse (< 50%) forest cover, elevations between 1 and 2 km; pumas preferred high forest cover, an elevation of ~ 1 km, and distances from water < 10 km; boreal caribou selected for intermediate to high forest cover, elevations near 500 m, and distances from water < 10 km; southern mountain caribou selected for dense forest cover, elevations near 2 km, and distances from water < 5 km; and grizzly bears selected for sparse forest cover (25-50%), elevation near 1 km, and distances from water < 2 km.
+Near 0$^\circ$C, wolves selected for dense forest cover ($\gtrapprox$ 50%), elevations near 1 km, and distances from water < 10 km; elk selected for forest cover near 50%, elevations between 1 and 2 km, and distances from water of 5-15 km; mountain goats selected for sparse (< 50%) forest cover, elevations between 1 and 2 km; cougars selected for dense forest cover, an elevation of ~ 1 km, and distances from water < 10 km; boreal caribou selected for intermediate to dense forest cover, elevations near 500 m, and distances from water < 10 km; southern mountain caribou selected for dense forest cover, elevations near 2 km, and distances from water < 5 km; and grizzly bears selected for sparse forest cover (25-50%), elevation near 1 km, and distances from water < 2 km.
 
 There was relatively strong agreement between models with and without temperature (Figs.
-B5 and S13), but including temperature always resulted in better fits.
-All analyses of deviance showed that temperature significantly affected movement behavior (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; see Table B1 and the analyses of deviance in Appendix B).
-The only exception to this was the mountain goat HRSF, where the addition of temperature resulted in a simpler model (lower model degrees of freedom) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.
+B3, and B13), but including temperature always resulted in better fits.
+All analyses of deviance showed that temperature significantly affected movement behavior (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; Appendix B).
+Accounting for temperature resulted in a simpler HRSF for mountain goats (lower model degrees of freedom) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.
 
@@
 
-\noindent The effects of temperature on movement rates varied in both direction and magnitude across species, even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; see Figs.
-B6-B8).
+\noindent The effects of temperature on movement rates varied in both direction and magnitude across species, even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; Figs.
+B4-B6).
 Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species.
-The models explained reasonably high proportions of the deviance (10.7% for the binomial model and 79.3% for the Gamma model) and had good in-sample prediction (Fig.
-B9).
+The models had good in-sample prediction (Fig.
+B7) and explained reasonably high proportions of the deviance (79.3% for the Gamma model and 10.7% for the binomial model, which is relatively high for a binomial model with binary responses: see ch.
+11 of @mcelreath_statistical_2020].
 All species adapted their daily and seasonal movement behaviour to changes in temperature (Fig.
 \@ref(fig:dist)).
-For example, wolves, elk, pumas, and boreal caribou reduced their mid-day and summer movement rates when temperatures were above 20$^\circ$C.
+The clearest example of this were cougars.
+In mid summer, they moved mostly in the evening if temperatures were cool and mostly in the early morning if temperatures were hot.
+Throughout the year, they tended to move more when it was colder, but what counted as "colder" depended on the time of year.
 However, uncertainties around the estimated effects were often higher at extreme temperatures due to data scarcity.
 All species' estimated probability of moving decreased with sampling intervals approximately above 1 hour, and most species' estimated speed when moving decreased monotonically with sampling interval, although the estimated trends were highly uncertain for some species (Fig.
-B10).
+B8).
 
@@
 Changes in RSS with temperature were also strongest for elevation and generally weakest distance from water, but there were no common trends across all species for any of the three resources.
-All species, with the exception of pumas, exhibited a clear temperature-dependent shift in their preference for forest cover.
+All species, with the exception of cougars, exhibited a clear temperature-dependent shift in their preference for forest cover.
 At higher temperatures, wolves, mountain goats, and grizzly bears became less selective for forest cover, while elk and caribou shifted towards more intermediate forest cover without much of a change in preference width.
 All species shifted elevationally with temperature, although boreal caribou did not exhibit as strong of a change.
-As temperatures rose, elk, mountain goats, and pumas increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation.
+As temperatures rose, elk, mountain goats, and cougars increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation.
 Most species generally remained within 5 km of water, and temperature did not affect their selection strength as much as for the other two resources.
 Again, estimated RSS values were generally most uncertain at extreme temperatures (Fig.
-B13).
+B14).
 
 ```{r hrsfs, fig.cap="\\textbf{Temeperature strongly affects mammals' habitat selection.} Estimated relative selection strength (RSS) for forest cover (\\%), elevation (km), and distance from water (km) as a function of temperature.
-The grey contours indicate the extent of each speices' observed locations.
+The grey contours indicate the extent of each species' observed locations.
 RSS values were re-centered and capped to $2^{\\pm 2}$ to improve readability."}
@@
 \noindent Species varied in both the direction and magnitude of predicted change in distance traveled, but worse climate-change scenarios always corresponded to greater absolute changes in behaviour (Figs.
-\@ref(fig:t-dist), S11, and S12).
-Under the best-case scenario, absolute changes by 2100 were small (approximately 0% to 4%), while under the worst-case scenario absolute changes ranged from ~2% (grizzly bear) to ~25% (pumas), although the models did not explicitly account for changes in hibernation phenology.
+\@ref(fig:t-dist), B9, and B10).
+Under the best-case scenario, absolute changes by 2100 were small (approximately 0% to 4%), while under the worst-case scenario absolute changes ranged from ~2% (grizzly bear) to ~25% (cougars), although the models did not explicitly account for changes in hibernation phenology.
 Throughout BC (Fig.
@@
 Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs).
-The projections only account for changes movement behaviour (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
+The projections only account for changes movement in behaviour (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
 knitr::include_graphics('../figures/distance-travelled-local-cc-predictions.png')
@@
 
-All species exhibited a decrease in RSS within their current range, irrespective of climate change scenario.
-Changes were often were only between -1% and -3% under the best-case scenario but they ranged from -8% to -14% under the worst-case scenario (\@ref(fig:t-hrsfs)).
+We do not predict any of the species to exhibit stronger selection for their current range, irrespective of climate change scenario.
+Changes in RSS by 2100 were between 0% and -3% under the best-case scenario, but they ranged from -8% to -14% under the worst-case scenario (Fig.
+\@ref(fig:t-hrsfs)).
 As with distance traveled, the change in RSS in 2100 varied throughout BC for all species (Fig.
@@
 Although our models do not account for explicit physiological or phenological changes, the tensor product terms in Fig.
-\@ref(fig:dist) suggest that warmer temperatures cause many species to alter their daily and seasonal activity patterns (most visible in pumas and grizzly bears).
-For example, when temperatures were above 0$^\circ$C, pumas moved most at night, but when temperatures were below 0$^\circ$C they tended to move more throughout the day.
+\@ref(fig:dist) suggest that warmer temperatures cause many species to alter their daily and seasonal activity patterns (most visible in cougars and grizzly bears).
+For example, when temperatures were above 0$^\circ$C, cougars moved most at night, but when temperatures were below 0$^\circ$C they tended to move more throughout the day.
 Throughout the year, they adapted their tolerance to temperature and moved less when it was relatively hot (for a given time of year), especially in spring and summer.
-The strong reduction in the mid-day movement rates of wolves, elk, pumas, and boreal caribou when summer temperatures were above 20$^\circ$C suggests that the increasingly common and intense heat waves across British Columbia will have community-wide impacts on movement rates, encounter rates, and potentially community structure [@martinez-garcia_how_2020].
+The strong reduction in the mid-day movement rates of wolves, elk, cougars, and boreal caribou when summer temperatures were above 20$^\circ$C suggests that the increasingly common and intense heat waves across British Columbia will have community-wide impacts on movement rates, encounter rates, and potentially community structure [@martinez-garcia_how_2020].
 More work is necessary on quantifying interspecific responses to temperature, including the effects of temperature on predation rates [but see: @cunningham_opportunity_2021; @glass_trade-off_2021; @brivio_seeking_2024].
@@
 
+## Species-specific considerations
+
+### *Canis lupus*
+
+### *Cervus canadensis*
+
+### *Oreamnos americanus*
+
+### *Puma concolor*
+
+### *Rangifer tarandus* (boreal)
+
+### *Rangifer tarandus* (s. mountain)
+
+### *Ursus arctos horribilis*
+
 ## Considerations for future studies
+
+**add a section about the importance of studying smaller mammals, too**
 

### 2025-06-11 writing/appendix-b-gams.Rmd
capped max SE to 3

@@
 3 in the main text).
-The grey contours indicate the extent of each speices' observed locations."}
+The grey contours indicate the extent of each speices' observed data.
+For ease of readability, the color bar values were capped to a maximum of 3 (original maximum was 7.00)."}
 knitr::include_graphics('../figures/hrsf-surface-plots-se.png')

### 2025-06-11 analysis/figures/temperature-hrsfs.R

@@ -197,6 +197,7 @@ p_se <-
   preds %>%
   filter((! too_far_2)) %>%
   select(species, lab, x, temperature_C, variable, se.fit, too_far_1) %>%
+  mutate(se.fit = if_else(se.fit > 3, 3, se.fit)) %>%
   ggplot() +
   facet_grid(variable ~ lab, scales = 'free', labeller = label_parsed,
              switch = 'y') +

### 2025-06-11 writing/manuscript.Rmd
addressed KH's comments (need to make some figs)

@@
 Previous studies have quantitatively estimated the effects of temperature on mammalian movement behaviour, but few have estimated the effects of future climate change.
-Consequently, it is still unclear how and where mammals will move in the 21^st^ century.
-We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on the probability of moving, the speed when moving, and the habitat selection of six species throughout British Columbia, Canada, between 1998 and 2023.
+Consequently, it is still unclear how mammals will adapt their movement behavior in the 21^st^ century.
+We address this gap by using continuous-time movement models and hierarchical generalized additive models to quantify the effects of temperature on the probability of moving, the speed when moving, and the habitat selection of six species throughout British Columbia, Canada, between 1998 and 2023.
 We show that temperature is a strong determinant of when, how, and where mammals move, and we predict that climate change will impact mammals' movement rates and habitat selection as temperatures warm throughout the 21^st^ century.
-We found the effects of temperature on movement rates varied across species, so we suggest that species' estimated responses not be interpreted individually when making inferences about how climate change will impact ecological communities.
+We found the effects of temperature on movement rates varied across species, so we suggest that species' estimated responses be interpreted holistically when making inferences about how climate change will impact ecological communities.
 In contrast, habitat selection strength decreased for all species within their observed ranges, suggesting that all six species may shift their range in the coming decades.
 As climate change exposes mammals to novel environmental conditions, predicting changes in animal behavior and community structure will become crucial for effective and proactive conservation.
-We thus conclude by providing suggestions for conservation and future research while calling for a more holistic and collaborative framework."
+We thus conclude by providing suggestions for conservation and future research, with particular attention to the study design, statistical modeling, and forecasting.
+We provide detailed `R` code and appendices so that other researchers can implement our methods on their own datasets."
 if(FALSE){
@@
 
-\noindent \textbf{References}: 141 (updated on 2025-05-16) <!-- counted by using movement ecology csl -->
+\noindent \textbf{References}: 142 (updated on 2025-06-11) <!-- counted by using movement ecology csl -->
 
@@
 The Canadian province of British Columbia is currently experiencing rapid widespread warming [@dierauer_climate_2021; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024], and more frequent and intense extreme events [@zhang_increased_2023], including forest fires [@zurowski_summer_2023].
-As we approach the deadline for the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022], decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness.
+As we approach the deadline for the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [@convention_on_biological_diversity_154_2022], decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness.
 Predicting how climate change will affect how, when, and where animals will move is necessary for identifying landscape dynamics and value will evolve throughout the century.
@@
 (B) DAG for Relative Selection Strength (RSS) for percent forest cover (F), elevation (E), and distance from water (W).
-The RSS for of a given habitat depends on all three resources, and the selection for each resource is independent of the other two resources and dependent on temperature.", fig.height=4, fig.width=9, out.width='\\linewidth'}
+The RSS for of a given habitat depends on all three resources, and the selection for each resource is independent of the other two resources and dependent on temperature.", fig.height=4, fig.width=10, out.width='\\linewidth'}
 library('dplyr')   # for data wrangling
@@
 Together, the binomial HGAM and the gamma HGAM inform us on an animal's long-term average speed, since it is the product of the probability of moving and its average speed when moving.
+
+\clearpage
 
@@
 
-```{r temp-qs, eval=FALSE}
+```{r temp-qs}
 library('dplyr')
-d <- readRDS('data/hgam-speed-data.rds')
+d <- readRDS('../data/hgam-speed-data.rds')
 get_perc <- function(bools) {
@@
     group_by(Species) %>%
-    summarize(`$T < -20^\\\\circ C$` = get_perc(temp_c < -20),
-              `$T > +20^\\\\circ C$` = get_perc(temp_c > +20)),
+    summarize(`T < -20\\textdegree{} C` = get_perc(temp_c < -20),
+              `T > +20\\textdegree{} C` = get_perc(temp_c > +20)),
   d %>%
     summarize(Species = 'Total',
-              `$T < -20^\\\\circ C$` = get_perc(temp_c < -20),
-              `$T > +20^\\\\circ C$` = get_perc(temp_c > +20))) %>%
+              `T < -20\\textdegree{} C` = get_perc(temp_c < -20),
+              `T > +20\\textdegree{} C` = get_perc(temp_c > +20))) %>%
   knitr::kable(format = 'markdown', caption = paste0("Percentage of observed GPS locations (after data cleaning) with temperatures below -20\U00B0", "C or above 20\U00B0", "C."))
@@
 \@ref(fig:dist)).
-The clearest example of this were cougars.
+The clearest example of this was cougars.
 In mid summer, they moved mostly in the evening if temperatures were cool and mostly in the early morning if temperatures were hot.
-Throughout the year, they tended to move more when it was colder, but what counted as "colder" depended on the time of year.
+Throughout the year, they tended to move more when it was colder, but what they perceived as "colder" depended on the time of year.
 However, uncertainties around the estimated effects were often higher at extreme temperatures due to data scarcity.
@@
 
-Species' relative selection strength (RSS) was generally strongest for elevation and weakest for forest cover, but species' RSS for resources depended on temperature.
+Species' relative selection strength (RSS) was generally strongest for elevation and weakest for forest cover, but species' RSS for resources depended on temperature (Fig.
+\@ref(fig:hrsfs)).
 Changes in RSS with temperature were also strongest for elevation and generally weakest distance from water, but there were no common trends across all species for any of the three resources.
@@
 At higher temperatures, wolves, mountain goats, and grizzly bears became less selective for forest cover, while elk and caribou shifted towards more intermediate forest cover without much of a change in preference width.
-All species shifted elevationally with temperature, although boreal caribou did not exhibit as strong of a change.
-As temperatures rose, elk, mountain goats, and cougars increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation.
+All species shifted elevationally with temperature, although boreal caribou's selection strength for elevation was less sensitive to elevation.
+Overall, as temperatures rose, elk, mountain goats, and cougars increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation.
 Most species generally remained within 5 km of water, and temperature did not affect their selection strength as much as for the other two resources.
@@
 ```{r hrsfs, fig.cap="\\textbf{Temeperature strongly affects mammals' habitat selection.} Estimated relative selection strength (RSS) for forest cover (\\%), elevation (km), and distance from water (km) as a function of temperature.
-The grey contours indicate the extent of each species' observed locations.
+The grey contours indicate the extent of each species' observed data.
 RSS values were re-centered and capped to $2^{\\pm 2}$ to improve readability."}
@@
 \@ref(fig:t-dist), B9, and B10).
-Under the best-case scenario, absolute changes by 2100 were small (approximately 0% to 4%), while under the worst-case scenario absolute changes ranged from ~2% (grizzly bear) to ~25% (cougars), although the models did not explicitly account for changes in hibernation phenology.
+Under the best-case scenario, absolute changes by 2100 were small (approximately 0% to 4%), while under the worst-case scenario absolute changes ranged from ~2% (grizzly bear) to ~25% (cougars).
+However, while the models estimated that grizzly bears would move substantially less (if at all) in winter (Figs.
+\@ref(fig:dist), B4-B6), the models did not explicitly account for changes in hibernation phenology.
+Consequently, the climate change projections do not explicitly account for changes in energetic needs or physiology.
 Throughout BC (Fig.
@@
 Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs).
-The projections only account for changes movement in behaviour (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
+The projections only account for changes in movement in behaviour (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
 knitr::include_graphics('../figures/distance-travelled-local-cc-predictions.png')
@@
 
-\noindent We have demonstrated that temperature is an important determinant of whether, how, and where mammals move, even while accounting for seasonality (e.g, changes in photoperiod, hormones, and resources).
+<!-- - address any questions presented in the introduction -->
+
+\noindent We have demonstrated that temperature is an important determinant of whether, how, and where large mammals move, even while accounting for seasonality (e.g, changes in photoperiod, hormones, and resources).
 Disentangling temperature from other drivers is important for predicting how changes in climate will affect mammalian movement when other drivers remain approximately constant.
 However, predicting mammals' responses to climate change becomes increasingly complex as habitats warm and animals are exposed to increasingly frequent conditions that are both extreme and novel.
+
 Predicting mammals' responses at the data-scarce fringes of the surface plots in Figs.
 \@ref(fig:dist) and \@ref(fig:hrsfs) comes with substantial uncertainty, particularly given that the responses are nonlinear.
-At unusually warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behaviour and that of their competitors, predators, and/or prey can often be substantially different from the behaviour at typical temperatures.
-As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @ipcc_climate_2023], mammals will be increasingly forced towards atypical behaviours that may alter community structures and behaviours, both during such events and afterwards [@zhang_increased_2023; @logares_black_2012; @anderson_black-swan_2017].
+At warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behaviour and that of their competitors, predators, and/or prey can often be substantially different from the behaviour at typical temperatures.
+As extreme temperature events become more frequent and prolonged throughout the year [@yao_emergence_2022; @ipcc_climate_2023], mammals will be increasingly forced towards atypical behaviours that may alter community structures and behaviours, both during such events and afterwards [@zhang_increased_2023; @logares_black_2012; @anderson_black-swan_2017].
 For instance, changes in climate and phenology impact the life history and behaviour of many hibernating mammals [@wells_life_2022], and hot weather can also affects mammals' sleep quality [@mortlock_sleep_2024] and likelihood to enter torpor [@fjelldal_nightly_2021].
-Such changes may alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024].
-At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023], increase their chances of finding food or being predated upon [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021].
+
+Changes in animal movement behavior may also alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024].
+At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023], increase their chances of finding food or being preyed on [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021].
 These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011].
 
-Our ability to respond to current and future changes in climate is contingent on our ability to prepare for and predict change.
+Our ability to respond changes in climate is contingent on our ability to prepare for and predict change.
 However, predicting animal behaviour becomes increasingly complicated as the conditions animals are exposed to deviate from current, typical conditions, especially when responses are nonlinear and data are sparse.
@@
 Instead, we hope they serve as a starting point to (1) demonstrate that mammals' movement rates and habitat selection depend on temperature and (2) how one can estimate mammals' changes in movement behaviour due to climate change.
-Additionally, communicating uncertainty in one's estimates is crucial in assessing risk probabilistically [@aven_implementing_2002; @ayre_bayesian_2012; @kluppelberg_bayesian_2014].
+Additionally, communicating uncertainty in one's estimates is crucial in assessing risk probabilistically, and appropriate measures of uncertainty require careful and robust modeling [@aven_implementing_2002; @ayre_bayesian_2012; @kluppelberg_bayesian_2014].
+
 Recent events in global politics, including global conflicts [@mcnutt_scientists_2022; @tollefson_what_2022] and the rapid rise in generative machine learning [more commonly: "AI", see @roumeliotis_chatgpt_2023; @van_noorden_chatgpt_2023; @deepseek-ai_deepseek-r1_2025] and its prohibitive energy and water demands [@li_making_2023] and carbon footprint [@luccioni_estimating_2024], have placed doubt on the feasibility of reaching climate change goals that seemed achievable in the past decade [@united_nations_environment_programme_emissions_2024].
 Mitigating climate change and its consequences will require extensive and widespread collaborative effort [@huang_achieving_2021].
-Similarly, protecting a third of the world's ecosystems [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022] appears less tangible as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023].
-Achieving the "30 by 30" goal will require active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023].
-Understanding the consequences of climate change on mammals' movement behaviour and spatial distribution is a first step towards anticipating and proactively responding to future changes in human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018].
+Similarly, protecting a third of the world's ecosystems [@convention_on_biological_diversity_154_2022] appears less achievable as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023].
+Achieving the "30 by 30" goal will require international collaboration as well as active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023].
+Understanding the consequences of climate change on mammals' movement behaviour is a first step towards proactively responding to how mammals will respond to human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018].
 In the following sections, we discuss the implications of our results in more detail.
@@
 
-\noindent The lack of common effects of temperature across all species' movement rates indicates that preparing for future changes will require a variety of physiological and behavioral adaptations as ecological communities respond in complex and interconnected ways.
-Although our models do not account for explicit physiological or phenological changes, the tensor product terms in Fig.
+\noindent The heterogeneity mammals' responses to temperature suggests that ecological communities will respond to change in complex and interconnected ways.
+Although our models do not account for explicit physiological or phenological changes, the surfaces in Fig.
 \@ref(fig:dist) suggest that warmer temperatures cause many species to alter their daily and seasonal activity patterns (most visible in cougars and grizzly bears).
@@
 
-As with the effects of temperature on movement rates, the lack of common effects of temperature across all species indicates that efficient and proactive habitat conservation will require taxonomic nuance.
-In particular, our HRSFs did not account for any explicit physiological responses to temperature that may impact movement, such as changes in the phenology of plants, ice, or mammalian physiology or behaviour (e.g., moulting, migration), other than as direct responses to temperature.
+As with the effects of temperature on movement rates, the hererogeneity of temperature effects on species' habitat selection indicates that efficient and proactive habitat conservation will require taxonomic nuance.
+In particular, our HRSFs did not account for any explicit physiological responses to temperature that may impact movement, such as changes in the phenology of plants, fire, ice, or mammalian physiology (e.g., moulting, fat storages) or behaviour (e.g., migration, food caching), other than as direct responses to temperature.
 Consequently, the behavioural changes we present should be interpreted carefully.
@@
 
-Achieving the goals laid out by the "30 by 30" conservation initiative [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022] will require careful planning but often also rapid action.
+Achieving the goals laid out by the "30 by 30" conservation initiative [@convention_on_biological_diversity_154_2022] will require careful planning but often also rapid action.
 In the case of Canada, only 13.7% of its terrestrial areas and inland waters were protected at the end of 2023 [@environment_and_climate_change_canada_canadian_2024; @minister_of_environment_and_climate_change_act_2024].
-Additionally, efficient and effective conservation will require collaboration with Indigenous and local communities to ensure Treaties are respected in the process [e.g., @falardeau_biophysical_2022; @lamb_braiding_2023; @wong_towards_2020; @turner_its_2009] while also actively preparing for future change [@desjardins_looking_2020; @hessami_indigenizing_2021].
-The large range in predicted changes by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in Fig.
+Additionally, efficient and effective conservation will require collaboration with Indigenous and local communities to ensure Treaties are respected in the process [@falardeau_biophysical_2022; @lamb_braiding_2023; @wong_towards_2020; @turner_its_2009] while also actively preparing for future change [@desjardins_looking_2020; @hessami_indigenizing_2021].
+
+The diversity in predicted changes in movement behavior by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in Fig.
 \@ref(fig:s-hrsfs) highlight the need for careful planning that incorporates not only reliable estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates.
-The width of the prediction intervals in Fig.
+For example, the width of the prediction intervals in Fig.
 \@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas.
-Accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage and prey availability [@mezzini_how_2025], encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@smith_densitydependent_2023], competitive pressure [@torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019].
-This complexity results in the great variation among the responses of different species (and individuals), especially as populations a rapid cascade of change [@botero_evolutionary_2015] and animals face increasing pressures from human activity, including habitat fragmentation, habitat loss, and greater human presence in wild spaces [@sih_evolution_2011; @tucker_moving_2018; @sawyer_identifying_2009; @weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022].
+Population-level means that ignore such variation can greatly limit the efficacy of conservation projects [@muff_accounting_2020; @mortlock_sleep_2024].
+Additionally, accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage and prey availability [@mezzini_how_2025], encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@smith_densitydependent_2023], competitive pressure [@torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019].
+This complexity results in the great variation among the responses of different species (and individuals), especially as populations undergo a rapid cascade of change [@botero_evolutionary_2015] and animals face increasing pressures from human activity, including habitat fragmentation, habitat loss, and greater human presence in wild spaces [@sih_evolution_2011; @tucker_moving_2018; @sawyer_identifying_2009; @weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022].
 As selection strength for current ranges changes (and likely decreases), some animals may disperse, relocate to a new habitat, or remain within the current range despite the reduced fitness and increased extinction risk [@logares_black_2012; @anderson_black-swan_2017; @duncan_predicting_2012].
@@
 
+Increasing temperatures increased RSS for both elevations near ~500 m as well as > 1,000 m, but such a selection is only possible if: (1) such habitats exist in the animals' range, and (2) crossing the 500 m to > 1,000 m regions is safe and attractive
+
 ### *Ursus arctos horribilis*
@@
 
-**add a section about the importance of studying smaller mammals, too**
+Our work highlights three central considerations for future research: (1) telemetry sampling rates should be designed primarily in relation to the movement timescales of the species of interest [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @pease_ecological_2024], (2) the number of individuals tracked and the length of each telemetry should depend not just on population size but also the rate of environmental change, and (3) predicting changes in movement behaviour becomes highly complex when responses are nonlinear, especially when changes depend on many interacting factors [@polazzo_measuring_2024] and one is interested in predicting responses in extreme conditions for which data are scarce [@steixner-kumar_strategies_2020].
 
-Our work highlights three central considerations for future research: (1) telemetry sampling rates should be designed primarily in relation to the movement timescales of the species of interest [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @pease_ecological_2024], (2) the number of individuals tracked and the length of each telemetry should depend not just on population size but also environmental stability, and (3) predicting changes in movement behaviour becomes highly complex when responses are nonlinear, especially when changes depend on many interacting factors [@polazzo_measuring_2024] and one is interested in predicting responses in extreme conditions for which data are scarce [@steixner-kumar_strategies_2020].
-We address each consideration in the following three paragraphs.
-
-While the `ctmm` models produced scale-independent estimates of speed [i.e., model interpretation is independent of sampling interval: @noonan_scale-insensitive_2019], the accuracy, size, and stability of speed estimates still depended on the GPS sampling frequency.
-This is because coarsely-sampled movement data contains information on large-scale movements (e.g., range crossings, migrations) but not fine-scale movements [e.g., first-order habitat selection *sensu* @johnson_comparison_1980; also see @nathan_big-data_2022].
+While the `ctmm` movement models produced scale-independent estimates of speed [i.e., model interpretation is independent of sampling interval: @noonan_scale-insensitive_2019], the accuracy, size, and stability of speed estimates still depended on the GPS sampling frequency.
+This dependency is because coarsely-sampled movement data contains information on large-scale movements (range crossings, migrations) but not fine-scale movements, including first-order habitat selection [@johnson_comparison_1980].
 Using the boreal caribou as an example, the 13-hour sampling interval allows us to reasonably estimate the caribou's movement path at a temporal scale of approximately 13 hours (or greater), but we cannot produce reasonable movement trajectories at a much finer (e.g., hourly) scale.
+@nathan_big-data_2022 provides additional examples of how the frequency of location data affects the results that can be inferred by modeling movemet behavior.
 Consequently, we suggest being cautious when comparing estimated movement behaviours across species, even though all predictions have been corrected to the hourly timescale by predicting for 1-hour time intervals (i.e., `dt_hours = 1`).
-Thus, sampling schedules should be fine enough to reconstruct animals' movement at a sufficiently fine scale.
-Good estimates of an animal's speed and its movement path require telemetry locations to be taken more often than the animal's directional persistence [@noonan_scale-insensitive_2019], so that, on average, the dataset contains multiple locations in between changes in direction.
+Ideally, sampling schedules should be fine enough to reconstruct animals' movement movement paths.
+Good estimates of an animal's speed and trajectory require telemetry locations to be taken more often than the animal's directional persistence [@noonan_scale-insensitive_2019], so that, on average, the dataset contains multiple locations in between changes in direction.
 What constitutes a change in direction depends on what movement scale one is investigating.
-Small-scale movements and first-order spatial selection will require more frequent sampling than migratory movement or second- and third-order spatial selection.
+Small-scale movements and first-order spatial selection will require more frequent sampling than migratory movement or second-order and third-order spatial selection.
 While `ctmm` movement models are scale-invariant in that any model can be scaled to larger or smaller areas and timescales, the model estimates are not independent of the scale of the sampling frequency.
 
-When landscapes are relatively predictable across years, a larger number of sampled individuals is likely preferable over longer, multi-year telemetries.
-This allows one to quantify the variance across individuals, including the range of conditions and environments that individual animals are in.
+When landscapes are relatively predictable across years, a larger number of sampled individuals is likely preferable over the duration of each tracking period.
+The greater number of tracked animals allows one to quantify the variance across individuals, including the range of conditions and environments that individual animals are in.
 A good estimate of inter-individual variance provides better coefficient estimates along with more appropriate measures of uncertainty.
-However, when conditions across years are relatively stochastic, multi-year telemetries allow one to better estimate inter-annual variation without conflating it with inter-individual differences.
-In either case, carefully-designed sampling schedules and (Bayesian) hierarchical models can provide good estimates of the effects of interested along with appropriate measures of uncertainty [@mcelreath_statistical_2020; @kluppelberg_bayesian_2014].
+However, when conditions across years are stochastic, multi-year telemetries allow one to better estimate inter-annual variation without conflating it with inter-individual differences.
+Given the widespread, rapid, and accelerating changes across many habitats (particulartly at high elevations and at high latitudes), we suggest researchers focus on long-term telemetry datasets to quantify the effects of intra-annal variability while increasing the ranges of temperatures each individual is observed in.
+Longer observational periods can also improve the chances of observing different community dynamics, such as shifts in predator-prey dinamics.
 
-Quantifying and communicating uncertainty become particularly important as we predict for conditions for which we have little to no data, such as during extreme events.
-Additionally, estimating nonlinear trends is particularly difficult near the edges of the data (e.g., extremely high temperatures), especially when data is scarce.
-Consequently, the estimated responses at the fringes of the surfaces in figures \@ref(fig:dist) and \@ref(fig:hrsfs) are more likely likely to be poorly estimated, particularly in the case of HRSFs, but increasing data density throughout the resource-temperature surfaces can ameliorate the issue.
-Preparing for future changes requires some degree of inevitable risk, so it is imperative that models be statistically sound and interpretable, and that results are communicated transparently and clearly to decision makers and the public alike.
+Temperature affects many aspects of mammalian physiology and behavior, including energetics, sleep, and movement behavior.
+Climatic changes during the last two centuries have exposed many mammals to novel and increasingly extreme environmental conditions that have led to visible changes that impact not only individuals but also ecological and human communities.
+However, quantifying the effects of climate change is often complex, especially as conditions become increasingly different and extreme events become increasingly common.
+Accurately quantifying the nonlinear effects of temperature on when, how, and where mammals move requires careful data collection and model design.
+We have presented a multi-species analysis of the effects of temperature on mammalian movement rates and habitat selection in hopes that other researchers can leverage this framework and models to expand our understanding of how temperature affects the movement behavior of other species, including smaller mammals, more elusive species, and non-mammal animals.
+Understanding of how temperature affects each species in a community will allow us to begin making community-level inferences of how temperature will affect intra-specific interactions from a movement behavior perspective [@nathan_movement_2008; @martinez-garcia_how_2020; @nathan_big-data_2022].
+While adopting such a perspective is surely not simple, it would result in more complete estimates of the effects of climate change on community ecology, particularly when species are exposed to rapid change in many important variables.
+Future work should explore the effects of temperature on movement behavior while accounting for finer-scale and species specific variables that were not accounted for in this study.
+Examples of these include temporally dynamic measures of forest type and age, canopy density, competitive pressures, forage availability, and predator avoidance, or environmental stochasticity.
+However, many of these data not available, so we also suggest that more work be done on quantifying such measures using spatiotemporally dynamic models and sufficiently fine resolutions.
 
-<!-- - address any questions presented in the introduction -->
+# Author contributions
+
+SM performed the data cleaning, ran the analyses, and wrote the majority of the manuscript.
+CHF wrote code for estimating instantaneous movement speeds.
+MJN conceived of the project idea and supervised SM throughout the project.
+All other authors contributed telemetry data and contributed to the interpretation of the results for their species of interest.
+All authors contributed to reviewing the manuscript.
 

### 2025-06-12 writing/manuscript.Rmd
renamed "displ" to "distance"

@@
 The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest cover."}
-knitr::include_graphics('../figures/bc-displ-2100.png')
+knitr::include_graphics('../figures/bc-distance-2100.png')
 ```

### 2025-06-12 analysis/figures/bc-distance-predictions.R

@@ -192,5 +192,28 @@ p <- bc_preds %>%
   labs(x = NULL, y = NULL) +
   theme(legend.position = 'top', legend.key.width = rel(3))
 
-ggsave('figures/bc-displ-2100.png', p, width = 15, height = 8,
+ggsave('figures/bc-distance-2100.png', p, width = 15, height = 8,
+       units = 'in', dpi = 600, bg = 'white', scale = 1.2)
+
+# make zoomed in version
+tels <- readRDS()
+
+p <- bc_preds %>%
+  filter(case_when(species == '')) %>%
+  # cap values for readability
+  # mutate(d = case_when(d < 2^min(z_breaks) ~ 2^min(z_breaks),
+  #                      d > 2^max(z_breaks) ~ 2^max(z_breaks),
+  #                      TRUE ~ d)) %>%
+  ggplot() +
+  facet_grid(scenario ~ lab, labeller = label_parsed) +
+  geom_raster(aes(x, y, fill = log2(d))) +
+  geom_sf(data = bc, fill = 'transparent') +
+  scale_fill_distiller(name = 'Relative change in distance travelled',
+                       palette = 'PuOr', limits = range(z_breaks),
+                       breaks = z_breaks, labels = \(x) round(2^x, 2),
+                       direction = 1) +
+  labs(x = NULL, y = NULL) +
+  theme(legend.position = 'top', legend.key.width = rel(3))
+
+ggsave('figures/ranges-dist-2100.png', p, width = 15, height = 8,
        units = 'in', dpi = 600, bg = 'white', scale = 1.2)

### 2025-06-12 analysis/figures/actws-2024-jasper/bc-resource-rasters.R
removed unused palettes

@@ -14,11 +14,6 @@ theme_set(theme_void() +
                   legend.text = element_text(face = 'bold'),
                   legend.title = element_text(face = 'bold')))
 
-# create custom color palettes for each predictor
-forest_pal <- colorRampPalette(c('#C2B280', 'darkgreen'))(100)
-elevation_pal <- colorRampPalette(c('white', 'brown4'))(100)
-water_pal <- c('transparent', 'blue')
-
 f <- rast('data/resource-rasters/bc-forest.tif') %>%
   project(crs(bc)) %>%
   as.data.frame(xy = TRUE) %>%

### 2025-06-12 analysis/figures/bc-distance-predictions.R
made local maps of distance and RSS

@@ -58,9 +58,9 @@ if(file.exists('data/cc-hgam-bc-projections-albers.rds')) {
     mutate(
       .,
       p = predict(m_1, newdata = ., type = 'response',se.fit = FALSE,
-                  discrete = TRUE, exclude = 's(animal)'),
+                  discrete = FALSE, exclude = 's(animal)'),
       s = predict(m_2, newdata = ., type = 'response', se.fit = FALSE,
-                  discrete = TRUE, exclude = 's(animal)'),
+                  discrete = FALSE, exclude = 's(animal)'),
       d = p * s) %>%
     # take the weighted mean of the approximated gaussian distributions 
     # after grouping all the 2025 scenarios together
@@ -193,27 +193,4 @@ p <- bc_preds %>%
   theme(legend.position = 'top', legend.key.width = rel(3))
 
 ggsave('figures/bc-distance-2100.png', p, width = 15, height = 8,
-       units = 'in', dpi = 600, bg = 'white', scale = 1.2)
-
-# make zoomed in version
-tels <- readRDS()
-
-p <- bc_preds %>%
-  filter(case_when(species == '')) %>%
-  # cap values for readability
-  # mutate(d = case_when(d < 2^min(z_breaks) ~ 2^min(z_breaks),
-  #                      d > 2^max(z_breaks) ~ 2^max(z_breaks),
-  #                      TRUE ~ d)) %>%
-  ggplot() +
-  facet_grid(scenario ~ lab, labeller = label_parsed) +
-  geom_raster(aes(x, y, fill = log2(d))) +
-  geom_sf(data = bc, fill = 'transparent') +
-  scale_fill_distiller(name = 'Relative change in distance travelled',
-                       palette = 'PuOr', limits = range(z_breaks),
-                       breaks = z_breaks, labels = \(x) round(2^x, 2),
-                       direction = 1) +
-  labs(x = NULL, y = NULL) +
-  theme(legend.position = 'top', legend.key.width = rel(3))
-
-ggsave('figures/ranges-dist-2100.png', p, width = 15, height = 8,
        units = 'in', dpi = 600, bg = 'white', scale = 1.2)

### 2025-06-12 analysis/figures/local-distance-predictions.R

@@ -1,0 +1,154 @@
+library('dplyr')     # for data wrangling
+library('tidyr')     # for data wrangling
+library('purrr')     # for functional programming
+library('lubridate') # for working with dates
+library('sf')        # for working with spatial data
+library('terra')     # for working with rasters
+library('mgcv')      # for generalized additive models
+library('gratia')    # for useful functions for generalized additive models
+library('ggplot2')   # for fancy figures
+library('khroma')    # for colorblind-friendly palettes
+library('stringr')   # for working with strings
+library('cowplot')   # for plots in grids (for free scales)
+source('analysis/figures/default-ggplot-theme.R') # for consistent theme
+source('functions/get_legend.R') # 
+source('data/bc-shapefile.R') # import shapefile of bc
+
+colorRampPalette(RColorBrewer::brewer.pal(11, 'PuOr'))(1e3) %>%
+  plot_scheme_colorblind()
+
+if(file.exists('data/cc-hgam-projections-local-2100.rds')) {
+  local_2100_proj <- readRDS('data/cc-hgam-projections-local-2100.rds')
+} else {
+  # import models
+  m_1 <- readRDS('models/binomial-gam.rds')
+  m_2 <- readRDS('models/gamma-gam.rds')
+  
+  # when predicting: only excluding animal-level random effect and choosing
+  # specific values for the rest
+  
+  # import prediction data for each species
+  #' from `analysis/figures/temperature-hgams-projections.R`
+  cc_newd <- readRDS('data/hgam-cc_newd.rds')
+  
+  cc_proj <- cc_newd %>%
+    filter(year == 2025 | year == 2100) %>%
+    mutate(.,
+           p = predict(m_1, newdata = ., type = 'response', se.fit = FALSE,
+                       discrete = TRUE, exclude = 's(animal)'),
+           s = predict(m_2, newdata = ., type = 'response', se.fit = FALSE,
+                       discrete = TRUE, exclude = 's(animal)'),
+           d = p * s) %>%
+    # average across day of year
+    group_by(scenario, year, species, long, lat) %>%
+    summarize(p = weighted.mean(p, w = weight),
+              s = weighted.mean(s, w = weight),
[…]

### 2025-06-12 analysis/figures/local-rss-predictions.R

@@ -1,0 +1,261 @@
+library('dplyr')     # for data wrangling
+library('tidyr')     # for data wrangling
+library('purrr')     # for functional programming
+library('lubridate') # for working with dates
+library('sf')        # for working with spatial data
+library('terra')     # for working with rasters
+library('mgcv')      # for generalized additive models
+library('gratia')    # for useful functions for generalized additive models
+library('ggplot2')   # for fancy figures
+library('khroma')    # for colorblind-friendly palettes
+library('stringr')   # for working with strings
+library('cowplot')   # for plots in grids (for free scales)
+source('analysis/figures/default-ggplot-theme.R') # for consistent theme
+source('functions/get_legend.R') # 
+source('data/bc-shapefile.R') # import shapefile of bc
+
+colorRampPalette(RColorBrewer::brewer.pal(11, 'PiYG'))(1e3) %>%
+  plot_scheme_colorblind()
+
+if(file.exists('data/cc-hrsf-projections-local-2100.rds')) {
+  local_2100_proj <- readRDS('data/cc-hrsf-projections-local-2100.rds')
+} else {
+  # import resource rasters
+  f <- rast('data/resource-rasters/forest.tif')
+  e <- rast('data/resource-rasters/bc-buffered-dem-z6.tif')
+  w <- rast('data/resource-rasters/distance-from-water.tif')
+  
+  # terms to exclude from the predictions
+  SM <- gratia::smooths(readRDS('models/rsf-Oreamnos americanus-2025-01-20.rds'))
+  EXCLUDE <- SM[grepl('animal', SM)]
+  
+  # find model files and drop those with no temperature smooths
+  MODEL_FILES <- list.files('models', '^rsf-', full.names = TRUE)
+  MODEL_FILES <- MODEL_FILES[! grepl('no-temp', MODEL_FILES)]
+  
+  cc_proj <-
+    map(MODEL_FILES, function(fn) {
+      rsf <- readRDS(fn)
+      sp <- tolower(fn) %>%
+        gsub('models/rsf-', '', .) %>%
+        gsub('-2.*', '', .) %>%
+        gsub(' ', '-', .) %>%
+        gsub('\\(', '', .) %>%
+        gsub('\\)', '', .) %>%
+        gsub('\\.', '', .)
+      
+      tibble(
+        species = fn %>%
+          gsub(pattern = 'models/rsf-', replacement = '', x = .) %>%
[…]

### 2025-06-12 functions/get_legend.R
added argument for legend position

@@ -2,7 +2,8 @@ library('ggplot2')
 library('cowplot')
 
 #' `get_legend()` for `{cowplot}` version 1.1.3 returns an empty plot 
-get_legend <- function(.plot) {
-  get_plot_component(.plot + theme(legend.position = 'top'),
-                     pattern = 'guide-box-top', return_all = TRUE)
+get_legend <- function(.plot, position = 'top') {
+  get_plot_component(.plot + theme(legend.position = position),
+                     pattern = paste0('guide-box-', position),
+                     return_all = TRUE)
 }

### 2025-06-12 writing/manuscript.Rmd
added comments for local distance map

--- 2025-06-11 writing/manuscript.Rmd
@@
 Consequently, the climate change projections do not explicitly account for changes in energetic needs or physiology.
-Throughout BC (Fig.
-\@ref(fig:s-dist)), all species showed little to no absolute change under the best-case scenario (approximately 0-4% relative to 2025, on average), and approximately 2-25% average absolute change under the worst-case scenario, but the predicted changes varied spatially due to heterogeneity in climate change.
-Again, absolute changes were smallest under the best-case scenario and greatest under the worst-case scenario for all species.
+Projected changes also varied spatially due to heterogeneity in projected climate change (Fig.
+\@ref(fig:s-dist)).
+Again, absolute changes were generally greatest under worse scenarios, but the direction of change at each location also varied across scenarios (most visible in cougars).
 
@@
 
-```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each location's average distance in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase.
-For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.72 to 1.29).
-The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest cover."}
-knitr::include_graphics('../figures/bc-displ-2100.png')
+```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each pixel's average distance in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase.
+For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.50 to 1.56).
+The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest density."}
+knitr::include_graphics('../figures/local-distance-2100.png')
 ```
@@
 The increase in boreal caribou's RSS for high elevations at warmer temperatures was highly uncertain and often not credible at the 95\\% level."}
-knitr::include_graphics('../figures/bc-rss-2100.png')
+knitr::include_graphics('../figures/local-rss-2100.png')
 ```

### 2025-06-12 analysis/figures/local-rss-predictions.R
forgot to save after img size change

@@ -257,5 +257,5 @@ plot_grid(
             rel_widths = c(1.32, 1.1, 1.67, 1.64, 1.57, 1.58, 1.64)),
   ncol = 1, rel_heights = rep(c(0.2, 1), 3))
 
-ggsave('figures/local-resources.png', width = 15, height = 7.25,
+ggsave('figures/local-resources.png', width = 15, height = 8.5,
        units = 'in', dpi = 600, bg = 'white', scale = 1.5)

### 2025-06-12 analysis/figures/temperature-hrsfs-projections.R
now quantifies rel chng at px-level ref (running)

@@ -89,13 +89,13 @@ if(file.exists('data/cc-hrsf-projections.rds')) {
       # not including CIs because averaging them is not straightforward
       group_by(lab, scenario, year) %>%
       summarize(l_05 = quantile(l, 0.05),
-                l_25 = quantile(l, 0.25),
                 l_median = quantile(l, 0.50),
-                l_75 = quantile(l, 0.75),
                 l_95 = quantile(l, 0.95),
                 .groups = 'drop') %>%
       # divide by mean of 2025 to find relative change since 2025
-      mutate(l_ref = mean(l_median[year == 2025])) %>%
+      mutate(l_05_ref = mean(l_05[year == 2025]),
+             l_median_ref = mean(l_median[year == 2025]),
+             l_95_ref = mean(l_95[year == 2025])) %>%
       mutate(scenario = case_when(grepl('126', scenario) ~ 'Best scenario (SSP 1-2.6)',
                                   grepl('245', scenario) ~ 'Good scenario (SSP 2-4.5)',
                                   grepl('370', scenario) ~ 'Bad scenario (SSP 3-7.0)',
@@ -110,6 +110,8 @@ if(file.exists('data/cc-hrsf-projections.rds')) {
   
   cc_proj <-
     list.files('data', 'cc-hrsf-projections-', full.names = TRUE) %>%
+    #' not including `data/cc-hrsf-projections-local-2100.rds`
+    grep(pattern='-local', invert = TRUE, value = TRUE) %>%
     map(readRDS) %>%
     bind_rows()
   saveRDS(cc_proj, 'data/cc-hrsf-projections.rds')
@@ -121,7 +123,7 @@ cc_proj %>%
   ggplot() +
   facet_wrap(~ lab, scales = 'fixed', labeller = label_parsed) +
   geom_hline(yintercept = 1, color = 'black', lty = 'dashed') +
-  geom_line(aes(year, l_median / l_ref, color = scenario), lwd = 1) +
+  geom_line(aes(year, l_median / l_median_ref, color = scenario), lwd = 1) +
   scale_x_continuous(NULL, breaks = c(2025, 2050, 2075, 2100)) +
   scale_y_continuous('Relative change in RSS') +
   scale_color_brewer('Climate change scenario', type = 'div', palette = 5,
@@ -133,15 +135,17 @@ ggsave('figures/rss-local-cc-predictions.png',
 
 cc_proj %>%
[…]

### 2025-06-13 analysis/figures/temperature-hgams-projections.R
fixed pixel-level change

@@ -60,25 +60,25 @@ if(file.exists('data/cc-hgam-projections.rds')) {
               s = weighted.mean(s, w = weight),
               d = weighted.mean(d, w = weight),
               .groups = 'drop') %>%
+    # divide by mean of 2025 to find pixel-level relative change since 2025
+    group_by(species, long, lat) %>%
+    mutate(p = p / mean(p[year == 2025]),
+           s = s / mean(s[year == 2025]),
+           d = d / mean(d[year == 2025])) %>%
+    ungroup() %>%
     # find median and 90% percentile interval of the predicted means
     # not including CIs because averaging them is not straightforward
     group_by(scenario, year, species) %>%
-    summarize(p_lwr_05 = quantile(p, 0.05),
+    summarize(p_lwr_05 = quantile(p, 0.05), # bottom 5%
               s_lwr_05 = quantile(s, 0.05),
               d_lwr_05 = quantile(d, 0.05),
-              p_median = quantile(p, 0.50),
+              p_median = quantile(p, 0.50), # median
               s_median = quantile(s, 0.50),
               d_median = quantile(d, 0.50),
-              p_upr_95 = quantile(p, 0.95),
+              p_upr_95 = quantile(p, 0.95), # top 5%
               s_upr_95 = quantile(s, 0.95),
               d_upr_95 = quantile(d, 0.95),
               .groups = 'drop') %>%
-    # divide by mean of 2025 to find relative change since 2025
-    group_by(species) %>%
-    mutate(p_ref = mean(p_median[year == 2025]),
-           s_ref = mean(s_median[year == 2025]),
-           d_ref = mean(d_median[year == 2025])) %>%
-    ungroup() %>%
     mutate(scenario = case_when(grepl('126', scenario) ~ 'Best scenario (SSP 1-2.6)',
                                 grepl('245', scenario) ~ 'Good scenario (SSP 2-4.5)',
                                 grepl('370', scenario) ~ 'Bad scenario (SSP 3-7.0)',
@@ -129,11 +129,13 @@ if(file.exists('data/cc-hgam-projections.rds')) {
 # make figures ----
 p_p_mov <-
   ggplot(cc_proj, aes(year, p_median, group = scenario)) +
[…]

### 2025-06-13 analysis/figures/temperature-hrsfs-projections.R

@@ -33,6 +33,11 @@ if(file.exists('data/cc-hrsf-projections.rds')) {
   MODEL_FILES <- list.files('models', '^rsf-', full.names = TRUE)
   MODEL_FILES <- MODEL_FILES[! grepl('no-temp', MODEL_FILES)]
   
+  # sort by file size
+  # cannot run in parallel because we are extracting values from rasters
+  MODEL_FILES <- MODEL_FILES[order(file.size(MODEL_FILES))]
+  
+  # find quantiles of percent change across habitat
   map_chr(MODEL_FILES, function(fn) {
     rsf <- readRDS(fn)
     sp <- tolower(fn) %>%
@@ -43,12 +48,12 @@ if(file.exists('data/cc-hrsf-projections.rds')) {
       gsub('\\)', '', .) %>%
       gsub('\\.', '', .)
     
-      tibble(
-        species = fn %>%
-          gsub(pattern = 'models/rsf-', replacement = '', x = .) %>%
-          gsub(pattern = '(-).*', replacement = '', x = ., perl = FALSE),
-        lab = gsub('southern mountain', '(s. mountain)', x = species) %>%
-          gsub('boreal', '(boreal)', x = .)) %>%
+    tibble(
+      species = fn %>%
+        gsub(pattern = 'models/rsf-', replacement = '', x = .) %>%
+        gsub(pattern = '(-).*', replacement = '', x = ., perl = FALSE),
+      lab = gsub('southern mountain', '(s. mountain)', x = species) %>%
+        gsub('boreal', '(boreal)', x = .)) %>%
       mutate(
         lab = SPECIES_LABS[map_int(lab, \(x) which(SPECIES == x))],
         preds = map(
@@ -73,6 +78,8 @@ if(file.exists('data/cc-hrsf-projections.rds')) {
               select(! species)
             
             #' use an `animal` from the model to use `discrete = TRUE`
+            #' even though discretization results in some approximation,
+            #' the projections still takes hours to run
             .nd %>%
               mutate(l = predict.bam(rsf, newdata = .nd, type = 'response',
                                      se.fit = FALSE, exclude = EXCLUDE,
@@ -83,19 +90,19 @@ if(file.exists('data/cc-hrsf-projections.rds')) {
       unnest(preds) %>%
       # calculate habitat suitability for each pixel in each year 
       group_by(lab, scenario, year, long, lat) %>%
[…]

### 2025-06-13 analysis/figures/local-distance-predictions.R
fixed spacing so scale doesn't overlap pixels

@@ -18,7 +18,7 @@ colorRampPalette(RColorBrewer::brewer.pal(11, 'PuOr'))(1e3) %>%
   plot_scheme_colorblind()
 
 if(file.exists('data/cc-hgam-projections-local-2100.rds')) {
-  local_2100_proj <- readRDS('data/cc-hgam-projections-local-2100.rds')
+  cc_proj <- readRDS('data/cc-hgam-projections-local-2100.rds')
 } else {
   # import models
   m_1 <- readRDS('models/binomial-gam.rds')
@@ -111,7 +111,7 @@ make_plot <- function(sp, y_facets = FALSE, get_legend = FALSE,
                          labels = \(x) round(2^x, 2),
                          direction = 1) +
     scale_x_continuous(NULL) +
-    scale_y_continuous(NULL) +
+    scale_y_continuous(NULL, expand = c(0.2, 0)) +
     ggspatial::annotation_scale(style = 'ticks', text_cex = 0.6,
                                 location = 'tr', text_face = 'bold') +
     theme(legend.position = 'none', axis.ticks = element_blank(),
@@ -150,5 +150,5 @@ plot_grid(
           rel_widths = c(1.32, 1.1, 1.67, 1.64, 1.57, 1.58, 1.64)),
   ncol = 1, rel_heights = c(0.05, 1))
 
-ggsave('figures/local-distance-2100.png', width = 15, height = 9.65,
-       units = 'in', dpi = 600, bg = 'white', scale = 1.2)
+ggsave('figures/local-distance-2100.png', width = 15, height = 10,
+       units = 'in', dpi = 600, bg = 'white', scale = 1.5)

### 2025-06-13 analysis/figures/local-rss-predictions.R

@@ -18,7 +18,7 @@ colorRampPalette(RColorBrewer::brewer.pal(11, 'PiYG'))(1e3) %>%
   plot_scheme_colorblind()
 
 if(file.exists('data/cc-hrsf-projections-local-2100.rds')) {
-  local_2100_proj <- readRDS('data/cc-hrsf-projections-local-2100.rds')
+  cc_proj <- readRDS('data/cc-hrsf-projections-local-2100.rds')
 } else {
   # import resource rasters
   f <- rast('data/resource-rasters/forest.tif')
@@ -216,7 +216,7 @@ plot_resources <- function(sp, resource = 'forest_perc') {
     facet_grid(. ~ lab, labeller = label_parsed) +
     geom_raster(aes(x, y, fill = z)) +
     scale_x_continuous(NULL) +
-    scale_y_continuous(NULL) +
+    scale_y_continuous(NULL, expand = c(0.2, 0)) +
     ggspatial::annotation_scale(style = 'ticks', text_cex = 0.6,
                                 location = 'tr', text_face = 'bold') +
     theme(legend.position = 'none', axis.ticks = element_blank(),
@@ -257,5 +257,5 @@ plot_grid(
             rel_widths = c(1.32, 1.1, 1.67, 1.64, 1.57, 1.58, 1.64)),
   ncol = 1, rel_heights = rep(c(0.2, 1), 3))
 
-ggsave('figures/local-resources.png', width = 15, height = 8.5,
+ggsave('figures/local-resources.png', width = 15, height = 10,
        units = 'in', dpi = 600, bg = 'white', scale = 1.5)

### 2025-06-13 writing/appendix-a-cleaning-telemetry-data.Rmd
updated manuscript and figures with new figures

@@
 title: "\\Large The heat is on: Rising temperatures alter when, how, and where large terrestrial mammals move"
+subtitle: "\\Large Appendix A: Cleaning Telemetry Data"
 author:

### 2025-06-13 writing/appendix-b-gams.Rmd
updated manuscript and figures with new figures

@@
 title: "\\Large The heat is on: Rising temperatures alter when, how, and where large terrestrial mammals move"
+subtitle: "\\Large Appendix B: Modeling"
 author:
@@
 
-```{r t-p-mov, fig.cap="\\textbf{The direction and magnitude of changes in probability of moving due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in probability of moving due to changes in temperature within the species' current extent.
-Shaded areas indicate the 90% prediction interval within the range.
-The dashed black lines indicate the mean probability of movement in 2025 across the four scenarios.
-The projections only account for changes movement behavior (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
+```{r t-p-mov, fig.cap="\\textbf{The direction and magnitude of changes in probability of moving due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in probability of moving due to changes in temperature changes in temperature within the area delimitated by the 99.9% Autocorrelated Densitity Kernel Estimates of space use.
+Shaded areas indicate the range of predicted values between the $95^\\text{th}$ and $5^\\text{th}$ percentiles.
+Changes are relative to the predicted mean probabilies of moving in 2025 across the four Shared Socioeconomic Pathways (SSPs).
+If the intervals are fully above the dashed line, at least 90% of the estimated means increased, relative to the each pixel's average of the four climate scenarios in 2025.
+Similar conclusions can be drawn regarding a decrease if the ribbons are below the dashed line.
+The projections only account for changes in movement behaviour and ignore changes in physiology or movement costs."}
 knitr::include_graphics('../figures/p-moving-local-cc-predictions.png')
@@
 
-```{r t-speed, fig.cap="\\textbf{The direction and magnitude of changes in speed due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in speed due to changes in temperature within the species' current extent.
-Shaded areas indicate the 90% prediction interval within the range.
-The dashed black lines indicate the mean speed in 2025 across the four scenarios.
-The projections only account for changes movement behavior (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
+```{r t-speed, fig.cap="\\textbf{The direction and magnitude of changes in speed due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in speed due to changes in temperature within the area delimitated by the 99.9% Autocorrelated Densitity Kernel Estimates of space use.
+Shaded areas indicate the range of predicted values between the $95^\\text{th}$ and $5^\\text{th}$ percentiles.
+Changes are relative to the predicted mean speeds in 2025 across the four Shared Socioeconomic Pathways (SSPs).
+If the intervals are fully above the dashed line, at least 90% of the estimated means increased, relative to the each pixel's average of the four climate scenarios in 2025.
+Similar conclusions can be drawn regarding a decrease if the ribbons are below the dashed line.
+The projections only account for changes in movement behaviour and ignore changes in physiology or movement costs."}
 knitr::include_graphics('../figures/speed-local-cc-predictions.png')
@@
 # Effects of temperature on habitat selection
+
+```{r kept-hrsf}
+readr::read_csv('../data/tracking-data/telemetry-metadata.csv',
+                show_col_types = FALSE) %>%
+  mutate(keep_for_RSF = if_else(keep_for_RSF == 'yes', 'yes', 'no')) %>%
+  group_by(species, keep_for_RSF) %>%
+  summarize(n = n()) %>%
+  mutate(prop = n / sum(n)) %>%
+  # fix species names
+  mutate(species = case_when(
+    species == 'Rangifer tarandus (s. mountain)' ~ 'R. tarandus (sm)',
+    species == 'Rangifer tarandus (boreal)' ~ 'R. tarandus (b)',
+    .default = species)) %>%
+  mutate(species = paste0('\\emph{', species, '}') %>%
+           gsub(')}', ')', .) %>%
+           gsub(' \\(', '} \\(', .)) %>%
+  bind_rows(.,
+            group_by(., keep_for_RSF) %>%
+              summarize(species = 'Total',
+                        n = sum(n)) %>%
+              mutate(prop = n / sum(n))) %>%
+  rename(Species = species, `Kept` = keep_for_RSF, Proportion = prop) %>%
+  knitr::kable(format = 'markdown', caption = "Number of animals (n) and proportion of animals (Proportion) used or excluded (Kept) when fitting the hierarchical resource selection function.
+The models only used animals that showed stationary range residency or two clear seasonal ranges within the tracking period.
+We excluded animals with insufficient data to estimate a range size and animals that migrated but did not establish clear ranges within the tracking period.")
+```
+
+```{r resources, fig.cap="Resource rasters used for each species' hierarchical resource selection function."}
+knitr::include_graphics('../figures/local-resources.png')
+```
 

### 2025-06-13 writing/manuscript.Rmd
updated manuscript and figures with new figures

@@
 Consequently, it is still unclear how mammals will adapt their movement behavior in the 21^st^ century.
-We address this gap by using continuous-time movement models and hierarchical generalized additive models to quantify the effects of temperature on the probability of moving, the speed when moving, and the habitat selection of six species throughout British Columbia, Canada, between 1998 and 2023.
+We address this gap by using continuous-time movement models and hierarchical generalized additive models to quantify the effects of temperature on the probability of moving, the speed when moving, and the habitat selection of six species throughout western Canada between 1998 and 2023.
 We show that temperature is a strong determinant of when, how, and where mammals move, and we predict that climate change will impact mammals' movement rates and habitat selection as temperatures warm throughout the 21^st^ century.
@@
 This paper provides a framework for quantifying the effects of proximal air temperature on animal movement using quantitative methods that are flexible and statistically robust.
-We present results for six large-bodied terrestrial mammal species in British Columbia (BC), Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
+We present results for six large-bodied terrestrial mammal species in western Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
 Using over 25 years of telemetry data throughout a large spatial range of British Columbia and hourly air temperature data, we estimate how these mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature.
@@
 
-\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)).
+\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public British Columbia Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)).
 ***MISSING INFO ON OTHER TELEMETRY DATA.*** From the full set of telemetry data, a total of 2396 GPS locations (0.16%, including collar calibration data) were removed using diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points.
@@
 
-We obtained rasters of projected monthly average temperature at a 0.08$^\circ$ resolution in BC from 2020 to 2100 under the different SSPs via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`.
-Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 data for BC from 1998 to 2023 [extremes included, @hersbach_era5_2023] to calculate within-month variance in temperature, which we defined as the variance within a given pixel, month, and year.
+We obtained rasters of projected monthly average temperature for the study region at a 0.08$^\circ$ spatial resolution from 2020 to 2100 under the different SSPs via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`.
+Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 data for western Canada from 1998 to 2023 [extremes included, @hersbach_era5_2023] to calculate within-month variance in temperature, which we defined as the variance within a given pixel, month, and year.
 We then modeled the estimated variance using a GAM for location and scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017].
@@
 \noindent We estimated the effects of temperature on each species' selection for percent forest cover (`forest_perc`), elevation (`elevation_m`), and distance from water (`dist_water_m`) by fitting a Hierarchical Resource Selection Function for each species using an HGAM with a Poisson family of distributions and log link function [Appendix B$;$ @aarts_estimating_2008].
-We accounted for the spatiotemporal autocorrelation in the telemetry locations by weighting each point based on the telemetry's Autocorrelated Kernel Density Estimate [@fleming_new_2017; @noonan_comprehensive_2019] to produce estimates of second- and third-order habitat selection [@johnson_comparison_1980].
+After removing non-resident individuals (Table B1), we accounted for the spatiotemporal autocorrelation in the telemetry locations by weighting each point based on the telemetry's Autocorrelated Kernel Density Estimate [@fleming_new_2017; @noonan_comprehensive_2019] to produce estimates of second- and third-order habitat selection [@johnson_comparison_1980].
 While we recognize there are other important drivers of habitat selection (e.g., forest age, forest type, terrain ruggedness, prey availability), we decided to only use these three proxies to produce results that are easily comparable across species.
@@
 The number of quadrature locations greatly outnumbered the number of observed locations (Fig.
-B11), especially after accounting for weighting based on the degree of autocorrelation and the number of home range crossings (Fig.
-B12).
+B12), especially after accounting for weighting based on the degree of autocorrelation and the number of home range crossings (Fig.
+B13).
 
@@
 There was relatively strong agreement between models with and without temperature (Figs.
-B3, and B13), but including temperature always resulted in better fits.
-All analyses of deviance showed that temperature significantly affected movement behavior (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; Appendix B).
+B3, and B14), but including temperature always resulted in better fits.
+All analyses of deviance showed that temperature significantly affected movement behavior (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; Table B2 and following pages in Appendix B).
 Accounting for temperature resulted in a simpler HRSF for mountain goats (lower model degrees of freedom) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.
@@
 Again, estimated RSS values were generally most uncertain at extreme temperatures (Fig.
-B14).
+B15).
 
@@
 
-\noindent Species varied in both the direction and magnitude of predicted change in distance traveled, but worse climate-change scenarios always corresponded to greater absolute changes in behaviour (Figs.
-\@ref(fig:t-dist), B9, and B10).
-Under the best-case scenario, absolute changes by 2100 were small (approximately 0% to 4%), while under the worst-case scenario absolute changes ranged from ~2% (grizzly bear) to ~25% (cougars).
-However, while the models estimated that grizzly bears would move substantially less (if at all) in winter (Figs.
+\noindent The predicted changes in distance traveled varied across species in both direction and magnitude, but worse climate-change SSPs always corresponded to greater absolute changes (Fig.
+\@ref(fig:t-dist)).
+Additionally, species that were predicted to move less often did not necessarily have lower speeds when moving, and vice versa (Figs.
+B9 and B10).
+Overall, absolute changes by 2100 under the best-case SSP were small (approximately 0% to +4%), while under the worst-case SSP absolute changes ranged from ~2% (grizzly bears) to ~24% (cougars).
+Notably, while the models estimated that grizzly bears would move substantially less (if at all) in winter (Figs.
 \@ref(fig:dist), B4-B6), the models did not explicitly account for changes in hibernation phenology.
@@
 \@ref(fig:s-dist)).
-Again, absolute changes were generally greatest under worse scenarios, but the direction of change at each location also varied across scenarios (most visible in cougars).
+Again, absolute changes were generally greatest under worse SSPs, but the direction of change at each location also varied across SSPs (most visible in cougars).
 
 ```{r, eval=FALSE, echo=FALSE}
-# local predictions
+library('dplyr')
+
 readRDS('../data/cc-hgam-projections.rds') %>%
@@
   filter(grepl('Best', scenario), year > 2095) %>%
-  mutate(d_median = d_median / d_ref) %>%
   summarise(mean = mean(d_median)) %>%
@@
   filter(grepl('Worst', scenario), year > 2095) %>%
-  mutate(d_median = d_median / d_ref) %>%
-  summarise(mean = mean(d_median)) %>%
-  mutate(mean = abs(mean - 1) * 100) %>%
-  arrange(mean)
-
-# BC predictions
-readRDS('../data/cc-hgam-bc-projections-albers.rds') %>%
-  group_by(lab) %>%
-  filter(grepl('Best', scenario)) %>%
-  summarise(mean = mean(d)) %>%
-  mutate(mean = abs(mean - 1) * 100) %>%
-  arrange(mean)
-
-readRDS('../data/cc-hgam-projections.rds') %>%
-  group_by(species) %>%
-  filter(grepl('Worst', scenario), year > 2095) %>%
-  mutate(d_median = d_median / d_ref) %>%
   summarise(mean = mean(d_median)) %>%
@@
 
-```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in distance travelled (probability of moving times speed when moving) due to changes in temperature within the species' current extent.
-Shaded areas indicate the 90% prediction interval within the range.
+```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in distance travelled (probability of moving times speed when moving) due to changes in temperature within the area delimitated by the 99.9% Autocorrelated Densitity Kernel Estimates of space use.
+Shaded areas indicate the range of predicted values between the $95^\\text{th}$ and $5^\\text{th}$ percentiles.
 Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs).
-The projections only account for changes in movement in behaviour (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
+If the intervals are fully above the dashed line, at least 90% of the estimated means increased, relative to the each pixel's average of the four climate SSPs in 2025.
+Similar conclusions can be drawn regarding a decrease if the ribbons are below the dashed line.
+The projections only account for changes in movement behaviour (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
 knitr::include_graphics('../figures/distance-travelled-local-cc-predictions.png')
@@
 
-```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each pixel's average distance in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase.
+```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each pixel's average distance in 2025 across all four SSPs, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase.
 For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.50 to 1.56).
 The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest density."}
+if(FALSE) range(readRDS('../data/cc-hgam-projections-local-2100.rds')$d)
 knitr::include_graphics('../figures/local-distance-2100.png')
@@
 
-We do not predict any of the species to exhibit stronger selection for their current range, irrespective of climate change scenario.
-Changes in RSS by 2100 were between 0% and -3% under the best-case scenario, but they ranged from -8% to -14% under the worst-case scenario (Fig.
-\@ref(fig:t-hrsfs)).
-As with distance traveled, the change in RSS in 2100 varied throughout BC for all species (Fig.
+Median RSS was projected to decrease over time within the each species' observed range, but changes were stronger under worse SSPs (Fig.
 \@ref(fig:s-hrsfs)).
-All species were predicted to exhibit a decrease in RSS throughout most of BC, with the exception of boreal caribou, which had a predicted increase in approximately 67% of the province, including their current range.
-The increase was in great part due to the estimated increase in selection for high-elevation areas that are well outside the population's current range.
+Decreases were most pronounced in the bottom fifth percentile and most severe for elk, southern mountain caribou, and mountain goats.
+Of all the species, only cougars showed a clear increase in RSS for areas within their current range.
+The change in RSS between 2025 and 2100 varied spatially for all species (Fig.
+\@ref(fig:s-hrsfs)).
+All species were predicted to exhibit a decrease in RSS throughout most of their observed habitats, although cougars were predicted to increase their selection for patches with higher altitude.
 
 ```{r, eval=FALSE}
-# find species that decrease in RSS in less than half of BC
 library('dplyr')
-readRDS('data/cc-rss-bc-projections-albers.rds') %>%
-  group_by(lab, scenario) %>%
-  summarize(p_decrease = mean(lambda < 1)) %>%
-  filter(p_decrease < 0.5) %>%
-  mutate(p_increase = 1 - p_decrease)
+
+readRDS('../data/cc-hrsf-projections.rds') %>%
+  group_by(lab) %>%
+  filter(grepl('Best', scenario), year > 2095) %>%
+  summarise(mean = mean(l_median)) %>%
+  mutate(mean = (mean - 1) * 100,
+         decreased = mean < 0) %>%
+  arrange(mean)
+
+readRDS('../data/cc-hrsf-projections.rds') %>%
+  group_by(lab) %>%
+  filter(grepl('Worst', scenario), year > 2095) %>%
+    summarise(mean = mean(l_median)) %>%
+  mutate(mean = (mean - 1) * 100,
+         decreased = mean < 0) %>%
+  arrange(mean)
 ```
 
-```{r t-hrsfs, fig.cap="\\textbf{All species are expected to exhibit a decrease in relative selection strength (RSS) for their current range, irrespective of RSS quantile or climate change scenario, but decreases are stronger in the worse-case scenario.} Lines indicate the projected change in RSS for the bottom 5\\%, median, and top 5\\% RSS values within the species' current extent.
-Changes are relative to the mean RSS for each quantile in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
-knitr::include_graphics('../figures/rss-local-cc-predictions-quantiles.png')
+```{r t-hrsfs, fig.cap="\\textbf{All species are expected to exhibit a decrease in relative selection strength (RSS) for their current range, irrespective of RSS quantile or climate change scenario, but decreases are stronger in the worse-case scenario.} Lines indicate the projected change in median RSS within the species' current extent, while the ribbons indicate the range between the $95^\\text{th}$ and $5^\\text{th}$ percentiles in RSS.
+Changes are relative to each location's mean RSS in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
+knitr::include_graphics('../figures/rss-local-cc-predictions.png')
 ```
 
-```{r s-hrsfs, fig.cap="\\textbf{Climate change will impact each species' habitat relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase.
-For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.26 to 117).
-The increase in boreal caribou's RSS for high elevations at warmer temperatures was highly uncertain and often not credible at the 95\\% level."}
+```{r s-hrsfs, fig.cap="\\textbf{Climate change will impact each species' habitat relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios, such that values < 1 indicate a decrease and values > 1 indicate an increase.
+For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.66 to 1.86)."}
+if(FALSE) range(readRDS('../data/cc-hrsf-projections-local-2100.rds')$l)
 knitr::include_graphics('../figures/local-rss-2100.png')
@@
 However, predicting animal behaviour becomes increasingly complicated as the conditions animals are exposed to deviate from current, typical conditions, especially when responses are nonlinear and data are sparse.
-Consequently, we do not present our results as a definitive guide to how mammals in BC will respond to climate change.
+Consequently, we do not present our results as a definitive guide to how mammals in western Canada will respond to climate change.
 Instead, we hope they serve as a starting point to (1) demonstrate that mammals' movement rates and habitat selection depend on temperature and (2) how one can estimate mammals' changes in movement behaviour due to climate change.
@@
 Recent events in global politics, including global conflicts [@mcnutt_scientists_2022; @tollefson_what_2022] and the rapid rise in generative machine learning [more commonly: "AI", see @roumeliotis_chatgpt_2023; @van_noorden_chatgpt_2023; @deepseek-ai_deepseek-r1_2025] and its prohibitive energy and water demands [@li_making_2023] and carbon footprint [@luccioni_estimating_2024], have placed doubt on the feasibility of reaching climate change goals that seemed achievable in the past decade [@united_nations_environment_programme_emissions_2024].
-Mitigating climate change and its consequences will require extensive and widespread collaborative effort [@huang_achieving_2021].
 Similarly, protecting a third of the world's ecosystems [@convention_on_biological_diversity_154_2022] appears less achievable as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023].
-Achieving the "30 by 30" goal will require international collaboration as well as active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023].
+Achieving the "30 by 30" goal will require international collaboration [@huang_achieving_2021] as well as active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023].
 Understanding the consequences of climate change on mammals' movement behaviour is a first step towards proactively responding to how mammals will respond to human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018].
@@
 
+attraction for disconnected patches of high elevation could lead to fragmentation or traps
+
 ### *Rangifer tarandus* (boreal)
+
+Increasing temperatures increased RSS for both elevations near ~500 m as well as > 1,000 m, but such a selection is only possible if: (1) such habitats exist in the animals' range, and (2) crossing the 500 m to > 1,000 m regions is safe and attractive
 
@@
 
-Increasing temperatures increased RSS for both elevations near ~500 m as well as > 1,000 m, but such a selection is only possible if: (1) such habitats exist in the animals' range, and (2) crossing the 500 m to > 1,000 m regions is safe and attractive
+HRSFs do not account for migratory behavior due to reproduction or predation risk
 

### 2025-08-28 writing/appendix-a-cleaning-telemetry-data.Rmd
removed cyan links from references

@@
 subparagraph: true # needed for \usepackage[small]{titlesec}
-urlcolor: cyan
 output:
@@
 In this appendix, we illustrate the cleaning procedure for the telemetry data.
-For the sake of brevity, we only show the code and outputs for some elk (*Cervus canadensis*); the full script is available at  [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R).
+For the sake of brevity, we only show the code and outputs for some elk (*Cervus canadensis*); the full script is available at  [\textcolor{cyan}{github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R}](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R).
 We suggest looking at the script since this appendix does not include the interactive maps that are generated by `plot_adj(..., map = TRUE)`.
@@
 We start by attaching the necessary packages and sourcing some custom functions.
-The custom functions are all available on GitHub in the `functions` folder of the repository located at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).
+The custom functions are all available on GitHub in the `functions` folder of the repository located at [\textcolor{cyan}{github.com/QuantitativeEcologyLab/bc-mammals-temperature}](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).
 While this appendix only shows the removal of a single outlier point, the appendix details the approach we used to clean the telemetry data.

### 2025-08-28 writing/appendix-b-gams.Rmd
re-knit appendix with updated figures

@@
 
-```{r pred-vs-pred-hrsfs, fig.cap="Hexplot of the fitted values from the HRSFs with and without including temperature."}
+```{r pred-vs-pred-hrsfs, fig.cap="Hexplot of the fitted values from the HRSFs with and without including temperature. For ease of visibility, values < -20 (0.41%) were set to -20."}
 knitr::include_graphics('../figures/hrsf-with-without-temp-prediction-agreement.png')
@@
 3 in the main text).
-The grey contours indicate the extent of each speices' observed data.
+The black contours indicate the extent of each speices' observed data.
 For ease of readability, the color bar values were capped to a maximum of 3 (original maximum was 7.00)."}

### 2025-08-28 writing/manuscript.Rmd
started reviewing manuscript

@@
 ---
-title: "\\Large The heat is on: Rising temperatures alter when, how, and where large terrestrial mammals move"
+title: "\\Large The heat is on: Rising temperatures alter how and where mammals move"
 author:
@@
 ```{r abstract, echo=FALSE}
-abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on fitness, habitat selection, and community structure.
-Previous studies have quantitatively estimated the effects of temperature on mammalian movement behaviour, but few have estimated the effects of future climate change.
+abstract <- "Widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness, habitat selection, and community structure.
+Previous studies have estimated the effects of temperature on mammalian movement behaviour, but many did not disentangle them from seasonal behavior cycles.
 Consequently, it is still unclear how mammals will adapt their movement behavior in the 21^st^ century.
-We address this gap by using continuous-time movement models and hierarchical generalized additive models to quantify the effects of temperature on the probability of moving, the speed when moving, and the habitat selection of six species throughout western Canada between 1998 and 2023.
-We show that temperature is a strong determinant of when, how, and where mammals move, and we predict that climate change will impact mammals' movement rates and habitat selection as temperatures warm throughout the 21^st^ century.
-We found the effects of temperature on movement rates varied across species, so we suggest that species' estimated responses be interpreted holistically when making inferences about how climate change will impact ecological communities.
-In contrast, habitat selection strength decreased for all species within their observed ranges, suggesting that all six species may shift their range in the coming decades.
-As climate change exposes mammals to novel environmental conditions, predicting changes in animal behavior and community structure will become crucial for effective and proactive conservation.
-We thus conclude by providing suggestions for conservation and future research, with particular attention to the study design, statistical modeling, and forecasting.
-We provide detailed `R` code and appendices so that other researchers can implement our methods on their own datasets."
+We address this gap by using continuous-time movement models and hierarchical generalized additive models to quantify the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species throughout western Canada between 1998 and 2023.
+We show that temperature is a strong determinant of how and where mammals move, and we predict that climate change will have significant impacts on mammals' movement rates and habitat selection throughout the 21^st^ century.
+We found that species did not respond uniformly to rising temperatures, so community-level responses will likely be complex as some species move more and others less.
+Median habitat selection strength decreased for all species, but some species showed increased selection strength for higher altitudes.
+Consequently, our results suggest that all six species may shift their range in the coming decades as their selection strength for their current ranges weakens.
+As climate change exposes mammals to novel environmental conditions, predicting changes in animal behavior and community structure will become crucial for effectively and proactively understanding mammalian movement behavior and protecting high-quality habitat.
+We thus present our work using a flexible approach that can be adapted for different species and spatiotemporal scales, and we conclude by providing suggestions for future research, with particular attention to the study design, statistical modeling, and forecasting."
 if(FALSE){
@@
 
+\
+
 \clearpage
@@
 
-<!--
+# other refs
 
-other refs
-
+* @gerlich_species_2025: effects of temperature on fly movement in the arctic
+* "Potential changes in climate indices in Alberta under projected global warming of 1.5–5 $^\circ$C" (https://www.sciencedirect.com/science/article/pii/S2214581823000770#fig0015)
 * Global vulnerability of marine mammals to global warming: https://doi.org/10.1038/s41598-019-57280-3
-
--->
+* marine turtles shift and lose habitat with temperature: https://www.science.org/doi/10.1126/sciadv.adw4495
 
@@
 \noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008].
-Animals move to maximize resource availability [e.g., food, water, shelter, see: @charnov_optimal_1976; @kacelnik_ideal_1992], optimize habitat use [@winter_forecasting_2024], and escape predation [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024].
+Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020].
 However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023].
-Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [e.g., @hall_microrefuges_2016].
-Additionally, temperature also has indirect effects on movement through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010], snow depth and density [@montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the frequency and intensity of extreme events [@zhang_increased_2023; @bunnell_global_2011], including forest fires, droughts, and novel phenology [@zurowski_summer_2023; @duncan_predicting_2012; @bunnell_global_2011; @merkle_large_2016; @slatyer_ecological_2022].
-Changes in temperature thus affects how animals move [e.g., movement frequency and speed -- their movement behaviour: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].
+Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [@hall_microrefuges_2016].
+
+Temperature also has indirect effects on movement through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010], snow depth and density [@montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], altered phenology [@bunnell_global_2011; ], and the frequency and intensity of extreme events [@zhang_increased_2023; @bunnell_global_2011], including forest fires, droughts, and extreme weather [@zurowski_summer_2023; @duncan_predicting_2012; @bunnell_global_2011; @merkle_large_2016; @slatyer_ecological_2022; @berger_climate_2018].
+Changes in temperature thus affect how animals move [e.g., movement frequency and speed -- their movement behaviour: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].
 
@@
 Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to respond to change in a proactive manner [@sawyer_identifying_2009; @mccain_assessing_2019].
-The Canadian province of British Columbia is currently experiencing rapid widespread warming [@dierauer_climate_2021; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024], and more frequent and intense extreme events [@zhang_increased_2023], including forest fires [@zurowski_summer_2023].
+Western Canada is currently experiencing rapid widespread warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme events [@zhang_increased_2023], including forest fires [@zurowski_summer_2023].
 As we approach the deadline for the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [@convention_on_biological_diversity_154_2022], decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness.
@@
 We present results for six large-bodied terrestrial mammal species in western Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
-Using over 25 years of telemetry data throughout a large spatial range of British Columbia and hourly air temperature data, we estimate how these mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature.
+Using over 25 years of telemetry data throughout a large spatial range of western Canada and hourly air temperature data, we estimate how these mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature.
 We then pair the estimated responses with climate change projections to forecast the species' movement during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017].
@@
 We obtained rasters of projected monthly average temperature for the study region at a 0.08$^\circ$ spatial resolution from 2020 to 2100 under the different SSPs via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`.
-Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 data for western Canada from 1998 to 2023 [extremes included, @hersbach_era5_2023] to calculate within-month variance in temperature, which we defined as the variance within a given pixel, month, and year.
+Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 reanalized data for western Canada from 1998 to 2023 [extremes included, @hersbach_era5_2023] to calculate within-month variance in temperature, which we defined as the variance within a given pixel, month, and year.
 We then modeled the estimated variance using a GAM for location and scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017].
@@
 Smooth effects of percent forest cover, elevation, and distance to water accounted for the species-level average selection strength for each resource.
-A random effect for each individual animal (`s(animal, bs = 're')`) corrected for uneven sampling across individuals, while factor smooth interaction terms (`bs = 'fs'`) accounted for individual-level resource selection (i.e., individual-level deviations from the species-level average).
+A random effect for each individual animal (`s(animal, bs = 're')`) corrected for uneven sampling across individuals, while factor smooth interaction terms (`bs = 'fs'`) accounted for individual-level resource selection [i.e., individual-level deviations from the species-level average$;$ @jeltsch_need_2025].
 Tensor interaction product terms (`ti()`) of the three resources and temperature estimated the change in resource selection at different temperatures.
@@
 
-```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in distance travelled (probability of moving times speed when moving) due to changes in temperature within the area delimitated by the 99.9% Autocorrelated Densitity Kernel Estimates of space use.
+```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in distance travelled (probability of moving times speed when moving) due to changes in temperature within each species' observed extent.
 Shaded areas indicate the range of predicted values between the $95^\\text{th}$ and $5^\\text{th}$ percentiles.
@@
 
-```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each pixel's average distance in 2025 across all four SSPs, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase.
+```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each pixel's average distance in 2025 across all four scenarios (i.e., not relative to other pixels).
+Values < 1 indicate a decrease, and values > 1 indicate an increase.
 For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.50 to 1.56).
-The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest density."}
+The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest density.
+All maps are shown with a BC Albers projection (EPSG:3005)."}
 if(FALSE) range(readRDS('../data/cc-hgam-projections-local-2100.rds')$d)
@@
 
-```{r t-hrsfs, fig.cap="\\textbf{All species are expected to exhibit a decrease in relative selection strength (RSS) for their current range, irrespective of RSS quantile or climate change scenario, but decreases are stronger in the worse-case scenario.} Lines indicate the projected change in median RSS within the species' current extent, while the ribbons indicate the range between the $95^\\text{th}$ and $5^\\text{th}$ percentiles in RSS.
+```{r t-hrsfs, fig.cap="\\textbf{All species are expected to exhibit a decrease in median relative selection strength (RSS) for their current range, irrespective of climate change scenario, but decreases are stronger in the worse-case scenario.} Lines indicate the projected change in median RSS within each species' observed extent within BC, while the ribbons indicate the range between the $95^\\text{th}$ and $5^\\text{th}$ percentiles in RSS.
 Changes are relative to each location's mean RSS in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
@@
 
-```{r s-hrsfs, fig.cap="\\textbf{Climate change will impact each species' habitat relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios, such that values < 1 indicate a decrease and values > 1 indicate an increase.
-For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.66 to 1.86)."}
+```{r s-hrsfs, fig.cap="\\textbf{Climate change will impact each species' habitat relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios (i.e., not relative to other pixels).
+Values < 1 indicate a decrease and values > 1 indicate an increase.
+For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.66 to 1.86).
+All maps are shown with a BC Albers projection (EPSG:3005)."}
 if(FALSE) range(readRDS('../data/cc-hrsf-projections-local-2100.rds')$l)
@@
 
-Recent events in global politics, including global conflicts [@mcnutt_scientists_2022; @tollefson_what_2022] and the rapid rise in generative machine learning [more commonly: "AI", see @roumeliotis_chatgpt_2023; @van_noorden_chatgpt_2023; @deepseek-ai_deepseek-r1_2025] and its prohibitive energy and water demands [@li_making_2023] and carbon footprint [@luccioni_estimating_2024], have placed doubt on the feasibility of reaching climate change goals that seemed achievable in the past decade [@united_nations_environment_programme_emissions_2024].
-Similarly, protecting a third of the world's ecosystems [@convention_on_biological_diversity_154_2022] appears less achievable as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023].
-Achieving the "30 by 30" goal will require international collaboration [@huang_achieving_2021] as well as active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023].
+*Achieving the "30 by 30" goal will require international collaboration [@huang_achieving_2021] as well as active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023].
 Understanding the consequences of climate change on mammals' movement behaviour is a first step towards proactively responding to how mammals will respond to human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018].
 In the following sections, we discuss the implications of our results in more detail.
-We then expand on consequences for conservation during the 21^st^ century and considerations for future studies.
+We then expand on consequences for conservation during the 21^st^ century and considerations for future studies.*
 
@@
 Throughout the year, they adapted their tolerance to temperature and moved less when it was relatively hot (for a given time of year), especially in spring and summer.
-The strong reduction in the mid-day movement rates of wolves, elk, cougars, and boreal caribou when summer temperatures were above 20$^\circ$C suggests that the increasingly common and intense heat waves across British Columbia will have community-wide impacts on movement rates, encounter rates, and potentially community structure [@martinez-garcia_how_2020].
+The strong reduction in the mid-day movement rates of wolves, elk, cougars, and boreal caribou when summer temperatures were above 20$^\circ$C suggests that the increasingly common and intense heat waves across western Canada will have community-wide impacts on movement rates, encounter rates, and potentially community structure [@martinez-garcia_how_2020].
 More work is necessary on quantifying interspecific responses to temperature, including the effects of temperature on predation rates [but see: @cunningham_opportunity_2021; @glass_trade-off_2021; @brivio_seeking_2024].
@@
 For example, the width of the prediction intervals in Fig.
-\@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas.
-Population-level means that ignore such variation can greatly limit the efficacy of conservation projects [@muff_accounting_2020; @mortlock_sleep_2024].
+\@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas [@jeltsch_need_2025].
+Population-level means that ignore such variation can greatly limit the efficacy of conservation projects [@muff_accounting_2020; @mortlock_sleep_2024; @silva_too_2025].
 Additionally, accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage and prey availability [@mezzini_how_2025], encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@smith_densitydependent_2023], competitive pressure [@torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019].

### 2025-08-29 writing/manuscript.Rmd
started cleaning up intro

@@
 * marine turtles shift and lose habitat with temperature: https://www.science.org/doi/10.1126/sciadv.adw4495
-
+* changes in temperature change phelogy and movement behavior as a consequence @gerlich_species_2025
+* @walker_global_2019: Global climate change and invariable photoperiods: {A} mismatch that jeopardizes animal fitness
 \clearpage
@@
 
-<!-- movement is a convenient way to respond to change, but movement depends on temperature -->
+<!-- movement is a convenient way for ANIMALS to respond to change, but movement depends on temperature -->
 
 \noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008].
-Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020].
-However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023].
+Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024; @schwerdt_ecological_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020].
+However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and the risk of hyperthermia [@hetem_responses_2014; @dyer_travel_2023].
 Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [@hall_microrefuges_2016].
+By altering mammals' movement speed and habitat use [jointly, their movement behavior: @nathan_movement_2008; @nathan_big-data_2022], ambient temperature also affects the consequences of movement behavior, such as encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, or competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024].
+Behavioral changes to temperature are exacerbated in extreme cold [@wilson_stenotherms_2001; @berger_climate_2018] and extreme heat [@powers_hovering_2017; @alston_temperature_2020; @giroux_activity_2023; @verzuh_behavioural_2023].
 
-Temperature also has indirect effects on movement through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010], snow depth and density [@montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], altered phenology [@bunnell_global_2011; ], and the frequency and intensity of extreme events [@zhang_increased_2023; @bunnell_global_2011], including forest fires, droughts, and extreme weather [@zurowski_summer_2023; @duncan_predicting_2012; @bunnell_global_2011; @merkle_large_2016; @slatyer_ecological_2022; @berger_climate_2018].
-Changes in temperature thus affect how animals move [e.g., movement frequency and speed -- their movement behaviour: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].
+<!-- mammals  -->
 
-<!-- since movement depends on temperature, the consequences of movement do, too -->
-
-By impacting animal movement behavior, temperature also affects the consequences of movement, such as encounter rates with resources [@mezzini_how_2025], other animals [@martinez-garcia_how_2020], and humans [@weststrate_how_2024], and these changes are exacerbated at extreme temperatures [@wilson_stenotherms_2001; @berger_climate_2018; @giroux_activity_2023].
-But what counts as "extreme" varies across species: Species that experience narrow temperature ranges tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while thermal generalists can adapt more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024].
-However, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019], affect many factors [e.g., temperature, phenology, and resource availability: @polazzo_measuring_2024], or are too rapid to respond within the span of a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].
+Mammals are particularly susceptible to adverse effects from excessive heat [@sherwood_adaptability_2010].
+While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more rapid and serious physiological damage [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021].
+Due to mammals' difficulty at dissipating heat, particularly when ambient temperatures are near or above their body temperature, they are particularly sensitive to extreme heat, particularly if they have a large body size [@dyer_travel_2023; @verzuh_behavioural_2023].
+Furthermore, species that experience narrow temperature ranges (i.e., $`$temperature specialists') tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while species that experience wide ranges of temperature variation (i.e., $`$temperature genralists') can adapt more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024].
+Still, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019; @rantanen_arctic_2022], affect many factors at once [e.g., temperature, snow depth, ecosystem phenology, and resource availability: @polazzo_measuring_2024], or are too rapid and severe to respond within a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].
 
@@
 Failing to understand how changes in temperature will affect mammalian movement behaviour and habitat selection will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.
+
+***tie to text above:***
+
+***...*** Temperature also has indirect effects on movement through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010], snow depth and density [@montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], altered phenology [@bunnell_global_2011; ], and the frequency and intensity of extreme events [@zhang_increased_2023; @bunnell_global_2011], including forest fires, droughts, and extreme weather [@zurowski_summer_2023; @duncan_predicting_2012; @bunnell_global_2011; @merkle_large_2016; @slatyer_ecological_2022; @berger_climate_2018].
+Changes in temperature thus affect how animals move [e.g., movement frequency and speed -- their movement behaviour: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].
 

### 2025-08-30 writing/manuscript.Rmd
finished editing intro

@@
 \noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008].
-Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024; @schwerdt_ecological_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020].
+Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @merkle_large_2016; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024; @schwerdt_ecological_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020].
 However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and the risk of hyperthermia [@hetem_responses_2014; @dyer_travel_2023].
 Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [@hall_microrefuges_2016].
-By altering mammals' movement speed and habitat use [jointly, their movement behavior: @nathan_movement_2008; @nathan_big-data_2022], ambient temperature also affects the consequences of movement behavior, such as encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, or competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024].
+By altering mammals' movement speed and habitat use [jointly, their movement behavior: @nathan_movement_2008; @nathan_big-data_2022], ambient temperature also affects the consequences of movement behavior [@alston_temperature_2020; @giroux_activity_2023], such as encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, or competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024].
 Behavioral changes to temperature are exacerbated in extreme cold [@wilson_stenotherms_2001; @berger_climate_2018] and extreme heat [@powers_hovering_2017; @alston_temperature_2020; @giroux_activity_2023; @verzuh_behavioural_2023].
@@
 Mammals are particularly susceptible to adverse effects from excessive heat [@sherwood_adaptability_2010].
-While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more rapid and serious physiological damage [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021].
+While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more rapid and serious physiological damage [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021; @newediuk_climate_2024].
 Due to mammals' difficulty at dissipating heat, particularly when ambient temperatures are near or above their body temperature, they are particularly sensitive to extreme heat, particularly if they have a large body size [@dyer_travel_2023; @verzuh_behavioural_2023].
@@
 In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@ipcc_climate_2023].
-Over the next 100 years, these changes are expected to worsen and continue to impact mammals' fitness, movement behaviour, and habitat selection as animals cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and more frequent and intense extreme events [@bunnell_global_2011], along with increasing anthropogenic pressure [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024].
-Consequently, it remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially when populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018].
-Uncertainty in future conditions and their consequences on ecosystems present a need for a comprehensive understanding of (1) the direct effects of temperature on animal movement behaviour and (2) the implications such changes and uncertainty will have on populations and species' adaptability and resiliency, as well as overall community structure.
+Over the next 100 years, these changes are expected to worsen and continue to impact mammals' fitness, movement behaviour, and habitat selection as animals cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and more frequent and intense extreme events [@bunnell_global_2011], along with growing anthropogenic habitat loss, modification, and fragmentation [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024].
+It remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially if populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018].
+Preparing for and responding to future changes require a comprehensive understanding of how temperature will affect how and where mammals move, as well as the implications such changes will have on species' adaptability and overall community structure.
 Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], movement behaviour [@melin_effect_2023], thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [e.g., body size, activity time, and elevational and latitudinal distribution @mccain_assessing_2019].
-However, it is still unclear how mammals will adapt their small-scale movement and habitat selection to changes in temperature during the 21^st^ century [@ipcc_climate_2023; but see @hetem_responses_2014; @winter_forecasting_2024].
+However, there remains a need for fine-scale estimates of the effects of temperature on mammals' movement behavior and quantitative predictions of how their behavior will change in the current century.
 
-<!-- why is understanding the present and predicting into the future necessary -->
+<!-- zoom in to western canada -->
 
-Understanding the direct and indirect impacts of temperature on mammalian movement is essential for decision makers to respond to change in a proactive manner [@sawyer_identifying_2009; @mccain_assessing_2019].
 Western Canada is currently experiencing rapid widespread warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme events [@zhang_increased_2023], including forest fires [@zurowski_summer_2023].
-As we approach the deadline for the "30 by 30" conservation initiative, a global effort to conserve 30% of the world's lands and oceans by 2030 [@convention_on_biological_diversity_154_2022], decision makers must understand which areas will be of greatest value for maximizing the project's effectiveness.
-Predicting how climate change will affect how, when, and where animals will move is necessary for identifying landscape dynamics and value will evolve throughout the century.
-Failing to understand how changes in temperature will affect mammalian movement behaviour and habitat selection will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.
-
-***tie to text above:***
-
-***...*** Temperature also has indirect effects on movement through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010], snow depth and density [@montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], altered phenology [@bunnell_global_2011; ], and the frequency and intensity of extreme events [@zhang_increased_2023; @bunnell_global_2011], including forest fires, droughts, and extreme weather [@zurowski_summer_2023; @duncan_predicting_2012; @bunnell_global_2011; @merkle_large_2016; @slatyer_ecological_2022; @berger_climate_2018].
-Changes in temperature thus affect how animals move [e.g., movement frequency and speed -- their movement behaviour: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].
+As we approach the deadline for the Kunming-Montreal Global Biodiversity Framework (i.e., the "30 by 30" initiative), a global effort to conserve 30% of the world's lands and oceans by 2030 [@convention_on_biological_diversity_154_2022], Canadian decision makers must understand which areas will be of greatest value for maximizing and protecting biodiversity not only for current conditions but future decades, too.
+Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.
+Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022].
 
@@
 
-This paper provides a framework for quantifying the effects of proximal air temperature on animal movement using quantitative methods that are flexible and statistically robust.
+This paper provides a quantifyies the effects of proximal air temperature on mammalian movement behavior, namely species' movement rates and habitat selection.
 We present results for six large-bodied terrestrial mammal species in western Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
 Using over 25 years of telemetry data throughout a large spatial range of western Canada and hourly air temperature data, we estimate how these mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature.
-We then pair the estimated responses with climate change projections to forecast the species' movement during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017].
-We disentangle the direct and indirect effects of temperature on the movement behaviour of terrestrial mammals and discuss the consequences of these changes at the individual, population, and community levels.
+We then pair the estimated responses with climate change projections to forecast the species' movement behavior during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017].
+We discuss the consequences of these changes at the individual, population, and community levels.
+We also provide suggestions for future research, including sampling design to overcome some common data limitations.
 

### 2025-09-02 writing/appendix-b-gams.Rmd
moved model code to appendix B

@@
 ```
+
+\noindent The two models for the probability of moving and speed when moving were fit using the code below:
+
+\footnotesize
+
+\singlespacing
+
+```{r hgam_1, eval=FALSE, echo=TRUE}
+m_moving <- bam(moving ~ # model estimating probability of moving
+                  # random intercept for each animal
+                  s(animal, bs = 're') +
+                  # fixed intercept for each species
+                  species +
+                  # to account for changes in behavior within days
+                  s(tod_pdt, by = species, k = 5, bs = 'cc') +
+                  # to account for changes in behavior within years
+                  s(doy, by = species, k = 5, bs = 'cc') +
+                  # species-level effect of temperature
+                  s(temp_c, by = species, k = 5, bs = 'tp') +
+                  # to account for seasonal changes in day length
+                  ti(doy, tod_pdt, by = species, k = 5, bs = c('cc', 'cc')) +
+                  # to account for changes in day nocturnality with temperature
+                  ti(temp_c, tod_pdt, by = species, k = 5, bs = c('tp', 'cc')) +
+                  # to account for changes in fur coats seasonally
+                  ti(temp_c, doy, by = species, k = 5, bs = c('tp', 'cc')) +
+                  # larger sampling intervals underestimate movement speed
+                  s(log(dt), k = 3) +
+                  s(log(dt), species, k = 3, bs = 'fs'),
+                family = binomial(link = 'logit'),
+                data = d,
+                method = 'fREML', # fast REML
+                discrete = TRUE, # discretize the covariates for faster computation
+                knots = list(tod_pdt = c(0, 1), doy = c(0.5, 366.5)), #' for `bs = 'cc'`
+                control = gam.control(trace = TRUE))
+```
+
+\clearpage
+
+```{r hgam_2, eval=FALSE, echo=TRUE}
+m_speed <- bam(speed_est ~ # model estimating speed when moving (i.e., speed > 0)
+                 # random intercept for each animal
+                 s(animal, bs = 're') +
+                 # fixed intercept for each species
+                 species +
+                 # to account for changes in behavior within days
+                 s(tod_pdt, by = species, k = 5, bs = 'cc') +
+                 # to account for changes in behavior within years
+                 s(doy, by = species, k = 5, bs = 'cc') +
+                 # species-level effect of temperature
+                 s(temp_c, by = species, k = 5, bs = 'tp') +
+                 # to account for seasonal changes in day length
+                 ti(doy, tod_pdt, by = species, k = 5, bs = c('cc', 'cc')) +
+                 # to account for changes in day nocturnality with temperature
+                 ti(temp_c, tod_pdt, by = species, k = 5, bs = c('tp', 'cc')) +
+                 # to account for changes in fur coats seasonally
+                 ti(temp_c, doy, by = species, k = 5, bs = c('tp', 'cc')) +
+                 # larger sampling intervals underestimate movement speed
+                 s(log(dt), k = 3) +
+                 s(log(dt), species, k = 3, bs = 'fs'),
+               family = Gamma(link = 'log'), # can use Gamma because no zeros
+               data = d_2,
+               method = 'fREML', # fast REML
+               discrete = TRUE, # discretize the covariates for faster computation
+               knots = list(tod_pdt = c(0, 1), doy = c(0.5, 366.5)), #' for `bs = 'cc'`
+               control = gam.control(trace = TRUE))
+```
+
+\normalsize
+
+\doublespacing
 
@@
 
+\noindent Each species had a Hierarchical Resource Selection Function fit with the code below.
+See @aarts_estimating_2008 for additional information on fitting Resource Selection Functions using Poisson models.
+
+\singlespacing
+
+\footnotesize
+
+```{r rsf, eval=FALSE, echo=TRUE}
+rsf <- bam(detected ~ # 1 for telemetry locations, 0 for quadrature points
+             # species-level average resource preference
+             s(forest_perc, k = 4, bs = 'cr') +
+             s(elevation_m, k = 4, bs = 'cr') +
+             s(dist_water_m, k = 4, bs = 'cr') +
+             # animal-level deviations from the species-level average
+             s(animal, bs = 're') +
+             s(forest_perc, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')) +
+             s(elevation_m, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')) +
+             s(dist_water_m, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')) +
+             # changes in preference with temperature
+             ti(forest_perc, temperature_C, k = 4, bs = 'cr') +
+             ti(elevation_m, temperature_C, k = 4, bs = 'cr') +
+             ti(dist_water_m, temperature_C, k = 4, bs = 'cr') +
+             # include marginals of temperature to post-stratify over afterwards
+             s(temperature_C, k = 4, bs = 'cr') +
+             s(temperature_C, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')),
+           family = poisson(link = 'log'),
+           data = d,
+           weights = weight, # based on AKDE
+           subset = species == sp, # subset to the species of interest
+           method = 'fREML',
+           discrete = TRUE)
+```
+
+\normalsize
+
+\doublespacing
+
+\clearpage
+
 ```{r kept-hrsf}
@@
   rename(Species = species, `Kept` = keep_for_RSF, Proportion = prop) %>%
-  knitr::kable(format = 'markdown', caption = "Number of animals (n) and proportion of animals (Proportion) used or excluded (Kept) when fitting the hierarchical resource selection function.
+  knitr::kable(format = 'markdown', digits = 3, caption = "Number of animals (n) and proportion of animals (Proportion) used or excluded (Kept) when fitting the hierarchical resource selection function.
 The models only used animals that showed stationary range residency or two clear seasonal ranges within the tracking period.

### 2025-09-02 writing/manuscript.Rmd
reviewed methods

@@
 abstract <- "Widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness, habitat selection, and community structure.
-Previous studies have estimated the effects of temperature on mammalian movement behaviour, but many did not disentangle them from seasonal behavior cycles.
-Consequently, it is still unclear how mammals will adapt their movement behavior in the 21^st^ century.
+Previous studies have estimated the effects of temperature on mammalian movement behaviour, but many did not disentangle them from seasonal behaviour cycles.
+Consequently, it is still unclear how mammals will adapt their movement behaviour in the 21^st^ century.
 We address this gap by using continuous-time movement models and hierarchical generalized additive models to quantify the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species throughout western Canada between 1998 and 2023.
@@
 Consequently, our results suggest that all six species may shift their range in the coming decades as their selection strength for their current ranges weakens.
-As climate change exposes mammals to novel environmental conditions, predicting changes in animal behavior and community structure will become crucial for effectively and proactively understanding mammalian movement behavior and protecting high-quality habitat.
+As climate change exposes mammals to novel environmental conditions, predicting changes in animal behaviour and community structure will become crucial for effectively and proactively understanding mammalian movement behaviour and protecting high-quality habitat.
 We thus present our work using a flexible approach that can be adapted for different species and spatiotemporal scales, and we conclude by providing suggestions for future research, with particular attention to the study design, statistical modeling, and forecasting."
@@
 * marine turtles shift and lose habitat with temperature: https://www.science.org/doi/10.1126/sciadv.adw4495
-* changes in temperature change phelogy and movement behavior as a consequence @gerlich_species_2025
+* changes in temperature change phelogy and movement behaviour as a consequence @gerlich_species_2025
 * @walker_global_2019: Global climate change and invariable photoperiods: {A} mismatch that jeopardizes animal fitness
@@
 Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [@hall_microrefuges_2016].
-By altering mammals' movement speed and habitat use [jointly, their movement behavior: @nathan_movement_2008; @nathan_big-data_2022], ambient temperature also affects the consequences of movement behavior [@alston_temperature_2020; @giroux_activity_2023], such as encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, or competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024].
-Behavioral changes to temperature are exacerbated in extreme cold [@wilson_stenotherms_2001; @berger_climate_2018] and extreme heat [@powers_hovering_2017; @alston_temperature_2020; @giroux_activity_2023; @verzuh_behavioural_2023].
+By altering mammals' movement speed and habitat use [jointly, their movement behaviour: @nathan_movement_2008; @nathan_big-data_2022], ambient temperature also affects the consequences of movement behaviour [@alston_temperature_2020; @giroux_activity_2023], such as encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, or competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024].
+Behavioural changes to temperature are exacerbated in extreme cold [@wilson_stenotherms_2001; @berger_climate_2018] and extreme heat [@powers_hovering_2017; @alston_temperature_2020; @giroux_activity_2023; @verzuh_behavioural_2023].
 
@@
 Mammals are particularly susceptible to adverse effects from excessive heat [@sherwood_adaptability_2010].
-While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more rapid and serious physiological damage [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021; @newediuk_climate_2024].
-Due to mammals' difficulty at dissipating heat, particularly when ambient temperatures are near or above their body temperature, they are particularly sensitive to extreme heat, particularly if they have a large body size [@dyer_travel_2023; @verzuh_behavioural_2023].
+While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more severe physiological damage in a shorter span of time [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021; @newediuk_climate_2024].
+Due to mammals' difficulty at dissipating heat, particularly when ambient temperatures are near or above their body temperature, they are particularly sensitive to extreme heat, especially with increasing body size [@dyer_travel_2023; @verzuh_behavioural_2023].
 Furthermore, species that experience narrow temperature ranges (i.e., $`$temperature specialists') tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while species that experience wide ranges of temperature variation (i.e., $`$temperature genralists') can adapt more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024].
@@
 
-In the last 200 years, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@ipcc_climate_2023].
-Over the next 100 years, these changes are expected to worsen and continue to impact mammals' fitness, movement behaviour, and habitat selection as animals cope with growing heat stress [@deb_modelling_2020; @woo-durand_increasing_2020] and more frequent and intense extreme events [@bunnell_global_2011], along with growing anthropogenic habitat loss, modification, and fragmentation [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024].
+In the last century, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@ipcc_climate_2023].
+As heat stress intensifies over the next 100 years, mammals will continue to suffer impacts on their fitness, movement rates, and habitat selection [@deb_modelling_2020; @woo-durand_increasing_2020], which will be exacerbated by more frequent and intense extreme events [@bunnell_global_2011] and anthropogenic habitat loss, modification, and fragmentation [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024].
 It remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially if populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018].
@@
 Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], movement behaviour [@melin_effect_2023], thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [e.g., body size, activity time, and elevational and latitudinal distribution @mccain_assessing_2019].
-However, there remains a need for fine-scale estimates of the effects of temperature on mammals' movement behavior and quantitative predictions of how their behavior will change in the current century.
+However, there remains a need for fine-scale estimates of the effects of temperature on mammals' movement behaviour and quantitative predictions of how their behaviour will respond to climate change in the current century [@ipcc_climate_2023].
 
@@
 
-This paper provides a quantifyies the effects of proximal air temperature on mammalian movement behavior, namely species' movement rates and habitat selection.
-We present results for six large-bodied terrestrial mammal species in western Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
+In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
 Using over 25 years of telemetry data throughout a large spatial range of western Canada and hourly air temperature data, we estimate how these mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature.
-We then pair the estimated responses with climate change projections to forecast the species' movement behavior during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017].
-We discuss the consequences of these changes at the individual, population, and community levels.
-We also provide suggestions for future research, including sampling design to overcome some common data limitations.
+We then pair the estimated responses with climate change projections to forecast the species' movement behaviour during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017].
+We discuss the consequences of these changes in movement behaviour at the individual, population, and community levels.
+We also provide directions for future research and solutions for overcoming some common data limitations.
 
@@
 
-knitr::kable(dt_data, format = 'markdown', caption = 'Start and end of the GPS telemetries along with the median sampling interval ($\\Delta t$; stratified by individual), number of animals (Animals), and number of animals with resolved speed (Has speed) for each of the species in this study.', escape = FALSE, longtable = FALSE, booktabs = TRUE)
+knitr::kable(dt_data, format = 'markdown', caption = "Summary statistics of each species' set of GPS data, including: the start and end of each set of GPS telemetries, the median sampling interval ($\\Delta t$; stratified by animal), number of animals (Animals), and number of animals with finite speed estimates (Has speed).", escape = FALSE, longtable = FALSE, booktabs = TRUE)
 ```
@@
 This study leverages four main datasets: (1) a multi-species collection of GPS telemetry data (Fig.
-\@ref(fig:bc-map)); (2) historical hourly reanalyzed air temperature from the ERA5 dataset from the European Center for Medium-range Weather Forecasting; (3) resource rasters of percent forest cover, elevation, and distance from water; and (4) climate change projections under four Shared Socioeconomic Pathways [SSPs, see @riahi_shared_2017].
-While we acknowledge that forest type, structure, and age are important variables, we do not include such variables in this study for the sake of parsimony and comparability across species.
-We detail the data sources and analyses in the sections below.
+\@ref(fig:bc-map)); (2) historical hourly air temperature from the ERA5 reanalyzed dataset [European Center for Medium-range Weather Forecasting$;$ @hersbach_era5_2023]; (3) temporally static rasters of percent forest cover, elevation, and distance from water; and (4) climate change projections under four Shared Socioeconomic Pathways [SSPs$;$ @riahi_shared_2017].
+While we acknowledge there are other important variables (e.g., forest type, structure, and age; terrain slope and aspect), we limited our models to only use three simple variables for the sake of parsimony and comparability across species.
+See the discussion section for additional detail on the advantages and limitations of using such simple models.
 
@@
 
-\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public British Columbia Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)).
-***MISSING INFO ON OTHER TELEMETRY DATA.*** From the full set of telemetry data, a total of 2396 GPS locations (0.16%, including collar calibration data) were removed using diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points.
-Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated stationary behaviour (Appendix A).
+\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022].
+Boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired from a data repository of the British Columbia Oil and Gas Research and Innovation Society (BC ORGIS) that was collected as part of the boreal caribou radio-collaring program of the BC OGRIS Research and Effectiveness Monitoring Board [REMB$;$ @bc_ogris_boreal_2018].
+***MISSING INFO ON OTHER TELEMETRY DATA.*** From the full set of telemetry data, we removed a total of 2396 problematic GPS locations (0.16%, including collar calibration data) after inspecting diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points.
+Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated relatively stationary behaviour (Appendix A).
 
-### Historical temperature data and temperature projections
+### Historical temperature data and climate change projections
 
 \noindent Rasters of hourly reanalyzed air temperature data were downloaded from the ERA5 database [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)).
-Proximal temperature was estimated for each location by extracting the value from the corresponding raster cell from the temporally nearest raster using the `extract()` function from the `terra` package [version 1.7-46, @hijmans_terra_2023] for `R` [@r_core_team_r_2024].
+Proximal air temperature was estimated for each location by extracting the value from the corresponding raster cell from the temporally nearest raster using the `extract()` function from the `terra` package [v.
+1.7-46, @hijmans_terra_2023] for `R` [@r_core_team_r_2024].
 
-We obtained rasters of projected monthly average temperature for the study region at a 0.08$^\circ$ spatial resolution from 2020 to 2100 under the different SSPs via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`.
-Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 reanalized data for western Canada from 1998 to 2023 [extremes included, @hersbach_era5_2023] to calculate within-month variance in temperature, which we defined as the variance within a given pixel, month, and year.
-We then modeled the estimated variance using a GAM for location and scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017].
-The linear predictor for the location (i.e., the mean) included smooth terms of the within-pixel monthly mean temperature (within each year), month (as a cyclic smooth), a two-dimensional smooth of space, and a tensor interaction product term of space and month.
-The linear predictor for the scale term (which governs the mean-variance relationship) included smooth terms of the monthly mean, month, and space.
-We did not include a smooth of year to avoid unrealistic projections when extrapolating beyond the range of historical data.
+We used the the `climatenaR` package [v.
+1.0, @burnett_climatenar_2023] for `R` and ClimateNA [v.
+7.4.2$;$ @wang_locally_2016; @mahony_global_2022] to obtain projected average monthly temperatures from 2020 to 2100 at a 0.08$^\circ$ spatial resolution.
+Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the historical data for western Canada from 1998 to 2023 (extremes included) to estimate the monthly variation in temperature, which we defined as $\sigma^2_T$, the variance at location $\langle x, y \rangle$ in month $m$ for each year (e.g., the variance at coordinates $\langle -119.40, 49.94 \rangle$ in April 2005).
+We then modeled the estimated variance using a Generalized Additive Model for Location and Scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017] using the `mgcv` package for `R` [v.
+1.9-1, @wood_generalized_2017].
+The linear predictor for the location (i.e., the mean) included smooth terms of: each year's estimated within-pixel monthly mean temperature ($\hat \mu_T$), month ($m$), a two-dimensional smooth of spatial coordinates $\langle x, y \rangle$, and a tensor product interaction term of month and space to allow for seasonal terms to vary smoothly over space.
+The linear predictor for the scale term, which governs the mean-variance relationship [table 3.1 on p.
+104 of @wood_generalized_2017], included smooth terms of the monthly mean temperature, month, and space.
+We did not include smooth terms of year to avoid unrealistic projections when extrapolating beyond past 2023.
+The model was thus
 
-```{r, eval=FALSE, echo=FALSE}
-m_gammals <- gam(list(
-  # linear predictor for the mean parameter
-  monthly_var ~
-    s(monthly_mean, bs = 'cr', k = 10) +
-    s(month, bs = 'cc', k = 10) + # using a cyclic cubic spline basis
-    s(long, lat, bs = 'ds', k = 500) +
-    ti(long, lat, month, bs = c('ds', 'cc'), d = c(2, 1), k = c(100, 10)),
-  # linear predictor for the scale parameter
-  ~ s(monthly_mean, bs = 'cr', k = 10) +
-    s(month, bs = 'cc', k = 10) +
-    s(long, lat, bs = 'ds', k = 100)),
-  family = gammals(),
-  data = d,
-  method = 'REML',
-  knots = list(month = c(0.5, 12.5)))
-```
+\begin{equation}\label{eq:temp-var-model}
+\begin{cases}
+\sigma^2_T \sim \text{Gamma}(\mu_{\sigma^2_T}, \nu_{\sigma^2_T}) \\
+\mathbb E(\sigma^2_T) = \mu_{\sigma^2_T} \\
+\mathbb V(\sigma^2_T) = (\mu_{\sigma^2_T})^2 / \nu_{\sigma^2_T} \\
+\log(\mu_{\sigma^2}) = \beta_{L,0} + f_{L,1}(\mu_T) + f_{L,2}(m) + f_{L,3}(x, y) + f_{L,4}(x, y, m) \\
+\log(\nu_{\sigma^2}) = \beta_{S,0} + f_{S,1}(\mu_T) + f_{S,2}(m) + f_{S,3}(x, y)
+\end{cases},
+\end{equation}
 
-We simulated hourly variation in future weather by assuming hourly temperature followed a normal distribution with mean specified by the monthly `climatenaR` climate projections and variance as specified by the Gamma GAMLS.
-For each month within each year from 2020 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantile by increments of 0.1, and we weighted each quantile proportionally to the (normalized) Gaussian probability density for each quantile.
+\noindent where $\mu_{\sigma^2_T}$ and $\nu_{\sigma^2_T}$ indicate the location and scale parameters of the gamma distribution of $\sigma^2_T$, and together they determine the variance of $\sigma^2_T$, indicated as $\mathbb V(\sigma^2_T)$.
+Functions $f_{L,j}$ and $f_{S,j}$ indicate the $j^\text{th}$ smooth functions for the location and scale parameters, respectively.
+To ensure the smooth functions of month, $f_{L,2}(m)$ and $f_{S,2}(m)$, joined smoothly between December and January, the terms use cyclic cubic spline bases.
+The spatial terms used two-dimensional Duchon splines, a generalization of thin-plate regression splines (p.
+221 of @wood_generalized_2017).
+The smoothness parameters were estimated via REstricted Maximum Likelihood [REML$;$ p.
+83 of @wood_generalized_2017].
+See the Data Availability Statement for additional information on the code used to fit the model.
+
+We simulated hourly variation in future years by assuming hourly temperature followed a normal distribution with mean specified by the monthly `climatenaR` climate projections and variance as specified by the gamma GAMLS.
+For each month within each year from 2025 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantiles by increments of 0.1, and we weighted each quantile proportionally to the (normalized) Gaussian probability density for each quantile.
 
@@
 We converted the raster of percent cover of open water (class 12) to a binary raster of pixels with water (cover > 0%) or without water (cover = 0%) and then calculated each pixel's distance from the nearest pixel with water using the `distance()` function from the `terra` package.
-Finally, we obtained two digital elevation models using the `get_elev_raster()` function from the `elevatr` package [version 0.99.0, @hollister_elevatr_2023]: We used a raster with a zoom of 6 (0.009$^\circ$) for model fitting and one with a zoom of 3 (0.08$^\circ$) for downloading climate change projections via `climatenaR`.
-All final rasters and code are available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).
+Finally, we obtained two digital elevation models using the `get_elev_raster()` function from the `elevatr` package [v.
+0.99.0, @hollister_elevatr_2023].
+We used a raster with a zoom of 6 (a resolution of 0.009$^\circ$) for model fitting and one with a zoom of 3 (a resolution of 0.08$^\circ$) for downloading climate change projections via `climatenaR`.
+All final rasters and code are available in the GitHub repository associated with this manuscript (see the Data Availability Statement).
 
@@
 
-\noindent We modeled each animal's movement using continuous-time movement models [@fleming_fine-scale_2014] via the `ctmm` package [version 1.2.0, @fleming_ctmm_2023] for `R`.
-We then estimated mammals' instantaneous speeds by applying the `ctmm::speeds()` function on all models with finite speed estimates [415 of 433, @fleming_fine-scale_2014; @noonan_scale-insensitive_2019].
-The 18 animals with insufficiently fine sampling to support movement models with finite speed [@noonan_scale-insensitive_2019] were for one mountain goat, 15 boreal caribou, and 2 southern mountain caribou (Table \@ref(tab:data-table)).
+\noindent We modeled each animal's movement using continuous-time movement models [@fleming_fine-scale_2014] via the `ctmm` package [v.
+1.2.0, @fleming_ctmm_2023] for `R`.
+We then estimated each mammal's instantaneous speed at each observed location by applying the `ctmm::speeds()` function on all models with finite speed estimates [415 of 433, @fleming_fine-scale_2014; @noonan_scale-insensitive_2019].
+The 18 animals with insufficiently fine sampling to support movement models with finite speed were for one mountain goat, 15 boreal caribou, and two southern mountain caribou (Table \@ref(tab:data-table)).
 
-Since `ctmm``s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed could be 0.
+Since $\texttt{ctmm}\text{'s}$ movement models assume a single moving state with stochastic but non-zero speed, we corrected data-informed speeds so that the minimum instantaneous speed could be 0.
 We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed.
-The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R).
-To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using a $k$-means algorithm with 2 clusters for each species' detrended speeds.
+The function we used is available on GitHub (see the Data Availability Statement).
+To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using a $k$-means algorithm with 2 clusters for each species' distribution of detrended speeds.
 When the algorithm clearly failed to discriminate between states, we estimated the split point using the inflection points in histograms of the detrended speeds (Fig.
@@
 
-## Estimating the effects of temperature on mammals' movement
+## Estimating the effects of temperature on mammals' movement behaviour
 
-\noindent The following sections detail the statistical framework and modelling we used to estimate the effect on temperature on mammals' movement.
-To assess the importance of including temperature as an explicit covariate (rather than including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the fits of the two sets of models.
+\noindent Ambient temperature is only one of the many drivers of mammalian movement behaviour (Fig.
+\@ref(fig:dags)).
+Many species adapt their movement rates (e.g., movement frequency and speed) daily or seasonally in response in factors such as solar time, photoperiod, forage availability, reproductive cycles, and predator avoidance.
+Similarly, ambient temperature also fluctuates throughout the day and across seasons.
+Therefore, estimating the effects of temperature on movement rates requires accounting for how mammals' response to temperature changes with time of day and day of year [Fig.
+\@ref(fig:dags)A$;$ @peron_periodic_2017; @buderman_time-varying_2018; @leclerc_determinants_2021; @xu_plasticity_2021].
+Similarly, mammals' selection strength for resources depends on ambient temperature, since higher temperatures can promote a selection for refuge from heat [e.g., thicker forest cover, higher elevation, proximity to water$;$ @attias_effects_2018; @giroux_activity_2023].
 
-### Disentangling direct and indirect effects
+The two following sections detail the statistical framework and models we used to estimate the effect on temperature on the six species' movement behavior.
+To assess the importance of including temperature as an explicit covariate (as ooposed to including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the fits of the two sets of models via analyses of deviance (a form of generalized likelihood ratio tests) following the methods of Section 3.3 in @wood_generalized_2017.
+See Appendix B for additional information.
 
-\noindent In this study, we separate the effects of temperature on mammalian movement into direct and indirect effects.
-We call "direct" all effects that impact movement behavior causally without first impacting another variable (Fig.
-\@ref(fig:dags)).
-For example, in Fig.
-\@ref(fig:dags)A, whether a grizzly bear is moving in a given moment directly depends on time of day and day of year, since it will not be moving when sleeping or hibernating.
-However, time of day and day of year also impact its movement behavior indirectly because its sensitivity to temperature will also depend on time of day and day of year.
-Thus, both time of day and day of year also impact movement frequency indirectly by altering the effect of temperature on movement frequency.
-Failing to disentangle the effects of these three variables will lead to incorrect and likely confusing estimates for the causal effect of temperature on movement behavior, since causal and non-causal relationships would be indistinguishable [@mcelreath_statistical_2020].
-
-Indirect effects of temperature also include all effects that can be controlled by altering an animal's habitat [its geographic space *sensu* @matthiopoulos_defining_2022].
-For example, in Fig.
-\@ref(fig:dags)B, an increase in temperature may push an animal to prefer higher forest cover in search of shade, and conservationists can mitigate these indirect effects of temperature by increasing the availability of forested areas, since the effect of temperature on movement is conditional on forest cover availability.
-In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022].
-In this case, it is not possible to mitigate the effects of temperature on movement behaviour (other than changing temperature directly).
-
-```{r dags,  message=FALSE, warning=FALSE, fig.cap="Directed Acyclical Graphs (DAGs) assumed for inferring the causal effects of temperature (Temp) on movement behavior.
-(A) DAG for the probability of moving, speed when moving, or distance traveled ($Y$) while accounting for the effects of time of day (TOD), day of year (DOY), and their interaction effects.
-Temperature directly affects $Y$, but the effects of temperature depend on the time of day and season.
-Time of day and day of year also affect $Y$ directly, but the effect of time of day changes throughout the year due to changes in day length and seasonality.
-(B) DAG for Relative Selection Strength (RSS) for percent forest cover (F), elevation (E), and distance from water (W).
-The RSS for of a given habitat depends on all three resources, and the selection for each resource is independent of the other two resources and dependent on temperature.", fig.height=4, fig.width=10, out.width='\\linewidth'}
+```{r dags,  message=FALSE, warning=FALSE, fig.cap="Directed Acyclical Graphs (DAGs) assumed for inferring the causal effects of temperature on each species' movement behaviour.
+(A) Ambient temperature affects mammals' movement rates (i.e.
+probability of moving, speed when moving, and distance traveled), but temperature is influenced by time of day and day of year.
+Additionally, the effects of temperature on movement rates depend on time of day and season, since animals may respond to temperatures differently at different times of day and or days of year.
+Finally, the effect of time of day changes seasonally due to changes in photoperiod.
+(B) Similarly, habitat selection depends on the availability and selection of resources (forest cover, elevation, and distance from water), but the selection strengh for each resource is conditional on temperature.
+For example, an animal may avoid forested areas unless it is seeking refuge from extreme temperatures.", fig.height=8, fig.width=20, out.width='\\linewidth'}
 library('dplyr')   # for data wrangling
 library('ggplot2') # for fancy plots
-library('ggdag')   # for directed acyclical graphs
 library('cowplot') # for plots in grids
 
-plot_grid(
-  dagify(Y ~ TOD + DOY + Temp,
-         Temp ~ TOD + DOY,
-         TOD ~ DOY,
-         outcome = 'Y',
-         coords = list(x = c(Y = 0, TOD = -1, DOY = 1, Temp = 0),
-                       y = c(Y = 1.2, TOD = 2, DOY = 2, Temp = 0))) %>%
-    tidy_dagitty() %>%
-    mutate(col = case_when(name == 'Y' ~ '1',
-                           name == 'TOD' ~ '0',
-                           name == 'DOY' ~ '0',
-                           name == 'Temp' ~ '2')) %>%
-    ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
-    geom_dag_point(aes(color = col), size = 20, alpha = 0.5) +
-    geom_dag_point(color = 'black', size = 20, shape = 1) +
-    geom_dag_edges() +
-    geom_dag_text(aes(label = name), color = 'black', size = 4.5) +
-    scale_color_manual(values = c('grey', 'darkorange', 'red')) +
-    theme_dag() +
-    theme(legend.position = 'none'),
-  dagify(RSS ~ `F` + E + W,
-         `F` ~ Temp,
-         E ~ Temp,
-         W ~ Temp,
-         outcome = 'RSS',
-         coords = list(x = c(RSS = 0, `F` = -1, E = 0, W = 1, Temp = 0),
-                       y = c(RSS = 2, `F` = 1, E = 1, W = 1, Temp = 0)))%>%
-    tidy_dagitty() %>%
-    mutate(col = case_when(name == 'RSS' ~ '1',
-                           name == 'F' ~ '0',
-                           name == 'E' ~ '0',
-                           name == 'W' ~ '0',
-                           name == 'Temp' ~ '2')) %>%
-    ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
-    geom_dag_point(aes(color = col), size = 20, alpha = 0.5) +
-    geom_dag_point(color = 'black', size = 20, shape = 1) +
-    geom_dag_edges() +
-    geom_dag_text(aes(label = name), color = 'black', size = 4.5) +
-    scale_color_manual(values = c('grey', 'darkorange', 'red')) +
-    theme_dag() +
-    theme(legend.position = 'none'), labels = 'AUTO', nrow = 1)
+# movement rates
+boxes_a <- tibble(xmin = c(0, 10, 5, 5),
+                  xmax = xmin + 4,
+                  ymin = c(10, 10, 5, 0),
+                  ymax = ymin + 1,
+                  fill = c('#BBBBBB80', '#BBBBBB80', '#FF8C00AA', '#FF0000CC'),
+                  label = c('Time of day', 'Day of year', 'Movement rates',
+                            'Temperature'))
+
+arrows_a <- tibble(id = sort(rep(1:6, 2)),
+                   x = c(10, 4, 4, 6, 10, 8, 2, 6, 7, 7, 12, 8),
+                   y = c(10.5, 10.5, 10, 6, 10, 6, 10, 1, 1, 5, 10, 1))
+
+p_a <-
+  ggplot() +
+  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
+            boxes_a, fill = boxes_a$fill, color = 'black') +
+  geom_path(aes(x, y, group = id), arrows_a, linewidth = 1,
+            arrow = arrow(type = 'closed')) +
+  geom_label(aes(x = (xmin + xmax) / 2, y = (ymin + ymax) / 2,
+                 label = label), boxes_a, fill = 'transparent',
+             fontface = 'bold', label.size = 0, size = 5) +
+  theme_void() +
+  theme(legend.position = 'none')
+
+# habitat selection
+boxes_b <- tibble(xmin = c(5, 0, 5, 10, 5),
+                  xmax = xmin + 4,
+                  ymin = c(10, 5, 5, 5, 0),
+                  ymax = ymin + 1,
+                  fill = c('#FF8C00AA', rep('#BBBBBB80', 3), '#FF0000CC'),
+                  label = c('Habitat selection', 'Forest cover',
+                            'Elevation', 'Distance from water',
+                            'Temperature'))
+
+arrows_b <- tibble(id = sort(rep(1:6, 2)),
+                   x = c(2, 6, 7, 7, 12, 8, 6, 2, 7, 7, 8, 12),
+                   y = c(6, 10, 6, 10, 6, 10, 1, 5, 1, 5, 1, 5))
+
+p_b <-
+  ggplot() +
+  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
+            boxes_b, fill = boxes_b$fill, color = 'black') +
+  geom_path(aes(x, y, group = id), arrows_b, linewidth = 1,
+            arrow = arrow(type = 'closed')) +
+  geom_label(aes(x = (xmin + xmax) / 2, y = (ymin + ymax) / 2,
+                 label = label), boxes_b, fill = 'transparent',
+             fontface = 'bold', label.size = 0, size = 5) +
+  theme_void() +
+  theme(legend.position = 'none')
+
+plot_grid(p_a, p_b, labels = 'AUTO', nrow = 1)
 ```
@@
 
-\noindent We estimated the effects of temperature ($^\circ$C) on mammals' instantaneous movement state (moving or not) and speed when moving using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019 and the code chunk below] with the `mgcv` package for `R` [version 1.9-1, @wood_generalized_2017].
+\noindent We estimated the effects of temperature on mammals' instantaneous movement state (moving or not) and speed when moving using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019 and Appendix B] with the `mgcv` package for `R`.
 The first HGAM estimated the probability that an animal was moving, $P(M)$, with a binomial family of distributions and logit link function.
-The response variable `moving` was coded as 1 if moving and 0 if not.
-The second HGAM estimated an animal's speed when moving (`speed`) with a gamma family of distributions and log link function.
-
-The HGAMs included random intercepts for each animal (`s(animal, bs = 're')`), fixed-effect intercepts for each species (`species`), and species-level `by` smooths (`s(..., by = species)`), which allowed each species' term to be estimated independent of other species [see model I in Figure 4 of @pedersen_hierarchical_2019].
-The `by` smooths accounted for trends in time of day (in Pacific Daylight Time; `tod_pdt`), day of year (`doy`), and temperature (`temp_c`).
-The models also had three tensor product interaction terms (`ti()`) `by` each species: (1) day of year and time of day, (2) temperature and time of day, and (3) temperature and day of year.
-These three terms accounted for: (1) seasonal changes in day length, (2) hourly changes in the response to temperature (e.g., changes in nocturnality), and (3) seasonal changes in the response to temperature (e.g., changes in coats and migration timing).
-Finally, smooth terms of log-transformed hours between GPS locations (`dt`) accounted for irregular sampling rates.
-A global term of `log(dt)` accounted for the overall effect of sampling interval, while a factor-smooth interaction term (`s(log(dt), species, bs = 'fs')`) allowed for species-level deviations from the global term while assuming a common smoothness parameter across species [see model GS in Figure 4 of @pedersen_hierarchical_2019].
-
-The HGAMs accounted for the cyclicity of time of day and day of year using cyclic cubic splines [`bs = 'cc'`, see p.
-202 of @wood_generalized_2017].
+The second HGAM estimated an animal's speed when moving with a gamma family of distributions and log link function.
 We fit the models with fast Restricted Maximum Likelihood (`'fREML'`) and discretized covariates (`discrete = TRUE`) to optimize computational efficiency with no appreciable losses to model performance [Appendix B$;$ @wood_generalized_2015; @wood_generalized_2017-1; @li_faster_2020].
@@
 
-\clearpage
+The HGAMs (equations \@ref(eq:p-moving-model) and \@ref(eq:speed-model)) included fixed-effect intercepts for each species ($\beta_\texttt s$), random intercepts for each animal ($Z_\texttt{a}$), and species-level `by` smooths that allowed independent smoothness parameters for each species [model I in Figure 4 of @pedersen_hierarchical_2019].
+The `by` smooths accounted for trends in time of day (in Pacific Daylight Time; `tod_pdt`), day of year (`doy`), and temperature (`temp_c`).
+To account for the cyclicity of time of day and day of year, the smooth terms used cyclic cubic splines [p.
+202 of @wood_generalized_2017].
+The models also had three tensor product interaction terms `by` each species: (1) day of year and time of day, (2) temperature and time of day, and (3) temperature and day of year.
+These three terms accounted for smooth changes in: (1) daily behavior across day of year, (2) the response to temperature over time of day (e.g., changes in nocturnality), and (3) the response to temperature over day of year (e.g., changes in coats and migration timing).
+Finally, two smooth terms of log-transformed sampling interval (`dt`; hours) corrected for biases in speed estimates arising from irregular GPS sampling intervals, since longer intervals result in lower speed estimates [@nathan_big-data_2022; @denicola_are_2025].
+A global smooth term of $\log(\texttt{dt})$ accounted for the overall effect of sampling interval, while a factor-smooth interaction term (`bs = 'fs'`) of $\log(\texttt{dt})$ and species accounted for species-level deviations from the global term while assuming a common smoothness parameter across species [model GS in Figure 4 of @pedersen_hierarchical_2019].
+Formally, the model for movement state $M$, with $M = 0$ indicating no movement and $M = 1$ indicating movement, was
 
-\singlespacing
+\begin{equation}\label{eq:p-moving-model}
+\begin{cases}
+M \sim \text{Bin}(p) \\
+\mathbb E(M) = p \\
+\mathbb V(M) = p (1-p) \\
+\begin{aligned}
+  \text{logit}(p) = & \beta_\texttt{s} + Z_\texttt{a} + f_{1,\texttt{s}}(\texttt{tod\_pdt}) + f_{2,\texttt s}(\texttt{doy}) + f_{3,\texttt s}(\texttt{temp\_c}) + \\
+  & f_{4,\texttt s}(\texttt{doy}, \texttt{tod\_pdt}) + f_{5, \texttt s}(\texttt{temp\_c}, \texttt{tod\_pdt}) + f_{6, \texttt s}(\texttt{temp\_c}, \texttt{doy}) + \\
+  & f_7(\log(\texttt{dt})) + f_{8,\texttt s}(\log(\texttt{dt}))
+\end{aligned}
+\end{cases},
+\end{equation}
 
-\scriptsize
+\noindent while the model for movement speed when moving (i.e., $M = 1$, indicated with $S$) was
 
-```{r hgams, eval=FALSE, echo=TRUE}
-m_moving <- bam(
-  moving ~
-    s(animal, bs = 're') +
-    species + #' necessary since `by` terms do not include intercept terms
-    s(tod_pdt, by = species, k = 5, bs = 'cc') +
-    s(doy, by = species, k = 5, bs = 'cc') +
-    s(temp_c, by = species, k = 5, bs = 'tp') +
-    ti(doy, tod_pdt, by = species, k = 5, bs = c('cc', 'cc')) +
-    ti(temp_c, tod_pdt, by = species, k = 5, bs = c('tp', 'cc')) +
-    ti(temp_c, doy, by = species, k = 5, bs = c('tp', 'cc')) +
-    s(log(dt), k = 3) +
-    s(log(dt), species, k = 3, bs = 'fs'),
-  data = d,
-  method = 'fREML', # fast REML
-  discrete = TRUE, # discretize the covariates for faster computation
-  knots = list(tod_pdt = c(0, 1), doy = c(0.5, 366.5))) #' for `bs = 'cc'`
+\begin{equation}\label{eq:speed-model}
+\begin{cases}
+S \sim \text{Gamma}(\mu_S, \nu_S) \\
+\mathbb E(S) = \mu_S \\
+\mathbb V(S) = \mu_S^2 / \nu_S \\
+\begin{aligned}
+  \text{log}(\mu_S) = & \beta_\texttt{s} + Z_\texttt{a} + f_{1,\texttt{s}}(\texttt{tod\_pdt}) + f_{2,\texttt s}(\texttt{doy}) + f_{3,\texttt s}(\texttt{temp\_c}) + \\
+  & f_{4,\texttt s}(\texttt{doy}, \texttt{tod\_pdt}) + f_{5, \texttt s}(\texttt{temp\_c}, \texttt{tod\_pdt}) + f_{6, \texttt s}(\texttt{temp\_c}, \texttt{doy}) + \\
+& f_7(\log(\texttt{dt})) + f_{8,\texttt s}(\log(\texttt{dt}))
+\end{aligned}
+\end{cases}.
+\end{equation}
 
-m_speed <- bam(
-  speed_est ~
-    s(animal, bs = 're') +
-    species +
-    s(tod_pdt, by = species, k = 5, bs = 'cc') +
-    s(doy, by = species, k = 5, bs = 'cc') +
-    s(temp_c, by = species, k = 5, bs = 'tp') +
-    ti(doy, tod_pdt, by = species, k = 5, bs = c('cc', 'cc')) +
-    ti(temp_c, tod_pdt, by = species, k = 5, bs = c('tp', 'cc')) +
-    ti(temp_c, doy, by = species, k = 5, bs = c('tp', 'cc')) +
-    s(log(dt), k = 3) +
-    s(log(dt), species, k = 3, bs = 'fs'),
-  family = Gamma(link = 'log'),
-  data = d_2, #' data filtered for only moving
-  method = 'fREML',
-  discrete = TRUE,
-  knots = list(tod_pdt = c(0, 1), doy = c(0.5, 366.5)))
-```
-
-\normalsize
-
-\doublespacing
-
-<!-- \clearpage -->
+\noindent In both models, $\beta_\texttt{s}$ indicates a fixed intercept for species \texttt{s}, $Z_\texttt a$ indicates a Gaussian random effect for animal \texttt{a} (of species \texttt{s}), $f_{j,\texttt{s}}$ indicates the $j^{\text{th}}$ smooth function for species \texttt{s}, and functions with two variables indicate tensor product interaction terms.
+The model code used to fit the models is available in Appendix B.
 
@@
 
-\noindent We estimated the effects of temperature on each species' selection for percent forest cover (`forest_perc`), elevation (`elevation_m`), and distance from water (`dist_water_m`) by fitting a Hierarchical Resource Selection Function for each species using an HGAM with a Poisson family of distributions and log link function [Appendix B$;$ @aarts_estimating_2008].
-After removing non-resident individuals (Table B1), we accounted for the spatiotemporal autocorrelation in the telemetry locations by weighting each point based on the telemetry's Autocorrelated Kernel Density Estimate [@fleming_new_2017; @noonan_comprehensive_2019] to produce estimates of second- and third-order habitat selection [@johnson_comparison_1980].
-While we recognize there are other important drivers of habitat selection (e.g., forest age, forest type, terrain ruggedness, prey availability), we decided to only use these three proxies to produce results that are easily comparable across species.
+\noindent We estimated the effects of temperature on each species' selection for percent forest cover (`forest_perc`), elevation (`elevation_m`), and distance from water (`dist_water_m`) by fitting a Hierarchical Resource Selection Function (HRSF) for each species.
+We fit each HRSF using an HGAM with a Poisson family of distributions and log link function [Appendix B$;$ @aarts_estimating_2008].
+After removing non-resident individuals (Table B1), we accounted for the spatiotemporal autocorrelation in the telemetry locations by weighting each point based on the telemetry's Autocorrelated Kernel Density Estimate [@fleming_new_2017; @noonan_comprehensive_2019; @alston_mitigating_2022] to produce estimates of second-order habitat selection [@johnson_comparison_1980].
+Quadrature points were determined using the raster cells in each animal's 99.9% AKDE percentile and given a weight of 1.
+The number of quadrature locations greatly outnumbered the number of observed locations (Fig.
+B12), especially after accounting for the AKDE weights (Fig.
+B13).
+
 Each species' model had the same structure:
 
-\singlespacing
+\begin{equation}\label{eq:hrsf}
+\begin{cases}
+D \sim \text{Pois}(\lambda) \\
+\mathbb E(D) = \mathbb V(D) = \lambda \\
+\begin{aligned}
+  \text{log}(\lambda) = & f_{1}(\texttt{forest\_perc}) + f_{2}(\texttt{elevation\_m}) + f_{3}(\texttt{dist\_water\_m}) + \\
+& Z_\texttt a + f_{4, \texttt{a}}(\texttt{forest\_perc}) + f_{5, \texttt{a}}(\texttt{elevation\_m}) + f_{6, \texttt{a}}(\texttt{dist\_water\_m}) + \\
+& f_{7}(\texttt{forest\_perc}, \texttt{temp\_c}) + f_{8}(\texttt{elevation\_m}, \texttt{temp\_c}) + \\
+& f_{9}(\texttt{dist\_water\_m}, \texttt{temp\_c}) + f_{10}(\texttt{temp\_c}) + f_{11,\texttt a}(\texttt{temp\_c}))
+\end{aligned}
+\end{cases},
+\end{equation}
 
-\scriptsize
-
-```{r rsf, eval=FALSE, echo=TRUE}
-rsf <- bam(
-  detected ~ # 1 for telemetry locations, 0 for quadrature points
-    # species-level average resource preference
-    s(forest_perc, k = 6, bs = 'tp') +
-    s(elevation_m, k = elev_k, bs = 'tp') +
-    s(dist_water_m, k = 6, bs = 'tp') +
-    # animal-level deviations from the species-level average
-    s(animal, bs = 're') +
-    s(forest_perc, animal, k = 6, bs = 'fs', xt = list(bc = 'cr')) +
-    s(elevation_m, animal, k = 6, bs = 'fs', xt = list(bc = 'cr')) +
-    s(dist_water_m, animal, k = 6, bs = 'fs', xt = list(bc = 'cr')) +
-    # changes in preference with temperature
-    ti(forest_perc, temp_c, k = 6, bs = 'tp') +
-    ti(elevation_m, temp_c, k = 6, bs = 'tp') +
-    ti(dist_water_m, temp_c, k = 6, bs = 'tp') +
-    # include marginals of temperature to remove sampling biases
-    s(temp_c, k = 4, bs = 'tp') +
-    s(temp_c, animal, k = 4, bs = 'fs', xt = list(bc = 'cr')),
-  family = poisson(link = 'log'),
-  data = d, # species-specific dataset
-  weights = weight, # based on AKDE
-  method = 'fREML',
-  discrete = TRUE)
-```
-
-\normalsize
-
-\doublespacing
-
-Smooth effects of percent forest cover, elevation, and distance to water accounted for the species-level average selection strength for each resource.
-A random effect for each individual animal (`s(animal, bs = 're')`) corrected for uneven sampling across individuals, while factor smooth interaction terms (`bs = 'fs'`) accounted for individual-level resource selection [i.e., individual-level deviations from the species-level average$;$ @jeltsch_need_2025].
-Tensor interaction product terms (`ti()`) of the three resources and temperature estimated the change in resource selection at different temperatures.
-Finally, we included marginal smooth terms of temperature to account for species- and individual-level sampling biases over temperature (e.g., sampling more during warm periods).
-Detections were weighted proportionally to their degree of independence from other temporally proximate detections [`weights = weight` -- Appendix B$;$ @alston_mitigating_2022], while quadrature points had a weight of 1.
-Quadrature points were obtained using all raster cells in the 99.9% AKDE percentile.
-The number of quadrature locations greatly outnumbered the number of observed locations (Fig.
-B12), especially after accounting for weighting based on the degree of autocorrelation and the number of home range crossings (Fig.
-B13).
+\noindent where the species-level indices are omitted for readability, but each term in the model can be assumed to be species-specific.
+Smooth effects of percent forest cover (\texttt{forest\_perc}), elevation (\texttt{elev\_m}), and distance to water (\texttt{dist\_water\_m}) accounted for the species-level selection strength for each resource.
+A Gaussian random effect for each individual animal ($Z_\texttt a$) corrected for uneven sampling across individuals, while factor-smooth interaction terms for each animal ($f_{j,\texttt a}$) accounted for animal-level resource selection [i.e., individual-level deviations from the species-level estimate$;$ @jeltsch_need_2025].
+Tensor product interaction terms of the three resources and temperature (\texttt{temp\_c}) estimated the smooth change in resource selection at different temperatures.
+Finally, marginal smooth terms of temperature and factor-smooth interaction terms of temperature and animal accounted for species- and individual-level sampling biases at different temperatures (e.g., sampling more during warm periods).
 
@@
 B3, and B14), but including temperature always resulted in better fits.
-All analyses of deviance showed that temperature significantly affected movement behavior (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; Table B2 and following pages in Appendix B).
+All analyses of deviance showed that temperature significantly affected movement behaviour (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; Table B2 and following pages in Appendix B).
 Accounting for temperature resulted in a simpler HRSF for mountain goats (lower model degrees of freedom) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.
@@
 The models had good in-sample prediction (Fig.
-B7) and explained reasonably high proportions of the deviance (79.3% for the Gamma model and 10.7% for the binomial model, which is relatively high for a binomial model with binary responses: see ch.
+B7) and explained reasonably high proportions of the deviance (79.3% for the gamma model and 10.7% for the binomial model, which is relatively high for a binomial model with binary responses: see ch.
 11 of @mcelreath_statistical_2020].
@@
 
-Changes in animal movement behavior may also alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024].
+Changes in animal movement behaviour may also alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024].
 At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023], increase their chances of finding food or being preyed on [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021].
@@
 
+* geographic space *sensu* @matthiopoulos_defining_2022
+* environmental space *sensu* @matthiopoulos_defining_2022
+* G space is the realization of the theoretical E space
+
 ## Predicted changes in movement during the 21^st^ century
@@
 
-The diversity in predicted changes in movement behavior by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in Fig.
+The diversity in predicted changes in movement behaviour by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in Fig.
 \@ref(fig:s-hrsfs) highlight the need for careful planning that incorporates not only reliable estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates.
@@
 
-HRSFs do not account for migratory behavior due to reproduction or predation risk
+HRSFs do not account for migratory behaviour due to reproduction or predation risk
 
@@
 
-Our work highlights three central considerations for future research: (1) telemetry sampling rates should be designed primarily in relation to the movement timescales of the species of interest [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @pease_ecological_2024], (2) the number of individuals tracked and the length of each telemetry should depend not just on population size but also the rate of environmental change, and (3) predicting changes in movement behaviour becomes highly complex when responses are nonlinear, especially when changes depend on many interacting factors [@polazzo_measuring_2024] and one is interested in predicting responses in extreme conditions for which data are scarce [@steixner-kumar_strategies_2020].
+Our work highlights three central considerations for future research: (1) telemetry sampling rates should be designed primarily in relation to the movement timescales of the species of interest [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @pease_ecological_2024], (2) the number of individuals tracked and the duration of each telemetry should depend not just on population size but also the rate of environmental change, and (3) predicting changes in movement behaviour becomes highly complex when responses are nonlinear, especially when changes depend on many interacting factors [@polazzo_measuring_2024] and one is interested in predicting responses in extreme conditions for which data are scarce [@steixner-kumar_strategies_2020].
 
@@
 Using the boreal caribou as an example, the 13-hour sampling interval allows us to reasonably estimate the caribou's movement path at a temporal scale of approximately 13 hours (or greater), but we cannot produce reasonable movement trajectories at a much finer (e.g., hourly) scale.
-@nathan_big-data_2022 provides additional examples of how the frequency of location data affects the results that can be inferred by modeling movemet behavior.
+@nathan_big-data_2022 provides additional examples of how the frequency of location data affects the results that can be inferred by modeling movement behaviour.
 Consequently, we suggest being cautious when comparing estimated movement behaviours across species, even though all predictions have been corrected to the hourly timescale by predicting for 1-hour time intervals (i.e., `dt_hours = 1`).
@@
 
-Temperature affects many aspects of mammalian physiology and behavior, including energetics, sleep, and movement behavior.
+Temperature affects many aspects of mammalian physiology and behaviour, including energetics, sleep, and movement behaviour.
 Climatic changes during the last two centuries have exposed many mammals to novel and increasingly extreme environmental conditions that have led to visible changes that impact not only individuals but also ecological and human communities.
@@
 Accurately quantifying the nonlinear effects of temperature on when, how, and where mammals move requires careful data collection and model design.
-We have presented a multi-species analysis of the effects of temperature on mammalian movement rates and habitat selection in hopes that other researchers can leverage this framework and models to expand our understanding of how temperature affects the movement behavior of other species, including smaller mammals, more elusive species, and non-mammal animals.
-Understanding of how temperature affects each species in a community will allow us to begin making community-level inferences of how temperature will affect intra-specific interactions from a movement behavior perspective [@nathan_movement_2008; @martinez-garcia_how_2020; @nathan_big-data_2022].
+We have presented a multi-species analysis of the effects of temperature on mammalian movement rates and habitat selection in hopes that other researchers can leverage this framework and models to expand our understanding of how temperature affects the movement behaviour of other species, including smaller mammals, more elusive species, and non-mammal animals.
+Understanding of how temperature affects each species in a community will allow us to begin making community-level inferences of how temperature will affect intra-specific interactions from a movement behaviour perspective [@nathan_movement_2008; @martinez-garcia_how_2020; @nathan_big-data_2022].
 While adopting such a perspective is surely not simple, it would result in more complete estimates of the effects of climate change on community ecology, particularly when species are exposed to rapid change in many important variables.
-Future work should explore the effects of temperature on movement behavior while accounting for finer-scale and species specific variables that were not accounted for in this study.
+Future work should explore the effects of temperature on movement behaviour while accounting for finer-scale and species specific variables that were not accounted for in this study.
 Examples of these include temporally dynamic measures of forest type and age, canopy density, competitive pressures, forage availability, and predator avoidance, or environmental stochasticity.

### 2025-09-02 writing/manuscript.Rmd
updated first paragraph and table of results

@@
 In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
-Using over 25 years of telemetry data throughout a large spatial range of western Canada and hourly air temperature data, we estimate how these mammals altered their probability of movement, speed when moving, and habitat selection in response to temperature.
+Using over 25 years of telemetry data throughout a large spatial range of western Canada and hourly air temperature data, we estimate how these mammals altered their probability of moving, speed when moving, and habitat selection in response to temperature.
 We then pair the estimated responses with climate change projections to forecast the species' movement behaviour during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017].
@@
 
-\noindent Overall, 2.6% of GPS locations had temperatures lower than $-20^\circ$C, while 6.5% had temperatures above $20^\circ$C, and temperature ranges differed across species (Table \@ref(tab:temp-qs), Fig.
+\noindent Overall, 2.6% of GPS locations had temperatures lower than $-20^\circ$C, while 6.5% had temperatures above $20^\circ$C, and temperature ranges differed across species (Table \@ref(tab:hgam-summary), Fig.
 B2).
-At $0^\circ$C, species differed in overall mean probabilities of movement (range: 0.05 -- 0.31), mean speed when moving (range: 0.42 -- 2.67 km/h), and mean distance traveled (i.e., $P(M) \times \mathbb E(S)$, range: 0.04 -- 0.61 km/h; Table \@ref(tab:gam-intercepts)).
-Grizzly bears tended to move least often ($P(M) \approx 0.05$), while wolves and cougars moved most often ($P(M) \ge 0.22$).
-When moving, mountain goats and southern mountain caribou moved the slowest ($\mathbb E(S|M) \approx 0.43$ km/h), while wolves had the highest mean speed when moving ($\mathbb E(S|M) \approx 2.64$ km/h).
-Consequently, wolves traveled, on average, 0.6 km/h -- 2.5 to 16.7 times further than other mammals at $0^\circ$C.
+Sampling interval affected estimates of probability as well of as speed when moving (Fig.
+B8).
+Consequently, we present all results while predicting specifically for one-hour sampling intervals.
+At $0^\circ$C, species differed in mean probabilities of moving ($\hat P(M)$; range: 0.05 -- 0.31), mean speed when moving ($\hat {\mathbb E}(S|M=1)$; range: 0.42 -- 2.67 km/h), and mean distance traveled (i.e., $\hat P(M) \times \hat {\mathbb E}(S|M=1)$, range: 0.04 -- 0.61 km/h; Table \@ref(tab:hgam-summary)).
+Grizzly bears had the lowest movement frequency ($\hat P(M) \approx 0.05$), while wolves and cougars moved most often ($\hat P(M) \ge 0.22$).
+Mountain goats and southern mountain caribou moved the slowest ($\hat{\mathbb E}(S|M = 1) \approx 0.43$ km/h), while wolves had the highest mean speed when moving ($\hat{\mathbb E}(S|M = 1) \approx 2.67$ km/h).
+Consequently, at $0^\circ$C, wolves traveled an average of $0.22 \times 2.67~\text{km/h} \approx 0.6$ km/h; 2.5 to 16.7 times further than other species.
 
-```{r temp-qs}
+\tiny
+
+```{r hgam-summary}
+setwd('~/GitHub/bc-mammals-temperature')
+library('mgcv')
 library('dplyr')
-d <- readRDS('../data/hgam-speed-data.rds')
+library('ctmm') #' for `%#%`
+source('analysis/figures/default-ggplot-theme.R')
+
+d <- readRDS('data/hgam-speed-data.rds')
 get_perc <- function(bools) {
-  out <- as.character(round(mean(bools) * 100, 1))
+  perc <- round(mean(bools) * 100, 1)
+  out <- as.character(perc)
   if(grepl('\\.', out)) {
@@
   }
+
+  if(perc < 10) out <- paste0(' ', out)
+
   return(out)
 }
-
-bind_rows(
-  d %>%
-    rename(Species = species) %>%
-    group_by(Species) %>%
-    summarize(`T < -20\\textdegree{} C` = get_perc(temp_c < -20),
-              `T > +20\\textdegree{} C` = get_perc(temp_c > +20)),
-  d %>%
-    summarize(Species = 'Total',
-              `T < -20\\textdegree{} C` = get_perc(temp_c < -20),
-              `T > +20\\textdegree{} C` = get_perc(temp_c > +20))) %>%
-  knitr::kable(format = 'markdown', caption = paste0("Percentage of observed GPS locations (after data cleaning) with temperatures below -20\U00B0", "C or above 20\U00B0", "C."))
-```
-
-\scriptsize
-
-```{r gam-intercepts}
-setwd('~/GitHub/bc-mammals-temperature')
-library('mgcv')
-library('dplyr')
-library('ctmm') #' for `%#%`
-source('analysis/figures/default-ggplot-theme.R')
 
@@
   Species = SPECIES,
-  'P(M)' = get_int(readRDS('models/binomial-gam-without-temperature.rds')),
-  'P(M|T)' = get_int(readRDS('models/binomial-gam.rds')),
+  'P(M=1)' = get_int(readRDS('models/binomial-gam-without-temperature.rds')),
+  'P(M=1|T)' = get_int(readRDS('models/binomial-gam.rds')),
   'E(S)' = 'km/h' %#% get_int(readRDS('models/gamma-gam-without-temperature.rds')),
   'E(S|T)' = 'km/h' %#% get_int(readRDS('models/gamma-gam.rds')),
-  'E(D)' = `P(M)` * (`E(S)`),
-  'E(D|T)' = `P(M|T)` * (`E(S|T)`)) %>%
+  'E(D)' = `P(M=1)` * (`E(S)`),
+  'E(D|T)' = `P(M=1|T)` * (`E(S|T)`)) %>%
   mutate(Species = case_when(
-    Species == 'Rangifer tarandus (s. mountain)' ~ 'R. tarandus (sm)',
-    Species == 'Ursus arctos horribilis' ~ 'U. arctos horr.',
-    Species == 'Puma concolor' ~ 'Puma concolor',
+    Species == 'Rangifer tarandus (s. mountain)' ~ 'R. tarandus; sm',
+    Species == 'Ursus arctos horribilis' ~ 'U. arctos h.',
+    Species == 'Puma concolor' ~ 'P. concolor',
     Species == 'Cervus canadensis' ~ 'C. canadensis',
-    Species == 'Rangifer tarandus (boreal)' ~ 'R. tarandus (b)',
+    Species == 'Rangifer tarandus (boreal)' ~ 'R. tarandus; b',
     Species == 'Canis lupus' ~ 'Canis lupus',
@@
 if(FALSE) {
-  hist(intercepts$`P(M)`)
-  hist(intercepts$`E(S)`)
-  round(range(max(intercepts$`E(D)`) / intercepts$`E(D)`[- which.max(intercepts$`E(D)`)]), 1)
+  # visualize the individual estimates
+  layout(1:3)
+  hist(intercepts$`P(M=1)`); hist(intercepts$`E(S)`); hist(intercepts$`E(D)`)
+  layout(1)
 
+  # find differences when accounting for temperature
   intercepts %>%
     transmute(Species,
-              p = round(`P(M|T)` - `P(M)`, 2),
+              p = round(`P(M=1|T)` - `P(M=1)`, 2),
               s = round(`E(S|T)` - `E(S)`, 2),
@@
 
+  # find difference in distance traveled, relative to furtherst distance
   intercepts %>%
@@
            displ_wolf_t = `E(D|T)`[grepl('Canis lupus', Species)] / `E(D|T)`)
+
+  # find difference in distance traveled, relative to furtherst distance
+  round(range(max(intercepts$`E(D|T)`) / intercepts$`E(D|T)`[- which.max(intercepts$`E(D|T)`)]), 1)
 }
 
-intercepts %>%
+options(knitr.kable.NA = '') # leave NAs blank
+
+bind_rows(
+  # species-level proportions
+  d %>%
+    rename(Species = species) %>%
+    group_by(Species) %>%
+    summarize(`T<-20\\textdegree{}C` = get_perc(temp_c < -20),
+              `T>+20\\textdegree{}C` = get_perc(temp_c > +20)),
+  # overall proportions
+  d %>%
+    summarize(Species = 'Total',
+              `T<-20\\textdegree{}C` = get_perc(temp_c < -20),
+              `T>+20\\textdegree{}C` = get_perc(temp_c > +20))) %>%
+  select(! Species) %>%
+  bind_cols(intercepts %>%
               arrange(Species) %>%
@@
               rename_with(\(.x) gsub('E', '\\\\hat{\\\\mathbb E}', .x)) %>%
-  knitr::kable(format = 'markdown', digits = 2, caption = paste0("Mean probability of movement (P(M)), speed ($\\mathbb E(S)$, km/h, assuming constant movement), and distance travelled ($P(M) \\times \\mathbb E(S) = \\mathbb E(D)$, km/h) as estimated by models without and with temperature ($\\mathbb E(...
-| T)$), after post-stratifting to a 1-hour sampling rate and a temperature of T = 0\U00B0", "C."))
+              bind_rows(tibble(Species = 'Total'))) %>%
+  relocate(Species, .before = 1) %>%
+  rename(`$\\hat{P}$(M = 1)` = `$\\hat{P}(M=1)$`) %>% # to avoid eq going to new line
+  knitr::kable(format = 'markdown', digits = 2, caption = paste0("Summary statistics for the observed temperatures and estimated mean movement rates at \U00B0", "C.
+The second and third columns indicate the percentage of GPS locations (after data cleaning) with temperature ($T$) below -20\U00B0", "C and above 20\U00B0", "C, respectively.
+The subsequent columns indicate the estimated mean probability of moving ($\\hat P(M=1)$), mean speed when moving ($\\hat{\\mathbb E}(S)$; km/h), and mean distance travelled ($\\hat P(M=1) \\times \\hat{\\mathbb E}(S) = \\hat{\\mathbb E}(D)$; km/h) as estimated by models without and with temperature (i.e., $(...
+| T)$), for a sampling interval of 1 hour and a temperature of T = 0\U00B0", "C."))
 ```

### 2025-09-03 analysis/figures/temperature-hrsfs.R
lowered offset for boreal caribou elev_m

@@ -164,7 +164,7 @@ p <-
          lambda = case_when(
            species == 'Cervus canadensis' & variable == 'bold(Elevation~"(km)")' ~ lambda / exp(1),
            species == 'Oreamnos americanus' & variable == 'bold(Elevation~"(km)")' ~ lambda * exp(0.25),
-           species == 'Rangifer tarandus (boreal)' & variable == 'bold(Elevation~"(km)")' ~ lambda / exp(4),
+           species == 'Rangifer tarandus (boreal)' & variable == 'bold(Elevation~"(km)")' ~ lambda / exp(2),
            species == 'Oreamnos americanus' & variable == 'bold(Distance~from~water~"(km)")' ~ lambda * exp(0.5),
            species == 'Rangifer tarandus (s. mountain)' & variable == 'bold(Distance~from~water~"(km)")' ~ lambda * exp(0.2),
            species == 'Ursus arctos horribilis' & variable == 'bold(Distance~from~water~"(km)")' ~ lambda * exp(0.5),

### 2025-09-03 analysis/figures/temperature-hrsfs-spatial-projections.R
created density plot of RSS change by 2100

@@ -179,6 +179,24 @@ plot_grid(make_plot(SPECIES_LABS[1], get_legend = TRUE),
 ggsave('figures/local-rss-2100.png',
        width = 17.8, height = 12.4, units = 'in', dpi = 600, bg = 'white')
 
+# make density plots of the current range
+cc_proj %>%
+  mutate(l = if_else(l > 1.3, 1.3, l)) %>%
+  mutate(scenario = factor(gsub('"', '', scenario),
+                           levels = gsub('"', '', levels(scenario)))) %>%
+  ggplot(aes(x = l, fill = scenario, color = scenario)) +
+  facet_wrap(~ lab, scales = 'free_y', labeller = label_parsed, nrow = 2) +
+  geom_density(alpha = 0.25) +
+  geom_vline(xintercept = 1, color = 'black', lty = 'dashed') +
+  scale_x_continuous('Pixel-level relative change in RSS in 2100') +
+  ylab('Density') +
+  scale_color_brewer('Climate change scenario', type = 'div', palette = 5,
+                     direction = -1, aesthetics = c('color', 'fill')) +
+  theme(legend.position = 'inside', legend.position.inside = c(7/8, 1/4))
+
+ggsave('figures/local-rss-2100-density.png',
+       width = 12, height = 6.67, dpi = 600, bg = 'white')
+
 # maps of resources ----
 plot_resources <- function(sp, resource) {
   .d <-

### 2025-09-03 analysis/figures/temperature-hrsfs-spatial-projections.R
simplified xlab

@@ -188,7 +188,7 @@ cc_proj %>%
   facet_wrap(~ lab, scales = 'free_y', labeller = label_parsed, nrow = 2) +
   geom_density(alpha = 0.25) +
   geom_vline(xintercept = 1, color = 'black', lty = 'dashed') +
-  scale_x_continuous('Pixel-level relative change in RSS in 2100') +
+  scale_x_continuous('Relative change in RSS in 2100') +
   ylab('Density') +
   scale_color_brewer('Climate change scenario', type = 'div', palette = 5,
                      direction = -1, aesthetics = c('color', 'fill')) +

### 2025-09-03 writing/appendix-b-gams.Rmd
added note on BC albers projection in fig cap

@@
 
-```{r resources, fig.cap="Resource rasters used for each species' hierarchical resource selection function."}
+```{r resources, fig.cap="Resource rasters used for each species' hierarchical resource selection function. All maps are shown in the BC Albers Equal Area Conic projection (EPSG:3005)."}
 knitr::include_graphics('../figures/local-resources.png')

### 2025-09-03 writing/manuscript.Rmd
reviewed results

@@
 \begin{cases}
-D \sim \text{Pois}(\lambda) \\
-\mathbb E(D) = \mathbb V(D) = \lambda \\
+C \sim \text{Pois}(\lambda) \\
+\mathbb E(C) = \mathbb V(C) = \lambda \\
 \begin{aligned}
@@
 
-\noindent where the species-level indices are omitted for readability, but each term in the model can be assumed to be species-specific.
+\noindent where $O$ indicates whether an animal was observed ($O = 1$) or not ($O = 0$), and the species-level indices are omitted for readability, but each term in the model can be assumed to be species-specific.
 Smooth effects of percent forest cover (\texttt{forest\_perc}), elevation (\texttt{elev\_m}), and distance to water (\texttt{dist\_water\_m}) accounted for the species-level selection strength for each resource.
@@
 
-Near 0$^\circ$C, wolves selected for dense forest cover ($\gtrapprox$ 50%), elevations near 1 km, and distances from water < 10 km; elk selected for forest cover near 50%, elevations between 1 and 2 km, and distances from water of 5-15 km; mountain goats selected for sparse (< 50%) forest cover, elevations between 1 and 2 km; cougars selected for dense forest cover, an elevation of ~ 1 km, and distances from water < 10 km; boreal caribou selected for intermediate to dense forest cover, elevations near 500 m, and distances from water < 10 km; southern mountain caribou selected for dense forest cover, elevations near 2 km, and distances from water < 5 km; and grizzly bears selected for sparse forest cover (25-50%), elevation near 1 km, and distances from water < 2 km.
+Across all species, Relative Selection Strength (RSS) was weakest for forest cover and strongest for elevation.
+At temperatures near 0$^\circ$C, wolves selected for forest cover ($\gtrapprox$ 50%), elevations near 1 km, and distances from water < 5 km; elk selected for intermediate forest cover ($\approx$ 50%), elevations between 1 and 2 km, and distances from water of 10-15 km; mountain goats selected for sparse forest cover (< 25%), elevations near 1.5 km, and distances from water < 5 km; cougars selected for dense forest cover (> 75%), an elevation of ~ 1 km, and distances from water < 7.5 km; boreal caribou selected for forest cover between 50% and 75%, elevations near 500 m, and distances from water < 10 km; southern mountain caribou selected for dense forest cover, elevations near 2 km, and distances from water $\lessapprox$ 5 km; and grizzly bears selected for relatively sparse forest cover (25-50%), elevation between 1 and 2 km, and distances from water < 3 km.
 
 There was relatively strong agreement between models with and without temperature (Figs.
-B3, and B14), but including temperature always resulted in better fits.
-All analyses of deviance showed that temperature significantly affected movement behaviour (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; Table B2 and following pages in Appendix B).
-Accounting for temperature resulted in a simpler HRSF for mountain goats (lower model degrees of freedom) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.
+B3, and B14), but including temperature always resulted in better fits (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; Table B2 and subsequent pages in Appendix B).
+Accounting for temperature even resulted in a simpler HRSF for mountain goats (lower model degrees of freedom) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.
 
@@
 
-\noindent The effects of temperature on movement rates varied in both direction and magnitude across species, even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; Figs.
+\noindent Species' changes in movement rates to temperature varied in both direction and magnitude (Figs.
+\@ref(fig:dist), B4-B6), even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; Figs.
 B4-B6).
@@
 The models had good in-sample prediction (Fig.
-B7) and explained reasonably high proportions of the deviance (79.3% for the gamma model and 10.7% for the binomial model, which is relatively high for a binomial model with binary responses: see ch.
+B7) and explained reasonably high proportions of the deviance [79.3% for the gamma model and 10.7% for the binomial model, which is relatively high for a binomial model with binary responses: see ch.
 11 of @mcelreath_statistical_2020].
@@
 The clearest example of this was cougars.
-In mid summer, they moved mostly in the evening if temperatures were cool and mostly in the early morning if temperatures were hot.
-Throughout the year, they tended to move more when it was colder, but what they perceived as "colder" depended on the time of year.
-However, uncertainties around the estimated effects were often higher at extreme temperatures due to data scarcity.
-All species' estimated probability of moving decreased with sampling intervals approximately above 1 hour, and most species' estimated speed when moving decreased monotonically with sampling interval, although the estimated trends were highly uncertain for some species (Fig.
+In late spring (June 1^st^), they moved from evening to early morning if temperatures were below $20^\circ$C, but if temperatures were above 20$^\circ$C they moved mostly between 3:00 and 6:00 AM.
+Throughout the year, they tended to move more when it was colder, but what they perceived as "colder" was relative to the time of year.
+All species' estimated probability of moving and speed when moving decreased with sampling intervals above 1 hour, except for cougars' speed, although the estimated trends were highly uncertain (Fig.
 B8).
+Uncertainties around the estimated effects were generally higher at extreme temperatures due to lower data availability (Figs.
+B4A, B5A, and B6A).
 
-```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals travel.} The fill indicates the estimated effects of temperature on the relative change in distance travelled (probability of moving times mean speed when moving) over time of day on June $1^{\\text{st}}$ (\\textbf{A}) and day of year at 12:00 PDT (\\textbf{B}).
+```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals travel.} The fill indicates the effects of temperature on the relative change in distance travelled (probability of moving times mean speed when moving) over time of day on June $1^{\\text{st}}$ (\\textbf{A}) and day of year at 12:00 PDT (\\textbf{B}).
 Predictions in the surface plots extend to 10% of the range away from each datum.
 The color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to $2^{\\pm 2}$ for ease of readability."}
-# not specifying dt, or doy for A and time for B because change is relative
+# not specifying dt and doy for A and time for B because change is relative
 knitr::include_graphics('../figures/distance.png')
@@
 
-Species' relative selection strength (RSS) was generally strongest for elevation and weakest for forest cover, but species' RSS for resources depended on temperature (Fig.
+Species' RSS was generally strongest for elevation and weakest for forest cover, but RSS depended significantly on temperature for all species (all p-values $< 2.2\times10^{-16}$; Fig.
 \@ref(fig:hrsfs)).
-Changes in RSS with temperature were also strongest for elevation and generally weakest distance from water, but there were no common trends across all species for any of the three resources.
-All species, with the exception of cougars, exhibited a clear temperature-dependent shift in their preference for forest cover.
-At higher temperatures, wolves, mountain goats, and grizzly bears became less selective for forest cover, while elk and caribou shifted towards more intermediate forest cover without much of a change in preference width.
-All species shifted elevationally with temperature, although boreal caribou's selection strength for elevation was less sensitive to elevation.
-Overall, as temperatures rose, elk, mountain goats, and cougars increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation.
-Most species generally remained within 5 km of water, and temperature did not affect their selection strength as much as for the other two resources.
-Again, estimated RSS values were generally most uncertain at extreme temperatures (Fig.
+Changes in RSS with temperature were strongest for elevation and generally weakest for distance from water, but there were no common trends across all species for any of the three resources.
+All species, with the exception of cougars, exhibited clear temperature-dependent shifts in their preference for forest cover.
+At higher temperatures, wolves relaxed their preference for forested areas, while mountain goats relaxed their preference open areas (cover < 50%).
+As temperatures warmed, elk and boreal caribou shifted towards more forest cover closer to 50%, while southern mountain caribou and grizzly bears selected for areas with 50% forest cover or less.
+All species shifted elevationally with temperature, although boreal caribou ha a much sharper change in RSS across elevation.
+Elk, mountain goats, and cougars increased in elevation with temperature, while wolves, southern mountain caribou, and grizzly bears decreased in elevation.
+Wolves, elk, and southern mountain caribou moved closer to water with temperature, while mountain goats, cougars, and grizzly bears moved somewhat further away from water but remained mainly within 5-10 of water.
+As with movement rates, estimated RSS was generally most uncertain at extreme temperatures, for which data was scarcer (Fig.
 B15).
 
-```{r hrsfs, fig.cap="\\textbf{Temeperature strongly affects mammals' habitat selection.} Estimated relative selection strength (RSS) for forest cover (\\%), elevation (km), and distance from water (km) as a function of temperature.
+```{r hrsfs, fig.cap="\\textbf{Mammals' habitat selection depends on ambient temperature.} Estimated relative selection strength (RSS) for forest cover (\\%), elevation (km), and distance from water (km) as a function of temperature.
 The grey contours indicate the extent of each species' observed data.
@@
 
-## Predicted changes in movement during the 21^st^ century
+## Predicted changes in movement behaviour during the 21^st^ century
 
-\noindent The predicted changes in distance traveled varied across species in both direction and magnitude, but worse climate-change SSPs always corresponded to greater absolute changes (Fig.
+\noindent Predicted changes in movement rates with future climate change varied across species in both magnitude and direction, but worse SSPs always corresponded to greater absolute changes (Fig.
 \@ref(fig:t-dist)).
@@
 B9 and B10).
-Overall, absolute changes by 2100 under the best-case SSP were small (approximately 0% to +4%), while under the worst-case SSP absolute changes ranged from ~2% (grizzly bears) to ~24% (cougars).
-Notably, while the models estimated that grizzly bears would move substantially less (if at all) in winter (Figs.
-\@ref(fig:dist), B4-B6), the models did not explicitly account for changes in hibernation phenology.
+Absolute relative changes in distance travelled were small under the best-case SSP (0-4% change in 2100 relative to 2025).
+Under the worst-case SSP, absolute changes by 2100 (relative to 2025) ranged from ~2% (grizzly bears) to ~24% (cougars).
+Notably, while the models estimated that grizzly bears would move substantially less (if at all) in late fall and winter (Figs.
+\@ref(fig:dist), B4), the models did not explicitly account for changes in hibernation phenology.
 Consequently, the climate change projections do not explicitly account for changes in energetic needs or physiology.
-Projected changes also varied spatially due to heterogeneity in projected climate change (Fig.
+Projected changes in 2100 varied spatially due to spatial heterogeneity in climate change projections (Fig.
 \@ref(fig:s-dist)).
@@
 
-```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in distance travelled (probability of moving times speed when moving) due to changes in temperature within each species' observed extent.
-Shaded areas indicate the range of predicted values between the $95^\\text{th}$ and $5^\\text{th}$ percentiles.
-Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs).
-If the intervals are fully above the dashed line, at least 90% of the estimated means increased, relative to the each pixel's average of the four climate SSPs in 2025.
-Similar conclusions can be drawn regarding a decrease if the ribbons are below the dashed line.
-The projections only account for changes in movement behaviour (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
+```{r t-dist, fig.cap="\\textbf{Species will adapt their movement rates differently in response to climate change, but worse climate-change scenarios will result in the greatest change.} Bold lines indicate the median change in distance travelled (probability of moving times speed when moving) due to predicted changes in temperature within each species' observed extent.
+Shaded areas indicate the $95^\\text{th}$ and $5^\\text{th}$ percentiles.
+Changes are relative to the mean predicted distance travelled at each location in 2025, averaged across the four Shared Socioeconomic Pathways (SSPs).
+Values > 1 indicate an increase, while values < 1 indicate a decrease.
+The projections only account for changes in movement frequency and speed, and they ignore changes in physiology or movement costs."}
 knitr::include_graphics('../figures/distance-travelled-local-cc-predictions.png')
@@
 
-```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each pixel's average distance in 2025 across all four scenarios (i.e., not relative to other pixels).
+```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently, but changes will also vary spatially.} The color of each pixel indicates the predicted changes in distance traveled in 2100 at that pixel, relative to the value in 2025, averaged across all four scenarios.
 Values < 1 indicate a decrease, and values > 1 indicate an increase.
-For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.50 to 1.56).
-The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest density.
-All maps are shown with a BC Albers projection (EPSG:3005)."}
+For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling).
+The predictions only account for the predicted temperature throughout the areas and ignore environmental factors such as terrain slope, soil type, and forest density.
+All maps are shown in the BC Albers Equal Area Conic projection (EPSG:3005)."}
 if(FALSE) range(readRDS('../data/cc-hgam-projections-local-2100.rds')$d)
@@
 
-Median RSS was projected to decrease over time within the each species' observed range, but changes were stronger under worse SSPs (Fig.
-\@ref(fig:s-hrsfs)).
-Decreases were most pronounced in the bottom fifth percentile and most severe for elk, southern mountain caribou, and mountain goats.
-Of all the species, only cougars showed a clear increase in RSS for areas within their current range.
-The change in RSS between 2025 and 2100 varied spatially for all species (Fig.
-\@ref(fig:s-hrsfs)).
-All species were predicted to exhibit a decrease in RSS throughout most of their observed habitats, although cougars were predicted to increase their selection for patches with higher altitude.
+Median RSS was projected to decrease over time within the each species' observed range, but, again, changes were stronger under worse SSPs (Fig.
+\@ref(fig:t-rss-cc)).
+Decreases were most pronounced in areas with the lowest RSS and most severe for elk, mountain goats, pumas, and southern mountain caribou.
+Changes for boreal wolves and caribou were relatively negligible.
+Elk, cougars, and grizzly bears were predicted to increase their selection strength for top-RSS areas, and elk, mountain goats, cougars, and southern mountain caribou were predicted to further decrease their selection for areas with low RSS.
+Unsurprisingly, the change in RSS between 2025 and 2100 also varied spatially for all species (Fig.
+\@ref(fig:s-rss-cc)).
+Overall, RSS decreased throughout most of each species' current range, although elk, cougars, and bears were predicted to increase their selection for higher-altitude habitats.
+Still, none of the species were projected to increase RSS throughout their habitat (\@ref(fig:density-hrsfs)).
 
@@
 
-```{r t-hrsfs, fig.cap="\\textbf{All species are expected to exhibit a decrease in median relative selection strength (RSS) for their current range, irrespective of climate change scenario, but decreases are stronger in the worse-case scenario.} Lines indicate the projected change in median RSS within each species' observed extent within BC, while the ribbons indicate the range between the $95^\\text{th}$ and $5^\\text{th}$ percentiles in RSS.
-Changes are relative to each location's mean RSS in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
+```{r t-rss-cc, fig.cap="\\textbf{Species will adapt their habitat selection differently in response to climate change, but worse climate-change scenarios will result in the greatest change.} Bold lines indicate the change in median RSS due to predicted changes in temperature within each species' observed extent.
+Shaded areas indicate the $95^\\text{th}$ and $5^\\text{th}$ percentiles in change in RSS.
+Values > 1 indicate an increase, while values < 1 indicate a decrease.
+Changes are relative to each location's mean RSS in 2025, averaged across the four Shared Socioeconomic Pathways (SSPs)."}
 knitr::include_graphics('../figures/rss-local-cc-predictions.png')
@@
 
-```{r s-hrsfs, fig.cap="\\textbf{Climate change will impact each species' habitat relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios (i.e., not relative to other pixels).
+```{r s-rss-cc, fig.cap="\\textbf{Climate change will impact each species' relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios (i.e., not relative to other pixels).
 Values < 1 indicate a decrease and values > 1 indicate an increase.
-For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.66 to 1.86).
-All maps are shown with a BC Albers projection (EPSG:3005)."}
+For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.8 and 1.25 ($\\approx 2^{\\pm 0.322}$; original data ranged 0.71 to 1.93).
+All maps are shown in the BC Albers Equal Area Conic projection (EPSG:3005)."}
 if(FALSE) range(readRDS('../data/cc-hrsf-projections-local-2100.rds')$l)
 knitr::include_graphics('../figures/local-rss-2100.png')
+```
+
+```{r density-hrsfs, fig.cap="\\textbf{Climate change will cause species to decrease their selection strength for their current habitats, but worse climate change scenarios will have the greatest impact.} The density plots indicate each species' change in RSS in 2100 for their current habitat (Fig.
+8), relative to each location's RSS in 2025.
+Values < 1 indicate a decrease and values > 1 indicate an increase.
+For ease of readability, values > 1.3 were changed to 1.3 (original maximum: 1.93)."}
+knitr::include_graphics('../figures/local-rss-2100-density.png')
 ```
@@
 
+***Overall, estimated selection strength was weakest for forest cover and strongest for elevation.
+Elevation is the only raster that doesn't change noticeably over time, so time-varying rasters of the other two would have likely given stronger (and better) estimates.***
+
 As with the effects of temperature on movement rates, the hererogeneity of temperature effects on species' habitat selection indicates that efficient and proactive habitat conservation will require taxonomic nuance.
@@
 The diversity in predicted changes in movement behaviour by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in Fig.
-\@ref(fig:s-hrsfs) highlight the need for careful planning that incorporates not only reliable estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates.
+\@ref(fig:s-rss-cc) highlight the need for careful planning that incorporates not only reliable estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates.
 For example, the width of the prediction intervals in Fig.

### 2025-09-04 writing/manuscript.Rmd
started editing the discussion

@@
 Consequently, our results suggest that all six species may shift their range in the coming decades as their selection strength for their current ranges weakens.
-As climate change exposes mammals to novel environmental conditions, predicting changes in animal behaviour and community structure will become crucial for effectively and proactively understanding mammalian movement behaviour and protecting high-quality habitat.
+As climate change exposes mammals to novel environmental conditions, predicting changes in mammalian behaviour and community structure will become crucial for effectively and proactively understanding mammalian movement behaviour and protecting high-quality habitat.
 We thus present our work using a flexible approach that can be adapted for different species and spatiotemporal scales, and we conclude by providing suggestions for future research, with particular attention to the study design, statistical modeling, and forecasting."
@@
 
-# other refs
-
-* @gerlich_species_2025: effects of temperature on fly movement in the arctic
-* "Potential changes in climate indices in Alberta under projected global warming of 1.5–5 $^\circ$C" (https://www.sciencedirect.com/science/article/pii/S2214581823000770#fig0015)
-* Global vulnerability of marine mammals to global warming: https://doi.org/10.1038/s41598-019-57280-3
-* marine turtles shift and lose habitat with temperature: https://www.science.org/doi/10.1126/sciadv.adw4495
-* changes in temperature change phelogy and movement behaviour as a consequence @gerlich_species_2025
-* @walker_global_2019: Global climate change and invariable photoperiods: {A} mismatch that jeopardizes animal fitness
-\clearpage
-
 # Introduction
@@
 Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @merkle_large_2016; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024; @schwerdt_ecological_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020].
-However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and the risk of hyperthermia [@hetem_responses_2014; @dyer_travel_2023].
+However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and extreme temperatures can reduce movement rates and cause physiological damage [@hetem_responses_2014; @dyer_travel_2023].
 Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [@hall_microrefuges_2016].
@@
 
-Western Canada is currently experiencing rapid widespread warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme events [@zhang_increased_2023], including forest fires [@zurowski_summer_2023].
+Western Canada is currently experiencing rapid widespread warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme events [@zhang_increased_2023].
 As we approach the deadline for the Kunming-Montreal Global Biodiversity Framework (i.e., the "30 by 30" initiative), a global effort to conserve 30% of the world's lands and oceans by 2030 [@convention_on_biological_diversity_154_2022], Canadian decision makers must understand which areas will be of greatest value for maximizing and protecting biodiversity not only for current conditions but future decades, too.
 Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.
-Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022].
+Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022;  @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].
 
@@
 
-\noindent We have demonstrated that temperature is an important determinant of whether, how, and where large mammals move, even while accounting for seasonality (e.g, changes in photoperiod, hormones, and resources).
-Disentangling temperature from other drivers is important for predicting how changes in climate will affect mammalian movement when other drivers remain approximately constant.
-However, predicting mammals' responses to climate change becomes increasingly complex as habitats warm and animals are exposed to increasingly frequent conditions that are both extreme and novel.
+\noindent We have demonstrated that temperature is an important determinant of how and where large mammals move, including how they alter seasonal and daily cycles in movement behavior.
+However, predicting mammals' responses to climate change becomes increasingly complex as habitats warm and animals are increasingly frequently exposed to extreme and novel conditions.
+
+***Temperature affects many aspects of mammalian physiology and behaviour, including energetics, sleep, and movement behaviour.
+Climatic changes during the last two centuries have exposed many mammals to novel and increasingly extreme environmental conditions that have led to visible changes that impact not only individuals but also ecological and human communities.
+However, quantifying the effects of climate change is often complex, especially as conditions become increasingly different and extreme events become increasingly common.
+Accurately quantifying the nonlinear effects of temperature on how and where mammals move requires careful data collection and model design.
+We have presented a multi-species analysis of the effects of temperature on mammalian movement rates and habitat selection.
+Understanding of how temperature affects each species in a community will allow us to begin making community-level inferences of how temperature will affect intra-specific interactions from a movement behaviour perspective [@nathan_movement_2008; @martinez-garcia_how_2020; @nathan_big-data_2022].
+While adopting such a perspective is surely not simple, it would result in more complete estimates of the effects of climate change on community ecology, particularly when species are exposed to rapid change in many important variables.
+Future work should explore the effects of temperature on movement behaviour while accounting for finer-scale and species specific variables that were not accounted for in this study.
+Examples of these include temporally dynamic measures of forest type and age, canopy density, competitive pressures, forage availability, and predator avoidance, or environmental stochasticity.
+However, many of these data not available, so we also suggest that more work be done on quantifying such measures using spatiotemporally dynamic models and sufficiently fine resolutions.***
 
@@
 \@ref(fig:dist) and \@ref(fig:hrsfs) comes with substantial uncertainty, particularly given that the responses are nonlinear.
-At warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behaviour and that of their competitors, predators, and/or prey can often be substantially different from the behaviour at typical temperatures.
+At warm temperatures, mammals overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behaviour and that of their competitors, predators, and/or prey can often be substantially different from the behaviour at typical temperatures.
 As extreme temperature events become more frequent and prolonged throughout the year [@yao_emergence_2022; @ipcc_climate_2023], mammals will be increasingly forced towards atypical behaviours that may alter community structures and behaviours, both during such events and afterwards [@zhang_increased_2023; @logares_black_2012; @anderson_black-swan_2017].
-For instance, changes in climate and phenology impact the life history and behaviour of many hibernating mammals [@wells_life_2022], and hot weather can also affects mammals' sleep quality [@mortlock_sleep_2024] and likelihood to enter torpor [@fjelldal_nightly_2021].
+For instance, changes in climate and phenology impact the life history and behaviour of many hibernating mammals [@wells_life_2022], and hot weather can also affect mammals' sleep quality and likelihood to enter torpor [@mortlock_sleep_2024; @fjelldal_nightly_2021].
 
-Changes in animal movement behaviour may also alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024].
+Changes in mammal movement behaviour may also alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024].
 At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023], increase their chances of finding food or being preyed on [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021].
@@
 
-Our ability to respond changes in climate is contingent on our ability to prepare for and predict change.
-However, predicting animal behaviour becomes increasingly complicated as the conditions animals are exposed to deviate from current, typical conditions, especially when responses are nonlinear and data are sparse.
-Consequently, we do not present our results as a definitive guide to how mammals in western Canada will respond to climate change.
-Instead, we hope they serve as a starting point to (1) demonstrate that mammals' movement rates and habitat selection depend on temperature and (2) how one can estimate mammals' changes in movement behaviour due to climate change.
-Additionally, communicating uncertainty in one's estimates is crucial in assessing risk probabilistically, and appropriate measures of uncertainty require careful and robust modeling [@aven_implementing_2002; @ayre_bayesian_2012; @kluppelberg_bayesian_2014].
-
-*Achieving the "30 by 30" goal will require international collaboration [@huang_achieving_2021] as well as active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023].
-Understanding the consequences of climate change on mammals' movement behaviour is a first step towards proactively responding to how mammals will respond to human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018].
-In the following sections, we discuss the implications of our results in more detail.
-We then expand on consequences for conservation during the 21^st^ century and considerations for future studies.*
+***Protecting a third of the world's ecosystems in accordance with the Kunming-Montreal Global Biodiversity Framework [@convention_on_biological_diversity_154_2022] appears less achievable as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023].
+Achieving the "30 by 30" goal will require international collaboration [@huang_achieving_2021] as well as active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023].
+Understanding the consequences of climate change on mammals' movement behaviour is a first step towards proactively responding to how mammals will respond to human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018].***
 
@@
 Additionally, it is important to stress the distinction between habitat selection and quality.
-While animals tend to select for high-quality habitats [@kacelnik_ideal_1992; @wright_survival_2024], high selection strength can also be a consequence of an ecological trap [@zuniga-palacios_what_2021; @swearer_overview_2021].
+While individuals tend to select for high-quality habitats [@kacelnik_ideal_1992; @wright_survival_2024], high selection strength can also be a consequence of an ecological trap [@zuniga-palacios_what_2021; @swearer_overview_2021].
 
@@
 Achieving the goals laid out by the "30 by 30" conservation initiative [@convention_on_biological_diversity_154_2022] will require careful planning but often also rapid action.
-In the case of Canada, only 13.7% of its terrestrial areas and inland waters were protected at the end of 2023 [@environment_and_climate_change_canada_canadian_2024; @minister_of_environment_and_climate_change_act_2024].
-Additionally, efficient and effective conservation will require collaboration with Indigenous and local communities to ensure Treaties are respected in the process [@falardeau_biophysical_2022; @lamb_braiding_2023; @wong_towards_2020; @turner_its_2009] while also actively preparing for future change [@desjardins_looking_2020; @hessami_indigenizing_2021].
+As of December 2024, only 13.8% of terrestrial Canada was within protected areas or other conserved areas [@environment_and_climate_change_canada_canadian_2025].
+Additionally, efficient and effective conservation will require collaboration with Indigenous and local communities to ensure Treaties are respected in the process [@falardeau_biophysical_2022; @lamb_braiding_2023; @wong_towards_2020; @turner_its_2009] and as we actively prepare for future change [@desjardins_looking_2020; @hessami_indigenizing_2021].
 
-The diversity in predicted changes in movement behaviour by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in Fig.
-\@ref(fig:s-rss-cc) highlight the need for careful planning that incorporates not only reliable estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates.
-For example, the width of the prediction intervals in Fig.
-\@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas [@jeltsch_need_2025].
+***...*** spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas [@jeltsch_need_2025].
 Population-level means that ignore such variation can greatly limit the efficacy of conservation projects [@muff_accounting_2020; @mortlock_sleep_2024; @silva_too_2025].
 Additionally, accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage and prey availability [@mezzini_how_2025], encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@smith_densitydependent_2023], competitive pressure [@torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019].
-This complexity results in the great variation among the responses of different species (and individuals), especially as populations undergo a rapid cascade of change [@botero_evolutionary_2015] and animals face increasing pressures from human activity, including habitat fragmentation, habitat loss, and greater human presence in wild spaces [@sih_evolution_2011; @tucker_moving_2018; @sawyer_identifying_2009; @weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022].
-As selection strength for current ranges changes (and likely decreases), some animals may disperse, relocate to a new habitat, or remain within the current range despite the reduced fitness and increased extinction risk [@logares_black_2012; @anderson_black-swan_2017; @duncan_predicting_2012].
+This complexity results in the great variation among the responses of different species (and individuals), especially as populations undergo a rapid cascade of change [@botero_evolutionary_2015] and mammals face increasing pressures from human activity, including habitat fragmentation, habitat loss, and greater human presence in wild spaces [@sih_evolution_2011; @tucker_moving_2018; @sawyer_identifying_2009; @weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022].
+As selection strength for current ranges changes (and likely decreases), some individuals may disperse, relocate to a new habitat, or remain within the current range despite the reduced fitness and increased risk of extirpation or extiction [@logares_black_2012; @anderson_black-swan_2017; @duncan_predicting_2012].
 
@@
 
-### *Canis lupus*
+* Cougars: attraction for disconnected patches of high elevation could lead to fragmentation or traps
 
-### *Cervus canadensis*
+* boreal caribou: Increasing temperatures increased RSS for both elevations near ~500 m as well as > 1,000 m, but such a selection is only possible if: (1) such habitats exist in the individuals' range, and (2) crossing the 500 m to > 1,000 m regions is safe and attractive
 
-### *Oreamnos americanus*
+* sm caribou: HRSFs do not account for migratory behaviour due to reproduction or predation risk
 
-### *Puma concolor*
-
-attraction for disconnected patches of high elevation could lead to fragmentation or traps
-
-### *Rangifer tarandus* (boreal)
-
-Increasing temperatures increased RSS for both elevations near ~500 m as well as > 1,000 m, but such a selection is only possible if: (1) such habitats exist in the animals' range, and (2) crossing the 500 m to > 1,000 m regions is safe and attractive
-
-### *Rangifer tarandus* (s. mountain)
-
-HRSFs do not account for migratory behaviour due to reproduction or predation risk
-
-### *Ursus arctos horribilis*
-
-## Considerations for future studies
+## Study limitation and considerations for future studies
 
@@
 While the `ctmm` movement models produced scale-independent estimates of speed [i.e., model interpretation is independent of sampling interval: @noonan_scale-insensitive_2019], the accuracy, size, and stability of speed estimates still depended on the GPS sampling frequency.
-This dependency is because coarsely-sampled movement data contains information on large-scale movements (range crossings, migrations) but not fine-scale movements, including first-order habitat selection [@johnson_comparison_1980].
+This dependency is because coarsely-sampled movement data contains information on large-scale movements (range crossings, migrations) but not fine-scale movements [@nathan_big-data_2022; @noonan_scale-insensitive_2019], including first-order habitat selection [@johnson_comparison_1980].
 Using the boreal caribou as an example, the 13-hour sampling interval allows us to reasonably estimate the caribou's movement path at a temporal scale of approximately 13 hours (or greater), but we cannot produce reasonable movement trajectories at a much finer (e.g., hourly) scale.
-@nathan_big-data_2022 provides additional examples of how the frequency of location data affects the results that can be inferred by modeling movement behaviour.
-Consequently, we suggest being cautious when comparing estimated movement behaviours across species, even though all predictions have been corrected to the hourly timescale by predicting for 1-hour time intervals (i.e., `dt_hours = 1`).
-Ideally, sampling schedules should be fine enough to reconstruct animals' movement movement paths.
-Good estimates of an animal's speed and trajectory require telemetry locations to be taken more often than the animal's directional persistence [@noonan_scale-insensitive_2019], so that, on average, the dataset contains multiple locations in between changes in direction.
+Consequently, we suggest being cautious when comparing estimated movement behaviours across species, even though all predictions have been corrected to the hourly timescale by predicting for 1-hour time intervals.
+Ideally, sampling schedules should be fine enough to reconstruct individuals' movement paths.
+***make data useful for future studies.
+recognize cost of GPS and suggest IoT [@wild_multi-species_2023], although station coverage can be a limitation [@denicola_monitoring_2025].*** Good estimates of an animal's speed and trajectory require telemetry locations to be taken more often than the animal's directional persistence [@noonan_scale-insensitive_2019], so that, on average, the dataset contains multiple locations in between changes in direction.
 What constitutes a change in direction depends on what movement scale one is investigating.
 Small-scale movements and first-order spatial selection will require more frequent sampling than migratory movement or second-order and third-order spatial selection.
-While `ctmm` movement models are scale-invariant in that any model can be scaled to larger or smaller areas and timescales, the model estimates are not independent of the scale of the sampling frequency.
+While `ctmm` movement models are scale-invariant in that any model can be scaled to larger or smaller areas and timescales [@noonan_scale-insensitive_2019], the model estimates are not independent of the scale of the sampling frequency [@denicola_are_2025].
+When tracking data are too coarse for accurate speed estimation, diffusion rates (measured in area per unit time) may be a useful alternative for quantifying animals' exploration rates [@denicola_are_2025].
+***also need better way to distinguish between sates.
+HMMs are an option, but there are limitations.***
 
-When landscapes are relatively predictable across years, a larger number of sampled individuals is likely preferable over the duration of each tracking period.
-The greater number of tracked animals allows one to quantify the variance across individuals, including the range of conditions and environments that individual animals are in.
-A good estimate of inter-individual variance provides better coefficient estimates along with more appropriate measures of uncertainty.
-However, when conditions across years are stochastic, multi-year telemetries allow one to better estimate inter-annual variation without conflating it with inter-individual differences.
-Given the widespread, rapid, and accelerating changes across many habitats (particulartly at high elevations and at high latitudes), we suggest researchers focus on long-term telemetry datasets to quantify the effects of intra-annal variability while increasing the ranges of temperatures each individual is observed in.
-Longer observational periods can also improve the chances of observing different community dynamics, such as shifts in predator-prey dinamics.
+# Other refs
 
-Temperature affects many aspects of mammalian physiology and behaviour, including energetics, sleep, and movement behaviour.
-Climatic changes during the last two centuries have exposed many mammals to novel and increasingly extreme environmental conditions that have led to visible changes that impact not only individuals but also ecological and human communities.
-However, quantifying the effects of climate change is often complex, especially as conditions become increasingly different and extreme events become increasingly common.
-Accurately quantifying the nonlinear effects of temperature on when, how, and where mammals move requires careful data collection and model design.
-We have presented a multi-species analysis of the effects of temperature on mammalian movement rates and habitat selection in hopes that other researchers can leverage this framework and models to expand our understanding of how temperature affects the movement behaviour of other species, including smaller mammals, more elusive species, and non-mammal animals.
-Understanding of how temperature affects each species in a community will allow us to begin making community-level inferences of how temperature will affect intra-specific interactions from a movement behaviour perspective [@nathan_movement_2008; @martinez-garcia_how_2020; @nathan_big-data_2022].
-While adopting such a perspective is surely not simple, it would result in more complete estimates of the effects of climate change on community ecology, particularly when species are exposed to rapid change in many important variables.
-Future work should explore the effects of temperature on movement behaviour while accounting for finer-scale and species specific variables that were not accounted for in this study.
-Examples of these include temporally dynamic measures of forest type and age, canopy density, competitive pressures, forage availability, and predator avoidance, or environmental stochasticity.
-However, many of these data not available, so we also suggest that more work be done on quantifying such measures using spatiotemporally dynamic models and sufficiently fine resolutions.
+* @eum_potential_2023 "Potential changes in climate indices in Alberta under projected global warming of 1.5–5 $^\circ$C"
+* climate change threatens biodiversity of marine mammals: https://doi.org/10.1038/s41598-019-57280-3
+* marine turtles shift and lose habitat with temperature: https://www.science.org/doi/10.1126/sciadv.adw4495
+* @walker_global_2019: Global climate change and invariable photoperiods: {A} mismatch that jeopardizes animal fitness
 

### 2025-09-08 writing/manuscript.Rmd
reviewed discussion; need to finish "future work"

@@
 Furthermore, species that experience narrow temperature ranges (i.e., $`$temperature specialists') tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while species that experience wide ranges of temperature variation (i.e., $`$temperature genralists') can adapt more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024].
-Still, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019; @rantanen_arctic_2022], affect many factors at once [e.g., temperature, snow depth, ecosystem phenology, and resource availability: @polazzo_measuring_2024], or are too rapid and severe to respond within a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].
+Still, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019; @rantanen_arctic_2022], affect many factors at once [e.g., temperature, snow depth, ecosystem phenology, and resource availability: @polazzo_measuring_2024], or are too rapid and severe to respond within a lifetime or a few generations [@levins_evolution_1974; @gienapp_climate_2008; @fuller_towards_2016; @xu_plasticity_2021].
 
@@
 It remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially if populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018].
-Preparing for and responding to future changes require a comprehensive understanding of how temperature will affect how and where mammals move, as well as the implications such changes will have on species' adaptability and overall community structure.
-Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], movement behaviour [@melin_effect_2023], thermoregulation [@mota-rojas_physiological_2021], and trait-based responses [e.g., body size, activity time, and elevational and latitudinal distribution @mccain_assessing_2019].
-However, there remains a need for fine-scale estimates of the effects of temperature on mammals' movement behaviour and quantitative predictions of how their behaviour will respond to climate change in the current century [@ipcc_climate_2023].
+Preparing for and responding to future changes require a comprehensive understanding of how temperature will affect how and where mammals move, as well as the implications such changes will have on species' adaptability and overall community structure [@cunningham_opportunity_2021].
+Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], thermoregulation [@mota-rojas_physiological_2021], and movement behaviour [@mccain_assessing_2019; @melin_effect_2023; @cunningham_opportunity_2021].
+However, there remains a need for fine-scale estimates of the effects of temperature on mammals' movement behaviour and quantitative predictions of changes in movement behaviour with climate change in the current century [@ipcc_climate_2023].
 
@@
 Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.
-Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022;  @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].
+Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022;  @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].
 
@@
 Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the historical data for western Canada from 1998 to 2023 (extremes included) to estimate the monthly variation in temperature, which we defined as $\sigma^2_T$, the variance at location $\langle x, y \rangle$ in month $m$ for each year (e.g., the variance at coordinates $\langle -119.40, 49.94 \rangle$ in April 2005).
-We then modeled the estimated variance using a Generalized Additive Model for Location and Scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017] using the `mgcv` package for `R` [v.
+We then modeled the estimated variance using a Generalized Additive Model (GAM) for Location and Scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017] using the `mgcv` package for `R` [v.
 1.9-1, @wood_generalized_2017].
@@
 B9 and B10).
-Absolute relative changes in distance travelled were small under the best-case SSP (0-4% change in 2100 relative to 2025).
+Absolute relative changes in average yearly distance travelled were small under the best-case SSP (0-4% change in 2100 relative to 2025).
 Under the worst-case SSP, absolute changes by 2100 (relative to 2025) ranged from ~2% (grizzly bears) to ~24% (cougars).
@@
 
-<!-- - address any questions presented in the introduction -->
+<!-- to do: address any questions presented in the introduction -->
 
-\noindent We have demonstrated that temperature is an important determinant of how and where large mammals move, including how they alter seasonal and daily cycles in movement behavior.
-However, predicting mammals' responses to climate change becomes increasingly complex as habitats warm and animals are increasingly frequently exposed to extreme and novel conditions.
+\noindent We have demonstrated that temperature is an important determinant of how and where large mammals move, and that mammals alter their seasonal and daily cycles in movement behavior in response to changes in temperature.
+Predicting mammals' responses to climate change remains a challenge, as habitats warm rapidly and mammals are exposed to increasingly novel and extreme conditions [@ipcc_climate_2023; @diffenbaugh_changes_2013; @botero_evolutionary_2015].
+We leveraged the flexibility and interptetability of GAMs to estimate the (nonlinear) responses  of terrestrial mammals to temperature.
+In particular, tensor product interaction terms were crucial for estimating smooth, nonlinear changes in daily and seasonal trends with temperature.
+By allowing interaction terms to be smooth bivariate functions, we were able to quantify changes in movement behavior without the need to discretize time into distinct windows (e.g., day/night and seasons).
+We were thus able to present changes using relatively simple and interpretable surface plots (Figs.
+\@ref(fig:dist) and \@ref(fig:hrsfs)) that conveyed detailed and nuanced estimates of change.
+This framework was essential for estimating how mammals behave in extreme conditions for which data are scarce.
 
-***Temperature affects many aspects of mammalian physiology and behaviour, including energetics, sleep, and movement behaviour.
-Climatic changes during the last two centuries have exposed many mammals to novel and increasingly extreme environmental conditions that have led to visible changes that impact not only individuals but also ecological and human communities.
-However, quantifying the effects of climate change is often complex, especially as conditions become increasingly different and extreme events become increasingly common.
-Accurately quantifying the nonlinear effects of temperature on how and where mammals move requires careful data collection and model design.
-We have presented a multi-species analysis of the effects of temperature on mammalian movement rates and habitat selection.
-Understanding of how temperature affects each species in a community will allow us to begin making community-level inferences of how temperature will affect intra-specific interactions from a movement behaviour perspective [@nathan_movement_2008; @martinez-garcia_how_2020; @nathan_big-data_2022].
-While adopting such a perspective is surely not simple, it would result in more complete estimates of the effects of climate change on community ecology, particularly when species are exposed to rapid change in many important variables.
-Future work should explore the effects of temperature on movement behaviour while accounting for finer-scale and species specific variables that were not accounted for in this study.
-Examples of these include temporally dynamic measures of forest type and age, canopy density, competitive pressures, forage availability, and predator avoidance, or environmental stochasticity.
-However, many of these data not available, so we also suggest that more work be done on quantifying such measures using spatiotemporally dynamic models and sufficiently fine resolutions.***
+As temperatures warm, mammals' risk of heperthermia increases nonlinearly [@sherwood_adaptability_2010], especially for large-bodied species [@alston_temperature_2020; @dyer_travel_2023; @verzuh_behavioural_2023].
+The recent increase in frequency and duration of extreme-temperature events [@yao_emergence_2022; @ipcc_climate_2023; @zhang_increased_2023; @bunnell_global_2011; @kienzle_has_2018] have forced many mammals towards increasingly atypical behaviours with long-term consequences, ranging from changes in the behavior and fitness of individuals [@brivio_seeking_2024; @cunningham_opportunity_2021] to mass die-offs [@ameca_y_juarez_natural_2012].
+A species' ability to respond to changing temperature and phenology depends mainly on the its exposure, sensitivity, and adaptability to changes, and growing evidence suggests that many mammals in western Canada will struggle to respond due limitations in all three factors [@ameca_y_juarez_natural_2012; @diffenbaugh_changes_2013].
+Firstly, exposure to warming temperatures is widespread throughout the area, and warming is expected to accelerate over the coming decades [@ipcc_climate_2023; @diffenbaugh_changes_2013; @bunnell_global_2011; @kienzle_has_2018; @eum_potential_2023].
+Secondly, large-bodied mammals are particularly sensitive to high heat [@dyer_travel_2023], but small-bodied mammals are also affected [@ratnayake_forecasting_2019; @krebs_impact_2019].
+Finally, species' adaptability is tested by the speed of climate change [@diffenbaugh_changes_2013] and the rigidity of many speices' photoperiod-dependent phenology [@walker_global_2019].
 
-Predicting mammals' responses at the data-scarce fringes of the surface plots in Figs.
-\@ref(fig:dist) and \@ref(fig:hrsfs) comes with substantial uncertainty, particularly given that the responses are nonlinear.
-At warm temperatures, mammals overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behaviour and that of their competitors, predators, and/or prey can often be substantially different from the behaviour at typical temperatures.
-As extreme temperature events become more frequent and prolonged throughout the year [@yao_emergence_2022; @ipcc_climate_2023], mammals will be increasingly forced towards atypical behaviours that may alter community structures and behaviours, both during such events and afterwards [@zhang_increased_2023; @logares_black_2012; @anderson_black-swan_2017].
-For instance, changes in climate and phenology impact the life history and behaviour of many hibernating mammals [@wells_life_2022], and hot weather can also affect mammals' sleep quality and likelihood to enter torpor [@mortlock_sleep_2024; @fjelldal_nightly_2021].
+<!-- complex response from each species -->
 
-Changes in mammal movement behaviour may also alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024].
-At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023], increase their chances of finding food or being preyed on [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021].
-These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011].
+Our results suggest that climate change will cause mammals to alter their seasonal movement behavior, but not all species will respond similarly.
+For example, hotter temperatures caused boreal caribou to travel longer distances in fall, winter, and spring but less in summer, mirroring the findings of @stien_congruent_2012, @leclerc_determinants_2021, and @lessard_local_2025.
+The increase in movement rates during the colder seasons are likely partly due to shallower snow depth that results in greater mobility [@pedersen_quantifying_2021], but rising temperatures during snowy seasons also increase the chances of rain-on-snow events, which limit forage availability and increase the ungulates' need to search for food [@stien_congruent_2012; @berger_climate_2018; @mezzini_how_2025].
+In contrast, the reduced movement rates during warmer summers are likely mainly due to caribou searching for shelter from the heat due to ungulates' elevated risk of hyperthermia [@alston_temperature_2020; @verzuh_behavioural_2023].
+The effects of extreme temperatures are exacerbated by phenological mismatches with seasonal photoperiod [@walker_global_2019], including the timing of molting and reproduction.
+Earlier growth seasons in boreal and arctic regions have resulted in lower calf birth and survival [@post_climate_2008], while the lengthening of the growth season has allowed moose (*Alces alces*) and deer (*Odocoileus* spp.) to encroach on boreal caribou habitat, along with increased density of coyotes (*Canis latrans*), cougars, and wolves [@barber_potential_2018; @demars_incorporating_2023].
+Thus, rising temperatures affect more than each individual species' movement behavior; they have complex and interconnected effects on an entire community and its trophic structure.
 
-***Protecting a third of the world's ecosystems in accordance with the Kunming-Montreal Global Biodiversity Framework [@convention_on_biological_diversity_154_2022] appears less achievable as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023].
-Achieving the "30 by 30" goal will require international collaboration [@huang_achieving_2021] as well as active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023].
-Understanding the consequences of climate change on mammals' movement behaviour is a first step towards proactively responding to how mammals will respond to human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018].***
+<!-- complex response from community -->
 
-## Effects of temperature on movement rates
+Our results suggest that rising temperatures will have similar effects on boreal wolves.
+Warmer temperatures caused wolves to travel further throughout the year, but, similarly to boreal caribou, they moved substantially less at temperatures above 25$^\circ$C in summer.
+At temperatures above 20$^\circ$, both boreal caribou and wolves tended to move less during the day and more between 17:00 and 24:00.
+This synchroniciy in movement patterns is likely to increase encounters between the two species, particularly when coupled with increasing human disturbances, since wolves leverage seismic lines and other anthropogenic linear features (e.g., roads) to reduce movement costs while increasing the chances of encountering prey [@whittington_caribou_2011; @dickie_faster_2017; @dickie_resource_2022].
+Additionally, caribou that attempt to reduce predation risk from wolves by avoiding wolf habitat may risk increasing predation pressure from bears [@leblond_caribou_2016] and other encroaching predators [@demars_incorporating_2023].
+As temperatures rise and allow new competitors and predators to encroach on boreal caribou habitat [@barber_potential_2018; @demars_incorporating_2023; @labadie_global_2023], climate change will have complex conseuqences not only for individual species but for entire communities and their trophic interactions.
 
-\noindent The heterogeneity mammals' responses to temperature suggests that ecological communities will respond to change in complex and interconnected ways.
-Although our models do not account for explicit physiological or phenological changes, the surfaces in Fig.
-\@ref(fig:dist) suggest that warmer temperatures cause many species to alter their daily and seasonal activity patterns (most visible in cougars and grizzly bears).
-For example, when temperatures were above 0$^\circ$C, cougars moved most at night, but when temperatures were below 0$^\circ$C they tended to move more throughout the day.
-Throughout the year, they adapted their tolerance to temperature and moved less when it was relatively hot (for a given time of year), especially in spring and summer.
-The strong reduction in the mid-day movement rates of wolves, elk, cougars, and boreal caribou when summer temperatures were above 20$^\circ$C suggests that the increasingly common and intense heat waves across western Canada will have community-wide impacts on movement rates, encounter rates, and potentially community structure [@martinez-garcia_how_2020].
+<!-- focus on habitat selection using changes in elevational preference of elk, goats, and cougars as the main example -->
+
+Considerations about changes in trophic interactions leads to an important caveat about the results we present.
+The estimated effects of temperature on movement behavior cannot be fully attributed to physiological responses to changes in temperature alone, since other aspects of habitats' seasonal phenology are (nonlinearly) correlated with temperature.
+For example, mountain goats' reduced movement rates at temperatures near 0$^\circ$C in winter may be due to increased movement difficulty, since melting snow and rain-on-snow events affect the energetic costs of movement [@white_population_2025] and encounter rates with predators [@sullender_defining_2023].
+However, milder temperatures can also increase plant growth, which, in turn, allows goats to spend less time searching for forage [@white_mountain_2025].
+Therefore, the figure \@ref(fig:dist) illustrates an estimate of the total effect of temperature on each species movement rates, but it does not explicitly account for energetic costs.
+Similarly, species' changes in habitat selection do not explicitly account for any changes in physiological or energetic costs.
+Both mountain goats and elk selected for higher elevation as temperatures warmed, but our models did not explicitly account for differences in forage availability at different altitudes.
+
+Forage quality and availability is often a limiting factor for mammals' ability to adapt to warming temperatures by moving to higher altitudes.
+Plants are slower to arrive in and establish in new habitats than mammals, and the advance of herbivores can further limit plants' ability to establish [@speed_elevational_2012; @diffenbaugh_changes_2013].
+Consequently, mammals that move to higher elevations may be forced to spend more time foraging [@mezzini_how_2025], which will increase their energetic expenditure as well as their chances of encountering predators (including cougars and grizzly bears) or competitors [@kohl_prey_2019; @martinez-garcia_how_2020; @torrez-herrera_monkeys_2020; @smith_densitydependent_2023].
+Additionally, plants at high elevations may be have weaker defenses against herbivory and be more susceptible to overgrazing [@callis-duehl_community-level_2017].
+The scarcity and susceptibility of forage may thus limit herbivores' ability to shift elevationally, particularly if predation rates remain relatively unchanged or increase.
+Furthermore, the physical upper limit to elevation can cause individuals to become trapped .
+If local peaks become unsuitable habitat, and moving to nearby higher peaks requires traveling across low-quality and high-risk habitat (Figure \@ref(fig:s-rss-cc)), habitat fragmentation may become too excessive to move across patches or alternative ranges [@white_mountain_2025].
+Translocating animals to locations with higher peaks may help avoid such a scenario, but the success of relocations can be highly species-dependent, and attempts are not always successful [@ranc_memory_2022; @white_mountain_2025].
+
+Changes in mammals' seasonal movement behavior are also likely to affect the timing and frequency of human-wildlife interactions.
+Since heat affects mammals' sleep quality and likelihood to enter torpor [@mortlock_sleep_2024; @fjelldal_nightly_2021; @wells_life_2022], future decades may see increased levels of incidents with bears that struggle to enter or remain in hibernation [@johnson_human_2018].
+The risk is further increased when bears' are unable to meet their increased energetic requirements through foraging or hunting.
+Projected increases in park attendance due to rising temperatures and human population density in Western Canada are expected to further increase frequency of human-wildlife interactions [@weststrate_how_2024].
+
+## Consequences for conservation
+
+Species' adaptability to rising temperatures is challenged by the aforementioned symultaneous, multifarious, and interconnected stressors caused by climate change, since responding to many concurrent changes at once is particularly difficult [@diffenbaugh_changes_2013; @polazzo_measuring_2024].
+Locations may thus not be able to support species and population sizes they supported historically, and species may thus not select for historical geograpical ranges in the future.
+It is therefore crucial for conservationists and other decision-makers to distinguish between a species' ideal environmental space [or E space, *sensu* @matthiopoulos_defining_2022], and its best available geographic space [or G space, *sensu* @matthiopoulos_defining_2022].
+Accelerating multufarious change in species' movement behavior and energetic costs require conservation measures to be proactive and anticipate future changes, even when future conditions are substantially different from historical conditions and historical data may not be available.
+To this end, the framework presented by @polazzo_measuring_2024 for estimating species' responses to multifarious change becomes essential, particularly when accounting for energetic and fitness costs, as argued by @cunningham_opportunity_2021.
+
+As of December 2024, only 13.8% of terrestrial Canada was within protected areas or other conserved areas [@environment_and_climate_change_canada_canadian_2025], so rapid action is necessary to protect and additional 16.2% of Canada's land by 2030 in accordance with Kunming-Montreal Global Biodiversity Framework [i.e., the "30 by 30 initiative"$;$ @convention_on_biological_diversity_154_2022].
+However, the choice of what habitats to protect will require careful consideration, since protected areas should support species not only in present conditions but also as species' needs and selection change over the coming decades [@simmons_tough_2021].
+As anthropogenic activity increasingly alters habitats and community structures [@sih_evolution_2011; @tucker_moving_2018; @rosenthal_impact_2022; @rice_conspicuous_2022], protecting biodiversity and the habitats it depends on will require widespread and collaboration with local stakeholders, especially Indigenous Peoples, to ensure that all decisions are culturally sound and respect Treaties [@turner_its_2009; @desjardins_looking_2020; @wong_towards_2020; @hessami_indigenizing_2021; @falardeau_biophysical_2022; @lamb_braiding_2023].
+Understanding the consequences of climate change in mammalian movement behaviour is a first step towards proactively responding to how mammals will respond to human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018].
+However, a better understanding of such consequences will require a more holistic approach that includes carefully data collection, model design, and interpretation and application of results.
+
+## Study limitations and considerations for future work
+
+***HERE***
+
+Estimated selection strength for forest cover was weak and did not change strongly with temperature. Time-varying rasters of forest cover would have likely given stronger (and better) estimates.
+
+* variation among individuals: @jeltsch_need_2025.
+Population-level means that ignore such variation can greatly limit the efficacy of conservation projects [@muff_accounting_2020; @mortlock_sleep_2024; @silva_too_2025].
+
+- we see clear signals despite the data issues and limitations, but more work needs to be done to disentangle important drivers
+
 More work is necessary on quantifying interspecific responses to temperature, including the effects of temperature on predation rates [but see: @cunningham_opportunity_2021; @glass_trade-off_2021; @brivio_seeking_2024].
@@
 
-## Effects of temperature on habitat selection
+Future work should explore the effects of temperature on movement behaviour while accounting for finer-scale and species specific variables that were not accounted for in this study.
+Examples of these include temporally dynamic measures of forest type and age, canopy density, competitive pressures, forage availability, and predator avoidance, or environmental stochasticity.
+However, many of these data not available, so we also suggest that more work be done on quantifying such measures using spatiotemporally dynamic models and sufficiently fine resolutions.
+***...*** animals in the study are generalists; we need more work on specialists and smaller species, but they can be harder to track
 
-***Overall, estimated selection strength was weakest for forest cover and strongest for elevation.
-Elevation is the only raster that doesn't change noticeably over time, so time-varying rasters of the other two would have likely given stronger (and better) estimates.***
+changes in the phenology of plants, fire, ice, or mammalian physiology (e.g., moulting, fat storages) or behaviour (e.g., migration, food caching), other than as direct responses to temperature.
 
-As with the effects of temperature on movement rates, the hererogeneity of temperature effects on species' habitat selection indicates that efficient and proactive habitat conservation will require taxonomic nuance.
-In particular, our HRSFs did not account for any explicit physiological responses to temperature that may impact movement, such as changes in the phenology of plants, fire, ice, or mammalian physiology (e.g., moulting, fat storages) or behaviour (e.g., migration, food caching), other than as direct responses to temperature.
-Consequently, the behavioural changes we present should be interpreted carefully.
-While some seasonal cycles do not depend directly on temperature and may thus not be impacted directly by climate change [such as those driven by photoperiod, see @liu_effects_2022], the predictions we present also rely on the assumption that mammals will not adapt these seasonal cycles to temperature-dependent cues, such as weather [@xu_frost_2019] or forage availability [@middleton_green-wave_2018], which may not be correct.
-Additionally, it is important to stress the distinction between habitat selection and quality.
-While individuals tend to select for high-quality habitats [@kacelnik_ideal_1992; @wright_survival_2024], high selection strength can also be a consequence of an ecological trap [@zuniga-palacios_what_2021; @swearer_overview_2021].
+- other confounding causes of movement (forest age, structure, fire, type, dist_water, etc.) change over the years but are assumed to be fixed
 
-* geographic space *sensu* @matthiopoulos_defining_2022
-* environmental space *sensu* @matthiopoulos_defining_2022
-* G space is the realization of the theoretical E space
+- the temporal resolution of climate change projections isn't great (many steps necessary to simulate weather)
 
-## Predicted changes in movement during the 21^st^ century
+***old text:***
 
-Achieving the goals laid out by the "30 by 30" conservation initiative [@convention_on_biological_diversity_154_2022] will require careful planning but often also rapid action.
-As of December 2024, only 13.8% of terrestrial Canada was within protected areas or other conserved areas [@environment_and_climate_change_canada_canadian_2025].
-Additionally, efficient and effective conservation will require collaboration with Indigenous and local communities to ensure Treaties are respected in the process [@falardeau_biophysical_2022; @lamb_braiding_2023; @wong_towards_2020; @turner_its_2009] and as we actively prepare for future change [@desjardins_looking_2020; @hessami_indigenizing_2021].
-
-***...*** spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas [@jeltsch_need_2025].
-Population-level means that ignore such variation can greatly limit the efficacy of conservation projects [@muff_accounting_2020; @mortlock_sleep_2024; @silva_too_2025].
-Additionally, accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage and prey availability [@mezzini_how_2025], encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@smith_densitydependent_2023], competitive pressure [@torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019].
-This complexity results in the great variation among the responses of different species (and individuals), especially as populations undergo a rapid cascade of change [@botero_evolutionary_2015] and mammals face increasing pressures from human activity, including habitat fragmentation, habitat loss, and greater human presence in wild spaces [@sih_evolution_2011; @tucker_moving_2018; @sawyer_identifying_2009; @weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022].
-As selection strength for current ranges changes (and likely decreases), some individuals may disperse, relocate to a new habitat, or remain within the current range despite the reduced fitness and increased risk of extirpation or extiction [@logares_black_2012; @anderson_black-swan_2017; @duncan_predicting_2012].
-
-## Species-specific considerations
-
-* Cougars: attraction for disconnected patches of high elevation could lead to fragmentation or traps
-
-* boreal caribou: Increasing temperatures increased RSS for both elevations near ~500 m as well as > 1,000 m, but such a selection is only possible if: (1) such habitats exist in the individuals' range, and (2) crossing the 500 m to > 1,000 m regions is safe and attractive
-
-* sm caribou: HRSFs do not account for migratory behaviour due to reproduction or predation risk
-
-## Study limitation and considerations for future studies
-
-Our work highlights three central considerations for future research: (1) telemetry sampling rates should be designed primarily in relation to the movement timescales of the species of interest [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @pease_ecological_2024], (2) the number of individuals tracked and the duration of each telemetry should depend not just on population size but also the rate of environmental change, and (3) predicting changes in movement behaviour becomes highly complex when responses are nonlinear, especially when changes depend on many interacting factors [@polazzo_measuring_2024] and one is interested in predicting responses in extreme conditions for which data are scarce [@steixner-kumar_strategies_2020].
+Our work highlights three central considerations for future research: (1) telemetry sampling rates should be designed primarily in relation to the movement timescales of the species of interest [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @pease_ecological_2024], (2) the number of individuals tracked and the duration of each telemetry should depend not just on population size but also the (accelerating) rate of environmental change, and (3) predicting changes in movement behaviour becomes highly complex when responses are nonlinear, especially when changes depend on many interacting factors [@polazzo_measuring_2024] and one is interested in predicting responses in extreme conditions for which data are scarce [@steixner-kumar_strategies_2020].
 
@@
 
-# Other refs
+# Other refs and miscellaneous maerial {-}
 
-* @eum_potential_2023 "Potential changes in climate indices in Alberta under projected global warming of 1.5–5 $^\circ$C"
-* climate change threatens biodiversity of marine mammals: https://doi.org/10.1038/s41598-019-57280-3
-* marine turtles shift and lose habitat with temperature: https://www.science.org/doi/10.1126/sciadv.adw4495
-* @walker_global_2019: Global climate change and invariable photoperiods: {A} mismatch that jeopardizes animal fitness
+GAMs' flexibility and complexity can also be a frustrating limitation if used too generously.
+As such, analysts should chose the basis size for each smooth term carefully.
+While penalized splines and REML can help avoid unruly smooth terms that would result in questionable estimates where data are scarce [@wood_generalized_2017], it is best to select the basis size of the smooth terms with parsimony, especially when datasets are large.
+Still, predicting mammals' responses at the data-scarce fringes of the surface plots in Figs.
+\@ref(fig:dist) and \@ref(fig:hrsfs) comes with greater uncertainty, particularly given that the changes are nonlinear.
 
@@
 
-SM performed the data cleaning, ran the analyses, and wrote the majority of the manuscript.
+SM performed the data cleaning, ran the analyses, and wrote the manuscript.
 CHF wrote code for estimating instantaneous movement speeds.
 MJN conceived of the project idea and supervised SM throughout the project.
-All other authors contributed telemetry data and contributed to the interpretation of the results for their species of interest.
+All other authors, which are listed in alphabetical order by last name, contributed telemetry data and reviewed the interpretation of the results for their species of interest.
 All authors contributed to reviewing the manuscript.

### 2025-09-09 writing/appendix-b-gams.Rmd
changed scientific names to common names

@@
 ---
-title: "\\Large The heat is on: Rising temperatures alter when, how, and where large terrestrial mammals move"
+title: "\\Large The heat is on: Rising temperatures alter how and where mammals move"
 subtitle: "\\Large Appendix B: Modeling"
@@
 
-```{r pred-vs-pred-hgams, fig.cap="Hexplot of the fitted values from the HGAMs with and without including temperature for the probability of movement (\\textbf{A}) and speed when moving (\\textbf{B})."}
+```{r pred-vs-pred-hgams, fig.cap="Hexplot of the fitted values from the HGAMs with and without including temperature for the probability of moving (\\textbf{A}) and speed when moving (\\textbf{B})."}
 knitr::include_graphics('../figures/hgam-with-without-temp-prediction-agreement.png')
@@
 
-```{r op, fig.cap="Relationsips between the observed and predicted values for probability of movement (\\textbf{A}) and speed given that an animal was moving (\\textbf{B}).
+```{r op, fig.cap="Relationsips between the observed and predicted values for probability of moving (\\textbf{A}) and speed given that an animal was moving (\\textbf{B}).
 The color in panel \\textbf{A} indicates the number of points used to calculate the estimate (on the $\\log_{2}$ scale), while the grey lines in both panels indicate perfect prediction (i.e., the 1:1 line)."}
@@
 
-knitr::kable(temp_comp, digits = 1, format = 'markdown', caption = "Changes in deviance expained ($\\Delta$DE) and Akaike Information Criterion ($\\Delta$AIC) from including temperature in the models estimating probability of movement ($P(M)$) and speed when moving (Speed) as well as in species-level Hierarchical Resource Selection Functions (HRSF).
+knitr::kable(temp_comp, digits = 1, format = 'markdown', caption = "Changes in deviance expained ($\\Delta$DE) and Akaike Information Criterion ($\\Delta$AIC) from including temperature in the models estimating probability of moving ($P(M = 1)$) and speed when moving (Speed) as well as in species-level Hierarchical Resource Selection Functions (HRSF).
 Unlike conventional marginal AIC [@akaike_new_1974], The AIC values provided by the `mgcv` package for `R` are calculated using the models' estimated effective degrees of freedom and the Bayesian variance-covariance matrices, which avoids issues related to random effects or Restricted Maximum Likelihood (REML).")

### 2025-09-09 analysis/06-temperature-hgams.R

@@ -66,11 +66,16 @@ if(file.exists('data/hgam-speed-data.rds')) {
   saveRDS(d, 'data/hgam-speed-data.rds')
 }
 
+# add common names
+d <- d %>%
+  mutate(name = COMMON_NAMES[map_int(species,
+                                     \(.s) which(sort(SPECIES) == .s))])
+
 #' `kmeans()` splits data poorly
 splits <-
   tibble(species = SPECIES,
-         lab = SPECIES_LABS,
-         manual = c(0.06, 0.05, 0.08, 0.07, 0.10, 0.20, 0.05),
+         lab = COMMON_NAMES,
+         manual = c(0.20, 0.07, 0.05, 0.08, 0.10, 0.06, 0.05),
          `k means` = map_dbl(species, \(sp) {
            km <- kmeans(filter(d, species == sp)$speed_est, centers = 2)
            max(min(filter(d, species == sp) %>%
@@ -83,7 +88,8 @@ splits <-
   ) %>%
   pivot_longer(c(manual, `k means`), values_to = 'split_m_s',
                names_to = 'method') %>%
-  mutate(split_km_h = 'km/h' %#% split_m_s)
+  mutate(split_km_h = 'km/h' %#% split_m_s) %>%
+  arrange(lab)
 
 hists <- list()
 for(i in (1:N_SPECIES)[order(SPECIES)]) {
@@ -107,14 +113,14 @@ for(i in (1:N_SPECIES)[order(SPECIES)]) {
                    lty = method), filter(splits, species == sp), lwd = 0.75) +
     coord_cartesian(xlim = XLIM, ylim = YLIM) +
     labs(x = 'Speed (km/h)', y = 'Count',
-         title = scales::parse_format()(SPECIES_LABS[i])) +
+         title = COMMON_NAMES[i]) +
     scale_color_bright(name = 'Method') +
     scale_linetype_manual(name = 'Method', values = c(2, 1)) +
     theme(legend.position = 'none')
   rm(sp, XLIM, YLIM)
 }
 
-hists[1:7] <- hists[order(SPECIES)]
+hists[1:7] <- hists[order(COMMON_NAMES)]
 hists[[8]] <-
   get_legend(
     ggplot() +
@@ -144,6 +150,15 @@ d %>%
   summarise(prop_moving = mean(moving),
             min_speed = min(speed_est[moving]))
 
+d <- mutate(d,
+            lab = case_when(species == SPECIES[1] ~ sort(SPECIES_LABS)[1],
[…]

### 2025-09-09 analysis/06-temperature-hrsfs.R

@@ -448,7 +448,8 @@ fits <- tibble(
     gsub('\\(s\\.', 'southern', .) %>%
     gsub('\\(', '', .) %>%
     gsub('\\)', '', .),
-  lab = SPECIES_LABS,
+  lab = sort(SPECIES_LABS),
+  name = COMMON_NAMES,
   w = map_chr(species, \(.sp) {
     list.files(path = 'models',
                pattern = paste0('rsf-', .sp, '-2025-'),
@@ -476,7 +477,7 @@ p_fits <-
   mutate(fits_wo = if_else(fits_wo < -15, -15, fits_wo),
          fits_w = if_else(fits_w < -15, -15, fits_w)) %>%
   ggplot(aes(fits_wo, fits_w)) +
-  facet_wrap(~ lab, labeller = label_parsed, scales = 'free', nrow = 2) +
+  facet_wrap(~ name, scales = 'free', nrow = 2) +
   geom_hex(aes(fill = log10(after_stat(count)))) +
   geom_abline(slope = 1, intercept = 0, color = 'black') +
   labs(x = 'log(RSS) without temperature',

### 2025-09-09 analysis/figures/default-ggplot-theme.R

@@ -14,3 +14,11 @@ SPECIES_LABS <- gsub(' ', '~', SPECIES) %>%
   gsub(')$', ')"', .) %>%
   paste0('bolditalic(', ., ')')
 N_SPECIES <- length(SPECIES)
+
+COMMON_NAMES <- c('Wolves', 'Elk', 'Mountain goats', 'Cougars',
+                  'Caribou (boreal)', 'Caribou (southern mountain)',
+                  'Grizzly bears')
+
+PAL <- PAL[order(COMMON_NAMES)] # rearrange based on order of common names
+SPECIES <- SPECIES[order(SPECIES)]
+COMMON_NAMES <- factor(COMMON_NAMES) # keep the vector in the species order

### 2025-09-09 analysis/figures/hrsfs-available-vs-used.R

@@ -16,23 +16,19 @@ d <- readRDS('data/tracking-data/rsf-data.rds') %>%
   pivot_longer(c(forest_perc, elevation_m, dist_water_m),
                names_to = 'variable') %>%
   mutate(
-    lab = case_when(species == 'Rangifer tarandus southern mountain' ~ SPECIES_LABS[1],
-                    species == 'Ursus arctos horribilis' ~ SPECIES_LABS[2],
-                    species == 'Puma concolor' ~ SPECIES_LABS[3],
-                    species == 'Cervus canadensis' ~ SPECIES_LABS[4],
-                    species == 'Rangifer tarandus boreal' ~ SPECIES_LABS[5],
-                    species == 'Canis lupus' ~ SPECIES_LABS[6],
-                    species == 'Oreamnos americanus' ~ SPECIES_LABS[7]),
+    species = gsub('southern mountain', '(s. mountain)', species) %>%
+      gsub('boreal', '(boreal)', .),
+    lab = COMMON_NAMES[map_int(species,
+                               \(.s) which(SPECIES == .s))],
     variable =
-      case_when(variable == 'forest_perc' ~ "bold(Forest~cover~'(%)')",
-                variable == 'elevation_m' ~ "bold(Elevation~'(km)')",
-                variable == 'dist_water_m' ~ "bold(Distance~from~water~'(km)')") %>%
+      case_when(variable == 'forest_perc' ~ 'Forest cover (%)',
+                variable == 'elevation_m' ~ 'Elevation (km)',
+                variable == 'dist_water_m' ~ 'Distance from water (km)') %>%
       factor(., levels = unique(.)))
 
 a_u <-
   ggplot(d) +
-  facet_grid(lab ~ variable, scales = 'free', labeller = label_parsed,
-             switch = 'x') +
+  facet_grid(lab ~ variable, scales = 'free', switch = 'x') +
   geom_histogram(aes(value, fill = factor(detected)),
                  position = 'identity', alpha = 0.5, bins = 30) +
   scale_fill_brewer(NULL, type = 'qual', palette = 6,
@@ -47,8 +43,7 @@ ggsave('figures/available-vs-used.png',
 
 a_u_w <-
   ggplot(d) +
-  facet_grid(lab ~ variable, scales = 'free', labeller = label_parsed,
-             switch = 'x') +
+  facet_grid(lab ~ variable, scales = 'free', switch = 'x') +
[…]

### 2025-09-09 analysis/figures/telemetry-ud-map.R

@@ -11,20 +11,14 @@ source('data/bc-shapefile.R')
 
 d <- readRDS('models/movement-models-akdes-2024-06-06.rds') %>%
   mutate(dataset_name = case_when(
-    dataset_name == 'Canis_lupus_boreal' ~ 'Canis lupus',
-    dataset_name == 'Rangifer_tarandus_boreal' ~ 'Rangifer tarandus (boreal)',
-    dataset_name == 'Rangifer_tarandus_southern_mountain' ~ 'Rangifer tarandus (southern mountain)',
-    dataset_name == 'Puma_concolor_2' ~ 'Puma concolor',
-    dataset_name == 'Puma_concolor_4' ~ 'Puma concolor',
-    dataset_name == 'Elk in southwestern Alberta' ~ 'Cervus canadensis',
-    dataset_name == 'Oreamnos_americanus' ~ 'Oreamnos americanus',
-    dataset_name == 'Ursus_arctos_horribilis' ~ 'Ursus arctos horribilis')) %>%
-  mutate(dataset_name = gsub(' ', '~', dataset_name) %>%
-           gsub('southern~', 'southern ', .) %>%
-           gsub('~\\(', '\\)~bold\\("(', .),
-         dataset_name = if_else(grepl('bold\\(', dataset_name),
-                                paste0('bolditalic(', dataset_name, '")'),
-                                paste0('bolditalic(', dataset_name, ')')))
+    dataset_name == 'Canis_lupus_boreal' ~ 'Wolves',
+    dataset_name == 'Rangifer_tarandus_boreal' ~ 'Caribou (boreal)',
+    dataset_name == 'Rangifer_tarandus_southern_mountain' ~ 'Caribou (southern mountain)',
+    dataset_name == 'Puma_concolor_2' ~ 'Cougars',
+    dataset_name == 'Puma_concolor_4' ~ 'Cougars',
+    dataset_name == 'Elk in southwestern Alberta' ~ 'Elk',
+    dataset_name == 'Oreamnos_americanus' ~ 'Mountain goats',
+    dataset_name == 'Ursus_arctos_horribilis' ~ 'Grizzly bears'))
 
 tels <- transmute(d,
                   dataset_name,
@@ -79,17 +73,18 @@ dem <- mutate(dem, elev_m = if_else(elev_m < 0, 0, elev_m))
 p_dem <-
   ggplot() +
   geom_raster(aes(x, y, fill = elev_m), dem) +
[…]

### 2025-09-09 analysis/figures/temperature-hgams-spatial-projections.R

@@ -76,27 +76,31 @@ if(file.exists('data/cc-hgam-projections-local-2100.rds')) {
   beepr::beep(2)
 }
 
+# add common species names
+cc_proj <- cc_proj %>%
+  mutate(lab = COMMON_NAMES[map_int(species, \(.s) {
+    which(sort(SPECIES_LABS) == .s)
+  })],
+  scenario = gsub('"', '', scenario) %>%
+    factor(., levels = unique(.)))
+
 # figure of estimated speeds for each species ----
 range(c(range(cc_proj$p), range(cc_proj$s), range(cc_proj$d)))
 
 z_breaks <- seq(log2(0.75), -log2(0.75), length.out = 5)
 
-make_plot <- function(sp, y_facets = FALSE, get_legend = FALSE,
+make_plot <- function(.lab, y_facets = FALSE, get_legend = FALSE,
                       reproject = TRUE, variable) {
+  sp_data <- filter(cc_proj, lab == .lab)
+  
   if(variable == 'p') {
-    sp_data <- cc_proj %>%
-      filter(species == sp) %>%
-      select(scenario, species, long, lat, p)
+    sp_data <- sp_data %>% select(scenario, lab, long, lat, p)
     z_lab <- 'probability of moving'
   } else if(variable == 's') {
-    sp_data <- cc_proj %>%
-      filter(species == sp) %>%
-      select(scenario, species, long, lat, s)
+    sp_data <- sp_data %>% select(scenario, lab, long, lat, s)
     z_lab <- 'speed when moving'
   } else if(variable == 'd') {
-    sp_data <- cc_proj %>%
-      filter(species == sp) %>%
-      select(scenario, species, long, lat, d)
+    sp_data <- sp_data %>% select(scenario, lab, long, lat, d)
     z_lab <- 'distance travelled'
   } else stop('Please choose a variable among p, s, or d.')
   
@@ -105,7 +109,7 @@ make_plot <- function(sp, y_facets = FALSE, get_legend = FALSE,
   
   if(reproject) {
     sp_data <- sp_data %>%
-      nest(rast = ! c(scenario, species)) %>%
+      nest(rast = ! c(scenario, lab)) %>%
       mutate(rast = map(rast, function(.r) {
         rast(.r) %>%
           `crs<-`('EPSG:4326') %>%
[…]

### 2025-09-09 analysis/figures/temperature-hgams-temporal-projections.R

@@ -96,14 +96,20 @@ if(file.exists('data/cc-hgam-projections.rds')) {
   beepr::beep(2)
 }
 
+# add common species names
+cc_proj <- cc_proj %>%
+  mutate(lab = COMMON_NAMES[map_int(species, \(.s) {
+    which(sort(SPECIES_LABS) == .s)
+  })],
+  scenario = gsub('"', '', scenario) %>%
+    factor(., levels = unique(.)))
+
 # make figures ----
 p_p_mov <-
   ggplot(cc_proj, aes(year, p_median, group = scenario)) +
-  facet_wrap(~ species, scales = 'fixed', drop = FALSE,
-             labeller = label_parsed, nrow = 2) +
+  facet_wrap(~ lab, scales = 'fixed', drop = FALSE, nrow = 2) +
   geom_hline(aes(yintercept = 1), lty = 'dashed') +
-  geom_ribbon(aes(ymin = p_lwr_05, ymax = p_upr_95,
-                  fill = factor(scenario, levels = rev(levels(scenario))),
+  geom_ribbon(aes(ymin = p_lwr_05, ymax = p_upr_95, fill = scenario,
                   color = scenario), linewidth = 0.2, alpha = 0.25) +
   geom_line(color = 'black', lwd = 1.5) +
   geom_line(aes(color = scenario), lwd = 1) +
@@ -119,10 +125,8 @@ ggsave('figures/p-moving-local-cc-predictions.png', p_p_mov,
 
 p_s <-
   ggplot(cc_proj, aes(year, s_median, group = scenario)) +
-  facet_wrap(~ species, scales = 'fixed', drop = FALSE,
-             labeller = label_parsed, nrow = 2) +
-  geom_ribbon(aes(ymin = s_lwr_05, ymax = s_upr_95,
-                  fill = factor(scenario, levels = rev(levels(scenario))),
+  facet_wrap(~ lab, scales = 'fixed', drop = FALSE, nrow = 2) +
+  geom_ribbon(aes(ymin = s_lwr_05, ymax = s_upr_95, fill = scenario,
                   color = scenario), linewidth = 0.2, alpha = 0.25) +
   geom_hline(aes(yintercept = 1), lty = 'dashed') +
   geom_line(color = 'black', lwd = 1.5) +
@@ -139,10 +143,8 @@ ggsave('figures/speed-local-cc-predictions.png', p_s,
 
 p_d <-
[…]

### 2025-09-09 analysis/figures/temperature-hgams.R

@@ -68,16 +68,15 @@ marginal <- function(newd, term) {
   
   # add columns of distance travelled
   bind_cols(newd, preds_1, preds_2) %>%
-    mutate(species = gsub(' ', '~', species),
-           species = gsub('~\\(', '\\)~bold\\((', species),
-           species = paste0('bolditalic(', species, ')'),
-           d_mu = p_mu * s_mu,
+    mutate(d_mu = p_mu * s_mu,
            d_lwr = p_lwr * s_lwr,
            d_upr = p_upr * s_upr) %>%
     group_by(species) %>%
     mutate(across(c(s_mu, s_lwr, s_upr), \(.col) .col / mean(s_mu)),
            across(c(d_mu, d_lwr, d_upr), \(.col) .col / mean(d_mu))) %>%
     ungroup() %>%
+    mutate(lab = COMMON_NAMES[map_int(species,
+                                      \(.s) which(SPECIES == .s))]) %>%
     return()
 }
 
@@ -104,16 +103,16 @@ p_mov_tod <-
   mutate(p_upr = if_else(p_upr > 0.75, 0.75, p_upr)) %>%
   ggplot() +
   coord_polar() +
-  facet_wrap(~ lab, nrow = 1, labeller = label_parsed) +
+  facet_wrap(~ lab, nrow = 1) +
   geom_area(aes(x, 0.75, group = g), tod, fill = 'black', alpha = 0.2) +
-  geom_line(aes(tod_pdt, p_mu, color = species), linewidth = 1) +
-  geom_ribbon(aes(tod_pdt, ymin = p_lwr, ymax = p_upr, fill = species),
+  geom_line(aes(tod_pdt, p_mu, color = lab), linewidth = 1) +
+  geom_ribbon(aes(tod_pdt, ymin = p_lwr, ymax = p_upr, fill = lab),
               alpha = 0.2) +
   scale_color_manual('Species', values = PAL,
                      aesthetics = c('color', 'fill')) +
   scale_x_continuous(expand = c(0, 0), limits = c(0, 24),
                      breaks = c(0, 6, 12, 18),
-                     labels = c('00:00', '06:00', '12:00', '18:00')) +
+                     labels = c('00:00', '06:00  ', '12:00', '  18:00')) +
   scale_y_continuous(limits = c(0, 0.75), expand = c(0, 0),
[…]

### 2025-09-09 analysis/figures/temperature-hrsfs-spatial-projections.R

@@ -109,18 +109,28 @@ if(file.exists('data/cc-hrsf-projections-local-2100.rds')) {
   beepr::beep(2)
 }
 
+cc_proj <- cc_proj %>%
+  mutate(name = COMMON_NAMES[map_int(lab, \(.lab) {
+    which(sort(SPECIES_LABS) == .lab)
+  })],
+  scenario = gsub('"', '', scenario) %>%
+    factor(., levels = unique(.)))
+
+slice(cc_proj, 1, .by = lab) %>%
+  select(lab, name)
+
 # figure of estimated speeds for each species ----
 z_breaks <- seq(-log2(1.25), log2(1.25), length.out = 5)
 
-make_plot <- function(sp, y_facets = FALSE, get_legend = FALSE,
+make_plot <- function(sn, y_facets = FALSE, get_legend = FALSE,
                       reproject = TRUE) {
   sp_data <- cc_proj %>%
-    filter(lab == sp) %>%
-    select(scenario, lab, long, lat, l)
+    filter(name == sn) %>%
+    select(scenario, name, long, lat, l)
   
   if(reproject) {
     sp_data <- sp_data %>%
-      nest(rast = ! c(scenario, lab)) %>%
+      nest(rast = ! c(scenario, name)) %>%
       mutate(rast = map(rast, function(.r) {
         rast(.r) %>%
           `crs<-`('EPSG:4326') %>%
@@ -139,7 +149,7 @@ make_plot <- function(sp, y_facets = FALSE, get_legend = FALSE,
                          TRUE ~ l)) %>%
     ggplot() +
     coord_sf(crs = 'EPSG:3005') +
-    facet_grid(scenario ~ lab, labeller = label_parsed) +
+    facet_grid(scenario ~ name) +
     geom_raster(aes(long, lat, fill = log2(l))) +
     scale_fill_distiller(name = 'Pixel-level relative change in selection strength',
                          palette = 'PiYG', limits = range(z_breaks),
@@ -166,13 +176,12 @@ make_plot <- function(sp, y_facets = FALSE, get_legend = FALSE,
   return(p)
 }
 
-plot_list <- map(sort(as.character(SPECIES_LABS)), make_plot)
-plot_list[[7]] <- make_plot('bolditalic(Ursus~arctos~horribilis)',
-                            y_facets = TRUE) # add vertical facet labels
+plot_list <- map(sort(COMMON_NAMES), make_plot)
[…]

### 2025-09-09 analysis/figures/temperature-hrsfs-temporal-projections.R

@@ -123,11 +123,17 @@ if(file.exists('data/cc-hrsf-projections.rds')) {
   saveRDS(cc_proj, 'data/cc-hrsf-projections.rds')
 }
 
+# add common species names
+cc_proj <- cc_proj %>%
+  mutate(lab = COMMON_NAMES[map_int(lab, \(.s) {
+    which(sort(SPECIES_LABS) == .s)
+  })])
+
 # figures of habitat selection strength ----
 cc_proj %>%
   filter(year >= 2025) %>% # 2025 is the reference year
   ggplot(aes(x = year, group = scenario)) +
-  facet_wrap(~ lab, scales = 'fixed', labeller = label_parsed, nrow = 2) +
+  facet_wrap(~ lab, scales = 'fixed', nrow = 2) +
   geom_hline(yintercept = 1, color = 'black', lty = 'dashed') +
   geom_line(aes(y = l_median, color = scenario), lwd = 0) +
   geom_ribbon(aes(ymin = l_05, ymax = l_95,
@@ -148,12 +154,12 @@ cc_proj %>%
   filter(year >= 2025) %>% # 2025 is the reference year
   pivot_longer(l_05:l_95, values_to = 'l', names_to = 'percentile',
                names_prefix = 'l_') %>%
-  mutate(percentile = case_when(percentile == '05' ~ '"Bottom 5% RSS"',
-                                percentile == 'median' ~ '"Median RSS"',
-                                percentile == '95' ~ '"Top 5% RSS"') %>%
+  mutate(percentile = case_when(percentile == '05' ~ 'Bottom 5% RSS',
+                                percentile == 'median' ~ 'Median RSS',
+                                percentile == '95' ~ 'Top 5% RSS') %>%
            factor(., levels = rev(sort(unique(.))))) %>%
   ggplot() +
-  facet_grid(percentile ~ lab, scales = 'fixed', labeller = label_parsed) +
+  facet_grid(percentile ~ lab, scales = 'fixed') +
   geom_line(aes(year, l, color = scenario), lwd = 0.5) +
   geom_hline(yintercept = 1, color = 'black', lty = 'dashed') +
   scale_x_continuous(NULL, breaks = c(2025, 2050, 2075, 2100)) +

### 2025-09-09 analysis/figures/temperature-hrsfs.R

@@ -138,11 +138,27 @@ if(file.exists('models/temperature-hrsf-preds.rds')) {
   saveRDS(preds, 'models/temperature-hrsf-preds.rds')
 }
 
+# add common species names
+preds <- preds %>%
+  mutate(lab = COMMON_NAMES[map_int(species, \(.s) {
+    which(sort(SPECIES) == .s)
+  })],
+  variable = variable %>%
+    gsub('bold\\(', '', .) %>%
+    gsub('~', ' ', .) %>%
+    gsub('"', '', .) %>%
+    gsub('\\)\\)', '\\)', .) %>%
+    factor(levels = c('Forest cover (%)',
+                      'Elevation (km)',
+                      'Distance from water (km)')))
+
+slice(preds, 1, .by = lab)
+
 # plot of RSS ----
 # need to divide by median lambda within each variable to get interpretable results
 preds %>%
   filter(! too_far_1) %>%
-  group_by(species, variable) %>%
+  group_by(lab, variable) %>%
   summarize(min_lambda = min(lambda),
             median_lambda = median(lambda),
             max_lambda = max(lambda))
@@ -152,22 +168,22 @@ LIM <- 2
 p <-
   preds %>%
   filter((! too_far_2)) %>%
-  select(species, lab, x, temperature_C, variable, lambda, too_far_1) %>%
+  select(lab, x, temperature_C, variable, lambda, too_far_1) %>%
   # make detection rates homogeneous across temperature
-  group_by(species, variable, temperature_C) %>%
+  group_by(lab, variable, temperature_C) %>%
   mutate(lambda = lambda / mean(lambda)) %>%
   # re-scale the center (lambda = 1) relative to the median in the data.
   # doing this because some estimated effects of distance from water and
   # elevation are extreme
-  group_by(species, variable) %>%
+  group_by(lab, variable) %>%
   mutate(lambda = lambda / median(lambda),
          lambda = case_when(
-           species == 'Cervus canadensis' & variable == 'bold(Elevation~"(km)")' ~ lambda / exp(1),
-           species == 'Oreamnos americanus' & variable == 'bold(Elevation~"(km)")' ~ lambda * exp(0.25),
-           species == 'Rangifer tarandus (boreal)' & variable == 'bold(Elevation~"(km)")' ~ lambda / exp(2),
[…]

### 2025-09-09 writing/appendix-a-cleaning-telemetry-data.Rmd
updated title

@@
 ---
-title: "\\Large The heat is on: Rising temperatures alter when, how, and where large terrestrial mammals move"
+title: "\\Large The heat is on: Rising temperatures alter how and where mammals move"
 subtitle: "\\Large Appendix A: Cleaning Telemetry Data"

### 2025-09-09 writing/manuscript.Rmd
updated based on some comments from SD and KH

@@
     - \usepackage{caption} # for more customization of captions
-    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
+    - \captionsetup[figure]{font=footnotesize, aboveskip=4pt, belowskip=-15pt}
     - \usepackage{hanging} # for hanging indents in references
@@
 ```
+
+# Alternative titles {-}
+
+* Rising temperatures alter the movement phenology of terrestrial mammals
+* Terrestrial mammals alter their movement behavior to rising temperatures
+
+\newpage
 
@@
 Due to mammals' difficulty at dissipating heat, particularly when ambient temperatures are near or above their body temperature, they are particularly sensitive to extreme heat, especially with increasing body size [@dyer_travel_2023; @verzuh_behavioural_2023].
-Furthermore, species that experience narrow temperature ranges (i.e., $`$temperature specialists') tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while species that experience wide ranges of temperature variation (i.e., $`$temperature genralists') can adapt more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024].
-Still, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019; @rantanen_arctic_2022], affect many factors at once [e.g., temperature, snow depth, ecosystem phenology, and resource availability: @polazzo_measuring_2024], or are too rapid and severe to respond within a lifetime or a few generations [@levins_evolution_1974; @gienapp_climate_2008; @fuller_towards_2016; @xu_plasticity_2021].
+Furthermore, species that experience narrow temperature ranges (i.e., $`$temperature specialists') tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while species that experience wide ranges of temperature variation (i.e., $`$temperature generalists') respond more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024].
+Still, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019; @rantanen_arctic_2022], affect many factors at once [e.g., temperature, snow depth, ecosystem phenology, and resource availability: @polazzo_measuring_2024], or are too rapid and severe to respond to within a lifetime or a few generations [@levins_evolution_1974; @gienapp_climate_2008; @fuller_towards_2016; @xu_plasticity_2021].
 
@@
 
-In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (wolves: *Canis lupus*, elk: *Cervus canadensis*, mountain goats: *Oreamnos americanus*, cougars: *Puma concolor*, boreal and southern mountain caribou: *Rangifer tarandus*, and grizzly bears: *Ursus arctos horribilis*; Table \@ref(tab:data-table)).
+In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou: *Rangifer tarandus*, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves: *Canis lupus*; Table \@ref(tab:data-table)).
 Using over 25 years of telemetry data throughout a large spatial range of western Canada and hourly air temperature data, we estimate how these mammals altered their probability of moving, speed when moving, and habitat selection in response to temperature.
@@
 ```{r data-table}
+setwd('..') # go to main directory
+source('analysis/figures/default-ggplot-theme.R')
+
 dt_data <-
-  readRDS('../models/movement-models-2024-06-06.rds') %>%
+  readRDS('models/movement-models-2024-06-06.rds') %>%
   transmute(
     Species = case_when(
-      dataset_name == 'Canis_lupus_boreal' ~ '\\emph{Canis lupus}',
-      dataset_name == 'Rangifer_tarandus_boreal' ~ '\\emph{Rangifer tarandus} (boreal)',
-      dataset_name == 'Rangifer_tarandus_southern_mountain' ~ '\\emph{Rangifer tarandus} (s. mountain)',
-      dataset_name == 'Puma_concolor_2' ~ '\\emph{Puma concolor}',
-      dataset_name == 'Puma_concolor_4' ~ '\\emph{Puma concolor}',
-      dataset_name == 'Elk in southwestern Alberta' ~ '\\emph{Cervus canadensis}',
-      dataset_name == 'Oreamnos_americanus' ~ '\\emph{Oreamnos americanus}',
-      dataset_name == 'Ursus_arctos_horribilis' ~ '\\emph{Ursus arctos horribilis}'),
+      dataset_name == 'Canis_lupus_boreal' ~ COMMON_NAMES[1],
+      dataset_name == 'Rangifer_tarandus_boreal' ~ COMMON_NAMES[2],
+      dataset_name == 'Rangifer_tarandus_southern_mountain' ~ COMMON_NAMES[3],
+      dataset_name == 'Puma_concolor_2' ~ COMMON_NAMES[4],
+      dataset_name == 'Puma_concolor_4' ~ COMMON_NAMES[4],
+      dataset_name == 'Elk in southwestern Alberta' ~ COMMON_NAMES[5],
+      dataset_name == 'Oreamnos_americanus' ~ COMMON_NAMES[6],
+      dataset_name == 'Ursus_arctos_horribilis' ~ COMMON_NAMES[7]),
     animal, # keep animal column for grouping later
@@
             End = max(as.Date(timestamp)),
+            n = n(),
             dt = median(timestamp - lag(timestamp), na.rm = TRUE),
@@
             End = max(End),
+            n = sum(n),
             dt = as.numeric(median(dt, na.rm = TRUE)),
@@
 
-knitr::kable(dt_data, format = 'markdown', caption = "Summary statistics of each species' set of GPS data, including: the start and end of each set of GPS telemetries, the median sampling interval ($\\Delta t$; stratified by animal), number of animals (Animals), and number of animals with finite speed estimates (Has speed).", escape = FALSE, longtable = FALSE, booktabs = TRUE)
+knitr::kable(dt_data, format = 'markdown', format.args = list(big.mark = ','), caption = "Summary statistics of each species' set of GPS data, namely: the start and end of each set of GPS telemetries, the number of GPS fixes (after data cleaning), the number the median sampling interval ($\\Delta t$; stratified by animal), the number of animals (Animals), and the number of animals with finite speed estimates (Has speed).", escape = FALSE, longtable = FALSE, booktabs = TRUE)
 ```
@@
 
-\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022].
-Boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired from a data repository of the British Columbia Oil and Gas Research and Innovation Society (BC ORGIS) that was collected as part of the boreal caribou radio-collaring program of the BC OGRIS Research and Effectiveness Monitoring Board [REMB$;$ @bc_ogris_boreal_2018].
+\noindent Elk data from @ciuti_human_2012 were downloaded from Movebank [@kays_movebank_2022].
+Boreal caribou and wolf telemetries were acquired from a data repository of the British Columbia Oil and Gas Research and Innovation Society (BC ORGIS) that was collected as part of the boreal caribou radio-collaring program of the BC OGRIS Research and Effectiveness Monitoring Board [REMB$;$ @bc_ogris_boreal_2018].
 ***MISSING INFO ON OTHER TELEMETRY DATA.*** From the full set of telemetry data, we removed a total of 2396 problematic GPS locations (0.16%, including collar calibration data) after inspecting diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points.
@@
 \@ref(fig:dags)).
-Many species adapt their movement rates (e.g., movement frequency and speed) daily or seasonally in response in factors such as solar time, photoperiod, forage availability, reproductive cycles, and predator avoidance.
+Many species alter their movement rates (e.g., movement frequency and speed) daily or seasonally in response in factors such as solar time, photoperiod, forage availability, reproductive cycles, and predator avoidance.
 Similarly, ambient temperature also fluctuates throughout the day and across seasons.
@@
 
-\noindent We estimated the effects of temperature on mammals' instantaneous movement state (moving or not) and speed when moving using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019 and Appendix B] with the `mgcv` package for `R`.
+\noindent We estimated the effects of temperature on mammals' instantaneous movement state (moving or not) and speed when moving using two Hierarchical Generalized Additive Models [HGAMs$;$ @pedersen_hierarchical_2019 and Appendix B] with the `mgcv` package for `R`.
 The first HGAM estimated the probability that an animal was moving, $P(M)$, with a binomial family of distributions and logit link function.
@@
 
-\noindent Overall, 2.6% of GPS locations had temperatures lower than $-20^\circ$C, while 6.5% had temperatures above $20^\circ$C, and temperature ranges differed across species (Table \@ref(tab:hgam-summary), Fig.
+\noindent Of the GPS fixes with finite speed estimates, 2.6% had temperatures lower than $-20^\circ$C, while 6.5% had temperatures above $20^\circ$C, and temperature ranges differed across species (Table \@ref(tab:hgam-summary), Fig.
 B2).
@@
 d <- readRDS('data/hgam-speed-data.rds')
+
 get_perc <- function(bools) {
@@
   if(grepl('\\.', out)) {
-    out <- paste0(out, ' %')
+    out <- paste0(out, '')
   } else {
-    out <- paste0(out, '.0 %')
+    out <- paste0(out, '.0')
   }
@@
   'E(D|T)' = `P(M=1|T)` * (`E(S|T)`)) %>%
-  mutate(Species = case_when(
-    Species == 'Rangifer tarandus (s. mountain)' ~ 'R. tarandus; sm',
-    Species == 'Ursus arctos horribilis' ~ 'U. arctos h.',
-    Species == 'Puma concolor' ~ 'P. concolor',
-    Species == 'Cervus canadensis' ~ 'C. canadensis',
-    Species == 'Rangifer tarandus (boreal)' ~ 'R. tarandus; b',
-    Species == 'Canis lupus' ~ 'Canis lupus',
-    Species == 'Oreamnos americanus' ~ 'O. americanus')) %>%
-  mutate(Species = paste0('\\emph{', Species, '}') %>%
-           gsub(')}', ')', .) %>%
-           gsub(' \\(', '} \\(', .))
+  mutate(Species = COMMON_NAMES)
 
@@
 
-  # find difference in distance traveled, relative to furtherst distance
+  # find difference in distance traveled, relative to furthest distance
   intercepts %>%
@@
 
-  # find difference in distance traveled, relative to furtherst distance
+  # find difference in distance traveled, relative to furthest distance
   round(range(max(intercepts$`E(D|T)`) / intercepts$`E(D|T)`[- which.max(intercepts$`E(D|T)`)]), 1)
@@
 
-bind_rows(
 # species-level proportions
@@
   group_by(Species) %>%
-    summarize(`T<-20\\textdegree{}C` = get_perc(temp_c < -20),
-              `T>+20\\textdegree{}C` = get_perc(temp_c > +20)),
-  # overall proportions
-  d %>%
-    summarize(Species = 'Total',
-              `T<-20\\textdegree{}C` = get_perc(temp_c < -20),
-              `T>+20\\textdegree{}C` = get_perc(temp_c > +20))) %>%
-  select(! Species) %>%
+  summarize(n = n(),
+            `T<-20\\textdegree{}C (%)` = get_perc(temp_c < -20),
+            `T>+20\\textdegree{}C (%)` = get_perc(temp_c > +20)) %>%
   bind_cols(intercepts %>%
@@
               rename_with(\(.x) gsub('E', '\\\\hat{\\\\mathbb E}', .x)) %>%
-              bind_rows(tibble(Species = 'Total'))) %>%
+              select(! Species)) %>%
+  mutate(Species = COMMON_NAMES[map_int(Species,
+                                        \(.s) which(SPECIES == .s))]) %>%
   relocate(Species, .before = 1) %>%
-  rename(`$\\hat{P}$(M = 1)` = `$\\hat{P}(M=1)$`) %>% # to avoid eq going to new line
-  knitr::kable(format = 'markdown', digits = 2, caption = paste0("Summary statistics for the observed temperatures and estimated mean movement rates at \U00B0", "C.
-The second and third columns indicate the percentage of GPS locations (after data cleaning) with temperature ($T$) below -20\U00B0", "C and above 20\U00B0", "C, respectively.
-The subsequent columns indicate the estimated mean probability of moving ($\\hat P(M=1)$), mean speed when moving ($\\hat{\\mathbb E}(S)$; km/h), and mean distance travelled ($\\hat P(M=1) \\times \\hat{\\mathbb E}(S) = \\hat{\\mathbb E}(D)$; km/h) as estimated by models without and with temperature (i.e., $(...
+  arrange(Species) %>% # arrange alphabetically by common name
+  mutate(Species = if_else(Species == 'Caribou (southern mountain)',
+                           'Caribou (s. mountain)', Species)) %>%
+  # add overall proportions
+  bind_rows(
+    d %>%
+      summarize(Species = 'Total',
+                n = n(),
+                `T<-20\\textdegree{}C (%)` = get_perc(temp_c < -20),
+                `T>+20\\textdegree{}C (%)` = get_perc(temp_c > +20))) %>%
+  knitr::kable(format = 'markdown', digits = 2, format.args = list(big.mark = ','), caption = paste0("Summary statistics for the GPS fixes with finite speed estimates, namely: the number fixes after data cleaning (n), the percentage of fixes with temperature ($T$) below -20\U00B0", "C and above 20\U00B0", "C, the estimated mean probability of moving ($\\hat P(M=1)$), the mean speed when moving ($\\hat{\\mathbb E}(S)$; km/h), and the mean distance travelled ($\\hat P(M=1) \\times \\hat{\\mathbb E}(S) = \\hat{\\mathbb E}(D)$; km/h), as estimated by models without and with temperature (i.e., $(...
 | T)$), for a sampling interval of 1 hour and a temperature of T = 0\U00B0", "C."))
@@
 There was relatively strong agreement between models with and without temperature (Figs.
-B3, and B14), but including temperature always resulted in better fits (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; Table B2 and subsequent pages in Appendix B).
+B3, and B14), but including temperature always resulted in better fits (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; Appendix B).
 Accounting for temperature even resulted in a simpler HRSF for mountain goats (lower model degrees of freedom) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.
@@
 11 of @mcelreath_statistical_2020].
-All species adapted their daily and seasonal movement behaviour to changes in temperature (Fig.
+All species altered their daily and seasonal movement behaviour to changes in temperature (Fig.
 \@ref(fig:dist)).
@@
 All species shifted elevationally with temperature, although boreal caribou ha a much sharper change in RSS across elevation.
-Elk, mountain goats, and cougars increased in elevation with temperature, while wolves, southern mountain caribou, and grizzly bears decreased in elevation.
+Elk, mountain goats, and cougars moved to higher elevations with temperature, while wolves, southern mountain caribou, and grizzly bears moved to lower elevations.
 Wolves, elk, and southern mountain caribou moved closer to water with temperature, while mountain goats, cougars, and grizzly bears moved somewhat further away from water but remained mainly within 5-10 of water.
@@
 
-```{r t-dist, fig.cap="\\textbf{Species will adapt their movement rates differently in response to climate change, but worse climate-change scenarios will result in the greatest change.} Bold lines indicate the median change in distance travelled (probability of moving times speed when moving) due to predicted changes in temperature within each species' observed extent.
+```{r t-dist, fig.cap="\\textbf{Species will alter their movement rates differently in response to climate change, but worse climate-change scenarios will result in the greatest change.} Bold lines indicate the median change in distance travelled (probability of moving times speed when moving) due to predicted changes in temperature within each species' observed extent.
 Shaded areas indicate the $95^\\text{th}$ and $5^\\text{th}$ percentiles.
@@
 
-```{r t-rss-cc, fig.cap="\\textbf{Species will adapt their habitat selection differently in response to climate change, but worse climate-change scenarios will result in the greatest change.} Bold lines indicate the change in median RSS due to predicted changes in temperature within each species' observed extent.
+```{r t-rss-cc, fig.cap="\\textbf{Species will alter their habitat selection differently in response to climate change, but worse climate-change scenarios will result in the greatest change.} Bold lines indicate the change in median RSS due to predicted changes in temperature within each species' observed extent.
 Shaded areas indicate the $95^\\text{th}$ and $5^\\text{th}$ percentiles in change in RSS.
@@
 
-changes in the phenology of plants, fire, ice, or mammalian physiology (e.g., moulting, fat storages) or behaviour (e.g., migration, food caching), other than as direct responses to temperature.
+changes in the phenology of plants, fire, ice, or mammalian physiology (e.g., moulting, fat storage) or behaviour (e.g., migration, food caching), other than as direct responses to temperature.
 
@@
 
-Our work highlights three central considerations for future research: (1) telemetry sampling rates should be designed primarily in relation to the movement timescales of the species of interest [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @pease_ecological_2024], (2) the number of individuals tracked and the duration of each telemetry should depend not just on population size but also the (accelerating) rate of environmental change, and (3) predicting changes in movement behaviour becomes highly complex when responses are nonlinear, especially when changes depend on many interacting factors [@polazzo_measuring_2024] and one is interested in predicting responses in extreme conditions for which data are scarce [@steixner-kumar_strategies_2020].
+Our work highlights three central considerations for future research:
+
+(1) tree cover data should vary over time to account for changes over time due to fires and logging
+
+(2) telemetry fixes should be frequent enough to reconstruct animals' movement paths and estimate speed [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @pease_ecological_2024],
+
+(3) the number of individuals tracked and the duration of each telemetry should depend not just on population size but also the (accelerating) rate of environmental change, and
+
+(4) predicting changes in movement behaviour becomes highly complex when responses are nonlinear, especially when changes depend on many interacting factors [@polazzo_measuring_2024] and one is interested in predicting responses in extreme conditions for which data are scarce [@steixner-kumar_strategies_2020].
 
@@
 
-# Author contributions
+# Author contributions {-}
 
-SM performed the data cleaning, ran the analyses, and wrote the manuscript.
+\noindent SM performed the data cleaning, ran the analyses, and wrote the manuscript.
 CHF wrote code for estimating instantaneous movement speeds.
@@
 All authors contributed to reviewing the manuscript.
+***TO FIX BASED ON COMMENTS FROM OTHER AUTHORS DURING REVIEW***
 

### 2025-09-12 writing/manuscript.Rmd
reviewed; need to finish discussion & abstract

@@
     institute: [ucf, smith]
+  - name: Karen E. Hodges
+    institute: [braes, biol]
   - name: Siobhan Darlington
@@
     institute: [forest]
-  - name: Karen E. Hodges
-    institute: [braes, biol]
   - name: Kirk Safford
@@
 * Rising temperatures alter the movement phenology of terrestrial mammals
-* Terrestrial mammals alter their movement behavior to rising temperatures
+* Terrestrial mammals alter their movement behaviour to rising temperatures
 
@@
 We address this gap by using continuous-time movement models and hierarchical generalized additive models to quantify the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species throughout western Canada between 1998 and 2023.
-We show that temperature is a strong determinant of how and where mammals move, and we predict that climate change will have significant impacts on mammals' movement rates and habitat selection throughout the 21^st^ century.
+We show that temperature is a strong driver of how and where mammals move, and we predict that climate change will have significant impacts on mammals' movement rates and habitat selection throughout the 21^st^ century.
 We found that species did not respond uniformly to rising temperatures, so community-level responses will likely be complex as some species move more and others less.
@@
 Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @merkle_large_2016; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024; @schwerdt_ecological_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020].
-However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and extreme temperatures can reduce movement rates and cause physiological damage [@hetem_responses_2014; @dyer_travel_2023].
-Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [@hall_microrefuges_2016].
+However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], and extreme temperatures can reduce movement rates and cause physiological damage [@hetem_responses_2014; @dyer_travel_2023].
+Animals may search for a short-term buffer from extreme heat or cold by moving to thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [@hall_microrefuges_2016].
 By altering mammals' movement speed and habitat use [jointly, their movement behaviour: @nathan_movement_2008; @nathan_big-data_2022], ambient temperature also affects the consequences of movement behaviour [@alston_temperature_2020; @giroux_activity_2023], such as encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, or competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024].
@@
 It remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially if populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018].
-Preparing for and responding to future changes require a comprehensive understanding of how temperature will affect how and where mammals move, as well as the implications such changes will have on species' adaptability and overall community structure [@cunningham_opportunity_2021].
+Preparing for and responding to future changes require a detailed understanding of how temperature will affect how and where different mammalian species move [@cunningham_opportunity_2021].
 Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], thermoregulation [@mota-rojas_physiological_2021], and movement behaviour [@mccain_assessing_2019; @melin_effect_2023; @cunningham_opportunity_2021].
@@
 
-Western Canada is currently experiencing rapid widespread warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme events [@zhang_increased_2023].
+Western Canada is currently experiencing accelerating and widespread but spatially heterogeneous warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme events [@zhang_increased_2023].
 As we approach the deadline for the Kunming-Montreal Global Biodiversity Framework (i.e., the "30 by 30" initiative), a global effort to conserve 30% of the world's lands and oceans by 2030 [@convention_on_biological_diversity_154_2022], Canadian decision makers must understand which areas will be of greatest value for maximizing and protecting biodiversity not only for current conditions but future decades, too.
@@
 
+Large-bodied mammals are most likely to show changes in movement behaviour with rising temperatures, since they tend to move long distances [@merkle_large_2016], are particularly sensitive to overheating [@fuller_towards_2016], and their movement rates are thus strongly limited by excessive heat [@dyer_travel_2023].
+Conveniently, their large body size also allows for larger tracking devices that have longer battery life and allow longer tracking periods and with finer temporal resolution [@kenward_manual_2001].
 In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou: *Rangifer tarandus*, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves: *Canis lupus*; Table \@ref(tab:data-table)).
 Using over 25 years of telemetry data throughout a large spatial range of western Canada and hourly air temperature data, we estimate how these mammals altered their probability of moving, speed when moving, and habitat selection in response to temperature.
-We then pair the estimated responses with climate change projections to forecast the species' movement behaviour during the 21^st^ century under different climate-change scenarios [referred to as Shared Socioeconomic Pathways, SSP, @riahi_shared_2017].
+We then pair the estimated responses with monthly climate change projections [@mahony_global_2022] to forecast each species' movement behaviour during the 21^st^ century under four different climate-change scenarios, namely Shared Socioeconomic Pathways (SSPs): 1-2.6 ("Sustainability -- Taking the Green Road"; low challenges to mitigation and adaptation), 2-4.5 ("Middle of the Road: medium challenges to mitigation and adaptation), 3-7.0 ("Regional Rivalry -- A Rocky Road"; high challenges to mitigation and adaptation), and 5-8.5 ("Fossil-fueled Development -- Taking the Highway"; high challenges to mitigation, low challenges to adaptation).
+@riahi_shared_2017 provide a detailed overview of SSPs.
 We discuss the consequences of these changes in movement behaviour at the individual, population, and community levels.
@@
 While we acknowledge there are other important variables (e.g., forest type, structure, and age; terrain slope and aspect), we limited our models to only use three simple variables for the sake of parsimony and comparability across species.
+The simplicity of the models also allowed for simple forecasts under the assumption of temporally invariant landscapes.
 See the discussion section for additional detail on the advantages and limitations of using such simple models.
@@
 
-\noindent Elk data from @ciuti_human_2012 were downloaded from Movebank [@kays_movebank_2022].
-Boreal caribou and wolf telemetries were acquired from a data repository of the British Columbia Oil and Gas Research and Innovation Society (BC ORGIS) that was collected as part of the boreal caribou radio-collaring program of the BC OGRIS Research and Effectiveness Monitoring Board [REMB$;$ @bc_ogris_boreal_2018].
-***MISSING INFO ON OTHER TELEMETRY DATA.*** From the full set of telemetry data, we removed a total of 2396 problematic GPS locations (0.16%, including collar calibration data) after inspecting diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points.
+\noindent Boreal caribou and wolf telemetries were acquired from a data repository of the British Columbia Oil and Gas Research and Innovation Society (BC ORGIS) that was collected as part of the boreal caribou radio-collaring program of the BC OGRIS Research and Effectiveness Monitoring Board [REMB$;$ @bc_ogris_boreal_2018].
+Southern mountain caribou data ***XXX***.
+Cougar data include telemetries from @darlington_southern_2025 and @serrouya_ref.
+Elk data from the work of @ciuti_human_2012 were downloaded from Movebank [@kays_movebank_2022].
+***MISSING INFO ON GRIZZLY BEAR AND GOAT TELEMETRY DATA.***
+
+From the full set of telemetry data, we removed a total of 2396 problematic GPS locations (0.16%, including collar calibration data) after inspecting diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points.
 Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated relatively stationary behaviour (Appendix A).
@@
 1.0, @burnett_climatenar_2023] for `R` and ClimateNA [v.
-7.4.2$;$ @wang_locally_2016; @mahony_global_2022] to obtain projected average monthly temperatures from 2020 to 2100 at a 0.08$^\circ$ spatial resolution.
+7.4.2$;$ @wang_locally_2016; @mahony_global_2022] to obtain projected average monthly temperatures from 2025 to 2100 at a 0.08$^\circ$ spatial resolution.
 Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the historical data for western Canada from 1998 to 2023 (extremes included) to estimate the monthly variation in temperature, which we defined as $\sigma^2_T$, the variance at location $\langle x, y \rangle$ in month $m$ for each year (e.g., the variance at coordinates $\langle -119.40, 49.94 \rangle$ in April 2005).
@@
 
-\noindent We estimated percent forest cover and distance from water using the rasters created by @tuanmu_global_2014.
-We calculated total forest cover by summing the temporally static rasters of evergreen/deciduous needle-leaf trees, evergreen broadleaf trees, deciduous broadleaf trees, and mixed/other trees (raster classes 1-4, respectively).
+\noindent We estimated percent forest cover and distance from water using the temporally static rasters created by @tuanmu_global_2014.
+We calculated total forest cover by summing the four rasters of evergreen/deciduous needle-leaf trees, evergreen broadleaf trees, deciduous broadleaf trees, and mixed/other trees (raster classes 1-4, respectively).
 We converted the raster of percent cover of open water (class 12) to a binary raster of pixels with water (cover > 0%) or without water (cover = 0%) and then calculated each pixel's distance from the nearest pixel with water using the `distance()` function from the `terra` package.
@@
 We used a raster with a zoom of 6 (a resolution of 0.009$^\circ$) for model fitting and one with a zoom of 3 (a resolution of 0.08$^\circ$) for downloading climate change projections via `climatenaR`.
-All final rasters and code are available in the GitHub repository associated with this manuscript (see the Data Availability Statement).
+All final rasters and code are available on GitHub (see the Data Availability Statement).
+For ease of interpretability and comparison with current conditions, we assumed resources would remain constant through to 2100, although we recognize that the spatial distribution of forest cover and open water will likely change throughout the 21^st^ century.
 
@@
 
-Since $\texttt{ctmm}\text{'s}$ movement models assume a single moving state with stochastic but non-zero speed, we corrected data-informed speeds so that the minimum instantaneous speed could be 0.
+Since $\texttt{ctmm}\text{'s}$ movement models assume a single moving state with stochastic but non-zero speed [@calabrese_ctmm_2016; @noonan_scale-insensitive_2019], we corrected data-informed speeds so that the minimum instantaneous speed could be 0.
 We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed.
@@
 
-The two following sections detail the statistical framework and models we used to estimate the effect on temperature on the six species' movement behavior.
+The two following sections detail the statistical framework and models we used to estimate the effect on temperature on the six species' movement behaviour.
 To assess the importance of including temperature as an explicit covariate (as ooposed to including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the fits of the two sets of models via analyses of deviance (a form of generalized likelihood ratio tests) following the methods of Section 3.3 in @wood_generalized_2017.
@@
 The models also had three tensor product interaction terms `by` each species: (1) day of year and time of day, (2) temperature and time of day, and (3) temperature and day of year.
-These three terms accounted for smooth changes in: (1) daily behavior across day of year, (2) the response to temperature over time of day (e.g., changes in nocturnality), and (3) the response to temperature over day of year (e.g., changes in coats and migration timing).
+These three terms accounted for smooth changes in: (1) daily behaviour across day of year, (2) the response to temperature over time of day (e.g., changes in nocturnality), and (3) the response to temperature over day of year (e.g., the timing of molting, migration, and hybernation).
 Finally, two smooth terms of log-transformed sampling interval (`dt`; hours) corrected for biases in speed estimates arising from irregular GPS sampling intervals, since longer intervals result in lower speed estimates [@nathan_big-data_2022; @denicola_are_2025].
@@
 
-\noindent We estimated the effects of temperature on each species' selection for percent forest cover (`forest_perc`), elevation (`elevation_m`), and distance from water (`dist_water_m`) by fitting a Hierarchical Resource Selection Function (HRSF) for each species.
+\noindent We estimated the effects of temperature on each species' selection for percent forest cover (`forest_perc`), elevation (`elevation_m`, in meters), and distance from water (`dist_water_m`, in meters) by fitting a Hierarchical Resource Selection Function (HRSF) for each species.
 We fit each HRSF using an HGAM with a Poisson family of distributions and log link function [Appendix B$;$ @aarts_estimating_2008].
@@
 \begin{cases}
-C \sim \text{Pois}(\lambda) \\
-\mathbb E(C) = \mathbb V(C) = \lambda \\
+O \sim \text{Pois}(\lambda) \\
+\mathbb E(O) = \mathbb V(O) = \lambda \\
 \begin{aligned}
@@
 \noindent where $O$ indicates whether an animal was observed ($O = 1$) or not ($O = 0$), and the species-level indices are omitted for readability, but each term in the model can be assumed to be species-specific.
-Smooth effects of percent forest cover (\texttt{forest\_perc}), elevation (\texttt{elev\_m}), and distance to water (\texttt{dist\_water\_m}) accounted for the species-level selection strength for each resource.
+Smooth effects of percent forest cover (\texttt{forest\_perc}), elevation (\texttt{elevation\_m}, in meters), and distance to water (\texttt{dist\_water\_m}, in meters) accounted for the species-level selection strength for each resource.
 A Gaussian random effect for each individual animal ($Z_\texttt a$) corrected for uneven sampling across individuals, while factor-smooth interaction terms for each animal ($f_{j,\texttt a}$) accounted for animal-level resource selection [i.e., individual-level deviations from the species-level estimate$;$ @jeltsch_need_2025].
@@
 Consequently, we present all results while predicting specifically for one-hour sampling intervals.
-At $0^\circ$C, species differed in mean probabilities of moving ($\hat P(M)$; range: 0.05 -- 0.31), mean speed when moving ($\hat {\mathbb E}(S|M=1)$; range: 0.42 -- 2.67 km/h), and mean distance traveled (i.e., $\hat P(M) \times \hat {\mathbb E}(S|M=1)$, range: 0.04 -- 0.61 km/h; Table \@ref(tab:hgam-summary)).
+At $0^\circ$C, species differed in estimated mean probabilities of moving ($\hat P(M = 1)$; range: 0.05 -- 0.31), mean speed when moving ($\hat {\mathbb E}(S|M=1)$; range: 0.42 -- 2.67 km/h), and mean distance traveled (i.e., $\hat P(M) \times \hat {\mathbb E}(S|M=1)$, range: 0.04 -- 0.61 km/h; Table \@ref(tab:hgam-summary)).
 Grizzly bears had the lowest movement frequency ($\hat P(M) \approx 0.05$), while wolves and cougars moved most often ($\hat P(M) \ge 0.22$).
@@
 library('dplyr')
+library('purrr')
 library('ctmm') #' for `%#%`
@@
   Species = SPECIES,
-  'P(M=1)' = get_int(readRDS('models/binomial-gam-without-temperature.rds')),
-  'P(M=1|T)' = get_int(readRDS('models/binomial-gam.rds')),
-  'E(S)' = 'km/h' %#% get_int(readRDS('models/gamma-gam-without-temperature.rds')),
-  'E(S|T)' = 'km/h' %#% get_int(readRDS('models/gamma-gam.rds')),
-  'E(D)' = `P(M=1)` * (`E(S)`),
-  'E(D|T)' = `P(M=1|T)` * (`E(S|T)`)) %>%
-  mutate(Species = COMMON_NAMES)
+  p = get_int(readRDS('models/binomial-gam.rds')),
+  s = 'km/h' %#% get_int(readRDS('models/gamma-gam.rds')),
+  d = p * s) %>%
+  rename('P(M=1|T=0^\\circ C)' = p,
+         'E(S|M=1,T=0^\\circ C)' = s,
+         'E(D|T=0^\\circ C)' = d) %>%
+  mutate(Species = COMMON_NAMES) %>%
+  arrange(Species) %>%
+  rename_with(\(.x) paste0('$', .x, '$'), ! Species) %>%
+  rename_with(\(.x) gsub('\\|', '\\\\vert ', .x)) %>%
+  rename_with(\(.x) gsub('P', '\\\\hat{P}', .x)) %>%
+  rename_with(\(.x) gsub('E', '\\\\hat{\\\\mathbb E}', .x))
 
@@
   layout(1:3)
-  hist(intercepts$`P(M=1)`); hist(intercepts$`E(S)`); hist(intercepts$`E(D)`)
+  hist(intercepts$`P(M=1)`); hist(intercepts$`E(S|M=1)`); hist(intercepts$`E(D)`)
   layout(1)
@@
               p = round(`P(M=1|T)` - `P(M=1)`, 2),
-              s = round(`E(S|T)` - `E(S)`, 2),
+              s = round(`E(S|M=1,T)` - `E(S|M=1)`, 2),
               d = round(`E(D|T)` - `E(D)`, 2))
@@
   group_by(Species) %>%
-  summarize(n = n(),
-            `T<-20\\textdegree{}C (%)` = get_perc(temp_c < -20),
-            `T>+20\\textdegree{}C (%)` = get_perc(temp_c > +20)) %>%
-  bind_cols(intercepts %>%
-              arrange(Species) %>%
-              rename_with(\(.x) paste0('$', .x, '$'), ! Species) %>%
-              rename_with(\(.x) gsub('\\|', '\\\\vert ', .x)) %>%
-              rename_with(\(.x) gsub('P', '\\\\hat{P}', .x)) %>%
-              rename_with(\(.x) gsub('E', '\\\\hat{\\\\mathbb E}', .x)) %>%
-              select(! Species)) %>%
+  summarize('$n$' = n(),
+            'T<-20\\textdegree{}C (%)' = get_perc(temp_c < -20),
+            'T>+20\\textdegree{}C (%)' = get_perc(temp_c > +20)) %>%
   mutate(Species = COMMON_NAMES[map_int(Species,
                                         \(.s) which(SPECIES == .s))]) %>%
-  relocate(Species, .before = 1) %>%
+  full_join(intercepts, by = 'Species') %>%
   arrange(Species) %>% # arrange alphabetically by common name
@@
       summarize(Species = 'Total',
-                n = n(),
-                `T<-20\\textdegree{}C (%)` = get_perc(temp_c < -20),
-                `T>+20\\textdegree{}C (%)` = get_perc(temp_c > +20))) %>%
-  knitr::kable(format = 'markdown', digits = 2, format.args = list(big.mark = ','), caption = paste0("Summary statistics for the GPS fixes with finite speed estimates, namely: the number fixes after data cleaning (n), the percentage of fixes with temperature ($T$) below -20\U00B0", "C and above 20\U00B0", "C, the estimated mean probability of moving ($\\hat P(M=1)$), the mean speed when moving ($\\hat{\\mathbb E}(S)$; km/h), and the mean distance travelled ($\\hat P(M=1) \\times \\hat{\\mathbb E}(S) = \\hat{\\mathbb E}(D)$; km/h), as estimated by models without and with temperature (i.e., $(...
-| T)$), for a sampling interval of 1 hour and a temperature of T = 0\U00B0", "C."))
+                '$n$' = n(),
+                'T<-20\\textdegree{}C (%)' = get_perc(temp_c < -20),
+                'T>+20\\textdegree{}C (%)' = get_perc(temp_c > +20))) %>%
+  knitr::kable(format = 'markdown', digits = 2, format.args = list(big.mark = ','), caption = "Summary statistics for each species' GPS fixes with finite speed estimates, namely: the number fixes after data cleaning ($n$), the percentage of fixes with temperature ($T$) below $-20^\\circ$C and above $20^\\circ$C, the estimated mean probability of moving ($\\hat P(M=1)$), the mean speed when moving ($\\hat{\\mathbb E}(S|M=1)$; km/h), and the mean distance travelled ($\\hat P(M=1) \\times \\hat{\\mathbb E}(S|M=1) = \\hat{\\mathbb E}(D)$; km/h), for a sampling interval of 1 hour and a temperature $T$ of $0^\\circ$C.")
 ```
@@
 There was relatively strong agreement between models with and without temperature (Figs.
-B3, and B14), but including temperature always resulted in better fits (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; Appendix B).
+B3, and B14), but including temperature always resulted in better fits (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\le$ -342; Appendix B).
 Accounting for temperature even resulted in a simpler HRSF for mountain goats (lower model degrees of freedom) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.
@@
 
-```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals travel.} The fill indicates the effects of temperature on the relative change in distance travelled (probability of moving times mean speed when moving) over time of day on June $1^{\\text{st}}$ (\\textbf{A}) and day of year at 12:00 PDT (\\textbf{B}).
+```{r dist, fig.cap="\\textbf{Temperature is a strong driver of how far and when mammals travel.} The fill indicates the effects of temperature on the relative change in distance travelled (probability of moving times mean speed when moving) over time of day on June $1^{\\text{st}}$ (\\textbf{A}) and day of year at 12:00 PDT (\\textbf{B}).
 Predictions in the surface plots extend to 10% of the range away from each datum.
@@
 B9 and B10).
+Estimated changes in average distance travaled were negligible for boreal caribou, and grizzly bears, although both species showed seasonal changes in seasonal movement rates.
+As temperatures warmed, boreal caribou moved further in winter, spring, and fall but less in summer, while grizzly bears showed a clear shift towards earlier emergence from hibernation (Fig.
+B4) and greater movement earlier in the year but less movement in early fall.
+Southern mountain caribou and mountain goats are projected to travel further by 2100, although the estimates for southern mountain caribou varied greatly over space (Fig.
+\@ref(fig:s-dist)).
+Cougars, elk, and wolves were projected to move less by 2100, with cougars showing as much as a 24% decrease in mean yearly distance travelled.
+
 Absolute relative changes in average yearly distance travelled were small under the best-case SSP (0-4% change in 2100 relative to 2025).
 Under the worst-case SSP, absolute changes by 2100 (relative to 2025) ranged from ~2% (grizzly bears) to ~24% (cougars).
-Notably, while the models estimated that grizzly bears would move substantially less (if at all) in late fall and winter (Figs.
-\@ref(fig:dist), B4), the models did not explicitly account for changes in hibernation phenology.
-Consequently, the climate change projections do not explicitly account for changes in energetic needs or physiology.
 Projected changes in 2100 varied spatially due to spatial heterogeneity in climate change projections (Fig.
@@
 The predictions only account for the predicted temperature throughout the areas and ignore environmental factors such as terrain slope, soil type, and forest density.
-All maps are shown in the BC Albers Equal Area Conic projection (EPSG:3005)."}
+All maps extend to each species' observed extent (Fig.
+1) and are shown in the BC Albers Equal Area Conic projection (EPSG:3005)."}
 if(FALSE) range(readRDS('../data/cc-hgam-projections-local-2100.rds')$d)
@@
 For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.8 and 1.25 ($\\approx 2^{\\pm 0.322}$; original data ranged 0.71 to 1.93).
-All maps are shown in the BC Albers Equal Area Conic projection (EPSG:3005)."}
+All maps extend to each species' observed extent (Fig.
+1) and are shown in the BC Albers Equal Area Conic projection (EPSG:3005)."}
 if(FALSE) range(readRDS('../data/cc-hrsf-projections-local-2100.rds')$l)
@@
 
-\noindent We have demonstrated that temperature is an important determinant of how and where large mammals move, and that mammals alter their seasonal and daily cycles in movement behavior in response to changes in temperature.
-Predicting mammals' responses to climate change remains a challenge, as habitats warm rapidly and mammals are exposed to increasingly novel and extreme conditions [@ipcc_climate_2023; @diffenbaugh_changes_2013; @botero_evolutionary_2015].
+\noindent We have demonstrated that temperature is an important driver of how and where large mammals move, and that mammals alter their seasonal and daily cycles in movement behaviour in response to changes in temperature.
+However, predicting mammals' responses to climate change remains a challenge, as habitats warm rapidly and mammals are exposed to increasingly novel and extreme conditions [@ipcc_climate_2023; @diffenbaugh_changes_2013; @botero_evolutionary_2015].
 We leveraged the flexibility and interptetability of GAMs to estimate the (nonlinear) responses  of terrestrial mammals to temperature.
 In particular, tensor product interaction terms were crucial for estimating smooth, nonlinear changes in daily and seasonal trends with temperature.
-By allowing interaction terms to be smooth bivariate functions, we were able to quantify changes in movement behavior without the need to discretize time into distinct windows (e.g., day/night and seasons).
+By allowing interaction terms to be smooth bivariate functions, we were able to quantify changes in movement behaviour without the need to discretize time into distinct windows (e.g., day/night and seasons).
 We were thus able to present changes using relatively simple and interpretable surface plots (Figs.
@@
 As temperatures warm, mammals' risk of heperthermia increases nonlinearly [@sherwood_adaptability_2010], especially for large-bodied species [@alston_temperature_2020; @dyer_travel_2023; @verzuh_behavioural_2023].
-The recent increase in frequency and duration of extreme-temperature events [@yao_emergence_2022; @ipcc_climate_2023; @zhang_increased_2023; @bunnell_global_2011; @kienzle_has_2018] have forced many mammals towards increasingly atypical behaviours with long-term consequences, ranging from changes in the behavior and fitness of individuals [@brivio_seeking_2024; @cunningham_opportunity_2021] to mass die-offs [@ameca_y_juarez_natural_2012].
+The recent increase in frequency and duration of extreme-temperature events [@yao_emergence_2022; @ipcc_climate_2023; @zhang_increased_2023; @bunnell_global_2011; @kienzle_has_2018] have forced many mammals towards increasingly atypical behaviours with long-term consequences, ranging from changes in the behaviour and fitness of individuals [@brivio_seeking_2024; @cunningham_opportunity_2021] to mass die-offs [@ameca_y_juarez_natural_2012].
 A species' ability to respond to changing temperature and phenology depends mainly on the its exposure, sensitivity, and adaptability to changes, and growing evidence suggests that many mammals in western Canada will struggle to respond due limitations in all three factors [@ameca_y_juarez_natural_2012; @diffenbaugh_changes_2013].
@@
 Secondly, large-bodied mammals are particularly sensitive to high heat [@dyer_travel_2023], but small-bodied mammals are also affected [@ratnayake_forecasting_2019; @krebs_impact_2019].
-Finally, species' adaptability is tested by the speed of climate change [@diffenbaugh_changes_2013] and the rigidity of many speices' photoperiod-dependent phenology [@walker_global_2019].
+Finally, species' adaptability is tested by the speed of climate change [@diffenbaugh_changes_2013] and the rigidity of many species' photoperiod-dependent phenology [@walker_global_2019].
 
@@
 
-Our results suggest that climate change will cause mammals to alter their seasonal movement behavior, but not all species will respond similarly.
+Our results suggest that climate change will cause mammals to alter their seasonal movement behaviour, but not all species will respond similarly.
 For example, hotter temperatures caused boreal caribou to travel longer distances in fall, winter, and spring but less in summer, mirroring the findings of @stien_congruent_2012, @leclerc_determinants_2021, and @lessard_local_2025.
@@
 Earlier growth seasons in boreal and arctic regions have resulted in lower calf birth and survival [@post_climate_2008], while the lengthening of the growth season has allowed moose (*Alces alces*) and deer (*Odocoileus* spp.) to encroach on boreal caribou habitat, along with increased density of coyotes (*Canis latrans*), cougars, and wolves [@barber_potential_2018; @demars_incorporating_2023].
-Thus, rising temperatures affect more than each individual species' movement behavior; they have complex and interconnected effects on an entire community and its trophic structure.
+Thus, rising temperatures affect more than each individual species' movement behaviour; they have complex and interconnected effects on an entire community and its trophic structure.
 
@@
 Considerations about changes in trophic interactions leads to an important caveat about the results we present.
-The estimated effects of temperature on movement behavior cannot be fully attributed to physiological responses to changes in temperature alone, since other aspects of habitats' seasonal phenology are (nonlinearly) correlated with temperature.
+The estimated effects of temperature on movement behaviour cannot be fully attributed to physiological responses to changes in temperature alone, since other aspects of habitats' seasonal phenology are (nonlinearly) correlated with temperature.
 For example, mountain goats' reduced movement rates at temperatures near 0$^\circ$C in winter may be due to increased movement difficulty, since melting snow and rain-on-snow events affect the energetic costs of movement [@white_population_2025] and encounter rates with predators [@sullender_defining_2023].
@@
 
-Changes in mammals' seasonal movement behavior are also likely to affect the timing and frequency of human-wildlife interactions.
+Changes in mammals' seasonal movement behaviour are also likely to affect the timing and frequency of human-wildlife interactions.
 Since heat affects mammals' sleep quality and likelihood to enter torpor [@mortlock_sleep_2024; @fjelldal_nightly_2021; @wells_life_2022], future decades may see increased levels of incidents with bears that struggle to enter or remain in hibernation [@johnson_human_2018].
@@
 It is therefore crucial for conservationists and other decision-makers to distinguish between a species' ideal environmental space [or E space, *sensu* @matthiopoulos_defining_2022], and its best available geographic space [or G space, *sensu* @matthiopoulos_defining_2022].
-Accelerating multufarious change in species' movement behavior and energetic costs require conservation measures to be proactive and anticipate future changes, even when future conditions are substantially different from historical conditions and historical data may not be available.
+Accelerating multufarious change in species' movement behaviour and energetic costs require conservation measures to be proactive and anticipate future changes, even when future conditions are substantially different from historical conditions and historical data may not be available.
 To this end, the framework presented by @polazzo_measuring_2024 for estimating species' responses to multifarious change becomes essential, particularly when accounting for energetic and fitness costs, as argued by @cunningham_opportunity_2021.
@@
 Understanding the consequences of climate change in mammalian movement behaviour is a first step towards proactively responding to how mammals will respond to human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018].
-However, a better understanding of such consequences will require a more holistic approach that includes carefully data collection, model design, and interpretation and application of results.
+However, a better understanding of such consequences will require a more holistic approach that includes careful data collection, model design, and interpretation and application of results.
 
-## Study limitations and considerations for future work
+## Study strengths and limitations
+
+Our results demonstrate that mammals respond to changes in temperature by altering how and where they move in complex and nonlinear ways (Figs.
+\@ref(fig:dist) and \@ref(fig:hrsfs)).
+Hierarchical Generalized Additive Models were crucial in producing such results.
+The hierarchical approach us to produce statistically sound population-level estimates by accounting for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mortlock_sleep_2024; @jeltsch_need_2025; @silva_too_2025], while the flexibility of GAMs estimated mammals' changes in movement behaviour without imposing rigid assumptions about the shape of the responses [@simpson_method_2025; @simpson_modelling_2018].
+Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously by inexperienced analysts.
+We suggest choosing the basis size (`k`) for each smooth term carefully.
+While penalized splines and REML can help avoid unruly smooth terms that would result in questionable estimates the penalty terms can struggle to restrict model complexity when datasets are particularly large [@wood_generalized_2017; @wood_generalized_2017-1].
+
+Despite the statistical soundness of the models, estimating mammals' response to projected climate change remained a challenge because climate change projections included mean monthly temperatures but no measures of the full distribution of temperature.
+Since mammals' responses to temperature are nonlinear, accurate estimates of the effects of climate change require estimates of the full distribution of temperatures that mammals will be exposed to -- estimated means alone are not sufficient.
+This is because, since the responses to temperature are nonlinear, the mean response across the distribution of temperatures is not the same as the response to mean temperature.
+This inequality is often referred to as Jensen's inequality or the fallacy of the average [@jensen_sur_1906; @denny_performance_2019], and it is a main motivator for Generalized Linear Models and GAMs [@denny_fallacy_2017; @simpson_method_2025].
+We approximated of the distribution of monthly temperatures by assuming monthly temperatures were normally distributed after estimating monthly variance in temperature, but more work is necessary on estimating changes in climate beyond montly mean values.
+
+A second advantage of our approach is the use of continous-time estimates of speed and distance traveled that are less impacted by irregular sampling rates than straight-line displacement [@noonan_scale-insensitive_2019].
+Still, our results indicate that the movement models fit by `ctmm` still suffer from two crucial issues.
+Firstly, although the models allow speeds to vary stochastically over time, they assume non-zero speeds and do not currently allow for multiple movement states [@augermethe_guide_2021].
+Secondly, while continuous-time estimates of speed are less sensitive to sampling rates than straight-line displacement estimates, our models indicate that longer sampling intervals still cause speed and travel distance to be underestimated [Fig.
+B8$;$ also see @denicola_are_2025].
+Shorter sampling intervals may help overcome this bias [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @pease_ecological_2024].
+To account for the occasional missed GPS fix, we suggest setting sampling intervals to five or more times smaller than the animal's expected directional persistence [@noonan_scale-insensitive_2019].
+Still, excessively high sampling can cause the animal's movement to be confounded with GPS error [@fleming_comprehensive_2020], so we suggest calibrating GPS error whenever possible.
+
+The final (and arguably most evident) limitation of our study was the use of temporally static estimates of forest cover.
+While the use of a single raster produced results and forecasts that are relatively simpler and easier to interpret, species' selection to estimated forest cover was relatively weak.
+Future studies should account for temporal variation in forest cover, especially given the growing severity of forest fires in Canada [@zurowski_summer_2023] as well widespread tree harvesting and its impact on wildlife [@betts_forest_2022; @lochhead_linking_2022].
+Additionally, a finer spatial scale may help detect changes in forest edge selection with greater precision.
+Future work would also benefit from accounting for other important forest properties, such as age, tree composition, and canopy height.
+However, accounting for more variables requires careful consideration about what values such variables should take when estimating changes in movement behaviour under future conditions.
+
+## Considerations for future work
 
@@
 
-Estimated selection strength for forest cover was weak and did not change strongly with temperature. Time-varying rasters of forest cover would have likely given stronger (and better) estimates.
+We have presented a multi-species analysis of the effects of temperature on mammalian movement rates and habitat selection in hopes that other researchers can leverage this framework and models to expand our understanding of how temperature affects the movement behavior of other species, including smaller mammals, more elusive species, and non-mammal animals.
 
-* variation among individuals: @jeltsch_need_2025.
-Population-level means that ignore such variation can greatly limit the efficacy of conservation projects [@muff_accounting_2020; @mortlock_sleep_2024; @silva_too_2025].
-
-- we see clear signals despite the data issues and limitations, but more work needs to be done to disentangle important drivers
+large mammals give clear responses, but they are also already studied most often
 
@@
 However, many of these data not available, so we also suggest that more work be done on quantifying such measures using spatiotemporally dynamic models and sufficiently fine resolutions.
-***...*** animals in the study are generalists; we need more work on specialists and smaller species, but they can be harder to track
+***...*** animals in the study are generalists; we need more work on specialists (i.e., species without wide species distributions) and smaller species, but they can be harder to track.
 
 changes in the phenology of plants, fire, ice, or mammalian physiology (e.g., moulting, fat storage) or behaviour (e.g., migration, food caching), other than as direct responses to temperature.
-
-- other confounding causes of movement (forest age, structure, fire, type, dist_water, etc.) change over the years but are assumed to be fixed
-
-- the temporal resolution of climate change projections isn't great (many steps necessary to simulate weather)
-
-***old text:***
-
-Our work highlights three central considerations for future research:
-
-(1) tree cover data should vary over time to account for changes over time due to fires and logging
-
-(2) telemetry fixes should be frequent enough to reconstruct animals' movement paths and estimate speed [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @pease_ecological_2024],
-
-(3) the number of individuals tracked and the duration of each telemetry should depend not just on population size but also the (accelerating) rate of environmental change, and
-
-(4) predicting changes in movement behaviour becomes highly complex when responses are nonlinear, especially when changes depend on many interacting factors [@polazzo_measuring_2024] and one is interested in predicting responses in extreme conditions for which data are scarce [@steixner-kumar_strategies_2020].
-
-While the `ctmm` movement models produced scale-independent estimates of speed [i.e., model interpretation is independent of sampling interval: @noonan_scale-insensitive_2019], the accuracy, size, and stability of speed estimates still depended on the GPS sampling frequency.
-This dependency is because coarsely-sampled movement data contains information on large-scale movements (range crossings, migrations) but not fine-scale movements [@nathan_big-data_2022; @noonan_scale-insensitive_2019], including first-order habitat selection [@johnson_comparison_1980].
-Using the boreal caribou as an example, the 13-hour sampling interval allows us to reasonably estimate the caribou's movement path at a temporal scale of approximately 13 hours (or greater), but we cannot produce reasonable movement trajectories at a much finer (e.g., hourly) scale.
-Consequently, we suggest being cautious when comparing estimated movement behaviours across species, even though all predictions have been corrected to the hourly timescale by predicting for 1-hour time intervals.
-Ideally, sampling schedules should be fine enough to reconstruct individuals' movement paths.
-***make data useful for future studies.
-recognize cost of GPS and suggest IoT [@wild_multi-species_2023], although station coverage can be a limitation [@denicola_monitoring_2025].*** Good estimates of an animal's speed and trajectory require telemetry locations to be taken more often than the animal's directional persistence [@noonan_scale-insensitive_2019], so that, on average, the dataset contains multiple locations in between changes in direction.
-What constitutes a change in direction depends on what movement scale one is investigating.
-Small-scale movements and first-order spatial selection will require more frequent sampling than migratory movement or second-order and third-order spatial selection.
-While `ctmm` movement models are scale-invariant in that any model can be scaled to larger or smaller areas and timescales [@noonan_scale-insensitive_2019], the model estimates are not independent of the scale of the sampling frequency [@denicola_are_2025].
-When tracking data are too coarse for accurate speed estimation, diffusion rates (measured in area per unit time) may be a useful alternative for quantifying animals' exploration rates [@denicola_are_2025].
-***also need better way to distinguish between sates.
-HMMs are an option, but there are limitations.***
-
-# Other refs and miscellaneous maerial {-}
-
-GAMs' flexibility and complexity can also be a frustrating limitation if used too generously.
-As such, analysts should chose the basis size for each smooth term carefully.
-While penalized splines and REML can help avoid unruly smooth terms that would result in questionable estimates where data are scarce [@wood_generalized_2017], it is best to select the basis size of the smooth terms with parsimony, especially when datasets are large.
-Still, predicting mammals' responses at the data-scarce fringes of the surface plots in Figs.
-\@ref(fig:dist) and \@ref(fig:hrsfs) comes with greater uncertainty, particularly given that the changes are nonlinear.
 
@@
 MJN conceived of the project idea and supervised SM throughout the project.
-All other authors, which are listed in alphabetical order by last name, contributed telemetry data and reviewed the interpretation of the results for their species of interest.
+All other authors contributed telemetry data and reviewed the interpretation of the results for their species of interest.
+KH and SD contributed substantially to manuscript review.
 All authors contributed to reviewing the manuscript.
-***TO FIX BASED ON COMMENTS FROM OTHER AUTHORS DURING REVIEW***
 

### 2025-09-27 analysis/figures/2025-tws-edmonton/hrsf-figures.R
fix fill lab; change fig direction; merge scripts

@@ -7,7 +7,15 @@ library('gratia')    # for useful convenience functions for GAMs
 library('ggplot2')   # for fancy plots
 library('khroma')    # for colorblind-friendly color palettes
 library('cowplot')   # for fancy multi-panel plots
+library('terra')   # for rasters
+library('sf')      # for spatial objects
 source('analysis/figures/default-ggplot-theme.R') # bold text and no grids
+source('functions/get_legend.R') # doesn't work in cowplot v.1.1.3.9000
+
+theme_set(theme_get() + theme(text = element_text(size = 15)))
+
+# boreal caribou RSF
+rsf_bc <- readRDS('H:/GitHub/bc-mammals-temperature/models/rsf-Rangifer tarandus boreal-2025-01-21.rds')
 
 #' created in `analysis/figures/temperature-hrsfs.R`
 preds <- readRDS('models/temperature-hrsf-preds.rds') %>%
@@ -61,7 +69,7 @@ p <-
   scale_x_continuous(paste0('Temperature (\U00B0', 'C)'), expand = c(0, 0),
                      breaks = c(-20, 0, 20)) +
   scale_y_continuous(NULL, expand = c(0, 0)) +
-  scale_fill_PRGn(name = 'Relative selection strength', midpoint = 0,
+  scale_fill_PRGn(name = 'Relative selection strength   ', midpoint = 0,
                   limits = c(-LIM, LIM), breaks = -LIM:LIM,
                   labels = \(x) 2^x) +
   theme(strip.placement = 'outside', strip.background.y = element_blank(),
@@ -70,4 +78,227 @@ p <-
         legend.key.width = rel(2))
 
 ggsave('figures/2025-tws-edmonton/boreal-caribou-hrsf-surface-plots.png',
-       p, width = 8, height = 3.5, units = 'in', dpi = 600, bg = 'white')
+       p, width = 10, height = 4.375, units = 'in', dpi = 600, bg = 'white')
+
+# habitat selection strength for the current habitat ----
+locs <- readRDS('data/tracking-data/all-tracking-data-cleaned-2024-02-22-13-49.rds') %>%
+  filter(dataset_name == 'Rangifer_tarandus_boreal') %>%
+  unnest(tel) %>%
+  select(location.long, location.lat) %>%
+  st_as_sf(coords = c('location.long', 'location.lat')) %>%
+  st_set_crs('+proj=longlat')
+
+bounds <- locs %>%
+  st_bbox() %>%
+  st_as_sfc() %>%
+  st_as_sf() %>%
[…]

### 2025-09-27 analysis/figures/2025-tws-edmonton/rsf-figures.R

@@ -1,227 +1,0 @@
-library('mgcv')    # for GAMs
-library('terra')   # for rasters
-library('sf')      # for spatial objects
-library('dplyr')   # for data wrangling
-library('tidyr')   # for data wrangling
-library('ggplot2') # for fancy plots
-source('analysis/figures/default-ggplot-theme.R') # bold text and no grids
-source('functions/get_legend.R') # doesn't work in cowplot v.1.1.3.9000
-
-theme_set(theme_get() + theme(text = element_text(size = 15)))
-
-rsf_bc <- readRDS('H:/GitHub/bc-mammals-temperature/models/rsf-Rangifer tarandus boreal-2025-01-21.rds')
-
-# habitat selection strength for the current habitat ----
-# create custom color palettes for each resource
-locs <- readRDS('data/tracking-data/all-tracking-data-cleaned-2024-02-22-13-49.rds') %>%
-  filter(dataset_name == 'Rangifer_tarandus_boreal') %>%
-  unnest(tel) %>%
-  select(location.long, location.lat) %>%
-  st_as_sf(coords = c('location.long', 'location.lat')) %>%
-  st_set_crs('+proj=longlat')
-
-bounds <- locs %>%
-  st_bbox() %>%
-  st_as_sfc() %>%
-  st_as_sf() %>%
-  st_transform('EPSG:3005')
-
-# projections are unnecessary since we are not making maps
-f_rast <- rast('data/resource-rasters/forest.tif') %>%
-  project('EPSG:3005') %>%
-  crop(bounds)
-
-f <- f_rast %>%
-  as.data.frame(xy = TRUE) %>%
-  rename(forest_perc = 3) %>%
-  filter(! is.na(forest_perc))
-
-e <- rast('data/resource-rasters/bc-buffered-dem-z6.tif') %>%
-  crop(st_buffer(st_transform(bounds, crs(.)), 1e5)) %>%
-  project(f_rast) %>%
-  crop(bounds) %>%
-  as.data.frame(xy = TRUE) %>%
-  rename(elevation_m = 3) %>%
-  filter(elevation_m > 0)
-
-w <- rast('data/resource-rasters/distance-from-water.tif') %>%
-  crop(st_buffer(st_transform(bounds, crs(.)), 1e5)) %>%
-  project(f_rast) %>%
-  crop(bounds) %>%
-  as.data.frame(xy = TRUE, na.rm = TRUE) %>%
[…]

### 2025-09-27 writing/manuscript.Rmd
minor edits to abstract; fixed species in table 1

@@
 ---
-title: "\\Large The heat is on: Rising temperatures alter how and where mammals move"
+title: "\\Large Rising temperatures alter how and where mammals move"
 author:
@@
 csl: 'global-change-biology.csl'
+# csl: 'movement-ecology.csl'
 fontsize: 12pt
@@
 
+\newpage
+
 # Alternative titles {-}
@@
 Previous studies have estimated the effects of temperature on mammalian movement behaviour, but many did not disentangle them from seasonal behaviour cycles.
-Consequently, it is still unclear how mammals will adapt their movement behaviour in the 21^st^ century.
-We address this gap by using continuous-time movement models and hierarchical generalized additive models to quantify the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species throughout western Canada between 1998 and 2023.
-We show that temperature is a strong driver of how and where mammals move, and we predict that climate change will have significant impacts on mammals' movement rates and habitat selection throughout the 21^st^ century.
-We found that species did not respond uniformly to rising temperatures, so community-level responses will likely be complex as some species move more and others less.
-Median habitat selection strength decreased for all species, but some species showed increased selection strength for higher altitudes.
-Consequently, our results suggest that all six species may shift their range in the coming decades as their selection strength for their current ranges weakens.
-As climate change exposes mammals to novel environmental conditions, predicting changes in mammalian behaviour and community structure will become crucial for effectively and proactively understanding mammalian movement behaviour and protecting high-quality habitat.
-We thus present our work using a flexible approach that can be adapted for different species and spatiotemporal scales, and we conclude by providing suggestions for future research, with particular attention to the study design, statistical modeling, and forecasting."
+Consequently, it is still unclear how mammals will adapt their movement behaviour and phenology in the 21^st^ century.
+We address this gap by using hierarchical generalized additive models to quantify the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species throughout western Canada between 1998 and 2023.
+We show that temperature significantly affects of how and where mammals move, and we predict that climate change will have significant impacts on mammals' movement rates and habitat selection throughout the 21^st^ century.
+However, species did not respond uniformly to rising temperatures, so community-level responses will likely be complex, as climate change will cause species to alter their seasonal movement behavior and increase their chances of hyperythermia.
+Median habitat selection strength decreased for most species, but some species showed increased selection strength for higher altitudes.
+Consequently, our results suggest that all six species may shift their range in the coming decades as they begin to select against their current ranges.
+As climate change exposes mammals to novel environmental conditions, proactively predicting changes in mammalian movement behaviour will become crucial for effectively and proactively understanding community-level responses and protecting high-quality habitat.
+We thus present our work using a flexible approach that can be adapted for different species and spatiotemporal scales, and we conclude by providing suggestions for future research."
 if(FALSE){
@@
 
-\noindent \textbf{References}: 142 (updated on 2025-06-11) <!-- counted by using movement ecology csl -->
+\noindent \textbf{References}: 165 (updated on 2025-09-25) <!-- counted by using movement ecology csl -->
 
@@
     Species = case_when(
-      dataset_name == 'Canis_lupus_boreal' ~ COMMON_NAMES[1],
-      dataset_name == 'Rangifer_tarandus_boreal' ~ COMMON_NAMES[2],
-      dataset_name == 'Rangifer_tarandus_southern_mountain' ~ COMMON_NAMES[3],
-      dataset_name == 'Puma_concolor_2' ~ COMMON_NAMES[4],
-      dataset_name == 'Puma_concolor_4' ~ COMMON_NAMES[4],
-      dataset_name == 'Elk in southwestern Alberta' ~ COMMON_NAMES[5],
-      dataset_name == 'Oreamnos_americanus' ~ COMMON_NAMES[6],
-      dataset_name == 'Ursus_arctos_horribilis' ~ COMMON_NAMES[7]),
+      dataset_name == 'Canis_lupus_boreal' ~ 'Wolves',
+      dataset_name == 'Rangifer_tarandus_boreal' ~ 'Caribou (boreal)',
+      dataset_name == 'Rangifer_tarandus_southern_mountain' ~ 'Caribou (southern mountain)',
+      dataset_name == 'Puma_concolor_2' ~ 'Cougars',
+      dataset_name == 'Puma_concolor_4' ~ 'Cougars',
+      dataset_name == 'Elk in southwestern Alberta' ~ 'Elk',
+      dataset_name == 'Oreamnos_americanus' ~ 'Mountain goats',
+      dataset_name == 'Ursus_arctos_horribilis' ~ 'Grizzly bears'),
     animal, # keep animal column for grouping later
@@
   group_by(Species, animal) %>%
-  summarize(Start = min(as.Date(timestamp)),
-            End = max(as.Date(timestamp)),
+  summarize(Start = min(as.Date(timestamp, tz = 'America/Vancouver')),
+            End = max(as.Date(timestamp, tz = 'America/Vancouver')),
             n = n(),

### 2025-09-27 writing/manuscript.Rmd
started addressing Mike's comments

@@
 
-# Alternative titles {-}
-
-* Rising temperatures alter the movement phenology of terrestrial mammals
-* Terrestrial mammals alter their movement behaviour to rising temperatures
-
-\newpage
-
 <!--
@@
 abstract <- "Widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness, habitat selection, and community structure.
-Previous studies have estimated the effects of temperature on mammalian movement behaviour, but many did not disentangle them from seasonal behaviour cycles.
-Consequently, it is still unclear how mammals will adapt their movement behaviour and phenology in the 21^st^ century.
-We address this gap by using hierarchical generalized additive models to quantify the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species throughout western Canada between 1998 and 2023.
-We show that temperature significantly affects of how and where mammals move, and we predict that climate change will have significant impacts on mammals' movement rates and habitat selection throughout the 21^st^ century.
-However, species did not respond uniformly to rising temperatures, so community-level responses will likely be complex, as climate change will cause species to alter their seasonal movement behavior and increase their chances of hyperythermia.
-Median habitat selection strength decreased for most species, but some species showed increased selection strength for higher altitudes.
-Consequently, our results suggest that all six species may shift their range in the coming decades as they begin to select against their current ranges.
-As climate change exposes mammals to novel environmental conditions, proactively predicting changes in mammalian movement behaviour will become crucial for effectively and proactively understanding community-level responses and protecting high-quality habitat.
-We thus present our work using a flexible approach that can be adapted for different species and spatiotemporal scales, and we conclude by providing suggestions for future research."
+Previous studies have estimated the effects of temperature on mammalian movement behaviour, but few disentangled them from seasonal behaviour cycles.
+Consequently, it is still uncertain how mammals will adapt their movement behaviour throughout the 21^st^ century.
+We address this gap by quantifying the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species throughout western Canada between 1998 and 2023.
+We show that temperature significantly affects of how and where these mammals move, and we predict that climate change will have significant impacts on mammals' movement behaviour throughout the 21^st^ century.
+Notably, we found that species did not respond uniformly to temperature.
+This suggests that community-level responses will be complex, as future climate change drives some species to move more and others less.
+Median habitat selection strength was predicted to decrease for most species, but some species are expected to show increased selection strength for higher altitudes.
+Consequently, our results suggest that individuals from all six species may shift their range in the coming decades as they begin to select against portions of their current ranges.
+The strong but varied responses to temperature suggest that future climate change will impact not only individuals' movement behaviour but also entire communities, as changes in movement behaviour will affect encounter rates, including predator prey-dynamics and human-wildlife interactions.
+As climate change exposes mammals to novel environmental conditions, proactively anticipating changes in mammalian movement behaviour will become crucial for effectively and proactively understanding community-level responses and selecting high-quality habitat for long-term conservation."
 if(FALSE){
@@
 \noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008].
-Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @merkle_large_2016; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024; @schwerdt_ecological_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020].
-However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], and extreme temperatures can reduce movement rates and cause physiological damage [@hetem_responses_2014; @dyer_travel_2023].
-Animals may search for a short-term buffer from extreme heat or cold by moving to thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [@hall_microrefuges_2016].
-By altering mammals' movement speed and habitat use [jointly, their movement behaviour: @nathan_movement_2008; @nathan_big-data_2022], ambient temperature also affects the consequences of movement behaviour [@alston_temperature_2020; @giroux_activity_2023], such as encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, or competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024].
-Behavioural changes to temperature are exacerbated in extreme cold [@wilson_stenotherms_2001; @berger_climate_2018] and extreme heat [@powers_hovering_2017; @alston_temperature_2020; @giroux_activity_2023; @verzuh_behavioural_2023].
+Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @merkle_large_2016; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020].
+How and where animals move thus determines encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, and competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024].
+However, ambient temperature affects animals' movement rates by altering the energetic cost of movement and thermoregulation [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], and extreme temperatures can reduce movement rates [@dyer_travel_2023] and even cause physiological damage and death [@hetem_responses_2014; @ratnayake_forecasting_2019; @schmidt_interplay_2020].
+Animals may search for a short-term buffer from extreme heat or cold by moving to thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], and shelter may even be preferred over forage abundance [@hall_microrefuges_2016].
+Changes in climate throughout the 21^st^ century and the simultaneous increase in frequency and intensity of extreme-heat events [@ipcc_climate_2023; @bunnell_global_2011] are expected to impact not only animals' movement behavior but also their physiology [@powers_hovering_2017; @schwerdt_ecological_2024], life history [@newediuk_climate_2024], and phenology [@carbeck_adaptation_2022; @johnson_human_2018].
 
@@
 While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more severe physiological damage in a shorter span of time [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021; @newediuk_climate_2024].
-Due to mammals' difficulty at dissipating heat, particularly when ambient temperatures are near or above their body temperature, they are particularly sensitive to extreme heat, especially with increasing body size [@dyer_travel_2023; @verzuh_behavioural_2023].
-Furthermore, species that experience narrow temperature ranges (i.e., $`$temperature specialists') tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while species that experience wide ranges of temperature variation (i.e., $`$temperature generalists') respond more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024].
-Still, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019; @rantanen_arctic_2022], affect many factors at once [e.g., temperature, snow depth, ecosystem phenology, and resource availability: @polazzo_measuring_2024], or are too rapid and severe to respond to within a lifetime or a few generations [@levins_evolution_1974; @gienapp_climate_2008; @fuller_towards_2016; @xu_plasticity_2021].
+Due to mammals' difficulty at dissipating heat, particularly when ambient temperatures are near or above their body temperature, they are particularly sensitive to extreme heat, especially with increasing body size.
+Large-bodied mammals are most likely to show changes in movement behaviour with rising temperatures, since they tend to move long distances [@merkle_large_2016], are particularly sensitive to overheating [@fuller_towards_2016; @verzuh_behavioural_2023], and their movement rates are thus strongly limited by excessive heat [@dyer_travel_2023].<!-- Furthermore, species that experience narrow temperature ranges (i.e., $`$temperature specialists') tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while species that experience wide ranges of temperature variation (i.e., $`$temperature generalists') respond more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024].
+Still, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019; @rantanen_arctic_2022], affect many factors at once [e.g., temperature, snow depth, ecosystem phenology, and resource availability: @polazzo_measuring_2024], or are too rapid and severe to respond to within a lifetime or a few generations [@levins_evolution_1974; @gienapp_climate_2008; @fuller_towards_2016; @xu_plasticity_2021].-->
 
@@
 It remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially if populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018].
-Preparing for and responding to future changes require a detailed understanding of how temperature will affect how and where different mammalian species move [@cunningham_opportunity_2021].
+Preparing for and responding to future changes will require a detailed understanding of how temperature will affect how and where different mammalian species move [@cunningham_opportunity_2021].
 Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], thermoregulation [@mota-rojas_physiological_2021], and movement behaviour [@mccain_assessing_2019; @melin_effect_2023; @cunningham_opportunity_2021].
@@
 
-<!-- zoom in to western canada -->
+<!-- introduce the study -->
 
-Western Canada is currently experiencing accelerating and widespread but spatially heterogeneous warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme events [@zhang_increased_2023].
-As we approach the deadline for the Kunming-Montreal Global Biodiversity Framework (i.e., the "30 by 30" initiative), a global effort to conserve 30% of the world's lands and oceans by 2030 [@convention_on_biological_diversity_154_2022], Canadian decision makers must understand which areas will be of greatest value for maximizing and protecting biodiversity not only for current conditions but future decades, too.
+***CLEAN UP PARAGRAPH BELOW***
+
+In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou: *Rangifer tarandus*, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves: *Canis lupus*; Table \@ref(tab:data-table)).
+Western Canada is currently experiencing accelerating and widespread but spatially heterogeneous warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme heat events [@zhang_increased_2023].
+Canadian decision makers must understand which areas will be of greatest value for maximizing and protecting biodiversity not only for current conditions but future decades, too.
 Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.
 Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022; @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].
-
-<!-- introduce the study -->
-
-Large-bodied mammals are most likely to show changes in movement behaviour with rising temperatures, since they tend to move long distances [@merkle_large_2016], are particularly sensitive to overheating [@fuller_towards_2016], and their movement rates are thus strongly limited by excessive heat [@dyer_travel_2023].
-Conveniently, their large body size also allows for larger tracking devices that have longer battery life and allow longer tracking periods and with finer temporal resolution [@kenward_manual_2001].
-In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou: *Rangifer tarandus*, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves: *Canis lupus*; Table \@ref(tab:data-table)).
-Using over 25 years of telemetry data throughout a large spatial range of western Canada and hourly air temperature data, we estimate how these mammals altered their probability of moving, speed when moving, and habitat selection in response to temperature.
-We then pair the estimated responses with monthly climate change projections [@mahony_global_2022] to forecast each species' movement behaviour during the 21^st^ century under four different climate-change scenarios, namely Shared Socioeconomic Pathways (SSPs): 1-2.6 ("Sustainability -- Taking the Green Road"; low challenges to mitigation and adaptation), 2-4.5 ("Middle of the Road: medium challenges to mitigation and adaptation), 3-7.0 ("Regional Rivalry -- A Rocky Road"; high challenges to mitigation and adaptation), and 5-8.5 ("Fossil-fueled Development -- Taking the Highway"; high challenges to mitigation, low challenges to adaptation).
-@riahi_shared_2017 provide a detailed overview of SSPs.
-We discuss the consequences of these changes in movement behaviour at the individual, population, and community levels.
-We also provide directions for future research and solutions for overcoming some common data limitations.
+We discuss the consequences of our findings at the individual, population, and community levels, and we also provide directions for future research and solutions for overcoming some common data limitations.
 
@@
 
-This study leverages four main datasets: (1) a multi-species collection of GPS telemetry data (Fig.
-\@ref(fig:bc-map)); (2) historical hourly air temperature from the ERA5 reanalyzed dataset [European Center for Medium-range Weather Forecasting$;$ @hersbach_era5_2023]; (3) temporally static rasters of percent forest cover, elevation, and distance from water; and (4) climate change projections under four Shared Socioeconomic Pathways [SSPs$;$ @riahi_shared_2017].
-While we acknowledge there are other important variables (e.g., forest type, structure, and age; terrain slope and aspect), we limited our models to only use three simple variables for the sake of parsimony and comparability across species.
-The simplicity of the models also allowed for simple forecasts under the assumption of temporally invariant landscapes.
+In this study, we estimate how temperature affected mammals' probability of moving, speed when moving, and habitat selection by leveraging three datasets: (1) a multi-species collection of 25 years of GPS telemetry data throughout a large spatial range of western Canada (Fig.
+\@ref(fig:bc-map)), (2) historical, hourly air temperature data from the ERA5 reanalyzed dataset [@hersbach_era5_2023], and (3) temporally static rasters of percent forest cover, elevation, and distance from water.
+We then forecast each species' movement behaviour thoughout the 21^st^ century using monthly climate change projections under four Shared Socioeconomic Pathways [SSPs$;$ @riahi_shared_2017; @mahony_global_2022]: SSP 1-2.6 ("Sustainability -- Taking the Green Road"; low challenges to mitigation and adaptation), SSP 2-4.5 ("Middle of the Road: medium challenges to mitigation and adaptation), SSP 3-7.0 ("Regional Rivalry -- A Rocky Road"; high challenges to mitigation and adaptation), and SSP 5-8.5 ("Fossil-fueled Development -- Taking the Highway"; high challenges to mitigation, low challenges to adaptation).
+
+We acknowledge that our models of habitat selection ignore many important variables of habitat selection (e.g., forest type, structure, and age; terrain slope and aspect; prey availability; population density).
+We limited our models to only use three simple and temporally static rasters to produce simple and easily interpretable estimates and forecasts under the assumption of temporally invariant landscapes.
 See the discussion section for additional detail on the advantages and limitations of using such simple models.
@@
 \noindent Boreal caribou and wolf telemetries were acquired from a data repository of the British Columbia Oil and Gas Research and Innovation Society (BC ORGIS) that was collected as part of the boreal caribou radio-collaring program of the BC OGRIS Research and Effectiveness Monitoring Board [REMB$;$ @bc_ogris_boreal_2018].
-Southern mountain caribou data ***XXX***.
+Southern mountain caribou data were obtained from the work of @ford_effects_2023.
 Cougar data include telemetries from @darlington_southern_2025 and @serrouya_ref.
@@
 
-\noindent Rasters of hourly reanalyzed air temperature data were downloaded from the ERA5 database [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)).
+\noindent Rasters of hourly air temperature data were downloaded from the ERA5 database [@hersbach_era5_2023] from the European Center for Medium-range Weather Forecasting server (ECMWF; [www.ecmwf.int](www.ecmwf.int); [https://cds.climate.copernicus.eu](https://cds.climate.copernicus.eu)).
 Proximal air temperature was estimated for each location by extracting the value from the corresponding raster cell from the temporally nearest raster using the `extract()` function from the `terra` package [v.
@@
 
-\noindent We estimated the effects of temperature on each species' selection for percent forest cover (`forest_perc`), elevation (`elevation_m`, in meters), and distance from water (`dist_water_m`, in meters) by fitting a Hierarchical Resource Selection Function (HRSF) for each species.
+\noindent We estimated the effects of temperature on each species' selection for percent forest cover (`forest_perc`), elevation (`elevation_m`, in meters), and distance from water (`dist_water_m`, in meters) by fitting a Hierarchical Resource Selection Function (HRSF) for each species [@mccabe_resource_2021].
 We fit each HRSF using an HGAM with a Poisson family of distributions and log link function [Appendix B$;$ @aarts_estimating_2008].
@@
 
-Changes in mammals' seasonal movement behaviour are also likely to affect the timing and frequency of human-wildlife interactions.
+Changes in mammals' seasonal movement behaviour are also likely to affect the timing and frequency of human-wildlife interactions [@abrahms_climate_2023; @weststrate_how_2024].
 Since heat affects mammals' sleep quality and likelihood to enter torpor [@mortlock_sleep_2024; @fjelldal_nightly_2021; @wells_life_2022], future decades may see increased levels of incidents with bears that struggle to enter or remain in hibernation [@johnson_human_2018].
@@
 Hierarchical Generalized Additive Models were crucial in producing such results.
-The hierarchical approach us to produce statistically sound population-level estimates by accounting for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mortlock_sleep_2024; @jeltsch_need_2025; @silva_too_2025], while the flexibility of GAMs estimated mammals' changes in movement behaviour without imposing rigid assumptions about the shape of the responses [@simpson_method_2025; @simpson_modelling_2018].
+The hierarchical approach us to produce statistically sound population-level estimates by accounting for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mccabe_resource_2021; @mortlock_sleep_2024; @jeltsch_need_2025; @silva_too_2025], while the flexibility of GAMs estimated mammals' changes in movement behaviour without imposing rigid assumptions about the shape of the responses [@simpson_method_2025; @simpson_modelling_2018].
 Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously by inexperienced analysts.
@@
 
-We have presented a multi-species analysis of the effects of temperature on mammalian movement rates and habitat selection in hopes that other researchers can leverage this framework and models to expand our understanding of how temperature affects the movement behavior of other species, including smaller mammals, more elusive species, and non-mammal animals.
+We have presented a multi-species analysis of the effects of temperature on mammalian movement rates and habitat selection in hopes that other researchers can leverage this framework and models to expand our understanding of how temperature affects the movement behaviour of other species, including smaller mammals, more elusive species, and non-mammal animals.
 

### 2025-09-29 writing/manuscript.Rmd
reviewed intro and abstract

@@
 
-<!--
-six-sentence abstract:
-
-1. introduce the topic
-2. state the problem
-3. summarize why the question hasn't been answered/what the consequences are
-4. explain how you answered the question
-5. explain the central findings
-6. explain the importance of your findings and their consequences
--->
-
 ```{r abstract, echo=FALSE}
-abstract <- "Widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness, habitat selection, and community structure.
+abstract <- "Widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness and community dynamics.
 Previous studies have estimated the effects of temperature on mammalian movement behaviour, but few disentangled them from seasonal behaviour cycles.
 Consequently, it is still uncertain how mammals will adapt their movement behaviour throughout the 21^st^ century.
-We address this gap by quantifying the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species throughout western Canada between 1998 and 2023.
-We show that temperature significantly affects of how and where these mammals move, and we predict that climate change will have significant impacts on mammals' movement behaviour throughout the 21^st^ century.
-Notably, we found that species did not respond uniformly to temperature.
-This suggests that community-level responses will be complex, as future climate change drives some species to move more and others less.
-Median habitat selection strength was predicted to decrease for most species, but some species are expected to show increased selection strength for higher altitudes.
-Consequently, our results suggest that individuals from all six species may shift their range in the coming decades as they begin to select against portions of their current ranges.
-The strong but varied responses to temperature suggest that future climate change will impact not only individuals' movement behaviour but also entire communities, as changes in movement behaviour will affect encounter rates, including predator prey-dynamics and human-wildlife interactions.
-As climate change exposes mammals to novel environmental conditions, proactively anticipating changes in mammalian movement behaviour will become crucial for effectively and proactively understanding community-level responses and selecting high-quality habitat for long-term conservation."
+We address this gap by quantifying the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species (boreal and southern mountain caribou, cougars, elk, grizzly bears, mountain goats, and wolves) throughout western Canada between 1998 and 2023.
+We show that temperature significantly affected of how and where these mammals moved, and we predict that climate change will impact mammals' movement behaviour throughout the 21^st^ century, although species did not respond uniformly to temperature.
+Projected responses to climate change suggested that rising temperatures will drive southern mountain caribou and mountain goats to move more, while cougars, elk, and wolves will move less.
+Boreal caribou and grizzly bears showed little change in projected yearly movement rates but clear shifts in seasonal phenology.
+We also predict that rising temperatures will reduce median habitat selection strength most of the six species, but some are expected to show increased selection strength for higher altitudes.
+As mammals increasingly alter their movement rates and select against portions of their current ranges, changes in individuals' movement behaviour will impact encounter rates, including predator-prey dynamics and human-wildlife interactions.
+Conservation efforts should therefore account for future changes in movement behaviour as well as the consequences such changes may have on communities.
+Proactively anticipating changes in mammalian movement behaviour will become crucial for effectively and proactively understanding community-level responses and selecting high-quality habitat for long-term conservation."
 if(FALSE){
@@
 
+\noindent \textbf{Running Title}: Temperature alters mammal movement behavior
+
 \noindent \textbf{Article type}: Research article
@@
 
-\noindent \textbf{Figures}: 8
+\noindent \textbf{Figures}: 9
 
@@
 
-\noindent \textbf{References}: 165 (updated on 2025-09-25) <!-- counted by using movement ecology csl -->
+\noindent \textbf{References}: 163 (updated on 2025-09-29) <!-- counted using movement ecology csl -->
 
@@
 
-<!-- using "behaviOR" in key words to increase audience -->
+\noindent \textbf{Key words:} climate change, temperature, mammals, animal movement, movement behaviour, habitat selection, projections
 
-\noindent \textbf{Key words:} climate change, temperature, mammals, animal movement, movement behavior, habitat selection
+\noindent \textbf{Contact Information:} *Corresponding author’s telephone, and email details*
 
@@
 Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @merkle_large_2016; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020].
-How and where animals move thus determines encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, and competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024].
-However, ambient temperature affects animals' movement rates by altering the energetic cost of movement and thermoregulation [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], and extreme temperatures can reduce movement rates [@dyer_travel_2023] and even cause physiological damage and death [@hetem_responses_2014; @ratnayake_forecasting_2019; @schmidt_interplay_2020].
-Animals may search for a short-term buffer from extreme heat or cold by moving to thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], and shelter may even be preferred over forage abundance [@hall_microrefuges_2016].
-Changes in climate throughout the 21^st^ century and the simultaneous increase in frequency and intensity of extreme-heat events [@ipcc_climate_2023; @bunnell_global_2011] are expected to impact not only animals' movement behavior but also their physiology [@powers_hovering_2017; @schwerdt_ecological_2024], life history [@newediuk_climate_2024], and phenology [@carbeck_adaptation_2022; @johnson_human_2018].
+How and where animals move thus also determines encounter rates with resources [@mezzini_how_2025], humans [@johnson_human_2018; @weststrate_how_2024], predators, prey, and competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024].
+However, ambient temperature affects animals' movement rates by altering the energetic cost of movement and thermoregulation [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], and extreme temperatures can impact animals' movement rates [@dyer_travel_2023], phenology [@carbeck_adaptation_2022; @johnson_human_2018], and life history [@newediuk_climate_2024], and even cause physiological damage and death [@hetem_responses_2014; @powers_hovering_2017; @ratnayake_forecasting_2019; @schmidt_interplay_2020; @schwerdt_ecological_2024].
+Animals may search for a short-term buffer from extreme temperatures by moving to thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], and shelter may even be preferred over forage abundance [@hall_microrefuges_2016].
+However, climate change during the last century [@ipcc_climate_2023] have caused animals to alter how and where they move as they respond to hotter temperatures and increasingly severe and frequent extreme-heat events [@ipcc_climate_2023; @bunnell_global_2011].
 
@@
 While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more severe physiological damage in a shorter span of time [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021; @newediuk_climate_2024].
-Due to mammals' difficulty at dissipating heat, particularly when ambient temperatures are near or above their body temperature, they are particularly sensitive to extreme heat, especially with increasing body size.
-Large-bodied mammals are most likely to show changes in movement behaviour with rising temperatures, since they tend to move long distances [@merkle_large_2016], are particularly sensitive to overheating [@fuller_towards_2016; @verzuh_behavioural_2023], and their movement rates are thus strongly limited by excessive heat [@dyer_travel_2023].<!-- Furthermore, species that experience narrow temperature ranges (i.e., $`$temperature specialists') tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while species that experience wide ranges of temperature variation (i.e., $`$temperature generalists') respond more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024].
-Still, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019; @rantanen_arctic_2022], affect many factors at once [e.g., temperature, snow depth, ecosystem phenology, and resource availability: @polazzo_measuring_2024], or are too rapid and severe to respond to within a lifetime or a few generations [@levins_evolution_1974; @gienapp_climate_2008; @fuller_towards_2016; @xu_plasticity_2021].-->
+Mammals' homeothermy can cause them to have difficulty dissipating heat, particularly when ambient temperatures are near or above body temperature.
+Large-bodied mammals are most likely to alter their movement behaviour with rising temperatures, since they tend to move long distances [@merkle_large_2016], are particularly sensitive to overheating [@fuller_towards_2016; @alston_temperature_2020; @verzuh_behavioural_2023], and their movement rates are thus strongly limited by excessive heat [@dyer_travel_2023].
+Global warming during the last century has resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extreme-heat events [@ipcc_climate_2023], with widespread changes in mammals' seasonal phenology, fitness, and life history [@botero_evolutionary_2015; @wells_life_2022; @newediuk_climate_2024].
+However, behavioral responses to heat vary substantially across species depending on body size, vagility, and activity times [@mccain_body_2014; @williams_range_2018].
+For example, warming temperatures have allowed white-tailed deer (*Odocoileus virginianus*) to shift northward, which has led to increased density of wolves (*Canis lupus*) and higher predation on caribou (*Rangifer tarandus caribou*) [@latham_invading_2011; @barber_potential_2018; @dickie_habitat_2024].
+Preparing for and responding to future changes will require a detailed understanding of not only how temperature affects how and where mammals move but also the cascade of effects that such changes will have at the population and community levels [@cunningham_opportunity_2021].
 
-<!-- we do not know how temperature will change -->
+<!-- responding to future changes -->
 
-In the last century, many ecosystems experienced widespread warming that resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extremely high temperatures [@ipcc_climate_2023].
-As heat stress intensifies over the next 100 years, mammals will continue to suffer impacts on their fitness, movement rates, and habitat selection [@deb_modelling_2020; @woo-durand_increasing_2020], which will be exacerbated by more frequent and intense extreme events [@bunnell_global_2011] and anthropogenic habitat loss, modification, and fragmentation [@sih_evolution_2011; @sawyer_identifying_2009; @weststrate_how_2024].
-It remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially if populations fail to adapt to changes [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @williams_range_2018].
-Preparing for and responding to future changes will require a detailed understanding of how temperature will affect how and where different mammalian species move [@cunningham_opportunity_2021].
-Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], thermoregulation [@mota-rojas_physiological_2021], and movement behaviour [@mccain_assessing_2019; @melin_effect_2023; @cunningham_opportunity_2021].
-However, there remains a need for fine-scale estimates of the effects of temperature on mammals' movement behaviour and quantitative predictions of changes in movement behaviour with climate change in the current century [@ipcc_climate_2023].
+As heat stress intensifies over the next 100 years, mammals will continue to suffer impacts on their fitness, movement rates, and habitat selection [@deb_modelling_2020; @woo-durand_increasing_2020].
+It remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially if populations fail to adapt [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @botero_evolutionary_2015; @williams_range_2018].
+Recent work has documented the effects of climate change on mammals' ranges [@leclerc_determinants_2021], thermoregulation [@mota-rojas_physiological_2021], and movement behaviour [@mccain_assessing_2019; @melin_effect_2023; @cunningham_opportunity_2021], but there remains a need for fine-scale estimates of the effects of temperature on mammals' movement behaviour and quantitative predictions of future changes in movement behaviour.
 
@@
 
-***CLEAN UP PARAGRAPH BELOW***
-
-In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou: *Rangifer tarandus*, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves: *Canis lupus*; Table \@ref(tab:data-table)).
+In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves: *Canis lupus*; Table \@ref(tab:data-table)).
 Western Canada is currently experiencing accelerating and widespread but spatially heterogeneous warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme heat events [@zhang_increased_2023].
-Canadian decision makers must understand which areas will be of greatest value for maximizing and protecting biodiversity not only for current conditions but future decades, too.
+Decision makers must understand which areas will best sustain and protect biodiversity not only for current conditions but future decades, too.
 Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.
@@
 
+Points to add:
+
+* thermal landscape?
+* comon trends:
+  * boreal species do not see much of a change in distance traveled, but this is in part due to opposite trends in the different seasons.
+  * habitat selection for boreal species flattens with temperatures above 0 but strengthens near zero.
+Therefore, warmer summers may result in lower encounter rates as ranges widen, but warmer winters may result in increased predation risk, which may impact reproductive success (since cows are pregrant in winter)
+  * SM caribou and mountain goats are projected to move more and see widespread decreases in their current habitats
+
 <!-- to do: address any questions presented in the introduction -->

### 2025-09-29 writing/manuscript.Rmd
addressed minor edits throughout

@@
 
-We used the the `climatenaR` package [v.
-1.0, @burnett_climatenar_2023] for `R` and ClimateNA [v.
-7.4.2$;$ @wang_locally_2016; @mahony_global_2022] to obtain projected average monthly temperatures from 2025 to 2100 at a 0.08$^\circ$ spatial resolution.
-Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the historical data for western Canada from 1998 to 2023 (extremes included) to estimate the monthly variation in temperature, which we defined as $\sigma^2_T$, the variance at location $\langle x, y \rangle$ in month $m$ for each year (e.g., the variance at coordinates $\langle -119.40, 49.94 \rangle$ in April 2005).
-We then modeled the estimated variance using a Generalized Additive Model (GAM) for Location and Scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017] using the `mgcv` package for `R` [v.
+To obtain projected average monthly temperatures from 2025 to 2100 at a 0.08$^\circ$ spatial resolution, we used the the `climatenaR` package [v.
+1.0, @burnett_climatenar_2023] for `R` and ClimateNA v.
+7.4.2 [@wang_locally_2016; @mahony_global_2022].
+However, the climate projections only included estimates of future monthly averages, a scale substantially coarser than that of our tracking data (Table \@ref(tab:data-table)).
+To estimate the distribtution of temperatures at an hourly scale within a month, we assumed values to be approximately normally distributed with mean $\mu_T$ and variance $\sigma^2_T$.
+We estimated $\mu_T$ using the ClimateNA projections, while we estimated $\sigma^2_T$ by modeling the observed variance in historical weather data for western Canada from 1998 to 2023 (inclusive).
+For each location $\langle x, y \rangle$ and month $m$ of each year (e.g., the observed variance at coordinates $\langle -119.40, 49.94 \rangle$ in April 2005), we calculated the variance in historical temperature data, for a total of $(2024 - 1998) \times 12 = 312$ observations per location.
+We then modeled $\sigma^2_T$ a Generalized Additive Model (GAM) for Location and Scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017] fitted with the `mgcv` package for `R` [v.
 1.9-1, @wood_generalized_2017].
@@
 We did not include smooth terms of year to avoid unrealistic projections when extrapolating beyond past 2023.
-The model was thus
+The complete model for the distribution of projected temperature was thus
 
@@
 \begin{cases}
+T \sim \text{Normal}(\mu_T, \sigma^2_T) \\
+\mu_T \approx \hat \mu_T \text{ as projected by ClimateNA} \\
 \sigma^2_T \sim \text{Gamma}(\mu_{\sigma^2_T}, \nu_{\sigma^2_T}) \\
@@
 
-### Resource rasters
+### Habitat rasters
 
@@
 
-The two following sections detail the statistical framework and models we used to estimate the effect on temperature on the six species' movement behaviour.
 To assess the importance of including temperature as an explicit covariate (as ooposed to including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the fits of the two sets of models via analyses of deviance (a form of generalized likelihood ratio tests) following the methods of Section 3.3 in @wood_generalized_2017.
@@
 After removing non-resident individuals (Table B1), we accounted for the spatiotemporal autocorrelation in the telemetry locations by weighting each point based on the telemetry's Autocorrelated Kernel Density Estimate [@fleming_new_2017; @noonan_comprehensive_2019; @alston_mitigating_2022] to produce estimates of second-order habitat selection [@johnson_comparison_1980].
-Quadrature points were determined using the raster cells in each animal's 99.9% AKDE percentile and given a weight of 1.
+Quadrature points were used to approximate the likelihood function of a Poisson point process through Monte-Carlo Markov chain-based integration [@baddeley_spatial_2015; @fithian_finite-sample_2013; @aarts_estimating_2008] and were determined using the raster cells in each animal's 99.9% AKDE percentile, with each raster cell having a weight of 1.
 The number of quadrature locations greatly outnumbered the number of observed locations (Fig.
@@
 The models had good in-sample prediction (Fig.
-B7) and explained reasonably high proportions of the deviance [79.3% for the gamma model and 10.7% for the binomial model, which is relatively high for a binomial model with binary responses: see ch.
-11 of @mcelreath_statistical_2020].
+B7) and explained reasonably high proportions of the deviance [79.3% for the gamma model and 10.7% for the binomial model, which is relatively high for a binomial model with binary responses].
 All species altered their daily and seasonal movement behaviour to changes in temperature (Fig.
@@
 The recent increase in frequency and duration of extreme-temperature events [@yao_emergence_2022; @ipcc_climate_2023; @zhang_increased_2023; @bunnell_global_2011; @kienzle_has_2018] have forced many mammals towards increasingly atypical behaviours with long-term consequences, ranging from changes in the behaviour and fitness of individuals [@brivio_seeking_2024; @cunningham_opportunity_2021] to mass die-offs [@ameca_y_juarez_natural_2012].
-A species' ability to respond to changing temperature and phenology depends mainly on the its exposure, sensitivity, and adaptability to changes, and growing evidence suggests that many mammals in western Canada will struggle to respond due limitations in all three factors [@ameca_y_juarez_natural_2012; @diffenbaugh_changes_2013].
+A species' ability to respond to changing temperature and phenology depends mainly on the its exposure, sensitivity, and adaptability to changes, and growing evidence suggests that many mammals in western Canada will struggle to respond due to limitations in all three factors [@ameca_y_juarez_natural_2012; @diffenbaugh_changes_2013].
 Firstly, exposure to warming temperatures is widespread throughout the area, and warming is expected to accelerate over the coming decades [@ipcc_climate_2023; @diffenbaugh_changes_2013; @bunnell_global_2011; @kienzle_has_2018; @eum_potential_2023].
@@
 Consequently, mammals that move to higher elevations may be forced to spend more time foraging [@mezzini_how_2025], which will increase their energetic expenditure as well as their chances of encountering predators (including cougars and grizzly bears) or competitors [@kohl_prey_2019; @martinez-garcia_how_2020; @torrez-herrera_monkeys_2020; @smith_densitydependent_2023].
-Additionally, plants at high elevations may be have weaker defenses against herbivory and be more susceptible to overgrazing [@callis-duehl_community-level_2017].
+Additionally, plants at high elevations may have weaker defenses against herbivory and be more susceptible to overgrazing [@callis-duehl_community-level_2017].
 The scarcity and susceptibility of forage may thus limit herbivores' ability to shift elevationally, particularly if predation rates remain relatively unchanged or increase.
@@
 The hierarchical approach us to produce statistically sound population-level estimates by accounting for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mccabe_resource_2021; @mortlock_sleep_2024; @jeltsch_need_2025; @silva_too_2025], while the flexibility of GAMs estimated mammals' changes in movement behaviour without imposing rigid assumptions about the shape of the responses [@simpson_method_2025; @simpson_modelling_2018].
-Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously by inexperienced analysts.
+Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously.
 We suggest choosing the basis size (`k`) for each smooth term carefully.

### 2025-09-30 writing/manuscript.Rmd
minor edits throughout; streamlined discussion

@@
 
-\noindent \textbf{References}: 163 (updated on 2025-09-29) <!-- counted using movement ecology csl -->
+\noindent \textbf{References}: 147 (updated on 2025-09-30) <!-- counted using movement ecology csl -->
 
@@
 How and where animals move thus also determines encounter rates with resources [@mezzini_how_2025], humans [@johnson_human_2018; @weststrate_how_2024], predators, prey, and competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024].
-However, ambient temperature affects animals' movement rates by altering the energetic cost of movement and thermoregulation [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], and extreme temperatures can impact animals' movement rates [@dyer_travel_2023], phenology [@carbeck_adaptation_2022; @johnson_human_2018], and life history [@newediuk_climate_2024], and even cause physiological damage and death [@hetem_responses_2014; @powers_hovering_2017; @ratnayake_forecasting_2019; @schmidt_interplay_2020; @schwerdt_ecological_2024].
+However, ambient temperature affects animals' movement rates by altering the energetic cost of movement and thermoregulation [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], and extreme temperatures can impact animals' movement rates [@dyer_travel_2023], phenology [@carbeck_adaptation_2022; @johnson_human_2018; @mclellan_effect_2015; @pigeon_drivers_2016], and life history [@newediuk_climate_2024], and even cause physiological damage and death [@hetem_responses_2014; @powers_hovering_2017; @ratnayake_forecasting_2019; @schmidt_interplay_2020; @schwerdt_ecological_2024].
 Animals may search for a short-term buffer from extreme temperatures by moving to thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], and shelter may even be preferred over forage abundance [@hall_microrefuges_2016].
@@
 Large-bodied mammals are most likely to alter their movement behaviour with rising temperatures, since they tend to move long distances [@merkle_large_2016], are particularly sensitive to overheating [@fuller_towards_2016; @alston_temperature_2020; @verzuh_behavioural_2023], and their movement rates are thus strongly limited by excessive heat [@dyer_travel_2023].
-Global warming during the last century has resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extreme-heat events [@ipcc_climate_2023], with widespread changes in mammals' seasonal phenology, fitness, and life history [@botero_evolutionary_2015; @wells_life_2022; @newediuk_climate_2024].
+Global warming during the last century has resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extreme-heat events [@ipcc_climate_2023], with widespread changes in mammals' seasonal phenology, fitness, and life history [@botero_evolutionary_2015; @wells_life_2022; @newediuk_climate_2024; @mclellan_effect_2015; @pigeon_drivers_2016].
 However, behavioral responses to heat vary substantially across species depending on body size, vagility, and activity times [@mccain_body_2014; @williams_range_2018].
-For example, warming temperatures have allowed white-tailed deer (*Odocoileus virginianus*) to shift northward, which has led to increased density of wolves (*Canis lupus*) and higher predation on caribou (*Rangifer tarandus caribou*) [@latham_invading_2011; @barber_potential_2018; @dickie_habitat_2024].
-Preparing for and responding to future changes will require a detailed understanding of not only how temperature affects how and where mammals move but also the cascade of effects that such changes will have at the population and community levels [@cunningham_opportunity_2021].
+For example, warming temperatures have allowed deer (*Odocoileus* spp.) to shift northward, which has led to increased density of wolves (*Canis lupus*) and higher predation on caribou (*Rangifer tarandus caribou*) [@latham_invading_2011; @barber_potential_2018; @dickie_habitat_2024].
+Preparing for and responding to future changes will require a detailed understanding of not only how how and where mammals move throughout the thermal landscape but also the cascade of effects that such changes will have at the population and community levels [@cunningham_opportunity_2021].
 
@@
 \@ref(fig:bc-map)), (2) historical, hourly air temperature data from the ERA5 reanalyzed dataset [@hersbach_era5_2023], and (3) temporally static rasters of percent forest cover, elevation, and distance from water.
-We then forecast each species' movement behaviour thoughout the 21^st^ century using monthly climate change projections under four Shared Socioeconomic Pathways [SSPs$;$ @riahi_shared_2017; @mahony_global_2022]: SSP 1-2.6 ("Sustainability -- Taking the Green Road"; low challenges to mitigation and adaptation), SSP 2-4.5 ("Middle of the Road: medium challenges to mitigation and adaptation), SSP 3-7.0 ("Regional Rivalry -- A Rocky Road"; high challenges to mitigation and adaptation), and SSP 5-8.5 ("Fossil-fueled Development -- Taking the Highway"; high challenges to mitigation, low challenges to adaptation).
+We then forecast each species' movement behaviour throughout the 21^st^ century using monthly climate change projections under four Shared Socioeconomic Pathways [SSPs$;$ @riahi_shared_2017; @mahony_global_2022]: SSP 1-2.6 ("Sustainability -- Taking the Green Road"; low challenges to mitigation and adaptation), SSP 2-4.5 ("Middle of the Road: medium challenges to mitigation and adaptation), SSP 3-7.0 ("Regional Rivalry -- A Rocky Road"; high challenges to mitigation and adaptation), and SSP 5-8.5 ("Fossil-fueled Development -- Taking the Highway"; high challenges to mitigation, low challenges to adaptation).
 
@@
 However, the climate projections only included estimates of future monthly averages, a scale substantially coarser than that of our tracking data (Table \@ref(tab:data-table)).
-To estimate the distribtution of temperatures at an hourly scale within a month, we assumed values to be approximately normally distributed with mean $\mu_T$ and variance $\sigma^2_T$.
+To estimate the distribution of temperatures at an hourly scale within a month, we assumed values to be approximately normally distributed with mean $\mu_T$ and variance $\sigma^2_T$.
 We estimated $\mu_T$ using the ClimateNA projections, while we estimated $\sigma^2_T$ by modeling the observed variance in historical weather data for western Canada from 1998 to 2023 (inclusive).
@@
 
-To assess the importance of including temperature as an explicit covariate (as ooposed to including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the fits of the two sets of models via analyses of deviance (a form of generalized likelihood ratio tests) following the methods of Section 3.3 in @wood_generalized_2017.
+To assess the importance of including temperature as an explicit covariate (as opposed to including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the fits of the two sets of models via analyses of deviance (a form of generalized likelihood ratio tests) following the methods of Section 3.3 in @wood_generalized_2017.
 See Appendix B for additional information.
@@
 The models also had three tensor product interaction terms `by` each species: (1) day of year and time of day, (2) temperature and time of day, and (3) temperature and day of year.
-These three terms accounted for smooth changes in: (1) daily behaviour across day of year, (2) the response to temperature over time of day (e.g., changes in nocturnality), and (3) the response to temperature over day of year (e.g., the timing of molting, migration, and hybernation).
+These three terms accounted for smooth changes in: (1) daily behaviour across day of year, (2) the response to temperature over time of day (e.g., changes in nocturnality), and (3) the response to temperature over day of year (e.g., the timing of molting, migration, and hibernation).
 Finally, two smooth terms of log-transformed sampling interval (`dt`; hours) corrected for biases in speed estimates arising from irregular GPS sampling intervals, since longer intervals result in lower speed estimates [@nathan_big-data_2022; @denicola_are_2025].
@@
 B9 and B10).
-Estimated changes in average distance travaled were negligible for boreal caribou, and grizzly bears, although both species showed seasonal changes in seasonal movement rates.
+Estimated changes in average distance traveled were negligible for boreal caribou, and grizzly bears, although both species showed seasonal changes in seasonal movement rates.
 As temperatures warmed, boreal caribou moved further in winter, spring, and fall but less in summer, while grizzly bears showed a clear shift towards earlier emergence from hibernation (Fig.
@@
 
-Points to add:
-
-* thermal landscape?
-* comon trends:
-  * boreal species do not see much of a change in distance traveled, but this is in part due to opposite trends in the different seasons.
-  * habitat selection for boreal species flattens with temperatures above 0 but strengthens near zero.
-Therefore, warmer summers may result in lower encounter rates as ranges widen, but warmer winters may result in increased predation risk, which may impact reproductive success (since cows are pregrant in winter)
-  * SM caribou and mountain goats are projected to move more and see widespread decreases in their current habitats
-
-<!-- to do: address any questions presented in the introduction -->
-
 \noindent We have demonstrated that temperature is an important driver of how and where large mammals move, and that mammals alter their seasonal and daily cycles in movement behaviour in response to changes in temperature.
 However, predicting mammals' responses to climate change remains a challenge, as habitats warm rapidly and mammals are exposed to increasingly novel and extreme conditions [@ipcc_climate_2023; @diffenbaugh_changes_2013; @botero_evolutionary_2015].
-We leveraged the flexibility and interptetability of GAMs to estimate the (nonlinear) responses  of terrestrial mammals to temperature.
-In particular, tensor product interaction terms were crucial for estimating smooth, nonlinear changes in daily and seasonal trends with temperature.
+We leveraged the flexibility and interpretability of HGAMs to estimate mammals' nonlinear responses to temperature without imposing rigid assumptions about the shape of the responses [@simpson_method_2025; @simpson_modelling_2018].
+In particular, tensor product interaction terms were crucial for estimating smooth, nonlinear changes in daily and seasonal trends with temperature with data from multiple individuals.
 By allowing interaction terms to be smooth bivariate functions, we were able to quantify changes in movement behaviour without the need to discretize time into distinct windows (e.g., day/night and seasons).
-We were thus able to present changes using relatively simple and interpretable surface plots (Figs.
+We were thus able to present results using relatively simple and interpretable surface plots (Figs.
 \@ref(fig:dist) and \@ref(fig:hrsfs)) that conveyed detailed and nuanced estimates of change.
-This framework was essential for estimating how mammals behave in extreme conditions for which data are scarce.
-
-As temperatures warm, mammals' risk of heperthermia increases nonlinearly [@sherwood_adaptability_2010], especially for large-bodied species [@alston_temperature_2020; @dyer_travel_2023; @verzuh_behavioural_2023].
-The recent increase in frequency and duration of extreme-temperature events [@yao_emergence_2022; @ipcc_climate_2023; @zhang_increased_2023; @bunnell_global_2011; @kienzle_has_2018] have forced many mammals towards increasingly atypical behaviours with long-term consequences, ranging from changes in the behaviour and fitness of individuals [@brivio_seeking_2024; @cunningham_opportunity_2021] to mass die-offs [@ameca_y_juarez_natural_2012].
-A species' ability to respond to changing temperature and phenology depends mainly on the its exposure, sensitivity, and adaptability to changes, and growing evidence suggests that many mammals in western Canada will struggle to respond due to limitations in all three factors [@ameca_y_juarez_natural_2012; @diffenbaugh_changes_2013].
-Firstly, exposure to warming temperatures is widespread throughout the area, and warming is expected to accelerate over the coming decades [@ipcc_climate_2023; @diffenbaugh_changes_2013; @bunnell_global_2011; @kienzle_has_2018; @eum_potential_2023].
-Secondly, large-bodied mammals are particularly sensitive to high heat [@dyer_travel_2023], but small-bodied mammals are also affected [@ratnayake_forecasting_2019; @krebs_impact_2019].
-Finally, species' adaptability is tested by the speed of climate change [@diffenbaugh_changes_2013] and the rigidity of many species' photoperiod-dependent phenology [@walker_global_2019].
-
-<!-- complex response from each species -->
+This framework was essential for estimating how mammals behave at extreme temperatures.
+Despite the recent increase in frequency of extreme weather events in western Canada [@zhang_increased_2023; @bunnell_global_2011; @kienzle_has_2018], such events remain relatively rare, so estimating mammals' nonlinear responses required the flexibility provided by GAMs [@wood_generalized_2017].
+Additionally, hierarchical models allowed us to produce statistically sound estimates that accounted for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mccabe_resource_2021; @jeltsch_need_2025; @silva_too_2025].
+Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously, so we suggest choosing each smooth term's basis size (`k`) carefully.
+While penalized splines and REML can help avoid unruly smooth terms, the penalty terms can struggle to restrict model complexity when datasets are particularly large [@wood_generalized_2017; @wood_generalized_2017-1].
 
@@
 For example, hotter temperatures caused boreal caribou to travel longer distances in fall, winter, and spring but less in summer, mirroring the findings of @stien_congruent_2012, @leclerc_determinants_2021, and @lessard_local_2025.
-The increase in movement rates during the colder seasons are likely partly due to shallower snow depth that results in greater mobility [@pedersen_quantifying_2021], but rising temperatures during snowy seasons also increase the chances of rain-on-snow events, which limit forage availability and increase the ungulates' need to search for food [@stien_congruent_2012; @berger_climate_2018; @mezzini_how_2025].
-In contrast, the reduced movement rates during warmer summers are likely mainly due to caribou searching for shelter from the heat due to ungulates' elevated risk of hyperthermia [@alston_temperature_2020; @verzuh_behavioural_2023].
-The effects of extreme temperatures are exacerbated by phenological mismatches with seasonal photoperiod [@walker_global_2019], including the timing of molting and reproduction.
-Earlier growth seasons in boreal and arctic regions have resulted in lower calf birth and survival [@post_climate_2008], while the lengthening of the growth season has allowed moose (*Alces alces*) and deer (*Odocoileus* spp.) to encroach on boreal caribou habitat, along with increased density of coyotes (*Canis latrans*), cougars, and wolves [@barber_potential_2018; @demars_incorporating_2023].
+The increase in movement rates during the colder seasons are likely partly due to shallower snow depth that results in greater mobility [@pedersen_quantifying_2021], but rising temperatures during snowy seasons also increase the risk of rain-on-snow events, which limit forage availability and increase time spent foraging [@stien_congruent_2012; @berger_climate_2018; @mezzini_how_2025].
+In contrast, the reduced movement rates during warmer summers are likely mainly due to caribou's reduced movement rates and use of thermal refugia, given ungulates' elevated risk of hyperthermia [@alston_temperature_2020; @verzuh_behavioural_2023].
+Therefore, although we predict that boreal caribou's yearly average movement rates will remain approximately unchanged throughout the 21^st^ century, the projections do not show the changes in seasonal movement phenology.
+However, this is an important caveat, since the effects of extreme temperatures are exacerbated by phenological mismatches with seasonal photoperiod [@walker_global_2019], including the timing of molting and reproduction.
+Earlier growth seasons in boreal and arctic regions have resulted in lower calf birth and survival [@post_climate_2008], while the lengthening of the growth season has allowed moose (*Alces alces*) and deer to encroach on boreal caribou habitat and increase the density of coyotes (*Canis latrans*), cougars, and wolves [@barber_potential_2018; @demars_incorporating_2023], whose movement behavior also depends on ambient temperature.
 Thus, rising temperatures affect more than each individual species' movement behaviour; they have complex and interconnected effects on an entire community and its trophic structure.
 
-<!-- complex response from community -->
+Work by @cunningham_opportunity_2021 discusses the importance of adopting a community-level perspective when assessing the effects of climate change on animals' movement behavior and phenology.
+For example, our results show that wolves responded to temperature by altering their movement rates similarly to caribou.
+Consequently, as tempeatures warm and both species increase their movement rates in fall, winter, and spring, the resulting increase in encounter rates [@martinez-garcia_how_2020] will likely increase predation risk when caribou are pregnant or with young calves.
+Additionally, while both species saw reduced movement rates during hot summer days, boreal wolves' use of anthropogenic linear features (e.g., roads, seismic lines) may allow them to reduce movement costs while increasing the chances of encountering heat-stressed prey [@whittington_caribou_2011; @dickie_faster_2017; @dickie_resource_2022].
+Furthermore, caribou that attempt to reduce predation risk from wolves by avoiding wolf habitat may still risk increasing predation pressure from bears [@leblond_caribou_2016] and other encroaching predators [@demars_incorporating_2023; @labadie_global_2023].
 
-Our results suggest that rising temperatures will have similar effects on boreal wolves.
-Warmer temperatures caused wolves to travel further throughout the year, but, similarly to boreal caribou, they moved substantially less at temperatures above 25$^\circ$C in summer.
-At temperatures above 20$^\circ$, both boreal caribou and wolves tended to move less during the day and more between 17:00 and 24:00.
-This synchroniciy in movement patterns is likely to increase encounters between the two species, particularly when coupled with increasing human disturbances, since wolves leverage seismic lines and other anthropogenic linear features (e.g., roads) to reduce movement costs while increasing the chances of encountering prey [@whittington_caribou_2011; @dickie_faster_2017; @dickie_resource_2022].
-Additionally, caribou that attempt to reduce predation risk from wolves by avoiding wolf habitat may risk increasing predation pressure from bears [@leblond_caribou_2016] and other encroaching predators [@demars_incorporating_2023].
-As temperatures rise and allow new competitors and predators to encroach on boreal caribou habitat [@barber_potential_2018; @demars_incorporating_2023; @labadie_global_2023], climate change will have complex conseuqences not only for individual species but for entire communities and their trophic interactions.
+Considerations about changes in trophic interactions leads to another important caveat about the results we present.
+The estimated changes movement behaviour cannot be fully attributed to physiological responses to temperature alone, since other aspects of habitats' seasonal phenology are (nonlinearly) correlated with temperature.
+For example, mountain goats' lower movement rates at temperatures near 0$^\circ$C in winter may be in part due to increased movement difficulty, since melting snow and rain-on-snow events increase the energetic costs of movement [@white_population_2025] and encounter rates with predators [@sullender_defining_2023].
+However, milder temperatures can also increase plant growth, which allows goats to spend more time foraging within and less time travelling across patches [@charnov_optimal_1976; @white_mountain_2025].
+Therefore, figure \@ref(fig:dist) illustrates an estimate of the total effects of temperature on each species movement rates, but it does not explicitly account for energetic costs.
+Similarly, species' changes in habitat selection do not explicitly account for any changes in physiological or energetic costs.
+Additionally, forest cover, elevation, and distance from water are often correlated, so the estimated effects should be interpreted carefully, and they may not be applicable outside our study areas.
+For example, both mountain goats and elk selected for higher elevation as temperatures warmed, but our models did not explicitly account for differences in forage availability at different altitudes.
 
-<!-- focus on habitat selection using changes in elevational preference of elk, goats, and cougars as the main example -->
+Forage quality and availability is often a limiting factor for mammals' ability to adapt to warming temperatures by moving to higher altitudes, since plants are slower to arrive in and establish in new habitats than mammals, and the advance of herbivores can further limit plants' ability to establish [@speed_elevational_2012; @diffenbaugh_changes_2013].
+Consequently, mammals that move to higher elevations may be forced to spend more time foraging [@mezzini_how_2025], which will increase their energetic expenditure as well as their chances of encountering predators (including cougars and grizzly bears) or competitors [@kohl_prey_2019; @martinez-garcia_how_2020; @torrez-herrera_monkeys_2020; @smith_densitydependent_2023].
+If local peaks become unsuitable habitat, and moving to nearby higher peaks requires traveling across low-quality and high-risk habitat (Figure \@ref(fig:s-rss-cc)), habitat fragmentation may prevent movement across ranges [@white_mountain_2025].
+Translocating animals to locations with higher peaks may help avoid such a scenario, but the success of relocations is species-dependent, and attempts are not always successful [@ranc_memory_2022; @white_mountain_2025].
 
-Considerations about changes in trophic interactions leads to an important caveat about the results we present.
-The estimated effects of temperature on movement behaviour cannot be fully attributed to physiological responses to changes in temperature alone, since other aspects of habitats' seasonal phenology are (nonlinearly) correlated with temperature.
-For example, mountain goats' reduced movement rates at temperatures near 0$^\circ$C in winter may be due to increased movement difficulty, since melting snow and rain-on-snow events affect the energetic costs of movement [@white_population_2025] and encounter rates with predators [@sullender_defining_2023].
-However, milder temperatures can also increase plant growth, which, in turn, allows goats to spend less time searching for forage [@white_mountain_2025].
-Therefore, the figure \@ref(fig:dist) illustrates an estimate of the total effect of temperature on each species movement rates, but it does not explicitly account for energetic costs.
-Similarly, species' changes in habitat selection do not explicitly account for any changes in physiological or energetic costs.
-Both mountain goats and elk selected for higher elevation as temperatures warmed, but our models did not explicitly account for differences in forage availability at different altitudes.
+Future work should develop models that account for covariates that are specific to the study species and area, such as forest age and disturbance [@demars_incorporating_2023], terrain ruggedness [@white_mountain_2017], anthropogenic linear features [@dickie_faster_2017], and growing human presence [@sih_evolution_2011; @rice_conspicuous_2022] and risk of human-wildlife conflict [@abrahms_climate_2023; @weststrate_how_2024].
+In particular, forests in western Canada are highly dynamic over time [@zhang_half-century_2015], and fires size and burn severity have increased in western Canada in recent decades [@zurowski_summer_2023; @parisien_abrupt_2023].
+Thus, while our use of a static raster of forest cover allowed for simpler models and climate change projections, future studies should account for mammals' responses to temporal forest dynamics, including the impacts of fire [@zurowski_summer_2023; @parisien_abrupt_2023], harvest [@lochhead_linking_2022] and other extractive industries [@dickie_faster_2017; @dickie_habitat_2024], as well as the subsequent forest regeneration.
+However, the inclusion of temporally dynamic variables will require that any future projections of species' behavior depend on temporally dynamic estimates of how such variables will change over time with changes in climate and anthropogenic activity.
 
-Forage quality and availability is often a limiting factor for mammals' ability to adapt to warming temperatures by moving to higher altitudes.
-Plants are slower to arrive in and establish in new habitats than mammals, and the advance of herbivores can further limit plants' ability to establish [@speed_elevational_2012; @diffenbaugh_changes_2013].
-Consequently, mammals that move to higher elevations may be forced to spend more time foraging [@mezzini_how_2025], which will increase their energetic expenditure as well as their chances of encountering predators (including cougars and grizzly bears) or competitors [@kohl_prey_2019; @martinez-garcia_how_2020; @torrez-herrera_monkeys_2020; @smith_densitydependent_2023].
-Additionally, plants at high elevations may have weaker defenses against herbivory and be more susceptible to overgrazing [@callis-duehl_community-level_2017].
-The scarcity and susceptibility of forage may thus limit herbivores' ability to shift elevationally, particularly if predation rates remain relatively unchanged or increase.
-Furthermore, the physical upper limit to elevation can cause individuals to become trapped .
-If local peaks become unsuitable habitat, and moving to nearby higher peaks requires traveling across low-quality and high-risk habitat (Figure \@ref(fig:s-rss-cc)), habitat fragmentation may become too excessive to move across patches or alternative ranges [@white_mountain_2025].
-Translocating animals to locations with higher peaks may help avoid such a scenario, but the success of relocations can be highly species-dependent, and attempts are not always successful [@ranc_memory_2022; @white_mountain_2025].
-
-Changes in mammals' seasonal movement behaviour are also likely to affect the timing and frequency of human-wildlife interactions [@abrahms_climate_2023; @weststrate_how_2024].
-Since heat affects mammals' sleep quality and likelihood to enter torpor [@mortlock_sleep_2024; @fjelldal_nightly_2021; @wells_life_2022], future decades may see increased levels of incidents with bears that struggle to enter or remain in hibernation [@johnson_human_2018].
-The risk is further increased when bears' are unable to meet their increased energetic requirements through foraging or hunting.
-Projected increases in park attendance due to rising temperatures and human population density in Western Canada are expected to further increase frequency of human-wildlife interactions [@weststrate_how_2024].
-
-## Consequences for conservation
-
-Species' adaptability to rising temperatures is challenged by the aforementioned symultaneous, multifarious, and interconnected stressors caused by climate change, since responding to many concurrent changes at once is particularly difficult [@diffenbaugh_changes_2013; @polazzo_measuring_2024].
-Locations may thus not be able to support species and population sizes they supported historically, and species may thus not select for historical geograpical ranges in the future.
-It is therefore crucial for conservationists and other decision-makers to distinguish between a species' ideal environmental space [or E space, *sensu* @matthiopoulos_defining_2022], and its best available geographic space [or G space, *sensu* @matthiopoulos_defining_2022].
-Accelerating multufarious change in species' movement behaviour and energetic costs require conservation measures to be proactive and anticipate future changes, even when future conditions are substantially different from historical conditions and historical data may not be available.
-To this end, the framework presented by @polazzo_measuring_2024 for estimating species' responses to multifarious change becomes essential, particularly when accounting for energetic and fitness costs, as argued by @cunningham_opportunity_2021.
-
-As of December 2024, only 13.8% of terrestrial Canada was within protected areas or other conserved areas [@environment_and_climate_change_canada_canadian_2025], so rapid action is necessary to protect and additional 16.2% of Canada's land by 2030 in accordance with Kunming-Montreal Global Biodiversity Framework [i.e., the "30 by 30 initiative"$;$ @convention_on_biological_diversity_154_2022].
-However, the choice of what habitats to protect will require careful consideration, since protected areas should support species not only in present conditions but also as species' needs and selection change over the coming decades [@simmons_tough_2021].
-As anthropogenic activity increasingly alters habitats and community structures [@sih_evolution_2011; @tucker_moving_2018; @rosenthal_impact_2022; @rice_conspicuous_2022], protecting biodiversity and the habitats it depends on will require widespread and collaboration with local stakeholders, especially Indigenous Peoples, to ensure that all decisions are culturally sound and respect Treaties [@turner_its_2009; @desjardins_looking_2020; @wong_towards_2020; @hessami_indigenizing_2021; @falardeau_biophysical_2022; @lamb_braiding_2023].
-Understanding the consequences of climate change in mammalian movement behaviour is a first step towards proactively responding to how mammals will respond to human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018].
-However, a better understanding of such consequences will require a more holistic approach that includes careful data collection, model design, and interpretation and application of results.
-
-## Study strengths and limitations
-
-Our results demonstrate that mammals respond to changes in temperature by altering how and where they move in complex and nonlinear ways (Figs.
-\@ref(fig:dist) and \@ref(fig:hrsfs)).
-Hierarchical Generalized Additive Models were crucial in producing such results.
-The hierarchical approach us to produce statistically sound population-level estimates by accounting for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mccabe_resource_2021; @mortlock_sleep_2024; @jeltsch_need_2025; @silva_too_2025], while the flexibility of GAMs estimated mammals' changes in movement behaviour without imposing rigid assumptions about the shape of the responses [@simpson_method_2025; @simpson_modelling_2018].
-Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously.
-We suggest choosing the basis size (`k`) for each smooth term carefully.
-While penalized splines and REML can help avoid unruly smooth terms that would result in questionable estimates the penalty terms can struggle to restrict model complexity when datasets are particularly large [@wood_generalized_2017; @wood_generalized_2017-1].
-
-Despite the statistical soundness of the models, estimating mammals' response to projected climate change remained a challenge because climate change projections included mean monthly temperatures but no measures of the full distribution of temperature.
-Since mammals' responses to temperature are nonlinear, accurate estimates of the effects of climate change require estimates of the full distribution of temperatures that mammals will be exposed to -- estimated means alone are not sufficient.
-This is because, since the responses to temperature are nonlinear, the mean response across the distribution of temperatures is not the same as the response to mean temperature.
-This inequality is often referred to as Jensen's inequality or the fallacy of the average [@jensen_sur_1906; @denny_performance_2019], and it is a main motivator for Generalized Linear Models and GAMs [@denny_fallacy_2017; @simpson_method_2025].
-We approximated of the distribution of monthly temperatures by assuming monthly temperatures were normally distributed after estimating monthly variance in temperature, but more work is necessary on estimating changes in climate beyond montly mean values.
-
-A second advantage of our approach is the use of continous-time estimates of speed and distance traveled that are less impacted by irregular sampling rates than straight-line displacement [@noonan_scale-insensitive_2019].
-Still, our results indicate that the movement models fit by `ctmm` still suffer from two crucial issues.
-Firstly, although the models allow speeds to vary stochastically over time, they assume non-zero speeds and do not currently allow for multiple movement states [@augermethe_guide_2021].
-Secondly, while continuous-time estimates of speed are less sensitive to sampling rates than straight-line displacement estimates, our models indicate that longer sampling intervals still cause speed and travel distance to be underestimated [Fig.
-B8$;$ also see @denicola_are_2025].
-Shorter sampling intervals may help overcome this bias [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @pease_ecological_2024].
-To account for the occasional missed GPS fix, we suggest setting sampling intervals to five or more times smaller than the animal's expected directional persistence [@noonan_scale-insensitive_2019].
-Still, excessively high sampling can cause the animal's movement to be confounded with GPS error [@fleming_comprehensive_2020], so we suggest calibrating GPS error whenever possible.
-
-The final (and arguably most evident) limitation of our study was the use of temporally static estimates of forest cover.
-While the use of a single raster produced results and forecasts that are relatively simpler and easier to interpret, species' selection to estimated forest cover was relatively weak.
-Future studies should account for temporal variation in forest cover, especially given the growing severity of forest fires in Canada [@zurowski_summer_2023] as well widespread tree harvesting and its impact on wildlife [@betts_forest_2022; @lochhead_linking_2022].
-Additionally, a finer spatial scale may help detect changes in forest edge selection with greater precision.
-Future work would also benefit from accounting for other important forest properties, such as age, tree composition, and canopy height.
-However, accounting for more variables requires careful consideration about what values such variables should take when estimating changes in movement behaviour under future conditions.
-
-## Considerations for future work
-
-***HERE***
-
-We have presented a multi-species analysis of the effects of temperature on mammalian movement rates and habitat selection in hopes that other researchers can leverage this framework and models to expand our understanding of how temperature affects the movement behaviour of other species, including smaller mammals, more elusive species, and non-mammal animals.
-
-large mammals give clear responses, but they are also already studied most often
-
-More work is necessary on quantifying interspecific responses to temperature, including the effects of temperature on predation rates [but see: @cunningham_opportunity_2021; @glass_trade-off_2021; @brivio_seeking_2024].
-The increase in intensity and frequency of extreme heat events [@bunnell_global_2011; @yao_emergence_2022] will likely also impact the occurrence and timing of hibernation [@wells_life_2022] and migration or seasonal range expansions [@carbeck_adaptation_2022; @morley_spatiotemporal_2021; @malpeli_ungulate_2022].
-However, not all species may be able to adapt at the current rate of climate change [@williams_range_2018; @hetem_responses_2014], especially since landscape changes will likely depend not only on the direct effect of temperature but also on cascading changes in the availability and unpredictability of resources [@mclellan_effect_2015; @pigeon_drivers_2016; @mezzini_how_2025].
-
-Future work should explore the effects of temperature on movement behaviour while accounting for finer-scale and species specific variables that were not accounted for in this study.
-Examples of these include temporally dynamic measures of forest type and age, canopy density, competitive pressures, forage availability, and predator avoidance, or environmental stochasticity.
-However, many of these data not available, so we also suggest that more work be done on quantifying such measures using spatiotemporally dynamic models and sufficiently fine resolutions.
-***...*** animals in the study are generalists; we need more work on specialists (i.e., species without wide species distributions) and smaller species, but they can be harder to track.
-
-changes in the phenology of plants, fire, ice, or mammalian physiology (e.g., moulting, fat storage) or behaviour (e.g., migration, food caching), other than as direct responses to temperature.
+Species' adaptability to change depends mainly on its exposure, sensitivity, and adaptability to changes, and growing evidence suggests that all three factors will limit mammals' ability to adapt to climate change [@ameca_y_juarez_natural_2012; @diffenbaugh_changes_2013].
+Exposure to warming temperatures is widespread throughout the western Canada [@bunnell_global_2011; @kienzle_has_2018; @eum_potential_2023], and changes in temperature and seasonal phenology may be too rapid for many species to respond to [@diffenbaugh_changes_2013].
+Species' adaptability is further challenged by the the difficulty in responding to many concurrent, multifarious, and interconnected changes [@diffenbaugh_changes_2013; @polazzo_measuring_2024].
+Landscapes may thus become unable to support species and population sizes they supported historically, and species may begin selecting against historical ranges in the future [@williams_range_2018].
+Therefore, it is crucial for conservationists to distinguish between a species' ideal environmental space and its best available geographic space [*sensu* @matthiopoulos_defining_2022].
+Accelerating multifarious change in species' movement behaviour and energetic costs require conservation measures to be proactive and anticipate future changes, especially when future conditions are projected to be substantially different from historical conditions.
+@polazzo_measuring_2024 present a framework for estimating species' responses to multifarious change using GAMs that is particularly useful when estimating species' responses to interactions between environmental variables (e.g., responses to changes in both temperature and limiting nutrients).
+By accounting for the effects of multiple variables at once, they demonstrate how a species' response to one variable can depend on other concurrent changes.
+Estimating responses to multifarious change is crucial in assessing the effects of the effects of rising temperatures on community dynamics, including species' selection for current habitats as well as changes in the habitats' carrying capacities and trophic interactions.
+Protected areas must thus adapt to support populations not only in present conditions but also as species' needs and behavior change over the coming decades [@simmons_tough_2021; @livingstone_integrative_2023].
+Recent evidence has questioned whether current protected areas will be effective in the long term [@loarie_velocity_2009; @diffenbaugh_changes_2013; @williams_range_2018], and increasingly more experts have been calling for proactive and long-term conservation.
+By identifying and actively protecting climate change refugia and corridors, we may be able to reduce the effects of climate change on mammalian movement behaviour and community-level responses and ensure that protected areas may be viable for decades to come [@stralberg_toward_2020; @hua_biodiversity_2022; @graham_prioritizing_2019; @michalak_distribution_2018].
 

### 2025-10-01 writing/manuscript.Rmd
tightened discussion and added closing paragraph

@@
 
-\noindent \textbf{Running Title}: Temperature alters mammal movement behavior
+\noindent \textbf{Running Title}: Temperature alters mammal movement behaviour
 
@@
 Global warming during the last century has resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extreme-heat events [@ipcc_climate_2023], with widespread changes in mammals' seasonal phenology, fitness, and life history [@botero_evolutionary_2015; @wells_life_2022; @newediuk_climate_2024; @mclellan_effect_2015; @pigeon_drivers_2016].
-However, behavioral responses to heat vary substantially across species depending on body size, vagility, and activity times [@mccain_body_2014; @williams_range_2018].
+However, behavioural responses to heat vary substantially across species depending on body size, vagility, and activity times [@mccain_body_2014; @williams_range_2018].
 For example, warming temperatures have allowed deer (*Odocoileus* spp.) to shift northward, which has led to increased density of wolves (*Canis lupus*) and higher predation on caribou (*Rangifer tarandus caribou*) [@latham_invading_2011; @barber_potential_2018; @dickie_habitat_2024].
@@
 
-\noindent We have demonstrated that temperature is an important driver of how and where large mammals move, and that mammals alter their seasonal and daily cycles in movement behaviour in response to changes in temperature.
-However, predicting mammals' responses to climate change remains a challenge, as habitats warm rapidly and mammals are exposed to increasingly novel and extreme conditions [@ipcc_climate_2023; @diffenbaugh_changes_2013; @botero_evolutionary_2015].
+\noindent We have demonstrated that temperature is an important driver of how and where large mammals move, and that it affects mammals' seasonal and daily movement behaviour in complex and nonlinear ways.
+However, predicting mammals' responses to climate change remains a challenge, as habitats are warming rapidly and mammals are exposed to increasingly novel and extreme conditions [@ipcc_climate_2023; @diffenbaugh_changes_2013; @botero_evolutionary_2015].
 We leveraged the flexibility and interpretability of HGAMs to estimate mammals' nonlinear responses to temperature without imposing rigid assumptions about the shape of the responses [@simpson_method_2025; @simpson_modelling_2018].
@@
 We were thus able to present results using relatively simple and interpretable surface plots (Figs.
-\@ref(fig:dist) and \@ref(fig:hrsfs)) that conveyed detailed and nuanced estimates of change.
+\@ref(fig:dist) and \@ref(fig:hrsfs)) that conveyed nuanced, continuous-time estimates of change.
 This framework was essential for estimating how mammals behave at extreme temperatures.
-Despite the recent increase in frequency of extreme weather events in western Canada [@zhang_increased_2023; @bunnell_global_2011; @kienzle_has_2018], such events remain relatively rare, so estimating mammals' nonlinear responses required the flexibility provided by GAMs [@wood_generalized_2017].
+In fact, despite the recent increase in frequency of extreme heat events in western Canada [@zhang_increased_2023; @bunnell_global_2011; @kienzle_has_2018], such events remain relatively rare, so estimating mammals' nonlinear responses required the flexibility provided by GAMs [@wood_generalized_2017].
 Additionally, hierarchical models allowed us to produce statistically sound estimates that accounted for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mccabe_resource_2021; @jeltsch_need_2025; @silva_too_2025].
@@
 
-Our results suggest that climate change will cause mammals to alter their seasonal movement behaviour, but not all species will respond similarly.
-For example, hotter temperatures caused boreal caribou to travel longer distances in fall, winter, and spring but less in summer, mirroring the findings of @stien_congruent_2012, @leclerc_determinants_2021, and @lessard_local_2025.
+Species varied greatly in how they adapted their movement behaviour to temperature.
+For boreal caribou, hotter temperatures increased movement rates in fall, winter, and spring but decreased them in summer, mirroring the findings of @stien_congruent_2012, @leclerc_determinants_2021, and @lessard_local_2025.
 The increase in movement rates during the colder seasons are likely partly due to shallower snow depth that results in greater mobility [@pedersen_quantifying_2021], but rising temperatures during snowy seasons also increase the risk of rain-on-snow events, which limit forage availability and increase time spent foraging [@stien_congruent_2012; @berger_climate_2018; @mezzini_how_2025].
-In contrast, the reduced movement rates during warmer summers are likely mainly due to caribou's reduced movement rates and use of thermal refugia, given ungulates' elevated risk of hyperthermia [@alston_temperature_2020; @verzuh_behavioural_2023].
+The reduced movement rates during warmer summers are likely mainly due to more frequent resting and use of thermal refugia, given ungulates' elevated risk of hyperthermia [@alston_temperature_2020; @verzuh_behavioural_2023].
 Therefore, although we predict that boreal caribou's yearly average movement rates will remain approximately unchanged throughout the 21^st^ century, the projections do not show the changes in seasonal movement phenology.
-However, this is an important caveat, since the effects of extreme temperatures are exacerbated by phenological mismatches with seasonal photoperiod [@walker_global_2019], including the timing of molting and reproduction.
-Earlier growth seasons in boreal and arctic regions have resulted in lower calf birth and survival [@post_climate_2008], while the lengthening of the growth season has allowed moose (*Alces alces*) and deer to encroach on boreal caribou habitat and increase the density of coyotes (*Canis latrans*), cougars, and wolves [@barber_potential_2018; @demars_incorporating_2023], whose movement behavior also depends on ambient temperature.
-Thus, rising temperatures affect more than each individual species' movement behaviour; they have complex and interconnected effects on an entire community and its trophic structure.
+This is an important caveat, since the effects of extreme temperatures are exacerbated by phenological mismatches with seasonal photoperiod [@walker_global_2019], including the timing of molting and reproduction.
+Earlier growth seasons in boreal and arctic regions have resulted in lower calf birth and survival [@post_climate_2008], while the lengthening of the growth season has allowed moose (*Alces alces*) and deer to encroach on boreal caribou habitat and increase the density of coyotes (*Canis latrans*), cougars, and wolves [@barber_potential_2018; @demars_incorporating_2023], whose movement behaviour also depends on ambient temperature.
+Boreal wolves responded to temperature similarly to boreal wolves, and habitat selection strength was not projected to change noticeably by 2100, but wolves were predicted to move less in future decades.
+In contrast, cougars showed markedly different seasonal responses to temperature, with relatively warmer temperatures always resulting in less travel, which resulted in strong predicted decreases in movement rates over the 21^st^ century.
+However, spatial heterogeneity in habitat, climate, and cougar physiology and behaviour may cause cougars in boreal regions to move differently from those in our study.
+Still, rising temperatures clearly affect more than each individual species' movement behaviour; they have complex and interconnected effects on entire communities and their trophic structures.
 
-Work by @cunningham_opportunity_2021 discusses the importance of adopting a community-level perspective when assessing the effects of climate change on animals' movement behavior and phenology.
-For example, our results show that wolves responded to temperature by altering their movement rates similarly to caribou.
-Consequently, as tempeatures warm and both species increase their movement rates in fall, winter, and spring, the resulting increase in encounter rates [@martinez-garcia_how_2020] will likely increase predation risk when caribou are pregnant or with young calves.
-Additionally, while both species saw reduced movement rates during hot summer days, boreal wolves' use of anthropogenic linear features (e.g., roads, seismic lines) may allow them to reduce movement costs while increasing the chances of encountering heat-stressed prey [@whittington_caribou_2011; @dickie_faster_2017; @dickie_resource_2022].
-Furthermore, caribou that attempt to reduce predation risk from wolves by avoiding wolf habitat may still risk increasing predation pressure from bears [@leblond_caribou_2016] and other encroaching predators [@demars_incorporating_2023; @labadie_global_2023].
+@cunningham_opportunity_2021 discuss the importance of adopting a community-level perspective when assessing the effects of climate change on animals' movement behaviour and phenology.
+For example, as temperatures warm and both boreal caribou and wolves increase their movement rates in fall, winter, and spring, they also increase their encounter rates [@martinez-garcia_how_2020], which will likely increase predation risk when caribou are pregnant or with young calves.
+Additionally, while both species saw reduced movement rates during hot summer days, boreal wolves' use of anthropogenic linear features (e.g., roads, seismic lines) may allow them to reduce the thermal costs of movement while increasing the chances of encountering heat-stressed prey [@whittington_caribou_2011; @dickie_faster_2017; @dickie_resource_2022].
+Furthermore, caribou that attempt to reduce predation risk from wolves by avoiding wolf habitat may still risk increasing predation pressure from bears [@leblond_caribou_2016] and the aforementioned encroaching predators [@demars_incorporating_2023; @labadie_global_2023].
 
@@
 Similarly, species' changes in habitat selection do not explicitly account for any changes in physiological or energetic costs.
-Additionally, forest cover, elevation, and distance from water are often correlated, so the estimated effects should be interpreted carefully, and they may not be applicable outside our study areas.
+Additionally, forest cover, elevation, and distance from water are often correlated, so the estimated effects should be interpreted carefully, and they may not be applicable outside each species' study area.
 For example, both mountain goats and elk selected for higher elevation as temperatures warmed, but our models did not explicitly account for differences in forage availability at different altitudes.
 
-Forage quality and availability is often a limiting factor for mammals' ability to adapt to warming temperatures by moving to higher altitudes, since plants are slower to arrive in and establish in new habitats than mammals, and the advance of herbivores can further limit plants' ability to establish [@speed_elevational_2012; @diffenbaugh_changes_2013].
-Consequently, mammals that move to higher elevations may be forced to spend more time foraging [@mezzini_how_2025], which will increase their energetic expenditure as well as their chances of encountering predators (including cougars and grizzly bears) or competitors [@kohl_prey_2019; @martinez-garcia_how_2020; @torrez-herrera_monkeys_2020; @smith_densitydependent_2023].
+Climate change has also had extensive impacts on plant growth and phenology [@cleland_shifting_2007; @denny_performance_2019; @tysor_countergradient_2025] with consequences on herbivores' behaviour and fitness [@post_climate_2008; @aikens_greenscape_2017].
+Forage quality and availability are often a limiting factor for herbivores' ability to escape hotter temperatures by moving to higher altitudes or latitudes, since plants are slower to arrive in and establish in new habitats, and the advance of herbivores can further limit plants' ability to establish [@speed_elevational_2012; @diffenbaugh_changes_2013].
+Consequently, mammals that move to higher elevations may be forced to spend more time foraging [@mezzini_how_2025], which will increase both their energetic expenditure and encounter rates with predators and competitors [@kohl_prey_2019; @martinez-garcia_how_2020; @torrez-herrera_monkeys_2020; @smith_densitydependent_2023].
 If local peaks become unsuitable habitat, and moving to nearby higher peaks requires traveling across low-quality and high-risk habitat (Figure \@ref(fig:s-rss-cc)), habitat fragmentation may prevent movement across ranges [@white_mountain_2025].
@@
 
-Future work should develop models that account for covariates that are specific to the study species and area, such as forest age and disturbance [@demars_incorporating_2023], terrain ruggedness [@white_mountain_2017], anthropogenic linear features [@dickie_faster_2017], and growing human presence [@sih_evolution_2011; @rice_conspicuous_2022] and risk of human-wildlife conflict [@abrahms_climate_2023; @weststrate_how_2024].
-In particular, forests in western Canada are highly dynamic over time [@zhang_half-century_2015], and fires size and burn severity have increased in western Canada in recent decades [@zurowski_summer_2023; @parisien_abrupt_2023].
-Thus, while our use of a static raster of forest cover allowed for simpler models and climate change projections, future studies should account for mammals' responses to temporal forest dynamics, including the impacts of fire [@zurowski_summer_2023; @parisien_abrupt_2023], harvest [@lochhead_linking_2022] and other extractive industries [@dickie_faster_2017; @dickie_habitat_2024], as well as the subsequent forest regeneration.
-However, the inclusion of temporally dynamic variables will require that any future projections of species' behavior depend on temporally dynamic estimates of how such variables will change over time with changes in climate and anthropogenic activity.
-
-Species' adaptability to change depends mainly on its exposure, sensitivity, and adaptability to changes, and growing evidence suggests that all three factors will limit mammals' ability to adapt to climate change [@ameca_y_juarez_natural_2012; @diffenbaugh_changes_2013].
-Exposure to warming temperatures is widespread throughout the western Canada [@bunnell_global_2011; @kienzle_has_2018; @eum_potential_2023], and changes in temperature and seasonal phenology may be too rapid for many species to respond to [@diffenbaugh_changes_2013].
-Species' adaptability is further challenged by the the difficulty in responding to many concurrent, multifarious, and interconnected changes [@diffenbaugh_changes_2013; @polazzo_measuring_2024].
-Landscapes may thus become unable to support species and population sizes they supported historically, and species may begin selecting against historical ranges in the future [@williams_range_2018].
+Growing evidence suggests that mammals' exposure, sensitivity, and adaptability to climate change will limit their ability to adapt to change [@ameca_y_juarez_natural_2012; @diffenbaugh_changes_2013].
+Exposure to warming temperatures is widespread throughout western Canada [@bunnell_global_2011; @kienzle_has_2018; @eum_potential_2023], and changes in temperature and seasonal phenology may be too rapid for many species to respond to [@diffenbaugh_changes_2013].
+Species' adaptability is further challenged by the difficulty in responding to concurrent, multifarious, and interconnected changes [@diffenbaugh_changes_2013; @polazzo_measuring_2024].
+In the future, landscapes may thus become unable to support species and population sizes they supported historically, and species may begin selecting against historical ranges [@williams_range_2018].
 Therefore, it is crucial for conservationists to distinguish between a species' ideal environmental space and its best available geographic space [*sensu* @matthiopoulos_defining_2022].
-Accelerating multifarious change in species' movement behaviour and energetic costs require conservation measures to be proactive and anticipate future changes, especially when future conditions are projected to be substantially different from historical conditions.
-@polazzo_measuring_2024 present a framework for estimating species' responses to multifarious change using GAMs that is particularly useful when estimating species' responses to interactions between environmental variables (e.g., responses to changes in both temperature and limiting nutrients).
+Accelerating multifarious change in species' movement behaviour and energetic costs require conservation measures that are proactive and anticipate future changes.
+@polazzo_measuring_2024 present a framework for estimating species' responses to multifarious change using GAMs that is particularly useful when estimating species' (nonlinear) responses to interactions between environmental variables (e.g., changes in both temperature and limiting nutrients).
 By accounting for the effects of multiple variables at once, they demonstrate how a species' response to one variable can depend on other concurrent changes.
-Estimating responses to multifarious change is crucial in assessing the effects of the effects of rising temperatures on community dynamics, including species' selection for current habitats as well as changes in the habitats' carrying capacities and trophic interactions.
-Protected areas must thus adapt to support populations not only in present conditions but also as species' needs and behavior change over the coming decades [@simmons_tough_2021; @livingstone_integrative_2023].
+Estimating responses to multifarious change is crucial in assessing mammals' responses to hotter temperatures, altered phenology, and novel community dynamics.
+Protected areas must support populations not only in present conditions but also as species' needs and behaviour change over the decades [@simmons_tough_2021; @livingstone_integrative_2023].
 Recent evidence has questioned whether current protected areas will be effective in the long term [@loarie_velocity_2009; @diffenbaugh_changes_2013; @williams_range_2018], and increasingly more experts have been calling for proactive and long-term conservation.
 By identifying and actively protecting climate change refugia and corridors, we may be able to reduce the effects of climate change on mammalian movement behaviour and community-level responses and ensure that protected areas may be viable for decades to come [@stralberg_toward_2020; @hua_biodiversity_2022; @graham_prioritizing_2019; @michalak_distribution_2018].
+
+Future work should develop models that account for covariates that are specific to the study species and area, such as forest age and disturbance [@demars_incorporating_2023], terrain ruggedness [@white_mountain_2017], anthropogenic linear features [@dickie_faster_2017], and growing human presence in wild spaces [@sih_evolution_2011; @rice_conspicuous_2022], including the growing risk of human-wildlife conflict [@abrahms_climate_2023; @weststrate_how_2024].
+In particular, forests in western Canada are highly dynamic over time [@zhang_half-century_2015], and fires size and burn severity have increased in western Canada in recent decades [@zurowski_summer_2023; @parisien_abrupt_2023].
+Thus, while our use of a static raster of forest cover allowed for simpler models and climate change projections, future studies should account for mammals' responses to temporal forest dynamics, including the impacts of fire [@zurowski_summer_2023; @parisien_abrupt_2023], logging [@lochhead_linking_2022] and other extractive industries [@dickie_faster_2017; @dickie_habitat_2024].
+However, the inclusion of temporally dynamic variables will require that any future projections of species' behaviour depend on temporally dynamic estimates of how such variables will change over time with changes in climate and anthropogenic activity.
+
+We have demonstrated that temperature has complex and nonlinear effects on how and where mammals move.
+Rising temperatures have impacted mammals' daily and seasonal movement phenology, and changes are expected to intensify throughout the 21^st^ century.
+Quantifying how temperature affects mammalian behavior and fitness is a first step towards developing an understanding of the effects of climate change on landscape dynamics and the effectiveness of protected areas in the future.
+By leveraging the flexibility and statistical power of Hierarchical Generalized Additive Models, we have provided researchers with malleable and data-driven methods for assessing species' responses to climate change.
+We hope that future work will develop more species-specific models with the intent of developing a community-level framework for quantifying and predicting how communities will respond to climate change.
 

### 2025-10-01 writing/manuscript.Rmd
updated refs; split phenology and time in DAGs

@@
 
-\noindent \textbf{References}: 147 (updated on 2025-09-30) <!-- counted using movement ecology csl -->
+\noindent \textbf{References}: 155 (updated on 2025-10-01) <!-- counted using movement ecology csl -->
 
@@
 Elk data from the work of @ciuti_human_2012 were downloaded from Movebank [@kays_movebank_2022].
-***MISSING INFO ON GRIZZLY BEAR AND GOAT TELEMETRY DATA.***
+Finally, grizzly bear telemetries are from the work of @ref, while mountain goat data are from @ref.
 
@@
 
-```{r dags,  message=FALSE, warning=FALSE, fig.cap="Directed Acyclical Graphs (DAGs) assumed for inferring the causal effects of temperature on each species' movement behaviour.
+```{r dags,  message=FALSE, warning=FALSE, fig.cap="Directed Acyclical Graphs (DAGs) assumed for inferring the causal effects of temperature (red) on each species' movement behaviour.
 (A) Ambient temperature affects mammals' movement rates (i.e.
-probability of moving, speed when moving, and distance traveled), but temperature is influenced by time of day and day of year.
-Additionally, the effects of temperature on movement rates depend on time of day and season, since animals may respond to temperatures differently at different times of day and or days of year.
-Finally, the effect of time of day changes seasonally due to changes in photoperiod.
-(B) Similarly, habitat selection depends on the availability and selection of resources (forest cover, elevation, and distance from water), but the selection strengh for each resource is conditional on temperature.
+probability of moving, speed when moving, and their product: distance traveled).
+The effects of temperature on mammals' movement rates depend on daily and seasonal phenology, since animals may respond to temperatures differently at different times of day and or days of year.
+Additionally, both temperature and phenology depend on time of day and day of year.
+Finally, daily phenology changes with seasonal phenology due to changes in photoperiod and life history (e.g, reproduction).
+(B) Similarly, habitat selection depends on the availability and selection of habitat variables (forest cover, elevation, and distance from water), but the selection strengh for each variable is conditional on temperature.
 For example, an animal may avoid forested areas unless it is seeking refuge from extreme temperatures.", fig.height=8, fig.width=20, out.width='\\linewidth'}
@@
 # movement rates
-boxes_a <- tibble(xmin = c(0, 10, 5, 5),
+boxes_a <- tibble(xmin = c(0, 10, 5, 5, 0, 10),
                   xmax = xmin + 4,
-                  ymin = c(10, 10, 5, 0),
+                  ymin = c(10, 10, 5, 0, 3, 3),
                   ymax = ymin + 1,
-                  fill = c('#BBBBBB80', '#BBBBBB80', '#FF8C00AA', '#FF0000CC'),
-                  label = c('Time of day', 'Day of year', 'Movement rates',
-                            'Temperature'))
+                  fill = c('#BBBBBB80', '#BBBBBB80', '#FF8C00AA', '#FF0000CC',
+                           '#BBBBBB80', '#BBBBBB80'),
+                  label = c('Daily phenology', 'Seasonal phenology', 'Movement rates',
+                            'Temperature', 'Time of day', 'Day of year'))
 
-arrows_a <- tibble(id = sort(rep(1:6, 2)),
-                   x = c(10, 4, 4, 6, 10, 8, 2, 6, 7, 7, 12, 8),
-                   y = c(10.5, 10.5, 10, 6, 10, 6, 10, 1, 1, 5, 10, 1))
+arrows_a <- tibble(id = sort(rep(1:10, 2)),
+                   x = c(10, 4, 4, 6, 10, 8, 6, 2, 7, 7, 8, 12, 2, 5, 1, 1, 12, 9, 13, 13),
+                   y = c(10.5, 10.5, 10, 6, 10, 6, 1, 10, 1, 5, 1, 10, 3, 0.5, 4, 10, 3, 0.5, 4, 10))
 

### 2025-10-03 writing/manuscript.Rmd
addressed some comments from KH

@@
 We address this gap by quantifying the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species (boreal and southern mountain caribou, cougars, elk, grizzly bears, mountain goats, and wolves) throughout western Canada between 1998 and 2023.
-We show that temperature significantly affected of how and where these mammals moved, and we predict that climate change will impact mammals' movement behaviour throughout the 21^st^ century, although species did not respond uniformly to temperature.
+We show that temperature significantly affected how and where these mammals moved, and we predict that climate change will impact mammals' movement behaviour throughout the 21^st^ century, although species did not respond uniformly to temperature.
 Projected responses to climate change suggested that rising temperatures will drive southern mountain caribou and mountain goats to move more, while cougars, elk, and wolves will move less.
 Boreal caribou and grizzly bears showed little change in projected yearly movement rates but clear shifts in seasonal phenology.
-We also predict that rising temperatures will reduce median habitat selection strength most of the six species, but some are expected to show increased selection strength for higher altitudes.
+We also predict that rising temperatures will reduce median habitat selection strength for most of the six species, but some are expected to show increased selection strength for higher altitudes.
 As mammals increasingly alter their movement rates and select against portions of their current ranges, changes in individuals' movement behaviour will impact encounter rates, including predator-prey dynamics and human-wildlife interactions.
 Conservation efforts should therefore account for future changes in movement behaviour as well as the consequences such changes may have on communities.
-Proactively anticipating changes in mammalian movement behaviour will become crucial for effectively and proactively understanding community-level responses and selecting high-quality habitat for long-term conservation."
+Anticipating changes in mammalian movement behaviour will become crucial for effectively and proactively understanding community-level responses and selecting high-quality habitat for long-term conservation."
 if(FALSE){
@@
 Animals may search for a short-term buffer from extreme temperatures by moving to thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], and shelter may even be preferred over forage abundance [@hall_microrefuges_2016].
-However, climate change during the last century [@ipcc_climate_2023] have caused animals to alter how and where they move as they respond to hotter temperatures and increasingly severe and frequent extreme-heat events [@ipcc_climate_2023; @bunnell_global_2011].
+However, climate change during the last century [@ipcc_climate_2023] has caused animals to alter how and where they move as they respond to hotter temperatures and increasingly severe and frequent extreme-heat events [@ipcc_climate_2023; @bunnell_global_2011].
 
@@
 Mammals' homeothermy can cause them to have difficulty dissipating heat, particularly when ambient temperatures are near or above body temperature.
-Large-bodied mammals are most likely to alter their movement behaviour with rising temperatures, since they tend to move long distances [@merkle_large_2016], are particularly sensitive to overheating [@fuller_towards_2016; @alston_temperature_2020; @verzuh_behavioural_2023], and their movement rates are thus strongly limited by excessive heat [@dyer_travel_2023].
+Larger-bodied mammals are more likely to alter their movement behaviour with rising temperatures, since they tend to move long distances [@merkle_large_2016], are particularly sensitive to overheating [@fuller_towards_2016; @alston_temperature_2020; @verzuh_behavioural_2023], and their movement rates are thus strongly limited by excessive heat [@dyer_travel_2023].
+
 Global warming during the last century has resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extreme-heat events [@ipcc_climate_2023], with widespread changes in mammals' seasonal phenology, fitness, and life history [@botero_evolutionary_2015; @wells_life_2022; @newediuk_climate_2024; @mclellan_effect_2015; @pigeon_drivers_2016].
@@
 
-As heat stress intensifies over the next 100 years, mammals will continue to suffer impacts on their fitness, movement rates, and habitat selection [@deb_modelling_2020; @woo-durand_increasing_2020].
+As heat stress intensifies throughout the 21^st^ century, mammals will continue to suffer impacts on their fitness, movement rates, and habitat selection [@deb_modelling_2020; @woo-durand_increasing_2020].
 It remains unclear how or whether species will be able to respond to climate change in the current century [@deb_modelling_2020; @woo-durand_increasing_2020; @verzuh_behavioural_2023], especially if populations fail to adapt [@sawyer_migratory_2019; @botero_evolutionary_2015] or are physiologically incapable to do so [@sherwood_adaptability_2010; @botero_evolutionary_2015; @williams_range_2018].
@@
 
-In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves: *Canis lupus*; Table \@ref(tab:data-table)).
+In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves; Table \@ref(tab:data-table)).
 Western Canada is currently experiencing accelerating and widespread but spatially heterogeneous warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme heat events [@zhang_increased_2023].
@@
 Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022; @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].
-We discuss the consequences of our findings at the individual, population, and community levels, and we also provide directions for future research and solutions for overcoming some common data limitations.
 
@@
   mutate(dt = round('hours' %#% dt, 2)) %>% # convert from seconds to hours
-  rename('\U0394t (hours)' = dt,
+  rename(Fixes = n,
+         '\U0394t (hours)' = dt,
          'Has speed' = n_missing)
 
-knitr::kable(dt_data, format = 'markdown', format.args = list(big.mark = ','), caption = "Summary statistics of each species' set of GPS data, namely: the start and end of each set of GPS telemetries, the number of GPS fixes (after data cleaning), the number the median sampling interval ($\\Delta t$; stratified by animal), the number of animals (Animals), and the number of animals with finite speed estimates (Has speed).", escape = FALSE, longtable = FALSE, booktabs = TRUE)
+knitr::kable(dt_data, format = 'markdown', format.args = list(big.mark = ','), caption = "Summary statistics of each species' set of GPS data, namely: the start and end of each set of GPS telemetry data, the number of GPS fixes (after data cleaning), the median sampling interval ($\\Delta t$; stratified by animal), the number of animals (Animals), and the number of animals with finite speed estimates (Has speed).", escape = FALSE, longtable = FALSE, booktabs = TRUE)
 ```
@@
 We limited our models to only use three simple and temporally static rasters to produce simple and easily interpretable estimates and forecasts under the assumption of temporally invariant landscapes.
-See the discussion section for additional detail on the advantages and limitations of using such simple models.
 
@@
 
-\noindent Boreal caribou and wolf telemetries were acquired from a data repository of the British Columbia Oil and Gas Research and Innovation Society (BC ORGIS) that was collected as part of the boreal caribou radio-collaring program of the BC OGRIS Research and Effectiveness Monitoring Board [REMB$;$ @bc_ogris_boreal_2018].
-Southern mountain caribou data were obtained from the work of @ford_effects_2023.
-Cougar data include telemetries from @darlington_southern_2025 and @serrouya_ref.
+\noindent Boreal caribou and wolf telemetry data were acquired from a data repository of the British Columbia Oil and Gas Research and Innovation Society (BC ORGIS) that was collected as part of the boreal caribou radio-collaring program of the BC OGRIS Research and Effectiveness Monitoring Board [REMB$;$ @bc_ogris_boreal_2018].
+Southern mountain caribou data were obtained from @ford_effects_2023.
+Cougar data are from @darlington_southern_2025 and @serrouya_ref.
 Elk data from the work of @ciuti_human_2012 were downloaded from Movebank [@kays_movebank_2022].
-Finally, grizzly bear telemetries are from the work of @ref, while mountain goat data are from @ref.
+Finally, grizzly bear telemetry data are from @ref, while mountain goat data are from @ref.
 
-From the full set of telemetry data, we removed a total of 2396 problematic GPS locations (0.16%, including collar calibration data) after inspecting diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points.
+From the full set of telemetry data, we removed 2,396 problematic GPS locations (0.16%, including collar calibration data) after inspecting diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points.
 Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated relatively stationary behaviour (Appendix A).
@@
 We simulated hourly variation in future years by assuming hourly temperature followed a normal distribution with mean specified by the monthly `climatenaR` climate projections and variance as specified by the gamma GAMLS.
-For each month within each year from 2025 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantiles by increments of 0.1, and we weighted each quantile proportionally to the (normalized) Gaussian probability density for each quantile.
+For each month within each year from 2025 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantiles by increments of 0.1, and we weighted each quantile proportionally to the normalized Gaussian probability density for each quantile.
 
@@
 All final rasters and code are available on GitHub (see the Data Availability Statement).
-For ease of interpretability and comparison with current conditions, we assumed resources would remain constant through to 2100, although we recognize that the spatial distribution of forest cover and open water will likely change throughout the 21^st^ century.
+For ease of interpretability and comparison with current conditions, we assumed resources would remain constant through to 2100, although we recognize that the spatial distribution of forest cover and open water will change throughout the 21^st^ century.
 
@@
 1.2.0, @fleming_ctmm_2023] for `R`.
-We then estimated each mammal's instantaneous speed at each observed location by applying the `ctmm::speeds()` function on all models with finite speed estimates [415 of 433, @fleming_fine-scale_2014; @noonan_scale-insensitive_2019].
-The 18 animals with insufficiently fine sampling to support movement models with finite speed were for one mountain goat, 15 boreal caribou, and two southern mountain caribou (Table \@ref(tab:data-table)).
+We then estimated each mammal's instantaneous speed at each observed location by applying the `ctmm::speeds()` function on all models with finite speed estimates [415 of 433$;$ @fleming_fine-scale_2014; @noonan_scale-insensitive_2019].
+The remaining 18 movement models had sampling rates that were too coarse, relative to the animals' directional persistance, to reconstruct the animals' movement trajectories [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @denicola_are_2025].
+The models were for one mountain goat, 15 boreal caribou, and two southern mountain caribou (Table \@ref(tab:data-table)).
 
@@
 (A) Ambient temperature affects mammals' movement rates (i.e.
-probability of moving, speed when moving, and their product: distance traveled).
+probability of moving, speed when moving, and their product: hourly distance traveled).
 The effects of temperature on mammals' movement rates depend on daily and seasonal phenology, since animals may respond to temperatures differently at different times of day and or days of year.
 Additionally, both temperature and phenology depend on time of day and day of year.
-Finally, daily phenology changes with seasonal phenology due to changes in photoperiod and life history (e.g, reproduction).
+Finally, daily phenology changes with seasonal phenology due to changes in photoperiod (e.g., the time of twilight changes throughout the year, affecting crepuscular activity).
 (B) Similarly, habitat selection depends on the availability and selection of habitat variables (forest cover, elevation, and distance from water), but the selection strengh for each variable is conditional on temperature.
-For example, an animal may avoid forested areas unless it is seeking refuge from extreme temperatures.", fig.height=8, fig.width=20, out.width='\\linewidth'}
+For example, an animal may select for more densely-forested areas at extreme temperatures.", fig.height=6, fig.width=15, out.width='\\linewidth'}
 library('dplyr')   # for data wrangling
@@
                   ymax = ymin + 1,
-                  fill = c('#BBBBBB80', '#BBBBBB80', '#FF8C00AA', '#FF0000CC',
-                           '#BBBBBB80', '#BBBBBB80'),
+                  bold = c(FALSE, FALSE, TRUE, TRUE, FALSE, FALSE),
                   label = c('Daily phenology', 'Seasonal phenology', 'Movement rates',
@@
   ggplot() +
-  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
-            boxes_a, fill = boxes_a$fill, color = 'black') +
+  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, lwd = bold),
+            boxes_a, color = 'black', fill = 'white') +
   geom_path(aes(x, y, group = id), arrows_a, linewidth = 1,
@@
              fontface = 'bold', label.size = 0, size = 5) +
+  scale_linewidth_manual(values = c(0.5, 1)) +
   theme_void() +
@@
                   ymax = ymin + 1,
-                  fill = c('#FF8C00AA', rep('#BBBBBB80', 3), '#FF0000CC'),
+                  bold = c(TRUE, FALSE, FALSE, FALSE, TRUE),
                   label = c('Habitat selection', 'Forest cover',
@@
   ggplot() +
-  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
-            boxes_b, fill = boxes_b$fill, color = 'black') +
+  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, lwd = bold),
+            boxes_b, color = 'black', fill = 'white') +
   geom_path(aes(x, y, group = id), arrows_b, linewidth = 1,
@@
 Consequently, we present all results while predicting specifically for one-hour sampling intervals.
-At $0^\circ$C, species differed in estimated mean probabilities of moving ($\hat P(M = 1)$; range: 0.05 -- 0.31), mean speed when moving ($\hat {\mathbb E}(S|M=1)$; range: 0.42 -- 2.67 km/h), and mean distance traveled (i.e., $\hat P(M) \times \hat {\mathbb E}(S|M=1)$, range: 0.04 -- 0.61 km/h; Table \@ref(tab:hgam-summary)).
+At $0^\circ$C, species differed in estimated mean probabilities of moving ($\hat P(M = 1)$; range: 0.05 -- 0.31), mean speed when moving ($\hat {\mathbb E}(S|M=1)$; range: 0.42 -- 2.67 km/h), and mean overall speed (i.e., $\hat P(M) \times \hat {\mathbb E}(S|M=1)$, range: 0.04 -- 0.61 km/h; Table \@ref(tab:hgam-summary)).
 Grizzly bears had the lowest movement frequency ($\hat P(M) \approx 0.05$), while wolves and cougars moved most often ($\hat P(M) \ge 0.22$).
@@
 
-  # find difference in distance traveled, relative to furthest distance
+  # find difference in hourly distance traveled, relative to furthest distance
   intercepts %>%
@@
 
-  # find difference in distance traveled, relative to furthest distance
+  # find difference in hourly distance traveled, relative to furthest distance
   round(range(max(intercepts$`E(D|T)`) / intercepts$`E(D|T)`[- which.max(intercepts$`E(D|T)`)]), 1)
@@
                 'T>+20\\textdegree{}C (%)' = get_perc(temp_c > +20))) %>%
-  knitr::kable(format = 'markdown', digits = 2, format.args = list(big.mark = ','), caption = "Summary statistics for each species' GPS fixes with finite speed estimates, namely: the number fixes after data cleaning ($n$), the percentage of fixes with temperature ($T$) below $-20^\\circ$C and above $20^\\circ$C, the estimated mean probability of moving ($\\hat P(M=1)$), the mean speed when moving ($\\hat{\\mathbb E}(S|M=1)$; km/h), and the mean distance travelled ($\\hat P(M=1) \\times \\hat{\\mathbb E}(S|M=1) = \\hat{\\mathbb E}(D)$; km/h), for a sampling interval of 1 hour and a temperature $T$ of $0^\\circ$C.")
+  knitr::kable(format = 'markdown', digits = 2, format.args = list(big.mark = ','), caption = "Summary statistics for each species' GPS fixes with finite speed estimates, namely: the number fixes after data cleaning ($n$), the percentage of fixes with temperature ($T$) below $-20^\\circ$C and above $20^\\circ$C, the estimated mean probability of moving ($\\hat P(M=1)$), the mean speed when moving ($\\hat{\\mathbb E}(S|M=1)$; km/h), and the mean hourly distance travelled ($\\hat P(M=1) \\times \\hat{\\mathbb E}(S|M=1) = \\hat{\\mathbb E}(D)$; km/h), for a sampling interval of 1 hour and a temperature $T$ of $0^\\circ$C.")
 ```
@@
 Across all species, Relative Selection Strength (RSS) was weakest for forest cover and strongest for elevation.
-At temperatures near 0$^\circ$C, wolves selected for forest cover ($\gtrapprox$ 50%), elevations near 1 km, and distances from water < 5 km; elk selected for intermediate forest cover ($\approx$ 50%), elevations between 1 and 2 km, and distances from water of 10-15 km; mountain goats selected for sparse forest cover (< 25%), elevations near 1.5 km, and distances from water < 5 km; cougars selected for dense forest cover (> 75%), an elevation of ~ 1 km, and distances from water < 7.5 km; boreal caribou selected for forest cover between 50% and 75%, elevations near 500 m, and distances from water < 10 km; southern mountain caribou selected for dense forest cover, elevations near 2 km, and distances from water $\lessapprox$ 5 km; and grizzly bears selected for relatively sparse forest cover (25-50%), elevation between 1 and 2 km, and distances from water < 3 km.
-
-There was relatively strong agreement between models with and without temperature (Figs.
-B3, and B14), but including temperature always resulted in better fits (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\le$ -342; Appendix B).
-Accounting for temperature even resulted in a simpler HRSF for mountain goats (lower model degrees of freedom) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.
+At temperatures near 0$^\circ$C, boreal caribou selected for forest cover between 50% and 75%, elevations near 500 m, and distances from water < 10 km, while southern mountain caribou selected for dense forest cover, elevations near 2 km, and distances from water $\lessapprox$ 5 km.
+Cougars selected for dense forest cover (> 75%), an elevation of ~ 1 km, and distances from water < 7.5 km.
+Elk selected for intermediate forest cover ($\approx$ 50%), elevations between 1 and 2 km, and distances from water of 10-15 km.
+Grizzly bears selected for relatively sparse forest cover (25-50%), elevation between 1 and 2 km, and distances from water < 3 km.
+Mountain goats selected for sparse forest cover (< 25%), elevations near 1.5 km, and distances from water < 5 km.
+Finally, wolves selected for forest cover ($\gtrapprox$ 50%), elevations near 1 km, and distances from water < 5 km.There was relatively strong agreement between models with and without temperature (Figs.
+B3, and B14), but including temperature always resulted in better fits (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; Appendix B).
+Accounting for temperature even resulted in a simpler HRSF for mountain goats (lower model degrees of freedom due to simpler smooth terms) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.
 
@@
 \@ref(fig:dist)).
-The clearest example of this was cougars.
-In late spring (June 1^st^), they moved from evening to early morning if temperatures were below $20^\circ$C, but if temperatures were above 20$^\circ$C they moved mostly between 3:00 and 6:00 AM.
+The response was most visible in cougars.
+In late spring (June 1^st^), they moved from evening to early morning if hourly temperatures were below $20^\circ$C, but if temperatures were above 20$^\circ$C they moved mostly between 3:00 and 6:00 AM.
 Throughout the year, they tended to move more when it was colder, but what they perceived as "colder" was relative to the time of year.
@@
 
-```{r dist, fig.cap="\\textbf{Temperature is a strong driver of how far and when mammals travel.} The fill indicates the effects of temperature on the relative change in distance travelled (probability of moving times mean speed when moving) over time of day on June $1^{\\text{st}}$ (\\textbf{A}) and day of year at 12:00 PDT (\\textbf{B}).
+```{r dist, fig.cap="\\textbf{Temperature is a strong driver of how far and when mammals travel.} The fill indicates the effects of temperature on the relative change in hourly distance travelled (probability of moving times mean speed when moving) over time of day on June $1^{\\text{st}}$ (\\textbf{A}) and day of year at 12:00 PDT (\\textbf{B}).
 Predictions in the surface plots extend to 10% of the range away from each datum.
@@
 Changes in RSS with temperature were strongest for elevation and generally weakest for distance from water, but there were no common trends across all species for any of the three resources.
-All species, with the exception of cougars, exhibited clear temperature-dependent shifts in their preference for forest cover.
-At higher temperatures, wolves relaxed their preference for forested areas, while mountain goats relaxed their preference open areas (cover < 50%).
+All species except cougars exhibited clear temperature-dependent shifts in their preference for forest cover.
+At higher temperatures, wolves relaxed their preference for forested areas, while mountain goats relaxed their preference for open areas (cover < 50%).
 As temperatures warmed, elk and boreal caribou shifted towards more forest cover closer to 50%, while southern mountain caribou and grizzly bears selected for areas with 50% forest cover or less.
@@
 Elk, mountain goats, and cougars moved to higher elevations with temperature, while wolves, southern mountain caribou, and grizzly bears moved to lower elevations.
-Wolves, elk, and southern mountain caribou moved closer to water with temperature, while mountain goats, cougars, and grizzly bears moved somewhat further away from water but remained mainly within 5-10 of water.
-As with movement rates, estimated RSS was generally most uncertain at extreme temperatures, for which data was scarcer (Fig.
+Wolves, elk, and southern mountain caribou moved closer to water with temperature, while mountain goats, cougars, and grizzly bears moved somewhat further away from water but remained mainly within 5-10 km of water.
+As with movement rates, estimated RSS was generally most uncertain at extreme temperatures, for which data were scarcer (Fig.
 B15).
@@
 B9 and B10).
-Estimated changes in average distance traveled were negligible for boreal caribou, and grizzly bears, although both species showed seasonal changes in seasonal movement rates.
-As temperatures warmed, boreal caribou moved further in winter, spring, and fall but less in summer, while grizzly bears showed a clear shift towards earlier emergence from hibernation (Fig.
+Estimated changes in average yearly distance traveled were negligible for boreal caribou and grizzly bears, although both species showed seasonal changes in seasonal movement rates.
+As temperatures warmed, boreal caribou were predicted to move more in winter, spring, and fall but less in summer (Fig.
+\@ref(fig:dist)), while grizzly bears were predicted to showed a clear shift towards earlier emergence from hibernation (Fig.
 B4) and greater movement earlier in the year but less movement in early fall.
@@
 
-Absolute relative changes in average yearly distance travelled were small under the best-case SSP (0-4% change in 2100 relative to 2025).
+Absolute relative changes in predicted yearly distance travelled were small under the best-case SSP (0-4% change in 2100 relative to 2025).
 Under the worst-case SSP, absolute changes by 2100 (relative to 2025) ranged from ~2% (grizzly bears) to ~24% (cougars).
@@
 
-```{r t-dist, fig.cap="\\textbf{Species will alter their movement rates differently in response to climate change, but worse climate-change scenarios will result in the greatest change.} Bold lines indicate the median change in distance travelled (probability of moving times speed when moving) due to predicted changes in temperature within each species' observed extent.
+```{r t-dist, fig.cap="\\textbf{Species are predicted to alter their movement rates differently in response to climate change, and worse climate-change scenarios will result in the greatest change.} Bold lines indicate the median change in yearly distance travelled (probability of moving times speed when moving) due to predicted changes in temperature within each species' sutdy area.
 Shaded areas indicate the $95^\\text{th}$ and $5^\\text{th}$ percentiles.
@@
 
-```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently, but changes will also vary spatially.} The color of each pixel indicates the predicted changes in distance traveled in 2100 at that pixel, relative to the value in 2025, averaged across all four scenarios.
+```{r s-dist, fig.cap="\\textbf{Climate change is predicted to impact each species' movement rates differently, but changes will also vary spatially.} The color of each pixel indicates the predicted changes in distance traveled in 2100 at that pixel, relative to the value in 2025, averaged across all four scenarios.
 Values < 1 indicate a decrease, and values > 1 indicate an increase.
 For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling).
-The predictions only account for the predicted temperature throughout the areas and ignore environmental factors such as terrain slope, soil type, and forest density.
-All maps extend to each species' observed extent (Fig.
+The predictions only account for the predicted temperature throughout the areas and ignore environmental factors such as terrain slope, soil type, and forest cover.
+All maps extend to each species' sutdy area (Fig.
 1) and are shown in the BC Albers Equal Area Conic projection (EPSG:3005)."}
@@
 
-```{r t-rss-cc, fig.cap="\\textbf{Species will alter their habitat selection differently in response to climate change, but worse climate-change scenarios will result in the greatest change.} Bold lines indicate the change in median RSS due to predicted changes in temperature within each species' observed extent.
+```{r t-rss-cc, fig.cap="\\textbf{Species are predicted to alter their habitat selection differently in response to climate change, and worse climate-change scenarios will result in the greatest change.} Bold lines indicate the change in median RSS due to predicted changes in temperature within each species' sutdy area.
 Shaded areas indicate the $95^\\text{th}$ and $5^\\text{th}$ percentiles in change in RSS.
@@
 For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.8 and 1.25 ($\\approx 2^{\\pm 0.322}$; original data ranged 0.71 to 1.93).
-All maps extend to each species' observed extent (Fig.
-1) and are shown in the BC Albers Equal Area Conic projection (EPSG:3005)."}
+All maps extend to each species' study area and are shown in the BC Albers Equal Area Conic projection (EPSG:3005)."}
 if(FALSE) range(readRDS('../data/cc-hrsf-projections-local-2100.rds')$l)
@@
 
-```{r density-hrsfs, fig.cap="\\textbf{Climate change will cause species to decrease their selection strength for their current habitats, but worse climate change scenarios will have the greatest impact.} The density plots indicate each species' change in RSS in 2100 for their current habitat (Fig.
+```{r density-hrsfs, fig.cap="\\textbf{Climate change is predicted to cause species to decrease their selection strength for their current habitats, and worse climate change scenarios will have the greatest impact.} The density plots indicate each species' change in RSS in 2100 for their current habitat (Fig.
 8), relative to each location's RSS in 2025.
@@
 In particular, tensor product interaction terms were crucial for estimating smooth, nonlinear changes in daily and seasonal trends with temperature with data from multiple individuals.
-By allowing interaction terms to be smooth bivariate functions, we were able to quantify changes in movement behaviour without the need to discretize time into distinct windows (e.g., day/night and seasons).
+By allowing interaction terms to be smooth bivariate functions, we were able to quantify changes in movement behaviour without the need to discretize time into windows (e.g., day/night and seasons).
 We were thus able to present results using relatively simple and interpretable surface plots (Figs.
@@
 Species varied greatly in how they adapted their movement behaviour to temperature.
-For boreal caribou, hotter temperatures increased movement rates in fall, winter, and spring but decreased them in summer, mirroring the findings of @stien_congruent_2012, @leclerc_determinants_2021, and @lessard_local_2025.
-The increase in movement rates during the colder seasons are likely partly due to shallower snow depth that results in greater mobility [@pedersen_quantifying_2021], but rising temperatures during snowy seasons also increase the risk of rain-on-snow events, which limit forage availability and increase time spent foraging [@stien_congruent_2012; @berger_climate_2018; @mezzini_how_2025].
+For boreal caribou, hotter temperatures increased movement rates in fall, winter, and spring but decreased them in summer, mirroring previous studies [@stien_congruent_2012; @leclerc_determinants_2021; @lessard_local_2025].
+The increase in movement rates with temperature during the cold months is likely partly due to shallower snow depth that results in greater mobility [@pedersen_quantifying_2021], but warmer temperatures during snowy seasons also increase the risk of rain-on-snow events, which limit forage availability and increase time spent foraging [@stien_congruent_2012; @berger_climate_2018; @mezzini_how_2025].
 The reduced movement rates during warmer summers are likely mainly due to more frequent resting and use of thermal refugia, given ungulates' elevated risk of hyperthermia [@alston_temperature_2020; @verzuh_behavioural_2023].
@@
 Earlier growth seasons in boreal and arctic regions have resulted in lower calf birth and survival [@post_climate_2008], while the lengthening of the growth season has allowed moose (*Alces alces*) and deer to encroach on boreal caribou habitat and increase the density of coyotes (*Canis latrans*), cougars, and wolves [@barber_potential_2018; @demars_incorporating_2023], whose movement behaviour also depends on ambient temperature.
-Boreal wolves responded to temperature similarly to boreal wolves, and habitat selection strength was not projected to change noticeably by 2100, but wolves were predicted to move less in future decades.
-In contrast, cougars showed markedly different seasonal responses to temperature, with relatively warmer temperatures always resulting in less travel, which resulted in strong predicted decreases in movement rates over the 21^st^ century.
+Boreal wolves responded to temperature similarly to boreal caribou, and habitat selection strength was not projected to change noticeably by 2100, but wolves were predicted to move less in future decades.
+In contrast, cougars showed markedly different seasonal responses to temperature, with warmer temperatures always resulting in less travel, which resulted in strong predicted decreases in movement rates over the 21^st^ century.
 However, spatial heterogeneity in habitat, climate, and cougar physiology and behaviour may cause cougars in boreal regions to move differently from those in our study.
@@
 Considerations about changes in trophic interactions leads to another important caveat about the results we present.
-The estimated changes movement behaviour cannot be fully attributed to physiological responses to temperature alone, since other aspects of habitats' seasonal phenology are (nonlinearly) correlated with temperature.
+The estimated changes in movement behaviour cannot be fully attributed to physiological responses to temperature alone, since other aspects of habitats' seasonal phenology are (nonlinearly) correlated with temperature.
 For example, mountain goats' lower movement rates at temperatures near 0$^\circ$C in winter may be in part due to increased movement difficulty, since melting snow and rain-on-snow events increase the energetic costs of movement [@white_population_2025] and encounter rates with predators [@sullender_defining_2023].
 However, milder temperatures can also increase plant growth, which allows goats to spend more time foraging within and less time travelling across patches [@charnov_optimal_1976; @white_mountain_2025].
-Therefore, figure \@ref(fig:dist) illustrates an estimate of the total effects of temperature on each species movement rates, but it does not explicitly account for energetic costs.
+Therefore, figure \@ref(fig:dist) illustrates an estimate of the total effects of temperature on each species' movement rates, but it does not explicitly account for energetic costs.
 Similarly, species' changes in habitat selection do not explicitly account for any changes in physiological or energetic costs.
 Additionally, forest cover, elevation, and distance from water are often correlated, so the estimated effects should be interpreted carefully, and they may not be applicable outside each species' study area.
-For example, both mountain goats and elk selected for higher elevation as temperatures warmed, but our models did not explicitly account for differences in forage availability at different altitudes.
+For example, both mountain goats and elk selected for higher elevation as temperatures warmed, but our models did not explicitly account for differences in forage availability at different elevations.
 

### 2025-10-03 analysis/figures/telemetry-ud-map.R
added inset in fig 1; rm unnecessary code

@@ -6,6 +6,7 @@ library('ctmm')    # for working with akdes
 library('ggplot2') # for fancy plots
 library('scales')  #' for `parse_format()`
 library('terra')   # for rasters
+library('cowplot') # for inset
 source('analysis/figures/default-ggplot-theme.R')
 source('data/bc-shapefile.R')
 
@@ -13,7 +14,7 @@ d <- readRDS('models/movement-models-akdes-2024-06-06.rds') %>%
   mutate(dataset_name = case_when(
     dataset_name == 'Canis_lupus_boreal' ~ 'Wolves',
     dataset_name == 'Rangifer_tarandus_boreal' ~ 'Caribou (boreal)',
-    dataset_name == 'Rangifer_tarandus_southern_mountain' ~ 'Caribou (southern mountain)',
+    dataset_name == 'Rangifer_tarandus_southern_mountain' ~ 'Caribou (s. mountain)',
     dataset_name == 'Puma_concolor_2' ~ 'Cougars',
     dataset_name == 'Puma_concolor_4' ~ 'Cougars',
     dataset_name == 'Elk in southwestern Alberta' ~ 'Elk',
@@ -34,25 +35,11 @@ tels <- transmute(d,
   unnest(tel) %>%
   st_as_sf()
 
-uds <- d %>%
-  transmute(
-    dataset_name,
-    akde =  map(akde, \(.a) {
-      SpatialPolygonsDataFrame.UD(.a, level = 0, level.UD = 0.95) %>%
-        st_as_sf() %>%
-        st_set_crs(.a@info$projection) %>%
-        filter(grepl(pattern = 'est', x = name)) %>%
-        st_transform('EPSG:3005')
-    }) %>%
-      bind_rows() %>%
-      pull(geometry))
-
 # create the figures ----
 us <- tigris::states(cb = TRUE, resolution = '5m') %>%
   st_geometry() %>%
   st_transform(crs(bc)) %>%
   st_as_sf()
-plot(us)
 
 na <- rbind(prov, us)
 
@@ -84,11 +71,26 @@ p_dem <-
   scale_fill_distiller(name = '\nElevation (m)', palette = 6) +
   labs(x = NULL, y = NULL) +
   theme_void() +
-  theme(legend.position = 'inside', legend.position.inside = c(-3, 0.05),
+  theme(legend.position = 'inside', legend.position.inside = c(-4.5, 0.2),
         legend.justification = c(1, 0), text = element_text(face = 'bold')) +
[…]

### 2025-10-03 analysis/figures/temperature-hrsfs.R
increased font size

@@ -205,11 +205,11 @@ p <-
                   limits = c(-LIM, LIM), breaks = -LIM:LIM,
                   labels = \(x) 2^x) +
   theme(strip.placement = 'outside', strip.background.y = element_blank(),
-        strip.text.y = element_text(size = 11), legend.position = 'top',
+        strip.text = element_text(size = 11), legend.position = 'top',
         panel.background = element_rect(fill = 'grey90'),
-        legend.key.width = rel(2))
+        legend.key.width = rel(2), legend.text = element_text(size = 11))
 
-ggsave('figures/hrsf-surface-plots.png', p, width = 17.5, height = 8,
+ggsave('figures/hrsf-surface-plots.png', p, width = 16, height = 7.2,
        units = 'in', dpi = 600, bg = 'white'); beepr::beep()
 
 # figure of standard error on log link scale ----
@@ -229,9 +229,9 @@ p_se <-
   scale_y_continuous(NULL, expand = c(0, 0)) +
   scale_fill_lajolla(name = 'Standard error in log(RSS)') +
   theme(strip.placement = 'outside', strip.background.y = element_blank(),
-        strip.text.y = element_text(size = 11), legend.position = 'top',
+        strip.text = element_text(size = 11), legend.position = 'top',
         panel.background = element_rect(fill = 'grey90'),
-        legend.key.width = rel(2))
+        legend.key.width = rel(2), legend.text = element_text(size = 11))
 
-ggsave('figures/hrsf-surface-plots-se.png', p_se, width = 17.5, height = 8,
+ggsave('figures/hrsf-surface-plots-se.png', p_se, width = 16, height = 7.2,
        units = 'in', dpi = 600, bg = 'white')

### 2025-10-03 writing/manuscript.Rmd
addressed more of the comments from KH

@@
 We address this gap by quantifying the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species (boreal and southern mountain caribou, cougars, elk, grizzly bears, mountain goats, and wolves) throughout western Canada between 1998 and 2023.
-We show that temperature significantly affected how and where these mammals moved, and we predict that climate change will impact mammals' movement behaviour throughout the 21^st^ century, although species did not respond uniformly to temperature.
+We show that temperature significantly affected how and where these mammals moved, and we predict that climate change will impact mammals' movement behaviour throughout the 21^st^ century.
 Projected responses to climate change suggested that rising temperatures will drive southern mountain caribou and mountain goats to move more, while cougars, elk, and wolves will move less.
 Boreal caribou and grizzly bears showed little change in projected yearly movement rates but clear shifts in seasonal phenology.
-We also predict that rising temperatures will reduce median habitat selection strength for most of the six species, but some are expected to show increased selection strength for higher altitudes.
+We also predict that rising temperatures will reduce median habitat selection strength for four of the species, although cougars and elk are expected to show increased selection strength for higher altitudes.
 As mammals increasingly alter their movement rates and select against portions of their current ranges, changes in individuals' movement behaviour will impact encounter rates, including predator-prey dynamics and human-wildlife interactions.
@@
 Animals may search for a short-term buffer from extreme temperatures by moving to thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], and shelter may even be preferred over forage abundance [@hall_microrefuges_2016].
-However, climate change during the last century [@ipcc_climate_2023] has caused animals to alter how and where they move as they respond to hotter temperatures and increasingly severe and frequent extreme-heat events [@ipcc_climate_2023; @bunnell_global_2011].
+However, climate change during the last century [@ipcc_climate_2023] has caused animals to alter how and where they move as they respond to hotter temperatures and increasingly severe and frequent extreme-heat events [@ipcc_climate_2023; @bunnell_global_2011; @thompson_2021_2022].
 
@@
 In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves; Table \@ref(tab:data-table)).
-Western Canada is currently experiencing accelerating and widespread but spatially heterogeneous warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [@tysor_countergradient_2025; @basu_phenological_2024; @kienzle_has_2018], and more frequent and intense extreme heat events [@zhang_increased_2023].
+Western Canada is currently experiencing accelerating and widespread but spatially heterogeneous warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [plants: @post_climate_2008; @tysor_countergradient_2025; ice: @basu_phenological_2024], and more frequent and intense extreme heat events [@zhang_increased_2023].
 Decision makers must understand which areas will best sustain and protect biodiversity not only for current conditions but future decades, too.
@@
 
-```{r bc-map, fig.cap="GPS telemetry data for the six species in this study (dots) over a digital elevation model for the study area.", out.width='0.75\\linewidth'}
+```{r bc-map, fig.cap="GPS telemetry data for the six species in this study (dots) over a digital elevation model for the study area.
+The inset in the bottom left shows the telemetry data within the north american continent.
+Both maps use the BC Albers Equal Area Conic projection (EPSG:3005).", out.width='\\linewidth'}
 knitr::include_graphics('../figures/tels-map-dem.png')
@@
 B8).
+All species' estimated probability of moving and speed when moving decreased with sampling intervals above 1 hour, except for cougars' speed, although the estimated trends were highly uncertain (Fig.
+B8).
 Consequently, we present all results while predicting specifically for one-hour sampling intervals.
@@
   d = p * s) %>%
-  rename('P(M=1|T=0^\\circ C)' = p,
-         'E(S|M=1,T=0^\\circ C)' = s,
-         'E(D|T=0^\\circ C)' = d) %>%
+  rename('P(M=1|T)' = p,
+         'E(S|M=1,T)' = s,
+         'E(D|T)' = d) %>%
   mutate(Species = COMMON_NAMES) %>%
@@
                 'T>+20\\textdegree{}C (%)' = get_perc(temp_c > +20))) %>%
-  knitr::kable(format = 'markdown', digits = 2, format.args = list(big.mark = ','), caption = "Summary statistics for each species' GPS fixes with finite speed estimates, namely: the number fixes after data cleaning ($n$), the percentage of fixes with temperature ($T$) below $-20^\\circ$C and above $20^\\circ$C, the estimated mean probability of moving ($\\hat P(M=1)$), the mean speed when moving ($\\hat{\\mathbb E}(S|M=1)$; km/h), and the mean hourly distance travelled ($\\hat P(M=1) \\times \\hat{\\mathbb E}(S|M=1) = \\hat{\\mathbb E}(D)$; km/h), for a sampling interval of 1 hour and a temperature $T$ of $0^\\circ$C.")
+  knitr::kable(format = 'markdown', digits = 2, format.args = list(big.mark = ','), caption = "Summary statistics for each species' GPS fixes with finite speed estimates, namely: the number fixes after data cleaning ($n$), the percentage of fixes with temperature ($T$) below $-20^\\circ$C and above $20^\\circ$C, the estimated mean probability of moving ($\\hat P(M=1)$), the mean speed when moving ($\\hat{\\mathbb E}(S|M=1)$; km/h), and the mean hourly distance travelled ($\\hat P(M=1) \\times \\hat{\\mathbb E}(S|M=1) = \\hat{\\mathbb E}(D)$; km/h), for a sampling interval of 1 hour and a temperature of $T = 0^\\circ$C.")
 ```
@@
 Throughout the year, they tended to move more when it was colder, but what they perceived as "colder" was relative to the time of year.
-All species' estimated probability of moving and speed when moving decreased with sampling intervals above 1 hour, except for cougars' speed, although the estimated trends were highly uncertain (Fig.
-B8).
-Uncertainties around the estimated effects were generally higher at extreme temperatures due to lower data availability (Figs.
+Overall, uncertainty around the estimated effects was generally higher at extreme temperatures due to lower data availability (Figs.
 B4A, B5A, and B6A).
@@
 The grey contours indicate the extent of each species' observed data.
-RSS values were re-centered and capped to $2^{\\pm 2}$ to improve readability."}
+RSS values were re-centered and capped to between 0.25 and 4 ($2^{\\pm 2}$) to improve readability."}
 knitr::include_graphics('../figures/hrsf-surface-plots.png')
@@
 
-Median RSS was projected to decrease over time within the each species' observed range, but, again, changes were stronger under worse SSPs (Fig.
+Median RSS was projected to decrease over time within each species' observed range, but, again, changes were stronger under worse SSPs (Fig.
 \@ref(fig:t-rss-cc)).
-Decreases were most pronounced in areas with the lowest RSS and most severe for elk, mountain goats, pumas, and southern mountain caribou.
-Changes for boreal wolves and caribou were relatively negligible.
+Decreases were most pronounced in areas with the lowest RSS and most severe for elk, mountain goats, cougars, and southern mountain caribou.
+Changes for boreal wolves and caribou were negligible.
 Elk, cougars, and grizzly bears were predicted to increase their selection strength for top-RSS areas, and elk, mountain goats, cougars, and southern mountain caribou were predicted to further decrease their selection for areas with low RSS.
-Unsurprisingly, the change in RSS between 2025 and 2100 also varied spatially for all species (Fig.
+Unsurprisingly, the predicted change in RSS between 2025 and 2100 also varied spatially for all species (Fig.
 \@ref(fig:s-rss-cc)).
@@
 This framework was essential for estimating how mammals behave at extreme temperatures.
-In fact, despite the recent increase in frequency of extreme heat events in western Canada [@zhang_increased_2023; @bunnell_global_2011; @kienzle_has_2018], such events remain relatively rare, so estimating mammals' nonlinear responses required the flexibility provided by GAMs [@wood_generalized_2017].
-Additionally, hierarchical models allowed us to produce statistically sound estimates that accounted for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mccabe_resource_2021; @jeltsch_need_2025; @silva_too_2025].
+In fact, despite the recent increase in frequency of extreme heat events in western Canada, including a severe heat wave in 2021 [@zhang_increased_2023; @thompson_2021_2022; @kienzle_has_2018], such events remain relatively rare, so estimating mammals' nonlinear responses required the flexibility provided by HGAMs [@wood_generalized_2017].
+Additionally, the hierarchical approach allowed us to leverage data from multiple individuals in a single model while accounting for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mccabe_resource_2021; @jeltsch_need_2025; @silva_too_2025].
 Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously, so we suggest choosing each smooth term's basis size (`k`) carefully.
-While penalized splines and REML can help avoid unruly smooth terms, the penalty terms can struggle to restrict model complexity when datasets are particularly large [@wood_generalized_2017; @wood_generalized_2017-1].
+While penalized splines and REML can help avoid unruly smooth terms, the penalty terms can struggle to restrict model complexity when datasets are as large as those presented here [@wood_generalized_2017; @wood_generalized_2017-1].
 
-Species varied greatly in how they adapted their movement behaviour to temperature.
-For boreal caribou, hotter temperatures increased movement rates in fall, winter, and spring but decreased them in summer, mirroring previous studies [@stien_congruent_2012; @leclerc_determinants_2021; @lessard_local_2025].
-The increase in movement rates with temperature during the cold months is likely partly due to shallower snow depth that results in greater mobility [@pedersen_quantifying_2021], but warmer temperatures during snowy seasons also increase the risk of rain-on-snow events, which limit forage availability and increase time spent foraging [@stien_congruent_2012; @berger_climate_2018; @mezzini_how_2025].
+The effects of warming temperatures on movement rates varied across species, seasons, and space due to differences in species' behavior and physiology as well as their habitats' phenology.
+For boreal caribou, hotter temperatures decreased movement rates summer but increased them otherwise, mirroring previous studies [@stien_congruent_2012; @leclerc_determinants_2021; @lessard_local_2025].
+The increase in movement rates with temperature during the cold months is likely partly due to shallower snow depth, which results in greater mobility [@pedersen_quantifying_2021], but warmer temperatures during snowy seasons also increase the risk of rain-on-snow events, which limit forage availability and increase time spent foraging [@stien_congruent_2012; @berger_climate_2018; @mezzini_how_2025].
 The reduced movement rates during warmer summers are likely mainly due to more frequent resting and use of thermal refugia, given ungulates' elevated risk of hyperthermia [@alston_temperature_2020; @verzuh_behavioural_2023].
@@
 Earlier growth seasons in boreal and arctic regions have resulted in lower calf birth and survival [@post_climate_2008], while the lengthening of the growth season has allowed moose (*Alces alces*) and deer to encroach on boreal caribou habitat and increase the density of coyotes (*Canis latrans*), cougars, and wolves [@barber_potential_2018; @demars_incorporating_2023], whose movement behaviour also depends on ambient temperature.
+
 Boreal wolves responded to temperature similarly to boreal caribou, and habitat selection strength was not projected to change noticeably by 2100, but wolves were predicted to move less in future decades.
 In contrast, cougars showed markedly different seasonal responses to temperature, with warmer temperatures always resulting in less travel, which resulted in strong predicted decreases in movement rates over the 21^st^ century.
-However, spatial heterogeneity in habitat, climate, and cougar physiology and behaviour may cause cougars in boreal regions to move differently from those in our study.
+However, spatial heterogeneity in habitat, climate, and cougar behaviour may cause boreal individuals to respond to temperature differently from those in our study.
 Still, rising temperatures clearly affect more than each individual species' movement behaviour; they have complex and interconnected effects on entire communities and their trophic structures.
@@
 For example, as temperatures warm and both boreal caribou and wolves increase their movement rates in fall, winter, and spring, they also increase their encounter rates [@martinez-garcia_how_2020], which will likely increase predation risk when caribou are pregnant or with young calves.
-Additionally, while both species saw reduced movement rates during hot summer days, boreal wolves' use of anthropogenic linear features (e.g., roads, seismic lines) may allow them to reduce the thermal costs of movement while increasing the chances of encountering heat-stressed prey [@whittington_caribou_2011; @dickie_faster_2017; @dickie_resource_2022].
-Furthermore, caribou that attempt to reduce predation risk from wolves by avoiding wolf habitat may still risk increasing predation pressure from bears [@leblond_caribou_2016] and the aforementioned encroaching predators [@demars_incorporating_2023; @labadie_global_2023].
+Additionally, while both species saw reduced movement rates during hot summer days, boreal wolves' use of anthropogenic linear features (e.g., roads, seismic lines) may allow them to reduce the total thermal costs of movement by moving for shorter periods while increasing the chances of encountering heat-stressed prey [@whittington_caribou_2011; @dickie_faster_2017; @dickie_resource_2022].
+Furthermore, caribou that attempt to reduce predation risk from wolves by avoiding wolf habitat may still risk increasing predation pressure from bears, cougars, and coyote [@leblond_caribou_2016; @demars_incorporating_2023; @labadie_global_2023].
 
@@
 Similarly, species' changes in habitat selection do not explicitly account for any changes in physiological or energetic costs.
-Additionally, forest cover, elevation, and distance from water are often correlated, so the estimated effects should be interpreted carefully, and they may not be applicable outside each species' study area.
-For example, both mountain goats and elk selected for higher elevation as temperatures warmed, but our models did not explicitly account for differences in forage availability at different elevations.
+Additionally, forest cover, elevation, and distance from water are often correlated, so the estimated effects should be interpreted carefully, and they may not be applicable outside the study areaes used here.
+For example, both mountain goats and elk selected for higher elevation as temperatures warmed, but our models did not account for differences in forage availability at different elevations.
 
 Climate change has also had extensive impacts on plant growth and phenology [@cleland_shifting_2007; @denny_performance_2019; @tysor_countergradient_2025] with consequences on herbivores' behaviour and fitness [@post_climate_2008; @aikens_greenscape_2017].
-Forage quality and availability are often a limiting factor for herbivores' ability to escape hotter temperatures by moving to higher altitudes or latitudes, since plants are slower to arrive in and establish in new habitats, and the advance of herbivores can further limit plants' ability to establish [@speed_elevational_2012; @diffenbaugh_changes_2013].
+Additionally, plants' limited ability to disperse to and establish in new habitats is often a limiting factor for herbivores' ability to escape hotter temperatures by moving to higher altitudes or latitudes, and the advance of herbivores can further limit plants' ability to establish [@speed_elevational_2012; @diffenbaugh_changes_2013].
 Consequently, mammals that move to higher elevations may be forced to spend more time foraging [@mezzini_how_2025], which will increase both their energetic expenditure and encounter rates with predators and competitors [@kohl_prey_2019; @martinez-garcia_how_2020; @torrez-herrera_monkeys_2020; @smith_densitydependent_2023].
 If local peaks become unsuitable habitat, and moving to nearby higher peaks requires traveling across low-quality and high-risk habitat (Figure \@ref(fig:s-rss-cc)), habitat fragmentation may prevent movement across ranges [@white_mountain_2025].
-Translocating animals to locations with higher peaks may help avoid such a scenario, but the success of relocations is species-dependent, and attempts are not always successful [@ranc_memory_2022; @white_mountain_2025].
 
 Growing evidence suggests that mammals' exposure, sensitivity, and adaptability to climate change will limit their ability to adapt to change [@ameca_y_juarez_natural_2012; @diffenbaugh_changes_2013].
-Exposure to warming temperatures is widespread throughout western Canada [@bunnell_global_2011; @kienzle_has_2018; @eum_potential_2023], and changes in temperature and seasonal phenology may be too rapid for many species to respond to [@diffenbaugh_changes_2013].
+Exposure to warming temperatures is widespread throughout western Canada [@bunnell_global_2011; @kienzle_has_2018; @eum_potential_2023], and changes in temperature and phenology may be too rapid for many species to respond to [@diffenbaugh_changes_2013].
 Species' adaptability is further challenged by the difficulty in responding to concurrent, multifarious, and interconnected changes [@diffenbaugh_changes_2013; @polazzo_measuring_2024].
@@
 Therefore, it is crucial for conservationists to distinguish between a species' ideal environmental space and its best available geographic space [*sensu* @matthiopoulos_defining_2022].
+
 Accelerating multifarious change in species' movement behaviour and energetic costs require conservation measures that are proactive and anticipate future changes.
@@
 Estimating responses to multifarious change is crucial in assessing mammals' responses to hotter temperatures, altered phenology, and novel community dynamics.
+
 Protected areas must support populations not only in present conditions but also as species' needs and behaviour change over the decades [@simmons_tough_2021; @livingstone_integrative_2023].
@@
 Future work should develop models that account for covariates that are specific to the study species and area, such as forest age and disturbance [@demars_incorporating_2023], terrain ruggedness [@white_mountain_2017], anthropogenic linear features [@dickie_faster_2017], and growing human presence in wild spaces [@sih_evolution_2011; @rice_conspicuous_2022], including the growing risk of human-wildlife conflict [@abrahms_climate_2023; @weststrate_how_2024].
-In particular, forests in western Canada are highly dynamic over time [@zhang_half-century_2015], and fires size and burn severity have increased in western Canada in recent decades [@zurowski_summer_2023; @parisien_abrupt_2023].
-Thus, while our use of a static raster of forest cover allowed for simpler models and climate change projections, future studies should account for mammals' responses to temporal forest dynamics, including the impacts of fire [@zurowski_summer_2023; @parisien_abrupt_2023], logging [@lochhead_linking_2022] and other extractive industries [@dickie_faster_2017; @dickie_habitat_2024].
+In particular, forests in western Canada are highly dynamic over time [@zhang_half-century_2015], and fires size and burn severity have increased in western Canada in recent decades [@whitman_climate-induced_2022; @parisien_abrupt_2023; @wang_canadian_2025].
+Thus, while our use of a static raster of forest cover allowed for simpler models and climate change projections, future studies should account for mammals' responses to temporal forest dynamics, including the impacts of fire [@whitman_climate-induced_2022; @parisien_abrupt_2023; @wang_canadian_2025], logging [@lochhead_linking_2022] and other extractive industries [@dickie_faster_2017; @dickie_habitat_2024].
 However, the inclusion of temporally dynamic variables will require that any future projections of species' behaviour depend on temporally dynamic estimates of how such variables will change over time with changes in climate and anthropogenic activity.
@@
 We have demonstrated that temperature has complex and nonlinear effects on how and where mammals move.
-Rising temperatures have impacted mammals' daily and seasonal movement phenology, and changes are expected to intensify throughout the 21^st^ century.
+Rising temperatures have impacted mammals' daily and seasonal movement cycles, and changes are expected to intensify throughout the 21^st^ century.
 Quantifying how temperature affects mammalian behavior and fitness is a first step towards developing an understanding of the effects of climate change on landscape dynamics and the effectiveness of protected areas in the future.

### 2025-10-14 analysis/figures/temperature-hrsfs-caribou-wolf-habitat.R
changed "tree cover" to "forest cover"

@@ -68,7 +68,7 @@ p_resources <-
       geom_raster() +
       scale_x_continuous(NULL, expand = c(0, 0)) +
       scale_y_continuous(NULL, expand = c(0, 0)) +
-      scale_fill_gradient('Tree cover (%)', low = 'white', na.value = NA,
+      scale_fill_gradient('Forest cover (%)', low = 'white', na.value = NA,
                           high = 'darkgreen', limits = c(0, 100),
                           breaks = c(0, 100)) +
       theme(legend.position = 'top'),
@@ -88,11 +88,11 @@ p_resources <-
       scale_x_continuous(NULL, expand = c(0, 0)) +
       scale_y_continuous(NULL, expand = c(0, 0)) +
       scale_fill_distiller(expression(bold(atop(Distance~from,
-                                                water~(km)~phantom(om)))),
+                                                water~'(km)'~phantom(om)))),
                            na.value = NA, values = c(0, 0.05, 1),
                            limits = c(0, 18), breaks = c(0, 18)) +
       theme(legend.position = 'top'),
-    nrow = 1, rel_widths = c(1, 1, 1.025))
+    nrow = 1)
 
 # predict habitat selection strength at 20 degrees C ----
 preds_habitat <-

### 2025-10-14 writing/manuscript.Rmd
added habitat figure for caribou and wolves

@@
 We then estimated each mammal's instantaneous speed at each observed location by applying the `ctmm::speeds()` function on all models with finite speed estimates [415 of 433$;$ @fleming_fine-scale_2014; @noonan_scale-insensitive_2019].
-The remaining 18 movement models had sampling rates that were too coarse, relative to the animals' directional persistance, to reconstruct the animals' movement trajectories [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @denicola_are_2025].
+The remaining 18 movement models had sampling rates that were too coarse, relative to the animals' directional persistence, to reconstruct the animals' movement trajectories [@noonan_scale-insensitive_2019; @nathan_big-data_2022; @denicola_are_2025].
 The models were for one mountain goat, 15 boreal caribou, and two southern mountain caribou (Table \@ref(tab:data-table)).
@@
 
+The three habitat variables were nonlinearly correlated.
+Within the boreal caribou range (Fig.
+\@ref(fig:boreal-rss)A), locations with an elevation of 0.9-1.1 km had a median forest cover of 100% and a median distance to water of 1.8 km.
+In contrast, elevations above 1750 m did not have forest cover above 90% and had a median distance to water of 1.5 km.
+Local predictions of RSS revealed that boreal caribou and wolves selected for similar habitats (Fig.
+\@ref(fig:boreal-rss)B), but caribou tended to avoid the river system that wolves strongly selected for.
+Both species selected for the same loop of low-elevation habitat (< 500 m), but their selection strength for the area varied with temperature.
+For caribou, selection for the area was stronger at -20$^\circ$C and weaker at warmer temperatures, while, for wolves, selection was stronger at +20$^\circ$C.
+Both species' HRSFs indicated selection against elevations approximately between 0.5 km and 0.9 km but selection for elevations approximately above 0.9 km.
+After data cleaning and the removal of non-range-resident animals, the highest observed elevations were 1 km for caribou and 1.733 km for wolves.
+Quadrature points had maximum elevations of 1.1 km for caribou and 1.3 km for wolves.
+
+```{r, eval=FALSE, echo=FALSE}
+median(filter(resources, elevation_m > 900, elevation_m < 1100)$forest_perc)
+median(filter(resources, elevation_m > 900, elevation_m < 1100)$dist_water_m)
+
+max(filter(resources, forest_perc > 90)$elevation_m)
+median(filter(resources, elevation_m > 1750)$dist_water_m)
+
+ggplot(resources, aes(elevation_m, forest_perc)) + geom_hex(bins = 20)
+ggplot(resources, aes(elevation_m, dist_water_m)) + geom_hex(bins = 20)
+ggplot(resources, aes(dist_water_m, forest_perc)) + geom_hex(bins = 20)
+
+e_rast <- rast('data/resource-rasters/bc-buffered-dem-z6.tif') %>%
+    crop(st_buffer(st_transform(bounds, crs(.)), 1e5)) %>%
+    project(f_rast) %>%
+    crop(bounds)
+
+plot(e_rast < 500)
+plot(e_rast < 900)
+plot(e_rast < 1000)
+
+d %>%
+  filter(species %in% c('Rangifer tarandus boreal', 'Canis lupus')) %>%
+  summarize(elevation_m = max(elevation_m), .by = c(species, detected))
+```
+
+```{r boreal-rss, fig.cap="\\textbf{Temperature affects boreal caribou's and wolves' habitat selection and, consequently, their encounter rates.} Panel A shows the rasters of the three habitat variables used in the hierarchical resource selection functions, while panel B shows the relative selection strength (RSS) for boreal caribou and wolves, as a function of temperature.
+RSS values were re-centered and capped to between 0.25 and 4 ($2^{\\pm 2}$) to improve readability."}
+knitr::include_graphics('../figures/boreal-caribou-wolf-habitat.png')
+```
+
 ## Predicted changes in movement behaviour during the 21^st^ century
@@
 Similarly, species' changes in habitat selection do not explicitly account for any changes in physiological or energetic costs.
-Additionally, forest cover, elevation, and distance from water are often correlated, so the estimated effects should be interpreted carefully, and they may not be applicable outside the study areaes used here.
+Additionally, forest cover, elevation, and distance from water are often correlated, so the estimated effects should be interpreted carefully, and they may not be applicable outside the study areas used here.
 For example, both mountain goats and elk selected for higher elevation as temperatures warmed, but our models did not account for differences in forage availability at different elevations.

### 2025-10-15 analysis/figures/temperature-hrsfs-caribou-wolf-habitat.R
added product of RSSs as proxy of encounter rates

@@ -66,32 +66,32 @@ p_resources <-
     ggplot(resources, aes(x, y, fill = forest_perc)) +
       coord_sf(crs = 'EPSG:3005') +
       geom_raster() +
-      scale_x_continuous(NULL, expand = c(0, 0)) +
+      scale_x_continuous(NULL, expand = c(0, 0), breaks = -c(124, 122, 120)) +
       scale_y_continuous(NULL, expand = c(0, 0)) +
       scale_fill_gradient('Forest cover (%)', low = 'white', na.value = NA,
                           high = 'darkgreen', limits = c(0, 100),
                           breaks = c(0, 100)) +
-      theme(legend.position = 'top'),
+      theme(legend.position = 'top', legend.key.width = rel(0.7)),
     ggplot(resources, aes(x, y, fill = elevation_m / 1e3)) +
       coord_sf(crs = 'EPSG:3005') +
       geom_raster() +
-      scale_x_continuous(NULL, expand = c(0, 0)) +
+      scale_x_continuous(NULL, expand = c(0, 0), breaks = -c(124, 122, 120)) +
       scale_y_continuous(NULL, expand = c(0, 0)) +
       scale_fill_distiller('Elevation (km) ', palette = 6, direction = 1,
                            breaks = round(range(resources$elevation_m / 1e3), 1),
                            labels = round(range(resources$elevation_m / 1e3), 1),
                            limits = round(range(resources$elevation_m / 1e3), 1)) +
-      theme(legend.position = 'top'),
+      theme(legend.position = 'top', legend.key.width = rel(0.7)),
     ggplot(resources, aes(x, y, fill = dist_water_m / 1e3)) +
       coord_sf(crs = 'EPSG:3005') +
       geom_raster() +
-      scale_x_continuous(NULL, expand = c(0, 0)) +
+      scale_x_continuous(NULL, expand = c(0, 0), breaks = -c(124, 122, 120)) +
       scale_y_continuous(NULL, expand = c(0, 0)) +
       scale_fill_distiller(expression(bold(atop(Distance~from,
                                                 water~'(km)'~phantom(om)))),
                            na.value = NA, values = c(0, 0.05, 1),
                            limits = c(0, 18), breaks = c(0, 18)) +
-      theme(legend.position = 'top'),
[…]

### 2025-10-15 writing/manuscript.Rmd
added results for boreal area

@@
     institute: [braes, biol]
-  - name: Chris H. Fleming
+  - name: Christen H. Fleming
     institute: [ucf, smith]
@@
 Projected responses to climate change suggested that rising temperatures will drive southern mountain caribou and mountain goats to move more, while cougars, elk, and wolves will move less.
-Boreal caribou and grizzly bears showed little change in projected yearly movement rates but clear shifts in seasonal phenology.
+Boreal caribou and grizzly bears showed little change in projected yearly movement rates but clear shifts in phenology.
 We also predict that rising temperatures will reduce median habitat selection strength for four of the species, although cougars and elk are expected to show increased selection strength for higher altitudes.
@@
 
-Global warming during the last century has resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extreme-heat events [@ipcc_climate_2023], with widespread changes in mammals' seasonal phenology, fitness, and life history [@botero_evolutionary_2015; @wells_life_2022; @newediuk_climate_2024; @mclellan_effect_2015; @pigeon_drivers_2016].
+Global warming during the last century has resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extreme-heat events [@ipcc_climate_2023], with widespread changes in mammals' phenologies, circadian rhythms, fitness, and life histories [@botero_evolutionary_2015; @wells_life_2022; @newediuk_climate_2024; @mclellan_effect_2015; @pigeon_drivers_2016].
 However, behavioural responses to heat vary substantially across species depending on body size, vagility, and activity times [@mccain_body_2014; @williams_range_2018].
@@
 Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.
-Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and habitat phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022; @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].
+Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022; @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].
 
@@
 probability of moving, speed when moving, and their product: hourly distance traveled).
-The effects of temperature on mammals' movement rates depend on daily and seasonal phenology, since animals may respond to temperatures differently at different times of day and or days of year.
+The effects of temperature on mammals' movement rates depend on circadian rhythm and phenology, since animals may respond to temperatures differently at different times of day and or days of year.
 Additionally, both temperature and phenology depend on time of day and day of year.
-Finally, daily phenology changes with seasonal phenology due to changes in photoperiod (e.g., the time of twilight changes throughout the year, affecting crepuscular activity).
+Finally, circadian rhythm changes with seasonal phenology due to changes in photoperiod (e.g., the time of twilight changes throughout the year, affecting crepuscular activity).
 (B) Similarly, habitat selection depends on the availability and selection of habitat variables (forest cover, elevation, and distance from water), but the selection strengh for each variable is conditional on temperature.
-For example, an animal may select for more densely-forested areas at extreme temperatures.", fig.height=6, fig.width=15, out.width='\\linewidth'}
+For example, an animal may select for more densely-forested areas at extreme temperatures.
+The resource selection functions for (B) also included marginal smooths of temperature to account for sampling biases across seasons.", fig.height=6, fig.width=15, out.width='\\linewidth'}
 library('dplyr')   # for data wrangling
@@
                   bold = c(FALSE, FALSE, TRUE, TRUE, FALSE, FALSE),
-                  label = c('Daily phenology', 'Seasonal phenology', 'Movement rates',
+                  label = c('Circadian rhythm', 'Phenology', 'Movement rates',
                             'Temperature', 'Time of day', 'Day of year'))
@@
             boxes_a, color = 'black', fill = 'white') +
-  geom_path(aes(x, y, group = id), arrows_a, linewidth = 1,
-            arrow = arrow(type = 'closed')) +
+  geom_path(aes(x, y, group = id), arrows_a, linewidth = 0.5,
+            arrow = arrow(type = 'closed', angle = 15, length = unit(0.2, 'in'))) +
   geom_label(aes(x = (xmin + xmax) / 2, y = (ymin + ymax) / 2,
@@
             boxes_b, color = 'black', fill = 'white') +
-  geom_path(aes(x, y, group = id), arrows_b, linewidth = 1,
-            arrow = arrow(type = 'closed')) +
+  geom_path(aes(x, y, group = id), arrows_b, linewidth = 0.5,
+            arrow = arrow(type = 'closed', angle = 15, length = unit(0.2, 'in'))) +
   geom_label(aes(x = (xmin + xmax) / 2, y = (ymin + ymax) / 2,
@@
 
-The three habitat variables were nonlinearly correlated.
+Fig.
+\@ref(fig:boreal-rss) presents the predicted habitat selection for boreal caribou and wolves.
+Overall, both species selected for the same low-elevation area (< 500 m; Fig.
+\@ref(fig:boreal-rss)A-B), but caribou generally avoided the river system that wolves were often found at, and their selection strength varied with temperature.
+As wolves' habitat selection strengthened at warmer temperatures, encounters at 20$^\circ$C were approximately four times more frequent than at -20$^\circ$C, independent of changes in the species' movement rates (Fig.
+\@ref(fig:boreal-rss)C).
+
+<!-- The three habitat variables were often nonlinearly correlated.
+Mean tree cover was above 75% for elevations below 1.1 km and between 1.7 and 1.8 km, while mean distance to water was generally near 2 km but peaked at 7 km at elevations near 1.4 km.
 Within the boreal caribou range (Fig.
-\@ref(fig:boreal-rss)A), locations with an elevation of 0.9-1.1 km had a median forest cover of 100% and a median distance to water of 1.8 km.
+\@ref(fig:boreal-rss)B), locations with an elevation of 0.9-1.1 km had a median forest cover of 100% and a median distance to water of 1.8 km.
 In contrast, elevations above 1750 m did not have forest cover above 90% and had a median distance to water of 1.5 km.
-Local predictions of RSS revealed that boreal caribou and wolves selected for similar habitats (Fig.
-\@ref(fig:boreal-rss)B), but caribou tended to avoid the river system that wolves strongly selected for.
-Both species selected for the same loop of low-elevation habitat (< 500 m), but their selection strength for the area varied with temperature.
 For caribou, selection for the area was stronger at -20$^\circ$C and weaker at warmer temperatures, while, for wolves, selection was stronger at +20$^\circ$C.
@@
 Quadrature points had maximum elevations of 1.1 km for caribou and 1.3 km for wolves.
+-->
 
 ```{r, eval=FALSE, echo=FALSE}
+d_1 <- filter(readRDS('data/tracking-data/rsf-data.rds'), detected == 1)
+
+ggplot(d_1, aes(elevation_m, forest_perc)) +
+  geom_hex(bins = 100) +
+  geom_smooth() +
+  scale_x_continuous(breaks = seq(500, 2000, by = 250)) +
+  scale_fill_viridis_c()
+
+ggplot(d_1, aes(elevation_m, dist_water_m)) +
+  geom_hex(bins = 100) +
+  geom_smooth() +
+  geom_hline(yintercept = 2000, color = 'red') +
+  geom_hline(yintercept = 7000, color = 'red') +
+  scale_x_continuous(breaks = seq(500, 2000, by = 250)) +
+  scale_fill_viridis_c()
+
+#' import `resources` from `analysis/figures/temperature-hrsfs-caribou-wolf-habitat.R`
 median(filter(resources, elevation_m > 900, elevation_m < 1100)$forest_perc)
@@
 
-```{r boreal-rss, fig.cap="\\textbf{Temperature affects boreal caribou's and wolves' habitat selection and, consequently, their encounter rates.} Panel A shows the rasters of the three habitat variables used in the hierarchical resource selection functions, while panel B shows the relative selection strength (RSS) for boreal caribou and wolves, as a function of temperature.
-RSS values were re-centered and capped to between 0.25 and 4 ($2^{\\pm 2}$) to improve readability."}
+```{r boreal-rss, fig.cap="\\textbf{Temperature affects boreal caribou's and wolves' habitat selection and, consequently, their encounter rates.} Panel (A) shows the rasters of the three habitat variables used in the hierarchical resource selection functions, while panel (B) shows the relative selection strength (RSS) for boreal caribou and wolves, as a function of temperature.
+RSS values were re-centered and capped to between 0.25 and 4 ($2^{\\pm 2}$) to improve readability.
+Finally, panel (C) shows the (scaled) product of the RSS values in panel (B) as a proxy four encounter rates between individuals of the two species."}
 knitr::include_graphics('../figures/boreal-caribou-wolf-habitat.png')
@@
 
-The effects of warming temperatures on movement rates varied across species, seasons, and space due to differences in species' behavior and physiology as well as their habitats' phenology.
+The effects of warming temperatures on movement rates varied across species, seasons, and space due to differences in species' behavior and physiology as well as their habitats' phenology (e.g., the timing of green-up).
 For boreal caribou, hotter temperatures decreased movement rates summer but increased them otherwise, mirroring previous studies [@stien_congruent_2012; @leclerc_determinants_2021; @lessard_local_2025].

### 2025-10-15 writing/manuscript.Rmd
specified "boreal mammals" in title and text

@@
 ---
-title: "\\Large Rising temperatures alter how and where mammals move"
+title: "\\Large Rising temperatures alter how and where boreal mammals move"
 author:
@@
 We address this gap by quantifying the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species (boreal and southern mountain caribou, cougars, elk, grizzly bears, mountain goats, and wolves) throughout western Canada between 1998 and 2023.
-We show that temperature significantly affected how and where these mammals moved, and we predict that climate change will impact mammals' movement behaviour throughout the 21^st^ century.
+We show that temperature significantly affected how and where these boreal mammals moved, and we predict that climate change will impact mammals' movement behaviour throughout the 21^st^ century.
 Projected responses to climate change suggested that rising temperatures will drive southern mountain caribou and mountain goats to move more, while cougars, elk, and wolves will move less.
@@
 B3, and B14), but including temperature always resulted in better fits (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; Appendix B).
-Accounting for temperature even resulted in a simpler HRSF for mountain goats (lower model degrees of freedom due to simpler smooth terms) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.
+Accounting for temperature even resulted in a simpler HRSF for mountain goats (lower model degrees of freedom due to simpler random smooth terms) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.
 
@@
 
-\noindent We have demonstrated that temperature is an important driver of how and where large mammals move, and that it affects mammals' seasonal and daily movement behaviour in complex and nonlinear ways.
+\noindent We have demonstrated that temperature is an important driver of how and where large boreal mammals move, and that it affects mammals' seasonal and daily movement behaviour in complex and nonlinear ways.
 However, predicting mammals' responses to climate change remains a challenge, as habitats are warming rapidly and mammals are exposed to increasingly novel and extreme conditions [@ipcc_climate_2023; @diffenbaugh_changes_2013; @botero_evolutionary_2015].
@@
 
-We have demonstrated that temperature has complex and nonlinear effects on how and where mammals move.
+We have demonstrated that temperature has complex and nonlinear effects on how and where boreal mammals move.
 Rising temperatures have impacted mammals' daily and seasonal movement cycles, and changes are expected to intensify throughout the 21^st^ century.

### 2025-10-16 writing/manuscript.Rmd
corrected typo; added a ref

@@
 
-Global warming during the last century has resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extreme-heat events [@ipcc_climate_2023], with widespread changes in mammals' phenologies, circadian rhythms, fitness, and life histories [@botero_evolutionary_2015; @wells_life_2022; @newediuk_climate_2024; @mclellan_effect_2015; @pigeon_drivers_2016].
+Global warming during the last century has resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extreme-heat events [@ipcc_climate_2023], with widespread changes in mammals' phenologies, circadian rhythms, fitness, and life histories [@botero_evolutionary_2015; @le_corre_weather_2016; @wells_life_2022; @newediuk_climate_2024; @mclellan_effect_2015; @pigeon_drivers_2016].
 However, behavioural responses to heat vary substantially across species depending on body size, vagility, and activity times [@mccain_body_2014; @williams_range_2018].
@@
 In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves; Table \@ref(tab:data-table)).
-Western Canada is currently experiencing accelerating and widespread but spatially heterogeneous warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [plants: @post_climate_2008; @tysor_countergradient_2025; ice: @basu_phenological_2024], and more frequent and intense extreme heat events [@zhang_increased_2023].
+Western Canada is currently experiencing accelerating and widespread but spatially heterogeneous warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [plants: @post_climate_2008; @tysor_countergradient_2025; ice: @basu_phenological_2024; caribou: @le_corre_weather_2016], and more frequent and intense extreme heat events [@zhang_increased_2023].
 Decision makers must understand which areas will best sustain and protect biodiversity not only for current conditions but future decades, too.
 Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.
-Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022; @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].
+Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @le_corre_weather_2016; @slatyer_ecological_2022; @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].
 
@@
 RSS values were re-centered and capped to between 0.25 and 4 ($2^{\\pm 2}$) to improve readability.
-Finally, panel (C) shows the (scaled) product of the RSS values in panel (B) as a proxy four encounter rates between individuals of the two species."}
+Finally, panel (C) shows the (scaled) product of the RSS values in panel (B) as a proxy for encounter rates between individuals of the two species."}
 knitr::include_graphics('../figures/boreal-caribou-wolf-habitat.png')

### 2025-10-24 analysis/figures/temperature-hrsfs-caribou-wolf-habitat.R
changed "encounter" to "co-occupancy"

@@ -149,19 +149,19 @@ p_enc <-
   pivot_wider(names_from = species, values_from = r) %>%
   mutate(r = map2(`Caribou (boreal)`, Wolves, \(.c, .w) {
     as.data.frame(.c * .w, xy = TRUE) %>%
-      rename(p_encounter = lambda)
+      rename(cooccupancy = lambda)
   })) %>%
   select(! c(`Caribou (boreal)`, Wolves)) %>%
   unnest(r) %>%
-  mutate(p_encounter = if_else(p_encounter > 4, 4, p_encounter),
-         p_encounter = if_else(p_encounter < 0.25, 0.25, p_encounter)) %>%
-  ggplot(aes(x, y, fill = p_encounter)) +
+  mutate(cooccupancy = if_else(cooccupancy > 4, 4, cooccupancy),
+         cooccupancy = if_else(cooccupancy < 0.25, 0.25, cooccupancy)) %>%
+  ggplot(aes(x, y, fill = cooccupancy)) +
   coord_sf(crs = 'EPSG:3005') +
   facet_grid(. ~ temperature_C) +
   geom_raster() +
   scale_x_continuous(NULL, expand = c(0, 0), breaks = -c(124, 122, 120)) +
   scale_y_continuous(NULL, expand = c(0, 0)) +
-  scale_fill_distiller('Relative encounter rate   ', type = 'div',
+  scale_fill_distiller('Relative co-occupancy rate   ', type = 'div',
                        palette = 2, direction = 1, trans = 'log2',
                        limits = c(0.25, 4),
                        breaks = c(0.25, 0.5, 1, 2, 4),

### 2025-10-24 writing/manuscript.Rmd
addressed remaining comments from KH

@@
 Elk data from the work of @ciuti_human_2012 were downloaded from Movebank [@kays_movebank_2022].
-Finally, grizzly bear telemetry data are from @ref, while mountain goat data are from @ref.
+Finally, grizzly bear telemetry data are from @grizzly_ref, while mountain goat data are from @goats_ref.
 
@@
 We calculated total forest cover by summing the four rasters of evergreen/deciduous needle-leaf trees, evergreen broadleaf trees, deciduous broadleaf trees, and mixed/other trees (raster classes 1-4, respectively).
-We converted the raster of percent cover of open water (class 12) to a binary raster of pixels with water (cover > 0%) or without water (cover = 0%) and then calculated each pixel's distance from the nearest pixel with water using the `distance()` function from the `terra` package.
+We converted the raster of percent cover of open water (class 12) to a binary raster of pixels with water (cover $\ge 1\% \approx 40,000 \text{m}^2$ for a $\approx 4 \text{km}^2$ pixel) or without water (cover $\le 1\%$) and then calculated each pixel's distance from the nearest pixel with water using the `distance()` function from the `terra` package.
+We recognize that although this approach may fail to account for small or ephemeral bodies of water, it should still capture large-scale preferences for proximity to water.
 Finally, we obtained two digital elevation models using the `get_elev_raster()` function from the `elevatr` package [v.
@@
 The effects of temperature on mammals' movement rates depend on circadian rhythm and phenology, since animals may respond to temperatures differently at different times of day and or days of year.
-Additionally, both temperature and phenology depend on time of day and day of year.
+Additionally, temperature varies with time of day and day of year.
 Finally, circadian rhythm changes with seasonal phenology due to changes in photoperiod (e.g., the time of twilight changes throughout the year, affecting crepuscular activity).
@@
 
-\tiny
+\scriptsize
 
@@
                 'T>+20\\textdegree{}C (%)' = get_perc(temp_c > +20))) %>%
-  knitr::kable(format = 'markdown', digits = 2, format.args = list(big.mark = ','), caption = "Summary statistics for each species' GPS fixes with finite speed estimates, namely: the number fixes after data cleaning ($n$), the percentage of fixes with temperature ($T$) below $-20^\\circ$C and above $20^\\circ$C, the estimated mean probability of moving ($\\hat P(M=1)$), the mean speed when moving ($\\hat{\\mathbb E}(S|M=1)$; km/h), and the mean hourly distance travelled ($\\hat P(M=1) \\times \\hat{\\mathbb E}(S|M=1) = \\hat{\\mathbb E}(D)$; km/h), for a sampling interval of 1 hour and a temperature of $T = 0^\\circ$C.")
+  knitr::kable(format = 'markdown', digits = 2, format.args = list(big.mark = ','), caption = "Summary statistics for each species' GPS fixes with finite speed estimates, namely: the number fixes after data cleaning ($n$), the percentage of fixes with temperature ($T$) below $-20^\\circ$C and above $20^\\circ$C, the estimated mean probability of moving ($\\hat P(M=1)$), the mean speed when moving ($\\hat{\\mathbb E}(S|M=1)$; km/h), and the mean hourly distance travelled ($\\hat P(M=1) \\times \\hat{\\mathbb E}(S|M=1) = \\hat{\\mathbb E}(D)$; km/h), for a sampling interval of 1 hour and a temperature of $T = 0^\\circ$C.", align = 'c')
 ```
@@
 As temperatures warmed, elk and boreal caribou shifted towards more forest cover closer to 50%, while southern mountain caribou and grizzly bears selected for areas with 50% forest cover or less.
-All species shifted elevationally with temperature, although boreal caribou ha a much sharper change in RSS across elevation.
-Elk, mountain goats, and cougars moved to higher elevations with temperature, while wolves, southern mountain caribou, and grizzly bears moved to lower elevations.
+All species shifted elevationally with temperature, but species varied in the magnitude, direction, and complexity of their responses.
+As temperatures warmed, elk, mountain goats, and cougars moved to higher elevations, while wolves, southern mountain caribou, and grizzly bears moved to lower elevations.
+Cougars' selection for higher elevation strengthened substantially at temperatures $\gtrapprox 20^\circ$C, while mountain goats and wolves showed strong switches in preferences near $10^\circ$C.
 Wolves, elk, and southern mountain caribou moved closer to water with temperature, while mountain goats, cougars, and grizzly bears moved somewhat further away from water but remained mainly within 5-10 km of water.
@@
 \@ref(fig:boreal-rss)A-B), but caribou generally avoided the river system that wolves were often found at, and their selection strength varied with temperature.
-As wolves' habitat selection strengthened at warmer temperatures, encounters at 20$^\circ$C were approximately four times more frequent than at -20$^\circ$C, independent of changes in the species' movement rates (Fig.
+As wolves' habitat selection strengthened at warmer temperatures, co-occupancy at 20$^\circ$C wes approximately four times more higher than at -20$^\circ$C, independent of changes in the species' movement rates (Fig.
 \@ref(fig:boreal-rss)C).
@@
 RSS values were re-centered and capped to between 0.25 and 4 ($2^{\\pm 2}$) to improve readability.
-Finally, panel (C) shows the (scaled) product of the RSS values in panel (B) as a proxy for encounter rates between individuals of the two species."}
+Finally, panel (C) shows the (scaled) product of the RSS values in panel (B) as a proxy for the co-occupancy of the two species."}
 knitr::include_graphics('../figures/boreal-caribou-wolf-habitat.png')
@@
 Additionally, the hierarchical approach allowed us to leverage data from multiple individuals in a single model while accounting for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mccabe_resource_2021; @jeltsch_need_2025; @silva_too_2025].
-Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously, so we suggest choosing each smooth term's basis size (`k`) carefully.
-While penalized splines and REML can help avoid unruly smooth terms, the penalty terms can struggle to restrict model complexity when datasets are as large as those presented here [@wood_generalized_2017; @wood_generalized_2017-1].
+Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously, so we suggest choosing each smooth term's basis size (`k`) carefully and confirming that estimated responses align with the spieces' ecology.
+While penalized splines and REML can help avoid unruly smooth terms, the penalty terms can struggle to restrict model complexity when datasets are as large as those presented here [@wood_generalized_2017; @wood_generalized_2017-1], so estimated responses should be aligned with biological relevance rather than statistical significance.
 
@@
 The increase in movement rates with temperature during the cold months is likely partly due to shallower snow depth, which results in greater mobility [@pedersen_quantifying_2021], but warmer temperatures during snowy seasons also increase the risk of rain-on-snow events, which limit forage availability and increase time spent foraging [@stien_congruent_2012; @berger_climate_2018; @mezzini_how_2025].
-The reduced movement rates during warmer summers are likely mainly due to more frequent resting and use of thermal refugia, given ungulates' elevated risk of hyperthermia [@alston_temperature_2020; @verzuh_behavioural_2023].
+The reduced movement rates during warmer summers are likely mainly due to more frequent resting and use of thermal refugia, given ungulates' documented risk of hyperthermia [@alston_temperature_2020; @verzuh_behavioural_2023].
 Therefore, although we predict that boreal caribou's yearly average movement rates will remain approximately unchanged throughout the 21^st^ century, the projections do not show the changes in seasonal movement phenology.
@@
 However, spatial heterogeneity in habitat, climate, and cougar behaviour may cause boreal individuals to respond to temperature differently from those in our study.
-Still, rising temperatures clearly affect more than each individual species' movement behaviour; they have complex and interconnected effects on entire communities and their trophic structures.
-
+Still, rising temperatures clearly affect more than each individual species' movement behaviour; they have complex effects on entire communities and their trophic structures.
 @cunningham_opportunity_2021 discuss the importance of adopting a community-level perspective when assessing the effects of climate change on animals' movement behaviour and phenology.

### 2025-10-30 analysis/figures/temperature-hrsfs.R
fix ordering of `SPECIES` and error w data check

@@ -14,8 +14,8 @@ if(file.exists('models/temperature-hrsf-preds.rds')) {
 } else {
   # import models for each species
   d <- tibble(
-    species = SPECIES,
-    lab = SPECIES_LABS,
+    species = sort(SPECIES),
+    lab = sort(SPECIES_LABS),
     file_name = map_chr(species, \(.sp) {
       fn <- list.files(path = 'H:/GitHub/bc-mammals-temperature/models/',
                        pattern = gsub('\\(s\\.', 'southern', .sp) %>%
@@ -71,7 +71,7 @@ if(file.exists('models/temperature-hrsf-preds.rds')) {
         bind_cols(.,
                   predict(object = m, newdata = newd, type = 'link',
                           se.fit = TRUE, discrete = TRUE,
-                          newdata.guaranteed = TRUE,
+                          # newdata.guaranteed = TRUE,
                           terms = c('s(forest_perc)',
                                     'ti(forest_perc,temperature_C)')) %>%
                     as.data.frame()),
@@ -91,7 +91,7 @@ if(file.exists('models/temperature-hrsf-preds.rds')) {
         bind_cols(.,
                   predict(object = m, newdata = newd, type = 'link',
                           se.fit = TRUE, discrete = TRUE,
-                          newdata.guaranteed = TRUE,
+                          # newdata.guaranteed = TRUE,
                           terms = c('s(elevation_m)',
                                     'ti(elevation_m,temperature_C)')) %>%
                     as.data.frame()),
@@ -111,7 +111,7 @@ if(file.exists('models/temperature-hrsf-preds.rds')) {
         bind_cols(.,
                   predict(object = m, newdata = newd, type = 'link',
                           se.fit = TRUE, discrete = TRUE,
-                          newdata.guaranteed = TRUE,
+                          # newdata.guaranteed = TRUE,
                           terms = c('s(dist_water_m)',
                                     'ti(dist_water_m,temperature_C)')) %>%
                     as.data.frame())) %>%

### 2025-10-30 analysis/figures/temperature-hrsfs.R
add code for diagnostic plots of raw data

@@ -235,3 +235,33 @@ p_se <-
 
 ggsave('figures/hrsf-surface-plots-se.png', p_se, width = 16, height = 7.2,
        units = 'in', dpi = 600, bg = 'white')
+
+# for comparing the models to the raw data
+if(FALSE) {
+  #' change index in `[[ ]]` to cycle across species
+  obs <- filter(d$rsf[[1]]$model, detected == 1)
+  
+  ggplot(obs, aes(temperature_C, forest_perc)) +
+    facet_wrap(~ animal) +
+    geom_hex() +
+    labs(x = expression(paste(bold('Temperature ('), bold(degree), bold('C)'))),
+         y = 'Forest cover (%)') +
+    scale_fill_lapaz(name = 'Count', trans = 'sqrt', reverse = TRUE) +
+    theme(legend.position = 'inside', legend.position.inside = c(0.9, 0.15))
+  
+  ggplot(obs, aes(temperature_C, elevation_m / 1e3)) +
+    facet_wrap(~ animal) +
+    geom_hex() +
+    labs(x = expression(paste(bold('Temperature ('), bold(degree), bold('C)'))),
+         y = 'Elevation (km)') +
+    scale_fill_lapaz(name = 'Count', trans = 'sqrt', reverse = TRUE) +
+    theme(legend.position = 'inside', legend.position.inside = c(0.9, 0.15))
+  
+  ggplot(obs, aes(temperature_C, dist_water_m / 1e3)) +
+    facet_wrap(~ animal) +
+    geom_hex() +
+    labs(x = expression(paste(bold('Temperature ('), bold(degree), bold('C)'))),
+         y = 'Distance from water (km)') +
+    scale_fill_lapaz(name = 'Count', trans = 'sqrt', reverse = TRUE) +
+    theme(legend.position = 'inside', legend.position.inside = c(0.9, 0.15))
+}

### 2025-10-30 writing/manuscript.Rmd
addressed comments from KS

@@
 Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @merkle_large_2016; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020].
-How and where animals move thus also determines encounter rates with resources [@mezzini_how_2025], humans [@johnson_human_2018; @weststrate_how_2024], predators, prey, and competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024].
+How and where animals move thus also determines encounter rates with resources [@mezzini_how_2025], humans [@johnson_human_2018; @weststrate_how_2024], and predators, prey, and competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024], including infected and parasitized individuals [@blanchong_respiratory_2018].
 However, ambient temperature affects animals' movement rates by altering the energetic cost of movement and thermoregulation [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], and extreme temperatures can impact animals' movement rates [@dyer_travel_2023], phenology [@carbeck_adaptation_2022; @johnson_human_2018; @mclellan_effect_2015; @pigeon_drivers_2016], and life history [@newediuk_climate_2024], and even cause physiological damage and death [@hetem_responses_2014; @powers_hovering_2017; @ratnayake_forecasting_2019; @schmidt_interplay_2020; @schwerdt_ecological_2024].
@@
 
-In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves; Table \@ref(tab:data-table)).
+In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats, and wolves; Table \@ref(tab:data-table)).
 Western Canada is currently experiencing accelerating and widespread but spatially heterogeneous warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [plants: @post_climate_2008; @tysor_countergradient_2025; ice: @basu_phenological_2024; caribou: @le_corre_weather_2016], and more frequent and intense extreme heat events [@zhang_increased_2023].
@@
 Elk data from the work of @ciuti_human_2012 were downloaded from Movebank [@kays_movebank_2022].
-Finally, grizzly bear telemetry data are from @grizzly_ref, while mountain goat data are from @goats_ref.
+Finally, grizzly bear telemetry data are from @grizzly_ref, while mountain goat data are from the work of @balyx_human_2022 and were provided by the British Columbia Ministry of Environment and Parks.
 
@@
 Across all species, Relative Selection Strength (RSS) was weakest for forest cover and strongest for elevation.
-At temperatures near 0$^\circ$C, boreal caribou selected for forest cover between 50% and 75%, elevations near 500 m, and distances from water < 10 km, while southern mountain caribou selected for dense forest cover, elevations near 2 km, and distances from water $\lessapprox$ 5 km.
+At temperatures near 0$^\circ$C, boreal caribou selected for forest cover between 50% and 75%, elevations near 500 m above sea level, and distances from water < 10 km, while southern mountain caribou selected for dense forest cover, elevations near 2 km, and distances from water $\lessapprox$ 5 km.
 Cougars selected for dense forest cover (> 75%), an elevation of ~ 1 km, and distances from water < 7.5 km.
@@
 Grizzly bears selected for relatively sparse forest cover (25-50%), elevation between 1 and 2 km, and distances from water < 3 km.
-Mountain goats selected for sparse forest cover (< 25%), elevations near 1.5 km, and distances from water < 5 km.
+Mountain goats selected for sparse forest cover (< 25%), elevations near 1.8 km, and distances from water < 5 km.
 Finally, wolves selected for forest cover ($\gtrapprox$ 50%), elevations near 1 km, and distances from water < 5 km.There was relatively strong agreement between models with and without temperature (Figs.
@@
 \@ref(fig:boreal-rss)A-B), but caribou generally avoided the river system that wolves were often found at, and their selection strength varied with temperature.
-As wolves' habitat selection strengthened at warmer temperatures, co-occupancy at 20$^\circ$C wes approximately four times more higher than at -20$^\circ$C, independent of changes in the species' movement rates (Fig.
+As wolves' habitat selection strengthened at warmer temperatures, co-occupancy at 20$^\circ$C was approximately four times more higher than at -20$^\circ$C, independent of changes in the species' movement rates (Fig.
 \@ref(fig:boreal-rss)C).
@@
 Additionally, the hierarchical approach allowed us to leverage data from multiple individuals in a single model while accounting for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mccabe_resource_2021; @jeltsch_need_2025; @silva_too_2025].
-Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously, so we suggest choosing each smooth term's basis size (`k`) carefully and confirming that estimated responses align with the spieces' ecology.
+Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously, so we suggest choosing each smooth term's basis size (`k`) carefully and confirming that estimated responses align with the species' ecology.
 While penalized splines and REML can help avoid unruly smooth terms, the penalty terms can struggle to restrict model complexity when datasets are as large as those presented here [@wood_generalized_2017; @wood_generalized_2017-1], so estimated responses should be aligned with biological relevance rather than statistical significance.
@@
 For boreal caribou, hotter temperatures decreased movement rates summer but increased them otherwise, mirroring previous studies [@stien_congruent_2012; @leclerc_determinants_2021; @lessard_local_2025].
-The increase in movement rates with temperature during the cold months is likely partly due to shallower snow depth, which results in greater mobility [@pedersen_quantifying_2021], but warmer temperatures during snowy seasons also increase the risk of rain-on-snow events, which limit forage availability and increase time spent foraging [@stien_congruent_2012; @berger_climate_2018; @mezzini_how_2025].
+The increase in movement rates with temperature during the cold months is likely partly due to shallower snow depth, which results in greater mobility [@pedersen_quantifying_2021], but warmer temperatures during snowy seasons also increase the risk of temperature fluctuations near $0^\circ$C and the risk of rain-on-snow events [@musselman_projected_2018], which limit forage availability and increase time spent foraging [@stien_congruent_2012; @berger_climate_2018; @mezzini_how_2025].
 The reduced movement rates during warmer summers are likely mainly due to more frequent resting and use of thermal refugia, given ungulates' documented risk of hyperthermia [@alston_temperature_2020; @verzuh_behavioural_2023].
@@
 The estimated changes in movement behaviour cannot be fully attributed to physiological responses to temperature alone, since other aspects of habitats' seasonal phenology are (nonlinearly) correlated with temperature.
-For example, mountain goats' lower movement rates at temperatures near 0$^\circ$C in winter may be in part due to increased movement difficulty, since melting snow and rain-on-snow events increase the energetic costs of movement [@white_population_2025] and encounter rates with predators [@sullender_defining_2023].
+For example, mountain goats' lower movement rates at temperatures near 0$^\circ$C in winter may be due to a multitude of factors, including: increased movement difficulty, since melting snow and rain-on-snow events increase the energetic costs of movement [@white_population_2025]; reduced movement during more frequent storms (Safford, pers.
+comm.) and encounter rates with predators [@sullender_defining_2023].
 However, milder temperatures can also increase plant growth, which allows goats to spend more time foraging within and less time travelling across patches [@charnov_optimal_1976; @white_mountain_2025].
@@
 Climate change has also had extensive impacts on plant growth and phenology [@cleland_shifting_2007; @denny_performance_2019; @tysor_countergradient_2025] with consequences on herbivores' behaviour and fitness [@post_climate_2008; @aikens_greenscape_2017].
-Additionally, plants' limited ability to disperse to and establish in new habitats is often a limiting factor for herbivores' ability to escape hotter temperatures by moving to higher altitudes or latitudes, and the advance of herbivores can further limit plants' ability to establish [@speed_elevational_2012; @diffenbaugh_changes_2013].
+Plants' limited ability to disperse to and establish in new habitats is often a limiting factor for herbivores' ability to escape hotter temperatures by moving to higher altitudes or latitudes, and the advance of herbivores can further limit plants' ability to establish [@speed_elevational_2012; @diffenbaugh_changes_2013].
 Consequently, mammals that move to higher elevations may be forced to spend more time foraging [@mezzini_how_2025], which will increase both their energetic expenditure and encounter rates with predators and competitors [@kohl_prey_2019; @martinez-garcia_how_2020; @torrez-herrera_monkeys_2020; @smith_densitydependent_2023].
@@
 By accounting for the effects of multiple variables at once, they demonstrate how a species' response to one variable can depend on other concurrent changes.
-Estimating responses to multifarious change is crucial in assessing mammals' responses to hotter temperatures, altered phenology, and novel community dynamics.
+Estimating responses to multifarious change is crucial in assessing mammals' responses to hotter temperatures, altered phenology, and novel community dynamics, including more frequent and severe exposure to disease and parasites as well as their effects on mammalian physiology and energetic needs [@pickles_predicting_2013; @barber_potential_2018; @gray_ticks_2021; @blanchong_respiratory_2018; @gude_demographic_2022].
 
@@
 
-Future work should develop models that account for covariates that are specific to the study species and area, such as forest age and disturbance [@demars_incorporating_2023], terrain ruggedness [@white_mountain_2017], anthropogenic linear features [@dickie_faster_2017], and growing human presence in wild spaces [@sih_evolution_2011; @rice_conspicuous_2022], including the growing risk of human-wildlife conflict [@abrahms_climate_2023; @weststrate_how_2024].
+Future work should develop models that account for covariates that are specific to the study species and area, such as forest age and disturbance [@demars_incorporating_2023], terrain ruggedness [@white_mountain_2017], anthropogenic linear features [@dickie_faster_2017], greater physiological stress from infection and parasitism [@pickles_predicting_2013; @gude_demographic_2022], and growing human presence in wild spaces [@sih_evolution_2011; @rice_conspicuous_2022], including the growing risk of human-wildlife conflict [@abrahms_climate_2023; @weststrate_how_2024].
 In particular, forests in western Canada are highly dynamic over time [@zhang_half-century_2015], and fires size and burn severity have increased in western Canada in recent decades [@whitman_climate-induced_2022; @parisien_abrupt_2023; @wang_canadian_2025].
 Thus, while our use of a static raster of forest cover allowed for simpler models and climate change projections, future studies should account for mammals' responses to temporal forest dynamics, including the impacts of fire [@whitman_climate-induced_2022; @parisien_abrupt_2023; @wang_canadian_2025], logging [@lochhead_linking_2022] and other extractive industries [@dickie_faster_2017; @dickie_habitat_2024].
-However, the inclusion of temporally dynamic variables will require that any future projections of species' behaviour depend on temporally dynamic estimates of how such variables will change over time with changes in climate and anthropogenic activity.
+Additionally, rising incidence of disease and parasite load over the coming decades may also alter increase mammals' energetic needs [@gude_demographic_2022] and time spent foraging or hunting [@mezzini_how_2025].
+However, the inclusion of temporally dynamic variables will require that any future projections of species' behaviour depend on temporally dynamic estimates of how such variables will change over the years.
 
@@
 
+\
+
 # Author contributions {-}
@@
 MJN conceived of the project idea and supervised SM throughout the project.
-All other authors contributed telemetry data and reviewed the interpretation of the results for their species of interest.
+All other authors contributed telemetry data and/or reviewed the interpretation of the results for their species of interest.
 KH and SD contributed substantially to manuscript review.

### 2025-10-31 writing/manuscript.Rmd
clarified some points based on feedback from KS

@@
 csl: 'global-change-biology.csl'
-# csl: 'movement-ecology.csl'
+# csl: 'movement-ecology.csl' # to count refs
 fontsize: 12pt
@@
 
-\noindent \textbf{References}: 155 (updated on 2025-10-01) <!-- counted using movement ecology csl -->
+\noindent \textbf{References}: 165 (updated on 2025-10-31) <!-- counted using movement ecology csl -->
 
@@
 Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.
-Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @le_corre_weather_2016; @slatyer_ecological_2022; @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].
+Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @stien_congruent_2012; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @le_corre_weather_2016; @slatyer_ecological_2022; @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].
 
@@
 
-The HGAMs (equations \@ref(eq:p-moving-model) and \@ref(eq:speed-model)) included fixed-effect intercepts for each species ($\beta_\texttt s$), random intercepts for each animal ($Z_\texttt{a}$), and species-level `by` smooths that allowed independent smoothness parameters for each species [model I in Figure 4 of @pedersen_hierarchical_2019].
+The HGAMs (equations \@ref(eq:p-moving-model) and \@ref(eq:speed-model)) included fixed-effect intercepts for each species ($\beta_\texttt s$), random intercepts for each animal ($Z_\texttt{a}$), and species-level `by` smooths that allowed independent smoothness parameters for each species [model I in Fig.
+4 of @pedersen_hierarchical_2019].
 The `by` smooths accounted for trends in time of day (in Pacific Daylight Time; `tod_pdt`), day of year (`doy`), and temperature (`temp_c`).
@@
 Finally, two smooth terms of log-transformed sampling interval (`dt`; hours) corrected for biases in speed estimates arising from irregular GPS sampling intervals, since longer intervals result in lower speed estimates [@nathan_big-data_2022; @denicola_are_2025].
-A global smooth term of $\log(\texttt{dt})$ accounted for the overall effect of sampling interval, while a factor-smooth interaction term (`bs = 'fs'`) of $\log(\texttt{dt})$ and species accounted for species-level deviations from the global term while assuming a common smoothness parameter across species [model GS in Figure 4 of @pedersen_hierarchical_2019].
+A global smooth term of $\log(\texttt{dt})$ accounted for the overall effect of sampling interval, while a factor-smooth interaction term (`bs = 'fs'`) of $\log(\texttt{dt})$ and species accounted for species-level deviations from the global term while assuming a common smoothness parameter across species [model GS in Fig.
+4 of @pedersen_hierarchical_2019].
 Formally, the model for movement state $M$, with $M = 0$ indicating no movement and $M = 1$ indicating movement, was
@@
 The estimated changes in movement behaviour cannot be fully attributed to physiological responses to temperature alone, since other aspects of habitats' seasonal phenology are (nonlinearly) correlated with temperature.
-For example, mountain goats' lower movement rates at temperatures near 0$^\circ$C in winter may be due to a multitude of factors, including: increased movement difficulty, since melting snow and rain-on-snow events increase the energetic costs of movement [@white_population_2025]; reduced movement during more frequent storms (Safford, pers.
-comm.) and encounter rates with predators [@sullender_defining_2023].
+For example, mountain goats' lower movement rates at temperatures near 0$^\circ$C in winter may be due to a multitude of factors, including: increased movement difficulty, since melting snow and rain-on-snow events increase the energetic costs of movement [@white_population_2025]; reduced movement during storms [@musselman_projected_2018]; and increased encounter risk with predators [@sullender_defining_2023].
 However, milder temperatures can also increase plant growth, which allows goats to spend more time foraging within and less time travelling across patches [@charnov_optimal_1976; @white_mountain_2025].
-Therefore, figure \@ref(fig:dist) illustrates an estimate of the total effects of temperature on each species' movement rates, but it does not explicitly account for energetic costs.
+Therefore, Fig.
+\@ref(fig:dist) illustrates an estimate of the total effects of temperature on each species' movement rates, but it does not explicitly account for energetic costs.
 Similarly, species' changes in habitat selection do not explicitly account for any changes in physiological or energetic costs.
@@
 Plants' limited ability to disperse to and establish in new habitats is often a limiting factor for herbivores' ability to escape hotter temperatures by moving to higher altitudes or latitudes, and the advance of herbivores can further limit plants' ability to establish [@speed_elevational_2012; @diffenbaugh_changes_2013].
-Consequently, mammals that move to higher elevations may be forced to spend more time foraging [@mezzini_how_2025], which will increase both their energetic expenditure and encounter rates with predators and competitors [@kohl_prey_2019; @martinez-garcia_how_2020; @torrez-herrera_monkeys_2020; @smith_densitydependent_2023].
-If local peaks become unsuitable habitat, and moving to nearby higher peaks requires traveling across low-quality and high-risk habitat (Figure \@ref(fig:s-rss-cc)), habitat fragmentation may prevent movement across ranges [@white_mountain_2025].
+Consequently, mammals that move to higher elevations may be forced to spend more time foraging if forage is scarce or unpredictable [@mezzini_how_2025], which will increase both their energetic expenditure and encounter rates with predators and competitors [@kohl_prey_2019; @martinez-garcia_how_2020; @torrez-herrera_monkeys_2020; @smith_densitydependent_2023].
+If local peaks become unsuitable habitat, and moving to nearby higher peaks requires traveling across low-quality and high-risk habitat (Fig.
+\@ref(fig:s-rss-cc)), habitat fragmentation may prevent movement across ranges [@white_mountain_2025].
 
@@
 By accounting for the effects of multiple variables at once, they demonstrate how a species' response to one variable can depend on other concurrent changes.
-Estimating responses to multifarious change is crucial in assessing mammals' responses to hotter temperatures, altered phenology, and novel community dynamics, including more frequent and severe exposure to disease and parasites as well as their effects on mammalian physiology and energetic needs [@pickles_predicting_2013; @barber_potential_2018; @gray_ticks_2021; @blanchong_respiratory_2018; @gude_demographic_2022].
+Estimating responses to multifarious change is crucial in assessing mammals' responses to hotter temperatures, altered phenology, and novel community dynamics.
+For instance, warming temperatures have increased exposure to disease and parasites, with impacts on mammalian physiology and energetic needs [@pickles_predicting_2013; @barber_potential_2018; @gray_ticks_2021; @blanchong_respiratory_2018; @gude_demographic_2022].
+In particular, severe tick loads can reduce mammals' ability to thermoregulate in the cold due to reduced body fat and fur [@pekins_metabolic_2020] and can even reduce movement rates by increasing time spent recumbent [@addison_influence_2023].
 

### 2025-11-03 writing/manuscript.Rmd
add ref for RS's cougar data; add missing `\beta_0`

@@
 Elk data from the work of @ciuti_human_2012 were downloaded from Movebank [@kays_movebank_2022].
-Finally, grizzly bear telemetry data are from @grizzly_ref, while mountain goat data are from the work of @balyx_human_2022 and were provided by the British Columbia Ministry of Environment and Parks.
+Finally, grizzly bear telemetry data are from @serrouya_grizzly_2011, while mountain goat data are from the work of @balyx_human_2022 and were provided by the British Columbia Ministry of Environment and Parks.
 
@@
 \begin{aligned}
-\text{log}(\lambda) = & f_{1}(\texttt{forest\_perc}) + f_{2}(\texttt{elevation\_m}) + f_{3}(\texttt{dist\_water\_m}) + \\
+\text{log}(\lambda) = & \beta_0 + f_{1}(\texttt{forest\_perc}) + f_{2}(\texttt{elevation\_m}) + f_{3}(\texttt{dist\_water\_m}) + \\
 & Z_\texttt a + f_{4, \texttt{a}}(\texttt{forest\_perc}) + f_{5, \texttt{a}}(\texttt{elevation\_m}) + f_{6, \texttt{a}}(\texttt{dist\_water\_m}) + \\

### 2025-11-07 analysis/figures/temperature-hrsfs.R
centered RRS values to medians at 0 degrees C

@@ -176,7 +176,7 @@ p <-
   # doing this because some estimated effects of distance from water and
   # elevation are extreme
   group_by(lab, variable) %>%
-  mutate(lambda = lambda / median(lambda),
+  mutate(lambda = lambda / median(lambda[round(temperature_C) == 0]),
          lambda = case_when(
            lab == 'Elk' & variable == 'bold(Elevation~"(km)")' ~ lambda / exp(1),
            lab == 'Mountain goats' & variable == 'bold(Elevation~"(km)")' ~ lambda * exp(0.25),

### 2025-11-07 writing/manuscript.Rmd
update based on comments from ATF

@@
 How and where animals move thus also determines encounter rates with resources [@mezzini_how_2025], humans [@johnson_human_2018; @weststrate_how_2024], and predators, prey, and competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024], including infected and parasitized individuals [@blanchong_respiratory_2018].
-However, ambient temperature affects animals' movement rates by altering the energetic cost of movement and thermoregulation [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], and extreme temperatures can impact animals' movement rates [@dyer_travel_2023], phenology [@carbeck_adaptation_2022; @johnson_human_2018; @mclellan_effect_2015; @pigeon_drivers_2016], and life history [@newediuk_climate_2024], and even cause physiological damage and death [@hetem_responses_2014; @powers_hovering_2017; @ratnayake_forecasting_2019; @schmidt_interplay_2020; @schwerdt_ecological_2024].
+However, ambient temperature affects animals' movement rates by altering the energetic cost of movement and thermoregulation [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], and extreme temperatures can impact animals' movement rates [@dyer_travel_2023], phenology [@carbeck_adaptation_2022; @johnson_human_2018; @mclellan_effect_2015; @pigeon_drivers_2016], and life history [@newediuk_climate_2024; @woodroffe_hot_2017], and even cause physiological damage and death [@hetem_responses_2014; @powers_hovering_2017; @ratnayake_forecasting_2019; @schmidt_interplay_2020; @schwerdt_ecological_2024; @rabaiotti_high_2021].
 Animals may search for a short-term buffer from extreme temperatures by moving to thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], and shelter may even be preferred over forage abundance [@hall_microrefuges_2016].
@@
 Larger-bodied mammals are more likely to alter their movement behaviour with rising temperatures, since they tend to move long distances [@merkle_large_2016], are particularly sensitive to overheating [@fuller_towards_2016; @alston_temperature_2020; @verzuh_behavioural_2023], and their movement rates are thus strongly limited by excessive heat [@dyer_travel_2023].
+Mammals can mitigate the effects of temperature by altering their movement behavior, including the timing of migration [@leclerc_determinants_2021; @xu_plasticity_2021] and their hunting strategies [@creel_hunting_2016; @rabaiotti_coping_2019], but a species' movement behavior is often constrained by its body size and vagility, including its main method of locomotion [@webber_heterothermy_2022].
 
-Global warming during the last century has resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extreme-heat events [@ipcc_climate_2023], with widespread changes in mammals' phenologies, circadian rhythms, fitness, and life histories [@botero_evolutionary_2015; @le_corre_weather_2016; @wells_life_2022; @newediuk_climate_2024; @mclellan_effect_2015; @pigeon_drivers_2016].
+Global warming during the last century has resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extreme-heat events [@ipcc_climate_2023], with widespread changes in mammals' phenologies, circadian rhythms, fitness, and life histories [@botero_evolutionary_2015; @le_corre_weather_2016; @wells_life_2022; @newediuk_climate_2024; @mclellan_effect_2015; @pigeon_drivers_2016; @woodroffe_hot_2017; @guiden_seasonal_2020].
 However, behavioural responses to heat vary substantially across species depending on body size, vagility, and activity times [@mccain_body_2014; @williams_range_2018].
@@
 Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term.
-Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @stien_congruent_2012; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @le_corre_weather_2016; @slatyer_ecological_2022; @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].
+Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @stien_congruent_2012; @carbeck_adaptation_2022; @cunningham_opportunity_2021; @pokallus_population_2015] and phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @le_corre_weather_2016; @slatyer_ecological_2022; @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].
 
@@
 Therefore, estimating the effects of temperature on movement rates requires accounting for how mammals' response to temperature changes with time of day and day of year [Fig.
-\@ref(fig:dags)A$;$ @peron_periodic_2017; @buderman_time-varying_2018; @leclerc_determinants_2021; @xu_plasticity_2021].
+\@ref(fig:dags)A$;$ @peron_periodic_2017; @buderman_time-varying_2018; @leclerc_determinants_2021; @xu_plasticity_2021; @guiden_seasonal_2020; @webb_measuring_2010].
 Similarly, mammals' selection strength for resources depends on ambient temperature, since higher temperatures can promote a selection for refuge from heat [e.g., thicker forest cover, higher elevation, proximity to water$;$ @attias_effects_2018; @giroux_activity_2023].
@@
 
-\noindent Of the GPS fixes with finite speed estimates, 2.6% had temperatures lower than $-20^\circ$C, while 6.5% had temperatures above $20^\circ$C, and temperature ranges differed across species (Table \@ref(tab:hgam-summary), Fig.
+\noindent Across species, approximately 10% of GPS fixes had absolute temperatures above $20^\circ$C.
+Of the fixes with finite speed estimates, 2.6% had temperatures lower than $-20^\circ$C, while 6.5% had temperatures above $20^\circ$C, but temperature ranges differed across species (Table \@ref(tab:hgam-summary), Fig.
 B2).
@@
 In late spring (June 1^st^), they moved from evening to early morning if hourly temperatures were below $20^\circ$C, but if temperatures were above 20$^\circ$C they moved mostly between 3:00 and 6:00 AM.
-Throughout the year, they tended to move more when it was colder, but what they perceived as "colder" was relative to the time of year.
+Throughout the year, they tended to move more when it was colder, relative to the range of temperatures at that time of year.
 Overall, uncertainty around the estimated effects was generally higher at extreme temperatures due to lower data availability (Figs.
@@
 ```{r dist, fig.cap="\\textbf{Temperature is a strong driver of how far and when mammals travel.} The fill indicates the effects of temperature on the relative change in hourly distance travelled (probability of moving times mean speed when moving) over time of day on June $1^{\\text{st}}$ (\\textbf{A}) and day of year at 12:00 PDT (\\textbf{B}).
-Predictions in the surface plots extend to 10% of the range away from each datum.
-The color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to $2^{\\pm 2}$ for ease of readability."}
+Predictions in the surface plots extend to 10% of the range away from each datum to avoid predicting for parameter spaces with no data (grey fill).
+The color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to $2^{\\pm 2}$ for ease of readability.
+Horizontal contour lines indicate that movement rates depend most on time of day and day of year (e.g., grizzly bears), while vertical lines indicate that temperature has the gretest effect, irrespective of time (e.g., mountain goats)."}
 # not specifying dt and doy for A and time for B because change is relative
@@
 The grey contours indicate the extent of each species' observed data.
-RSS values were re-centered and capped to between 0.25 and 4 ($2^{\\pm 2}$) to improve readability."}
+To mitigate the sampling bias across temperature (e.g., grizzly bears have few to no while hibernating in winter), RSS values at each temperature were divided by the average RSS value at that temperature within the respective panel.
+To further improve readability, each pannel's RSS values were centered by dividing by the median RSS value at $0^\\circ$C and capped to between 0.25 and 4 ($2^{\\pm 2}$)."}
 knitr::include_graphics('../figures/hrsf-surface-plots.png')
@@
 As wolves' habitat selection strengthened at warmer temperatures, co-occupancy at 20$^\circ$C was approximately four times more higher than at -20$^\circ$C, independent of changes in the species' movement rates (Fig.
-\@ref(fig:boreal-rss)C).
+\@ref(fig:boreal-rss)C-D).
 
@@
 
-```{r boreal-rss, fig.cap="\\textbf{Temperature affects boreal caribou's and wolves' habitat selection and, consequently, their encounter rates.} Panel (A) shows the rasters of the three habitat variables used in the hierarchical resource selection functions, while panel (B) shows the relative selection strength (RSS) for boreal caribou and wolves, as a function of temperature.
+```{r boreal-rss, fig.cap="\\textbf{Temperature affects boreal caribou's and wolves' habitat selection and, consequently, their encounter rates.} Panel (A) shows the relative selection strength (RSS) for boreal caribou and wolves, as a function of temperature.
 RSS values were re-centered and capped to between 0.25 and 4 ($2^{\\pm 2}$) to improve readability.
+Panel (B) shows the rasters of the three habitat variables used in the hierarchical resource selection functions used to calculate the values in (A).
 Finally, panel (C) shows the (scaled) product of the RSS values in panel (B) as a proxy for the co-occupancy of the two species."}
@@
 Consequently, mammals that move to higher elevations may be forced to spend more time foraging if forage is scarce or unpredictable [@mezzini_how_2025], which will increase both their energetic expenditure and encounter rates with predators and competitors [@kohl_prey_2019; @martinez-garcia_how_2020; @torrez-herrera_monkeys_2020; @smith_densitydependent_2023].
-If local peaks become unsuitable habitat, and moving to nearby higher peaks requires traveling across low-quality and high-risk habitat (Fig.
-\@ref(fig:s-rss-cc)), habitat fragmentation may prevent movement across ranges [@white_mountain_2025].
+If local high-elevation habitats become unsuitable, and moving to nearby higher peaks requires traveling across low-quality and high-risk habitat (Fig.
+\@ref(fig:s-rss-cc)), habitat fragmentation may prevent movement between ranges [@white_mountain_2025].
+Consequently, proactive corridor planning will be crucial in allowing species to mitigate the effects of climate change, and corridors should include a wide range of habitat [@sawyer_identifying_2009; @stralberg_toward_2020; @morley_spatiotemporal_2021].
 
@@
 Exposure to warming temperatures is widespread throughout western Canada [@bunnell_global_2011; @kienzle_has_2018; @eum_potential_2023], and changes in temperature and phenology may be too rapid for many species to respond to [@diffenbaugh_changes_2013].
-Species' adaptability is further challenged by the difficulty in responding to concurrent, multifarious, and interconnected changes [@diffenbaugh_changes_2013; @polazzo_measuring_2024].
+Species' adaptability is further challenged by the difficulty in responding to concurrent, multifarious, and interconnected changes [@polazzo_measuring_2024], including changes in: climate, land and sea use, direct exploitation of organisms, pollution, and invasion of alien species [@ipbes_summary_2019; @diffenbaugh_changes_2013].
 In the future, landscapes may thus become unable to support species and population sizes they supported historically, and species may begin selecting against historical ranges [@williams_range_2018].

### 2025-11-11 writing/manuscript.Rmd
rm KS as author; rm extra from main text
Kirk asked to be removed because he didn't believe his contributions were sufficient for authorship

@@
     institute: [forest]
-  - name: Kirk Safford
-    institute: bcp
   - name: Robert Serrouya
@@
   - forest: British Columbia Ministry of Forests, Lands, Natural Resource Operations and Rural Development, Penticton, BC, Canada.
-  - bcp: British Columbia Ministry of Environment and Parks, Penticton, BC, Canada.
   - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
@@
 
-\noindent \textbf{Words in main text}: `r suppressMessages(wordcountaddin::word_count())`
+\noindent \textbf{Words in main text}: `r suppressMessages(wordcountaddin::word_count()) - 80` <!-- does not include abstract or figure captions; "- 80" removes the stuff on page 2 -->
 
@@
 
-\noindent \textbf{References}: 165 (updated on 2025-10-31) <!-- counted using movement ecology csl -->
+\noindent \textbf{References}: 178 (updated on 2025-11-11, including +1 for @serrouya_ref) <!-- counted using movement ecology csl -->
 
@@
 
-\
+# Acknowledgements {-}
+
+```{r}
+ackn <- "\\noindent We would like to thank Kirk Safford from the British Columbia Ministry of Environment and Parks (BC Parks) in Penticton, BC, Canada for providing the mountain goat telemetry data.
+SM would also like to thank BC Parks and Biodiversity Pathways for financial suppport, and also all those who provided feedback following presentations and poster sessions."
+```
+
+`r ackn`
 
@@
 
-\noindent SM performed the data cleaning, ran the analyses, and wrote the manuscript.
+```{r}
+contrib <- "\\noindent SM performed the data cleaning, ran the analyses, and wrote the manuscript.
 CHF wrote code for estimating instantaneous movement speeds.
@@
 KH and SD contributed substantially to manuscript review.
-All authors contributed to reviewing the manuscript.
+All authors contributed to reviewing the manuscript."
+```
+
+`r contrib`
 

### 2025-12-10 writing/cover-letter.tex
wrote cover letter

\documentclass{article}

\usepackage[dvipsnames]{xcolor}
% Measurements are taken directly from the guide
\usepackage[top=1.7in,left=1in,bottom=1in,right=1in]{geometry}
\usepackage{graphicx}
\usepackage[colorlinks=true,
            pdfborder={0 0 0},
            ]{hyperref}
\usepackage{lipsum}
\usepackage[absolute]{textpos}
\usepackage{tikz}
\usetikzlibrary{calc}

% No paragraph indentation
\parindent0pt
\setlength{\parskip}{0.8\baselineskip}
\raggedright
\pagestyle{empty}
% Define SI official colors
\definecolor{UBC_blue}{RGB}{13,40,86}
\definecolor{SIgray}{HTML}{5e6a71}
% Ensure consistency in the footer
\urlstyle{rm}

\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyfoot{}
\fancyhead{}
\fancyfoot[C]{ \thepage}

\setcounter{page}{1}
\renewcommand{\thepage}{Noonan, Cover Letter--- \arabic{page}}

\renewcommand{\footrulewidth}{0pt}
\renewcommand{\headrulewidth}{0pt}
\fancyfoot{}
\fancyfoot[L]{%
    {\footnotesize\color{SIgray}
Department of Biology, Irving K. Barber Faculty of Science, University of British Columbia, Kelowna, B.C., Canada \\[-0.1\baselineskip]
T: 1(250) 258-1118  $\mid$  \href{mailto:michael.noonan@ubc.ca}{michael.noonan@ubc.ca} $\mid$ \url{https://biology.ok.ubc.ca/about/contact/michael-j-noonan/}
}\color{black}}

\begin{document}

% -------------------------------------------------------
% Add logo, the text under the blue line, and the line itself
\begin{textblock*}{2in}[0.3066,0.39](1.6in,1.05in)
    \includegraphics[width=2in]{Lab_Logo_Trans.png}
\end{textblock*}
\begin{textblock*}{6.375in}(1.5in,1.1in)   % 6.375=8.5 - 1.5 - 0.625
    \hfill \color{SIgray} Stefano Mezzini, PhD Biology Candidate\\
    \hfill Department of Biology, UBC-Okanagan

\end{textblock*}
\begin{tikzpicture}[remember picture,overlay]
    \draw[color=UBC_blue,line width=1pt] (current page.north west)+(1.05in,-1.6in) -- ($(-0.625in,-1.6in)+(current page.north east)$);
\end{tikzpicture}

\color{black}
\vspace{10pt}
\hfill Kelowna, Canada, \today{}

Dear Editors,

We are pleased to submit our manuscript entitled ``\emph{Rising temperatures alter how and where boreal mammals move}'' for consideration as a research article in Global Change Biology.

Climate change is rapidly reshaping ecosystems around the world. A growing body of evidence is demonstrating how widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness and community dynamics. Although research on the effects of temperature on mammalian movement behaviour is growing, most studies have focused on documenting changes in only a single species, and few have disentangled temperature effects from seasonal behaviour cycles. Consequently, it is still uncertain how mammals will adapt their movement behaviour throughout the 21\textsuperscript{st} century, nor what the community-level implications of such changes might be. More work is needed on quantfying how different species respond to temperature and the consequences for ecological communities.

Our study addresses this gap by modelling the effects of hourly, proximal temperature on the movement rates and habitat selection of 434 individuals across six mammal species throughout North-Western Canada, namely: boreal and southern mountain caribou, cougars, elk, grizzly bears, mountain goats, and wolves between 1998 and 2023. We employ the flexibility and interpretability of Generalized Additive Models to show that temperature has nonlinear and complex effects on mammalian movement behaviour, but that responses vary greatly across species. We then leverage the estimates to produce quantitative predictions of how climate change will affect each species' movement behaviour throughout the 21\textsuperscript{st} century. We show that, as temperatures rise, many individuals will begin selecting against their current ranges as they adapt to changing phenology and community dynamics. Finally, we present a case study using data on boreal caribou and wolves to show how rising temperatures (and concomitant changes in habitat selection) are likely to increase encounter rates between the two species due to changes in habitat selection, but that the changes are not uniform in space, and there are regions where interactions may decrease, versus others where they will increase. Consequently, rising temperatures are likely to impact wolves' predation of caribou and the effectiveness of wolf reduction strategies.

Our findings offer new insight into the effects of climate change on the movement behaviour of multiple boreal mammalian species, that suggest conservation efforts should account for future changes in movement behaviour as well as any cascading effects on biological communities. Anticipating changes in mammalian movement behaviour will become crucial for effectively and proactively understanding community-level responses and selecting high-quality habitat for long-term conservation. We believe the manuscript will be of strong interest to researchers working at the intersection of climate adaptation, biodiversity protection, and ecological forecasting. All code and data are openly available via GitHub (\url{https://github.com/QuantitativeEcologyLab/bc-mammals-temperature}) to support transparent and reproducible science.

We thank you for considering our submission and look forward to your response.

\vspace{10pt}
Kind regards,

Stefano Mezzini and Prof. Michael Noonan on behalf of all co-authors

\end{document}
